# Prerequisites

This chapter is just a quick review of [Matrix Theory] and [Probability Theory]  

If you feel you do not need to brush up on these theories, you can jump right into [Descriptive Statistics]

## Matrix Theory

\begin{equation}
\begin{split}
A=
\left[\begin{array}{c}
a_{11} & a_{12} \\
a_{21} & a_{22} \\
\end{array}
\right]
\end{split}
\end{equation}


\begin{equation}
\begin{split}
A' =
\left[\begin{array}{c}
a_{11} & a_{21} \\
a_{12} & a_{22} \\
\end{array}
\right]
\end{split}
\end{equation}


$$
\mathbf{(ABC)'=C'B'A'}
$$

<br>

\begin{equation}
\begin{split}
\mathbf{A} &= 
\left(\begin{array}{cccc} 
a_{11} & a_{12} & a_{13} \\ 
a_{21} & a_{22} & a_{23} \\ 
\end{array}\right)
\left(\begin{array}{c}
b_{11} & b_{12} & b_{13} \\
b_{21} & b_{22} & b_{23} \\
b_{31} & b_{32} & b_{33} \\
\end{array}\right) \\
&= 
\left(\begin{array}{c}
a_{11}b_{11}+a_{12}b_{21}+a_{13}b_{31} & \sum_{i=1}^{3}a_{1i}b_{i2} &  \sum_{i=1}^{3}a_{1i}b_{i3} \\
\sum_{i=1}^{3}a_{2i}b_{i1} & \sum_{i=1}^{3}a_{2i}b_{i2} & \sum_{i=1}^{3}a_{2i}b_{i3} \\
\end{array}\right) 
\end{split}
\end{equation}

Let $\mathbf{a}$ be a 3 x 1 vector, then the quadratic form is 
$$
\mathbf{a'Ba} = \sum_{i=1}^{3}\sum_{i=1}^{3}a_i b_{ij} a_{j}
$$

### Rank 

 * Dimension of space spanned by its columns (or its rows).  
 * Number of linearly indepdent columns/rows  
 
 For a n x k matrix **A** and k x k matrix **B**  
 
 * $rank(A)\leq min(n,k)$
 * $rank(A) = rank(A') = rank(A'A)=rank(AA')$  
 * $rank(AB)=min(rank(A),rank(B))$  
 * **B** is invertible if and only if rank(B) = k (non-singular)  

<br>

### Inverse
In scalar, a = 0 then 1/a does not exist.
In matrix, a matrix is invertible when it's a non-zero matrix. 

A non-singular square matrix A is invertible if there exists a non-singular square matrix B such that, 
$$AB=I$$
Then $A^{-1}=B$. For a 2x2 matrix, 

$$
A =
\left(\begin{array}{c}
a & b \\
c & d \\
\end{array}
\right)
$$

$$
A^{-1}=
\frac{1}{ad-bc}
\left(\begin{array}{c}
d & -b \\
-c & a \\
\end{array}
\right)
$$

For the partition matrix, 

\begin{equation}
\begin{split}
\left[\begin{array}{c}
A & B \\
C & D \\
\end{array}
\right]^{-1}
 =
\left[\begin{array}{c}
\mathbf{(A-BD^{-1}C)^{-1}} & \mathbf{-(A-BD^{-1}C)^{-1}BD^-1}\\
\mathbf{-DC(A-BD^{-1}C)^{-1}} & \mathbf{D^{-1}+D^{-1}C(A-BD^{-1}C)^{-1}BD^{-1}}\ \\
\end{array}
\right]
\end{split}
\end{equation}

<br>

Properties for a non-singular square matrix  

 * $\mathbf{A^{-1}}=A$  
 * for a non-zero scalar b, $\mathbf{(bA)^{-1}=b^{-1}A^{-1}}$  
 * for a matrix B, $\mathbf(BA)^{-1}=B^{-1}A^{-1}$ only if B is non-singular
 * $\mathbf{(A^{-1})'=(A')^{-1}}$  
 * Never notate $\mathbf{1/A}$

<br>

### Definiteness

A symmetric square k x k matrix, $\mathbf{A}$, is Positive Semi-Definite if for any non-zero k x 1 vector $\mathbf{x}$, 
$$\mathbf{x'Ax \geq 0 }$$

A symmetric square k x k matrix, $\mathbf{A}$, is Negative Semi-Definite if for any non-zero k x 1 vector $\mathbf{x}$
$$\mathbf{x'Ax \leq 0 }$$

$\mathbf{A}$ is indefinite if it is neither positive semi-definite or negative semi-definite. 

The identity matrix is positive definite

**Example**
Let $\mathbf{x} =(x_1 x_2)'$, then for a 2 x 2 identity matrix, 

\begin{equation}
\begin{split}
\mathbf{x'Ix} 
&= (x_1 x_2) 
\left(\begin{array}{c}
1 & 0 \\
0 & 1 \\
\end{array}
\right)
\left(\begin{array}{c}
x_1 \\
x_2 \\
\end{array}
\right) \\
&=
(x_1 x_2)
\left(\begin{array}{c}
x_1 \\
x_2 \\
\end{array}
\right) \\
&=
x_1^2 + x_2^2 >0
\end{split}
\end{equation}

Definiteness gives us the ability to compare matrices
$\mathbf{A-B}$ is PSD
This property also helps us show efficiency (which variance covariance matrix of one estimator is smaller than another)  

Properties  

 * any variance matrix is PSD  
 * a matrix $\mathbf{A}$ is PSD if and only if there exists a matrix $\mathbf{B}$ such that $\mathbf{A=B'B}$  
 * if $\mathbf{A}$ is PSD, then $\mathbf{B'AB}$ is PSD  
 * if A and C are non-singular, then A-C is PSD if and only if $\mathbf{C^{-1}-A^{-1}}$  
 * if A is PD (ND) then $A^{-1}$ is PD (ND)  


**Note**  

 * Indefinite A is neither PSD nor NSD. There is no comparable concept in scalar.
 * If a square matrix is PSD and invertible then it is PD
 

Example: 

 1. Invertible / Indefinite

$$
\left[
\begin{array}{a}
-1 & 0 \\
0 & 10 \\
\end{array}
\right]
$$

 2. Non-invertible/ Indefinite

$$
\left[
\begin{array}{a}
0 & 1 \\
0 & 0 \\
\end{array}
\right]
$$

 3. Invertible / PSD

$$
\left[
\begin{array}{a}
1 & 0 \\
0 & 1 \\
\end{array}
\right]
$$


 4. Non-Invertible / PSD

$$
\left[
\begin{array}{a}
0 & 0 \\
0 & 1 \\
\end{array}
\right]
$$


### Matrix Calculus
$y=f(x_1,x_2,...,x_k)=f(x)$
where x is a 1 x k row vector. The Gradient (first order derivative with respect to a vector) is, 

$$
\frac{\partial{f(x)}}{\partial{x}}=
\left(\begin{array}{c}
\frac{\partial{f(x)}}{\partial{x_1}} \\
\frac{\partial{f(x)}}{\partial{x_2}} \\
... \\
\frac{\partial{f(x)}}{\partial{x_k}}
\end{array}
\right)
$$

The **Hessian** (second order derivative with respect to a vector) is, 

$$
\frac{\partial^2{f(x)}}{\partial{x}\partial{x'}}=
\left(\begin{array}{c}
\frac{\partial^2{f(x)}}{\partial{x_1}\partial{x_1}} & \frac{\partial^2{f(x)}}{\partial{x_1}\partial{x_2}} & ... & \frac{\partial^2{f(x)}}{\partial{x_1}\partial{x_k}} \\
\frac{\partial^2{f(x)}}{\partial{x_1}\partial{x_2}} & \frac{\partial^2{f(x)}}{\partial{x_2}\partial{x_2}} & ... & \frac{\partial^2{f(x)}}{\partial{x_2}\partial{x_k}} \\
... & ...& & ...\\
\frac{\partial^2{f(x)}}{\partial{x_k}\partial{x_1}} & \frac{\partial^2{f(x)}}{\partial{x_k}\partial{x_2}} & ... & \frac{\partial^2{f(x)}}{\partial{x_k}\partial{x_k}}
\end{array}
\right)
$$

### Optimization

| |**Scalar Optimization**|**Vector Optimization**|
|---|:-:|:-:|
|First Order Condition|$$\frac{\partial{f(x_0)}}{\partial{x}}=0$$|$$\frac{\partial{f(x_0)}}{\partial{x}}=\left(\begin{array}{c}0 \\ .\\ .\\ .\\ 0\end{array}\right)$$|
|Second Order Condition \ **Convex** $\rightarrow$ **Min**|$$\frac{\partial^2{f(x_0)}}{\partial{x^2}} > 0$$|$$\frac{\partial^2{f(x_0)}}{\partial{xx'}}>0$$|
**Concave** $\rightarrow$ **Max** |$$\frac{\partial^2{f(x_0)}}{\partial{x^2}} < 0$$|$$\frac{\partial^2{f(x_0)}}{\partial{xx'}}<0$$|




## Probability Theory
### Axiom and Theorems of Probability

 1. Let S denote a sample space of an experiment P[S]=1
 2. $P[A] \ge 0$ for every event A
 3. Let $A_1,A_2,A_3,...$ be a finite or an infinite collection of mutually exclusive events. Then $P[A_1\cup A_2 \cup A_3 ...]=P[A_1]+P[A_2]+P[A_3]+...$
 4. $P[\emptyset]=0$
 5. $P[A']=1-P[A]$
 6. $P[A_1 \cup A_2] = P[A_1] + P[A_2] - P[A_1 \cap A_2]$

<br>

**Conditional Probability**
$$
P[A|B]=\frac{A \cap B}{P[B]}
$$

<br>

**Independent Events **
Two events A and B are independent if and only if:  

 1. $P[A\cap B]=P[A]P[B]$
 2. $P[A|B]=P[A]$
 3. $P[B|A]=P[B]$

A finite collection of events $A_1, A_2, ..., A_n$ is independent if and only if any subcollection is independent.

<br>

**Multiplication Rule**
$P[A \cap B] = P[A|B]P[B] = P[B|A]P[A]$

<br>

**Bayes' Theorem**
Let $A_1, A_2, ..., A_n$ be a collection of mutually exclusive events whose union is S.  
Let b be an event such that $P[B]\neq0$  
Then for any of the events $A_j$, j = 1,2,...,n
$$
P[A_|B]=\frac{P[B|A_j]P[A_j]}{\sum_{i=1}^{n}P[B|A_j]P[A_i]}
$$

**Jensen's Inequality**

 * If g(x) is convex $E(g(X)) \ge g(E(X))$
 * If g(x) is concave $E(g(X)) \le g(E(X))$



#### Law of Iterated Expectations
$E(Y)=E(E(Y|X))$

#### Correlation and Independence
**Independence**

 * $f(x,y)=f_X(x)f_Y(y)$
 * $f_{Y|X}(y|x)=f_Y(y)$ and $f_{X|Y}(x|y)=f_X(x)$
 * $E(g_1(X)g_2(Y))=E(g_1(X))E(g_2(Y))$

**Mean Independence** (implied by independence)

 * Y is mean independent of X if and only if $E(Y|X)=E(Y)$
 * $E(Xg(Y))=E(X)E(g(Y))$
 
**Uncorrelated** (implied by independence and mean independence)

 * $Cov(X,Y)=0$
 * $Var(X+Y)=Var(X) + Var(Y)$
 * $E(XY)=E(X)E(Y)$ 

$$
Strongest \\
\downarrow \\
Independence \\
\downarrow \\
Mean Independence \\
\downarrow \\
Uncorrelated \\ 
\downarrow \\
Weakest\\
$$

<br>

### Central Limit Theorem

Let $X_1, X_2,...,X_n$ be a random sample of size n from a distribution (not necessarily normal) X with mean $\mu$ and variance $\sigma^2$. then for large n ($n \ge 25$),  

 1. $\bar{X}$ is approximately normal with with mean $\mu_{\bar{X}}=\mu$ and variance $\sigma^2_{\bar{X}} = Var(\bar{X})= \frac{\sigma^2}{n}$  
 2. $\hat{p}$is approximately normal with $\mu_{\hat{p}} = p, \sigma^2_{\hat{p}} = \frac{p(1-p)}{n}$
 3. $\hat{p_1} - \hat{p_2}$ is approximately normal with $\mu_{\hat{p_1} - \hat{p_2}} = p_1 - p_2, \sigma^2_{\hat{p_1} - \hat{p_2}}=\frac{p_1(1-p)}{n_1} + \frac{p_2(1-p)}{n_2}$
 4. $\bar{X_1} - \bar{X_2}$ is approximately normal with $\mu_{\bar{X_1} - \bar{X_2}} = \mu_1 - \mu_2, \sigma^2_{\bar{X_1} - \bar{X_2}} = \frac{\sigma^2_1}{n_1}+\frac{\sigma^2_2}{n_2}$
 5. The following random variables are approximately standard normal:  
        + $\frac{\bar{X}-\mu}{\sigma/\sqrt{n}}$  
        + $\frac{\hat{p}-p}{\sqrt{\frac{p(1-p)}{n}}}$  
        + $\frac{(\hat{p_1}-\hat{p_2})-(p_1-p_2)}{\sqrt{\frac{p_1(1-p_1)}{n_1}-\frac{p_2(1-p_2)}{n_2}}}$  
        + $\frac{(\bar{X_1}-\bar{X_2})-(\mu_1-\mu_2)}{\sqrt{\frac{\sigma^2_1}{n_1}-\frac{\sigma^2_2}{n_2}}}$  



If $\{x_i\}_{i=1}^{n}$ is an iid random sample from a probability distribution with finite mean $\mu$ and finite variance $\sigma^2$ then the sample mean $\bar{x}=n^{-1}\sum_{i=1}^{n}x_i$ scaled by $\sqrt{n}$ has the following limiting distribution

$$
\sqrt{n}(\bar{x}-\mu) \to^d N(0,\sigma^2)
$$
or if we were to standardize the sample mean, 

$$
\frac{\sqrt{n}(\bar{x}-\mu)}{\sigma} \to^d N(0,1)
$$


 * holds for most random sample from any distribution (continuous, discrete, unknown).  
 * extends to multivariate case: random sample of a random vector converges to a multivariate normal.  
 * Variance from the limiting distribution is the asymptotic variance (Avar)  

$$
Avar(\sqrt{n}(\bar{x}-\mu)) = \sigma^2 \\
\lim_{n \to \infty} Var(\sqrt{n}(\bar{x}-\mu)) = \sigma^2 \\
Avar(.) \neq lim_{n \to \infty} Var(.)
$$

### Random variable

| | Discrete Variable | Continuous Variable|
|--- | --- | ---|
|**Definition**|A random variable is discrete if it can assume at most a finite or countably infinite number of possible values| A random variable is continuous if it can assume any value in some interval or intervals of real numbers and the probability that it assumes any specific value is 0|
|**Density Function**|A function f is called a density for X if: <br> (1) $f(x) \ge 0$ <br> (2) $\sum_{all~x}f(x)=1$ <br> (3) $f(x)=P(X=x)$ for x real |A function f is called a density for X if: <br> (1) $f(x) \ge 0$ for x real <br> (2) $\int_{-\infty}^{\infty} f(x) \; dx=1$ <br> (3) $P[a \le X \le] =\int_{a}^{b} f(x) \; dx$ for a and b real|
|**Cumulative Distribution Function** <br> for x real| $F(x)=P[X \le x]$|$F(x)=P[X \le x]=\int_{-\infty}^{\infty}f(t)dt$|
| $E[H(X)]$ | $\sum_{all ~x}H(x)f(x)$ | $\int_{-\infty}^{\infty}H(x)f(x)$ |
|$\mu=E[X]$ | $\sum_{all ~ x}xf(x)$ | $\int_{-\infty}^{\infty}xf(x)$|
|**Ordinary Moments** <br> the kth ordinary moment for variable X is defined as: $E[X^k]$| $\sum_{all ~ x \in X}(x^kf(x))$| $\int_{-\infty}^{\infty}(x^kf(x))$|
|**Moment generating function (mgf)** <br> $m_X(t)=E[e^{tX}]$| $\sum_{all ~ x \in X}(e^{tx}f(x))$| $\int_{-\infty}^{\infty}(e^{tx}f(x)dx)$|

<br>

Expected Value Properties:  

 * E[c] = c for any constant c  
 * E[cX] = cE[X] for any constant c  
 * E[X+Y] = E[X] = E[Y]  
 * E[XY] = E[X].E[Y] (if X and Y are independent)

Expected Variance Properties:  

 * $Var(c) = 0$ for any constant c  
 * $Var(cX) = c^2Var(X)$ for any constant c  
 * $Var(X) \ge 0$
 * $Var(X) = E(X^2) - (E(X))^2$
 * $Var(X+c)=Var(X)$
 * $Var (X+Y) = Var(X) + Var(Y)$ (if X and Y are independent)

Standard deviation 
$\sigma=\sqrt(\sigma^2)=\sqrt(Var X)$

Moment generating function properties:  

 (a) $\frac{d^k(m_X(t))}{dt^k}|_{t=0}=E[X^k]$  
 (b) $\mu=E[X]=m_X'(0)$  
 (c) $E[X^2]=m_X''(0)$  


**mgf Theorems**  

Let $X_1,X_2,...X_n,Y$ be random variables with moment-generating functions $m_{X_1}(t),m_{X_2}(t),...,m_{X_n}(t),m_{Y}(t)$  

 1. If $m_{X_1}(t)=m_{X_2}(t)$ for all t in some open interval about 0, then $X_1$ and $X_2$ have the same distribution
 2. If $Y = \alpha + \beta X_1$, then $m_{Y}(t)= e^{\alpha t}m_{X_1}(\beta t)$
 3. If $X_1,X_2,...X_n$ are independent and $Y = \alpha_0 + \alpha_1 X_1 + \alpha_2 X_2 + ... + \alpha_n X_n$ (where $\alpha_0, ... ,\alpha_n$ are real numbers), then $m_{Y}(t)=e^{\alpha_0 t}m_{X_1}(\alpha_1t)m_{X_2}(\alpha_2 t)...m_{X_n}(\alpha_nt)$  
 4. Suppose $X_1,X_2,...X_n$ are independent normal random variables with means $\mu_1,\mu_2,...\mu_n$ and variances $\sigma^2_1,\sigma^2_2,...,\sigma^2_n$. If $Y = \alpha_0 + \alpha_1 X_1 + \alpha_2 X_2 + ... + \alpha_n X_n$ (where $\alpha_0, ... ,\alpha_n$ are real numbers), then Y is normally distributed with mean $\mu_Y = \alpha_0 + \alpha_1 \mu_1 +\alpha_2 \mu_2 + ... + \alpha_n \mu_n$ and variance $\sigma^2_Y = \alpha_1^2 \sigma_1^2 + \alpha_2^2 \sigma_2^2 + ... + \alpha_n^2 \sigma_n^2$

### Moment
Moment | Uncentered | Centered 
---|---|---
1st | $E(X)=\mu=Mean(X)$| 
2nd | $E(X^2)$ | $E((X-\mu)^2)=Var(X)=\sigma^2$
3rd | $E(X^3)$ | $E((X-\mu)^3)$
4th | $E(X^4)$ | $E((X-\mu)^4)$

Skewness(X) = $E((X-\mu)^3)/\sigma^3$  

Kurtosis(X) = $E((X-\mu)^4)/\sigma^4$  


**Conditional Moments**
$$
E(Y|X=x)=
\begin{cases}
\sum_yyf_Y(y|x) & \text{for discrete RV}\\
\int_yyf_Y(y|x)dy & \text{for continous RV}\\
\end{cases}
$$



$$
Var(Y|X=x)=
\begin{cases}
\sum_y(y-E(Y|x))^2f_Y(y|x) & \text{for discrete RV}\\
\int_y(y-E(Y|x))^2f_Y(y|x)dy & \text{for continous RV}\\
\end{cases}
$$



#### Multivariate Moments

$$
\begin{equation}
E=
\left(
\begin{array}{c}
X \\
Y \\
\end{array}
\right)
=
\left(
\begin{array}{c}
E(X) \\
E(Y) \\
\end{array}
\right)
=
\left(
\begin{array}{c}
\mu_X \\
\mu_Y \\
\end{array}
\right)
\end{equation}
$$


$$
\begin{equation}
\begin{split}
Var
\left(
\begin{array}{c}
X \\
Y \\
\end{array}
\right)
&=
\left(
\begin{array}{c}
Var(X) & Cov(X,Y) \\
Cov(X,Y) & Var(Y) \\
\end{array}
\right) \\
&=
\left(
\begin{array}{c}
E((X-\mu_X)^2) & E((X-\mu_X)(Y-\mu_Y)) \\
E((X-\mu_X)(Y-\mu_Y)) & E((Y-\mu_Y)^2) \\
\end{array}
\right)
\end{split}
\end{equation}
$$
<br>

**Properties**

 * $E(aX + bY + c)=aE(X) +bE(Y) + c$
 * $Var(aX + bY + c) = a^2Var(X) + b^2Var(Y) + 2abCov(X,Y)$
 * $Cov(aX + bY, cX + bY) =acVar(X)+bdVar(Y) + (ad+bc)Cov(X,Y)$
 * Correlation: $\rho_{XY} = \frac{Cov(X,Y)}{\sigma_X\sigma_Y}$




### Distributions

**Conditional Distributions**
$$
f_{X|Y}(X|Y=y)=\frac{f(X,Y)}{f_Y(y)}
$$
$f_{X|Y}(X|Y=y)=f_X(X)$ if X and Y are independent


#### Discrete

CDF: Cumulative Density Function   
MGF: Moment Generating Function


##### Bernoulli
$Bernoulli(p)$

**PDF**

```{r}
hist(mc2d::rbern(100000, prob=.5))
```


##### Binomial
$B(n,p)$  

 * the experiment consists of a fixed number (n) of Bernoulli trials, each of which results in a success (s) or failure (f).  
 * The trials are identical and independent, and probability of success (p) and probability of failure (q = 1- p) remains the same for all trials. 
 * The random variable X denotes the number of successes obtained in the n trials. 

**Density**

$$
f(x)={{n}\choose{x}}p^xq^{n-x}
$$

**CDF**  
You have to use table   

**PDF**

```{r}
# Histogram of 100000 random values from a sample of 100 with probability of 0.5
hist(rbinom(100000, size = 100, prob = 0.5))
```


**MGF**
$$
m_X(t) =(q+pe^t)^n
$$
**Mean**
$$
\mu = E(x) = np
$$

**Variance**
$$
\sigma^2 =Var(X) = npq
$$

##### Poisson
$Pois(\lambda)$  

 * Arises with Poisson process, which involves observing discrete events in a continuous "interval" of time, length, or space. 
 * The random variable X is the number of occurrences of the event within an interval of s units
 * The parameter $\lambda$ is the average number of occurrences of the event in question per measurement unit. For the distribution, we use the parameter $k=\lambda s$

**Density**
$$
f(x) = \frac{e^{-k}k^x}{x!}
$$
,k > 0, x =0,1,...

**CDF**  
Use table  

**PDF**  

```{r}
# Poisson dist with mean of 5 or Poisson(5)
hist(rpois(10000, lambda = 5))
```


**MGF**
$$
m_X(t)=e^{k(e^t-1)}
$$ 

**Mean**
$$
\mu = E(X) = k
$$

**Variance**
$$
\sigma^2 = Var(X) = k
$$

##### Geometric

 * The experiment consists of a series of trails. The outcome of each trial can be classed as being either a "success" (s) or "failure" (f). (This is called a Bernoulli trial).
 * The trials are identical and independent in the sense that the outcome of one trial has no effect on the outcome of any other. The probability of success (p) and probability of failure (q=1-p) remains the same from trial to trial. 
 * lack of memory
 * X: the number of trials needed to obtain the first success.

**Density**

$$
f(x)=pq^{x-1}
$$

**CDF**  
$$
F(x) = 1- q^x
$$

**PDF**  

```{r}
# hist of Geometric distribution with probability of success = 0.5
hist(rgeom(n = 10000, prob = 0.5))
```


**MGF** 
$$
m_X(t) = \frac{pe^t}{1-qe^t}
$$
for $t < -ln(q)$

**Mean**
$$
\mu = \frac{1}{p}
$$

**Variance**
$$
\sigma^2 = Var(X) = \frac{q}{p^2}
$$

##### Hypergeometric

 * The experiment consists of drawing a random sample of size n without replacement and without regard to order from a collection of N objects. 
 * Of the N objects, r have a trait of interest; N-r do not have the trait 
 * X is the number of objects in the sample with the trait.

**Density**
$$
f(x)=\frac{{{r}\choose{x}}{{N-r}\choose{n-x}}}{{{N}\choose{n}}}
$$ 
where $max[0,n-(N-r)] \le x \le min(n,r)$  

**PDF**  

```{r}
# hist of hypergeometric distribution with the number of white balls = 50, and the number of black balls = 20, and number of balls drawn = 30. 
hist(rhyper(nn = 10000 , m=50, n=20, k=30))
```


**Mean**
$$
\mu = E(x)= \frac{nr}{N}
$$

**Variance**
$$
\sigma^2 = var(X) = n (\frac{r}{N})(\frac{N-r}{N})(\frac{N-n}{N-1})
$$

**Note**
 For large N (if $\frac{n}{N} \le 0.05$), this distribution can be approximated using a Binomial distribution with $p = \frac{r}{N}$

##### 

##### 
##### 

##### 

##### 

##### 

##### 







<br>



#### Continuous

##### Uniform

 * Defined over an interval (a,b) in which the probabilities are "equally likely" for subintervals of equal length. 

**Density**
$$
f(x)=\frac{1}{b-a}
$$ 
for a < x < b

**CDF**

\begin{cases}
0 & \text{if x <a } \\
\frac{x-a}{b-a} & \text{if $a \le x \le b$ }\\
1 & \text{if x >b}\\
\end{cases}  

**PDF**  

```{r}
hist(runif(100000, min = 0, max = 1))
```


**MGF**


\begin{cases}
\frac{e^{tb} - e^{ta}}{t(b-a)}&\text{ if $t \neq 0$}\\
1&\text{if $ t \neq 0$}\\
\end{cases}

**Mean**

$$
\mu = E(X) = \frac{a +b}{2}
$$

**Variance**
$$
\sigma^2 = Var(X) = \frac{(b-a)^2}{12}
$$


##### Gamma 

 * is used to define the exponential and chi-squared distributions
 * The gamma function is defined as:

$$
\Gamma(\alpha) = \int_0^{\infty} z^{\alpha-1}e^{-z}dz
$$
where $\alpha > 0$

 * Properties of The Gamma function:  
        + $\Gamma(1) = 1$
        + For $\alpha >1$, $\Gamma(\alpha)=(\alpha-1)\Gamma(\alpha-1)$
        + If n is an integer and $n>1$, then $\Gamma(n) = (n-1)!$

**Density**
$$
f(x) = \frac{1}{\Gamma(\alpha)\beta^{\alpha}}x^{\alpha-1}e^{-x/\beta}
$$

**CDF**
$$
F(x,n,\beta) = 1 -\sum_{k=0}^{n-1} \frac{(\frac{x}{\beta})^k e^{-x/\beta}}{k!}
$$
for x>0, and $\alpha = n$ (a positive integer)  

**PDF**  

```{r}
hist(rgamma(n = 10000, shape = 5, rate = 1))
```


**MGF**
$$
m_X(t) = (1-\beta t)^{-\alpha}
$$
where $t < \frac{1}{\beta}$

**Mean**
$$
\mu = E(X) = \alpha \beta
$$

**Variance**
$$
\sigma^2 = Var(X) = \alpha \beta^2
$$


##### Normal

$N(\mu,\sigma^2)$  

 * is symmetric, bell-shaped curve with parameters $\mu$ and $\sigma^2$
 * also known as Gaussian.

**Density**
$$
f(x) = \frac{1}{\sqrt{2\pi \sigma}}e^{-\frac{1}{2}(\frac{x-\mu}{\sigma})^2}
$$
for $-\infty < x, \mu< \infty, \sigma > 0$

**CDF**  

Use table 

**PDF**  

```{r}
hist(rnorm(1000000, mean = 0, sd = 1))
```


**MGF**
$$
m_X(t) = e^{\mu t + \frac{\sigma^2 t^2}{2}}
$$

**Mean**
$$
\mu = E(X)
$$

**Variance**
$$
\sigma^2 = Var(X)
$$

**Standard Normal Random Variable**  

 * The normal random variable Z with mean $\mu = 0$ and standard deviation $\sigma =1$ is called standard normal
 * Any normal random variable X with mean $\mu$ and standard deviation $\sigma$ can be converted to the standard normal random variable $Z = \frac{X-\mu}{\sigma}$

**Normal Approximation to the Binomial Distribution**  

 Let X be binomial with parameters n and p. For large n (so that $(A)p \le .5$ and $np > 5$ or (B) $p>.5$ and $nq>5$), X is approximately normally distributed with mean $\mu = np$ and standard deviation $\sigma = \sqrt{npq}$

When using the normal approximation, add or subtract 0.5 as needed for the continuity correction 

Discrete | Approximate Normal (corrected)
:-:|:-:
P(X = c) | P(c -0.5 < Y < c + 0.5)
P(X < c) |P(Y < c - 0.5)
P(X \le c) | P(Y < c + 0.5)
P(X > c) | P(Y > c + 0.5)
P(X \ge c) | P(Y > c - 0.5)


**Normal Probability Rule**

If X is normally distributed with parameters $\mu$ and $\sigma$, then  
 * $P(-\sigma < X - \mu < \sigma) \approx .68$
 * $P(-2\sigma < X - \mu < 2\sigma) \approx .95$
 * $P(-3\sigma < X - \mu < 3\sigma) \approx .997$


##### Logistic
$Logistic(\mu,s)$

**PDF**  

```{r}
hist(rlogis(n = 100000, location = 0, scale = 1))
```


##### Lognomral
$lognormal(\mu,\sigma^2)$

**PDF**  

```{r}
hist(rlnorm(n = 10000, meanlog = 0, sdlog = 1))
```


##### Exponential 
$Exp(\lambda)$

 * A special case of the gamma distribution with $\alpha = 1$
 * Lack of memory 
 * $\lambda$ = rate
 Within a Poisson process with parameter $\lambda$, if W is the waiting tine until the occurrence of the first event, then W has an exponential distribution with $\beta = 1/\alpha$

**Density**
$$
f(x) = \frac{1}{\beta} e^{-x/\beta}
$$
for $x,\beta > 0$

**CDF**

\begin{equation}
F(x) = 
\begin{cases}
0 & \text{ if $x \le 0$}\\
1 - e^{-x/\beta} & \text{if $x > 0$}\\
\end{cases}
\end{equation}

**PDF**  

```{r}
hist(rexp(n = 100000, rate = 1))
```



**MGF**
$$
m_X(t) = (1-\beta t)^{-1}
$$
for $t < 1/\beta$

**Mean**
$$
\mu = E(X) = \beta
$$

**Variance**
$$
\sigma^2 = Var(X) =\beta^2
$$

##### Chi-squared
$\chi^2=\chi^2(k)$

 * A special case of the gamma distribution with $\beta =2$, and $\alpha = \gamma /2$ for a positive integer $\gamma$
 * The random variable X is denoted $\chi_{\gamma}^2$ and is said to have a chi-squared distribution with $\gamma$ degrees of freedom.

**Density**
Use density for Gamma Distribution with $\beta = 2$ and $\alpha = \gamma/2$

**CDF**
Use table   

**PDF**  

```{r}
hist(rchisq(n = 10000, df=2, ncp = 0))
```


**MGF**
$$
m_X(t) = (1-2t)^{-\gamma/2}
$$

**Mean**
$$
\mu = E(X) = \gamma
$$

**Variance**
$$
\sigma^2 = Var(X) = 2\gamma
$$



##### Student T
$T(v)$

 * $T=\frac{Z}{\sqrt{\chi_{\gamma}^2/\gamma}}$, where Z is standard normal follows a student-t distribution with $\gamma$ dof 
 * The distribution is symmetric, bell-shaped , with a mean of $\mu=0$  


```{r}
hist(rt(n = 100000, df=2, ncp =1))
```



##### F-Distribution
$F(d_1,d_2)$

 * F distribution is strictly positive
 * $F=\frac{\chi_{\gamma_1}^2/\gamma_1}{\chi_{\gamma_2^2}/\gamma_2}$ follows an F distribution with dof $\gamma_1$ and $\gamma_2$, where $\chi_{\gamma_1}^2$ and $\chi_{\gamma_2}^2$ are independent chi-squared random variables. 
 * The distribution is asymmetric and never negative.


**PDF**  

```{r}
hist(rf(n = 100000, df1=2, df2=3, ncp=1))
```


##### Cauchy

[Central Limit Theorem] and [Weak Law] do not apply to Cauchy because it does not have finite mean and finite variance  

**PDF**  

```{r}
hist(rcauchy(n = 100000, location = 0, scale = 1))
```


##### 

##### 

##### 

##### 

## General Math 

**Chebyshev's Inequality**
Let X be a random variable with mean $\mu$ and standard deviation $\sigma$. Then for any postiive number k: 

$$
P(|X-\mu| < k\sigma) \ge 1 - \frac{1}{k^2}
$$
Chebyshev's Inequality does not require that X be normally distributed


**Maclaurin series expansion for **
$$
e^z = 1 + z + \frac{z^2}{2!} + \frac{z^3}{3!} + ...
$$

<br>

**Geometric series: **
$$
s_n=\sum_{k=1}^{n}ar^{n-1}=\frac{a(1-r^n)}{1-r}
$$
if |r| < 1
$$
s=\sum_{k=1}^{\infty}ar^{n-1}=\frac{a}{1-r}
$$  

### Law of large numbers

Let $X_1,X_2,...$ be an infinite sequence of independent and identically distributed (i.i.d) 
Then, the sample average is

$$
\bar{X}_n =\frac{1}{n} (X_1 + ... + X_n)
$$
converges to the expected value ($\bar{X}_n \rightarrow \mu$) as $n \rightarrow \infty$

$$
Var(X_i) = Var(\frac{1}{n}(X_1 + ... + X_n)) = \frac{1}{n^2}Var(X_1 + ... + X_n)= \frac{n\sigma^2}{n^2}=\frac{\sigma^2}{n}
$$

The difference between [Weak Law] and [Strong Law] regards the mode of convergence 

#### Weak Law

The sample average converges in probability towards the expected value 

$$
\bar{X}_n \rightarrow^{p} \mu
$$
when $n \rightarrow \infty$

$$
\lim_{n\to \infty}P(|\bar{X}_n - \mu| > \epsilon) = 0
$$

The sample mean from a iid random sample ($\{ x_i \}_{i=1}^n$) from any population with a finite mean and finite variance $\sigma^2$ is ca consistent estimation for the population mean $\mu$

$$
plim(\bar{x})=plim(n^{-1}\sum_{i=1}^{n}x_i) =\mu
$$


#### Strong Law

The sample average converges almost surely to the expected value

$$
\bar{X}_n \rightarrow^{a.s} \mu 
$$
when $n \rightarrow \infty$

Equivalently,

$$
P(\lim_{n\to \infty}\bar{X}_n =\mu) =1
$$



### Law of Iterated Expectation
Let X, Y be random variables. Then, 

$$
E(X) = E(E(X|Y))
$$
means that the expected value of X can be calculated from the probability distribution of X|Y and Y



### Convergence
#### Convergence in Probability

 * $n \rightarrow \infty$, an estimator (random variable) that is close to the true value.
 * The random variable $\theta_n$ converges in probability to a constant c if 

$$
\lim_{n\to \infty}P(|\theta_n - c| \ge \epsilon) = 0
$$
for any positive $\epsilon$

Notation
$$
plim(\theta_n)=c 
$$
Equivalently, 
$$
\theta_n \rightarrow^p c
$$

**Properties of Convergence in Probability**

 * Slutsky's Theorem: for a continuous function g(.), if $plim(\theta_n)= \theta$ then $plim(g(\theta_n)) = g(\theta)$
 * if $\gamma_n \rightarrow^p \gamma$ then  
        + $plim(\theta_n + \gamma_n)=\theta + \gamma$
        + $plim(\theta_n \gamma_n) = \theta \gamma$
        + $plim(\theta_n/\gamma_n) = \theta/\gamma$ if $\gamma \neq 0$
 * Also hold for random vectors/ matrices

#### Convergence in Distribution

 * As $n \rightarrow \infty$, the distribution of a random variable may converge towards another ("fixed") distribution.
 * The random variable $X_n$ with CDF $F_n(x)$ converges in distribution to a random variable X with CDF $F(X)$ if 

$$
\lim_{n\to \infty}|F_n(x) - F(x)| = 0
$$
at all points of continuity of $F(X)$

Notation
F(x) is the limiting distribution of $X_n$ or $X_n \rightarrow^d X$

 * E(X) is the limiting mean (asymptotic mean)
 * Var(X) is the limiting variance (asymptotic variance)

**Note**
$$
E(X) \neq \lim_{n\to \infty}E(X_n) \\
Avar(X_n) \neq \lim_{n\to \infty}Var(X_n)
$$

**Properties of Convergence in Distribution**

 * Continuous Mapping Theorem: for a continuous function g(.), if $X_n \to^{d} g(X)$ then $g(X_n) \to^{d} g(X)$ 
 * if $Y_n\to^{d} c$, then  
 
        + $X_n + Y_n \to^{d} X + c$
        + $Y_nX_n \to^{d} cX$
        + $X_nY_n \to^{d} X/c$ if $c \neq 0$
 * also hold for random vectors/matrices

#### Summary
Properties of Convergence

Probability | Distribution
---|---
Slutsky's Theorem: for a continuous function g(.), if $plim(\theta_n)= \theta$ then $plim(g(\theta_n)) = g(\theta)$ | Continuous Mapping Theorem: for a continuous function g(.), if $X_n \to^{d} g(X)$ then $g(X_n) \to^{d} g(X)$ 
if $\gamma_n \rightarrow^p \gamma$ then |if $Y_n\to^{d} c$, then
$plim(\theta_n + \gamma_n)=\theta + \gamma$ | $X_n + Y_n \to^{d} X + c$
$plim(\theta_n \gamma_n) = \theta \gamma$ | $Y_nX_n \to^{d} cX$
$plim(\theta_n/\gamma_n) = \theta/\gamma$ if $\gamma \neq 0$ | $X_nY_n \to^{d} X/c$ if $c \neq 0$

[Convergence in Probability] is stronger than [Convergence in Distribution]. However, [Convergence in Distribution] does not guarantee [[Convergence in Probability]]



### Sufficient Statistics

**Likelihood**  

 * describes the extent to which the sample provides support for any particular parameter value.
 * Higher support corresponds to a higher value for the likelihood
 * The exact value of any likelihood is **meaningless**, 
 * The relative value, (i.e., comparing two values of $\theta$), is **informative**.

$$
L(\theta_0; y) = P(Y = y | \theta = \theta_0) = f_Y(y;\theta_0)
$$

**Likelihood Ratio**

$$
\frac{L(\theta_0;y)}{L(\theta_1;y)}
$$


**Likelihood Function**

For a given sample, you can create likelihoods for all possible values of $\theta$, which is called *likelihood function*

$$
L(\theta) = L(\theta; y) = f_Y(y;\theta)
$$

In a sample of size n, the likelihood function takes the form of a product

$$
L(\theta) = \prod_{i=1}^{n}f_i (y_i;\theta)
$$

Equivalently, the log likelihood function

$$
l(\theta) = \sum_{i=1}^{n} logf_i(y_i;\theta)
$$

**Sufficient statistics**

 * A statistic, T(y), is any quantity that can be calculated purely from  a sample (independent of $\theta$)
 * A statistic is **sufficient** if it conveys all the available information about the parameter. 

$$
L(\theta; y) = c(y)L^*(\theta;T(y))
$$


**Nuisance parameters**
If we are interested in a parameter (e.g., mean). Other parameters requiring estimation (e.g., standard deviation) are **nuisance** parameters. We can replace nuisance parameters in likelihood function with their estimates to create a **profile likelihood*. 



### Parameter transformations

log-odds transformation

$$
Log odds = g(\theta)= ln[\frac{\theta}{1-\theta}]
$$

log transformation

$$

$$





## Methods

Trade-off between parametric and non-parametric


