### Multiple Linear Regression {#sec-multiple-linear-regression}

The geometry of least squares regression involves projecting the response vector $\mathbf{y}$ onto the space spanned by the columns of the design matrix $\mathbf{X}$. The fitted values $\mathbf{\hat{y}}$ can be expressed as:

$$
\begin{aligned}
\mathbf{\hat{y}} &= \mathbf{Xb} \\
&= \mathbf{X(X'X)^{-1}X'y} \\
&= \mathbf{Hy},
\end{aligned}
$$

where:

-   $\mathbf{H} = \mathbf{X(X'X)^{-1}X'}$ is the projection operator (sometimes denoted as $\mathbf{P}$).
-   $\mathbf{\hat{y}}$ is the projection of $\mathbf{y}$ onto the linear space spanned by the columns of $\mathbf{X}$ (the model space).

The dimension of the model space is equal to the rank of $\mathbf{X}$ (i.e., the number of linearly independent columns in $\mathbf{X}$).

------------------------------------------------------------------------

Properties of the Projection Matrix $\mathbf{H}$

1.  Symmetry:
    -   The projection matrix $\mathbf{H}$ is symmetric: $$
        \mathbf{H} = \mathbf{H}'.
        $$
2.  Idempotence:
    -   Applying $\mathbf{H}$ twice gives the same result: $$
        \mathbf{HH} = \mathbf{H}.
        $$ Proof: $$
        \begin{aligned}
        \mathbf{HH} &= \mathbf{X(X'X)^{-1}X'X(X'X)^{-1}X'} \\
        &= \mathbf{X(X'X)^{-1}IX'} \\
        &= \mathbf{X(X'X)^{-1}X'} \\
        &= \mathbf{H}.
        \end{aligned}
        $$
3.  Dimensionality:
    -   $\mathbf{H}$ is an $n \times n$ matrix (where $n$ is the number of observations).
    -   The rank of $\mathbf{H}$ is equal to the rank of $\mathbf{X}$, which is typically the number of predictors (including the intercept).
4.  Orthogonal Complement:
    -   The matrix $\mathbf{(I - H)}$, where: $$
        \mathbf{I - H} = \mathbf{I - X(X'X)^{-1}X'},
        $$ is also a projection operator.
    -   It projects onto the orthogonal complement of the space spanned by the columns of $\mathbf{X}$ (i.e., the space orthogonal to the model space).
5.  Orthogonality of Projections:
    -   $\mathbf{H}$ and $\mathbf{(I - H)}$ are orthogonal: $$
        \mathbf{H(I - H)} = \mathbf{0}.
        $$
    -   Similarly: $$
        \mathbf{(I - H)H} = \mathbf{0}.
        $$

------------------------------------------------------------------------

Intuition for $\mathbf{H}$ and $\mathbf{(I - H)}$

-   $\mathbf{H}$: Projects $\mathbf{y}$ onto the model space, giving the fitted values $\mathbf{\hat{y}}$.
-   $\mathbf{I - H}$: Projects $\mathbf{y}$ onto the residual space, giving the residuals $\mathbf{e}$: $$
    \mathbf{e} = \mathbf{(I - H)y}.
    $$
-   $\mathbf{H}$ and $\mathbf{(I - H)}$ divide the response vector $\mathbf{y}$ into two components: $$
    \mathbf{y} = \mathbf{\hat{y}} + \mathbf{e}.
    $$
    -   $\mathbf{\hat{y}} = \mathbf{Hy}$ (fitted values).
    -   $\mathbf{e} = \mathbf{(I - H)y}$ (residuals).
-   The properties of $\mathbf{H}$ (symmetry, idempotence, dimensionality) reflect its role as a linear transformation that projects vectors onto the model space.

This geometric perspective provides insight into the mechanics of least squares regression, particularly how the response variable $\mathbf{y}$ is decomposed into fitted values and residuals.

------------------------------------------------------------------------

Similar to simple regression, the total sum of squares in multiple regression analysis can be partitioned into components corresponding to the regression (model fit) and the residuals (errors).

The uncorrected total sum of squares is:

$$
\mathbf{y'y} = \mathbf{\hat{y}'\hat{y} + e'e},
$$

where:

-   $\mathbf{\hat{y} = Hy}$ (fitted values, projected onto the model space).

-   $\mathbf{e = (I - H)y}$ (residuals, projected onto the orthogonal complement of the model space).

Expanding this using projection matrices:

$$
\begin{aligned}
\mathbf{y'y} &= \mathbf{(Hy)'(Hy) + ((I-H)y)'((I-H)y)} \\
&= \mathbf{y'H'Hy + y'(I-H)'(I-H)y} \\
&= \mathbf{y'Hy + y'(I-H)y}.
\end{aligned}
$$

This equation shows the partition of $\mathbf{y'y}$ into components explained by the model ($\mathbf{\hat{y}}$) and the unexplained variation (residuals).

For the corrected total sum of squares, we adjust for the mean (using the projection matrix $\mathbf{H_1}$):

$$
\mathbf{y'(I-H_1)y = y'(H-H_1)y + y'(I-H)y}.
$$

Here:

-   $\mathbf{H_1} = \frac{1}{n} \mathbf{J}$, where $\mathbf{J}$ is an $n \times n$ matrix of ones.

-   $\mathbf{H - H_1}$ projects onto the subspace explained by the predictors after centering.

+---------------------+--------------------------------------------------------------------+--------------------------------------------------------------------------------+
| Aspect              | Uncorrected Total Sum of Squares ($\mathbf{y'y}$)                  | Corrected Total Sum of Squares ($\mathbf{y'(I-H_1)y}$)                         |
+=====================+====================================================================+================================================================================+
| Definition          | Total variation in $y$ relative to the origin.                     | Total variation in $y$ relative to its mean (centered data).                   |
+---------------------+--------------------------------------------------------------------+--------------------------------------------------------------------------------+
| Adjustment          | No adjustment for the mean of $y$.                                 | Adjusts for the mean of $y$ by centering it.                                   |
+---------------------+--------------------------------------------------------------------+--------------------------------------------------------------------------------+
| Equation            | $\mathbf{y'y} = \mathbf{\hat{y}'\hat{y}} + \mathbf{e'e}$           | $\mathbf{y'(I-H_1)y} = \mathbf{y'(H-H_1)y} + \mathbf{y'(I-H)y}$                |
+---------------------+--------------------------------------------------------------------+--------------------------------------------------------------------------------+
| Projection Matrices | $\mathbf{H}$: Projects onto model space.                           | $\mathbf{H_1} = \frac{1}{n} \mathbf{J}$: Adjusts for the mean.                 |
|                     |                                                                    |                                                                                |
|                     | $\mathbf{I-H}$: Projects onto residuals.                           | $\mathbf{H-H_1}$: Projects onto predictors after centering.                    |
|                     |                                                                    |                                                                                |
|                     |                                                                    | $\mathbf{I-H}$: Projects onto residuals.                                       |
+---------------------+--------------------------------------------------------------------+--------------------------------------------------------------------------------+
| Interpretation      | Includes variation due to the mean of $y$.                         | Focuses on variation in $y$ around its mean.                                   |
+---------------------+--------------------------------------------------------------------+--------------------------------------------------------------------------------+
| Usage               | Suitable for raw, uncentered data.                                 | Common in regression and ANOVA to isolate variability explained by predictors. |
+---------------------+--------------------------------------------------------------------+--------------------------------------------------------------------------------+
| Application         | Measures total variability in $y$, including overall level (mean). | Measures variability explained by predictors relative to the mean.             |
+---------------------+--------------------------------------------------------------------+--------------------------------------------------------------------------------+

: Comparison of Uncorrected and Corrected Total Sum of Squares in Linear Regression

Why the Correction Matters

-   In ANOVA and regression, removing the contribution of the mean helps isolate the variability explained by predictors from the overall level of the response variable.

-   Corrected sums of squares are more common when comparing models or computing $R^2$, which requires centering to ensure consistency in proportionate variance explained.

The corrected total sum of squares can be decomposed into the sum of squares for regression (SSR) and the sum of squares for error (SSE):

| Source     | SS                                      | df      | MS                  | F           |
|------------|-----------------------------------------|---------|---------------------|-------------|
| Regression | $SSR = \mathbf{y'(H - \frac{1}{n} J)y}$ | $p - 1$ | $MSR = SSR / (p-1)$ | $MSR / MSE$ |
| Error      | $SSE = \mathbf{y'(I - H)y}$             | $n - p$ | $MSE = SSE / (n-p)$ |             |
| Total      | $SST = \mathbf{y'(I - H_1)y}$           | $n - 1$ |                     |             |

: ANOVA Decomposition of the Corrected Total Sum of Squares

Where:

-   $p$: Number of parameters (including intercept).

-   $n$: Number of observations.

------------------------------------------------------------------------

Alternatively, the regression model can be expressed as:

$$
\mathbf{Y = X\hat{\beta} + (Y - X\hat{\beta})},
$$

where:

-   $\mathbf{\hat{Y} = X\hat{\beta}}$: Vector of fitted values (in the subspace spanned by $\mathbf{X}$).

-   $\mathbf{e = Y - X\hat{\beta}}$: Vector of residuals (in the orthogonal complement of the subspace spanned by $\mathbf{X}$).

-   $\mathbf{Y}$ is an $n \times 1$ vector in the $n$-dimensional space $\mathbb{R}^n$.

-   $\mathbf{X}$ is an $n \times p$ full-rank matrix, with its columns generating a $p$-dimensional subspace of $\mathbb{R}^n$. Hence, any estimator $\mathbf{X\hat{\beta}}$ is also in this subspace.

In linear regression, the Ordinary Least Squares estimator $\hat{\beta}$ minimizes the squared Euclidean distance $\|\mathbf{Y} - \mathbf{X}\beta\|^2$ between the observed response vector $\mathbf{Y}$ and the fitted values $\mathbf{X}\beta$. This minimization corresponds to the orthogonal projection of $\mathbf{Y}$ onto the column space of $\mathbf{X}$.

We solve the optimization problem:

$$
\min_{\beta} \|\mathbf{Y} - \mathbf{X}\beta\|^2.
$$

The objective function can be expanded as:

$$
\|\mathbf{Y} - \mathbf{X}\beta\|^2 
= (\mathbf{Y} - \mathbf{X}\beta)^\top (\mathbf{Y} - \mathbf{X}\beta).
$$

Perform the multiplication:

$$
\begin{aligned}
(\mathbf{Y} - \mathbf{X}\beta)^\top (\mathbf{Y} - \mathbf{X}\beta) 
&= \mathbf{Y}^\top \mathbf{Y} 
- \mathbf{Y}^\top \mathbf{X}\beta 
- \beta^\top \mathbf{X}^\top \mathbf{Y} 
+ \beta^\top (\mathbf{X}^\top \mathbf{X}) \beta.
\end{aligned}
$$

Since $\mathbf{Y}^\top \mathbf{X}\beta$ is a scalar, it equals $\beta^\top \mathbf{X}^\top \mathbf{Y}$. Therefore, the expanded expression becomes:

$$
\|\mathbf{Y} - \mathbf{X}\beta\|^2 
= \mathbf{Y}^\top \mathbf{Y} 
- 2\beta^\top \mathbf{X}^\top \mathbf{Y} 
+ \beta^\top (\mathbf{X}^\top \mathbf{X}) \beta.
$$

To find the $\beta$ that minimizes this expression, take the derivative with respect to $\beta$ and set it to 0:

$$
\frac{\partial}{\partial \beta} 
\left[
  \mathbf{Y}^\top \mathbf{Y} 
  - 2\beta^\top \mathbf{X}^\top \mathbf{Y} 
  + \beta^\top (\mathbf{X}^\top \mathbf{X}) \beta
\right]
= 0
$$

Computing the gradient:

$$
\frac{\partial}{\partial \beta} = -2\mathbf{X}^\top \mathbf{Y} + 2(\mathbf{X}^\top \mathbf{X})\beta.
$$

Setting this to zero:

$$
-2\mathbf{X}^\top \mathbf{Y} + 2\mathbf{X}^\top \mathbf{X}\beta = 0.
$$

Simplify:

$$
\mathbf{X}^\top \mathbf{X}\beta = \mathbf{X}^\top \mathbf{Y}.
$$

If $\mathbf{X}^\top \mathbf{X}$ is invertible, the solution is:

$$
\hat{\beta} = (\mathbf{X}^\top \mathbf{X})^{-1} \mathbf{X}^\top \mathbf{Y}.
$$

------------------------------------------------------------------------

Orthogonal Projection Interpretation

The fitted values are:

$$
\hat{\mathbf{Y}} = \mathbf{X}\hat{\beta}.
$$

From the normal equations, $\mathbf{X}^\top(\mathbf{Y} - \mathbf{X}\hat{\beta}) = 0$, which implies that the residual vector $\mathbf{Y} - \hat{\mathbf{Y}}$ is orthogonal to every column of $\mathbf{X}$. Therefore:

-   $\hat{\mathbf{Y}} = \mathbf{X}\hat{\beta}$ is the orthogonal projection of $\mathbf{Y}$ onto $\mathrm{Col}(\mathbf{X})$.
-   $\mathbf{Y} - \mathbf{X}\hat{\beta}$ lies in the orthogonal complement of $\mathrm{Col}(\mathbf{X})$.

------------------------------------------------------------------------

Pythagoras Decomposition

The geometric interpretation gives us the decomposition:

$$
\mathbf{Y} = \mathbf{X}\hat{\beta} + (\mathbf{Y} - \mathbf{X}\hat{\beta}),
$$

where:

-   $\mathbf{X}\hat{\beta}$ is the projection of $\mathbf{Y}$ onto the column space of $\mathbf{X}$.

-   $(\mathbf{Y} - \mathbf{X}\hat{\beta})$ is the residual vector, orthogonal to $\mathbf{X}\hat{\beta}$.

Since the two components are orthogonal, their squared norms satisfy:

$$
\begin{aligned}\|\mathbf{Y}\|^2 &= \mathbf{Y}^\top \mathbf{Y} && \text{(definition of norm squared)} \\
&= (\mathbf{Y} - \mathbf{X}\hat{\beta} + \mathbf{X}\hat{\beta})^\top(\mathbf{Y} - \mathbf{X}\hat{\beta} + \mathbf{X}\hat{\beta}) && \text{(add and subtract } \mathbf{X}\hat{\beta} \text{)} \\
&= (\mathbf{Y} - \mathbf{X}\hat{\beta})^\top(\mathbf{Y} - \mathbf{X}\hat{\beta}) + 2(\mathbf{X}\hat{\beta})^\top(\mathbf{Y} - \mathbf{X}\hat{\beta}) + (\mathbf{X}\hat{\beta})^\top(\mathbf{X}\hat{\beta}) && \text{(expand } (a+b)^\top(a+b) \text{)} \\
&= \|\mathbf{Y} - \mathbf{X}\hat{\beta}\|^2 + 2(\mathbf{X}\hat{\beta})^\top(\mathbf{Y} - \mathbf{X}\hat{\beta}) + \|\mathbf{X}\hat{\beta}\|^2 && \text{(rewrite each quadratic form as a norm)} \\
&= \|\mathbf{Y} - \mathbf{X}\hat{\beta}\|^2 + \|\mathbf{X}\hat{\beta}\|^2 && \text{(orthogonality: } (\mathbf{X}\hat{\beta})^\top(\mathbf{Y}-\mathbf{X}\hat{\beta}) = 0\text{)} \\
&= \|\mathbf{X}\hat{\beta}\|^2 + \|\mathbf{Y} - \mathbf{X}\hat{\beta}\|^2.\end{aligned}
$$

where the norm of a vector $\mathbf{a}$ in $\mathbb{R}^p$ is defined as:

$$
\|\mathbf{a}\| = \sqrt{\mathbf{a}^\top \mathbf{a}} = \sqrt{\sum_{i=1}^p a_i^2}.
$$

We are saying that $\mathbf{Y}$ is decomposed into two orthogonal components:

1.  $\mathbf{X}\hat{\beta}$ (the projection onto $\mathrm{Col}(\mathbf{X})$

    -   $\|\mathbf{X}\hat{\beta}\|$ measures the part of $\mathbf{Y}$ explained by the model.

2.  $\mathbf{Y} - \mathbf{X}\hat{\beta}$ (the residual lying in the orthogonal complement).

    -   $\|\mathbf{Y} - \mathbf{X}\hat{\beta}\|$ measures the residual error.

This geometric interpretation (projection plus orthogonal remainder) is exactly why we call $\mathbf{X}\hat{\beta}$ the *orthogonal projection* of $\mathbf{Y}$ onto the column space of $\mathbf{X}$. This decomposition also underlies the analysis of variance (ANOVA) in regression.

------------------------------------------------------------------------

The coefficient of multiple determination, denoted $R^2$, measures the proportion of the total variation in the response variable ($\mathbf{Y}$) that is explained by the regression model. It is defined as:

$$
R^2 = \frac{SSR}{SSTO} = 1 - \frac{SSE}{SSTO},
$$

where:

-   $SSR$: Regression sum of squares (variation explained by the model).

-   $SSE$: Error sum of squares (unexplained variation).

-   $SSTO$: Total sum of squares (total variation in $\mathbf{Y}$).

The adjusted $R^2$ adjusts $R^2$ for the number of predictors in the model, penalizing for adding predictors that do not improve the model's fit substantially. It is defined as:

$$
R^2_a = 1 - \frac{SSE/(n-p)}{SSTO/(n-1)} = 1 - \frac{(n-1)SSE}{(n-p)SSTO},
$$

where:

-   $n$: Number of observations.

-   $p$: Number of parameters (including the intercept).

------------------------------------------------------------------------

| Aspect                   | $R^2$                                                                                                                  | $R^2_a$                                                                                                                  |
|--------------------------|------------------------------------------------------------------------------------------------------------------------|--------------------------------------------------------------------------------------------------------------------------|
| Behavior with Predictors | Always increases (or remains constant) when more predictors are added, even if they are not statistically significant. | Includes a penalty for the number of predictors. May decrease if added predictors do not improve the model sufficiently. |
| Interpretation           | Proportion of the total variation in $\mathbf{Y}$ explained by the regression model.                                   | Adjusted measure of explained variance, accounting for model complexity.                                                 |
| Range                    | Ranges between $0$ and $1$.                                                                                            | Can be lower than $R^2$, particularly when the model includes irrelevant predictors.                                     |
| Usefulness               | Useful for understanding the overall fit of the model.                                                                 | Useful for comparing models with different numbers of predictors.                                                        |

: Key Differences Between $R^2$ and $R^2_a$

In multiple regression, $R^2_a$ provides a more reliable measure of model fit, especially when comparing models with different numbers of predictors.

------------------------------------------------------------------------

In a regression model with coefficients $\beta = (\beta_0, \beta_1, \dots, \beta_{p-1})^\top$, the sums of squares are used to evaluate the contribution of predictors to explaining the variation in the response variable.

1.  Model Sums of Squares:
    -   $SSM$: Total model sum of squares, capturing the variation explained by all predictors: $$
        SSM = SS(\beta_0, \beta_1, \dots, \beta_{p-1}).
        $$
2.  Marginal Contribution:
    -   $SSM_m$: Conditional model sum of squares, capturing the variation explained by predictors after accounting for others: $$
        SSM_m = SS(\beta_0, \beta_1, \dots, \beta_{p-1} | \beta_0).
        $$

------------------------------------------------------------------------

Decompositions of $SSM_m$

1.  Sequential Sums of Squares (Type I SS)

-   Definition:

    -   Sequential SS depends on the order in which predictors are added to the model.
    -   It represents the additional contribution of each predictor given only the predictors that precede it in the sequence.

-   Formula: $$
    SSM_m = SS(\beta_1 | \beta_0) + SS(\beta_2 | \beta_0, \beta_1) + \dots + SS(\beta_{p-1} | \beta_0, \dots, \beta_{p-2}).
    $$

-   Key Points:

    -   Sequential SS is not unique; it depends on the order of the predictors.
    -   Default in many statistical software functions (e.g., `anova()` in R).

2.  Marginal Sums of Squares (Type II SS)

-   Definition:
    -   Marginal SS evaluates the contribution of a predictor after accounting for all other predictors except those with which it is collinear.
    -   It ignores hierarchical relationships or interactions, focusing on independent contributions.
-   Formula: $SSM_m = SS(\beta_j | \beta_1, \dots, \beta_{j-1}, \beta_{j + 1}, \dots, \beta_{p-1})$ where Type II SS evaluates the contribution of $\beta_j$ while excluding any terms collinear with $\beta_j$.
-   Key Points:
    -   Type II SS is independent of predictor order.
    -   Suitable for models without interaction terms or when predictors are balanced.

3.  Partial Sums of Squares (Type III SS)

-   Definition:

    -   Partial SS evaluates the contribution of each predictor after accounting for all other predictors in the model.
    -   It quantifies the unique contribution of a predictor, controlling for the presence of others.

-   Formula: $$
    SSM_m = SS(\beta_1 | \beta_0, \beta_2, \dots, \beta_{p-1}) + \dots + SS(\beta_{p-1} | \beta_0, \beta_1, \dots, \beta_{p-2}).
    $$

-   Key Points:

    -   Partial SS is unique for a given model.
    -   More commonly used in practice for assessing individual predictor importance.

------------------------------------------------------------------------

| Aspect         | Sequential SS (Type I)                                          | Marginal SS (Type II)                                               | Partial SS (Type III)                                 |
|----------------|-----------------------------------------------------------------|---------------------------------------------------------------------|-------------------------------------------------------|
| Dependency     | Depends on the order in which predictors are entered.           | Independent of order; adjusts for non-collinear predictors.         | Independent of order; evaluates unique contributions. |
| Usage          | Default in software functions like `anova()` (Type I SS).       | Models without interactions or hierarchical dependencies.           | Commonly used for hypothesis testing.                 |
| Interpretation | Measures the additional contribution of predictors in sequence. | Measures the contribution of a predictor, ignoring collinear terms. | Measures the unique contribution of each predictor.   |
| Uniqueness     | Not unique; changes with predictor order.                       | Unique for a given model without interactions.                      | Unique for a given model.                             |

: Comparison of Sequential, Marginal, and Partial SS

------------------------------------------------------------------------

Practical Notes

-   Use Type III SS (Partial SS) when:
    -   The focus is on individual predictor contributions while accounting for all others.
    -   Conducting hypothesis tests on predictors in complex models with interactions or hierarchical structures.
-   Use Type II SS (Marginal SS) when:
    -   Working with balanced datasets or models without interaction terms.
    -   Ignoring interactions and focusing on independent effects.
-   Use Type I SS (Sequential SS) when:
    -   Interested in understanding the incremental contribution of predictors based on a specific order of entry (e.g., stepwise regression).

------------------------------------------------------------------------

#### OLS Assumptions {#ols-assumptions}

-   [A1 Linearity](#a1-linearity)
-   [A2 Full Rank](#a2-full-rank)
-   [A3 Exogeneity of Independent Variables](#a3-exogeneity-of-independent-variables)
-   [A4 Homoskedasticity](#a4-homoskedasticity)
-   [A5 Data Generation (Random Sampling)](#a5-data-generation-random-sampling)
-   [A6 Normal Distribution](#a6-normal-distribution)

------------------------------------------------------------------------

##### A1 Linearity {#a1-linearity}

The linear regression model is expressed as:

$$
A1: y = \mathbf{x}\beta + \epsilon
$$

This assumption is not restrictive since $x$ can include nonlinear transformations (e.g., interactions, natural logarithms, quadratic terms).

However, when combined with [A3 Exogeneity of Independent Variables](#a3-exogeneity-of-independent-variables), linearity can become restrictive.

------------------------------------------------------------------------

###### Log Model Variants

Logarithmic transformations of variables allow for flexible modeling of nonlinear relationships. Common log model forms include:

| Model       | Form                                           | Interpretation of $\beta$             | In Words                                                              |
|-------------|------------------------------------------------|---------------------------------------|-----------------------------------------------------------------------|
| Level-Level | $y = \beta_0 + \beta_1x + \epsilon$            | $\Delta y = \beta_1 \Delta x$         | A unit change in $x$ results in a $\beta_1$ unit change in $y$.       |
| Log-Level   | $\ln(y) = \beta_0 + \beta_1x + \epsilon$       | $\% \Delta y = 100 \beta_1 \Delta x$  | A unit change in $x$ results in a $100 \beta_1 \%$ change in $y$.     |
| Level-Log   | $y = \beta_0 + \beta_1 \ln(x) + \epsilon$      | $\Delta y = (\beta_1/100)\% \Delta x$ | A 1% change in $x$ results in a $(\beta_1 / 100)$ unit change in $y$. |
| Log-Log     | $\ln(y) = \beta_0 + \beta_1 \ln(x) + \epsilon$ | $\% \Delta y = \beta_1 \% \Delta x$   | A 1% change in $x$ results in a $\beta_1 \%$ change in $y$.           |

: Interpretation of Coefficients in Different Functional Forms of Regression Models

------------------------------------------------------------------------

###### Higher-Order Models

Higher-order terms allow the effect of $x_1$ on $y$ to depend on the level of $x_1$. For example:

$$
y = \beta_0 + x_1\beta_1 + x_1^2\beta_2 + \epsilon
$$

The partial derivative of $y$ with respect to $x_1$ is:

$$
\frac{\partial y}{\partial x_1} = \beta_1 + 2x_1\beta_2
$$

-   The effect of $x_1$ on $y$ depends on the value of $x_1$.
-   Partial Effect at the Average: $\beta_1 + 2E(x_1)\beta_2$.
-   Average Partial Effect: $E(\beta_1 + 2x_1\beta_2)$.

------------------------------------------------------------------------

###### Interaction Terms

Interactions capture the joint effect of two variables. For example:

$$
y = \beta_0 + x_1\beta_1 + x_2\beta_2 + x_1x_2\beta_3 + \epsilon
$$

-   $\beta_1$ is the average effect of a unit change in $x_1$ on $y$ when $x_2 = 0$.
-   The partial effect of $x_1$ on $y$, which depends on the level of $x_2$, is:

$$
\beta_1 + x_2\beta_3.
$$

------------------------------------------------------------------------

##### A2 Full Rank {#a2-full-rank}

The full rank assumption ensures the uniqueness and existence of the parameter estimates in the population regression equation. It is expressed as:

$$
A2: \text{rank}(E(\mathbf{x'x})) = k
$$

This assumption is also known as the identification condition.

**Key Points**

1.  No Perfect Multicollinearity:
    -   The columns of $\mathbf{x}$ (the matrix of predictors) must be linearly independent.
    -   No column in $\mathbf{x}$ can be written as a linear combination of other columns.
2.  Implications:
    -   Ensures that each parameter in the regression equation is identifiable and unique.
    -   Prevents computational issues, such as the inability to invert $\mathbf{x'x}$, which is required for estimating $\hat{\beta}$.

Example of Violation

If two predictors, $x_1$ and $x_2$, are perfectly correlated (e.g., $x_2 = 2x_1$), the rank of $\mathbf{x}$ is reduced, and $\mathbf{x'x}$ becomes singular. In such cases:

-   The regression coefficients cannot be uniquely estimated.

-   The model fails to satisfy the full rank assumption.

------------------------------------------------------------------------

##### A3 Exogeneity of Independent Variables {#a3-exogeneity-of-independent-variables}

The exogeneity assumption ensures that the independent variables ($\mathbf{x}$) are not systematically related to the error term ($\epsilon$). It is expressed as:

$$
A3: E[\epsilon | x_1, x_2, \dots, x_k] = E[\epsilon | \mathbf{x}] = 0
$$

This assumption is often referred to as strict exogeneity or mean independence (see [Correlation and Independence].

Key Points

1.  Strict Exogeneity:
    -   Independent variables carry no information about the error term $\epsilon$.
    -   By the [Law of Iterated Expectations], $E(\epsilon) = 0$, which can be satisfied by always including an intercept in the regression model.
2.  Implication:
    -   A3 implies: $$
        E(y | \mathbf{x}) = \mathbf{x}\beta,
        $$ meaning the conditional mean function is a linear function of $\mathbf{x}$. This aligns with [A1 Linearity](#a1-linearity).
3.  Relationship with Independence:
    -   Also referred to as mean independence, which is a weaker condition than full independence (see [Correlation and Independence]).

------------------------------------------------------------------------

###### A3a: Weaker Exogeneity Assumption {#a3a-weak-exogeneity}

A weaker version of the exogeneity assumption is:

$$
A3a: E(\mathbf{x_i'}\epsilon_i) = 0
$$

This implies:

-   The independent variables ($\mathbf{x}_i$) are uncorrelated with the error term ($\epsilon_i$).

-   Weaker than mean independence in A3.

------------------------------------------------------------------------

| Aspect         | A3 (Strict Exogeneity)                              | A3a (Weaker Exogeneity)                      |
|----------------|-----------------------------------------------------|----------------------------------------------|
| Definition     | $E(\epsilon | \mathbf{x}) = 0$.                     | $E(\mathbf{x}_i'\epsilon_i) = 0$.            |
| Strength       | Stronger assumption; implies A3a.                   | Weaker assumption; does not imply A3.        |
| Interpretation | Predictors provide no information about $\epsilon$. | Predictors are uncorrelated with $\epsilon$. |
| Causality      | Enables causal interpretation.                      | Does not allow causal interpretations.       |

: Comparison Between [A3](#a3-exogeneity-of-independent-variables) and [A3a](#a3a-weak-exogeneity)

------------------------------------------------------------------------

Notes on Practical Relevance

-   Checking for Exogeneity:
    -   Strict exogeneity cannot be tested directly, but violations can manifest as omitted variable bias, endogeneity, or measurement error.
    -   Including all relevant predictors and ensuring accurate measurement can help satisfy this assumption.
-   Violations of Exogeneity:
    -   If [A3](#a3-exogeneity-of-independent-variables) is violated, standard OLS estimates are biased and inconsistent.
    -   In such cases, instrumental variable (IV) methods or other approaches may be required to correct for endogeneity.

------------------------------------------------------------------------

##### A4 Homoskedasticity {#a4-homoskedasticity}

The homoskedasticity assumption ensures that the variance of the error term ($\epsilon$) is constant across all levels of the independent variables ($\mathbf{x}$). It is expressed as:

$$
A4: \text{Var}(\epsilon | \mathbf{x}) = \text{Var}(\epsilon) = \sigma^2
$$

Key Points

1.  Definition:
    -   The variance of the disturbance term $\epsilon$ is the same for all observations, regardless of the values of the predictors $\mathbf{x}$.
2.  Practical Implication:
    -   Homoskedasticity ensures that the errors do not systematically vary with the predictors.
    -   This is critical for valid inference, as the standard errors of the coefficients rely on this assumption.
3.  Violation (Heteroskedasticity):
    -   If the variance of $\epsilon$ depends on $\mathbf{x}$, the assumption is violated.
    -   Common signs include funnel-shaped patterns in residual plots or varying error sizes.

------------------------------------------------------------------------

##### A5 Data Generation (Random Sampling) {#a5-data-generation-random-sampling}

The random sampling assumption ensures that the observations $(y_i, x_{i1}, \dots, x_{ik-1})$ are drawn independently and identically distributed (iid) from the joint distribution of $(y, \mathbf{x})$. It is expressed as:

$$
A5: \{y_i, x_{i1}, \dots, x_{ik-1} : i = 1, \dots, n\}
$$

Key Points

1.  Random Sampling:
    -   The dataset is assumed to be a random sample from the population.
    -   Each observation is independent of others and follows the same probability distribution.
2.  Implications:
    -   With [A3 (Exogeneity of Independent Variables)](#a3-exogeneity-of-independent-variables) and [A4 (Homoskedasticity)](#a4-homoskedasticity), random sampling implies:
        -   Strict Exogeneity: $$
            E(\epsilon_i | x_1, \dots, x_n) = 0
            $$ Independent variables do not contain information for predicting $\epsilon$.
        -   Non-Autocorrelation: $$
            E(\epsilon_i \epsilon_j | x_1, \dots, x_n) = 0 \quad \text{for } i \neq j
            $$ The error terms are uncorrelated across observations, conditional on the independent variables.
        -   Variance of Errors: $$
            \text{Var}(\epsilon | \mathbf{X}) = \text{Var}(\epsilon) = \sigma^2 \mathbf{I}_n
            $$
3.  When A5 May Not Hold:
    -   In time series data, where observations are often autocorrelated.
    -   In spatial data, where neighboring observations may not be independent.

Practical Considerations

-   Time Series Data:
    -   Use methods such as autoregressive models or generalized least squares (GLS) to address dependency in observations.
-   Spatial Data:
    -   Spatial econometric models may be required to handle correlation across geographic locations.
-   Checking Random Sampling:
    -   While true randomness cannot always be verified, exploratory analysis of the residuals (e.g., for patterns or autocorrelation) can help detect violations.

------------------------------------------------------------------------

###### A5a: Stationarity in Stochastic Processes {#A5a-stationarity-in-stochastic-processes}

A stochastic process $\{x_t\}_{t=1}^T$ is stationary if, for every collection of time indices $\{t_1, t_2, \dots, t_m\}$, the joint distribution of:

$$
x_{t_1}, x_{t_2}, \dots, x_{t_m}
$$

is the same as the joint distribution of:

$$
x_{t_1+h}, x_{t_2+h}, \dots, x_{t_m+h}
$$

for any $h \geq 1$.

Key Points on Stationarity

1.  Definition:
    -   A stationary process has statistical properties (mean, variance, and covariance) that are invariant over time.
    -   For example, the joint distribution for the first ten observations is identical to the joint distribution for the next ten observations, regardless of their position in time.
2.  Implication:
    -   Stationarity ensures that the relationships observed in the data remain consistent over time.
    -   Independent draws automatically satisfy the stationarity condition.

Weak Stationarity

A stochastic process $\{x_t\}_{t=1}^T$ is weakly stationary if:

-   The covariance between $x_t$ and $x_{t+h}$ depends only on the lag $h$ and not on $t$.
-   As $h \to \infty$, the covariance diminishes, meaning $x_t$ and $x_{t+h}$ become "almost independent."

------------------------------------------------------------------------

| Aspect               | Stationarity                                           | Weak Stationarity                                           |
|----------------------|--------------------------------------------------------|-------------------------------------------------------------|
| Joint Distribution   | Entire joint distribution remains unchanged over time. | Focuses only on the first two moments: mean and covariance. |
| Dependence Over Time | Observations at all lags are equally distributed.      | Observations far apart are "almost independent."            |
| Application          | Ensures strong consistency in time-series processes.   | More practical for many time-series applications.           |

: Differences Between Stationarity and Weak Stationarity

Weak stationarity is often sufficient for many time-series analyses, especially when focusing on correlations and trends rather than full distributions.

------------------------------------------------------------------------

**Common Weakly Dependent Processes**

1\. Moving Average Process of Order 1 (MA(1))

An MA(1) process models the dependent variable $y_t$ as a function of the current and one-period lagged stochastic error term:

$$
y_t = u_t + \alpha_1 u_{t-1},
$$

where $u_t$ is white noise, independently and identically distributed (iid) with variance $\sigma^2$.

Key Properties

1.  Mean: $$
    E(y_t) = E(u_t) + \alpha_1E(u_{t-1}) = 0
    $$

2.  Variance: $$
    \begin{aligned}
    \text{Var}(y_t) &= \text{Var}(u_t) + \alpha_1^2 \text{Var}(u_{t-1}) \\
    &= \sigma^2 + \alpha_1^2 \sigma^2 \\
    &= \sigma^2 (1 + \alpha_1^2)
    \end{aligned}
    $$ An increase in the absolute value of $\alpha_1$ increases the variance.

3.  Autocovariance:

    -   For lag 1: $$
        \text{Cov}(y_t, y_{t-1}) = \alpha_1 \text{Var}(u_{t-1}) = \alpha_1 \sigma^2.
        $$
    -   For lag 2 or greater: $$
        \text{Cov}(y_t, y_{t-2}) = 0.
        $$

The MA(1) process is invertible if $|\alpha_1| < 1$, allowing it to be rewritten as an autoregressive (AR) representation:

$$
u_t = y_t - \alpha_1 u_{t-1}.
$$

Invertibility implies that we can express the current observation in terms of past observations.

------------------------------------------------------------------------

An MA(q) process generalizes the MA(1) process to include $q$ lags:

$$
y_t = u_t + \alpha_1 u_{t-1} + \dots + \alpha_q u_{t-q},
$$

where $u_t \sim WN(0, \sigma^2)$ (white noise with mean 0 and variance $\sigma^2$).

Key Characteristics:

1.  Covariance Stationary:
    -   An MA(q) process is covariance stationary irrespective of the parameter values.
2.  Invertibility:
    -   An MA(q) process is invertible if the parameters satisfy certain conditions (e.g., $|\alpha_i| < 1$).
3.  Autocorrelations:
    -   The autocorrelations are nonzero for lags up to $q$ but are 0 for lags beyond $q$.
4.  Conditional Mean:
    -   The conditional mean of $y_t$ depends on the $q$ lags, indicating "long-term memory."

Example:

1.  **Autocovariance of an MA(1)**

<!-- -->

1.  Lag 1: $$
    \begin{aligned}
    \text{Cov}(y_t, y_{t-1}) &= \text{Cov}(u_t + \alpha_1 u_{t-1}, u_{t-1} + \alpha_1 u_{t-2}) \\
    &= \alpha_1 \text{Var}(u_{t-1}) \\
    &= \alpha_1 \sigma^2.
    \end{aligned}
    $$

2.  Lag 2: $$
    \begin{aligned}
    \text{Cov}(y_t, y_{t-2}) &= \text{Cov}(u_t + \alpha_1 u_{t-1}, u_{t-2} + \alpha_1 u_{t-3}) \\
    &= 0.
    \end{aligned}
    $$

An MA process captures a linear relationship between the dependent variable $y_t$ and the current and past values of a stochastic error term $u_t$. Its properties make it useful for modeling time-series data with limited memory and short-term dependencies.

2.  **Auto-Regressive Process of Order 1 (AR(1))**

An auto-regressive process of order 1 (AR(1)) is defined as:

$$
y_t = \rho y_{t-1} + u_t, \quad |\rho| < 1
$$

where $u_t$ represents independent and identically distributed (i.i.d.) random noise over $t$ with variance $\sigma^2$.

Covariance at lag 1:

$$
\begin{aligned}
Cov(y_t, y_{t-1}) &= Cov(\rho y_{t-1} + u_t, y_{t-1}) \\
&= \rho Var(y_{t-1}) \\
&= \rho \frac{\sigma^2}{1-\rho^2}.
\end{aligned}
$$

Covariance at lag $h$:

$$
Cov(y_t, y_{t-h}) = \rho^h \frac{\sigma^2}{1-\rho^2}.
$$

Stationarity implies that the distribution of $y_t$ does not change over time, requiring constant mean and variance. For this process:

1.  Mean Stationarity:\
    Assuming $E(y_t) = 0$, we have: $$
    y_t = \rho^t y_0 + \rho^{t-1}u_1 + \rho^{t-2}u_2 + \dots + \rho u_{t-1} + u_t.
    $$

    If the initial observation $y_0 = 0$, then $y_t$ is simply a weighted sum of the random shocks $u_t$ from all prior time periods. Thus, $E(y_t) = 0$ for all $t$.

2.  Variance Stationarity:\
    The variance is computed as: $$
    Var(y_t) = Var(\rho y_{t-1} + u_t).
    $$

    Expanding and simplifying: $$
    Var(y_t) = \rho^2 Var(y_{t-1}) + Var(u_t) + 2\rho Cov(y_{t-1}, u_t).
    $$ Since $u_t$ is independent of $y_{t-1}$, $Cov(y_{t-1}, u_t) = 0$, giving: $$
    Var(y_t) = \rho^2 Var(y_{t-1}) + \sigma^2.
    $$

    Solving recursively, we find: $$
    Var(y_t) = \frac{\sigma^2}{1-\rho^2}.
    $$

For the variance to remain constant over time, it is required that $|\rho| < 1$ and $p \notin \{1,-1\}$.

Key Insights on Stationarity

1.  Stationarity Requirement:\
    The condition $|\rho| < 1$ ensures stationarity, as this guarantees that both the mean and variance of the process are constant over time.

2.  Weak Dependence:\
    As $|\rho| < 1$, the dependency between observations diminishes with increasing lag $h$, as seen from the covariance: $$
    Cov(y_t, y_{t-h}) = \rho^h \frac{\sigma^2}{1-\rho^2}.
    $$

To estimate an AR(1) process, we utilize the Yule-Walker equations, which relate the autocovariances of the process to its parameters.

Starting with the AR(1) process: $$
y_t = \epsilon_t + \phi y_{t-1},
$$ multiplying both sides by $y_{t-\tau}$ and taking expectations, we get: $$
y_t y_{t-\tau} = \epsilon_t y_{t-\tau} + \phi y_{t-1} y_{t-\tau}.
$$

For $\tau \geq 1$, the autocovariance $\gamma(\tau)$ satisfies: $$
\gamma(\tau) = \phi \gamma(\tau - 1).
$$

Dividing through by the variance $\gamma(0)$, we obtain the autocorrelation: $$
\rho_\tau = \phi^\tau.
$$

Thus, the autocorrelations decay geometrically as $\phi^\tau$, where $|\phi| < 1$ ensures stationarity and decay over time.

------------------------------------------------------------------------

**Generalizing to AR(p)**

An AR(p) process extends the AR(1) structure to include $p$ lags: $$
y_t = \phi_1 y_{t-1} + \phi_2 y_{t-2} + \dots + \phi_p y_{t-p} + \epsilon_t.
$$

Here, $\epsilon_t$ is white noise with $E(\epsilon_t) = 0$ and $Var(\epsilon_t) = \sigma^2$.

The AR(p) process is covariance stationary if the roots of the characteristic equation lie outside the unit circle: $$
1 - \phi_1 z - \phi_2 z^2 - \dots - \phi_p z^p = 0.
$$

For the AR(p) process, the autocorrelations $\rho_\tau$ decay more complexly compared to the AR(1). However, they still diminish over time, ensuring weak dependence among distant observations.

------------------------------------------------------------------------

The Yule-Walker equations for an AR(p) process provide a system of linear equations to estimate the parameters $\phi_1, \phi_2, \dots, \phi_p$: $$
\gamma(\tau) = \phi_1 \gamma(\tau - 1) + \phi_2 \gamma(\tau - 2) + \dots + \phi_p \gamma(\tau - p), \quad \tau \geq 1.
$$

This system can be written in matrix form for $\tau = 1, \dots, p$ as:

$$
\begin{bmatrix}
\gamma(1) \\
\gamma(2) \\
\vdots \\
\gamma(p)
\end{bmatrix}
=
\begin{bmatrix}
\gamma(0) & \gamma(1) & \dots & \gamma(p-1) \\
\gamma(1) & \gamma(0) & \dots & \gamma(p-2) \\
\vdots & \vdots & \ddots & \vdots \\
\gamma(p-1) & \gamma(p-2) & \dots & \gamma(0)
\end{bmatrix}
\begin{bmatrix}
\phi_1 \\
\phi_2 \\
\vdots \\
\phi_p
\end{bmatrix}.
$$

This system is solved to estimate the coefficients $\phi_1, \phi_2, \dots, \phi_p$.

------------------------------------------------------------------------

3.  **ARMA(p, q) Process**

An ARMA(p, q) process combines autoregressive (AR) and moving average (MA) components to model time series data effectively. The general form is:

$$
y_t = \phi_1 y_{t-1} + \phi_2 y_{t-2} + \dots + \phi_p y_{t-p} + \epsilon_t + \alpha_1 \epsilon_{t-1} + \alpha_2 \epsilon_{t-2} + \dots + \alpha_q \epsilon_{t-q}.
$$

A simple case of ARMA(1, 1) is given by: $$
y_t = \phi y_{t-1} + \epsilon_t + \alpha \epsilon_{t-1},
$$ where:

-   $\phi$ captures the autoregressive behavior,

-   $\alpha$ controls the moving average component,

-   $\epsilon_t$ represents white noise.

ARMA processes can capture seasonality and more complex dependencies than pure AR or MA models.

------------------------------------------------------------------------

4.  **Random Walk Process**

A **random walk** is a non-stationary process defined as: $$
y_t = y_0 + \sum_{s=1}^t u_s,
$$ where $u_s$ are i.i.d. random variables.

Properties:

1.  **Non-Stationarity**:

    -   If $y_0 = 0$, then $E(y_t) = 0$, but the variance increases over time: $$
        Var(y_t) = t \sigma^2.
        $$

2.  **Not Weakly Dependent**:

    -   The covariance of the process does not diminish with increasing lag $h$: $$
        Cov\left(\sum_{s=1}^t u_s, \sum_{s=1}^{t-h} u_s\right) = (t-h)\sigma^2.
        $$

    As $h$ increases, the covariance remains large, violating the condition for weak dependence.

------------------------------------------------------------------------

###### A5a: Stationarity and Weak Dependence in Time Series {#A5a-stationarity-and-weak-dependence-in-time-series}

For time series data, the set $\{y_t, x_{t1}, \dots, x_{tk-1}\}$, where $t = 1, \dots, T$, must satisfy the conditions of **stationarity** and **weak dependence**. These properties are essential to ensure the consistency and efficiency of estimators in time-series regression models.

1.  **Stationarity**: A stationary process has statistical properties (e.g., mean, variance, autocovariance) that remain constant over time. This ensures that the relationships in the data do not change as $t$ progresses, making it possible to draw meaningful inferences.
2.  **Weak Dependence**: Weak dependence implies that observations far apart in time are "almost independent." While there may be short-term correlations, these diminish as the time lag increases. This property ensures that the sample averages are representative of the population mean.

The [Weak Law of Large Numbers] provides a foundation for the consistency of sample means. If $\{z_t\}$ is a weakly dependent, stationary process with $E(|z_t|) < \infty$ and $E(z_t) = \mu$, then:

$$
\frac{1}{T} \sum_{t=1}^T z_t \xrightarrow{p} \mu.
$$

Interpretation:

-   As the sample size $T \to \infty$, the sample mean $\bar{z} = \frac{1}{T} \sum_{t=1}^T z_t$ converges in probability to the true mean $\mu$.

-   This ensures the **consistency** of estimators based on time-series data.

The [Central Limit Theorem] extends the WLLN by describing the distribution of the sample mean. Under additional regularity conditions (e.g., finite variance) [@greene1990gamma], the sample mean $\bar{z}$ satisfies:

$$
\sqrt{T}(\bar{z} - \mu) \xrightarrow{d} N(0, B),
$$

where:

$$
B = \text{Var}(z_t) + 2\sum_{h=1}^\infty \text{Cov}(z_t, z_{t-h}).
$$

**Interpretation**:

-   The sample mean $\bar{z}$ is approximately normally distributed for large $T$.

-   The variance of the limiting distribution, $B$, depends not only on the variance of $z_t$ but also on the covariances between $z_t$ and its past values.

------------------------------------------------------------------------

##### A6 Normal Distribution {#a6-normal-distribution}

A6: $\epsilon|\mathbf{x} \sim N(0, \sigma^2 I_n)$

The assumption here implies that the error term $\epsilon$ is normally distributed with mean zero and variance $\sigma^2 I_n$. This assumption is fundamental for statistical inference in linear regression models.

Using assumptions [A1 Linearity](#a1-linearity), [A2 Full Rank](#a2-full-rank), and [A3 Exogeneity of Independent Variables](#a3-exogeneity-of-independent-variables), we derive the identification (or orthogonality condition) for the population parameter $\beta$:

$$
\begin{aligned}
y &= x\beta + \epsilon && \text{(A1: Model Specification)} \\
x'y &= x'x\beta + x'\epsilon && \text{(Multiply both sides by $x'$)} \\
E(x'y) &= E(x'x)\beta + E(x'\epsilon) && \text{(Taking expectation)} \\
E(x'y) &= E(x'x)\beta && \text{(A3: Exogeneity, $E(x'\epsilon) = 0$)} \\
[E(x'x)]^{-1}E(x'y) &= [E(x'x)]^{-1}E(x'x)\beta && \text{(Invertibility of $E(x'x)$, A2)} \\
[E(x'x)]^{-1}E(x'y) &= \beta && \text{(Simplified solution for $\beta$)}
\end{aligned}
$$

Thus, $\beta$ is identified as the vector of parameters that minimizes the expected squared error.

To find $\beta$, we minimize the expected value of the squared error:

$$
\underset{\gamma}{\operatorname{argmin}} \ E\big((y - x\gamma)^2\big)
$$

The first-order condition is derived by taking the derivative of the objective function with respect to $\gamma$ and setting it to zero:

$$
\begin{aligned}
\frac{\partial E\big((y - x\gamma)^2\big)}{\partial \gamma} &= 0 \\
-2E(x'(y - x\gamma)) &= 0 \\
E(x'y) - E(x'x\gamma) &= 0 \\
E(x'y) &= E(x'x)\gamma \\
(E(x'x))^{-1}E(x'y) &= \gamma
\end{aligned}
$$

This confirms that $\gamma = \beta$.

The second-order condition ensures that the solution minimizes the objective function. Taking the second derivative:

$$
\frac{\partial^2 E\big((y - x\gamma)^2\big)}{\partial \gamma'^2} = 0 = 2E(x'x) 
$$

If assumption [A3 Exogeneity of Independent Variables](#a3-exogeneity-of-independent-variables) holds, $E(x'x)$ is positive semi-definite (PSD). Thus, $2E(x'x)$ is also PSD, ensuring a minimum.

##### Hierarchy of OLS Assumptions

This table summarizes the hierarchical nature of assumptions required to derive different properties of the OLS estimator.

+-----------------------------------------------------------------------------------+-------------------------------------+------------------------------+-----------------------------------------------------------------------------------------------+---------------------------------------------------------+
| Assumption                                                                        | **Identification Data Description** | **Unbiasedness Consistency** | [**Gauss-Markov**](#gauss-markov-theorem) **(BLUE) Asymptotic Inference (z and Chi-squared)** | **Classical LM (BUE) Small-sample Inference (t and F)** |
+===================================================================================+=====================================+==============================+===============================================================================================+=========================================================+
| [A2 Full Rank](#a2-full-rank)                                                     | $\checkmark$                        | $\checkmark$                 | $\checkmark$                                                                                  | $\checkmark$                                            |
|                                                                                   |                                     |                              |                                                                                               |                                                         |
| Variation in $\mathbf{X}$                                                         |                                     |                              |                                                                                               |                                                         |
+-----------------------------------------------------------------------------------+-------------------------------------+------------------------------+-----------------------------------------------------------------------------------------------+---------------------------------------------------------+
| [A5 Data Generation (Random Sampling)](#a5-data-generation-random-sampling)       |                                     | $\checkmark$                 | $\checkmark$                                                                                  | $\checkmark$                                            |
|                                                                                   |                                     |                              |                                                                                               |                                                         |
| Random Sampling                                                                   |                                     |                              |                                                                                               |                                                         |
+-----------------------------------------------------------------------------------+-------------------------------------+------------------------------+-----------------------------------------------------------------------------------------------+---------------------------------------------------------+
| [A1 Linearity](#a1-linearity)                                                     |                                     | $\checkmark$                 | $\checkmark$                                                                                  | $\checkmark$                                            |
|                                                                                   |                                     |                              |                                                                                               |                                                         |
| Linearity in Parameters                                                           |                                     |                              |                                                                                               |                                                         |
+-----------------------------------------------------------------------------------+-------------------------------------+------------------------------+-----------------------------------------------------------------------------------------------+---------------------------------------------------------+
| [A3 Exogeneity of Independent Variables](#a3-exogeneity-of-independent-variables) |                                     | $\checkmark$                 | $\checkmark$                                                                                  | $\checkmark$                                            |
|                                                                                   |                                     |                              |                                                                                               |                                                         |
| Zero Conditional Mean                                                             |                                     |                              |                                                                                               |                                                         |
+-----------------------------------------------------------------------------------+-------------------------------------+------------------------------+-----------------------------------------------------------------------------------------------+---------------------------------------------------------+
| [A4 Homoskedasticity](#a4-homoskedasticity)                                       |                                     |                              | $\checkmark$                                                                                  | $\checkmark$                                            |
|                                                                                   |                                     |                              |                                                                                               |                                                         |
| $\mathbf{H}$ homoskedasticity                                                     |                                     |                              |                                                                                               |                                                         |
+-----------------------------------------------------------------------------------+-------------------------------------+------------------------------+-----------------------------------------------------------------------------------------------+---------------------------------------------------------+
| [A6 Normal Distribution](#a6-normal-distribution)                                 |                                     |                              |                                                                                               | $\checkmark$                                            |
|                                                                                   |                                     |                              |                                                                                               |                                                         |
| Normality of Errors                                                               |                                     |                              |                                                                                               |                                                         |
+-----------------------------------------------------------------------------------+-------------------------------------+------------------------------+-----------------------------------------------------------------------------------------------+---------------------------------------------------------+

: Mapping Linear Model Assumptions to Estimator Properties and Inference Validity

Usage of Assumptions

-   **Identification Data Description:** Ensures the model is identifiable and coefficients can be estimated.
-   **Unbiasedness Consistency:** Guarantees that OLS estimates are unbiased and converge to the true parameter values as the sample size increases.
-   **Gauss-Markov (BLUE) and Asymptotic Inference:** Requires additional assumptions (e.g., homoskedasticity) to ensure minimum variance of estimators and valid inference using large-sample tests (z and chi-squared).
-   **Classical LM (BUE) Small-sample Inference:** Builds on all previous assumptions and adds normality of errors for valid t and F tests in finite samples.

------------------------------------------------------------------------

#### Theorems

##### Frisch--Waugh--Lovell Theorem {#Frisch--Waugh--Lovell Theorem}

The Frisch--Waugh--Lovell (FWL) Theorem is a fundamental result in linear regression that allows for a deeper understanding of how coefficients are computed in a multiple regression setting [@lovell2008simple]. Informally, it states:

> When estimating the effect of a subset of variables ($X_1$) on $y$ in the presence of other variables ($X_2$), you can "partial out" the influence of $X_2$ from both $y$ and $X_1$. Then, regressing the residuals of $y$ on the residuals of $X_1$ produces coefficients for $X_1$ that are identical to those obtained from the full multiple regression.

Consider the multiple linear regression model:

$$ \mathbf{y = X\beta + \epsilon = X_1\beta_1 + X_2\beta_2 + \epsilon} $$

where:

-   $y$ is an $n \times 1$ vector of the dependent variable.
-   $X_1$ is an $n \times k_1$ matrix of regressors of interest.
-   $X_2$ is an $n \times k_2$ matrix of additional regressors.
-   $\beta_1$ and $\beta_2$ are coefficient vectors of sizes $k_1 \times 1$ and $k_2 \times 1$, respectively.
-   $\epsilon$ is an $n \times 1$ error term vector.

This can be equivalently represented in partitioned matrix form as:

$$ \left( \begin{array}{cc} X_1'X_1 & X_1'X_2 \\ X_2'X_1 & X_2'X_2 \end{array} \right) \left( \begin{array}{c} \hat{\beta_1} \\ \hat{\beta_2} \end{array} \right) = \left( \begin{array}{c} X_1'y \\ X_2'y \end{array} \right) $$

The OLS estimator for the vector $\begin{pmatrix} \beta_1 \\ \beta_2 \end{pmatrix}$ is:

$$
\begin{pmatrix}
\hat{\beta}_1 \\
\hat{\beta}_2
\end{pmatrix}
=
\begin{pmatrix}
X_1'X_1 & X_1'X_2 \\
X_2'X_1 & X_2'X_2
\end{pmatrix}^{-1}
\begin{pmatrix}
X_1'y \\
X_2'y
\end{pmatrix}.
$$

If we only want the coefficients on $X_1$, a known result from partitioned-inversion gives:

$$
\hat{\beta}_1 
= 
\left(X_1' M_2 X_1\right)^{-1} 
X_1' M_2 y,
$$

where

$$
M_2 
= 
I 
- 
X_2 \left(X_2'X_2\right)^{-1} X_2'.
$$

The matrix $M_2$ is often called the **residual-maker** or **annihilator** matrix for $X_2$. It is an $n \times n$ symmetric, idempotent projection matrix that projects any vector in $\mathbb{R}^n$ onto the orthogonal complement of the column space of $X_2$. $M_2$ satisfies $M_2^2 = M_2$, and $M_2 = M_2'$.

Intuitively, $M_2$ captures the part of $y$ (and any other vector) that is orthogonal to the columns of $X_2$. This "partialling out" of $X_2$ from both $y$ and $X_1$ lets us isolate $\hat{\beta}_1$.

Equivalently, we can also represent $\hat{\beta_1}$ as:

$$ \mathbf{\hat{\beta_1} = (X_1'X_1)^{-1}X_1'y - (X_1'X_1)^{-1}X_1'X_2\hat{\beta_2}} $$

From this equation, we can see that

1.  **Betas from Multiple vs. Simple Regressions:**
    -   The coefficients ($\beta$) from a multiple regression are generally **not the same** as the coefficients from separate individual simple regressions.
2.  **Impact of Additional Variables (**$X_2$):
    -   The inclusion of different sets of explanatory variables ($X_2$) affects all coefficient estimates, even for those in $X_1$.
3.  **Special Cases:**
    -   If $X_1'X_2 = 0$ (orthogonality between $X_1$ and $X_2$) or $\hat{\beta_2} = 0$, the above points (1 and 2) do not hold. In such cases, there is no interaction between the coefficients in $X_1$ and $X_2$, making the coefficients in $X_1$ unaffected by $X_2$.

**Steps in FWL:**

1.  **Partial Out** $X_2$ from $y$: Regress $y$ on $X_2$ to obtain residuals:

    $$
    \tilde{y} = M_2y.
    $$

2.  **Partial Out** $X_2$ from $X_1$: For each column of $X_1$, regress it on $X_2$ to obtain residuals:

    $$
    \tilde{X}_1 = M_2X_1.
    $$

3.  **Regression of Residuals:** Regress $\tilde{y}$ on $\tilde{X}_1$:

    $$
    \tilde{y} = \tilde{X}_1\beta_1 + \text{error}.
    $$

The coefficients $\beta_1$ obtained here are identical to those from the full model regression:

$$
y = X_1\beta_1 + X_2\beta_2 + \epsilon.
$$

Why It Matters

1.  **Interpretation of Partial Effects:** The FWL Theorem provides a way to interpret $\beta_1$ as the effect of $X_1$ on $y$ after removing any linear dependence on $X_2$.

2.  **Computational Simplicity:** It allows the decomposition of a large regression problem into smaller, computationally simpler pieces.

```{r}
# Simulate data
set.seed(123)
n <- 100
X1 <- matrix(rnorm(n * 2), n, 2)  # Two regressors of interest
X2 <- matrix(rnorm(n * 2), n, 2)  # Two additional regressors
beta1 <- c(2,-1)  # Coefficients for X1
beta2 <- c(1, 0.5)  # Coefficients for X2undefined
u <- rnorm(n)  # Error term
y <- X1 %*% beta1 + X2 %*% beta2 + u  # Generate dependent variable

# Full regression
full_model <- lm(y ~ X1 + X2)
summary(full_model)

# Step 1: Partial out X2 from y
y_residual <- residuals(lm(y ~ X2))

# Step 2: Partial out X2 from X1
X1_residual <- residuals(lm(X1 ~ X2))

# Step 3: Regress residuals
fwl_model <- lm(y_residual ~ X1_residual - 1)
summary(fwl_model)

# Comparison of coefficients
cat("Full model coefficients (X1):", coef(full_model)[2:3], "\n")
cat("FWL model coefficients:", coef(fwl_model), "\n")
```

------------------------------------------------------------------------

##### Gauss-Markov Theorem {#gauss-markov-theorem}

For a linear regression model:

$$
\mathbf{y = X\beta + \epsilon},
$$

under the assumptions:

1.  [**A1**](#a1-linearity): Linearity of the model.
2.  [**A2**](#a2-full-rank): Full rank of $\mathbf{X}$.
3.  [**A3**](#a3-exogeneity-of-independent-variables): Exogeneity of $\mathbf{X}$.
4.  [**A4**](#a4-homoskedasticity): Homoskedasticity of $\epsilon$.

The [Ordinary Least Squares](#ordinary-least-squares) estimator:

$$
\hat{\beta} = \mathbf{(X'X)^{-1}X'y},
$$

is the **Best Linear Unbiased Estimator (BLUE)**. This means that $\hat{\beta}$ has the minimum variance among all linear unbiased estimators of $\beta$.

**1. Unbiasedness**

Suppose we consider **any** linear estimator of $\beta$ of the form:

$$
\tilde{\beta} = \mathbf{Cy},
$$

where

-   $\mathbf{y}$ is the $n \times 1$ vector of observations,
-   $\mathbf{C}$ is a $k \times n$ matrix (with $k$ the dimension of $\beta$) that depends **only** on the design matrix $\mathbf{X}$.

Our regression model is

$$
\mathbf{y} = \mathbf{X}\beta + \mathbf{\epsilon}, 
\quad 
E[\mathbf{\epsilon} \mid \mathbf{X}] = \mathbf{0}, 
\quad 
\mathrm{Var}(\mathbf{\epsilon} \mid \mathbf{X}) = \sigma^2 \mathbf{I}.
$$

We say $\tilde{\beta}$ is **unbiased** if its conditional expectation (given $\mathbf{X}$) equals the true parameter $\beta$:

$$
E(\tilde{\beta} \mid \mathbf{X})
= 
E(\mathbf{Cy} \mid \mathbf{X})
= 
\beta.
$$

1.  Substitute $\mathbf{y} = \mathbf{X}\beta + \mathbf{\epsilon}$:

$$
E(\mathbf{Cy} \mid \mathbf{X})
=
E\left(\mathbf{C}(\mathbf{X}\beta + \mathbf{\epsilon}) \mid \mathbf{X}\right)
=
\mathbf{CX}\beta 
+
\mathbf{C}E(\mathbf{\epsilon} \mid \mathbf{X})
=
\mathbf{CX}\beta.
$$

2.  For this to hold for *all* $\beta$, we require

$$
\mathbf{CX} = \mathbf{I}.
$$

In other words, $\mathbf{C}$ must be a "right-inverse" of $\mathbf{X}$.

On the other hand, the OLS estimator $\hat{\beta}$ is given by

$$
\hat{\beta} 
= 
(\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{y}.
$$

You can verify:

-   Let $\mathbf{C}_{\text{OLS}} = (\mathbf{X}'\mathbf{X})^{-1} \mathbf{X}'$.
-   Then $$
    \mathbf{C}_{\text{OLS}}\mathbf{X}
    =
    (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{X}
    =
    \mathbf{I}.
    $$
-   By the argument above, this makes $\hat{\beta}$ unbiased.

Hence, any linear estimator $\tilde{\beta} = \mathbf{Cy}$ that is unbiased must satisfy $\mathbf{CX} = \mathbf{I}$.

**2. Minimum Variance (Gauss--Markov Part)**

Among all estimators of the form $\tilde{\beta} = \mathbf{Cy}$ that are unbiased (so $\mathbf{CX} = \mathbf{I}$), **OLS** achieves the smallest covariance matrix.

-   Variance of a General Unbiased Estimator

If $\tilde{\beta} = \mathbf{Cy}$ with $\mathbf{CX} = \mathbf{I}$, then:

$$
\mathrm{Var}(\tilde{\beta} \mid \mathbf{X})
=
\mathrm{Var}(\mathbf{Cy} \mid \mathbf{X})
=
\mathbf{C}\mathrm{Var}(\mathbf{y} \mid \mathbf{X})\mathbf{C}'
=
\mathbf{C}\left(\sigma^2 \mathbf{I}\right)\mathbf{C}'
=
\sigma^2\mathbf{C}\mathbf{C}'.
$$

-   Variance of the OLS Estimator

For [OLS](#ordinary-least-squares), $\hat{\beta} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{y}$. Thus,

$$
\mathrm{Var}(\hat{\beta} \mid \mathbf{X})
=
\sigma^2(\mathbf{X}'\mathbf{X})^{-1}.
$$

-   Comparing $\mathrm{Var}(\tilde{\beta})$ to $\mathrm{Var}(\hat{\beta})$

We want to show:

$$
\mathrm{Var}(\tilde{\beta} \mid \mathbf{X}) 
- 
\mathrm{Var}(\hat{\beta} \mid \mathbf{X})
$$

is positive semi-definite.

1.  Since both $\tilde{\beta}$ and $\hat{\beta}$ are unbiased, we know: $$
    \mathbf{CX} = \mathbf{I},
    \quad
    (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{X} = \mathbf{I}.
    $$

2.  One can show algebraically (as in the proof provided above) that $$
    \mathrm{Var}(\tilde{\beta} \mid \mathbf{X})
    -
    \mathrm{Var}(\hat{\beta} \mid \mathbf{X})
    =
    \sigma^2 \left[\mathbf{C}\mathbf{C}' - (\mathbf{X}'\mathbf{X})^{-1}\right].
    $$ Under the condition $\mathbf{CX} = \mathbf{I}$, the difference $\mathbf{C}\mathbf{C}' - (\mathbf{X}'\mathbf{X})^{-1}$ is positive semi-definite.

3.  Positive semi-definite difference means

$$
\mathbf{v}' \left(\mathbf{C}\mathbf{C}' 
- 
(\mathbf{X}'\mathbf{X})^{-1}\right)\mathbf{v} 
\ge 
0 
\quad 
\text{for all vectors } \mathbf{v}.
$$

Hence, $\hat{\beta}$ has the smallest variance (in the sense of covariance matrices) among all linear unbiased estimators $\tilde{\beta} = \mathbf{Cy}$.

**Summary of the Key Points**

1.  **Unbiasedness:**\
    A linear estimator $\tilde{\beta} = \mathbf{Cy}$ is unbiased if $E(\tilde{\beta}\mid \mathbf{X}) = \beta$.\
    This forces $\mathbf{CX} = \mathbf{I}$.

2.  **OLS is Unbiased:**\
    The OLS estimator $\hat{\beta} = (X'X)^{-1} X'  y$ satisfies $(X'X)^{-1} X'  X = I$, hence is unbiased.

3.  **OLS has Minimum Variance:**\
    Among all $\mathbf{C}$ that satisfy $\mathbf{CX} = \mathbf{I}$, the matrix $(\mathbf{X}'\mathbf{X})^{-1}$ gives the smallest possible $\mathrm{Var}(\tilde{\beta})$.\
    In matrix form, $\mathrm{Var}(\tilde{\beta}) - \mathrm{Var}(\hat{\beta})$ is positive semi-definite, showing OLS is optimal (the **Best Linear Unbiased Estimator**, BLUE).

------------------------------------------------------------------------

#### Finite Sample Properties

The finite sample properties of an estimator are considered when the sample size $n$ is fixed (not asymptotically large). Key properties include **bias**, **distribution**, and **standard deviation** of the estimator.

**Bias** measures how close an estimator is, on average, to the true parameter value $\beta$. It is defined as:

$$
\text{Bias} = E(\hat{\beta}) - \beta
$$

Where:

-   $\beta$: True parameter value.
-   $\hat{\beta}$: Estimator for $\beta$.

**Unbiased Estimator**: An estimator is unbiased if:

$$
\text{Bias} = E(\hat{\beta}) - \beta = 0 \quad \text{or equivalently} \quad E(\hat{\beta}) = \beta
$$

This means the estimator will produce estimates that are, on average, equal to the value it is trying to estimate.

An estimator is a function of random variables (data). Its **distribution** describes how the estimates vary across repeated samples. Key aspects include:

-   **Center**: Mean of the distribution, which relates to bias.
-   **Spread**: Variability of the estimator, captured by its standard deviation or variance.

The **standard deviation** of an estimator measures the spread of its sampling distribution. It indicates the variability of the estimator across different samples.

------------------------------------------------------------------------

##### Ordinary Least Squares Properties

Under the standard assumptions for OLS:

1.  [A1](#a1-linearity): The relationship between $Y$ and $X$ is linear.
2.  [A2](#a2-full-rank): The matrix $\mathbf{X'X}$ is invertible.
3.  [A3](#a3-exogeneity-of-independent-variables): $E(\epsilon|X) = 0$ (errors are uncorrelated with predictors).

**OLS is unbiased** under these assumptions. The proof is as follows:

$$
\begin{aligned}
E(\hat{\beta}) &= E(\mathbf{(X'X)^{-1}X'y}) && \text{A2}\\
               &= E(\mathbf{(X'X)^{-1}X'(X\beta + \epsilon)}) && \text{A1}\\
               &= E(\mathbf{(X'X)^{-1}X'X\beta + (X'X)^{-1}X'\epsilon}) \\
               &= E(\beta + \mathbf{(X'X)^{-1}X'\epsilon}) \\
               &= \beta + E(\mathbf{(X'X)^{-1}X'\epsilon}) \\
               &= \beta + E(E(\mathbf{(X'X)^{-1}X'\epsilon}|X)) && \text{LIE (Law of Iterated Expectation)} \\
               &= \beta + E(\mathbf{(X'X)^{-1}X'}E(\epsilon|X)) \\
               &= \beta + E(\mathbf{(X'X)^{-1}X'} \cdot 0) && \text{A3}\\
               &= \beta
\end{aligned}
$$

**Key Points**:

1.  **Linearity of Expectation**: Used to separate terms involving $\beta$ and $\epsilon$.
2.  [Law of Iterated Expectation] **(LIE)**: Simplifies nested expectations.
3.  **Exogeneity of Errors ([A3](#a3-exogeneity-of-independent-variables))**: Ensures $E(\epsilon|X) = 0$, eliminating bias.

------------------------------------------------------------------------

Implications of Unbiasedness

-   OLS estimators are centered around the true value $\beta$ across repeated samples.
-   In small samples, OLS estimators may exhibit variability, but their expected value remains $\beta$.

If the assumption of exogeneity ([A3](#a3-exogeneity-of-independent-variables)) is violated, the OLS estimator becomes **biased**. Specifically, omitted variables or endogeneity can introduce systematic errors into the estimation.

From the [Frisch-Waugh-Lovell Theorem](#Frisch–Waugh–Lovell%20Theorem):

-   If an omitted variable $\hat{\beta}_2 \neq 0$ (non-zero effect) and the omitted variable is correlated with the included predictors ($\mathbf{X_1'X_2} \neq 0$), then the OLS estimator will be biased.
-   This bias arises because the omitted variable contributes to the variation in the dependent variable, but its effect is incorrectly attributed to other predictors.

------------------------------------------------------------------------

##### Conditional Variance of OLS Estimator

Under assumptions [A1](#a1-linearity), [A2](#a2-full-rank), [A3](#a3-exogeneity-of-independent-variables), and [A4](#a4-homoskedasticity), the **conditional variance of the OLS estimator** is:

$$
\begin{aligned}
Var(\hat{\beta}|\mathbf{X}) &= Var(\beta + \mathbf{(X'X)^{-1}X'\epsilon|X}) && \text{A1-A2} \\
    &= Var((\mathbf{X'X)^{-1}X'\epsilon|X}) \\
    &= \mathbf{(X'X)^{-1}X'} Var(\epsilon|\mathbf{X})\mathbf{X(X'X)^{-1}} \\
    &= \mathbf{(X'X)^{-1}X'} \sigma^2 I \mathbf{X(X'X)^{-1}} && \text{A4} \\
    &= \sigma^2 \mathbf{(X'X)^{-1}}
\end{aligned}
$$

This result shows that the variance of $\hat{\beta}$ depends on:

1.  $\sigma^2$: The variance of the errors.
2.  $\mathbf{X'X}$: The information content in the design matrix $\mathbf{X}$.

------------------------------------------------------------------------

##### Sources of Variation in OLS Estimator

1.  **Unexplained Variation in the Dependent Variable**: $\sigma^2 = Var(\epsilon_i|\mathbf{X})$

-   Large $\sigma^2$ indicates that the amount of unexplained variation (noise) is high relative to the explained variation ($\mathbf{x_i \beta}$).
-   This increases the variance of the OLS estimator.

2.  **Small Variation in Predictor Variables**

-   If the variance of predictors ($Var(x_{i1}), Var(x_{i2}), \dots$) is small, the design matrix $\mathbf{X}$ lacks information, leading to:
    -   High variability in $\hat{\beta}$.
    -   Potential issues in estimating coefficients accurately.
-   **Small sample size** exacerbates this issue, as fewer observations reduce the robustness of parameter estimates.

3.  **Correlation Between Explanatory Variables (Collinearity)**

-   **Strong correlation** among explanatory variables creates problems:
    -   $x_{i1}$ being highly correlated with a linear combination of $1, x_{i2}, x_{i3}, \dots$ contributes to inflated standard errors for $\hat{\beta}_1$.
    -   Including many irrelevant variables exacerbates this issue.
-   **Perfect Collinearity**:
    -   If $x_1$ is perfectly determined by a linear combination of other predictors, the matrix $\mathbf{X'X}$ becomes singular.
    -   This violates [A2](#a2-full-rank), making OLS impossible to compute.
-   **Multicollinearity**:
    -   If $x_1$ is highly correlated (but not perfectly) with a linear combination of other variables, the variance of $\hat{\beta}_1$ increases.
    -   Multicollinearity does not violate OLS assumptions but weakens inference by inflating standard errors.

------------------------------------------------------------------------

##### Standard Errors

Standard errors measure the variability of an estimator, specifically the standard deviation of $\hat{\beta}$. They are crucial for inference, as they quantify the uncertainty associated with parameter estimates.

------------------------------------------------------------------------

The variance of the OLS estimator $\hat{\beta}$ is:

$$
Var(\hat{\beta}|\mathbf{X}) = \sigma^2 \mathbf{(X'X)^{-1}}
$$

Where:

-   $\sigma^2$: Variance of the error terms.
-   $\mathbf{(X'X)^{-1}}$: Inverse of the design matrix product, capturing the geometry of the predictors.

------------------------------------------------------------------------

**Estimation of** $\sigma^2$

Under assumptions [A1](#a1-linearity) through [A5](#a5-homoskedasticity), we can estimate $\sigma^2$ as:

$$
s^2 = \frac{1}{n-k} \sum_{i=1}^{n} e_i^2 = \frac{1}{n-k} SSR
$$

Where:

-   $n$: Number of observations.
-   $k$: Number of predictors, including the intercept.
-   $e_i$: Residuals from the regression model ($e_i = y_i - \hat{y}_i$).
-   $SSR$: Sum of squared residuals ($\sum e_i^2$).

The degrees of freedom adjustment ($n-k$) accounts for the fact that residuals $e_i$ are not true errors $\epsilon_i$. Since the regression model uses $k$ parameters, we lose $k$ degrees of freedom in estimating variance.

The standard error for $\sigma$ is:

$$
s = \sqrt{s^2}
$$

However, $s$ is a biased estimator of $\sigma$ due to [Jensen's Inequality].

------------------------------------------------------------------------

The standard error of each regression coefficient $\hat{\beta}_{j-1}$ is:

$$
SE(\hat{\beta}_{j-1}) = s \sqrt{[(\mathbf{X'X})^{-1}]_{jj}}
$$

Alternatively, it can be expressed in terms of $SST_{j-1}$ and $R_{j-1}^2$:

$$
SE(\hat{\beta}_{j-1}) = \frac{s}{\sqrt{SST_{j-1}(1 - R_{j-1}^2)}}
$$

Where:

-   $SST_{j-1}$: Total sum of squares for $x_{j-1}$ from a regression of $x_{j-1}$ on all other predictors.
-   $R_{j-1}^2$: Coefficient of determination for the same regression.

This formulation highlights the role of multicollinearity, as $R_{j-1}^2$ reflects the correlation between $x_{j-1}$ and other predictors.

------------------------------------------------------------------------

##### Summary of Finite Sample Properties of OLS Under Different Assumptions

1.  **Under [A1-A3](#a1-linearity)**: [OLS](#ordinary-least-squares) is unbiased. $$ E(\hat{\beta}) = \beta $$

2.  **Under [A1-A4](#a4-homoskedasticity)**: The variance of the OLS estimator is: $$ Var(\hat{\beta}|\mathbf{X}) = \sigma^2 \mathbf{(X'X)^{-1}} $$

3.  **Under [A1-A4, A6](#a6-normality-of-errors)**: The OLS estimator is normally distributed: $$ \hat{\beta} \sim N(\beta, \sigma^2 \mathbf{(X'X)^{-1}}) $$

4.  **Under [A1-A4, Gauss-Markov Theorem](#gauss-markov-theorem)**: OLS is BLUE (Best Linear Unbiased Estimator).

5.  **Under [A1-A5](#a5-homoskedasticity)**: The standard errors for $\hat{\beta}$ are unbiased estimators of the standard deviation of $\hat{\beta}$.

------------------------------------------------------------------------

#### Large Sample Properties

Large sample properties provide a framework to evaluate the quality of estimators when finite sample properties are either uninformative or computationally infeasible. This perspective becomes crucial in modern data analysis, especially for methods like GLS or MLE, where assumptions for finite sample analysis may not hold.

| **Aspect**               | **Finite Sample Properties**                                     | **Large Sample Properties**                                 |
|--------------------------|------------------------------------------------------------------|-------------------------------------------------------------|
| **Applicability**        | Limited to fixed sample sizes                                    | Relevant for $n \to \infty$                                 |
| **Exactness**            | Exact results (e.g., distributions, unbiasedness)                | Approximate results                                         |
| **Assumptions**          | May require stronger assumptions (e.g., normality, independence) | Relies on asymptotic approximations (e.g., CLT, LLN)        |
| **Estimator Behavior**   | Performance may vary significantly                               | Estimators stabilize and improve in accuracy                |
| **Ease of Use**          | Often complex due to reliance on exact distributions             | Simplifies analysis by leveraging asymptotic approximations |
| **Real-World Relevance** | More realistic for small datasets                                | More relevant for large datasets                            |

: When to Use Finite vs. Large Sample Analysis

-   **Finite Sample Analysis**:

    -   Small sample sizes (e.g., $n < 30$).

    -   Critical for studies where exact results are needed.

    -   Useful in experimental designs and case studies.

-   **Large Sample Analysis**:

    -   Large datasets (e.g., $n > 100$).

    -   Necessary when asymptotic approximations improve computational simplicity.

------------------------------------------------------------------------

**Key Concepts:**

1.  **Consistency**:
    -   Consistency ensures that an estimator converges in probability to the true parameter value as the sample size increases.
    -   Mathematically, an estimator $\hat{\theta}$ is consistent for $\theta$ if: $$
        \hat{\theta}_n \to^p \theta \quad \text{as } n \to \infty.
        $$
    -   Consistency does not imply unbiasedness, and unbiasedness does not guarantee consistency.
2.  **Asymptotic Distribution**:
    -   The limiting distribution describes the shape of the scaled estimator as $n \to \infty$.
    -   Asymptotic distributions often follow normality due to the [Central Limit Theorem], which underpins much of inferential statistics.
3.  **Asymptotic Variance**:
    -   Represents the spread of the estimator with respect to its limiting distribution.
    -   Smaller asymptotic variance implies greater precision in large samples.

------------------------------------------------------------------------

**Motivation**

[Finite Sample Properties], such as unbiasedness, rely on strong assumptions like:

-   [A1 Linearity](#a1-linearity)
-   [A3 Exogeneity of Independent Variables](#a3-exogeneity-of-independent-variables)
-   [A4 Homoskedasticity](#a4-homoskedasticity)
-   [A6 Normal Distribution](#a6-normal-distribution)

When these assumptions are violated or impractical to verify, finite sample properties lose relevance. In such cases, [Large Sample Properties]serve as an essential alternative for evaluating estimators.

For example, let the conditional expectation function (CEF) be: $$
\mu(\mathbf{X}) = E(y | \mathbf{X}),
$$ which represents the **minimum mean squared predictor** over all possible functions $f(\mathbf{X})$: $$
\min_f E((y - f(\mathbf{X}))^2).
$$

Under the assumptions A1 and A3, the CEF simplifies to: $$
\mu(\mathbf{X}) = \mathbf{X}\beta.
$$

The **linear projection** is given by: $$
L(y | 1, \mathbf{X}) = \gamma_0 + \mathbf{X}\text{Var}(\mathbf{X})^{-1}\text{Cov}(\mathbf{X}, y),
$$ where: $$
\gamma = \mathbf{X}\text{Var}(\mathbf{X})^{-1}\text{Cov}(\mathbf{X}, y).
$$

This linear projection minimizes the mean squared error: $$
(\gamma_0, \gamma) = \arg\min_{(a, b)} E\left[\left(E(y|\mathbf{X}) - \left(a + \mathbf{X}b\right)\right)^2\right].
$$

Implications for [OLS](#ordinary-least-squares)

-   **Consistency**: OLS is always consistent for the linear projection, ensuring convergence to the true parameter value as $n \to \infty$.
-   **Causal Interpretation**: The linear projection has no inherent causal interpretation---it approximates the conditional mean function.
-   **Assumption Independence**: Unlike the CEF, the linear projection does not depend on assumptions A1 and A3.

------------------------------------------------------------------------

**Evaluating Estimators via Large Sample Properties**

1.  **Consistency**:
    -   Measures the estimator's centrality to the true value.
    -   A consistent estimator ensures that with larger samples, estimates become arbitrarily close to the population parameter.
2.  **Limiting Distribution**:
    -   Helps infer the sampling behavior of the estimator as $n$ grows.
    -   Often approximated by a normal distribution for practical use in hypothesis testing and confidence interval construction.
3.  **Asymptotic Variance**:
    -   Quantifies the dispersion of the estimator around its limiting distribution.
    -   Smaller variance is desirable for greater reliability.

An estimator $\hat{\theta}$ is **consistent** for a parameter $\theta$ if, as the sample size $n$ increases, $\hat{\theta}$ converges in probability to $\theta$:

$$
\hat{\theta}_n \to^p \theta \quad \text{as } n \to \infty.
$$

1.  **Convergence in Probability**:

    -   The probability that $\hat{\theta}$ deviates from $\theta$ by more than a small margin (no matter how small) approaches zero as $n$ increases.

    Formally: $$
    \forall \epsilon > 0, \quad P(|\hat{\theta}_n - \theta| > \epsilon) \to 0 \quad \text{as } n \to \infty.
    $$

2.  **Interpretation**: Consistency ensures that the estimator becomes arbitrarily close to the true population parameter $\theta$ as the sample size grows.

3.  **Asymptotic Behavior**: Large sample properties rely on consistency to provide valid approximations of an estimator's behavior in finite samples.

------------------------------------------------------------------------

**Relationship Between Consistency and Unbiasedness**

-   **Unbiasedness**:
    -   An estimator $\hat{\theta}$ is unbiased if its expected value equals the true parameter: $$
        E(\hat{\theta}) = \theta.
        $$
    -   Unbiasedness is a finite-sample property and does not depend on the sample size.
-   **Consistency**:
    -   Consistency is a large-sample property and requires $\hat{\theta}$ to converge to $\theta$ as $n \to \infty$.

------------------------------------------------------------------------

**Important Distinctions**

1.  **Unbiasedness Does Not Imply Consistency**:
    -   Example: Consider an unbiased estimator with extremely high variance that does not diminish as $n$ increases. Such an estimator does not converge to $\theta$ in probability.
2.  **Consistency Does Not Imply Unbiasedness**:
    -   Example: $\hat{\theta}_n = \frac{n-1}{n}\theta$ is biased for all finite $n$, but as $n \to \infty$, $\hat{\theta}_n \to^p \theta$, making it consistent.

------------------------------------------------------------------------

From the OLS formula: $$
\hat{\beta} = \mathbf{(X'X)^{-1}X'y},
$$ we can expand as: $$
\hat{\beta} = \mathbf{(\sum_{i=1}^{n}x_i'x_i)^{-1} \sum_{i=1}^{n}x_i'y_i},
$$ or equivalently: $$
\hat{\beta} = (n^{-1}\mathbf{\sum_{i=1}^{n}x_i'x_i)^{-1}} n^{-1}\mathbf{\sum_{i=1}^{n}x_i'y_i}.
$$

Taking the probability limit under the assumptions [A2](#a2-full-rank) and [A5](#a5-data-generation-random-sampling), we apply the [Weak Law of Large Numbers] (for a random sample, averages converge to expectations as $n \to \infty$): $$
plim(\hat{\beta}) = plim((n^{-1}\mathbf{\sum_{i=1}^{n}x_i'x_i)^{-1}})plim(n^{-1}\mathbf{\sum_{i=1}^{n}x_i'y_i}),
$$ which simplifies to: $$
\begin{aligned}
plim(\hat{\beta}) 
  &= (E(\mathbf{x_i'x_i}))^{-1}E(\mathbf{x_i'y_i}) & \text{A1}\\
  &= (E(\mathbf{x_i'x_i}))^{-1} \left(E(\mathbf{x_i'x_i} \beta) + E(\mathbf{x_i\epsilon_i})\right) & (A3a) \\
  &= (E(\mathbf{x_i' x_i}))^{-1}E(\mathbf{x_i' x_i}) \beta & (A2)\\
  &= \beta 
\end{aligned}
$$

**Proof:**

For a model of $y_i = x_i \beta + \epsilon_i$, where $\beta$ is the true parameter vector, $\epsilon_i$ is the random error:

Expanding $E(x_i y_i)$:

$$
E(x_i'y_i) = E(x_i'(x_i\beta + \epsilon_i))
$$

By the linearity of expectation:

$$
E(x_i'y_i) = E(x_i'x_i \beta) + E(x_i \epsilon_i)
$$

The second term $E(x_i \epsilon_i) = 0$ under assumption [A3a](#a3a).

Thus, $$
E(x_i'y_i) = E(x_i'x_i)\beta.
$$

------------------------------------------------------------------------

**Consistency of OLS**

Hence, in short, under the assumptions:

-   [A1 Linearity](#a1-linearity)
-   [A2 Full Rank](#a2-full-rank)
-   [A3a: Weaker Exogeneity Assumption](#a3a-weak-exogeneity)
-   [A5 Data Generation (Random Sampling)](#a5-data-generation-random-sampling)

the term $E(\mathbf{x_i'\epsilon_i}) = 0$, and the OLS estimator is **consistent**:

$$
plim(\hat{\beta}) = \beta.
$$

However, [OLS](#ordinary-least-squares) consistency does not guarantee unbiasedness in small samples.

------------------------------------------------------------------------

##### Asymptotic Distribution of OLS

Under the same assumptions :

-   [A1 Linearity](#a1-linearity)
-   [A2 Full Rank](#a2-full-rank)
-   [A3a: Weaker Exogeneity Assumption](#a3a-weak-exogeneity)
-   [A5 Data Generation (Random Sampling)](#a5-data-generation-random-sampling)

and if $\mathbf{x_i'x_i}$ has finite first and second moments (required for the [Central Limit Theorem]), we have:

1.  Convergence of $n^{-1}\sum_{i=1}^n \mathbf{x_i'x_i}$: $$
    n^{-1}\sum_{i=1}^n \mathbf{x_i'x_i} \to^p E(\mathbf{x_i'x_i}).
    $$

2.  Asymptotic normality of $\sqrt{n}(n^{-1}\sum_{i=1}^n \mathbf{x_i'\epsilon_i})$: $$
    \sqrt{n}(n^{-1}\sum_{i=1}^n \mathbf{x_i'\epsilon_i}) \to^d N(0, \mathbf{B}),
    $$ where $\mathbf{B} = Var(\mathbf{x_i'\epsilon_i})$.

From these results, the scaled difference between $\hat{\beta}$ and $\beta$ follows: $$
\sqrt{n}(\hat{\beta} - \beta) = (n^{-1}\sum_{i=1}^n \mathbf{x_i'x_i})^{-1} \sqrt{n}(n^{-1}\sum_{i=1}^n \mathbf{x_i'\epsilon_i}).
$$

By the [Central Limit Theorem]: $$
\sqrt{n}(\hat{\beta} - \beta) \to^d N(0, \Sigma),
$$ where: $$
\Sigma = (E(\mathbf{x_i'x_i}))^{-1} \mathbf{B} (E(\mathbf{x_i'x_i}))^{-1}.
$$

The sandwich form is $\Sigma$ is standard.

**Implications for Homoskedasticity (A4) vs. Heteroskedasticity**

1.  No Homoskedasticity ([A4 Homoskedasticity](#a4-homoskedasticity)) needed:

    -   the CLT and the large-sample distribution of $\hat{\beta}$ do *not* require homoskedasticity. The only place homoskedasticity would simplify things is that $$
        \mathbf{B} = Var(\mathbf{x_i'\epsilon_i}) = \sigma^2 E(\mathbf{x_i'x_i}),
        $$

    only if $Var(\epsilon_i | \mathbf{x}_i) \sigma^2$

    Then $$
    \Sigma = \sigma^2 (E(\mathbf{x_i'x_i}))^{-1}.
    $$

2.  Adjusting for Heteroskedasticity:

    -   In practice, $\sigma_i^2$ can vary with $\mathbf{x}_i$​, leading to heteroskedasticity.
    -   The standard OLS formula for $Var(\hat{\beta})$ is inconsistent under heteroskedasticity, so one uses *robust (White) standard errors*.
    -   Heteroskedasticity can arise from (but not limited to):
        -   Limited dependent variables.
        -   Dependent variables with large or skewed ranges.

------------------------------------------------------------------------

##### Derivation of Asymptotic Variance

The asymptotic variance of the OLS estimator is derived as follows:

$$
\Sigma = (E(\mathbf{x_i'x_i}))^{-1}\mathbf{B}(E(\mathbf{x_i'x_i}))^{-1}.
$$

1.  Substituting $\mathbf{B} = Var(\mathbf{x_i'}\epsilon_i)$: $$
    \Sigma = (E(\mathbf{x_i'x_i}))^{-1}Var(\mathbf{x_i'}\epsilon_i)(E(\mathbf{x_i'x_i}))^{-1}.
    $$

2.  Using the definition of variance: $$
    \Sigma = (E(\mathbf{x_i'x_i}))^{-1}E[(\mathbf{x_i'}\epsilon_i - 0)(\mathbf{x_i'}\epsilon_i - 0)'](E(\mathbf{x_i'x_i}))^{-1}.
    $$

3.  By the [Law of Iterated Expectations] and [A3a: Weaker Exogeneity Assumption](#a3a-weak-exogeneity), we have: $$
    E[(\mathbf{x_i'}\epsilon_i)(\mathbf{x_i'}\epsilon_i)'] = E[E(\epsilon_i^2|\mathbf{x_i})\mathbf{x_i'x_i}],
    $$

4.  Assuming homoskedasticity ([A4 Homoskedasticity](#a4-homoskedasticity)), $E(\epsilon_i^2|\mathbf{x_i}) = \sigma^2$, so: $$
    \Sigma = (E(\mathbf{x_i'x_i}))^{-1}\sigma^2E(\mathbf{x_i'x_i})(E(\mathbf{x_i'x_i}))^{-1}.
    $$

5.  Simplifying: $$
    \Sigma = \sigma^2(E(\mathbf{x_i'x_i}))^{-1}.
    $$

Hence, under the assumptions:

-   [A1 Linearity](#a1-linearity)

-   [A2 Full Rank](#a2-full-rank)

-   [A3a: Weaker Exogeneity Assumption](#a3a-weak-exogeneity)

-   [A4 Homoskedasticity](#a4-homoskedasticity)

-   [A5 Data Generation (Random Sampling)](#a5-data-generation-random-sampling)

we have$$
\sqrt{n}(\hat{\beta} - \beta) \to^d N(0, \sigma^2(E(\mathbf{x_i'x_i}))^{-1}).
$$

The asymptotic variance provides an approximation of the scaled estimator's variance for large $n$. This leads to:

$$
Avar(\sqrt{n}(\hat{\beta} - \beta)) = \sigma^2(E(\mathbf{x_i'x_i}))^{-1}.
$$

The finite sample variance of an estimator can be approximated using the asymptotic variance for large sample sizes:

$$
\begin{aligned}
Avar(\sqrt{n}(\hat{\beta}-\beta)) &\approx Var(\sqrt{n}(\hat{\beta}-\beta)) \\
Avar(\sqrt{n}(\hat{\beta}-\beta))/n &\approx Var(\sqrt{n}(\hat{\beta}-\beta))/n = Var(\hat{\beta})
\end{aligned}
$$

However, it is critical to note that **asymptotic variance** ($Avar(.)$) does not behave in the same manner as finite sample variance ($Var(.)$). This distinction is evident in the following expressions:

$$
\begin{aligned}
Avar(\sqrt{n}(\hat{\beta}-\beta))/n &\neq Avar(\sqrt{n}(\hat{\beta}-\beta)/\sqrt{n}) \\
&\neq Avar(\hat{\beta})
\end{aligned}
$$

Thus, while $Avar(.)$ provides a useful approximation for large samples, its conceptual properties differ from finite sample variance.

------------------------------------------------------------------------

In [Finite Sample Properties], the standard errors are calculated as estimates of the conditional standard deviation:

$$
SE_{fs}(\hat{\beta}_{j-1}) = \sqrt{\hat{Var}(\hat{\beta}_{j-1}|\mathbf{X})} = \sqrt{s^2 \cdot [\mathbf{(X'X)}^{-1}]_{jj}},
$$

where:

-   $s^2$ is the estimator of the error variance,

-   $[\mathbf{(X'X)}^{-1}]_{jj}$ represents the $j$th diagonal element of the inverse design matrix.

In contrast, in [Large Sample Properties], the standard errors are calculated as estimates of the square root of the asymptotic variance:

$$
SE_{ls}(\hat{\beta}_{j-1}) = \sqrt{\hat{Avar}(\sqrt{n} \hat{\beta}_{j-1})/n} = \sqrt{s^2 \cdot [\mathbf{(X'X)}^{-1}]_{jj}}.
$$

Interestingly, the standard error estimator is **identical** for both finite and large samples:

-   The expressions for $SE_{fs}$ and $SE_{ls}$ are mathematically the same.
-   However, they are conceptually estimating two **different quantities**:
    -   **Finite Sample Standard Error**: An estimate of the conditional standard deviation of $\hat{\beta}_{j-1}$ given $\mathbf{X}$.
    -   **Large Sample Standard Error**: An estimate of the square root of the asymptotic variance of $\hat{\beta}_{j-1}$.

The assumptions required for these estimators to be valid differ in their stringency:

1.  **Finite Sample Variance (Conditional Variance):**
    -   Requires stronger assumptions (A1-A5).
2.  **Asymptotic Variance:**
    -   Valid under weaker assumptions ([A1](#a1-linearity), [A2](#a2-full-rank), [A3a](#a3a-weak-exogeneity), [A4](#a4-homoskedasticity), [A5](#a5-data-generation-random-sampling)).

This distinction highlights the utility of asymptotic properties in providing robust approximations when finite sample assumptions may not hold.

------------------------------------------------------------------------

#### Diagnostics

##### Normality of Errors

Ensuring the normality of errors is a critical assumption in many regression models. Deviations from this assumption can impact inference and model interpretation. For diagnoses assessing normality, see [Normality Assessment].

Plots are invaluable for visual inspection of normality. One common approach is the Q-Q plot, which compares the quantiles of the residuals against those of a standard normal distribution:

```{r fig-qqplot-example, fig.cap="Q-Q Plot for Assessing Normality of Sample Data", fig.alt="Q-Q plot comparing sample quantiles to theoretical quantiles. The plot shows data points forming an approximately linear pattern, indicating that the sample distribution is similar to the theoretical distribution. The x-axis represents theoretical quantiles, and the y-axis represents sample quantiles, with values ranging from -2 to 2 on the x-axis and 0 to 100 on the y-axis.", out.width="100%", fig.align="center"}
# Example Q-Q plot
set.seed(123) # For reproducibility
y <- 1:100
x <- rnorm(100) # Generating random normal data
qqplot(x,
       y,
       main = "Q-Q Plot",
       xlab = "Theoretical Quantiles",
       ylab = "Sample Quantiles")
```

##### Influential Observations and Outliers

Identifying influential observations or outliers is essential for robust regression modeling. The **hat matrix** ($\mathbf{H}$) plays a key role in diagnosing influence.

###### Hat Matrix: Outliers in X-Space

The hat matrix is primarily concerned with **leverage**, which reflects how far an observation's predictor values ($X$-space) are from the centroid of the predictor space.

-   **What it measures:** The diagonal elements of the hat matrix quantify **leverage**, not residual size or model fit. High leverage suggests that an observation has an unusual predictor configuration and might disproportionately influence the regression line, irrespective of the response variable.

-   **What it doesn't measure:** It doesn't directly account for outliers in $Y$-space or residuals.

The hat matrix, defined as:

$$
\mathbf{H} = \mathbf{X}(\mathbf{X}'\mathbf{X})^{-1}
$$

has the following properties:

-   **Fitted Values**: $\hat{\mathbf{Y}} = \mathbf{H}\mathbf{Y}$.

-   **Residuals**: $\mathbf{e} = (\mathbf{I} - \mathbf{H})\mathbf{Y}$.

-   **Variance of Residuals**: $\text{var}(\mathbf{e}) = \sigma^2 (\mathbf{I} - \mathbf{H})$.

Diagonal Elements of the Hat Matrix ($h_{ii}$)

-   $h_{ii}$ is the $i$-th element on the main diagonal of $\mathbf{H}$. It must satisfy $0 \leq h_{ii} \leq 1$.

-   $\sum_{i=1}^{n} h_{ii} = p$, where $p$ is the number of parameters (including the intercept).

-   The variance of residuals for observation $i$ is given by $\sigma^2(e_i) = \sigma^2 (1 - h_{ii})$.

-   The covariance between residuals for observations $i$ and $j$ ($i \neq j$) is $-h_{ij}\sigma^2$.

For large datasets, the off-diagonal elements ($h_{ij}$) tend to have small covariance if model assumptions hold.

Estimations Using MSE

-   Variance of residuals: $s^2(e_i) = MSE (1 - h_{ii})$, where $MSE$ is the mean squared error.

-   Covariance of residuals: $\hat{\text{cov}}(e_i, e_j) = -h_{ij}(MSE)$.

Interpretation of $h_{ii}$

If $\mathbf{x}_i = [1, X_{i,1}, \ldots, X_{i,p-1}]'$ represents the vector of predictor values for observation $i$, then:

$$
h_{ii} = \mathbf{x}_i' (\mathbf{X}'\mathbf{X})^{-1} \mathbf{x}_i
$$

The value of $h_{ii}$ depends on the relative positions of the design points $X_{i,1}, \ldots, X_{i,p-1}$. Observations with high $h_{ii}$ are more influential and warrant closer inspection for leverage or outlier behavior.

###### Studentized Residuals: Outliers in Y-Space

Residuals focus on discrepancies between observed ($Y$) and predicted ($\hat{Y}$) values, helping to identify outliers in $Y$-space.

-   **What they measure:**

    -   Standardized or studentized residuals assess how far an observation's response is from the regression line, adjusted for variance.

    -   **Externally studentized residuals** are more robust because they exclude the $i$-th observation when estimating variance.

    -   Large studentized residuals (e.g., $>2$ or $>3$) indicate observations that are unusual in $Y$-space.

-   **What they don't measure:** They do not consider leverage or the $X$-space configuration. A point with a large residual could have low leverage, making it less influential overall.

Studentized residuals, also known as standardized residuals, adjust for the variance of residuals by dividing the residuals by their standard error:

$$
\begin{aligned}
r_i &= \frac{e_i}{s(e_i)} \\
r_i &\sim N(0,1),
\end{aligned}
$$

where $s(e_i) = \sqrt{MSE(1-h_{ii})}$, and $r_i$ accounts for the varying variances of residuals. These residuals allow for a better comparison of model fit across observations.

-   **Semi-Studentized Residuals**: In contrast, the semi-studentized residuals are defined as:

$$
e_i^* = \frac{e_i}{\sqrt{MSE}}
$$

This approach does not adjust for the heterogeneity in variances of residuals, as $e_i^*$ assumes equal variance for all residuals.

------------------------------------------------------------------------

To assess the influence of individual observations, we consider the model without a particular value. When the $i$-th observation is removed, the **deleted residual** is defined as:

$$
\begin{aligned}
d_i &= Y_i - \hat{Y}_{i(i)} \\
    &= \frac{e_i}{1-h_{ii}},
\end{aligned}
$$

where $Y_i$ is the actual observation, and $\hat{Y}_{i(i)}$ is the predicted value for the $i$-th observation, computed using the regression model fitted to the remaining $n-1$ observations. Importantly, we do not need to refit the regression model for each observation to compute $d_i$.

-   As $h_{ii}$ (leverage) increases, $d_i$ also increases, indicating higher influence of the observation.

The variance of the deleted residual is given by:

$$
s^2(d_i) = \frac{MSE_{(i)}}{1-h_{ii}},
$$

where $MSE_{(i)}$ is the mean squared error when the $i$-th case is omitted.

------------------------------------------------------------------------

The **studentized deleted residual** accounts for variability and follows a $t$-distribution with $n-p-1$ degrees of freedom:

$$
t_i = \frac{d_i}{s(d_i)} = \frac{e_i}{\sqrt{MSE_{(i)}(1-h_{ii})}},
$$

where $t_i$ helps identify outliers more effectively.

We can compute $t_i$ without fitting the regression model multiple times. Using the relationship:

$$
(n-p)MSE = (n-p-1)MSE_{(i)} + \frac{e_i^2}{1-h_{ii}},
$$

we derive:

$$
t_i = e_i \sqrt{\frac{n-p-1}{SSE(1-h_{ii}) - e_i^2}}.
$$

This formulation avoids the need for recalculating $MSE_{(i)}$ explicitly for each case.

------------------------------------------------------------------------

Outlying $Y$-observations are those with large studentized deleted residuals in absolute value. To handle multiple testing when there are many residuals, we use a Bonferroni-adjusted critical value:

$$
t_{1-\alpha/2n; n-p-1},
$$

where $\alpha$ is the desired significance level, $n$ is the sample size, and $p$ is the number of parameters in the model. Observations exceeding this threshold are flagged as potential outliers.

```{r}
# Example R Code for Demonstrating Residual Diagnostics
set.seed(123) # For reproducibility

# Simulate some data
n <- 100
x <- rnorm(n, mean = 10, sd = 3)
y <- 5 + 2 * x + rnorm(n, mean = 0, sd = 2)

# Fit a linear regression model
model <- lm(y ~ x)


# Extract residuals, fitted values, and hat values
residuals <- resid(model)
hat_values <- lm.influence(model)$hat
mse <- mean(residuals ^ 2)
n <- length(residuals)
p <- length(coefficients(model)) # Number of parameters

# Compute studentized residuals
studentized_residuals <- residuals / sqrt(mse * (1 - hat_values))

# Compute deleted residuals
deleted_residuals <- residuals / (1 - hat_values)

# Compute studentized deleted residuals
studentized_deleted_residuals <-
    residuals * sqrt((n - p - 1)
                     / (sum(residuals ^ 2) * (1 - hat_values) - residuals ^ 2))

# Flag potential outliers using Bonferroni-adjusted critical value
alpha <- 0.05
bonferroni_threshold <- qt(1 - alpha / (2 * n), df = n - p - 1)
outliers <-
    abs(studentized_deleted_residuals) > bonferroni_threshold

# Print results
results <- data.frame(
    Residuals                     = residuals,
    Hat_Values                    = hat_values,
    Studentized_Residuals         = studentized_residuals,
    Deleted_Residuals             = deleted_residuals,
    Studentized_Deleted_Residuals = studentized_deleted_residuals,
    Outlier                       = outliers
)

causalverse::nice_tab(head(results))

```

```{r fig-studentized-residuals, fig.cap="Studentized Deleted Residuals for Outlier Detection", fig.alt="Scatter plot titled 'Studentized Deleted Residuals' with residuals on the y-axis and observations on the x-axis. Most points fall between -2 and 2, indicating normal residual behavior. A single red point above the threshold near 3 suggests a potential outlier. A horizontal blue dashed line marks the detection threshold. The legend identifies 'Potential Outliers' in red and 'Threshold' in blue.", out.width="100%", fig.align="center"}
# Plot studentized deleted residuals for visualization
plot(
    studentized_deleted_residuals,
    main = "Studentized Deleted Residuals",
    xlab = "Observation",
    ylab = "Studentized Deleted Residuals",
    pch = 16,
    col = ifelse(outliers, "red", "black")
)
abline(
    h = c(-bonferroni_threshold, bonferroni_threshold),
    col = "blue",
    lty = 2
)
legend(
    "topright",
    legend = c("Potential Outliers", "Threshold"),
    col = c("red", "blue"),
    pch = c(16, NA),
    lty = c(NA, 2)
)
```

------------------------------------------------------------------------

##### Identifying Influential Cases

By **influential**, we refer to observations whose exclusion causes major changes in the fitted regression model. Note that not all outliers are influential.

Types of Influence Measures

-   **Influence on Single Fitted Values**: [DFFITS](#dffits)
-   **Influence on All Fitted Values**: [Cook's D](#cooks-d)
-   **Influence on the Regression Coefficients**: [DFBETAS](#dfbetas)

Measures like Cook's D, DFFITS, and DFBETAS combine leverage (from the hat matrix) and residual size (from studentized residuals) to assess the **influence** of an observation on the model as a whole. Hence, these effectively combine impact of $X$-space and $Y$-space.

###### DFFITS {#dffits}

DFFITS measures the **influence on single fitted values**. It is defined as:

$$
\begin{aligned}
(DFFITS)_i &= \frac{\hat{Y}_i - \hat{Y}_{i(i)}}{\sqrt{MSE_{(i)}h_{ii}}} \\
&= t_i \left(\frac{h_{ii}}{1-h_{ii}}\right)^{1/2}
\end{aligned}
$$

Where:

-   $\hat{Y}_i$ = fitted value for observation $i$ using all data.
-   $\hat{Y}_{i(i)}$ = fitted value for observation $i$ with the $i$th case removed.
-   $MSE_{(i)}$ = mean squared error with observation $i$ excluded.
-   $h_{ii}$ = leverage of the $i$th observation.
-   $t_i$ = studentized deleted residual.

High DFFITS values occur when leverage and residuals are jointly significant.

DFFITS captures the **standardized difference between the fitted value for observation** $i$ with and without the $i$th case in the model. It is a product of:

1.  The **studentized deleted residual**.
2.  A scaling factor based on the leverage of the $i$th observation, $h_{ii}$.

An observation is considered influential based on DFFITS if:

-   **Small to medium data sets**: $|DFFITS| > 1$
-   **Large data sets**: $|DFFITS| > 2 \sqrt{p/n}$

Where:

-   $p$ = number of predictors (including the intercept).
-   $n$ = total number of observations.

This provides a practical threshold for detecting influential observations in different dataset sizes.

```{r}
# Load necessary package
library(car)

# Fit a linear model
model <- lm(mpg ~ hp + wt, data = mtcars)

# Compute DFFITS
dffits_values <- dffits(model)

# Display influential observations based on the threshold for a large dataset
threshold <- 2 * sqrt(length(coefficients(model)) / nrow(mtcars))
influential_obs <- which(abs(dffits_values) > threshold)

# Results
list(
    DFFITS                   = dffits_values,
    Threshold                = threshold,
    Influential_Observations = influential_obs
)
```

###### Cook's D {#cooks-d}

**Cook's D** measures the **influence of the** $i$th case on all fitted values in a regression model. It is defined as:

$$
\begin{aligned}
D_i &= \frac{\sum_{j=1}^{n}(\hat{Y}_j - \hat{Y}_{j(i)})^2}{p(MSE)} \\
&= \frac{e^2_i}{p(MSE)}\left(\frac{h_{ii}}{(1-h_{ii})^2}\right)
\end{aligned}
$$

Where:

-   $\hat{Y}_j$ = fitted value for observation $j$ using all data.
-   $\hat{Y}_{j(i)}$ = fitted value for observation $j$ with the $i$th case removed.
-   $e_i$ = residual for observation $i$.
-   $h_{ii}$ = leverage of the $i$th observation.
-   $p$ = number of predictors (including the intercept).
-   $MSE$ = mean squared error of the model.

Key Insights

-   **Cook's D quantifies the overall influence** of the $i$th observation on the entire set of fitted values.
-   If either the residual $e_i$ increases or the leverage $h_{ii}$ increases, then $D_i$ also increases, indicating higher influence.
-   Observations with both high leverage and large residuals are flagged as influential.

Threshold for Influence

-   $D_i$ can be interpreted as a percentile of an $F_{(p,n-p)}$ distribution.
-   Practical thresholds:
    -   **If** $D_i > 4/n$, the $i$th case has major influence.
    -   Alternatively, cases where $D_i$ exceeds the 50th percentile of the $F$-distribution may also be considered influential.

```{r}
# Load necessary package
library(car)

# Fit a linear model
model <- lm(mpg ~ hp + wt, data = mtcars)

# Compute Cook's D
cooks_d_values <- cooks.distance(model)

# Display influential observations based on the threshold
threshold <- 4 / nrow(mtcars)
influential_obs <- which(cooks_d_values > threshold)

# Results
list(
  Cooks_D                  = cooks_d_values,
  Threshold                = threshold,
  Influential_Observations = influential_obs
)
```

###### DFBETAS {#dfbetas}

**DFBETAS** measures the **influence of the** $i$th observation on each regression coefficient in a regression model. It is defined as:

$$
(DFBETAS)_{k(i)} = \frac{b_k - b_{k(i)}}{\sqrt{MSE_{(i)}c_{kk}}}
$$

Where:

-   $b_k$ = regression coefficient for the $k$th predictor using all observations.
-   $b_{k(i)}$ = regression coefficient for the $k$th predictor with the $i$th observation omitted.
-   $MSE_{(i)}$ = mean squared error with the $i$th observation excluded.
-   $c_{kk}$ = $k$th diagonal element of $\mathbf{X'X}^{-1}$, representing the variance of $b_k$.

Key Insights

-   **DFBETAS quantifies the change in each regression coefficient (**$b_k$) caused by omitting the $i$th observation.
-   The sign of DFBETAS indicates whether the inclusion of an observation increases or decreases the regression coefficient.
    -   Positive DFBETAS: Inclusion increases $b_k$.
    -   Negative DFBETAS: Inclusion decreases $b_k$.
-   High DFBETAS indicate that a single observation disproportionately affects one or more predictors.

The thresholds for identifying influential observations based on DFBETAS are:

-   **Small data sets**: $|DFBETAS| > 1$
-   **Large data sets**: $|DFBETAS| > 2 / \sqrt{n}$

Where $n$ is the total number of observations.

```{r}
# Load necessary package
library(car)

# Fit a linear model
model <- lm(mpg ~ hp + wt, data = mtcars)

# Compute DFBETAS
dfbetas_values <- dfbetas(model)

# Display influential observations based on the threshold for each predictor
threshold <- 2 / sqrt(nrow(mtcars))
influential_obs <-
    apply(dfbetas_values, 2, function(x)
        which(abs(x) > threshold))

# Results
list(
    DFBETAS                  = dfbetas_values,
    Threshold                = threshold,
    Influential_Observations = influential_obs
)
```

##### Collinearity

**Collinearity** (or multicollinearity) refers to the correlation among explanatory variables in a regression model. It can lead to various issues, including:

-   Large changes in the estimated regression coefficients when a predictor variable is added or removed, or when observations are altered.
-   Non-significant results for individual tests on regression coefficients of important predictor variables.
-   Regression coefficients with signs opposite to theoretical expectations or prior experience.
-   Large coefficients of simple correlation between pairs of predictor variables in the correlation matrix.
-   Wide confidence intervals for the regression coefficients representing important predictor variables.

When some $X$ variables are highly correlated, the inverse $(X'X)^{-1}$ either does not exist or is computationally unstable. This can result in:

1.  **Non-interpretability of parameters**: $$\mathbf{b = (X'X)^{-1}X'y}$$
2.  **Infinite sampling variability**: $$\mathbf{s^2(b) = MSE (X'X)^{-1}}$$

If some predictor variables ($X$) are "perfectly" correlated, the system becomes undetermined, leading to an infinite number of models that fit the data. Specifically:

-   If $X'X$ is singular, then $(X'X)^{-1}$ does not exist.
-   This results in poor parameter estimation and invalid statistical inference.

------------------------------------------------------------------------

###### Variance Inflation Factors (VIFs)

The **Variance Inflation Factor (VIF)** quantifies how much the variance of a regression coefficient is inflated due to multicollinearity. It is defined as:

$$
VIF_k = \frac{1}{1-R^2_k}
$$

Where:

-   $R^2_k$ is the coefficient of multiple determination when $X_k$ is regressed on the other $p-2$ predictor variables in the model.

------------------------------------------------------------------------

**Interpretation of VIFs**

-   Large $VIF_k$ values indicate that near collinearity is inflating the variance of $b_k$. The relationship is given by: $$ var(b_k) \propto \sigma^2 (VIF_k) $$
-   **Thresholds**:
    -   $VIF > 4$: Investigate the cause of multicollinearity.
    -   $VIF_k > 10$: Serious multicollinearity problem that can lead to poor parameter estimates.
-   The **mean VIF** provides an estimate of the degree of multicollinearity:
    -   If $avg(VIF) >> 1$, serious multicollinearity is present.
-   **Multicollinearity and VIF:**
    -   High VIFs with indicator variables are normal and not problematic.

    -   VIF is generally not useful for detecting multicollinearity concerns in models with fixed effects.
-   **Overemphasis on Multicollinearity:**
    -   Multicollinearity inflates standard errors and widens confidence intervals but does not bias results.

    -   If key variables have narrow confidence intervals, multicollinearity is not an issue.
-   **Goldberger's Insight** [@goldberger1991course]**:**
    -   Multicollinearity is akin to small sample size ("micronumerosity").

    -   Large standard errors are expected with highly correlated independent variables.
-   **Practical Implications:**
    -   Evaluate whether confidence intervals for key variables are sufficiently narrow.

    -   If not, the study is inconclusive, and a larger dataset or redesigned study is needed.

------------------------------------------------------------------------

```{r}
# Load necessary packages
library(car)

# Fit a regression model
model <- lm(mpg ~ hp + wt + disp, data = mtcars)

# Compute Variance Inflation Factors
vif_values <- vif(model)

# Check for high multicollinearity
threshold <- 10
high_vif <- which(vif_values > threshold)

# Results
list(
  VIFs = vif_values,
  High_VIF_Threshold = threshold,
  High_VIF_Indices = high_vif
)
```

###### Condition Number

**Condition Number** is a diagnostic measure for detecting multicollinearity, derived from the spectral decomposition of the matrix $\mathbf{X'X}$.

The spectral decomposition of $\mathbf{X'X}$ is:

$$
\mathbf{X'X}= \sum_{i=1}^{p} \lambda_i \mathbf{u_i u_i'}
$$

Where:

-   $\lambda_i$: Eigenvalue associated with the $i$th eigenvector.
-   $\mathbf{u}_i$: Eigenvector associated with $\lambda_i$.
-   $\lambda_1 > \lambda_2 > \dots > \lambda_p$ (ordered eigenvalues).
-   The eigenvectors are orthogonal: $$ 
    \begin{cases}
    \mathbf{u_i'u_j} = 0 & \text{for } i \neq j \\
    \mathbf{u_i'u_j} = 1 & \text{for } i = j
    \end{cases}
    $$

------------------------------------------------------------------------

Definition of the Condition Number

The **Condition Number** is defined as:

$$
k = \sqrt{\frac{\lambda_{\text{max}}}{\lambda_{\text{min}}}}
$$

Where:

-   $\lambda_{\text{max}}$: Largest eigenvalue.
-   $\lambda_{\text{min}}$: Smallest eigenvalue.

Interpretation

-   $k > 30$: Cause for concern.
-   $30 < k < 100$: Moderate dependencies among predictors.
-   $k > 100$: Strong collinearity, indicating serious multicollinearity.

------------------------------------------------------------------------

The **Condition Index** for the $i$th eigenvalue is defined as:

$$
\delta_i = \sqrt{\frac{\lambda_{\text{max}}}{\lambda_i}}
$$

Where $i = 1, \dots, p$.

------------------------------------------------------------------------

**Variance proportions** can be used to identify collinearity issues. The proportion of total variance associated with the $k$th regression coefficient and the $i$th eigen mode is given by:

$$
\frac{u_{ik}^2/\lambda_i}{\sum_j \left(u^2_{jk}/\lambda_j\right)}
$$

**Key Indicators**:

-   A **large condition index** $\delta_i$ suggests potential collinearity.
-   Variance proportions $> 0.5$ for at least two regression coefficients indicate serious collinearity problems.

------------------------------------------------------------------------

```{r}
# Load necessary package
library(car)

# Fit a regression model
model <- lm(mpg ~ hp + wt + disp, data = mtcars)

# Compute eigenvalues and eigenvectors of the correlation matrix of predictors
cor_matrix <- cor(mtcars[, c("hp", "wt", "disp")])
eigen_decomp <- eigen(cor_matrix)

# Extract eigenvalues and compute the Condition Number
eigenvalues <- eigen_decomp$values
condition_number <- sqrt(max(eigenvalues) / min(eigenvalues))

# Compute Condition Indices
condition_indices <- sqrt(max(eigenvalues) / eigenvalues)

# Results
list(
  Condition_Number = condition_number,
  Condition_Indices = condition_indices
)
```

1.  **Condition Number**: A single value indicating the degree of multicollinearity.
2.  **Condition Indices**: A vector showing the relative dependency associated with each eigenvalue.

##### Constancy of Error Variance

Testing for the constancy of error variance (homoscedasticity) ensures that the assumptions of linear regression are met.

Below are two commonly used tests to assess error variance.

------------------------------------------------------------------------

###### Brown-Forsythe Test (Modified Levene Test) {#brown-forsythe-test-modified-levene-test}

The **Brown-Forsythe Test** does not depend on the normality of errors and is suitable when error variance increases or decreases with $X$.

Procedure

1.  Split the residuals into two groups: $$ e_{i1}, i = 1, \dots, n_1 \quad \text{and} \quad e_{j2}, j = 1, \dots, n_2 $$

2.  Compute absolute deviations from the group medians: $$ d_{i1} = |e_{i1} - \tilde{e}_1| \quad \text{and} \quad d_{j2} = |e_{j2} - \tilde{e}_2| $$ where $\tilde{e}_1$ and $\tilde{e}_2$ are the medians of groups 1 and 2, respectively.

3.  Perform a two-sample t-test on $d_{i1}$ and $d_{j2}$: $$
    t_L = \frac{\bar{d}_1 - \bar{d}_2}{s\sqrt{\frac{1}{n_1} + \frac{1}{n_2}}}
    $$ where $$
    s^2 = \frac{\sum_i (d_{i1} - \bar{d}_1)^2 + \sum_j (d_{j2} - \bar{d}_2)^2}{n - 2}
    $$

4.  Reject the null hypothesis of constant error variance if: $$ |t_L| > t_{1-\alpha/2; n-2} $$

```{r}
# Load necessary package
library(car)

# Fit a linear model
model <- lm(mpg ~ hp + wt, data = mtcars)

# Perform the Brown-Forsythe Test
levene_test <- leveneTest(model$residuals ~ cut(mtcars$hp, 2))  # Split HP into 2 groups

# Results
levene_test
```

The p-value determines whether to reject the null hypothesis of constant variance.

------------------------------------------------------------------------

###### Breusch-Pagan Test (Cook-Weisberg Test) {#breusch-pagan-test-cook-weisberg-test}

The **Breusch-Pagan Test** assumes independent and normally distributed errors. It tests the hypothesis:

-   $H_0: \gamma_1 = 0$ (Constant variance).
-   $H_1: \gamma_1 \neq 0$ (Non-constant variance).

Procedure

1.  Assume the variance of the error terms is modeled as: $$ \sigma^2_i = \gamma_0 + \gamma_1 X_i $$

2.  Regress the squared residuals ($e_i^2$) on $X_i$:

    -   Obtain the **regression sum of squares** ($SSR^*$).

3.  Compute the Breusch-Pagan statistic: $$
    X^2_{BP} = \frac{SSR^*/2}{\left(\frac{SSE}{n}\right)^2}
    $$ where $SSE$ is the error sum of squares from the regression of $Y$ on $X$.

4.  Compare $X^2_{BP}$ to the critical value from the $\chi^2$ distribution with 1 degree of freedom:

    -   Reject $H_0$ (homogeneous variance) if: $$ X^2_{BP} > \chi^2_{1-\alpha;1} $$

```{r}
# Load necessary package
library(lmtest)

# Perform the Breusch-Pagan Test
bp_test <- bptest(model)

# Results
bp_test
```

If the p-value is below the chosen significance level (e.g., 0.05), reject \$H_0\$ and conclude non-constant variance.

##### Independence

Testing for independence ensures that the residuals of a regression model are uncorrelated. Violation of this assumption may lead to biased or inefficient parameter estimates. Below, we discuss three primary methods for diagnosing dependence: plots, the Durbin-Watson test, and specific methods for time-series and spatial data.

------------------------------------------------------------------------

###### Plots

A residual plot can help detect dependence in the residuals:

1.  Plot residuals ($e_i$) versus the predicted values ($\hat{Y}_i$).
2.  Patterns such as systematic waves, trends, or clusters indicate possible dependence.
3.  Independence is suggested if residuals are randomly scattered without clear patterns.

```{r fig-residual-plot, fig.cap="Residual Plot for Linear Model Diagnostics", fig.alt="Scatter plot titled 'Residual Plot' with residuals on the y-axis and fitted values on the x-axis. The data points are scattered around a horizontal red line at zero, illustrating the distribution of residuals. The x-axis ranges approximately from 10 to 30, and the y-axis from -4 to 6. This plot is used to assess the adequacy of model fit and check for patterns such as heteroscedasticity or non-linearity.", out.width="100%", fig.align="center"}
# Fit a regression model
model <- lm(mpg ~ hp + wt, data = mtcars)

# Residual plot
plot(model$fitted.values, model$residuals,
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residual Plot")
abline(h = 0, col = "red")
```

###### Durbin-Watson Test

The **Durbin-Watson test** specifically detects autocorrelation in residuals, especially in time-series data.

-   **Hypotheses**:

    -   $H_0$: Residuals are uncorrelated.

    -   $H_1$: Residuals are autocorrelated.

-   The Durbin-Watson statistic ($d$) is calculated as: $d = \frac{\sum_{t=2}^n (e_t - e_{t-1})^2}{\sum_{t=1}^n e_t^2}$

    -   $d \approx 2$: No autocorrelation.

    -   $d < 2$: Positive autocorrelation.

    -   $d > 2$: Negative autocorrelation.

```{r}
# Load necessary package
library(lmtest)

# Perform the Durbin-Watson test
dw_test <- dwtest(model)

# Results
dw_test
```

###### Time-Series Autocorrelation

For time-series data, autocorrelation often occurs due to the temporal structure. Key diagnostics include:

1.  **Autocorrelation Function (ACF)**:

    -   Shows the correlation of residuals with their lagged values.

    -   Significant spikes in the ACF plot indicate autocorrelation.

2.  **Partial Autocorrelation Function (PACF)**:

    -   Identifies the correlation of residuals after removing the influence of intermediate lags.

```{r fig-acf-residuals, fig.cap="ACF of Residuals", fig.alt="Autocorrelation function (ACF) plot of residuals with lag values on the x-axis and autocorrelation coefficients on the y-axis. The plot displays vertical bars at each lag, with most bars within the blue dashed confidence bounds. A large spike at lag 0 is visible, and smaller values appear at higher lags, indicating little autocorrelation beyond lag 1. The plot is used to assess independence of residuals.", out.width="100%", fig.align='center'}
# Load necessary package
library(forecast)

# Residuals from a time-series model
time_series_res <- ts(model$residuals)

# Plot ACF
acf(time_series_res, main = "ACF of Residuals")
```

```{r fig-pacf-residuals, fig.cap="PACF of Residuals", fig.alt="Partial autocorrelation function (PACF) plot of residuals showing lags on the x-axis and partial autocorrelation coefficients on the y-axis. Most bars fall within the horizontal blue dashed confidence bounds, except for a notable spike at lag 1 and minor spikes at lags 4 and 8. The plot is used to detect the presence of partial autocorrelation in residuals after fitting a model.", out.width="100%", fig.align='center'}
# Plot PACF
pacf(time_series_res, main = "PACF of Residuals")
```

###### Spatial Statistics

Spatial dependence occurs when residuals are correlated across geographical or spatial locations. Two primary tests are used to diagnose spatial autocorrelation:

------------------------------------------------------------------------

**Moran's I** measures global spatial autocorrelation. It determines whether similar values cluster spatially. The statistic is defined as:

$$
I = \frac{n}{W} \cdot \frac{\sum_i \sum_j w_{ij} (x_i - \bar{x})(x_j - \bar{x})}{\sum_i (x_i - \bar{x})^2}
$$

Where:

-   $n$: Number of observations.
-   $W$: Sum of all spatial weights $w_{ij}$.
-   $w_{ij}$: Spatial weight between locations $i$ and $j$.
-   $x_i$: Residual value at location $i$.
-   $\bar{x}$: Mean of the residuals.

Spatial Weight Matrix ($W$): $W$ defines the spatial relationship between observations. It is often derived from:

-   **Distance-based methods**: E.g., k-nearest neighbors or distance bands.
-   **Adjacency methods**: Based on shared boundaries in geographic data.

**Interpretation**:

-   $I > 0$: Positive spatial autocorrelation (similar values cluster together).

-   $I < 0$: Negative spatial autocorrelation (dissimilar values are neighbors).

-   $I \approx 0$: Random spatial distribution.

```{r}
# Load necessary packages
library(sf)
library(spdep)

# Simulate spatial data (example with mtcars dataset)
coords <- cbind(mtcars$hp, mtcars$wt)  # Coordinates based on two variables

# Add small jitter to avoid duplicate coordinates
set.seed(123)  # For reproducibility
coords_jittered <-
    coords + matrix(runif(length(coords),-0.01, 0.01), ncol = 2)

# Find nearest neighbors
neighbors <- knn2nb(knearneigh(coords_jittered, k = 3))

# Create spatial weights matrix
weights <- nb2listw(neighbors, style = "W")  # Row-standardized weights

# Fit the linear model
model <- lm(mpg ~ hp + wt, data = mtcars)

# Check lengths of residuals and weights
length(model$residuals)  # Should be 32
length(weights$neighbours)  # Should also be 32

# Compute Moran's I for residuals
moran_test <- moran.test(model$residuals, weights)
print(moran_test)
```

```{r fig-moran-scatterplot, fig.cap="Moran's Scatterplot", fig.alt="Scatterplot titled 'Moran's Scatterplot' showing the relationship between raw model residuals on the x-axis and spatially lagged residuals on the y-axis. A fitted diagonal line indicates the degree of spatial autocorrelation. Most points are clustered around the origin, with a few labeled outliers such as points 10, 19, 23, and 26. Dashed vertical and horizontal lines represent zero on each axis. The plot is used to assess spatial dependence in model residuals.", out.width="100%", fig.align='center'}
# Moran's scatterplot
moran.plot(model$residuals, weights, main = "Moran's Scatterplot")
```

Significant Moran's I: Indicates global clustering of similar residuals, suggesting spatial dependence.

------------------------------------------------------------------------

**Geary's C** measures spatial autocorrelation at a local level, emphasizing differences between neighboring observations. The statistic is defined as:

$$
C = \frac{n - 1}{2W} \cdot \frac{\sum_i \sum_j w_{ij} (x_i - x_j)^2}{\sum_i (x_i - \bar{x})^2}
$$

Where:

-   $C$ ranges from 0 to 2:
    -   $C \approx 0$: High positive spatial autocorrelation.
    -   $C \approx 1$: Spatial randomness.
    -   $C \approx 2$: High negative spatial autocorrelation.

**Comparison of Moran's I and Geary's C**:

-   Moran's I is more global and measures the overall pattern of autocorrelation.

-   Geary's C is more sensitive to local spatial autocorrelation, detecting small-scale patterns.

------------------------------------------------------------------------

```{r}
# Compute Geary's C for residuals
geary_test <- geary.test(model$residuals, weights)

# Results
geary_test
```

Significant Geary's C: Highlights local spatial autocorrelation, useful for identifying specific regions or groups of observations where dependence is strong.

------------------------------------------------------------------------
