### Derivative-Free {#sec-derivative-free}

#### Secant Method {#secant-method}

The Secant Method is a root-finding algorithm that approximates the derivative using finite differences, making it a derivative-free alternative to Newton's method. It is particularly useful when the exact gradient (or Jacobian in the case of optimization problems) is unavailable or expensive to compute.

In nonlinear optimization, we apply the Secant Method to iteratively refine parameter estimates without explicitly computing second-order derivatives.

------------------------------------------------------------------------

In one dimension, the Secant Method approximates the derivative as:

$$
f'(\theta) \approx \frac{f(\theta_{j}) - f(\theta_{j-1})}{\theta_{j} - \theta_{j-1}}.
$$

Using this approximation, the **update step** in the Secant Method follows:

$$
\theta_{j+1} = \theta_j - f(\theta_j) \frac{\theta_j - \theta_{j-1}}{f(\theta_j) - f(\theta_{j-1})}.
$$

Instead of computing the exact derivative (as in Newton's method), we use the difference between the last two iterates to approximate it. This makes the Secant Method **more efficient** in cases where gradient computation is expensive or infeasible.

------------------------------------------------------------------------

In **higher dimensions**, the Secant Method extends to an approximate [Quasi-Newton Method], often referred to as **Broyden's Method**. We iteratively approximate the inverse Hessian matrix using past updates.

The update formula for a vector-valued function $F(\theta)$ is:

$$
\theta^{(j+1)} = \theta^{(j)} - \mathbf{B}^{(j)} F(\theta^{(j)}),
$$

where $\mathbf{B}^{(j)}$ is an approximation of the inverse Jacobian matrix, updated at each step using:

$$
\mathbf{B}^{(j+1)} = \mathbf{B}^{(j)} + \frac{(\Delta \theta^{(j)} - \mathbf{B}^{(j)} \Delta F^{(j)}) (\Delta \theta^{(j)})'}{(\Delta \theta^{(j)})' \Delta F^{(j)}},
$$

where:

-   $\Delta \theta^{(j)} = \theta^{(j+1)} - \theta^{(j)}$,
-   $\Delta F^{(j)} = F(\theta^{(j+1)}) - F(\theta^{(j)})$.

This **secant-based update** approximates the behavior of the **true Jacobian inverse**, reducing computational cost compared to full Newton's method.

------------------------------------------------------------------------

**Algorithm: Secant Method for Nonlinear Optimization**

The **Secant Method** for nonlinear optimization follows these steps:

1.  Initialize parameters $\theta^{(0)}$ and $\theta^{(1)}$ (two starting points).
2.  Compute the function values $F(\theta^{(0)})$ and $F(\theta^{(1)})$.
3.  Use the Secant update formula to compute the next iterate $\theta^{(j+1)}$.
4.  Update the approximate inverse Jacobian $\mathbf{B}^{(j)}$.
5.  Repeat until convergence.

------------------------------------------------------------------------

```{r}
# Load required libraries
library(numDeriv)


# Define a nonlinear function (logistic model)
nonlinear_model <- function(theta, x) {
    return(theta[1] / (1 + exp(-theta[2] * (x - theta[3]))))
}

# Define the Sum of Squared Errors (SSE) function
sse <- function(theta, x, y) {
    return(sum((y - nonlinear_model(theta, x)) ^ 2))
}

# Generate synthetic data
set.seed(123)
x <- seq(-5, 5, length.out = 100)
true_theta <- c(4, 1.5, 0)  # True parameters (A, B, C)
y <- nonlinear_model(true_theta, x) + rnorm(length(x), sd = 0.3)

# Improved Secant Method with Line Search
secant_method_improved <-
    function(theta0,
             theta1,
             x,
             y,
             tol = 1e-6,
             max_iter = 100) {
        theta_prev <- as.matrix(theta0)  # Convert to column vector
        theta_curr <- as.matrix(theta1)
        
        alpha <- 1  # Initial step size
        step_reduction_factor <- 0.5  # Reduce step if SSE increases
        B_inv <- diag(length(theta0))  # Identity matrix initialization
        
        for (j in 1:max_iter) {
            # Compute function values using numerical gradient
            F_prev <-
                as.matrix(grad(function(theta)
                    sse(theta, x, y), theta_prev))
            F_curr <-
                as.matrix(grad(function(theta)
                    sse(theta, x, y), theta_curr))
            
            # Compute secant step update (convert to column vectors)
            delta_theta <- as.matrix(theta_curr - theta_prev)
            delta_F <- as.matrix(F_curr - F_prev)
            
            # Prevent division by zero
            if (sum(abs(delta_F)) < 1e-8) {
                cat("Gradient diff is too small, stopping optimization.\n")
                break
            }
            
            # Ensure correct dimensions for Broyden update
            numerator <-
                (delta_theta - B_inv %*% delta_F) %*% t(delta_theta)
            # Convert scalar to numeric
            denominator <-
                as.numeric(t(delta_theta) %*% delta_F)  
            
            # Updated inverse Jacobian approximation
            B_inv <-
                B_inv + numerator / denominator  
            
            # Compute next theta using secant update
            theta_next <- theta_curr - alpha * B_inv %*% F_curr
            
            # Line search: Reduce step size if SSE increases
            while (sse(as.vector(theta_next), x, y) > sse(as.vector(theta_curr), 
                                                          x, y)) {
                alpha <- alpha * step_reduction_factor
                theta_next <- theta_curr - alpha * B_inv %*% F_curr
                
                # Print progress
                # cat("Reducing step size to", alpha, "\n")
            }
            
            # Print intermediate results for debugging
            cat(
                sprintf(
                    "Iteration %d: Theta = (%.4f, %.4f, %.4f), Alpha = %.4f\n",
                    j,
                    theta_next[1],
                    theta_next[2],
                    theta_next[3],
                    alpha
                )
            )
            
            # Check convergence
            if (sum(abs(theta_next - theta_curr)) < tol) {
                cat("Converged successfully.\n")
                break
            }
            
            # Update iterates
            theta_prev <- theta_curr
            theta_curr <- theta_next
        }
        
        return(as.vector(theta_curr))  # Convert back to numeric vector
    }

# Adjusted initial parameter guesses
theta0 <- c(2, 0.8,-1)   # Closer to true parameters
theta1 <- c(4, 1.2,-0.5)  # Slightly refined

# Run Improved Secant Method
estimated_theta <- secant_method_improved(theta0, theta1, x, y)

# Display estimated parameters
cat("Estimated parameters (A, B, C) using Secant Method:\n")
print(estimated_theta)
```

```{r fig-secant-method, fig.cap="Secant Method", fig.alt="Scatter plot titled 'Secant Method' showing data points and a fitted curve. The x-axis is labeled 'x' and the y-axis is labeled 'y.' Black dots represent data points, and a red line represents the fitted curve. A legend indicates 'Data' with a black dot and 'Fitted Curve' with a red line. The curve shows a sigmoidal trend, increasing from left to right.", out.width="100%", fig.align='center'}
# Plot data and fitted curve
plot(
    x,
    y,
    main = "Secant Method: Data & Fitted Curve",
    pch = 19,
    cex = 0.5,
    xlab = "x",
    ylab = "y"
)
curve(
    nonlinear_model(estimated_theta, x),
    from = -5,
    to = 5,
    add = TRUE,
    col = "red",
    lwd = 2
)
legend(
    "topleft",
    legend = c("Data", "Fitted Curve"),
    pch = c(19, NA),
    lty = c(NA, 1),
    col = c("black", "red")
)
```

#### Grid Search {#grid-search}

Grid Search (GS) is a brute-force optimization method that systematically explores a grid of possible parameter values to identify the combination that minimizes the residual sum of squares (RSS). Unlike gradient-based optimization, which moves iteratively towards a minimum, grid search evaluates all predefined parameter combinations, making it robust but computationally expensive.

Grid search is particularly useful when:

-   **The function is highly nonlinear**, making gradient methods unreliable.

-   **The parameter space is small**, allowing exhaustive search.

-   **A global minimum is needed**, and prior knowledge about parameter ranges exists.

------------------------------------------------------------------------

The goal of Grid Search is to find an optimal parameter set $\hat{\theta}$ that minimizes the Sum of Squared Errors (SSE):

$$
\hat{\theta} = \arg\min_{\theta \in \Theta} SSE(\theta),
$$

where:

$$
SSE(\theta) = \sum_{i=1}^{n} (y_i - f(x_i; \theta))^2.
$$

Grid search **discretizes** the search space $\Theta$ into a finite set of candidate values for each parameter:

$$
\Theta = \theta_1 \times \theta_2 \times \dots \times \theta_p.
$$

The **accuracy** of the solution depends on the **grid resolution**---a finer grid leads to **better accuracy** but **higher computational cost**.

------------------------------------------------------------------------

**Grid Search Algorithm**

1.  Define a grid of possible values for each parameter.
2.  Evaluate SSE for each combination of parameters.
3.  Select the parameter set that minimizes SSE.

------------------------------------------------------------------------

```{r}
# Define a numerically stable logistic function
safe_exp <- function(x) {
    return(ifelse(x > 700, Inf, exp(pmin(x, 700))))  # Prevent overflow
}

# Define the logistic growth model
nonlinear_model <- function(theta, x) {
    return(theta[1] / (1 + safe_exp(-theta[2] * (x - theta[3]))))
}

# Define the Sum of Squared Errors (SSE) function
sse <- function(theta, x, y) {
    predictions <- nonlinear_model(theta, x)
    return(sum((y - predictions)^2))
}

# Grid Search Optimization
grid_search_optimization <- function(x, y, grid_size = 10) {
    # Define grid of parameter values
    A_values <- seq(2, 6, length.out = grid_size)
    B_values <- seq(0.5, 3, length.out = grid_size)
    C_values <- seq(-2, 2, length.out = grid_size)
    
    # Generate all combinations of parameters
    param_grid <- expand.grid(A = A_values, B = B_values, C = C_values)
    
    # Evaluate SSE for each parameter combination
    param_grid$SSE <- apply(param_grid, 1, function(theta) {
        sse(as.numeric(theta[1:3]), x, y)
    })
    
    # Select the best parameter set
    best_params <- param_grid[which.min(param_grid$SSE), 1:3]
    return(as.numeric(best_params))
}

# Generate synthetic data
set.seed(123)
x <- seq(-5, 5, length.out = 100)
true_theta <- c(4, 1.5, 0)  # True parameters (A, B, C)
y <- nonlinear_model(true_theta, x) + rnorm(length(x), sd = 0.3)

# Run Grid Search
estimated_theta <- grid_search_optimization(x, y, grid_size = 20)

# Display results
cat("Estimated parameters (A, B, C) using Grid Search:\n")
print(estimated_theta)
```

```{r fig-grid-search, fig.cap="Grid Search: Nonlinear Regression Optimization", fig.alt="Scatter plot titled 'Grid Search: Nonlinear Regression Optimization' showing data points and a fitted curve. The x-axis is labeled 'x' and the y-axis is labeled 'y.' Black dots represent data points, and a red line represents the fitted curve. A legend indicates 'Data' with a black dot and 'Fitted Curve' with a red line. The curve demonstrates a nonlinear relationship between x and y.", out.width="100%", fig.align='center'}
# Plot data and fitted curve
plot(
    x, y,
    main = "Grid Search: Nonlinear Regression Optimization",
    pch = 19,
    cex = 0.5,
    xlab = "x",
    ylab = "y"
)
curve(
    nonlinear_model(estimated_theta, x),
    from = min(x),
    to = max(x),
    add = TRUE,
    col = "red",
    lwd = 2
)
legend(
    "topleft",
    legend = c("Data", "Fitted Curve"),
    pch = c(19, NA),
    lty = c(NA, 1),
    col = c("black", "red")
)
```

------------------------------------------------------------------------

#### Nelder-Mead (Simplex) {#nelder-mead}

The Nelder-Mead algorithm, also known as the Simplex method, is a derivative-free optimization algorithm that iteratively adjusts a geometric shape (simplex) to find the minimum of an objective function. It is particularly useful for nonlinear regression when gradient-based methods fail due to non-differentiability or noisy function evaluations.

Nelder-Mead is particularly useful when:

-   The function is non-differentiable or noisy.

-   Gradient-based methods are unreliable.

-   The parameter space is low-dimensional.

------------------------------------------------------------------------

The goal of Nelder-Mead is to find an optimal parameter set $\hat{\theta}$ that minimizes the Sum of Squared Errors (SSE):

$$
\hat{\theta} = \arg\min_{\theta} SSE(\theta),
$$

where:

$$
SSE(\theta) = \sum_{i=1}^{n} (y_i - f(x_i; \theta))^2.
$$

**1. Simplex Representation**

The algorithm maintains a simplex, a geometric shape with $p+1$ vertices for a $p$-dimensional parameter space.

Each vertex represents a parameter vector:

$$
S = \{ \theta_1, \theta_2, \dots, \theta_{p+1} \}.
$$

**2. Simplex Operations**

At each iteration, the algorithm **updates the simplex** based on the objective function values at each vertex:

1.  **Reflection**: Reflect the worst point $\theta_h$ across the centroid.

    $$
    \theta_r = \theta_c + \alpha (\theta_c - \theta_h)
    $$

2.  **Expansion**: If reflection improves the objective, expand the step.

    $$
    \theta_e = \theta_c + \gamma (\theta_r - \theta_c)
    $$

3.  **Contraction**: If reflection worsens the objective, contract towards the centroid.

    $$
    \theta_c = \theta_c + \rho (\theta_h - \theta_c)
    $$

4.  **Shrink**: If contraction fails, shrink the simplex.

    $$
    \theta_i = \theta_1 + \sigma (\theta_i - \theta_1), \quad \forall i > 1
    $$

where:

-   $\alpha = 1$ (reflection coefficient),

-   $\gamma = 2$ (expansion coefficient),

-   $\rho = 0.5$ (contraction coefficient),

-   $\sigma = 0.5$ (shrink coefficient).

The process continues until convergence.

------------------------------------------------------------------------

**Nelder-Mead Algorithm**

1.  Initialize a simplex with $p+1$ vertices.
2.  Evaluate SSE at each vertex.
3.  Perform reflection, expansion, contraction, or shrink operations.
4.  Repeat until convergence.

------------------------------------------------------------------------

```{r}
# Load required library
library(stats)


# Define a numerically stable logistic function
safe_exp <- function(x) {
    return(ifelse(x > 700, Inf, exp(pmin(x, 700))))  # Prevent overflow
}

# Define the logistic growth model
nonlinear_model <- function(theta, x) {
    return(theta[1] / (1 + safe_exp(-theta[2] * (x - theta[3]))))
}

# Define the Sum of Squared Errors (SSE) function
sse <- function(theta, x, y) {
    predictions <- nonlinear_model(theta, x)
    return(sum((y - predictions) ^ 2))
}

# Nelder-Mead Optimization
nelder_mead_optimization <- function(x, y) {
    # Initial guess for parameters
    initial_guess <- c(2, 1, 0)
    
    # Run Nelder-Mead optimization
    result <- optim(
        par = initial_guess,
        fn = sse,
        x = x,
        y = y,
        method = "Nelder-Mead",
        control = list(maxit = 500)
    )
    
    return(result$par)  # Return optimized parameters
}

# Generate synthetic data
set.seed(123)
x <- seq(-5, 5, length.out = 100)
true_theta <- c(4, 1.5, 0)  # True parameters (A, B, C)
y <- nonlinear_model(true_theta, x) + rnorm(length(x), sd = 0.3)

# Run Nelder-Mead Optimization
estimated_theta <- nelder_mead_optimization(x, y)

# Display results
cat("Estimated parameters (A, B, C) using Nelder-Mead:\n")
print(estimated_theta)
```

```{r fig-nelder-mead, fig.cap="Nelder–Mead: Nonlinear Regression Optimization", fig.alt="Scatter plot titled 'Nelder–Mead: Nonlinear Regression Optimization' showing data points and a fitted curve. The x-axis ranges from -5 to 5, and the y-axis ranges from 0 to 4. Black dots represent data points, and a red line represents the fitted curve. A legend indicates 'Data' for black dots and 'Fitted Curve' for the red line.", out.width="100%", fig.align='center'}
# Plot data and fitted curve
plot(
    x,
    y,
    main = "Nelder-Mead: Nonlinear Regression Optimization",
    pch = 19,
    cex = 0.5,
    xlab = "x",
    ylab = "y"
)
curve(
    nonlinear_model(estimated_theta, x),
    from = min(x),
    to = max(x),
    add = TRUE,
    col = "red",
    lwd = 2
)
legend(
    "topleft",
    legend = c("Data", "Fitted Curve"),
    pch = c(19, NA),
    lty = c(NA, 1),
    col = c("black", "red")
)
```

------------------------------------------------------------------------

#### Powell's Method {#powells-method}

Powell's Method is a derivative-free optimization algorithm that minimizes a function without using gradients. It works by iteratively refining a set of search directions to efficiently navigate the parameter space. Unlike [Nelder-Mead (Simplex)](#nelder-mead), which adapts a simplex, Powell's method builds an orthogonal basis of search directions.

Powell's method is particularly useful when:

-   The function is non-differentiable or noisy.

-   Gradient-based methods are unreliable.

-   The parameter space is low-to-moderate dimensional.

------------------------------------------------------------------------

The goal of Powell's Method is to find an optimal parameter set $\hat{\theta}$ that minimizes the Sum of Squared Errors (SSE):

$$
\hat{\theta} = \arg\min_{\theta} SSE(\theta),
$$

where:

$$
SSE(\theta) = \sum_{i=1}^{n} (y_i - f(x_i; \theta))^2.
$$

**1. Search Directions**

Powell's method **maintains a set of search directions** $\mathbf{d}_1, \mathbf{d}_2, \dots, \mathbf{d}_p$:

$$
D = \{ \mathbf{d}_1, \mathbf{d}_2, ..., \mathbf{d}_p \}.
$$

Initially, these directions are chosen as **unit basis vectors** (each optimizing a single parameter independently).

**2. Line Minimization**

For each direction $\mathbf{d}_j$, Powell's method **performs a 1D optimization**:

$$
\theta' = \theta + \lambda \mathbf{d}_j,
$$

where $\lambda$ is chosen to minimize $SSE(\theta')$.

**3. Updating Search Directions**

After optimizing along all directions:

1.  The **largest improvement** direction $\mathbf{d}_{\max}$ is replaced with:

$$
\mathbf{d}_{\text{new}} = \theta_{\text{final}} - \theta_{\text{initial}}.
$$

2.  The **new direction set is orthogonalized** using Gram-Schmidt.

This ensures **efficient exploration** of the parameter space.

**4. Convergence Criteria**

Powell's method **stops when function improvement is below a tolerance level**:

$$
|SSE(\theta_{t+1}) - SSE(\theta_t)| < \epsilon.
$$

------------------------------------------------------------------------

**Powell's Method Algorithm**

1.  Initialize search directions (standard basis vectors).
2.  Perform 1D line searches along each direction.
3.  Update the search directions based on the largest improvement.
4.  Repeat until convergence.

------------------------------------------------------------------------

```{r}
# Load required library
library(stats)

# Define a numerically stable logistic function
safe_exp <- function(x) {
    return(ifelse(x > 700, Inf, exp(pmin(x, 700))))  # Prevent overflow
}

# Define the logistic growth model
nonlinear_model <- function(theta, x) {
    return(theta[1] / (1 + safe_exp(-theta[2] * (x - theta[3]))))
}

# Define the Sum of Squared Errors (SSE) function
sse <- function(theta, x, y) {
    predictions <- nonlinear_model(theta, x)
    return(sum((y - predictions) ^ 2))
}

# Powell's Method Optimization
powell_optimization <- function(x, y) {
    # Initial guess for parameters
    initial_guess <- c(2, 1, 0)
    
    # Run Powell’s optimization (via BFGS without gradient)
    result <- optim(
        par = initial_guess,
        fn = sse,
        x = x,
        y = y,
        method = "BFGS",
        control = list(maxit = 500),
        gr = NULL  # No gradient used
    )
    
    return(result$par)  # Return optimized parameters
}

# Generate synthetic data
set.seed(123)
x <- seq(-5, 5, length.out = 100)
true_theta <- c(4, 1.5, 0)  # True parameters (A, B, C)
y <- nonlinear_model(true_theta, x) + rnorm(length(x), sd = 0.3)

# Run Powell's Method Optimization
estimated_theta <- powell_optimization(x, y)

# Display results
cat("Estimated parameters (A, B, C) using Powell’s Method:\n")
print(estimated_theta)
```

```{r fig-powerll-method, fig.cap="Powell's Method: Nonlinear Regression Optimization", fig.alt="Scatter plot titled 'Powell's Method: Nonlinear Regression Optimization' showing data points and a fitted curve. The x-axis is labeled 'x' and the y-axis is labeled 'y.' Black dots represent data points, and a red line represents the fitted curve. A legend indicates 'Data' and 'Fitted Curve.'", out.width="100%", fig.align='center'}
# Plot data and fitted curve
plot(
    x,
    y,
    main = "Powell's Method: Nonlinear Regression Optimization",
    pch = 19,
    cex = 0.5,
    xlab = "x",
    ylab = "y"
)
curve(
    nonlinear_model(estimated_theta, x),
    from = min(x),
    to = max(x),
    add = TRUE,
    col = "red",
    lwd = 2
)
legend(
    "topleft",
    legend = c("Data", "Fitted Curve"),
    pch = c(19, NA),
    lty = c(NA, 1),
    col = c("black", "red")
)
```

------------------------------------------------------------------------

#### Random Search {#random-search}

Random Search (RS) is a simple yet effective optimization algorithm that explores the search space by randomly sampling candidate solutions. Unlike [grid search](#grid-search), which evaluates all predefined parameter combinations, random search selects a random subset, reducing computational cost.

Random search is particularly useful when:

-   The search space is large, making grid search impractical.

-   Gradient-based methods fail due to non-differentiability or noisy data.

-   The optimal region is unknown, making global exploration essential.

------------------------------------------------------------------------

The goal of Random Search is to find an optimal parameter set $\hat{\theta}$ that minimizes the Sum of Squared Errors (SSE):

$$
\hat{\theta} = \arg\min_{\theta \in \Theta} SSE(\theta),
$$

where:

$$
SSE(\theta) = \sum_{i=1}^{n} (y_i - f(x_i; \theta))^2.
$$

Unlike [grid search](#grid-search), **random search does not evaluate all parameter combinations** but instead **randomly samples a subset**:

$$
\Theta_{\text{sampled}} \subset \Theta.
$$

The **accuracy** of the solution depends on the **number of random samples**---a higher number increases **the likelihood of finding a good solution**.

------------------------------------------------------------------------

**Random Search Algorithm**

1.  Define a random sampling range for each parameter.
2.  Randomly sample $N$ parameter sets.
3.  Evaluate SSE for each sampled set.
4.  Select the parameter set that minimizes SSE.

------------------------------------------------------------------------

```{r}
# Load required library
library(stats)

# Define a numerically stable logistic function
safe_exp <- function(x) {
  return(ifelse(x > 700, Inf, exp(pmin(x, 700))))  # Prevent overflow
}

# Define the logistic growth model
nonlinear_model <- function(theta, x) {
  return(theta[1] / (1 + safe_exp(-theta[2] * (x - theta[3]))))
}

# Define the Sum of Squared Errors (SSE) function
sse <- function(theta, x, y) {
  predictions <- nonlinear_model(theta, x)
  return(sum((y - predictions) ^ 2))
}

# Random Search Optimization
random_search_optimization <- function(x, y, num_samples = 100) {
  # Define parameter ranges
  A_range <-
    runif(num_samples, 2, 6)   # Random values between 2 and 6
  B_range <-
    runif(num_samples, 0.5, 3) # Random values between 0.5 and 3
  C_range <-
    runif(num_samples,-2, 2)  # Random values between -2 and 2
  
  # Initialize best parameters
  best_theta <- NULL
  best_sse <- Inf
  
  # Evaluate randomly sampled parameter sets
  for (i in 1:num_samples) {
    theta <- c(A_range[i], B_range[i], C_range[i])
    current_sse <- sse(theta, x, y)
    
    if (current_sse < best_sse) {
      best_sse <- current_sse
      best_theta <- theta
    }
  }
  
  return(best_theta)
}

# Generate synthetic data
set.seed(123)
x <- seq(-5, 5, length.out = 100)
true_theta <- c(4, 1.5, 0)  # True parameters (A, B, C)
y <- nonlinear_model(true_theta, x) + rnorm(length(x), sd = 0.3)

# Run Random Search
estimated_theta <-
  random_search_optimization(x, y, num_samples = 500)

# Display results
cat("Estimated parameters (A, B, C) using Random Search:\n")
print(estimated_theta)
```

```{r fig-random-search, fig.cap="Random Search: Nonlinear Regression Optimization", fig.alt="Scatter plot titled 'Random Search: Nonlinear Regression Optimization' showing data points and a fitted curve. The x-axis is labeled 'x' and the y-axis is labeled 'y.' Black dots represent data points, and a red line represents the fitted curve. A legend indicates 'Data' with a black dot and 'Fitted Curve' with a red line.", out.width="100%", fig.align='center'}
# Plot data and fitted curve
plot(
  x,
  y,
  main = "Random Search: Nonlinear Regression Optimization",
  pch = 19,
  cex = 0.5,
  xlab = "x",
  ylab = "y"
)
curve(
  nonlinear_model(estimated_theta, x),
  from = min(x),
  to = max(x),
  add = TRUE,
  col = "red",
  lwd = 2
)
legend(
  "topleft",
  legend = c("Data", "Fitted Curve"),
  pch = c(19, NA),
  lty = c(NA, 1),
  col = c("black", "red")
)
```

------------------------------------------------------------------------

#### Hooke-Jeeves Pattern Search {#hooke-jeeves}

Hooke-Jeeves Pattern Search is a derivative-free optimization algorithm that searches for an optimal solution by exploring and adjusting parameter values iteratively. Unlike gradient-based methods, it does not require differentiability, making it effective for non-smooth and noisy functions.

Hooke-Jeeves is particularly useful when:

-   The function is non-differentiable or noisy.

-   Gradient-based methods are unreliable.

-   The parameter space is low-to-moderate dimensional.

------------------------------------------------------------------------

The goal of Hooke-Jeeves Pattern Search is to find an optimal parameter set $\hat{\theta}$ that minimizes the Sum of Squared Errors (SSE):

$$
\hat{\theta} = \arg\min_{\theta} SSE(\theta),
$$

where:

$$
SSE(\theta) = \sum_{i=1}^{n} (y_i - f(x_i; \theta))^2.
$$

**1. Exploratory Moves**

At each iteration, the algorithm **perturbs each parameter** to find a lower SSE:

$$
\theta' = \theta \pm \delta.
$$

If a new parameter $\theta'$ reduces SSE, it becomes the new base point.

**2. Pattern Moves**

After an improvement, the algorithm **accelerates movement** in the promising direction:

$$
\theta_{\text{new}} = \theta_{\text{old}} + (\theta_{\text{old}} - \theta_{\text{prev}}).
$$

This **speeds up convergence** towards an optimum.

**3. Step Size Adaptation**

If no improvement is found, the step size $\delta$ is reduced:

$$
\delta_{\text{new}} = \beta \cdot \delta.
$$

where $\beta < 1$ is a reduction factor.

**4. Convergence Criteria**

The algorithm **stops when step size is sufficiently small**:

$$
\delta < \epsilon.
$$

------------------------------------------------------------------------

**Hooke-Jeeves Algorithm**

1.  Initialize a starting point $\theta$ and step size $\delta$.
2.  Perform exploratory moves in each parameter direction.
3.  If improvement is found, perform pattern moves.
4.  Reduce step size if no improvement is found.
5.  Repeat until convergence.

------------------------------------------------------------------------

```{r}
# Load required library
library(stats)

# Define a numerically stable logistic function
safe_exp <- function(x) {
    return(ifelse(x > 700, Inf, exp(pmin(x, 700))))  # Prevent overflow
}

# Define the logistic growth model
nonlinear_model <- function(theta, x) {
    return(theta[1] / (1 + safe_exp(-theta[2] * (x - theta[3]))))
}

# Define the Sum of Squared Errors (SSE) function
sse <- function(theta, x, y) {
    predictions <- nonlinear_model(theta, x)
    return(sum((y - predictions) ^ 2))
}

# Hooke-Jeeves Pattern Search Optimization
hooke_jeeves_optimization <-
    function(x,
             y,
             step_size = 0.5,
             tol = 1e-6,
             max_iter = 500) {
        # Initial guess for parameters
        theta <- c(2, 1, 0)
        best_sse <- sse(theta, x, y)
        step <- step_size
        iter <- 0
        
        while (step > tol & iter < max_iter) {
            iter <- iter + 1
            improved <- FALSE
            new_theta <- theta
            
            # Exploratory move in each parameter direction
            for (i in 1:length(theta)) {
                for (delta in c(step,-step)) {
                    theta_test <- new_theta
                    theta_test[i] <- theta_test[i] + delta
                    sse_test <- sse(theta_test, x, y)
                    
                    if (sse_test < best_sse) {
                        best_sse <- sse_test
                        new_theta <- theta_test
                        improved <- TRUE
                    }
                }
            }
            
            # Pattern move if improvement found
            if (improved) {
                theta <- 2 * new_theta - theta
                best_sse <- sse(theta, x, y)
            } else {
                step <- step / 2  # Reduce step size
            }
        }
        
        return(theta)
    }

# Generate synthetic data
set.seed(123)
x <- seq(-5, 5, length.out = 100)
true_theta <- c(4, 1.5, 0)  # True parameters (A, B, C)
y <- nonlinear_model(true_theta, x) + rnorm(length(x), sd = 0.3)

# Run Hooke-Jeeves Optimization
estimated_theta <- hooke_jeeves_optimization(x, y)

# Display results
cat("Estimated parameters (A, B, C) using Hooke-Jeeves:\n")
print(estimated_theta)
```

```{r fig-hooke-jeeves, fig.cap="Hooke–Jeeves: Nonlinear Regression Optimization", fig.alt="Scatter plot titled 'Hooke–Jeeves: Nonlinear Regression Optimization' showing data points and a fitted curve. The x-axis is labeled 'x' and the y-axis is labeled 'y.' The legend indicates black dots represent data and a red line represents the fitted curve. The curve demonstrates a nonlinear relationship, increasing from left to right.", out.width="100%", fig.align='center'}
# Plot data and fitted curve
plot(
    x,
    y,
    main = "Hooke-Jeeves: Nonlinear Regression Optimization",
    pch = 19,
    cex = 0.5,
    xlab = "x",
    ylab = "y"
)
curve(
    nonlinear_model(estimated_theta, x),
    from = min(x),
    to = max(x),
    add = TRUE,
    col = "red",
    lwd = 2
)
legend(
    "topleft",
    legend = c("Data", "Fitted Curve"),
    pch = c(19, NA),
    lty = c(NA, 1),
    col = c("black", "red")
)
```

------------------------------------------------------------------------

#### Bisection Method {#bisection-method}

The **Bisection Method** is a fundamental numerical technique primarily used for **root finding**, but it can also be adapted for **optimization problems** where the goal is to minimize or maximize a nonlinear function.

In nonlinear regression, optimization often involves **finding the parameter values that minimize the sum of squared errors (SSE)**:

$$
\hat{\theta} = \arg\min_{\theta} SSE(\theta)
$$

where:

$$
SSE(\theta) = \sum_{i=1}^{n} \left( y_i - f(x_i; \theta) \right)^2.
$$

Since the minimum of a function occurs where the derivative equals zero, we apply the **Bisection Method** to the **derivative of the SSE function**:

$$
\frac{d}{d\theta} SSE(\theta) = 0.
$$

This transforms the **optimization problem** into a **root-finding problem**.

------------------------------------------------------------------------

The **Bisection Method** is based on the **Intermediate Value Theorem**, which states:

> If a continuous function $f(\theta)$ satisfies $f(\theta_a) \cdot f(\theta_b) < 0$,\
> then there exists at least one root in the interval $(\theta_a, \theta_b)$.

For **optimization**, we apply this principle to the **first derivative** of the objective function $Q(\theta)$:

$$
Q'(\theta) = 0.
$$

**Step-by-Step Algorithm for Optimization**

1.  **Choose an interval** $[\theta_a, \theta_b]$ such that: $$ Q'(\theta_a) \cdot Q'(\theta_b) < 0. $$
2.  **Compute the midpoint**: $$
    \theta_m = \frac{\theta_a + \theta_b}{2}.
    $$
3.  **Evaluate** $Q'(\theta_m)$:
    -   If $Q'(\theta_m) = 0$, then $\theta_m$ is the optimal point.
    -   If $Q'(\theta_a) \cdot Q'(\theta_m) < 0$, set $\theta_b = \theta_m$.
    -   Otherwise, set $\theta_a = \theta_m$.
4.  **Repeat** until convergence: $$
    |\theta_b - \theta_a| < \epsilon.
    $$

**Determining the Nature of the Critical Point**

Since the Bisection Method finds a **stationary point**, we use the **second derivative test** to classify it:

-   If $Q''(\theta) > 0$, the point is a **local minimum**.
-   If $Q''(\theta) < 0$, the point is a **local maximum**.

For **nonlinear regression**, we expect $Q(\theta) = SSE(\theta)$, so the solution found by Bisection should be a **minimum**.

------------------------------------------------------------------------

```{r}
# Load required library
library(stats)

# Define a numerically stable logistic function
safe_exp <- function(x) {
    return(ifelse(x > 700, Inf, exp(pmin(x, 700))))  # Prevent overflow
}

# Define the logistic growth model
nonlinear_model <- function(theta, x) {
    return(theta[1] / (1 + safe_exp(-theta[2] * (x - theta[3]))))
}

# Define the Sum of Squared Errors (SSE) function for optimization
sse <- function(theta, x, y) {
    predictions <- nonlinear_model(theta, x)
    return(sum((y - predictions)^2))
}

# Optimize all three parameters simultaneously
find_optimal_parameters <- function(x, y) {
    # Initial guess for parameters (based on data)
    initial_guess <- c(max(y), 1, median(x))  

    # Bounds for parameters
    lower_bounds <- c(0.1, 0.01, min(x))  # Ensure positive scaling
    upper_bounds <- c(max(y) * 2, 10, max(x))

    # Run optim() with L-BFGS-B (bounded optimization)
    result <- optim(
        par = initial_guess,
        fn = sse,
        x = x, y = y,
        method = "L-BFGS-B",
        lower = lower_bounds,
        upper = upper_bounds
    )
    
    return(result$par)  # Extract optimized parameters
}

# Generate synthetic data
set.seed(123)
x <- seq(-5, 5, length.out = 100)
true_theta <- c(4, 1.5, 0)  # True parameters (A, B, C)
y <- nonlinear_model(true_theta, x) + rnorm(length(x), sd = 0.3)

# Find optimal parameters using optim()
estimated_theta <- find_optimal_parameters(x, y)

# Display results
cat("Estimated parameters (A, B, C) using optim():\n")
print(estimated_theta)
```

```{r fig-optim, fig.cap="Optim(): Nonlinear Regression Optimization", fig.alt="Scatter plot titled 'Optim(): Nonlinear Regression Optimization' showing data points and a fitted curve. The x-axis is labeled 'x' and the y-axis is labeled 'y.' Black dots represent data points, and a red line represents the fitted curve. A legend indicates the symbols for 'Data' and 'Fitted Curve.' ", out.width="100%", fig.align='center'}
# Plot data and fitted curve
plot(
    x, y,
    main = "Optim(): Nonlinear Regression Optimization",
    pch = 19,
    cex = 0.5,
    xlab = "x",
    ylab = "y"
)
curve(
    nonlinear_model(estimated_theta, x),
    from = min(x),
    to = max(x),
    add = TRUE,
    col = "red",
    lwd = 2
)
legend(
    "topleft",
    legend = c("Data", "Fitted Curve"),
    pch = c(19, NA),
    lty = c(NA, 1),
    col = c("black", "red")
)
```

------------------------------------------------------------------------

### Stochastic Heuristic {#stochastic-heuristic-nolinear-regression}

#### Differential Evolution Algorithm {#differential-evolution-algorithm}

The **Differential Evolution (DE) Algorithm** is a **stochastic, population-based optimization algorithm** that is widely used for solving complex **global optimization problems**. Unlike gradient-based methods such as Newton's method or the [Secant method](#secant-method), DE does not require derivatives and is well-suited for optimizing **non-differentiable, nonlinear, and multimodal functions**.

------------------------------------------------------------------------

**Key Features of Differential Evolution**

-   **Population-based approach:** Maintains a population of candidate solutions instead of a single point.
-   **Mutation and crossover:** Introduces variations to explore the search space.
-   **Selection mechanism:** Retains the best candidates for the next generation.
-   **Global optimization:** Avoids local minima by using **stochastic search strategies**.

------------------------------------------------------------------------

**Mathematical Formulation of Differential Evolution**

Differential Evolution operates on a **population** of candidate solutions $\{\theta_i\}$, where each $\theta_i$ is a **vector of parameters**. The algorithm iteratively updates the population using three main operations:

**1. Mutation**

For each candidate solution $\theta_i$, a **mutant vector** $\mathbf{v}_i^{(j)}$ is generated as:

$$
\mathbf{v}_i^{(j)} = \mathbf{\theta}_{r_1}^{(j)} + F \cdot (\mathbf{\theta}_{r_2}^{(j)} - \mathbf{\theta}_{r_3}^{(j)})
$$

where:

-   $\mathbf{\theta}_{r_1}, \mathbf{\theta}_{r_2}, \mathbf{\theta}_{r_3}$ are **randomly selected distinct vectors** from the population.

-   $F \in (0,2)$ is the **mutation factor** controlling the step size.

**2. Crossover**

A trial vector $\mathbf{u}_i^{(j)}$ is generated by combining the mutant vector $\mathbf{v}_i^{(j)}$ with the original solution $\mathbf{\theta}_i^{(j)}$:

$$
u_{i,k}^{(j)} =
\begin{cases}
v_{i,k}^{(j)}  & \text{if } rand_k \leq C_r \text{ or } k = k_{\text{rand}}, \\
\theta_{i,k}^{(j)}  & \text{otherwise}.
\end{cases}
$$

where:

-   $C_r \in (0,1)$ is the **crossover probability**.

-   $rand_k$ is a **random value between 0 and 1**.

-   $k_{\text{rand}}$ ensures at least **one parameter is mutated**.

**3. Selection**

The new candidate solution is accepted only if it improves the objective function:

$$
\mathbf{\theta}_i^{(j+1)} =
\begin{cases}
\mathbf{u}_i^{(j)} & \text{if } Q(\mathbf{u}_i^{(j)}) < Q(\mathbf{\theta}_i^{(j)}), \\
\mathbf{\theta}_i^{(j)} & \text{otherwise}.
\end{cases}
$$

where $Q(\theta)$ is the **objective function** (e.g., sum of squared errors in regression problems).

------------------------------------------------------------------------

**Algorithm: Differential Evolution for Nonlinear Optimization**

1.  Initialize a population of candidate solutions.
2.  Evaluate the objective function for each candidate.
3.  Mutate individuals using a difference strategy.
4.  Apply crossover to create trial solutions.
5.  Select individuals based on their fitness (objective function value).
6.  Repeat until convergence (or a stopping criterion is met).

------------------------------------------------------------------------

```{r}
# Load required library
library(DEoptim)

# Define a nonlinear function (logistic model)
nonlinear_model <- function(theta, x) {
    return(theta[1] / (1 + exp(-theta[2] * (x - theta[3]))))
}

# Define the Sum of Squared Errors (SSE) function
sse <- function(theta, x, y) {
    return(sum((y - nonlinear_model(theta, x))^2))
}

# Generate synthetic data
set.seed(123)
x <- seq(-5, 5, length.out = 100)
true_theta <- c(4, 1.5, 0)  # True parameters (A, B, C)
y <- nonlinear_model(true_theta, x) + rnorm(length(x), sd = 0.3)

# Define the objective function for DEoptim
objective_function <- function(theta) {
    return(sse(theta, x, y))
}

# Define parameter bounds
lower_bounds <- c(0, 0, -5)
upper_bounds <- c(10, 5, 5)

# Run Differential Evolution Algorithm
de_result <-
    DEoptim(
        objective_function,
        lower_bounds,
        upper_bounds,
        DEoptim.control(
            NP = 50,
            itermax = 100,
            F = 0.8,
            CR = 0.9, 
            trace = F
        )
    )

# Extract optimized parameters
estimated_theta <- de_result$optim$bestmem

# Display estimated parameters
cat("Estimated parameters (A, B, C) using Differential Evolution:\n")
print(estimated_theta)
```

```{r fig-differential-evoluion, fig.cap="Differential Evolution", fig.alt="Scatter plot titled 'Differential Evolution' showing data points and a fitted curve. The x-axis is labeled 'x' and the y-axis is labeled 'y.' Black dots represent data points, and a red line represents the fitted curve. A legend indicates black dots as 'Data' and the red line as 'Fitted Curve.' The curve shows a sigmoidal trend.", out.width="100%", fig.align='center'}
# Plot data and fitted curve
plot(
    x,
    y,
    main = "Differential Evolution: Data & Fitted Curve",
    pch = 19,
    cex = 0.5,
    xlab = "x",
    ylab = "y"
)
curve(
    nonlinear_model(estimated_theta, x),
    from = -5,
    to = 5,
    add = TRUE,
    col = "red",
    lwd = 2
)
legend(
    "topleft",
    legend = c("Data", "Fitted Curve"),
    pch = c(19, NA),
    lty = c(NA, 1),
    col = c("black", "red")
)
```

------------------------------------------------------------------------

#### Simulated Annealing {#simulated-annealing}

Simulated Annealing (SA) is a probabilistic global optimization algorithm inspired by annealing in metallurgy, where a material is heated and slowly cooled to remove defects. In optimization, SA gradually refines a solution by exploring the search space, allowing occasional jumps to escape local minima, before converging to an optimal solution.

Simulated Annealing is particularly useful when:

-   The function is highly nonlinear and multimodal.

-   Gradient-based methods struggle due to non-differentiability or poor initialization.

-   A global minimum is needed, rather than a local one.

------------------------------------------------------------------------

**1. Energy Function (Objective Function)**

The goal of SA is to minimize an objective function $Q(\theta)$. For nonlinear regression, this is the Sum of Squared Errors (SSE):

$$
Q(\theta) = SSE(\theta) = \sum_{i=1}^{n} (y_i - f(x_i; \theta))^2.
$$

**2. Probability of Acceptance**

At each step, SA randomly perturbs the parameters $\theta$ to create a **new candidate solution** $\theta'$ and evaluates the change in SSE:

$$
\Delta Q = Q(\theta') - Q(\theta).
$$

The **Metropolis Criterion** determines whether to accept the new solution:

$$
P(\text{accept}) =
\begin{cases}
1, & \text{if } \Delta Q < 0 \quad \text{(new solution improves fit)} \\
\exp\left( -\frac{\Delta Q}{T} \right), & \text{if } \Delta Q \geq 0 \quad \text{(accept with probability)}.
\end{cases}
$$

where:

-   $T$ is the **temperature** that **gradually decreases** over iterations.

-   **Worse solutions** are accepted **with small probability** to escape local minima.

**3. Cooling Schedule**

The **temperature** follows a cooling schedule:

$$
T_k = \alpha T_{k-1},
$$

where $\alpha \in (0,1)$ is a decay factor that controls cooling speed.

------------------------------------------------------------------------

**Simulated Annealing Algorithm**

1.  Initialize parameters $\theta$ randomly.
2.  Set an initial temperature $T_0$ and cooling rate $\alpha$.
3.  Repeat for max iterations:
    -   Generate perturbed candidate $\theta'$.
    -   Compute $\Delta Q = Q(\theta') - Q(\theta)$.
    -   Accept if $\Delta Q < 0$ or with probability $\exp(-\Delta Q / T)$.
    -   Reduce temperature: $T \leftarrow \alpha T$.
4.  Return the best solution found.

------------------------------------------------------------------------

```{r}
# Load required library
library(stats)

# Define a numerically stable logistic function
safe_exp <- function(x) {
    return(ifelse(x > 700, Inf, exp(pmin(x, 700))))  # Prevent overflow
}

# Define the logistic growth model
nonlinear_model <- function(theta, x) {
    return(theta[1] / (1 + safe_exp(-theta[2] * (x - theta[3]))))
}

# Define the Sum of Squared Errors (SSE) function
sse <- function(theta, x, y) {
    predictions <- nonlinear_model(theta, x)
    return(sum((y - predictions) ^ 2))
}

# Simulated Annealing Algorithm
simulated_annealing <-
    function(x,
             y,
             initial_theta,
             T_init = 1.0,
             alpha = 0.99,
             max_iter = 5000) {
        # Initialize parameters
        theta <- initial_theta
        best_theta <- theta
        best_sse <- sse(theta, x, y)
        T <- T_init  # Initial temperature
        
        for (iter in 1:max_iter) {
            # Generate new candidate solution (small random perturbation)
            theta_new <- theta + rnorm(length(theta), mean = 0, sd = T)
            
            # Compute new SSE
            sse_new <- sse(theta_new, x, y)
            
            # Compute change in SSE
            delta_Q <- sse_new - best_sse
            
            # Acceptance criteria
            if (delta_Q < 0 || runif(1) < exp(-delta_Q / T)) {
                theta <- theta_new
                best_sse <- sse_new
                best_theta <- theta_new
            }
            
            # Reduce temperature
            T <- alpha * T
            
            # Stopping condition (very low temperature)
            if (T < 1e-6)
                break
        }
        
        return(best_theta)
    }

# Generate synthetic data
set.seed(123)
x <- seq(-5, 5, length.out = 100)
true_theta <- c(4, 1.5, 0)  # True parameters (A, B, C)
y <- nonlinear_model(true_theta, x) + rnorm(length(x), sd = 0.3)

# Initial guess
initial_theta <-
    c(runif(1, 1, 5), runif(1, 0.1, 3), runif(1,-2, 2))

# Run Simulated Annealing
estimated_theta <- simulated_annealing(x, y, initial_theta)

# Display results
cat("Estimated parameters (A, B, C) using Simulated Annealing:\n")
print(estimated_theta)
```

```{r fig-simulated-annealing, fig.cap="Simulated Annealing: Nonlinear Regression Optimization", fig.alt="Scatter plot titled 'Simulated Annealing: Nonlinear Regression Optimization' showing data points and a fitted curve. The x-axis is labeled 'x' and the y-axis is labeled 'y.' Black dots represent data points, and a red line represents the fitted curve, illustrating the optimization process. A legend indicates 'Data' and 'Fitted Curve.'", out.width="100%", fig.align='center'}
# Plot data and fitted curve
plot(
    x,
    y,
    main = "Simulated Annealing: Nonlinear Regression Optimization",
    pch = 19,
    cex = 0.5,
    xlab = "x",
    ylab = "y"
)
curve(
    nonlinear_model(estimated_theta, x),
    from = min(x),
    to = max(x),
    add = TRUE,
    col = "red",
    lwd = 2
)
legend(
    "topleft",
    legend = c("Data", "Fitted Curve"),
    pch = c(19, NA),
    lty = c(NA, 1),
    col = c("black", "red")
)
```

------------------------------------------------------------------------

#### Genetic Algorithm {#genetic-algorithm}

Genetic Algorithms (GA) are a class of evolutionary algorithms inspired by the principles of natural selection and genetics. Unlike deterministic optimization techniques, GA evolves a population of candidate solutions over multiple generations, using genetic operators such as selection, crossover, and mutation.

GA is particularly useful when:

-   The function is nonlinear, non-differentiable, or highly multimodal.

-   Gradient-based methods fail due to rugged function landscapes.

-   A global minimum is required, rather than a local one.

The goal of a Genetic Algorithm is to find an optimal solution \$\\hat{\\theta}\$ that minimizes an objective function:

$$
\hat{\theta} = \arg\min_{\theta} SSE(\theta),
$$

where:

$$
SSE(\theta) = \sum_{i=1}^{n} (y_i - f(x_i; \theta))^2.
$$

**1. Population Representation**

Each candidate solution (individual) is represented as a **chromosome**, which is simply a vector of parameters: $$
\theta = (\theta_1, \theta_2, \theta_3)
$$

An entire **population** consists of multiple such solutions.

**2. Selection**

Each individual's fitness is evaluated using:

$$
\text{Fitness}(\theta) = -SSE(\theta)
$$

We use **Tournament Selection** or **Roulette Wheel Selection** to choose parents for reproduction.

**3. Crossover (Recombination)**

A new solution $\theta'$ is generated by **combining** two parents:

$$\theta' = \alpha \theta_{\text{parent1}} + (1 - \alpha) \theta_{\text{parent2}}, \quad \alpha \sim U(0,1).$$

**4. Mutation**

Random small changes are introduced to **increase diversity**:

$$\theta_i' = \theta_i + \mathcal{N}(0, \sigma),$$

where $\mathcal{N}(0, \sigma)$ is a **small Gaussian perturbation**.

**5. Evolutionary Cycle**

The **algorithm iterates** through:

1.  Selection

2.  Crossover

3.  Mutation

4.  Survival of the fittest

5.  Termination when convergence is reached.

```{r}
# Load required library
library(GA)

# Define a numerically stable logistic function
safe_exp <- function(x) {
    return(ifelse(x > 700, Inf, exp(pmin(x, 700))))  # Prevent overflow
}

# Define the logistic growth model
nonlinear_model <- function(theta, x) {
    return(theta[1] / (1 + safe_exp(-theta[2] * (x - theta[3]))))
}

# Define the Sum of Squared Errors (SSE) function for optimization
sse <- function(theta, x, y) {
    predictions <- nonlinear_model(theta, x)
    return(sum((y - predictions) ^ 2))
}

# Genetic Algorithm for Optimization
ga_optimization <- function(x, y) {
    # Define fitness function (negative SSE for maximization)
    fitness_function <- function(theta) {
        # GA maximizes fitness, so we use negative SSE
        return(-sse(theta, x, y))  
    }
    
    # Set parameter bounds
    lower_bounds <- c(0.1, 0.01, min(x))  # Ensure positive scaling
    upper_bounds <- c(max(y) * 2, 10, max(x))
    
    # Run GA optimization
    ga_result <- ga(
        type = "real-valued",
        fitness = fitness_function,
        lower = lower_bounds,
        upper = upper_bounds,
        popSize = 50,
        # Population size
        maxiter = 200,
        # Max generations
        pmutation = 0.1,
        # Mutation probability
        monitor = FALSE
    )
    
    return(ga_result@solution)  # Return optimized parameters
}

# Generate synthetic data
set.seed(123)
x <- seq(-5, 5, length.out = 100)
true_theta <- c(4, 1.5, 0)  # True parameters (A, B, C)
y <- nonlinear_model(true_theta, x) + rnorm(length(x), sd = 0.3)

# Run Genetic Algorithm
estimated_theta <- ga_optimization(x, y)

# Display results
cat("Estimated parameters (A, B, C) using Genetic Algorithm:\n")
print(estimated_theta)
```

```{r fig-genetic-algo, fig.cap="Genetic Algorithm: Nonlinear Regression Optimization", fig.alt="Scatter plot titled 'Genetic Algorithm: Nonlinear Regression Optimization' showing data points and a fitted curve. The x-axis is labeled 'x' and the y-axis is labeled 'y.' Black dots represent data points, and a red line represents the fitted curve, illustrating the optimization process. A legend indicates 'Data' and 'Fitted Curve.'", out.width="100%", fig.align='center'}
# Plot data and fitted curve
plot(
    x,
    y,
    main = "Genetic Algorithm: Nonlinear Regression Optimization",
    pch = 19,
    cex = 0.5,
    xlab = "x",
    ylab = "y"
)
curve(
    nonlinear_model(estimated_theta, x),
    from = min(x),
    to = max(x),
    add = TRUE,
    col = "red",
    lwd = 2
)
legend(
    "topleft",
    legend = c("Data", "Fitted Curve"),
    pch = c(19, NA),
    lty = c(NA, 1),
    col = c("black", "red")
)
```

------------------------------------------------------------------------

#### Particle Swarm Optimization {#particle-swarm-optimization}

Particle Swarm Optimization (PSO) is a population-based global optimization algorithm inspired by the social behavior of birds and fish schools. Instead of using genetic operators (like in Genetic Algorithms), PSO models particles (solutions) flying through the search space, adjusting their position based on their own experience and the experience of their neighbors.

PSO is particularly useful when:

-   The function is nonlinear, noisy, or lacks smooth gradients.

-   Gradient-based methods struggle due to non-differentiability.

-   A global minimum is needed, rather than a local one.

------------------------------------------------------------------------

The goal of PSO is to find an optimal solution $\hat{\theta}$ that minimizes an objective function:

$$
\hat{\theta} = \arg\min_{\theta} SSE(\theta),
$$

where:

$$
SSE(\theta) = \sum_{i=1}^{n} (y_i - f(x_i; \theta))^2.
$$

**1. Particle Representation**

Each **particle** represents a candidate solution:

$$
\theta_i = (\theta_{i1}, \theta_{i2}, \theta_{i3})
$$

where $\theta_{ij}$ is the $j^{th}$ parameter of particle $i$.

**2. Particle Velocity and Position Updates**

Each particle moves in the search space with **velocity** $v_i$, which is updated as:

$$
v_i^{(t+1)} = \omega v_i^{(t)} + c_1 r_1 (p_i - \theta_i^{(t)}) + c_2 r_2 (g - \theta_i^{(t)})
$$

where:

-   $\omega$ is the **inertia weight** (controls exploration vs. exploitation),

-   $c_1, c_2$ are **acceleration coefficients**,

-   $r_1, r_2 \sim U(0,1)$ are random numbers,

-   $p_i$ is the **particle's personal best position**,

-   $g$ is the **global best position**.

Then, the **position update** is:

$$
\theta_i^{(t+1)} = \theta_i^{(t)} + v_i^{(t+1)}
$$

This process continues until **convergence criteria** (like a max number of iterations or minimum error) is met.

------------------------------------------------------------------------

**Particle Swarm Optimization Algorithm**

1.  Initialize particles randomly within search bounds.
2.  Set random initial velocities.
3.  Evaluate SSE for each particle.
4.  Update the personal and global best solutions.
5.  Update velocities and positions using the update equations.
6.  Repeat until convergence.

------------------------------------------------------------------------

```{r}
# Load required library
library(pso)

# Define a numerically stable logistic function
safe_exp <- function(x) {
    return(ifelse(x > 700, Inf, exp(pmin(x, 700))))  # Prevent overflow
}

# Define the logistic growth model
nonlinear_model <- function(theta, x) {
    return(theta[1] / (1 + safe_exp(-theta[2] * (x - theta[3]))))
}

# Define the Sum of Squared Errors (SSE) function for optimization
sse <- function(theta, x, y) {
    predictions <- nonlinear_model(theta, x)
    return(sum((y - predictions) ^ 2))
}

# Particle Swarm Optimization (PSO) for Nonlinear Regression
pso_optimization <- function(x, y) {
    # Define fitness function (minimize SSE)
    fitness_function <- function(theta) {
        return(sse(theta, x, y))
    }
    
    # Set parameter bounds
    lower_bounds <- c(0.1, 0.01, min(x))  # Ensure positive scaling
    upper_bounds <- c(max(y) * 2, 10, max(x))
    
    # Run PSO optimization
    pso_result <- psoptim(
        par = c(1, 1, 0),
        # Initial guess
        fn = fitness_function,
        lower = lower_bounds,
        upper = upper_bounds,
        control = list(maxit = 200, s = 50)  # 200 iterations, 50 particles
    )
    
    return(pso_result$par)  # Return optimized parameters
}

# Generate synthetic data
set.seed(123)
x <- seq(-5, 5, length.out = 100)
true_theta <- c(4, 1.5, 0)  # True parameters (A, B, C)
y <- nonlinear_model(true_theta, x) + rnorm(length(x), sd = 0.3)

# Run Particle Swarm Optimization
estimated_theta <- pso_optimization(x, y)

# Display results
cat("Estimated parameters (A, B, C) using Particle Swarm Optimization:\n")
print(estimated_theta)
```

```{r fig-particle-swarm-opti, fig.cap="Particle Swarm Optimization: Nonlinear Regression Optimization", fig.alt="Scatter plot titled 'Particle Swarm Optimization: Nonlinear Regression Optimization' showing data points and a fitted curve. The x-axis is labeled 'x' and the y-axis is labeled 'y.' Black dots represent data, and a red line represents the fitted curve. A legend indicates the symbols for data and the fitted curve.", out.width="100%", fig.align='center'}
# Plot data and fitted curve
plot(
    x,
    y,
    main = "Particle Swarm Optimization: Nonlinear Regression Optimization",
    pch = 19,
    cex = 0.5,
    xlab = "x",
    ylab = "y"
)
curve(
    nonlinear_model(estimated_theta, x),
    from = min(x),
    to = max(x),
    add = TRUE,
    col = "red",
    lwd = 2
)
legend(
    "topleft",
    legend = c("Data", "Fitted Curve"),
    pch = c(19, NA),
    lty = c(NA, 1),
    col = c("black", "red")
)
```

------------------------------------------------------------------------

#### Evolutionary Strategies {#evolutionary-strategies}

Evolutionary Strategies (ES) are a class of evolutionary optimization algorithms that improve solutions by mutating and selecting individuals based on fitness. Unlike [Genetic Algorithm](#genetic-algorithm), ES focuses on self-adaptive mutation rates and selection pressure rather than crossover. This makes ES particularly robust for continuous optimization problems like nonlinear regression.

ES is particularly useful when:

-   The function is complex, noisy, or lacks smooth gradients.

-   Gradient-based methods fail due to non-differentiability.

-   An adaptive approach to exploration and exploitation is needed.

------------------------------------------------------------------------

The goal of ES is to find an optimal solution $\hat{\theta}$ that minimizes an objective function:

$$
\hat{\theta} = \arg\min_{\theta} SSE(\theta),
$$

where:

$$
SSE(\theta) = \sum_{i=1}^{n} (y_i - f(x_i; \theta))^2.
$$

**1. Population Representation**

Each **individual** is a solution $\theta_i$ in the parameter space:

$$
\theta_i = (\theta_{i1}, \theta_{i2}, \theta_{i3}).
$$

The **population** consists of multiple individuals, each representing different candidate parameters.

**2. Mutation**

New candidate solutions are generated by **adding random noise**:

$$
\theta'_i = \theta_i + \sigma \mathcal{N}(0, I),
$$

where:

-   $\sigma$ is the **mutation step size**, which adapts over time.

-   $\mathcal{N}(0, I)$ is a standard normal distribution.

**3. Selection**

ES employs $(\mu, \lambda)$-selection:

-   $(\mu, \lambda)$-ES: Select the best $\mu$ solutions from $\lambda$ offspring.

-   $(\mu + \lambda)$-ES: Combine parents and offspring, selecting the top $\mu$.

**4. Step-Size Adaptation**

Mutation strength $\sigma$ **self-adapts** using the **1/5 success rule**:

$$
\sigma_{t+1} =
\begin{cases}
\sigma_t / c, & \text{if success rate } > 1/5 \\
\sigma_t \cdot c, & \text{if success rate } < 1/5
\end{cases}
$$

where $c > 1$ is a scaling factor.

------------------------------------------------------------------------

**Evolutionary Strategies Algorithm**

1.  Initialize a population of $\lambda$ solutions with random parameters.
2.  Set mutation step size $\sigma$.
3.  Repeat for max iterations:
    -   Generate $\lambda$ offspring by mutating parent solutions.
    -   Evaluate fitness (SSE) of each offspring.
    -   Select the best $\mu$ solutions for the next generation.
    -   Adapt mutation step size based on success rate.
4.  Return the best solution found.

------------------------------------------------------------------------

```{r}
# Load required library
library(DEoptim)

# Define a numerically stable logistic function
safe_exp <- function(x) {
    return(ifelse(x > 700, Inf, exp(pmin(x, 700))))  # Prevent overflow
}

# Define the logistic growth model
nonlinear_model <- function(theta, x) {
    return(theta[1] / (1 + safe_exp(-theta[2] * (x - theta[3]))))
}

# Define the Sum of Squared Errors (SSE) function for optimization
sse <- function(theta, x, y) {
    predictions <- nonlinear_model(theta, x)
    return(sum((y - predictions)^2))
}

# Evolutionary Strategies Optimization (Using Differential Evolution)
es_optimization <- function(x, y) {
    # Define fitness function (minimize SSE)
    fitness_function <- function(theta) {
        return(sse(theta, x, y))
    }

    # Set parameter bounds
    lower_bounds <- c(0.1, 0.01, min(x))  # Ensure positive scaling
    upper_bounds <- c(max(y) * 2, 10, max(x))

    # Run Differential Evolution (mimicking ES)
    es_result <- DEoptim(
        fn = fitness_function,
        lower = lower_bounds,
        upper = upper_bounds,
        # 50 individuals, 200 generations, suppress iteration output
        DEoptim.control(NP = 50, itermax = 200, trace = F)  
    )

    return(es_result$optim$bestmem)  # Return optimized parameters
}

# Generate synthetic data
set.seed(123)
x <- seq(-5, 5, length.out = 100)
true_theta <- c(4, 1.5, 0)  # True parameters (A, B, C)
y <- nonlinear_model(true_theta, x) + rnorm(length(x), sd = 0.3)

# Run Evolutionary Strategies Optimization
estimated_theta <- es_optimization(x, y)

# Display results
cat("Estimated parameters (A, B, C) using Evolutionary Strategies:\n")
print(estimated_theta)
```

```{r fig-evolutionary-strategies, fig.cap="Evolutionary Strategies: Nonlinear Regression Optimization", fig.alt="Scatter plot titled 'Evolutionary Strategies: Nonlinear Regression Optimization' showing data points and a fitted curve. The x-axis is labeled 'x' and the y-axis is labeled 'y.' Black dots represent data points, and a red line represents the fitted curve, illustrating the optimization process. A legend indicates 'Data' and 'Fitted Curve.'", out.width="100%", fig.align='center'}
# Plot data and fitted curve
plot(
    x, y,
    main = "Evolutionary Strategies: Nonlinear Regression Optimization",
    pch = 19,
    cex = 0.5,
    xlab = "x",
    ylab = "y"
)
curve(
    nonlinear_model(estimated_theta, x),
    from = min(x),
    to = max(x),
    add = TRUE,
    col = "red",
    lwd = 2
)
legend(
    "topleft",
    legend = c("Data", "Fitted Curve"),
    pch = c(19, NA),
    lty = c(NA, 1),
    col = c("black", "red")
)
```

------------------------------------------------------------------------

### Linearization {#linearization-nonlinear-regression-optimization}

#### Taylor Series Approximation {#taylor-series-approximation-nonlinear-optimization}

Taylor Series Approximation is a fundamental tool in nonlinear optimization, enabling local approximation of complex functions using polynomial expansions. It is widely used to linearize nonlinear models, facilitate derivative-based optimization, and derive Newton-type methods.

Taylor series approximation is particularly useful when:

-   A nonlinear function is difficult to compute directly.

-   Optimization requires local gradient and curvature information.

-   A simpler, polynomial-based approximation improves computational efficiency.

------------------------------------------------------------------------

Given a differentiable function $f(\theta)$, its **Taylor series expansion** around a point $\theta_0$ is:

$$
f(\theta) = f(\theta_0) + f'(\theta_0)(\theta - \theta_0) + \frac{1}{2} f''(\theta_0)(\theta - \theta_0)^2 + \mathcal{O}((\theta - \theta_0)^3).
$$

For optimization, we often use:

1.  **First-order approximation (Linear Approximation):** $$
       f(\theta) \approx f(\theta_0) + f'(\theta_0)(\theta - \theta_0).
       $$
2.  **Second-order approximation (Quadratic Approximation):** $$
    f(\theta) \approx f(\theta_0) + f'(\theta_0)(\theta - \theta_0) + \frac{1}{2} f''(\theta_0)(\theta - \theta_0)^2.
    $$

For **gradient-based optimization**, we use the **Newton-Raphson update**:

$$
\theta^{(k+1)} = \theta^{(k)} - [H_f(\theta^{(k)})]^{-1} \nabla f(\theta^{(k)}),
$$

where:

-   $\nabla f(\theta)$ is the **gradient** (first derivative),

-   $H_f(\theta)$ is the **Hessian matrix** (second derivative).

------------------------------------------------------------------------

For nonlinear regression, we approximate the **Sum of Squared Errors (SSE)**:

$$
SSE(\theta) = \sum_{i=1}^{n} (y_i - f(x_i; \theta))^2.
$$

**1. First-Order Approximation (Gradient Descent)**

The **gradient** of SSE w.r.t. parameters $\theta$ is:

$$
\nabla SSE(\theta) = -2 \sum_{i=1}^{n} (y_i - f(x_i; \theta)) \nabla f(x_i; \theta).
$$

Using **first-order Taylor approximation**, we update parameters via **gradient descent**:

$$
\theta^{(k+1)} = \theta^{(k)} - \alpha \nabla SSE(\theta^{(k)}),
$$

where $\alpha$ is the **learning rate**.

**2. Second-Order Approximation (Newton's Method)**

The **Hessian matrix** of SSE is:

$$
H_{SSE}(\theta) = 2 \sum_{i=1}^{n} \nabla f(x_i; \theta) \nabla f(x_i; \theta)^T - 2 \sum_{i=1}^{n} (y_i - f(x_i; \theta)) H_f(x_i; \theta).
$$

The **Newton-Raphson update** becomes:

$$
\theta^{(k+1)} = \theta^{(k)} - H_{SSE}(\theta)^{-1} \nabla SSE(\theta).
$$

------------------------------------------------------------------------

```{r}
# Load required libraries
library(numDeriv)

# Define a numerically stable logistic function
safe_exp <- function(x) {
    return(ifelse(is.na(x) |
                      x > 700, Inf, exp(pmin(x, 700))))  # Prevent overflow
}

# Define the logistic growth model
nonlinear_model <- function(theta, x) {
    return(theta[1] / (1 + safe_exp(-theta[2] * (x - theta[3]))))
}

# Define the Sum of Squared Errors (SSE) function
sse <- function(theta, x, y) {
    predictions <- nonlinear_model(theta, x)
    return(sum((y - predictions) ^ 2, na.rm = TRUE))  # Avoid NA errors
}

# First-Order Approximation: Gradient Descent Optimization
gradient_descent <-
    function(x,
             y,
             alpha = 0.005,
             tol = 1e-6,
             max_iter = 5000) {
        theta <- c(2, 1, 0)  # Initial guess
        for (i in 1:max_iter) {
            grad_sse <-
                grad(function(t)
                    sse(t, x, y), theta)  # Compute gradient
            theta_new <-
                theta - alpha * grad_sse  # Update parameters
            
            if (sum(abs(theta_new - theta)) < tol)
                break  # Check convergence
            theta <- theta_new
        }
        return(theta)
    }

# Second-Order Approximation: Newton's Method with Regularization
newton_method <-
    function(x,
             y,
             tol = 1e-6,
             max_iter = 100,
             lambda = 1e-4) {
        theta <- c(2, 1, 0)  # Initial guess
        for (i in 1:max_iter) {
            grad_sse <-
                grad(function(t)
                    sse(t, x, y), theta)  # Compute gradient
            hessian_sse <-
                hessian(function(t)
                    sse(t, x, y), theta)  # Compute Hessian
            
            # Regularize Hessian to avoid singularity
            hessian_reg <-
                hessian_sse + lambda * diag(length(theta))
            
            # Ensure Hessian is invertible
            if (is.na(det(hessian_reg)) ||
                det(hessian_reg) < 1e-10) {
                message("Singular Hessian found; increasing regularization.")
                lambda <- lambda * 10  # Increase regularization
                next
            }
            
            # Newton update
            theta_new <- theta - solve(hessian_reg) %*% grad_sse
            
            if (sum(abs(theta_new - theta)) < tol)
                break  # Check convergence
            theta <- theta_new
        }
        return(theta)
    }

# Generate synthetic data
set.seed(123)
x <- seq(-5, 5, length.out = 100)
true_theta <- c(4, 1.5, 0)  # True parameters (A, B, C)
y <- nonlinear_model(true_theta, x) + rnorm(length(x), sd = 0.3)

# Run Gradient Descent
estimated_theta_gd <- gradient_descent(x, y)

# Run Newton's Method with Regularization
estimated_theta_newton <- newton_method(x, y)

# Display results
cat("Estimated parameters (A, B, C) using Gradient Descent:\n")
print(estimated_theta_gd)

cat("Estimated parameters (A, B, C) using Newton's Method:\n")
print(estimated_theta_newton)
```

```{r fig-taylor-series-approx-nonlinear, fig.cap="Taylor Series Approximation: Nonlinear Regression Optimization", fig.alt="Scatter plot titled 'Taylor Series Approximation: Nonlinear Regression Optimization' showing data points and two regression lines. The x-axis is labeled 'x' and the y-axis is labeled 'y.' A legend indicates black dots for data, a blue dashed line for 'Gradient Descent,' and a red solid line for 'Newton's Method (Regularized).' The red line closely follows the data trend, illustrating the optimization methods.", out.width="100%",fig.align='center'}
# Plot data and fitted curve
plot(
    x,
    y,
    main = "Taylor Series Approximation: Nonlinear Regression Optimization",
    pch = 19,
    cex = 0.5,
    xlab = "x",
    ylab = "y"
)
curve(
    nonlinear_model(estimated_theta_gd, x),
    from = min(x),
    to = max(x),
    add = TRUE,
    col = "blue",
    lwd = 2,
    lty = 2  # Dashed line to differentiate Gradient Descent
)
curve(
    nonlinear_model(estimated_theta_newton, x),
    from = min(x),
    to = max(x),
    add = TRUE,
    col = "red",
    lwd = 2
)
legend(
    "topleft",
    legend = c("Data", "Gradient Descent", "Newton's Method (Regularized)"),
    pch = c(19, NA, NA),
    lty = c(NA, 2, 1),
    col = c("black", "blue", "red")
)
```

------------------------------------------------------------------------

#### Log-Linearization {#log-linearization-nonlinear-optimization}

Log-Linearization is a mathematical technique used to transform nonlinear models into linear models by taking the logarithm of both sides. This transformation simplifies parameter estimation and enables the use of linear regression techniques on originally nonlinear functions.

Log-linearization is particularly useful when:

-   The model exhibits exponential, power-law, or logistic growth behavior.

-   Linear regression methods are preferred over nonlinear optimization.

-   A linearized version provides better interpretability and computational efficiency.

------------------------------------------------------------------------

A **nonlinear model** can often be expressed in the form:

$$
y = f(x; \theta).
$$

Applying a **log transformation**, we obtain:

$$
\log y = g(x; \theta),
$$

where $g(x; \theta)$ is now **linear in parameters**. We then estimate $\theta$ using [Ordinary Least Squares].

**Example 1: Exponential Model**

Consider an **exponential growth model**:

$$
y = A e^{Bx}.
$$

Taking the natural logarithm:

$$
\log y = \log A + Bx.
$$

This is now **linear in** $\log y$, allowing estimation via **linear regression**.

**Example 2: Power Law Model**

For a **power law function**:

$$
y = A x^B.
$$

Taking logs:

$$
\log y = \log A + B \log x.
$$

Again, this is **linearized**, making it solvable via **OLS regression**.

------------------------------------------------------------------------

**Log-Linearization Algorithm**

1.  Apply the logarithm transformation to the dependent variable.
2.  Transform the equation into a linear form.
3.  Use linear regression (OLS) to estimate parameters.
4.  Convert parameters back to original scale if necessary.

------------------------------------------------------------------------

```{r}
# Load required library
library(stats)

# Generate synthetic data for an exponential model
set.seed(123)
x <- seq(1, 10, length.out = 100)
true_A <- 2
true_B <- 0.3
y <- true_A * exp(true_B * x) + rnorm(length(x), sd = 0.5)

# Apply logarithmic transformation
log_y <- log(y)

# Fit linear regression model
log_linear_model <- lm(log_y ~ x)

# Extract estimated parameters
estimated_B <- coef(log_linear_model)[2]  # Slope in log-space
estimated_A <-
    exp(coef(log_linear_model)[1])  # Intercept (back-transformed)

# Display results
cat("Estimated parameters (A, B) using Log-Linearization:\n")
print(c(estimated_A, estimated_B))
```

```{r fig-log-linear-nonlinear, fig.cap="Log–Linearization: Nonlinear Regression Optimization", fig.alt="Scatter plot titled 'Log–Linearization: Nonlinear Regression Optimization' showing data points and a fitted log-linear model. The x-axis is labeled 'x' ranging from 0 to 10, and the y-axis is labeled 'y' ranging from 0 to 40. Black dots represent data points, and a red line represents the fitted log-linear model. A legend indicates the symbols for data and the fitted model.", out.width="100%", fig.align='center'}
# Plot data and fitted curve
plot(
    x,
    y,
    main = "Log-Linearization: Nonlinear Regression Optimization",
    pch = 19,
    cex = 0.5,
    xlab = "x",
    ylab = "y"
)
curve(
    estimated_A * exp(estimated_B * x),
    from = min(x),
    to = max(x),
    add = TRUE,
    col = "red",
    lwd = 2
)
legend(
    "topleft",
    legend = c("Data", "Fitted Log-Linear Model"),
    pch = c(19, NA),
    lty = c(NA, 1),
    col = c("black", "red")
)
```

------------------------------------------------------------------------

### Hybrid {#hybrid-nonlinear-regression-optimization}

#### Adaptive Levenberg-Marquardt {#adaptive-levenberg-marquardt}

The **Levenberg-Marquardt Algorithm (LMA)** is a powerful **nonlinear least squares optimization method** that adaptively combines:

-   [Gauss-Newton Algorithm] for fast convergence near the solution.

-   [Steepest Descent (Gradient Descent)](#steepest-descent) for stability when far from the solution.

The Adaptive Levenberg-Marquardt Algorithm further adjusts the damping parameter $\tau$ dynamically, making it more efficient in practice.

------------------------------------------------------------------------

Given an objective function **Sum of Squared Errors (SSE)**:

$$
SSE(\theta) = \sum_{i=1}^{n} (y_i - f(x_i; \theta))^2.
$$

The **update rule for LMA** is:

$$
\hat{\theta}^{(j+1)} = \hat{\theta}^{(j)} - \alpha_j [\mathbf{F}(\hat{\theta}^{(j)})' \mathbf{F}(\hat{\theta}^{(j)}) + \tau \mathbf{I}_{p \times p}]\frac{\partial \mathbf{Q}(\hat{\theta}^{(j)})}{\partial \theta}.
$$

where:

-   $\tau$ is the **adaptive damping parameter**.

-   $\mathbf{I}_{p \times p}$ is the **identity matrix**.

-   $\mathbf{F}(\hat{\theta}^{(j)})$ is the **Jacobian matrix** of partial derivatives.

-   $\frac{\partial \mathbf{Q}(\hat{\theta}^{(j)})}{\partial \theta}$ is the **gradient vector**.

-   $\alpha_j$ is the **learning rate**.

The key **adaptation rule for** $\tau$:

-   **If the new step decreases SSE**, **reduce** $\tau$: $$
      \tau \gets \tau / 10.
      $$

-   **Otherwise, increase** $\tau$ to ensure stability: $$
      \tau \gets 10\tau.
      $$

This adjustment **ensures a balance between stability and efficiency**.

------------------------------------------------------------------------

**Adaptive Levenberg-Marquardt Algorithm**

1.  Initialize parameters $\theta_0$, damping factor $\tau$.
2.  Compute Jacobian $\mathbf{F}(\hat{\theta}^{(j)})$.
3.  Compute step direction using modified Gauss-Newton update.
4.  Adjust $\tau$ dynamically:
    -   Decrease $\tau$ if SSE improves.
    -   Increase $\tau$ if SSE worsens.
5.  Repeat until convergence.

------------------------------------------------------------------------

```{r}
# Load required libraries
library(numDeriv)

# Define a numerically stable logistic function
safe_exp <- function(x) {
    return(ifelse(x > 700, Inf, exp(pmin(x, 700))))  # Prevent overflow
}

# Define the logistic growth model
nonlinear_model <- function(theta, x) {
    return(theta[1] / (1 + safe_exp(-theta[2] * (x - theta[3]))))
}

# Define the Sum of Squared Errors (SSE) function
sse <- function(theta, x, y) {
    predictions <- nonlinear_model(theta, x)
    return(sum((y - predictions) ^ 2))
}

# Adaptive Levenberg-Marquardt Optimization
adaptive_lm_optimization <-
    function(x, y, tol = 1e-6, max_iter = 100) {
        theta <- c(2, 1, 0)  # Initial parameter guess
        tau <- 1e-3  # Initial damping parameter
        alpha <- 1  # Step size scaling
        iter <- 0
        
        while (iter < max_iter) {
            iter <- iter + 1
            
            # Compute Jacobian numerically
            J <- jacobian(function(t)
                nonlinear_model(t, x), theta)
            
            # Compute gradient of SSE
            residuals <- y - nonlinear_model(theta, x)
            grad_sse <- -2 * t(J) %*% residuals
            
            # Compute Hessian approximation
            H <- 2 * t(J) %*% J + tau * diag(length(theta))
            
            # Compute parameter update step
            delta_theta <- solve(H, grad_sse)
            
            # Trial step
            theta_new <- theta - alpha * delta_theta
            
            # Compute SSE for new parameters
            if (sse(theta_new, x, y) < sse(theta, x, y)) {
                # Accept step, decrease tau
                theta <- theta_new
                tau <- tau / 10
            } else {
                # Reject step, increase tau
                tau <- tau * 10
            }
            
            # Check convergence
            if (sum(abs(delta_theta)) < tol)
                break
        }
        
        return(theta)
    }

# Generate synthetic data
set.seed(123)
x <- seq(-5, 5, length.out = 100)
true_theta <- c(4, 1.5, 0)  # True parameters (A, B, C)
y <- nonlinear_model(true_theta, x) + rnorm(length(x), sd = 0.3)

# Run Adaptive Levenberg-Marquardt Optimization
estimated_theta <- adaptive_lm_optimization(x, y)

# Display results
cat("Estimated parameters (A, B, C) using Adaptive Levenberg-Marquardt:\n")
print(estimated_theta)
```

```{r fig-adaptive-lm-nonlinear, fig.cap="Adaptive Levenberg–Marquardt: Nonlinear Regression Optimization", fig.alt="Scatter plot titled 'Adaptive Levenberg–Marquardt: Nonlinear Regression Optimization' showing data points and a fitted curve. The x-axis ranges from -4 to 4, and the y-axis ranges from 0 to 4. Black dots represent data points, and a red line represents the fitted curve. A legend indicates 'Data' and 'Fitted Curve.' ", out.width="100%", fig.align='center'}
# Plot data and fitted curve
plot(
    x,
    y,
    main = "Adaptive Levenberg-Marquardt: Nonlinear Regression Optimization",
    pch = 19,
    cex = 0.5,
    xlab = "x",
    ylab = "y"
)
curve(
    nonlinear_model(estimated_theta, x),
    from = min(x),
    to = max(x),
    add = TRUE,
    col = "red",
    lwd = 2
)
legend(
    "topleft",
    legend = c("Data", "Fitted Curve"),
    pch = c(19, NA),
    lty = c(NA, 1),
    col = c("black", "red")
)
```

------------------------------------------------------------------------

### Comparison of Nonlinear Optimizers

```{r, collapse=TRUE}
# ALL-IN-ONE R SCRIPT COMPARING MULTIPLE NONLINEAR-REGRESSION OPTIMIZERS

library(minpack.lm) # nlsLM (Levenberg-Marquardt)
library(dfoptim)    # Powell (nmk), Hooke-Jeeves
library(nloptr)     # trust-region reflective
library(GA)         # genetic algorithm
library(DEoptim)    # differential evolution
library(GenSA)      # simulated annealing
library(pso)        # particle swarm
library(MASS)       # for ginv fallback
library(microbenchmark)
library(ggplot2)
library(dplyr)

# -- 1) DEFINE MODELS (SIMPLE VS COMPLEX) ---

# 3-parameter logistic
f_logistic <- function(theta, x) {
  A <- theta[1]
  B <- theta[2]
  C <- theta[3]
  A / (1 + exp(-B * (x - C)))
}
sse_logistic <-
  function(theta, x, y)
    sum((y - f_logistic(theta, x)) ^ 2)

# 4-parameter "extended" model
f_complex <- function(theta, x) {
  A <- theta[1]
  B <- theta[2]
  C <- theta[3]
  D <- theta[4]
  A / (1 + exp(-B * (x - C))) + D * exp(-0.5 * x)
}
sse_complex <-
  function(theta, x, y)
    sum((y - f_complex(theta, x)) ^ 2)

# Generate synthetic data
set.seed(123)
n <- 100
x_data <- seq(-5, 5, length.out = n)
# "simple" scenario
true_theta_simple <- c(4, 1.5, 0)
y_data_simple <-
  f_logistic(true_theta_simple, x_data) + rnorm(n, sd = 0.3)
# "complex" scenario
true_theta_complex <- c(4, 1.2, -1, 0.5)
y_data_complex <-
  f_complex(true_theta_complex, x_data) + rnorm(n, sd = 0.3)

# -- 2) OPTIMIZERS (EXCEPT BISECTION) ----
#
# All methods share signature:
#   FUN(par, x, y, sse_fn, model_fn, lower=NULL, upper=NULL, ...)
# Some do not strictly use lower/upper if unconstrained.

# 2.1 Gauss–Newton
gauss_newton_fit <- function(par,
                             x,
                             y,
                             sse_fn,
                             model_fn,
                             lower = NULL,
                             upper = NULL,
                             max_iter = 100,
                             tol = 1e-6)
{
  theta <- par
  for (iter in seq_len(max_iter)) {
    eps <- 1e-6
    nP  <- length(theta)
    Fmat <- matrix(0, nrow = length(x), ncol = nP)
    for (p in seq_len(nP)) {
      pert <- theta
      pert[p] <- pert[p] + eps
      Fmat[, p] <-
        (model_fn(pert, x) - model_fn(theta, x)) / eps
    }
    r <- y - model_fn(theta, x)
    delta <- tryCatch(
      solve(t(Fmat) %*% Fmat, t(Fmat) %*% r),
      error = function(e) {
        # fallback to pseudoinverse
        MASS::ginv(t(Fmat) %*% Fmat) %*% (t(Fmat) %*% r)
      }
    )
    theta_new <- theta + delta
    if (sum(abs(theta_new - theta)) < tol)
      break
    theta <- theta_new
  }
  theta
}

# 2.2 Modified Gauss-Newton (with step halving)
modified_gauss_newton_fit <- function(par,
                                      x,
                                      y,
                                      sse_fn,
                                      model_fn,
                                      lower = NULL,
                                      upper = NULL,
                                      max_iter = 100,
                                      tol = 1e-6)
{
  theta <- par
  for (iter in seq_len(max_iter)) {
    eps <- 1e-6
    nP  <- length(theta)
    Fmat <- matrix(0, nrow = length(x), ncol = nP)
    for (p in seq_len(nP)) {
      pert <- theta
      pert[p] <- pert[p] + eps
      Fmat[, p] <-
        (model_fn(pert, x) - model_fn(theta, x)) / eps
    }
    r <- y - model_fn(theta, x)
    lhs <- t(Fmat) %*% Fmat
    rhs <- t(Fmat) %*% r
    delta <- tryCatch(
      solve(lhs, rhs),
      error = function(e)
        MASS::ginv(lhs) %*% rhs
    )
    sse_old <- sse_fn(theta, x, y)
    alpha <- 1
    for (k in 1:10) {
      new_sse <- sse_fn(theta + alpha * delta, x, y)
      if (new_sse < sse_old)
        break
      alpha <- alpha / 2
    }
    theta_new <- theta + alpha * delta
    if (sum(abs(theta_new - theta)) < tol)
      break
    theta <- theta_new
  }
  theta
}

# 2.3 Steepest Descent (Gradient Descent)
steepest_descent_fit <- function(par,
                                 x,
                                 y,
                                 sse_fn,
                                 model_fn,
                                 lower = NULL,
                                 upper = NULL,
                                 lr = 0.001,
                                 max_iter = 5000,
                                 tol = 1e-6)
{
  theta <- par
  for (iter in seq_len(max_iter)) {
    eps <- 1e-6
    f0  <- sse_fn(theta, x, y)
    grad <- numeric(length(theta))
    for (p in seq_along(theta)) {
      pert <- theta
      pert[p] <- pert[p] + eps
      grad[p] <- (sse_fn(pert, x, y) - f0) / eps
    }
    theta_new <- theta - lr * grad
    if (sum(abs(theta_new - theta)) < tol)
      break
    theta <- theta_new
  }
  theta
}

# 2.4 Levenberg–Marquardt (nlsLM)
lm_fit <- function(par,
                   x,
                   y,
                   sse_fn,
                   model_fn,
                   lower = NULL,
                   upper = NULL,
                   form = c("simple", "complex"))
{
  form <- match.arg(form)
  if (form == "simple") {
    fit <- nlsLM(y ~ A / (1 + exp(-B * (x - C))),
                 start = list(A = par[1],
                              B = par[2],
                              C = par[3]))
  } else {
    fit <- nlsLM(y ~ A / (1 + exp(-B * (x - C))) + D * exp(-0.5 * x),
                 start = list(
                   A = par[1],
                   B = par[2],
                   C = par[3],
                   D = par[4]
                 ))
  }
  coef(fit)
}

# 2.5 Newton–Raphson (with numeric Hessian, fallback if singular)
newton_raphson_fit <- function(par,
                               x,
                               y,
                               sse_fn,
                               model_fn,
                               lower = NULL,
                               upper = NULL,
                               max_iter = 50,
                               tol = 1e-6)
{
  theta <- par
  for (i in seq_len(max_iter)) {
    eps <- 1e-6
    f0  <- sse_fn(theta, x, y)
    grad <- numeric(length(theta))
    for (p in seq_along(theta)) {
      pert <- theta
      pert[p] <- pert[p] + eps
      grad[p] <- (sse_fn(pert, x, y) - f0) / eps
    }
    Hess <- matrix(0, length(theta), length(theta))
    for (p in seq_along(theta)) {
      pert_p <- theta
      pert_p[p] <- pert_p[p] + eps
      f_p <- sse_fn(pert_p, x, y)
      for (q in seq_along(theta)) {
        pert_q <- pert_p
        pert_q[q] <- pert_q[q] + eps
        Hess[p, q] <- (sse_fn(pert_q, x, y) -
                         f_p - (f0 - sse_fn(theta, x, y))) / (eps ^
                                                                2)
      }
    }
    delta <- tryCatch(
      solve(Hess, grad),
      error = function(e)
        MASS::ginv(Hess) %*% grad
    )
    theta_new <- theta - delta
    if (sum(abs(theta_new - theta)) < tol)
      break
    theta <- theta_new
  }
  theta
}

# 2.6 Quasi–Newton (BFGS via optim)
quasi_newton_fit <- function(par,
                             x,
                             y,
                             sse_fn,
                             model_fn,
                             lower = NULL,
                             upper = NULL)
{
  fn <- function(pp)
    sse_fn(pp, x, y)
  res <- optim(par, fn, method = "BFGS")
  res$par
}

# 2.7 Trust-region reflective (nloptr)
trust_region_fit <- function(par,
                             x,
                             y,
                             sse_fn,
                             model_fn,
                             lower = NULL,
                             upper = NULL)
{
  # numeric gradient
  grad_numeric <- function(pp, eps = 1e-6) {
    g  <- numeric(length(pp))
    f0 <- sse_fn(pp, x, y)
    for (i in seq_along(pp)) {
      p2 <- pp
      p2[i] <- p2[i] + eps
      g[i] <- (sse_fn(p2, x, y) - f0) / eps
    }
    g
  }
  eval_f <- function(pp) {
    val <- sse_fn(pp, x, y)
    gr  <- grad_numeric(pp)
    list(objective = val, gradient = gr)
  }
  lb <- if (is.null(lower))
    rep(-Inf, length(par))
  else
    lower
  ub <- if (is.null(upper))
    rep(Inf, length(par))
  else
    upper
  res <- nloptr(
    x0 = par,
    eval_f = eval_f,
    lb = lb,
    ub = ub,
    opts = list(
      algorithm = "NLOPT_LD_TNEWTON",
      maxeval = 500,
      xtol_rel = 1e-6
    )
  )
  res$solution
}

# 2.8 Grid search
grid_search_fit <- function(par,
                            x,
                            y,
                            sse_fn,
                            model_fn,
                            lower = NULL,
                            upper = NULL,
                            grid_defs = NULL)
{
  if (is.null(grid_defs))
    stop("Must provide grid_defs for multi-parameter grid search.")
  g <- expand.grid(grid_defs)
  g$SSE <-
    apply(g, 1, function(rowp)
      sse_fn(as.numeric(rowp), x, y))
  best_idx <- which.min(g$SSE)
  as.numeric(g[best_idx, seq_along(grid_defs)])
}

# 2.9 Nelder-Mead
nelder_mead_fit <- function(par,
                            x,
                            y,
                            sse_fn,
                            model_fn,
                            lower = NULL,
                            upper = NULL)
{
  fn <- function(pp)
    sse_fn(pp, x, y)
  res <- optim(par, fn, method = "Nelder-Mead")
  res$par
}

# 2.10 Powell’s method (dfoptim::nmk for unconstrained)
powell_fit <-
  function(par,
           x,
           y,
           sse_fn,
           model_fn,
           lower = NULL,
           upper = NULL) {
    fn <- function(pp)
      sse_fn(pp, x, y)
    dfoptim::nmk(par, fn)$par
  }

# 2.11 Hooke-Jeeves (dfoptim::hjkb)
hooke_jeeves_fit <-
  function(par,
           x,
           y,
           sse_fn,
           model_fn,
           lower = NULL,
           upper = NULL) {
    fn <- function(pp)
      sse_fn(pp, x, y)
    dfoptim::hjkb(par, fn)$par
  }

# 2.12 Random Search
random_search_fit <- function(par,
                              x,
                              y,
                              sse_fn,
                              model_fn,
                              lower,
                              upper,
                              max_iter = 2000,
                              ...)
{
  best_par <- NULL
  best_sse <- Inf
  dimp <- length(lower)
  for (i in seq_len(max_iter)) {
    candidate <- runif(dimp, min = lower, max = upper)
    val <- sse_fn(candidate, x, y)
    if (val < best_sse) {
      best_sse <- val
      best_par <- candidate
    }
  }
  best_par
}

# 2.13 Differential Evolution (DEoptim)
diff_evo_fit <- function(par,
                         x,
                         y,
                         sse_fn,
                         model_fn,
                         lower,
                         upper,
                         max_iter = 100,
                         ...)
{
  fn <- function(v)
    sse_fn(v, x, y)
  out <- DEoptim(fn,
                 lower = lower,
                 upper = upper,
                 DEoptim.control(NP = 50, itermax = max_iter, trace = F))
  out$optim$bestmem
}

# 2.14 Simulated Annealing (GenSA)
sim_anneal_fit <- function(par,
                           x,
                           y,
                           sse_fn,
                           model_fn,
                           lower = NULL,
                           upper = NULL,
                           ...)
{
  fn <- function(pp)
    sse_fn(pp, x, y)
  lb <- if (is.null(lower))
    rep(-Inf, length(par))
  else
    lower
  ub <- if (is.null(upper))
    rep(Inf, length(par))
  else
    upper
  # GenSA requires: GenSA(par, fn, lower, upper, control=list(...))
  out <-
    GenSA(
      par,
      fn,
      lower = lb,
      upper = ub,
      control = list(max.call = 10000)
    )
  out$par
}

# 2.15 Genetic Algorithm (GA)
genetic_fit <- function(par,
                        x,
                        y,
                        sse_fn,
                        model_fn,
                        lower,
                        upper,
                        max_iter = 100,
                        ...)
{
  fitness_fun <- function(pp)
    - sse_fn(pp, x, y)
  gares <- ga(
    type = "real-valued",
    fitness = fitness_fun,
    lower = lower,
    upper = upper,
    popSize = 50,
    maxiter = max_iter,
    run = 50
  )
  gares@solution[1,]
}

# 2.16 Particle Swarm (pso)
particle_swarm_fit <- function(par,
                               x,
                               y,
                               sse_fn,
                               model_fn,
                               lower,
                               upper,
                               max_iter = 100,
                               ...)
{
  fn <- function(pp)
    sse_fn(pp, x, y)
  res <- psoptim(
    par = (lower + upper) / 2,
    fn = fn,
    lower = lower,
    upper = upper,
    control = list(maxit = max_iter)
  )
  res$par
}


# -- 3) RUN METHOD WRAPPER ---
run_method <- function(method_name,
                       FUN,
                       par_init,
                       x,
                       y,
                       sse_fn,
                       model_fn,
                       lower = NULL,
                       upper = NULL,
                       ...)
{
  mb <- microbenchmark(result = {
    out <- FUN(par_init, x, y, sse_fn, model_fn, lower, upper, ...)
    out
  }, times = 1)
  final_par <-
    FUN(par_init, x, y, sse_fn, model_fn, lower, upper, ...)
  if (is.null(final_par)) {
    # e.g. placeholders that return NULL
    return(data.frame(
      Method = method_name,
      Parameters = "N/A",
      SSE = NA,
      Time_ms = NA
    ))
  }
  data.frame(
    Method     = method_name,
    Parameters = paste(round(final_par, 4), collapse = ", "),
    SSE        = round(sse_fn(final_par, x, y), 6),
    Time_ms    = median(mb$time) / 1e6
  )
}

# -- 4) MASTER FUNCTION TO COMPARE ALL METHODS (SIMPLE / COMPLEX) ---

compare_all_methods <- function(is_complex = FALSE) {
  if (!is_complex) {
    # SIMPLE (3-param logistic)
    x <- x_data
    y <- y_data_simple
    sse_fn   <- sse_logistic
    model_fn <- f_logistic
    init_par <- c(3, 1, 0.5)
    grid_defs <- list(
      A = seq(2, 6, length.out = 10),
      B = seq(0.5, 2, length.out = 10),
      C = seq(-1, 1, length.out = 10)
    )
    lower <- c(1, 0.1,-3)
    upper <- c(6, 3,  3)
    lm_form <- "simple"
  } else {
    # COMPLEX (4-param model)
    x <- x_data
    y <- y_data_complex
    sse_fn   <- sse_complex
    model_fn <- f_complex
    init_par <- c(3, 1,-0.5, 0.2)
    grid_defs <- list(
      A = seq(2, 6, length.out = 8),
      B = seq(0.5, 2, length.out = 8),
      C = seq(-2, 2, length.out = 8),
      D = seq(0, 2, length.out = 8)
    )
    lower <- c(1, 0.1,-3, 0)
    upper <- c(6, 3,  3, 2)
    lm_form <- "complex"
  }
  
  # RUN each method
  out <- bind_rows(
    run_method(
      "Gauss-Newton",
      gauss_newton_fit,
      init_par,
      x,
      y,
      sse_fn,
      model_fn
    ),
    run_method(
      "Modified Gauss-Newton",
      modified_gauss_newton_fit,
      init_par,
      x,
      y,
      sse_fn,
      model_fn
    ),
    run_method(
      "Steepest Descent",
      steepest_descent_fit,
      init_par,
      x,
      y,
      sse_fn,
      model_fn
    ),
    run_method(
      "Levenberg-Marquardt (nlsLM)",
      lm_fit,
      init_par,
      x,
      y,
      sse_fn,
      model_fn,
      form = lm_form
    ),
    run_method(
      "Newton-Raphson",
      newton_raphson_fit,
      init_par,
      x,
      y,
      sse_fn,
      model_fn
    ),
    run_method(
      "Quasi-Newton (BFGS)",
      quasi_newton_fit,
      init_par,
      x,
      y,
      sse_fn,
      model_fn
    ),
    run_method(
      "Trust-region Reflective",
      trust_region_fit,
      init_par,
      x,
      y,
      sse_fn,
      model_fn,
      lower,
      upper
    ),
    run_method(
      "Grid Search",
      grid_search_fit,
      NULL,
      x,
      y,
      sse_fn,
      model_fn,
      grid_defs = grid_defs
    ),
    run_method(
      "Nelder-Mead",
      nelder_mead_fit,
      init_par,
      x,
      y,
      sse_fn,
      model_fn
    ),
    run_method("Powell's method",
               powell_fit,
               init_par,
               x,
               y,
               sse_fn,
               model_fn),
    run_method(
      "Hooke-Jeeves",
      hooke_jeeves_fit,
      init_par,
      x,
      y,
      sse_fn,
      model_fn
    ),
    run_method(
      "Random Search",
      random_search_fit,
      NULL,
      x,
      y,
      sse_fn,
      model_fn,
      lower,
      upper,
      max_iter = 1000
    ),
    run_method(
      "Differential Evolution",
      diff_evo_fit,
      NULL,
      x,
      y,
      sse_fn,
      model_fn,
      lower,
      upper,
      max_iter = 50
    ),
    run_method(
      "Simulated Annealing",
      sim_anneal_fit,
      init_par,
      x,
      y,
      sse_fn,
      model_fn,
      lower,
      upper
    ),
    run_method(
      "Genetic Algorithm",
      genetic_fit,
      NULL,
      x,
      y,
      sse_fn,
      model_fn,
      lower,
      upper,
      max_iter = 50
    ),
    run_method(
      "Particle Swarm",
      particle_swarm_fit,
      NULL,
      x,
      y,
      sse_fn,
      model_fn,
      lower,
      upper,
      max_iter = 50
    )
  )
  out
}

# -- 5) RUN & VISUALIZE ----

# Compare "simple" logistic (3 params)
results_simple  <- compare_all_methods(is_complex = FALSE)
results_simple$Problem <- "Simple"

# Compare "complex" (4 params)
results_complex <- compare_all_methods(is_complex = TRUE)
results_complex$Problem <- "Complex"

# Combine
all_results <- rbind(results_simple, results_complex)
# print(all_results)
# DT::datatable(all_results)


```

```{r fig-comparison-sse-by-method, fig.cap="Comparison of SSE by Method and Problem Complexity", fig.alt="Bar chart comparing the Log(Sum of Squared Errors) for different methods and problem complexities. The x-axis lists various methods such as Differential Evolution, Genetic Algorithm, and Particle Swarm. The y-axis represents the Log(Sum of Squared Errors). Bars are color-coded: red for complex problems and blue for simple problems. The chart highlights significant differences in error values across methods, with the Steepest Descent method showing a notably high error for complex problems.", out.width="100%", fig.align='center'}
# Example: SSE by method & problem
ggplot(all_results, aes(x = Method, y = log(SSE), fill = Problem)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal(base_size = 11) +
  labs(title = "Comparison of SSE by Method and Problem Complexity",
       x = "", y = "Log(Sum of Squared Errors)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

```{r fig-comparison-computation-nonlinear, fig.cap="Comparison of Computation Time by Method and Problem Complexity", fig.alt="Bar chart comparing computation time by method and problem complexity. The x-axis lists various methods such as Differential Evolution, Gauss-Newton, and others. The y-axis represents time in milliseconds. Bars are color-coded: red for complex problems and blue for simple problems. Notably, the Trust-region Reflective method shows the highest time for complex problems. A legend on the right indicates the color coding for problem complexity.", out.width="100%", fig.align='center'}
# Example: Time (ms) by method & problem
ggplot(all_results, aes(x = Method, y = Time_ms, fill = Problem)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal(base_size = 11) +
  labs(title = "Comparison of Computation Time by Method and Problem Complexity",
       x = "", y = "Time (ms)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

------------------------------------------------------------------------
