# Variable Transformation

## Continuous Variables

Transforming continuous variables can be useful for various reasons, including:

-   **Changing the scale of variables** to make them more interpretable or comparable.
-   **Reducing skewness** to approximate a normal distribution, which can improve statistical inference.
-   **Stabilizing variance** in cases of heteroskedasticity.
-   **Enhancing interpretability** in business applications (e.g., logarithmic transformations for financial data).

------------------------------------------------------------------------

### Standardization (Z-score Normalization)

A common transformation to center and scale data:

$$
x_i' = \frac{x_i - \bar{x}}{s}
$$

where:

-   $x_i$ is the original value,
-   $\bar{x}$ is the sample mean,
-   $s$ is the sample standard deviation.

**When to Use:**

-   When variables have different units of measurement and need to be on a common scale.

-   When a few large numbers dominate the dataset.

------------------------------------------------------------------------

### Min-Max Scaling (Normalization)

Rescales data to a fixed range, typically $[0,1]$:

$$
x_i' = \frac{x_i - x_{\min}}{x_{\max} - x_{\min}}
$$

**When to Use:**

-   When working with fixed-interval data (e.g., percentages, proportions).

-   When preserving relative relationships between values is important.

-   Caution: This method is sensitive to outliers, as extreme values determine the range.

------------------------------------------------------------------------

### Square Root and Cube Root Transformations

Useful for handling **positive skewness** and **heteroskedasticity**:

-   **Square root:** Reduces moderate skewness and variance.
-   **Cube root:** Works on more extreme skewness and allows negative values.

**Common Use Cases:**

-   **Frequency count data** (e.g., website visits, sales transactions).

-   **Data with many small values or zeros** (e.g., income distributions in microfinance).

------------------------------------------------------------------------

### Logarithmic Transformation {#sec-logarithmic-transformation}

Logarithmic transformations are particularly useful for handling highly skewed data. They compress large values while expanding small values, which helps with heteroskedasticity and normality assumptions.

#### Common Log Transformations

| Formula                                             | When to Use                         |
|----------------------------------------|--------------------------------|
| $x_i' = \log(x_i)$                                  | When all values are positive.       |
| $x_i' = \log(x_i + 1)$                              | When data contains zeros.           |
| $x_i' = \log(x_i + c)$                              | Choosing $c$ depends on context.    |
| $x_i' = \frac{x_i}{|x_i|} \log |x_i|$               | When data contains negative values. |
| $x_i'^\lambda = \log(x_i + \sqrt{x_i^2 + \lambda})$ | Generalized log transformation.     |

Selecting the constant $c$ is critical:

-   If $c$ is too large, it can obscure the true nature of the data.
-   If $c$ is too small, the transformation might not effectively reduce skewness.

From a statistical modeling perspective:

-   **For inference-based models**, the choice of $c$ can significantly impact the fit. See [@ekwaru2018overlooked].
-   **In causal inference** (e.g., DID, IV), improper log transformations (e.g., logging zero values) can introduce bias [@chen2024logs].

#### When is Log Transformation Problematic?

-   When **zero values have a meaningful interpretation** (e.g., income of unemployed individuals).
-   When **data are censored** (e.g., income data truncated at reporting thresholds).
-   When **measurement error exists** (e.g., rounding errors from survey responses).

If zeros are **small but meaningful** (e.g., revenue from startups), then using $\log(x + c)$ may be acceptable.

------------------------------------------------------------------------

```{r}
library(tidyverse)

# Load dataset
cars = datasets::cars

# Original values
head(cars$speed)

# Log transformation (basic)
log(cars$speed) %>% head()

# Log transformation for zero-inflated data
log1p(cars$speed) %>% head()
```

### Exponential Transformation

The exponential transformation is useful when data exhibit negative skewness or when an underlying logarithmic trend is suspected, such as in survival analysis and decay models.

When to Use:

-   Negatively skewed distributions.

-   Processes that follow an exponential trend (e.g., population growth, depreciation of assets).

------------------------------------------------------------------------

### Power Transformation

Power transformations help adjust skewness, particularly for negatively skewed data.

**When to Use:**

-   When variables have a negatively skewed distribution.

-   When the relationship between variables is non-linear.

Common power transformations include:

-   Square transformation: $x^2$ (moderate adjustment).

-   Cubic transformation: $x^3$ (stronger adjustment).

-   Fourth-root transformation: $x^{1/4}$ (more subtle than square root).

------------------------------------------------------------------------

### Inverse (Reciprocal) Transformation

The inverse transformation is useful for handling platykurtic (flat) distributions or positively skewed data.

**Formula:**

$$
x_i' = \frac{1}{x_i}
$$

**When to Use:**

-   Reducing extreme values in positively skewed distributions.

-   Ratio data (e.g., speed = distance/time).

-   When the variable has a natural lower bound (e.g., time to completion).

```{r}
# data(cars)
cars = datasets::cars

# Original distribution
head(cars$dist)
plot(cars$dist)

# Reciprocal transformation
plot(1 / cars$dist)
```

### Hyperbolic Arcsine Transformation

The arcsinh (inverse hyperbolic sine) transformation is useful for handling proportion variables (0-1) and skewed distributions. It behaves similarly to the logarithmic transformation but has the advantage of handling zero and negative values.

**Formula:**

$$
\text{arcsinh}(Y) = \log(\sqrt{1 + Y^2} + Y)
$$

**When to Use:**

-   Proportion variables (e.g., market share, probability estimates).

-   Data with extreme skewness where log transformation is problematic.

-   Variables containing zeros or negative values (unlike log, arcsinh handles zeros naturally).

-   Alternative to log transformation for handling zeros.

```{r}
# Visualize original distribution 
cars$dist %>% hist() 
# Alternative histogram  
cars$dist %>% MASS::truehist()  

# Apply arcsinh transformation 
as_dist <- bestNormalize::arcsinh_x(cars$dist) 
as_dist
as_dist$x.t %>% hist()

```

| Paper                  | Interpretation |
|------------------------|----------------|
| @azoulay2019does       | Elasticity     |
| @faber2019tourism      | Percentage     |
| @hjort2019arrival      | Percentage     |
| @johnson2020regulation | Percentage     |
| @beerli2021abolition   | Percentage     |
| @norris2021effects     | Percentage     |
| @berkouwer2022credit   | Percentage     |
| @cabral2022demand      | Elasticity     |
| @carranza2022job       | Percentage     |
| @mirenda2022economic   | Percentage     |

Consider a simple regression model: $$ Y = \beta X + \epsilon $$ When both $Y$ and $X$ are transformed:

-   The coefficient estimate $\beta$ represents **elasticity**: *A 1% increase in* $X$ leads to a $\beta$% change in $Y$.

When only $Y$ is transformed:

-   The coefficient estimate represents a **percentage change in** $Y$ for a one-unit change in $X$.

This makes the arcsinh transformation particularly valuable for **log-linear models** where zero values exist.

### Ordered Quantile Normalization (Rank-Based Transformation)

The Ordered Quantile Normalization (OQN) technique transforms data into a normal distribution using rank-based methods [@bartlett1947use].

**Formula:**

$$
x_i' = \Phi^{-1} \left( \frac{\text{rank}(x_i) - 1/2}{\text{length}(x)} \right)
$$

where $\Phi^{-1}$ is the inverse normal cumulative distribution function.

**When to Use:**

-   When data are **heavily skewed** or **contain extreme values**.

-   When normality is required for **parametric tests**.

```{r}
ord_dist <- bestNormalize::orderNorm(cars$dist)
ord_dist
ord_dist$x.t %>% hist()

```

### Lambert W x F Transformation

The Lambert W transformation is a more advanced method that normalizes data by removing skewness and heavy tails.

**When to Use:**

-   When traditional transformations (e.g., log, Box-Cox) fail.

-   When dealing with heavy-tailed distributions.

```{r}
cars = datasets::cars
head(cars$dist)
cars$dist %>% hist()

# Apply Lambert W transformation
l_dist <- LambertW::Gaussianize(cars$dist)
l_dist %>% hist()
```

### Inverse Hyperbolic Sine Transformation

The Inverse Hyperbolic Sine (IHS) transformation is similar to the log transformation but handles zero and negative values [@johnson1949].

**Formula:**

$$
f(x,\theta) = \frac{\sinh^{-1} (\theta x)}{\theta} = \frac{\log(\theta x + (\theta^2 x^2 + 1)^{1/2})}{\theta}
$$

**When to Use:**

-   When data contain zeros or negative values.

-   Alternative to log transformation in economic and financial modeling.

### Box-Cox Transformation {#sec-box-cox-transformation}

The Box-Cox transformation is a power transformation designed to improve linearity and normality [@manly1976exponential; @bickel1981analysis; @box1981analysis].

**Formula:**

$$
x_i'^\lambda = \begin{cases} \frac{x_i^\lambda-1}{\lambda} & \text{if } \lambda \neq 0\\ \log(x_i) & \text{if } \lambda = 0 \end{cases}
$$

**When to Use:**

-   To fix non-linearity in the error terms of regression models.

-   When data are strictly positive

```{r}
library(MASS)
# data(cars)
cars = datasets::cars
mod <- lm(cars$speed ~ cars$dist, data = cars)

# Check residuals
plot(mod)

# Find optimal lambda
bc <- boxcox(mod, lambda = seq(-3, 3))
best_lambda <- bc$x[which.max(bc$y)]

# Apply transformation
mod_lambda = lm(cars$speed ^ best_lambda ~ cars$dist, data = cars)
plot(mod_lambda)
```

For the **two-parameter Box-Cox transformation**, we use:

$$
x_i' (\lambda_1, \lambda_2) = \begin{cases} \frac{(x_i + \lambda_2)^{\lambda_1}-1}{\lambda_1} & \text{if } \lambda_1 \neq 0 \\ \log(x_i + \lambda_2) & \text{if } \lambda_1 = 0 \end{cases}
$$

```{r}
# Two-parameter Box-Cox transformation
two_bc <- geoR::boxcoxfit(cars$speed)
two_bc
plot(two_bc)
```

### Yeo-Johnson Transformation

Similar to [Box-Cox](#sec-box-cox-transformation) (when $\lambda = 1$), but allows for negative values.

**Formula:**

$$
x_i'^\lambda = \begin{cases} \frac{(x_i+1)^\lambda -1}{\lambda} & \text{if } \lambda \neq0, x_i \ge 0 \\ \log(x_i + 1) & \text{if } \lambda = 0, x_i \ge 0 \\ \frac{-[(-x_i+1)^{2-\lambda}-1]}{2 - \lambda} & \text{if } \lambda \neq 2, x_i <0 \\ -\log(-x_i + 1) & \text{if } \lambda = 2, x_i <0 \end{cases}
$$

```{r}
# data(cars)
cars = datasets::cars
yj_speed <- bestNormalize::yeojohnson(cars$speed)
yj_speed$x.t %>% hist()

```

### RankGauss Transformation

A rank-based transformation that maps values to a normal distribution.

When to Use:

-   To handle skewed data while preserving rank order.

### Automatically Choosing the Best Transformation

The `bestNormalize` package selects the **best transformation** for a given dataset.

```{r}
bestdist <- bestNormalize::bestNormalize(cars$dist)
bestdist$x.t %>% hist()
```

## Categorical Variables

Transforming categorical variables into numerical representations is essential for machine learning models and statistical analysis. The key objectives include:

-   Converting categorical data into a format suitable for numerical models.
-   Improving model interpretability and performance.
-   Handling high-cardinality categorical variables efficiently.

------------------------------------------------------------------------

There are multiple ways to transform categorical variables, each with its advantages and use cases. The choice depends on factors like **cardinality**, **ordinality**, and **model type**.

### One-Hot Encoding (Dummy Variables)

Creates binary indicator variables for each category.

**Formula:**\
For a categorical variable with $k$ unique values, create $k$ binary columns:

$$
x_i' = 
\begin{cases}
1 & \text{if } x_i = \text{category} \\
0 & \text{otherwise}
\end{cases}
$$

**When to Use:**

-   Low-cardinality categorical variables (e.g., "Red", "Blue", "Green").

-   Tree-based models (e.g., Random Forest, XGBoost).

-   Linear regression models (dummy variables prevent information loss).

```{r}
library(caret)

data(iris)
dummy_vars <- dummyVars(~ Species, data = iris)
one_hot_encoded <- predict(dummy_vars, newdata = iris)
head(one_hot_encoded)
```

### Label Encoding

Assigns integer values to categories.

**Formula:**\
If a categorical variable has $k$ unique values:

$$
\text{Category } \rightarrow \text{Integer}
$$

Example:

| Category | Encoded Value |
|----------|---------------|
| Red      | 1             |
| Blue     | 2             |
| Green    | 3             |

**When to Use:**

-   Ordinal categorical variables (e.g., "Low", "Medium", "High").

-   Neural networks (use embeddings instead of one-hot).

-   Memory-efficient encoding for high-cardinality features.

```{r}
iris$Species_encoded <- as.numeric(factor(iris$Species))
head(iris$Species_encoded)

```

### Feature Hashing (Hash Encoding)

Maps categories to a **fixed number of hash bins**, reducing memory usage.

**When to Use:**

-   High-cardinality categorical variables (e.g., user IDs, URLs).

-   Scenarios where an exact category match isn't needed.

-   Sparse models (e.g., text data in NLP).

```{r}
library(text2vec)
library(Matrix)

data(iris)

# Convert the 'Species' factor to character tokens
tokens <- word_tokenizer(as.character(iris$Species))

# Create an iterator over tokens
it <- itoken(tokens, progressbar = FALSE)

# Define the hash_vectorizer with a specified hash size (8 in this case)
vectorizer <- hash_vectorizer(hash_size = 8)

# Create a Document-Term Matrix (DTM) using the hashed features
hashed_dtm <- create_dtm(it, vectorizer)

# Inspect the first few rows of the hashed feature matrix
head(hashed_dtm)
```

-   **word_tokenizer**: This function splits the character vector into tokens. Since `iris$Species` is already a categorical variable with values like `"setosa"`, `"versicolor"`, and `"virginica"`, each value becomes a token.

-   **itoken**: Creates an iterator over the tokens.

-   **hash_vectorizer**: Sets up a hashing vectorizer that transforms tokens into a sparse feature space of size `2^3 = 8` (because `hash_size = 8` means $2^8$ bins; if you intend exactly 8 bins, you might adjust the parameter accordingly).

-   **create_dtm**: Builds the document-term matrix (which in this case is analogous to a feature matrix for each observation).

### Binary Encoding

Converts categories to binary representations and distributes them across multiple columns.

**Example:**\
For four categories ("A", "B", "C", "D"):

| Category | Binary Code | Encoded Columns |
|----------|-------------|-----------------|
| A        | 00          | 0, 0            |
| B        | 01          | 0, 1            |
| C        | 10          | 1, 0            |
| D        | 11          | 1, 1            |

**When to Use:**

-   High-cardinality categorical features (less memory than one-hot encoding).

-   Tree-based models (preserves some ordinal information).

```{r}
library(mltools)
library(data.table)

# Convert the Species column to a data.table and perform one-hot encoding
binary_encoded <- one_hot(as.data.table(iris[, "Species"]))
head(binary_encoded)
```

### Base-N Encoding (Generalized Binary Encoding)

Expands Binary Encoding to base $N$ instead of binary.

When to Use:

-   Similar to Binary Encoding, but allows for greater flexibility.

### Frequency Encoding

Replaces each category with its frequency (proportion) in the dataset.

**Formula:** $$
x_i' = \frac{\text{count}(x_i)}{\text{total count}}
$$ **When to Use:**

-   High-cardinality categorical variables.

-   Feature engineering for boosting algorithms (e.g., LightGBM).

```{r}
freq_encoding <- table(iris$Species) / length(iris$Species)
iris$Species_freq <-
    iris$Species %>% as.character() %>% map_dbl(~ freq_encoding[.])
head(iris$Species_freq)
```

### Target Encoding (Mean Encoding)

Encodes categories using the mean of the target variable.

**Formula:** $$
x_i' = E[Y | X = x_i]
$$ **When to Use:**

-   Predictive models with categorical features strongly correlated with the target.

-   High-cardinality categorical variables.

-   Risk: Can lead to data leakage (use cross-validation).

```{r}
library(data.table)
iris_dt <- as.data.table(iris)
iris_dt[, Species_mean := mean(Sepal.Length), by = Species]
head(iris_dt$Species_mean)

```

### Ordinal Encoding

Maps categories to **ordered integer values** based on logical ranking.

**Example:**

| Category | Ordinal Encoding |
|----------|------------------|
| Low      | 1                |
| Medium   | 2                |
| High     | 3                |

**When to Use:**

-   Ordinal variables with meaningful order (e.g., satisfaction ratings).

```{r}
iris$Species_ordinal <-
    as.numeric(factor(iris$Species, 
                      levels = c("setosa", "versicolor", "virginica")))
head(iris$Species_ordinal)

```

### Weight of Evidence Encoding

-   **Concept:**\
    WoE is a method to convert categorical data into numerical values that capture the strength of the relationship between a feature (or category) and a binary outcome (like default vs. non-default).

-   **The Formula:** $$
    \text{WoE} = \log \left( \frac{P(X_i | Y=1)}{P(X_i | Y=0)} \right)
    $$

    -   $(X_i | Y=1)$: The probability (or proportion) of observing category $X_i$ given the positive outcome (e.g., a "good" credit event).

    -   $P(X_i | Y=0)$: The probability of observing category $X_i$ given the negative outcome (e.g., a "bad" credit event).

    -   **Logarithm:** Taking the log of the ratio gives us a symmetric scale where:

        -   A positive WoE indicates the category is more associated with the positive outcome.

        -   A negative WoE indicates the category is more associated with the negative outcome.

When and Why to Use WoE Encoding?

-   **Logistic Regression in Credit Scoring:**\
    Logistic regression models predict probabilities in terms of log-odds. WoE encoding aligns well with this because it essentially expresses how the odds of the positive outcome change with different categories. This is why it's popular in credit scoring models.

-   **Interpretability:**\
    The WoE transformation makes it easier to understand and interpret the relationship between each category of a variable and the outcome. Each category's WoE value tells you whether it increases or decreases the odds of a particular event (e.g., default).

Imagine you have a feature "Employment Status" with categories "Employed" and "Unemployed":

1.  **Calculate Proportions:**

    -   $P(\text{Employed} | Y=1) = 0.8$ (80% of good credit cases are employed)

    -   $P(\text{Employed} | Y=0) = 0.4$ (40% of bad credit cases are employed)

2.  **Compute WoE for "Employed":** $$
    \text{WoE}_{\text{Employed}} = \log \left( \frac{0.8}{0.4} \right) = \log(2) \approx 0.693
    $$ A positive value indicates that being employed increases the odds of a good credit outcome.

3.  **Repeat for "Unemployed":**\
    Suppose:

    -   $P(\text{Unemployed} | Y=1) = 0.2$

    -   $P(\text{Unemployed} | Y=0) = 0.6$ $$
        \text{WoE}_{\text{Unemployed}} = \log \left( \frac{0.2}{0.6} \right) = \log\left(\frac{1}{3}\right) \approx -1.099
        $$ A negative value indicates that being unemployed is associated with a higher likelihood of a bad credit outcome.

Why is WoE Valuable?

-   **Linear Relationship:**\
    When you plug these WoE values into a logistic regression, the model essentially adds these values linearly, which fits nicely with how logistic regression models the log-odds.

-   **Stability & Handling of Missing Values:**\
    WoE can also help in smoothing out fluctuations in categorical data, especially when there are many levels or some levels with few observations.

-   **Regulatory Acceptance:**\
    In industries like banking, WoE is widely accepted because of its clear interpretability, which is crucial for compliance and transparency in credit risk modeling.

```{r}
# Load required packages
library(dplyr)
library(knitr)

# Create a sample dataset
# We assume 100 good credit cases and 100 bad credit cases
# Good credit: 80 "Employed" and 20 "Unemployed"
# Bad credit: 40 "Employed" and 60 "Unemployed"
data <- data.frame(
  employment_status = c(rep("Employed", 80), rep("Unemployed", 20), 
                        rep("Employed", 40), rep("Unemployed", 60)),
  credit = c(rep(1, 100), rep(0, 100))
)

# Calculate counts for each category
woe_table <- data %>%
  group_by(employment_status) %>%
  summarise(
    good = sum(credit == 1),
    bad = sum(credit == 0)
  ) %>%
  # Calculate the distribution for good and bad credit cases
  mutate(
    dist_good = good / sum(good),
    dist_bad = bad / sum(bad),
    WoE = log(dist_good / dist_bad)
  )

# Print the WoE table
kable(woe_table)

# Merge the WoE values into the original data
data_woe <- data %>%
  left_join(woe_table %>% dplyr::select(employment_status, WoE), by = "employment_status")

head(data_woe)


# Fit a logistic regression model using WoE as predictor
model <- glm(credit ~ WoE, data = data_woe, family = binomial)

# Summary of the model
summary(model)
```

When you fit a logistic regression using the WoE-encoded variable, the model is essentially: $$
\log\left(\frac{P(Y=1)}{P(Y=0)}\right) = \beta_0 + \beta_1 \cdot \text{WoE}
$$ Here, **WoE** represents the Weight of Evidence value for a given category.

-   **Log Odds Change:**\
    $\beta_1$ indicates how much the log odds of a good credit outcome change for a one-unit increase in WoE. For example, if $\beta_1 = 0.5$, then a one-unit increase in WoE is associated with an increase of 0.5 in the log odds of having a good credit outcome.

-   **Odds Ratio:**\
    If you exponentiate $\beta_1$, you get the odds ratio. For instance, if $\beta_1 = 0.5$, then $\exp(0.5) \approx 1.65$. This means that for each one-unit increase in WoE, the odds of having a good credit outcome are multiplied by about 1.65.

Why is This Meaningful?

-   **Direct Link to the Data:**\
    The WoE value itself is a transformation of the original categorical variable that reflects the ratio of the proportions of good to bad outcomes for that category. By using WoE, you're directly incorporating this information into the model.

-   **Interpretability:**\
    The interpretation becomes intuitive:

    -   A **positive** WoE indicates that the category is more associated with a good outcome.

    -   A **negative** WoE indicates that the category is more associated with a bad outcome.

    Thus, if $\beta_1$ is positive, it suggests that as the category moves to one with a higher WoE (i.e., more favorable for a good outcome), the likelihood of a good outcome increases.

### Helmert Encoding

Compares each category against the mean of previous categories.

**When to Use:**

-   ANOVA models and categorical regression.

### Probability Ratio Encoding

Encodes categories using the probability ratio of the target variable.

### Backward Difference Encoding

Compares each category against the mean of all remaining categories.

### Leave-One-Out Encoding

Similar to target encoding, but excludes the current observation to avoid bias.

### James-Stein Encoding

A smoothed version of target encoding, reducing overfitting.

### M-Estimator Encoding

Uses a Bayesian prior to smooth target encoding.

### Thermometer Encoding

Similar to one-hot encoding, but retains ordinal structure.

### Choosing the Right Encoding Method

| Encoding Method    | Best for Low Cardinality | Best for High Cardinality | Handles Ordinality | Suitable for Tree Models | Suitable for Linear Models |
|------------|------------|------------|------------|------------|------------|
| One-Hot Encoding   | ✅ Yes                   | ❌ No                     | ❌ No              | ✅ Yes                   | ✅ Yes                     |
| Label Encoding     | ✅ Yes                   | ✅ Yes                    | ✅ Yes             | ❌ No                    | ✅ Yes                     |
| Target Encoding    | ✅ Yes                   | ✅ Yes                    | ❌ No              | ✅ Yes                   | ✅ Yes                     |
| Frequency Encoding | ✅ Yes                   | ✅ Yes                    | ❌ No              | ✅ Yes                   | ✅ Yes                     |
| Binary Encoding    | ✅ Yes                   | ✅ Yes                    | ❌ No              | ✅ Yes                   | ✅ Yes                     |
