---
title: "Untitled"
author: "Mike"
date: "11/11/2020"
---

# Bookdown cheat sheet
Took from professor Jenny to 

Here's where I park _little_ *examples* **for myself** about bookdown mechanics that I keep forgetting.

The bookdown book: <https://bookdown.org/yihui/bookdown/>  


## Math Expresssion/ Syntax

[Full list](http://tug.ctan.org/info/symbols/comprehensive/symbols-a4.pdf)

Aligning equations

```{r eval=FALSE}
\begin{align*}
a & = b \\
X &\sim {\sf Norm}(10, 3) \\
5 & \le 10
\end{align*}
```



\begin{align*}
a & = b \\
X &\sim {\sf Norm}(10, 3) \\
5 & \le 10
\end{align*}

Syntax | Notation
---|---
**Math** | 
`$\pm$`|$\pm$
`$\ge$`|$\ge$
`$\le$`|$\le$
`$\neq$`|$\neq$
`$\equiv$`|$\equiv$
`$^\circ$`|$^\circ$
`$\times$`|$\times$
`$\cdot$`|$\cdot$
`$\leq$`|$\leq$
`$\geq$`|$\geq$
`$\subset$`|$\subset$
`$\subseteq$`|$\subseteq$
`$\leftarrow$`|$\leftarrow$
`$\rightarrow$`|$\rightarrow$
`$\Leftarrow$`|$\Leftarrow$
`$\Rightarrow$`|$\Rightarrow$
`$\approx$`|$\approx$
`$\mathbb{R}$`|$\mathbb{R}$
`$\sum_{n=1}^{10} n^2$`|$\sum_{n=1}^{10} n^2$
`$$\sum_{n=1}^{10} n^2$$`|$$\sum_{n=1}^{10} n^2$$
`$x^{n}$`|$x^{n}$
`$x_{n}$`|$x_{n}$
`$\overline{x}$`|$\overline{x}$
`$\hat{x}$`|$\hat{x}$
`$\tilde{x}$`|$\tilde{x}$
`\underset{\gamma}{\operatorname{argmin}}`|$\underset{\gamma}{\operatorname{argmin}}$
`$\frac{a}{b}$`|$\frac{a}{b}$ 
`$\frac{a}{b}$`|$\frac{a}{b}$
`$\displaystyle \frac{a}{b}$`|$\displaystyle \frac{a}{b}$
`$\binom{n}{k}$`|$\binom{n}{k}$
`$x_{1} + x_{2} + \cdots + x_{n}$`|$x_{1} + x_{2} + \cdots + x_{n}$
`$x_{1}, x_{2}, \dots, x_{n}$`|$x_{1}, x_{2}, \dots, x_{n}$
`\mathbf{x} = \langle x_{1}, x_{2}, \dots, x_{n}\rangle$`|$\mathbf{x} = \langle x_{1}, x_{2}, \dots, x_{n}\rangle$
`$x \in A$`|$x \in A$
`$|A|$`|$|A|$
`$x \in A$`|$x \in A$
`$x \subset B$`|$x \subset B$
`$x \subseteq B$`|$x \subseteq B$
`$A \cup B$`|$A \cup B$
`$A \cap B$`|$A \cap B$
`$X \sim {\sf Binom}(n, \pi)$`|$X \sim {\sf Binom}(n, \pi)$
`$\mathrm{P}(X \le x) = {\tt pbinom}(x, n, \pi)$`|$\mathrm{P}(X \le x) = {\tt pbinom}(x, n, \pi)$
`$P(A \mid B)$`|$P(A \mid B)$
`$\mathrm{P}(A \mid B)$`|$\mathrm{P}(A \mid B)$
`$\{1, 2, 3\}$`|$\{1, 2, 3\}$
`$\sin(x)$`|$\sin(x)$
`$\log(x)$`|$\log(x)$
`$\int_{a}^{b}$`|$\int_{a}^{b}$
`$\left(\int_{a}^{b} f(x) \; dx\right)$`|$\left(\int_{a}^{b} f(x) \; dx\right)$
`$\left[\int_{\-infty}^{\infty} f(x) \; dx\right]$`|$\left[\int_{-\infty}^{\infty} f(x) \; dx\right]$
`$\left. F(x) \right|_{a}^{b}$`|$\left. F(x) \right|_{a}^{b}$
`$\sum_{x = a}^{b} f(x)$`|$\sum_{x = a}^{b} f(x)$
`$\prod_{x = a}^{b} f(x)$`|$\prod_{x = a}^{b} f(x)$
`$\lim_{x \to \infty} f(x)$`|$\lim_{x \to \infty} f(x)$
`$\displaystyle \lim_{x \to \infty} f(x)$`|$\displaystyle \lim_{x \to \infty} f(x)$
**Greek Letters** | 
`$\alpha A$`|$\alpha A$
`$\beta B$`|$\beta B$
`$\gamma \Gamma$`|$\gamma \Gamma$ 
`$\delta \Delta$`|$\delta \Delta$ 
`$\epsilon \varepsilon E$`|$\epsilon \varepsilon E$
`$\zeta Z \sigma \,\!$`|$\zeta Z \sigma \,\!$
`$\eta H$`|$\eta H$
`$\theta \vartheta \Theta$`|$\theta \vartheta \Theta$
`$\iota I$`|$\iota I$
`$\kappa K$`|$\kappa K$
`$\lambda \Lambda$`|$\lambda \Lambda$
`$\mu M$`|$\mu M$
`$\nu N$`|$\nu N$
`$\xi\Xi$`|$\xi\Xi$
`$o O$`|$o O$
`$\pi \Pi$`|$\pi \Pi$
`$\rho\varrho P$`|$\rho\varrho P$
`$\sigma \Sigma$`|$\sigma \Sigma$
`$\tau T$`|$\tau T$
`$\upsilon \Upsilon$`|$\upsilon \Upsilon$
`$\phi \varphi \Phi$`|$\phi \varphi \Phi$
`$\chi X$`|$\chi X$
`$\psi \Psi$`|$\psi \Psi$
`$\omega \Omega$`|$\omega \Omega$


Matrices
```{r eval = FALSE}
$$\begin{array}
{rrr}
1 & 2 & 3 \\
4 & 5 & 6 \\
7 & 8 & 9
\end{array}
$$
```

$$\begin{array}
{rrr}
1 & 2 & 3 \\
4 & 5 & 6 \\
7 & 8 & 9
\end{array}
$$


```{r eval=FALSE}
$$\mathbf{X} = \left[\begin{array}
{rrr}
1 & 2 & 3 \\
4 & 5 & 6 \\
7 & 8 & 9
\end{array}\right]
$$
```
$$\mathbf{X} = \left[\begin{array}
{rrr}
1 & 2 & 3 \\
4 & 5 & 6 \\
7 & 8 & 9
\end{array}\right]
$$


Aligning Equations


Aligning Equations with Comments

```{r eval=FALSE}
\begin{align}
    3+x &=4 && \text{(Solve for} x \text{.)}\\
    x &=4-3 && \text{(Subtract 3 from both sides.)}\\
    x &=1   && \text{(Yielding the solution.)}
\end{align}

```


\begin{align}
    3+x &=4 && \text{(Solve for} x \text{.)}\\
    x &=4-3 && \text{(Subtract 3 from both sides.)}\\
    x &=1   && \text{(Yielding the solution.)}
\end{align}



### Statistics Notation
```{r eval = FALSE}
$$f(y|N,p) = \frac{N!}{y!(N-y)!}\cdot p^y \cdot (1-p)^{N-y} = {{N}\choose{y}} \cdot p^y \cdot (1-p)^{N-y}$$
```
$$f(y|N,p) = \frac{N!}{y!(N-y)!}\cdot p^y \cdot (1-p)^{N-y} = {{N}\choose{y}} \cdot p^y \cdot (1-p)^{N-y}$$
```{r eval=FALSE}
\begin{cases}
\frac{1}{b-a}&\text{for $x\in[a,b]$}\\
0&\text{otherwise}\\
\end{cases}
```



\begin{cases}
\frac{1}{b-a}&\text{for $x\in[a,b]$}\\
0&\text{otherwise}\\
\end{cases}

## Table
```{r eval = FALSE}
+---------------+---------------+--------------------+
| Fruit         | Price         | Advantages         |
+===============+===============+====================+
| *Bananas*     | $1.34         | - built-in wrapper |
|               |               | - bright color     |
+---------------+---------------+--------------------+
| Oranges       | $2.10         | - cures scurvy     |
|               |               | - **tasty**        |
+---------------+---------------+--------------------+
```

+---------------+---------------+--------------------+
| Fruit         | Price         | Advantages         |
+===============+===============+====================+
| *Bananas*     | $1.34         | - built-in wrapper |
|               |               | - bright color     |
+---------------+---------------+--------------------+
| Oranges       | $2.10         | - cures scurvy     |
|               |               | - **tasty**        |
+---------------+---------------+--------------------+


```{r eval=FALSE}
(\mathbf{x}^T\mathbf{x})^{-1}\mathbf{x}^T\mathbf{y}

```
$(\mathbf{x}^T\mathbf{x})^{-1}\mathbf{x}^T\mathbf{y}$

## heading

## About labelling things {#id-example}

You can label chapter and section titles using `{#label}` after them, e.g., we can reference Section \@ref(id-example). If you do not manually label them, there will be automatic labels anyway, e.g., this reference to the unlabelled heading \@ref(heading) uses the automatically generated label `\@ref(heading)`.

## Cross-references

Add an explicit label by adding `{#label}` to the end of the section header. If you know you're going to refer to something, this is probably a good idea.

To refer to in a chapter- or section-number-y way, use `\@ref(label)`.

  * `\@ref(install-git)` example: In chapter \@ref(intro)

If you are happy with the section header as the link text, use it inside a single set of square brackets:

  * `[A picture is worth a thousand words]`: example "A picture is worth a thousand words" via [A picture is worth a thousand words]

There are two ways to specify custom link text:

  * `[link text][Section header text]`, e.g., "pic = 1000 words" via [pic = 1000 words][A picture is worth a thousand words]
  * `[link text](#label)`, e.g., "RStudio, meet Git" via [RStudio, meet Git](#rstudio-see-git)
  
The Pandoc documentation provides more details on automatic section IDs and implicit header references.

## Figures, tables, citations

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@knitr-book].

## How the square bracket links work

  * There are three ways to address a section when creating links within your book:
    - **Explicit identifier**: In `# My header {#foo}` the explicit identifier is `foo`.
    - **Automatically generated identifier**: `my-header` is the auto-identifier for `# My header`. Pandoc creates auto-identifiers according to rules laid out in [Extension: auto_identifiers](http://pandoc.org/README.html#extension-auto_identifiers).
    - The header text, e.g., `My header` be used verbatim as an **implicit header reference**. See [Extension: implicit_header_references](http://pandoc.org/README.html#extension-implicit_header_references) for more.
  * All 3 forms can be used to create cross-references but you build the links differently.
  * Advantage of explicit identification: You are less likely to update the section header and then forget to make matching edits to references elsewhere in the book.

How to make text-based links using explicit identifiers, automatic identifiers, and implicit references:

  * Use implicit reference alone to get a link where the text is exactly the section header:
    - `[Introduce yourself to Git]` [Introduce yourself to Git]
    - `[Success and operating systems]` [Success and operating systems]
  * You can provide custom text for the link with all 3 methods of addressing a section:
    - Implicit header reference: `[link text][Recommended Git clients]` [link text][Recommended Git clients]    
    - Explicit identifier: `[hello git! I'm Jenny](#hello-git)` [hello git! I'm Jenny](#hello-git)
    - Automatic identifier: `[Any text you want](#recommended-git-clients)` [Any text you want](#recommended-git-clients)