# Report

Structure

-   Exploratory analysis

    -   plots
    -   preliminary results
    -   interesting structure/features in the data
    -   outliers

-   Model

    -   Assumptions
    -   Why this model/ How is this model the best one?
    -   Consideration: interactions, collinearity, dependence

-   Model Fit

    -   How well does it fit?

    -   Are the model assumptions met?

        -   Residual analysis

-   Inference/ Prediction

    -   Are there different way to support your inference?

-   Conclusion

    -   Recommendation

    -   Limitation of the analysis

    -   How to correct those in the future

<br>

This chapter is based on the `jtools` package. More information can be found [here.](https://www.rdocumentation.org/packages/jtools/versions/2.1.0)

## One summary table

Packages for reporting:

Summary Statistics Table:

 * [qwraps2](https://cran.r-project.org/web/packages/qwraps2/vignettes/summary-statistics.html)
 * [vtable](https://cran.r-project.org/web/packages/vtable/vignettes/sumtable.html)
 * [gtsummary](http://www.danieldsjoberg.com/gtsummary/)
 * [apaTables](https://cran.r-project.org/web/packages/apaTables/apaTables.pdf)
 * [stargazer](https://cran.r-project.org/web/packages/stargazer/vignettes/stargazer.pdf)
 
Regression Table

 * [gtsummary](http://www.danieldsjoberg.com/gtsummary/) 
 * [sjPlot,sjmisc, sjlabelled](https://cran.r-project.org/web/packages/sjPlot/vignettes/tab_model_estimates.html)
 * [stargazer](https://cran.r-project.org/web/packages/stargazer/vignettes/stargazer.pdf): recommended ([Example](https://www.jakeruss.com/cheatsheets/stargazer/))
 * [modelsummary](https://github.com/vincentarelbundock/modelsummary#a-simple-example)
 

```{r}
library(jtools)
data(movies)
fit <- lm(metascore ~ budget + us_gross + year, data = movies)
summ(fit)
summ(fit, scale = TRUE, vifs = TRUE, part.corr = TRUE, confint = TRUE, pvals = FALSE) #notice that scale here is TRUE

#obtain clsuter-robust SE
data("PetersenCL", package = "sandwich")
fit2 <- lm(y ~ x, data = PetersenCL)
summ(fit2, robust = "HC3", cluster = "firm") 
```

Model to Equation

```{r}
# install.packages("equatiomatic")
fit <- lm(metascore ~ budget + us_gross + year, data = movies)
# show the theoretical model
equatiomatic::extract_eq(fit)
# display the actual coefficients
equatiomatic::extract_eq(fit, use_coefs = TRUE)
```



## Model Comparison

```{r}
fit <- lm(metascore ~ log(budget), data = movies)
fit_b <- lm(metascore ~ log(budget) + log(us_gross), data = movies)
fit_c <- lm(metascore ~ log(budget) + log(us_gross) + runtime, data = movies)
coef_names <- c("Budget" = "log(budget)", "US Gross" = "log(us_gross)",
                "Runtime (Hours)" = "runtime", "Constant" = "(Intercept)")
export_summs(fit, fit_b, fit_c, robust = "HC3", coefs = coef_names)
```

Another package is `stargazer`

```{r}
library("stargazer")
stargazer(attitude)
## 2 OLS models
linear.1 <- lm(rating ~ complaints + privileges + learning + raises + critical,data = attitude)
linear.2 <- lm(rating ~ complaints + privileges + learning, data = attitude)
## create an indicator dependent variable, and run a probit model
attitude$high.rating <- (attitude$rating > 70)
probit.model <-
    glm(
        high.rating ~ learning + critical + advance,
        data = attitude,
        family = binomial(link = "probit")
    )
stargazer(linear.1,
          linear.2,
          probit.model,
          title = "Results",
          align = TRUE)
```

```{r eval = FALSE}
# Latex
stargazer(
    linear.1,
    linear.2,
    probit.model,
    title = "Regression Results",
    align = TRUE,
    dep.var.labels = c("Overall Rating", "High Rating"),
    covariate.labels = c(
        "Handling of Complaints",
        "No Special Privileges",
        "Opportunity to Learn",
        "Performance-Based Raises",
        "Too Critical",
        "Advancement"
    ),
    omit.stat = c("LL", "ser", "f"),
    no.space = TRUE
)
```

```{r}
# ASCII text output
stargazer(
    linear.1,
    linear.2,
    type = "text",
    title = "Regression Results",
    dep.var.labels = c("Overall Rating", "High Rating"),
    covariate.labels = c(
        "Handling of Complaints",
        "No Special Privileges",
        "Opportunity to Learn",
        "Performance-Based Raises",
        "Too Critical",
        "Advancement"
    ),
    omit.stat = c("LL", "ser", "f"),
    ci = TRUE,
    ci.level = 0.90,
    single.row = TRUE
)
```

```{r eval = FALSE}
stargazer(
    linear.1,
    linear.2,
    probit.model,
    title = "Regression Results",
    align = TRUE,
    dep.var.labels = c("Overall Rating", "High Rating"),
    covariate.labels = c(
        "Handling of Complaints",
        "No Special Privileges",
        "Opportunity to Learn",
        "Performance-Based Raises",
        "Too Critical",
        "Advancement"
    ),
    omit.stat = c("LL", "ser", "f"),
    no.space = TRUE
)
```

Correlation Table

```{r eval = FALSE}
correlation.matrix <- cor(attitude[,c("rating","complaints","privileges")])
stargazer(correlation.matrix, title="Correlation Matrix")
```

## Changes in an estimate

```{r}
coef_names <- coef_names[1:3] # Dropping intercept for plots
plot_summs(fit, fit_b, fit_c, robust = "HC3", coefs = coef_names)
plot_summs(fit_c, robust = "HC3", coefs = coef_names, plot.distributions = TRUE)
```
