## Modern Estimators for Staggered Adoption {#sec-modern-estimators-for-staggered-adoption}

Notation in this section follows the conventions of Table \@ref(tab:quasi-exp-notation) in the foundations chapter.

::: {.rmdcaution}
Why modern estimators are needed. When treatment is adopted at different times by different units (staggered adoption), the conventional two-way fixed-effects (TWFE) specification

$$
Y_{it} = \alpha_i + \lambda_t + \beta D_{it} + \varepsilon_{it}
$$

does not in general recover a simple average of treatment effects. @goodman2021difference shows that the TWFE coefficient $\hat\beta$ is a weighted average of all possible 2×2 DiD comparisons across treatment-timing cohorts, and that *some of those weights can be negative* whenever treatment effects are heterogeneous across cohorts or across time within a cohort. Early-treated units end up serving as "controls" for later-treated units even though they are themselves treated (the so-called *forbidden comparison*). The consequence is that $\hat\beta$ can be smaller than every underlying ATT, of the wrong sign, or otherwise uninterpretable, even when parallel trends holds.

The estimators introduced below address this problem directly: Callaway and Sant'Anna [@callaway2021difference], Sun and Abraham [@sun2021estimating], de Chaisemartin and d'Haultfœuille [@de2020two], and Borusyak, Jaravel and Spiess [@borusyak2024revisiting]. Each explicitly estimates group-time ATTs with non-negative, interpretable weights. Prefer them over TWFE whenever treatment timing is staggered and effect heterogeneity is plausible, which in practice is almost always.
:::

### Group-Time Average Treatment Effects [@callaway2021difference] {#sec-group-time-average-treatment-effects-callaway2021difference}

**Notation Recap**

-   $Y_{it}(0)$: Potential outcome for unit $i$ at time $t$ in the absence of treatment.

-   $Y_{it}(g)$: Potential outcome for unit $i$ at time $t$ if first treated in period $g$.

-   $Y_{it}$: Observed outcome for unit $i$ at time $t$.

    $$
    Y_{it} =
    \begin{cases}
    Y_{it}(0), & \text{if unit } i \text{ never treated ( } C_i = 1 \text{)} \\
    1\{G_i > t\} Y_{it}(0) + 1\{G_i \le t\} Y_{it}(G_i), & \text{otherwise}
    \end{cases}
    $$

-   $G_i$: Group assignment, i.e., the time period when unit $i$ first receives treatment.

-   $C_i = 1$: Indicator that unit $i$ never receives treatment (the never-treated group).

-   $D_{it} = 1\{G_i \le t\}$: Indicator that unit $i$ has been treated by time $t$.

------------------------------------------------------------------------

**Assumptions**

Identifying $ATT(g, t)$ in a staggered design requires several restrictions on the data-generating process. Each plays a distinct role: some pin down the counterfactual, some restrict the assignment mechanism, and some ensure that the relevant comparison groups exist in finite samples. Reviewing them as a package, rather than as a checklist, clarifies which empirical settings are most likely to satisfy them and which diagnostics matter most.

1.  **Staggered Treatment Adoption**\
    Once treated, a unit remains treated in all subsequent periods, so $D_{it}$ is non-decreasing in $t$. This is what makes the design "staggered" rather than "switching": absorbing treatment lets each cohort be summarized by a single adoption date $G_i$. Empirically the assumption fails whenever units can opt out, lose eligibility, or cycle in and out of policy regimes (think laws that sunset, subsidies that expire, or firms that adopt and later abandon a practice). When that happens, the cohort indexing breaks down and one must turn to estimators that explicitly accommodate transitions in both directions, such as the switching DiD framework discussed in Section \@ref(sec-switching-difference-in-differences-estimator-de2020two) and the panel-match approach in Section \@ref(sec-panel-match-did-estimator-with-in-and-out-treatment-conditions).

2.  **Parallel Trends Assumptions** (Conditional or Unconditional on Covariates)

    Parallel trends is the central identifying restriction across all DiD designs (Section \@ref(sec-difference-in-differences)). In the staggered setting it must hold for *each* cohort $g$ relative to whichever comparison group is used. Two variants are standard, and they differ in which units serve as controls:

    -   **Parallel trends based on never-treated units**: $$
        \mathbb{E}[Y_t(0) - Y_{t-1}(0) | G_i = g] = \mathbb{E}[Y_t(0) - Y_{t-1}(0) | C_i = 1]
        $$ The cohort first treated at $g$ would, absent treatment, have evolved on the same path as units never exposed. This is the cleanest comparison because controls are by construction unaffected by treatment, but it presumes that a meaningful never-treated population exists and is comparable to eventual adopters. In policy settings with eventual universal rollout, the never-treated set is empty or unrepresentative.

    -   **Parallel trends based on not-yet-treated units**: $$
        \mathbb{E}[Y_t(0) - Y_{t-1}(0) | G_i = g] = \mathbb{E}[Y_t(0) - Y_{t-1}(0) | D_{is} = 0, G_i \ne g]
        $$ Units that have not yet been treated by time $s$ serve as controls for cohort $g$. This is the workhorse variant when never-treated units are scarce or suspected to differ systematically. The cost is that controls eventually become treated, so the effective comparison sample shrinks at later horizons, and unmodeled anticipation by future cohorts can contaminate the trend.

    Either variant can be imposed **conditional on covariates** $X$, leveraging the conditional ignorability logic discussed in Section \@ref(sec-conditional-ignorability-assumption):

    $$
    \mathbb{E}[Y_t(0) - Y_{t-1}(0) | X_i, G_i = g] = \mathbb{E}[Y_t(0) - Y_{t-1}(0) | X_i, C_i = 1]
    $$

    The conditional version is appropriate when adoption timing correlates with observables that themselves move outcomes (firm size, baseline income, sector). Empirically, parallel trends fails when cohorts are differentially exposed to shocks (selection on shocks), when adoption is driven by anticipated outcome changes (Ashenfelter dips), or when underlying trends genuinely differ across cohorts. Pre-trend tests, the placebo strategies in Section \@ref(sec-placebo-test-did), and the partial-identification tools of Sections \@ref(sec-partial-identification-did), \@ref(sec-relative-magnitude-restriction-honestdid), and \@ref(sec-smoothness-restriction-honestdid) help diagnose and bound the consequences of violations.

3.  **No Anticipatory Effects**\
    Outcomes prior to $g$ should not respond to the impending treatment, so that pre-treatment periods are valid baselines. This is what makes the standard $g - 1$ reference period meaningful. Anticipation tends to appear when adoption is preceded by announcements, application processes, or strategic positioning by treated units; visible pre-trends in event-study plots are the usual telltale. When anticipation is unavoidable, researchers either shift the reference period further back or explicitly model the anticipation window.

4.  **Random Sampling**\
    Units are drawn independently and identically from the target population. This sampling assumption underlies the standard inference machinery, including the multiplier-bootstrap procedures used in the @callaway2021difference framework. Clustered or network-correlated sampling violates it and typically requires cluster-robust inference or design-based alternatives such as the Fisher randomization tests demonstrated later in this section.

5.  **Irreversibility of Treatment**\
    Once treated, units do not revert to untreated status. This is the population-level analogue of the staggered adoption assumption above: it pins down what $Y_{it}(g)$ even means once $t \ge g$. It also rules out the kind of treatment cycling that would otherwise demand a switching estimator (Section \@ref(sec-switching-difference-in-differences-estimator-de2020two)).

6.  **Overlap (Positivity)**\
    For each group $g$ and each covariate cell, the probability of belonging to that cohort lies strictly inside the unit interval: $$
    0 < \mathbb{P}(G_i = g | X_i) < 1
    $$ Overlap (Section \@ref(sec-overlap-positivity-assumption)) ensures that conditional comparisons are not built on extrapolation. In small samples or with rich covariates the assumption is easy to violate by accident: a single covariate cell with no never-treated units forces the estimator to extrapolate, and inverse-probability weights blow up near the boundary. Trimming, coarsening covariates, or switching to outcome-regression-only adjustments are the usual remedies.

------------------------------------------------------------------------

The Group-Time [ATT](#sec-average-treatment-effect-on-the-treated), $ATT(g, t)$, measures the average treatment effect for units first treated in period $g$, evaluated at time $t$.

$$
ATT(g, t) = \mathbb{E}[Y_t(g) - Y_t(0) | G_i = g]
$$

Interpretation:

-   $g$ indexes when the group first receives treatment.

-   $t$ is the time period when the effect is evaluated.

-   $ATT(g, t)$ captures how treatment effects evolve over time, following adoption at time $g$.

------------------------------------------------------------------------

Identification of $ATT(g, t)$

Each parallel trends variant maps directly to a sample analogue. The choice is driven by what counts as a credible control: a never-treated reservoir if one exists, the not-yet-treated otherwise, with covariate adjustment layered on whenever observables predict adoption timing.

1.  **Using Never-Treated Units as Controls**: $$
    ATT(g, t) = \mathbb{E}[Y_t - Y_{g-1} | G_i = g] - \mathbb{E}[Y_t - Y_{g-1} | C_i = 1], \quad \forall t \ge g
    $$ Preferred when the never-treated population is large and plausibly comparable. It avoids any contamination from future treatment effects but is only as good as the comparability of the two groups.

2.  **Using Not-Yet-Treated Units as Controls**: $$
    ATT(g, t) = \mathbb{E}[Y_t - Y_{g-1} | G_i = g] - \mathbb{E}[Y_t - Y_{g-1} | D_{it} = 0, G_i \ne g], \quad \forall t \ge g
    $$ Useful when a never-treated group is small, missing, or visibly different from adopters. It expands the effective control sample but requires that future cohorts not anticipate treatment in ways that distort their pre-period outcomes.

3.  **Conditional Parallel Trends (with Covariates)**:\
    When treatment assignment depends on covariates $X_i$, identification requires the conditional version of parallel trends and a corresponding adjustment, typically through outcome regression, inverse-probability weighting, or a doubly robust combination of the two:

    -   **Never-treated controls**: $$
        ATT(g, t) = \mathbb{E}[Y_t - Y_{g-1} | X_i, G_i = g] - \mathbb{E}[Y_t - Y_{g-1} | X_i, C_i = 1], \quad \forall t \ge g
        $$
    -   **Not-yet-treated controls**: $$
        ATT(g, t) = \mathbb{E}[Y_t - Y_{g-1} | X_i, G_i = g] - \mathbb{E}[Y_t - Y_{g-1} | X_i, D_{it} = 0, G_i \ne g], \quad \forall t \ge g
        $$

    Covariate adjustment matters most when adoption is endogenous to characteristics that themselves drive outcome trends. The reweighting machinery here parallels the inverse-probability ideas in Section \@ref(sec-reshaped-inverse-probability-weighting-twfe-estimator), and overlap (Section \@ref(sec-overlap-positivity-assumption)) is the binding constraint on how aggressively one can condition.

------------------------------------------------------------------------

Aggregating $ATT(g, t)$: Common Parameters of Interest

The full grid $\{ATT(g, t)\}$ is rarely the object an applied researcher reports. There are typically too many cells, each estimated with limited precision, and policy questions usually demand a single summary or a small set of summaries. The aggregation step is therefore a substantive choice: it embeds a particular question about *which* treated units, *which* exposure horizons, and *which* calendar windows the reader should care about. The following parameters are the standard menu, and the right pick depends on whether the goal is a cohort-level effect, an overall effect, an exposure profile, or a time-series narrative.

1.  [Average Treatment Effect] per Group ($\theta_S(g)$):\
    Average effect over all post-treatment periods for cohort $g$: $$
    \theta_S(g) = \frac{1}{\tau - g + 1} \sum_{t = g}^{\tau} ATT(g, t)
    $$

    -   $\tau$: Last time period in the panel.

    Use $\theta_S(g)$ when the substantive interest is in *which cohorts* responded most. It is also the natural diagnostic for heterogeneity across adoption timing, the very source of bias documented in the @goodman2021difference decomposition (Section \@ref(sec-goodman-bacon-decomposition)). The drawback is that early cohorts are averaged over many post-periods while late cohorts are averaged over few, so $\theta_S(g)$ values are not directly comparable across $g$.

2.  Overall [Average Treatment Effect on the Treated] (ATT) ($\theta_S^O$):\
    Weighted average of $\theta_S(g)$ across cohorts, with weights proportional to cohort size: $$
    \theta_S^O = \sum_{g=2}^{\tau} \theta_S(g) \cdot \mathbb{P}(G_i = g)
    $$

    This is the closest analogue to the scalar $\beta$ that TWFE was supposed to deliver, but it is built from non-negative weights and so is interpretable as the average effect on a treated unit drawn at random. Report it when a single headline number is required; pair it with $\theta_S(g)$ or the dynamic profile below to avoid masking heterogeneity.

3.  **Dynamic Treatment Effects** ($\theta_D(e)$):\
    Average effect $e$ periods after adoption, averaged over cohorts that have been observed at least $e$ periods post-treatment: $$
    \theta_D(e) = \sum_{g=2}^{\tau} \mathbb{1}\{g + e \le \tau\} \cdot ATT(g, g + e) \cdot \mathbb{P}(G_i = g | g + e \le \tau)
    $$

    The dynamic profile is the natural target when treatment effects build up, decay, or peak over time. It is the foundation of event studies (Section \@ref(sec-event-studies)) and supports the visualization conventions discussed in Section \@ref(sec-visualization-did). Note that the composition of cohorts entering $\theta_D(e)$ shifts with $e$: at long horizons only early-adopting cohorts remain, so apparent dynamics may reflect cohort composition as much as time-since-treatment.

4.  **Calendar Time Treatment Effects** ($\theta_C(t)$):\
    Average treatment effect at calendar time $t$ across all cohorts already treated by $t$: $$
    \theta_C(t) = \sum_{g=2}^{\tau} \mathbb{1}\{g \le t\} \cdot ATT(g, t) \cdot \mathbb{P}(G_i = g | g \le t)
    $$

    Choose $\theta_C(t)$ when the question is *how the policy is performing as time passes*, for example whether a treatment loses bite during a recession or strengthens after a complementary policy goes into effect. It is also the right object when a macro shock plausibly hits all treated units simultaneously.

5.  **Average Calendar Time Treatment Effect** ($\theta_C$):\
    Average of $\theta_C(t)$ across post-treatment periods: $$
    \theta_C = \frac{1}{\tau - 1} \sum_{t=2}^{\tau} \theta_C(t)
    $$

    A useful counterpart to $\theta_S^O$: where the latter weights treated units, $\theta_C$ weights calendar periods. The two coincide under effect homogeneity and diverge whenever cohort sizes are unbalanced or treatment dynamics differ across calendar windows.

The `staggered()` function exposes the same logic through a small set of estimand names. They are not different identifiers of a single quantity; they are different summaries that answer different questions, and reporting more than one is often informative:

-   **Simple**: Equally weighted across all $(g, t)$ cells in the post-treatment region. A useful baseline because it does not let any one cohort dominate, but it gives the same weight to a cohort with a few units as to one with many.

-   **Cohort**: Weights by treated-cohort size, so $(g, t)$ cells are aggregated by how many treated units they actually represent. This is the natural estimand when the goal is the average effect on a treated unit, and it lines up with $\theta_S^O$ above.

-   **Calendar**: Weights by the number of observations in each calendar period. Appropriate when the policy question is calendar-anchored (a fiscal year's effect, a post-shock window) rather than cohort-anchored.

A practical rule of thumb: lead with the simple or cohort estimand for the headline ATT, use the dynamic event-study profile to communicate timing, and bring in the calendar variant when temporal context (business cycle, complementary policies) matters for interpretation.

```{r}
library(staggered) 
library(fixest)
data("base_stagg")

# Simple weighted average ATT
staggered(
    df = base_stagg,
    i = "id",
    t = "year",
    g = "year_treated",
    y = "y",
    estimand = "simple"
)

# Cohort weighted ATT (i.e., by treatment cohort size)
staggered(
    df = base_stagg,
    i = "id",
    t = "year",
    g = "year_treated",
    y = "y",
    estimand = "cohort"
)

# Calendar weighted ATT (i.e., by year)
staggered(
    df = base_stagg,
    i = "id",
    t = "year",
    g = "year_treated",
    y = "y",
    estimand = "calendar"
)
```

To visualize treatment dynamics around the time of adoption, the event study specification estimates dynamic treatment effects relative to the time of treatment (Figure \@ref(fig:fig-event-study-dynamic-effects)).

```{r}
res <- staggered(
    df = base_stagg,
    i = "id",
    t = "year",
    g = "year_treated",
    y = "y",
    estimand = "eventstudy", 
    eventTime = -9:8
)
```

```{r  fig-event-study-dynamic-effects, fig.cap='Event-study dynamic treatment effects.', fig.alt='A dot-and-whisker plot showing ATT estimates over event time. Points before treatment hover near zero, while post-treatment estimates increase steadily, with confidence intervals widening over time.'}
# Plotting the event study with pointwise confidence intervals
library(ggplot2)
library(dplyr)

ggplot(
    res |> mutate(
        ymin_ptwise = estimate - 1.96 * se,
        ymax_ptwise = estimate + 1.96 * se
    ),
    aes(x = eventTime, y = estimate)
) +
    geom_pointrange(aes(ymin = ymin_ptwise, ymax = ymax_ptwise)) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    xlab("Event Time") +
    ylab("ATT Estimate") +
    ggtitle("Event Study: Dynamic Treatment Effects") +
    causalverse::ama_theme()
```

The `staggered` package also includes direct implementations of alternative estimators:

-   `staggered_cs()` implements the @callaway2021difference estimator.

-   `staggered_sa()` implements the @sun2021estimating estimator, which adjusts for bias from comparisons involving already-treated units.

```{r}
# Callaway and Sant’Anna estimator
staggered_cs(
    df = base_stagg,
    i = "id",
    t = "year",
    g = "year_treated",
    y = "y",
    estimand = "simple"
)

# Sun and Abraham estimator
staggered_sa(
    df = base_stagg,
    i = "id",
    t = "year",
    g = "year_treated",
    y = "y",
    estimand = "simple"
)
```

To assess statistical significance under the sharp null hypothesis $H_0: \text{TE} = 0$, the `staggered` package includes an option for Fisher's randomization (permutation) test. This approach tests whether the observed estimate could plausibly occur under a random reallocation of treatment timings.

```{r}
# Fisher Randomization Test
staggered(
    df = base_stagg,
    i = "id",
    t = "year",
    g = "year_treated",
    y = "y",
    estimand = "simple",
    compute_fisher = TRUE,
    num_fisher_permutations = 100
)
```

This test provides a non-parametric method for inference and is particularly useful when the number of groups is small or standard errors are unreliable due to clustering or heteroskedasticity.

------------------------------------------------------------------------

### Cohort Average Treatment Effects [@sun2021estimating] {#sec-cohort-average-treatment-effects-sun2021estimating}

@sun2021estimating propose a solution to the TWFE problem in staggered adoption settings by introducing an interaction-weighted estimator for dynamic treatment effects. This estimator is based on the concept of **Cohort Average Treatment Effects on the Treated (CATT)**, which accounts for variation in treatment timing and dynamic treatment responses.

Traditional TWFE estimators implicitly assume **homogeneous treatment effects** and often rely on treated units serving as controls for later-treated units. When treatment effects vary over time or across groups, this leads to **contaminated comparisons**, especially in event-study specifications.

@sun2021estimating address this issue by:

-   Estimating cohort-specific treatment effects relative to time since treatment.

-   Using never-treated units as controls, or in their absence, the last-treated cohort.

#### Defining the Parameter of Interest: CATT

Let $G_i = g$ denote the period when unit $i$ first receives treatment. The **cohort-specific average treatment effect on the treated** (CATT) is defined as: $$
CATT_{g, l} = \mathbb{E}[Y_{i, g + l} - Y_{i, g + l}(0) \mid G_i = g]
$$ Where:

-   $l$ is the relative period (e.g., $l = -1$ is one year before treatment, $l = 0$ is the treatment year).

-   $Y_{i, g + l}(0)$ is the potential outcome without treatment.

-   $Y_{i, g + l}$ is the observed outcome.

This formulation allows one to trace out the dynamic effect of treatment for each cohort, relative to their treatment start time.

@sun2021estimating extend the interaction-weighted idea to panel settings, originally introduced by @gibbons2018broken in a cross-sectional context.

They propose regressing the outcome on:

-   Relative time indicators constructed by interacting treatment cohort ($G_i$) with time ($t$).

-   Unit and time fixed effects.

This method explicitly estimates $CATT_{g, l}$ terms, avoiding the contaminating influence of already-treated units that TWFE models often suffer from.

**Relative Period Bin Indicator**

$$ D_{it}^l = \mathbb{1}(t - G_i = l) $$

-   $G_i$: The time period when unit $i$ first receives treatment.
-   $l$: The relative time period: how many periods have passed since treatment began.

1.  **Static Specification**

$$ Y_{it} = \alpha_i + \lambda_t + \mu_g \sum_{l \ge 0} D_{it}^l + \epsilon_{it} $$

-   $\alpha_i$: Unit fixed effects.
-   $\lambda_t$: Time fixed effects.
-   $\mu_g$: Effect for group $g$.
-   Excludes periods prior to treatment.

2.  **Dynamic Specification**

$$ Y_{it} = \alpha_i + \lambda_t + \sum_{\substack{l = -K \\ l \neq -1}}^{L} \mu_l D_{it}^l + \epsilon_{it} $$

-   Includes leads and lags of treatment indicators $D_{it}^l$.
-   Excludes one period (typically $l = -1$) to avoid perfect collinearity.
-   Tests for pre-treatment parallel trends rely on the leads ($l < 0$).

------------------------------------------------------------------------

#### Identifying Assumptions

1.  **Parallel Trends**

For identification, it is assumed that untreated potential outcomes follow parallel trends across cohorts in the absence of treatment: $$
\mathbb{E}[Y_{it}(0) - Y_{i, t-1}(0) \mid G_i = g] = \text{constant across } g
$$ This allows us to use **never-treated** or **not-yet-treated** units as valid counterfactuals.

2.  **No Anticipatory Effects**

Treatment should not influence outcomes before it is implemented. That is: $$
CATT_{g, l} = 0 \quad \text{for all } l < 0
$$ This ensures that any pre-trends are not driven by behavioral changes in anticipation of treatment.

3.  **Treatment Effect Homogeneity (Optional)**

The treatment effect is consistent across cohorts for each relative period. Each adoption cohort should have the same path of treatment effects. In other words, the trajectory of each treatment cohort is similar.

Although @sun2021estimating allow treatment effect heterogeneity, some settings may assume homogeneous effects within cohorts and periods:

-   Each cohort has the same pattern of response over time.

-   This is relaxed in their design but assumed in simpler TWFE settings.

#### Comparison to Other Designs

Different DiD designs make distinct assumptions about how treatment effects vary (Table \@ref(tab:did-comparison-design-assumptions))

+-------------------------+----------------+---------------------+------------------------------------------------------------------------------------------------------+
| Study                   | Vary Over Time | Vary Across Cohorts | Notes                                                                                                |
+=========================+================+=====================+======================================================================================================+
| @sun2021estimating      | ✓              | ✓                   | Allows full heterogeneity                                                                            |
+-------------------------+----------------+---------------------+------------------------------------------------------------------------------------------------------+
| @callaway2021difference | ✓              | ✓                   | Estimates group × time ATTs                                                                          |
+-------------------------+----------------+---------------------+------------------------------------------------------------------------------------------------------+
| @borusyak2024revisiting | ✓              | ✗                   | Homogeneous across cohorts                                                                           |
|                         |                |                     |                                                                                                      |
|                         |                |                     | Heterogeneity over time                                                                              |
+-------------------------+----------------+---------------------+------------------------------------------------------------------------------------------------------+
| @athey2022design        | ✗              | ✓                   | Heterogeneity only across adoption cohorts                                                           |
+-------------------------+----------------+---------------------+------------------------------------------------------------------------------------------------------+
| @de2023two              | ✓              | ✓                   | Complete heterogeneity                                                                               |
+-------------------------+----------------+---------------------+------------------------------------------------------------------------------------------------------+
| @goodman2021difference  | ✓ or ✗         | ✗ or ✓              | Restricts one dimension                                                                              |
|                         |                |                     |                                                                                                      |
|                         |                |                     | Heterogeneity either "vary across units but not over time" or "vary over time but not across units". |
+-------------------------+----------------+---------------------+------------------------------------------------------------------------------------------------------+

: (#tab:did-comparison-design-assumptions) Comparison of designs' assumptions.

#### Sources of Treatment Effect Heterogeneity

Several forces can generate heterogeneous treatment effects:

-   **Calendar Time Effects**: Macro events (e.g., recessions, policy changes) affect cohorts differently.

-   **Selection into Timing**: Units self-select into early/late treatment based on anticipated effects.

-   **Composition Differences**: Adoption cohorts may differ in observed or unobserved ways.

Such heterogeneity can bias TWFE estimates, which often average effects across incomparable groups.

------------------------------------------------------------------------

#### Technical Issues

When using an event-study TWFE regression to estimate dynamic treatment effects in staggered adoption settings, one must exclude certain relative time indicators to avoid perfect multicollinearity. This arises because relative period indicators are linearly dependent due to the presence of unit and time fixed effects.

Specifically, the following two terms must be addressed:

-   **The period immediately before treatment** ($l = -1$): This period is typically omitted and serves as the baseline for comparison. This normalization has been standard practice in event study regressions prior to @sun2021estimating .

-   **A distant post-treatment period** (e.g., $l = +5$ or $l = +10$): @sun2021estimating clarified that in addition to the baseline period, at least one other relative time indicator, typically from the far tail of the post-treatment distribution, must be dropped, binned, or trimmed to avoid multicollinearity among the relative time dummies. This issue emerges because fixed effects absorb much of the within-unit and within-time variation, reducing the effective rank of the design matrix.

Dropping certain relative periods (especially pre-treatment periods) introduces an implicit normalization: the estimates for included periods are now interpreted relative to the omitted periods. If treatment effects are present in these omitted periods (e.g., due to anticipation or early effects), this will contaminate the estimates of included relative periods.

To avoid this contamination, researchers often assume that all pre-treatment periods have zero treatment effect, i.e.,

$$
CATT_{g, l} = 0 \quad \text{for all } l < 0
$$

This assumption ensures that excluded pre-treatment periods form a valid counterfactual, and estimates for $l \geq 0$ are not biased due to normalization.

------------------------------------------------------------------------

@sun2021estimating resolve the issues of weighting and aggregation by using a clean weighting scheme that avoids contamination from excluded periods. Their method produces a weighted average of cohort- and time-specific treatment effects ($CATT_{g, l}$), where the weights are:

-   **Non-negative**
-   **Sum to one**
-   **Interpretable as the fraction of treated units who are observed** $l$ periods after treatment, normalized over the number of available periods $g$

This interaction-weighted estimator ensures that the estimated average treatment effect reflects a convex combination of dynamic treatment effects from different cohorts and times.

In this way, their aggregation logic closely mirrors that of @callaway2021difference, who also construct average treatment effects from group-time ATTs using interpretable weights that align with the sampling structure.

------------------------------------------------------------------------

```{r}
library(fixest)
data("base_stagg")

# Estimate Sun & Abraham interaction-weighted model
res_sa20 <- feols(
  y ~ x1 + sunab(year_treated, year) | id + year,
  data = base_stagg
)

```

Use `iplot()` to visualize the estimated dynamic treatment effects across relative time (Figure \@ref(fig:event-study-effect-on-y))

```{r event-study-effect-on-y, fig.cap='Estimated dynamic treatment effects on $y$ across relative time.', fig.alt='Dot plot of treatment effects over time with 95% confidence intervals. Pre-treatment estimates hover near zero, while post-treatment effects show a rising trend, indicating increasing impact.'}
iplot(res_sa20)
```

You can summarize the results using different aggregation options:

```{r}
# Overall average ATT
summary(res_sa20, agg = "att")

# Aggregation across post-treatment periods (excluding leads)
summary(res_sa20, agg = c("att" = "year::[^-]"))

# Aggregate post-treatment effects from l = 0 to 8
summary(res_sa20, agg = c("att" = "year::[012345678]")) |> 
  etable(digits = 2)

```

The `fwlplot` package provides diagnostics for how much variation is explained by fixed effects or covariates (Figures \@ref(fig:scatter-regression-x1-y), \@ref(fig:residualized-scatter-x1-y) and \@ref(fig:residualized-scatter-by-treatment)).

```{r}
library(fwlplot)
```

```{r scatter-regression-x1-y, fig.cap='Scatter plot of $y$ against $x_1$.', fig.alt='A scatter plot showing a dense cloud of black points with a light blue regression line sloping upward, indicating a positive linear relationship between x1 and y, along with a shaded confidence band.'}
# Simple FWL plot
fwl_plot(y ~ x1, data = base_stagg)
```

```{r residualized-scatter-x1-y, fig.cap='Scatter plot of residualized $y$ against residualized $x_1$.', fig.alt='Scatter plot of residualized y against residualized x1, with a fitted linear regression line and shaded confidence band. The plot reveals a positive partial relationship between the variables.'}
# With fixed effects
fwl_plot(y ~ x1 | id + year,
         data = base_stagg,
         n_sample = 100)
```

```{r residualized-scatter-by-treatment, fig.cap='Scatter plot of residualized $y$ against residualized $x_1$ by treatment status.', fig.alt='Three scatter plots displaying residualized y versus residualized x1. The left panel shows the full sample, the middle panel shows untreated units, and the right panel shows treated units. Each includes a positively sloped regression line with confidence intervals.'}
# Splitting by treatment status
fwl_plot(
    y ~ x1 |
        id + year,
    data = base_stagg,
    n_sample = 100,
    fsplit = ~ treated
)
```

------------------------------------------------------------------------

### Stacked Difference-in-Differences {#sec-stacked-difference-in-differences}

The **Stacked DiD** approach addresses key limitations of standard TWFE models in staggered adoption designs, particularly treatment effect heterogeneity and timing variations. By constructing sub-experiments around each treatment event, researchers can isolate cleaner comparisons and reduce contamination from improperly specified control groups.

Basic TWFE Specification

$$
Y_{it} = \beta_{FE} D_{it} + A_i + B_t + \epsilon_{it}
$$

-   $Y_{it}$: Outcome for unit $i$ at time $t$.
-   $D_{it}$: Treatment indicator (1 if treated, 0 otherwise).
-   $A_i$: Unit (group) fixed effects.
-   $B_t$: Time period fixed effects.
-   $\epsilon_{it}$: Idiosyncratic error term.

------------------------------------------------------------------------

Steps in the Stacked DiD Procedure

#### Choose an Event Window

Define:

-   $\kappa_a$: Number of **pre-treatment** periods to include in the event window (lead periods).
-   $\kappa_b$: Number of **post-treatment** periods to include in the event window (lag periods).

**Implication**:\
Only events where sufficient **pre- and post-treatment periods** exist will be included (i.e., excluding those events that do not meet this criteria).

------------------------------------------------------------------------

#### Enumerate Sub-Experiments

Define:

-   $T_1$: First period in the panel.
-   $T_T$: Last period in the panel.
-   $\Omega_A$: The set of treatment adoption periods that fit within the event window:

$$
\Omega_A = \left\{ A_i \middle| T_1 + \kappa_a \le A_i \le T_T - \kappa_b \right\}
$$

-   Each $A_i$ represents an adoption period for unit $i$ that has enough time on both sides of the event.

Let $d = 1, \dots, D$ index the sub-experiments in $\Omega_A$.

-   $\omega_d$: The event (adoption) date of the $d$-th sub-experiment.

------------------------------------------------------------------------

#### Define Inclusion Criteria

**Valid Treated Units**

-   In sub-experiment $d$, treated units have adoption date exactly equal to $\omega_d$.
-   A unit may only be treated in one sub-experiment to avoid duplication.

**Clean Control Units**

-   Controls are units where $A_i > \omega_d + \kappa_b$, i.e.,
    -   They are **never treated**, or
    -   They are **treated in the far future** (beyond the post-event window).
-   A control unit can appear in multiple sub-experiments, but this requires correcting standard errors (see below).

**Valid Time Periods**

-   Only observations where\
    $$
    \omega_d - \kappa_a \le t \le \omega_d + \kappa_b
    $$\
    are included.
-   This ensures the analysis is centered on the event window.

#### Specify Estimating Equation

Basic DiD Specification in the Stacked Dataset

$$
Y_{itd} = \beta_0 + \beta_1 T_{id}  + \beta_2 P_{td} + \beta_3 (T_{id} \times P_{td}) + \epsilon_{itd}
$$

Where:

-   $i$: Unit index

-   $t$: Time index

-   $d$: Sub-experiment index

-   $T_{id}$: Indicator for treated units in sub-experiment $d$

-   $P_{td}$: Indicator for post-treatment periods in sub-experiment $d$

-   $\beta_3$: Captures the DiD estimate of the treatment effect.

Equivalent Form with Fixed Effects

$$
Y_{itd} = \beta_3 (T_{id} \times P_{td}) + \theta_{id} + \gamma_{td} + \epsilon_{itd}
$$

where

-   $\theta_{id}$: Unit-by-sub-experiment fixed effect.

-   $\gamma_{td}$: Time-by-sub-experiment fixed effect.

Note:

-   $\beta_3$ summarizes the average treatment effect across all sub-experiments but does not allow for dynamic effects by time since treatment.

#### Stacked Event Study Specification

Define Time Since Event ($YSE_{td}$):

$$
YSE_{td} = t- \omega_d
$$

where

-   Measures time since the event (relative time) in sub-experiment $d$.

-   $YSE_{td} \in [-\kappa_a, \dots, 0, \dots, \kappa_b]$ in every sub-experiment.

**Event-Study Regression (Sub-Experiment Level)**

$$
Y_{it}^d = \sum_{j = -\kappa_a}^{\kappa_b} \beta_j^d . 1 (YSE_{td} = j) + \sum_{j = -\kappa_a}^{\kappa_b} \delta_j^d (T_{id} . 1 (YSE_{td} = j)) + \theta_i^d + \epsilon_{it}^d
$$

where

-   Separate coefficients for each sub-experiment $d$.

-   $\delta_j^d$: Captures treatment effects at relative time $j$ within sub-experiment $d$.

**Pooled Stacked Event-Study Regression**

$$
Y_{itd} = \sum_{j = -\kappa_a}^{\kappa_b} \beta_j \cdot \mathbb{1}(YSE_{td} = j) + \sum_{j = -\kappa_a}^{\kappa_b} \delta_j \left( T_{id} \cdot \mathbb{1}(YSE_{td} = j) \right) + \theta_{id} + \epsilon_{itd}
$$

-   Pooled coefficients $\delta_j$ reflect average treatment effects by event time $j$ across sub-experiments.

#### Clustering in Stacked DID

-   **Cluster at Unit × Sub-Experiment Level** [@cengiz2019effect]: Accounts for units appearing multiple times across sub-experiments.

-   **Cluster at Unit Level** [@deshpande2019screened]: Appropriate when units are uniquely identified and do not appear in multiple sub-experiments.

------------------------------------------------------------------------

```{r}
library(did)
library(tidyverse)
library(fixest)

# Load example data
data(base_stagg)

# Get treated cohorts (exclude never-treated units coded as 10000)
cohorts <- base_stagg %>%
    filter(year_treated != 10000) %>%
    distinct(year_treated) %>%
    pull()

# Function to generate data for each sub-experiment
getdata <- function(j, window) {
    base_stagg %>%
        filter(
            year_treated == j |               # treated units in cohort j
            year_treated > j + window         # controls not treated soon after
        ) %>%
        filter(
            year >= j - window &
            year <= j + window                # event window bounds
        ) %>%
        mutate(df = j)                        # sub-experiment indicator
}

# Generate the stacked dataset
stacked_data <- map_df(cohorts, ~ getdata(., window = 5)) %>%
    mutate(
        rel_year = if_else(df == year_treated, time_to_treatment, NA_real_)
    ) %>%
    fastDummies::dummy_cols("rel_year", ignore_na = TRUE) %>%
    mutate(across(starts_with("rel_year_"), ~ replace_na(., 0)))

# Estimate fixed effects regression on the stacked data
stacked_result <- feols(
    y ~ `rel_year_-5` + `rel_year_-4` + `rel_year_-3` + `rel_year_-2` +
        rel_year_0 + rel_year_1 + rel_year_2 + rel_year_3 +
        rel_year_4 + rel_year_5 |
        id ^ df + year ^ df,
    data = stacked_data
)

# Extract coefficients and standard errors
stacked_coeffs <- stacked_result$coefficients
stacked_se <- stacked_result$se

# Insert zero for the omitted period (usually -1)
stacked_coeffs <- c(stacked_coeffs[1:4], 0, stacked_coeffs[5:10])
stacked_se <- c(stacked_se[1:4], 0, stacked_se[5:10])
```

```{r}
# Plotting estimates from three methods: Callaway & Sant'Anna, Sun & Abraham, and Stacked DiD

cs_out <- att_gt(
    yname = "y",
    data = base_stagg,
    gname = "year_treated",
    idname = "id",
    # xformla = "~x1",
    tname = "year"
)
cs <-
    aggte(
        cs_out,
        type = "dynamic",
        min_e = -5,
        max_e = 5,
        bstrap = FALSE,
        cband = FALSE
    )



res_sa20 = feols(y ~ sunab(year_treated, year) |
                     id + year, base_stagg)
sa = tidy(res_sa20)[5:14, ] %>% pull(estimate)
sa = c(sa[1:4], 0, sa[5:10])

sa_se = tidy(res_sa20)[6:15, ] %>% pull(std.error)
sa_se = c(sa_se[1:4], 0, sa_se[5:10])

compare_df_est = data.frame(
    period = -5:5,
    cs = cs$att.egt,
    sa = sa,
    stacked = stacked_coeffs
)

compare_df_se = data.frame(
    period = -5:5,
    cs = cs$se.egt,
    sa = sa_se,
    stacked = stacked_se
)

compare_df_longer <- compare_df_est %>%
    pivot_longer(!period, names_to = "estimator", values_to = "est") %>%
    full_join(compare_df_se %>%
                  pivot_longer(!period, names_to = "estimator", values_to = "se")) %>%
    mutate(upper = est +  1.96 * se,
           lower = est - 1.96 * se)
```

Figure \@ref(fig:dynamic-treatment-comparison-att) shows a comparison between three estimators (i.e., [CS](#sec-group-time-average-treatment-effects-callaway2021difference), [SA](#sec-cohort-average-treatment-effects-sun2021estimating), and [stacked](#sec-stacked-difference-in-differences))

```{r dynamic-treatment-comparison-att, fig.cap='Comparison of dynamic average treatment effects across CS, SA, and stacked estimators.', fig.alt='Line plot showing estimated ATT over event time for three estimators: cs (red), sa (green), and stacked (blue). All estimators show a dip around event time zero, followed by increasing effects. Confidence bands are shown as shaded regions.'}
ggplot(compare_df_longer) +
    geom_ribbon(aes(
        x = period,
        ymin = lower,
        ymax = upper,
        group = estimator
    ), alpha = 0.2) +
    geom_line(aes(
        x = period,
        y = est,
        group = estimator,
        color = estimator
    ),
    linewidth = 1.2) +
    
    labs(
        title = "Comparison of Dynamic Treatment Effects",
        x = "Event Time (Periods since Treatment)",
        y = "Estimated ATT",
        color = "Estimator"
    ) + 
    causalverse::ama_theme()
```

------------------------------------------------------------------------
