### Nonlinear Difference-in-Differences {#sec-nonlinear-did}

Notation in this section follows the conventions of Table \@ref(tab:quasi-exp-notation) in the foundations chapter.

The standard [Difference-in-Differences](#sec-difference-in-differences) estimator assumes that the untreated potential outcome trends linearly across groups: in the absence of treatment, the treated and control groups would have moved on parallel lines. This is a defensible assumption for continuous outcomes that range freely over the real line (log earnings, log expenditure, voter share), but it breaks down when the outcome is bounded, discrete, or otherwise constrained. A linear trend that produces an "untreated employment rate" of $-0.05$ or a "predicted churn proportion" of $1.13$ has not actually predicted anything, and an estimator built on top of such a prediction is mis-specified in a way the linear specification will not flag.

The cases where this matters are common in applied work:

-   **Binary outcomes**, such as employment status, default indicators, or whether a customer renewed a contract. The outcome is bounded in $\{0, 1\}$ and the conditional mean lies in $[0, 1]$.
-   **Fractional outcomes**, such as the proportion of a firm's customers who churned, the share of a portfolio invested in a particular asset, or the fraction of households in poverty. The outcome lies in $[0, 1]$.
-   **Count outcomes**, such as the number of crimes in a neighborhood, the number of patents filed by a firm, or the number of insurance claims per policyholder. The outcome is a non-negative integer with a discrete distribution.

For each of these, the natural model of the conditional mean is *nonlinear*: a logistic or probit link for binary and fractional outcomes, an exponential link for counts. Linear parallel trends imposed on such outcomes can produce qualitatively wrong inferences, including effects with the wrong sign in regions of the covariate space where the linear approximation is at its worst. The framework developed below, drawing on @wooldridge2023simple, replaces linear parallel trends with a *parallel trends in a transformed index*, which respects the constrained nature of the outcome while preserving the DiD logic of canceling unit and time fixed effects through differencing.

The remainder of this section develops the framework in three parts: an overview of the nonlinear conditional-mean structure, the identifying assumptions (a conditional no-anticipation condition and a conditional index-parallel-trends condition), and two equivalent estimators (an imputation estimator and a pooled QMLE).

------------------------------------------------------------------------

#### Overview of Framework

We consider a panel dataset where units are observed over $T$ time periods. Units become treated at various times (staggered rollout), and the goal is to estimate [Average Treatment Effect on the Treated] (ATT) at different times.

Let $Y_{it}(g)$ denote the potential outcome at time $t$ if unit $i$ were first treated in period $g$, where $g = q, \ldots, T$ for treated cohorts. Let $Y_{it}(0)$ denote the never-treated potential outcome. Define the ATT for cohort $g$ at time $r \geq g$ as:

$$
\tau_{gr} = \mathbb{E}\left[Y_{ir}(g) - Y_{ir}(0) \mid D_g = 1\right]
$$

Here, $D_g = 1$ indicates that unit $i$ was first treated in period $g$.

Rather than assuming linear conditional expectations of untreated outcomes, we posit a **nonlinear conditional mean** using a known, strictly increasing function $G(\cdot)$:

$$
\mathbb{E}[Y_{it}(0) \mid D, X] = G\left( \alpha + \sum_{g=q}^{T} \beta_g D_g + X \kappa + \sum_{g=q}^{T} (D_g \cdot X)\eta_g + \gamma_t + X \pi_t \right)
$$

This formulation nests logistic and Poisson mean structures, and allows us to handle various limited dependent variables.

------------------------------------------------------------------------

#### Assumptions

Identification of the nonlinear DiD ATT requires two assumptions, each a natural extension of the standard linear DiD assumptions to the nonlinear setting.

The first is **Conditional No Anticipation**: prior to treatment, the conditional mean of the eventually-treated cohort does not differ systematically from the conditional mean it would have had under permanent non-treatment. Formally,

$$
\mathbb{E}[Y_{it}(g) \mid D_g = 1, X] = \mathbb{E}[Y_{it}(0) \mid D_g = 1, X], \quad \forall t < g.
$$

In words, the cohort treated at $g$ does not adjust its behavior in anticipation of treatment in the periods before $g$. This is the same assumption as in linear DiD; the only difference is that it is now phrased in terms of the conditional mean of the *transformed* outcome rather than the level. Anticipation, where present, manifests as a divergence of the cohort's pre-treatment conditional means from the untreated counterfactual, and it can be diagnosed in the same way as in standard event-study plots.

The second is **Conditional Index Parallel Trends**: the untreated mean trends are parallel after passing through a known monotone transformation $G(\cdot)$. Formally,

$$
\mathbb{E}[Y_{it}(0) \mid D, X] = G(\text{linear index in } D, X, t),
$$

where $G$ is strictly increasing and known up to its functional form (e.g., $G(\cdot) = \exp(\cdot)$ for Poisson outcomes, or the logistic CDF $G(\cdot) = \Lambda(\cdot)$ for binary outcomes). The choice of $G$ is dictated by the outcome: count outcomes call for the exponential, binary outcomes for the logistic or probit, and so on. The substantive content of the assumption is that on the *index* scale (the argument of $G$), trends are parallel across groups; on the outcome scale they are parallel only after the transformation, which is what allows the model to respect the bounds on the outcome.

These assumptions are weaker than linear parallel trends in two ways. First, they accommodate bounded outcomes without imposing implausible counterfactuals. Second, they allow for cross-group differences in the transformation that the linear specification would force into the residual. The trade-off is that the analyst must commit to a specific link function $G$, and a mis-specified link can re-introduce some of the bias that the framework was meant to remove.

------------------------------------------------------------------------

#### Estimation

**Step 1: Imputation Estimator**

1.  Estimate Parameters Using Untreated Observations Only:

Use all $(i,t)$ such that unit $i$ is untreated at $t$ (i.e., $W_{it} = 0$). Fit the nonlinear regression model: $$ Y_{it} = G\left(\alpha + \sum_g \beta_g D_g + X_i \kappa + D_g X_i \eta_g + \gamma_t + X_i \pi_t\right) + \varepsilon_{it} $$

2.  Impute Counterfactual Outcomes for Treated Observations:

For treated observations $(i,t)$ with $W_{it}=1$, predict $\widehat{Y}_{it}(0)$ using the model from Step 1.

3.  Compute ATT for Each Cohort $g$ and Time $r$:

$$ \hat{\tau}_{gr} = \frac{1}{N_{gr}} \sum_{i: D_g=1} \left( Y_{ir} - \widehat{Y}_{ir}(0) \right) $$

------------------------------------------------------------------------

**Step 2: Pooled QMLE Estimator (Equivalent When Using Canonical Link)**

1.  Fit Model Using All Observations:

Fit the pooled nonlinear model across all units and time: $$ Y_{it} = G\left(\alpha + \sum_g \beta_g D_g + X_i \kappa + D_g X_i \eta_g + \gamma_t + X_i \pi_t + \delta_r \cdot W_{it} + W_{it} X_i \xi \right) + \varepsilon_{it} $$

Where:

-   $W_{it} = 1$ if unit $i$ is treated at time $t$

-   $W_{it} = 0$ otherwise

2.  Estimate $\delta_r$ as the ATT for cohort $g$ in period $r$:

-   $\delta_r$ is interpreted as an event-time ATT
-   This estimator is consistent when $G^{-1}(\cdot)$ is the canonical link (e.g., log link for Poisson)

3.  Average Partial Effect (APE) for ATT:

$$ \hat{\tau}_{gr} = \frac{1}{N_g} \sum_{i: D_g=1} \left[ G\left( X_i\beta + \delta_r + \ldots \right) - G\left( X_i\beta + \ldots \right) \right] $$

**Canonical Links in Practice**

| Conditional Mean              | LEF Density | Suitable For                               |
|-------------------------------|-------------|--------------------------------------------|
| $G(z) = z$                    | Normal      | Any response                               |
| $G(z) = \exp(z)$              | Poisson     | Nonnegative/counts, no natural upper bound |
| $G(z) = \text{logit}^{-1}(z)$ | Binomial    | Nonnegative, known upper bound             |
| $G(z) = \text{logit}^{-1}(z)$ | Bernoulli   | Binary or fractional responses             |

------------------------------------------------------------------------

#### Inference

-   Standard errors can be obtained via the delta method or bootstrap
-   Cluster-robust standard errors by unit are preferred
-   When using QMLE, the estimates are valid under correct mean specification, regardless of higher moments

When Do Imputation and Pooled Methods Match?

-   They are numerically identical when:
    -   Estimating with the canonical link function
    -   Model is correctly specified
    -   Same data used for both (i.e., `W_it = 0` and pooled)

------------------------------------------------------------------------

#### Application Using `etwfe`

The `etwfe` package provides a unified, user-friendly interface for estimating staggered treatment effects using **generalized linear models**. It is particularly well-suited for nonlinear outcomes, such as **binary**, **fractional**, or **count** data.

We'll now demonstrate how to apply `etwfe` to estimate [Average Treatment Effect on the Treated] (ATT) under a nonlinear DiD framework using a Poisson model. This aligns with the exponential conditional mean assumption discussed earlier.

1.  Install and load packages

```{r}
# --- 1) Load packages ---
# install.packages("fixest")
# install.packages("marginaleffects")
# install.packages("etwfe")
# install.packages("ggplot2")
# install.packages("modelsummary")

library(etwfe)
library(fixest)
library(marginaleffects)
library(ggplot2)
library(modelsummary)
set.seed(12345)
```

2.  Simulate a known data-generating process

Imagine a **multi-period business panel** where each "unit" is a *regional store* or *branch* of a large retail chain. Half of these stores eventually receive a new *marketing analytics platform* at some known time, which in principle changes their performance metric (e.g., weekly log sales). The other half *never* receive the platform, functioning as a "never-treated" control group.

-   We have $N=200$ stores (half eventually treated, half never treated).

-   Each store is observed over $T=10$ time periods (e.g., quarters or years).

-   The *true* "treatment effect on the treated" is constant at $\delta = -0.05$ for all *post*-treatment times. (Interpretation: the new marketing platform *reduced* log-sales by about 5 percent, though in real life one might expect a positive effect!)

-   Some stores are "staggered" in the sense that they adopt in different periods. We'll randomly draw their adoption date from $\{4,5,6\}$. Others never adopt at all.

-   We include store-level intercepts, time intercepts, and idiosyncratic noise to make it more realistic.

```{r}
# --- 2) Simulate Data ---
N <- 200   # number of stores
T <- 10    # number of time periods
id   <- rep(1:N, each = T)
time <- rep(1:T, times = N)

# Mark half of them as eventually treated, half never
treated_ids <- sample(1:N, size = N/2, replace = FALSE)
is_treated  <- id %in% treated_ids

# Among the treated, pick an adoption time 4,5, or 6 at random
adopt_time_vec <- sample(c(4,5,6), size = length(treated_ids), replace = TRUE)
adopt_time     <- rep(0, N) # 0 means "never"
adopt_time[treated_ids] <- adopt_time_vec

# Store effects, time effects, control variable, noise
alpha_i <- rnorm(N, mean = 2, sd = 0.5)[id]
gamma_t <- rnorm(T, mean = 0, sd = 0.2)[time]
xvar    <- rnorm(N*T, mean = 1, sd = 0.3)
beta    <- 0.10
noise   <- rnorm(N*T, mean = 0, sd = 0.1)

# True treatment effect = -0.05 for time >= adopt_time
true_ATT <- -0.05
D_it     <- as.numeric((adopt_time[id] != 0) & (time >= adopt_time[id]))

# Final outcome in logs:
y <- alpha_i + gamma_t + beta*xvar + true_ATT*D_it + noise

# Put it all in a data frame
simdat <- data.frame(
    id         = id,
    time       = time,
    adopt_time = adopt_time[id],
    treat      = D_it,
    xvar       = xvar,
    logY       = y
)

head(simdat)
```

In this business setting, `logY` represents the natural log of revenue, sales, or another KPI, and `xvar` represents a log of local population, the number of competitor stores in the region, or a comparable covariate.

3.  Estimate with `etwfe`

We want to test whether the new marketing analytics platform has changed the log outcome. We will use `etwfe`:

-   `fml = logY ~ xvar` says that `logY` is the outcome, `xvar` is a control.

-   `tvar = time` is the time variable.

-   `gvar = adopt_time` is the *group/cohort* variable (the "first treatment time" or 0 if never).

-   `vcov = ~id` clusters standard errors at the store level.

-   `cgroup = "never"`: We specify that the never-treated units form our comparison group. This ensures we can see *pre-treatment* and *post-treatment* dynamic effects in an event-study plot.

```{r}
# --- 3) Estimate with etwfe ---
mod <- etwfe(
    fml    = logY ~ xvar,
    tvar   = time,
    gvar   = adopt_time,
    data   = simdat,
    # xvar   = moderator, # Heterogenous Treatment Effects
    vcov   = ~id,
    cgroup = "never"  # so that never-treated are the baseline
)
```

The raw coefficient list is uninformative on its own because the model is fully saturated with interactions. The substantive output is the aggregated treatment effects, which we obtain next.

4.  Recover the [ATT](#sec-average-treatment-effect-on-the-treated)

Here's a *single-number* estimate of the overall average effect on the treated, across all times and cohorts:

```{r}
# --- 4) Single-number ATT ---
ATT_est <- emfx(mod, type = "simple")
print(ATT_est)
```

You should see an estimate near the *true* $-0.05$.

5.  Recover an event-study pattern of dynamic effects

To check pre- and post-treatment dynamics, we ask for an **event study** via `type = "event"`. This shows how the outcome evolves *around* the adoption time. Negative "event" values correspond to pre-treatment, while nonnegative "event" values are post-treatment.

```{r}
# --- 5) Event-study estimates ---
mod_es <- emfx(mod, type = "event")
mod_es


# Renaming function to replace ".Dtreat" with something more meaningful
rename_fn = function(old_names) {
  new_names = gsub(".Dtreat", "Period post treatment =", old_names)
  setNames(new_names, old_names)
}

modelsummary(
  list(mod_es),
  shape       = term:event:statistic ~ model,
  coef_rename = rename_fn,
  gof_omit    = "Adj|Within|IC|RMSE",
  stars       = TRUE,
  title       = "Event study",
  notes       = "Std. errors are clustered at the id level"
)
```

-   By default, this will return events from (roughly) the earliest pre-treatment period up to the maximum possible post-treatment period in your data, using *never-treated* as the comparison group.

-   Inspect the estimates and confidence intervals. Ideally, pre-treatment estimates should be near 0, and post-treatment estimates near $-0.05$.

6.  Plot the estimated event-study vs. the true effect

In a business or marketing study, a useful final step is a chart showing the point estimates (with confidence bands) plus the known *true* effect as a reference.

Construct the "true" dynamic effect curve

-   Pre-treatment periods: effect = 0

-   Post-treatment periods: effect = $\delta=-0.05$

Below we will:

1.  Extract the estimated event effects from `mod_es`.

2.  Build a **reference** dataset with the same event times.

3.  Plot both on the same figure.

```{r}
# --- 6) Plot results vs. known effect ---
est_df <- as.data.frame(mod_es)

range_of_event <- range(est_df$event)
event_breaks   <- seq(range_of_event[1], range_of_event[2], by = 1)
true_fun <- function(e) ifelse(e < 0, 0, -0.05)
event_grid <- seq(range_of_event[1], range_of_event[2], by = 1)
true_df <- data.frame(
    event       = event_grid,
    true_effect = sapply(event_grid, true_fun)
)

ggplot() +
    # Confidence interval ribbon (put it first so it's behind everything)
    geom_ribbon(
        data = est_df,
        aes(x = event, ymin = conf.low, ymax = conf.high),
        fill = "grey60",   # light gray fill
        alpha = 0.3
    ) +
    # Estimated effect line
    geom_line(
        data = est_df,
        aes(x = event, y = estimate),
        color = "black",
        size = 1
    ) +
    # Estimated effect points
    geom_point(
        data = est_df,
        aes(x = event, y = estimate),
        color = "black",
        size = 2
    ) +
    # Known true effect (dashed red line)
    geom_line(
        data = true_df,
        aes(x = event, y = true_effect),
        color = "red",
        linetype = "dashed",
        linewidth = 1
    ) +
    # Horizontal zero line
    geom_hline(yintercept = 0, linetype = "dotted") +
    # Vertical line at event = 0 for clarity
    geom_vline(xintercept = 0, color = "gray40", linetype = "dashed") +
    # Make sure x-axis breaks are integers
    scale_x_continuous(breaks = event_breaks) +
    labs(
        title = "Event-Study Plot vs. Known True Effect",
        subtitle = "Business simulation with new marketing platform adoption",
        x = "Event time (periods relative to adoption)",
        y = "Effect on log-outcome (ATT)",
        caption = "Dashed red line = known true effect; Shaded area = 95% CI"
    ) +
    causalverse::ama_theme()
```

-   **Solid line** and shaded region: the ETWFE *point estimates* and their 95% confidence intervals, for each event time relative to adoption.

-   **Dashed red line**: the *true* effect that we built into the DGP.

If the estimation works well (and your sample is big enough), the estimated event-study effects should hover near the dashed red line *post*-treatment, and near zero *pre*-treatment.

Alternatively, we could also the `plot` function to produce a quick plot.

```{r}
plot(
    mod_es,
    type = "ribbon",
    # col  = "",# color
    xlab = "",
    main = "",
    sub  = "",
    # file = "event-study.png", width = 8, height = 5. # save file
)
```

7.  Double-check in a regression table (optional)

If you like to see a clean numeric summary of the *dynamic* estimates by period, you can pipe your event-study object into `modelsummary`:

```{r, eval = FALSE}
# --- 7) Optional table for dynamic estimates ---
modelsummary(
    list("Event-Study" = mod_es),
    shape     = term + statistic ~ model + event,
    gof_map   = NA,
    coef_map  = c(".Dtreat" = "ATT"),
    title     = "ETWFE Event-Study by Relative Adoption Period",
    notes     = "Std. errors are clustered by store ID"
)
```

------------------------------------------------------------------------
