## Modern Concerns in DiD {#sec-modern-concerns-in-did}

@roth2023whats provide a comprehensive synthesis of the recent DiD literature; readers may consult that paper as a roadmap to many of the techniques discussed in this chapter, including staggered adoption, heterogeneous treatment effects, parallel-trends sensitivity analysis, and conditional/relaxed identifying assumptions.

A useful way to organize the rest of this chapter is to recognize that "what could go wrong with DiD" decomposes into a small number of recurring failure modes, and that most of the modern toolkit is a targeted response to one (or several) of them. Heterogeneous and time-varying treatment effects break the implicit weighting in two-way fixed effects, motivating the staggered-adoption estimators introduced in Section \@ref(sec-modern-estimators-for-staggered-adoption) and the [Goodman-Bacon decomposition](#sec-goodman-bacon-decomposition). Departures from parallel trends motivate visual diagnostics in Section \@ref(sec-visualization-did), the [placebo tests](#sec-placebo-test-did) below, and the [partial-identification machinery](#sec-partial-identification-did) at the end of this chapter. Selection on observables that vary across treated and control units pulls in [matching](#sec-matching-methods) and the [conditional ignorability](#sec-conditional-ignorability-assumption) and [overlap](#sec-overlap-positivity-assumption) assumptions. Functional-form sensitivity, especially with zero-valued outcomes, motivates the [changes-in-changes](#sec-changes-in-changes) estimator and the proportional-effects approach below. Reading the rest of the chapter as a menu indexed by failure mode (rather than as a list of unrelated diagnostics) makes it easier to know which tool is the right one to reach for, and the closing sections show how a credible DiD analysis layers several of them.

------------------------------------------------------------------------

### Multiple Treatments

In some settings, researchers encounter **two (or more) treatments** rather than a single treatment-and-control comparison. This is more than a notational nuisance: with multiple treatments operating on overlapping populations, the standard DiD machinery can collapse identification onto contrasts the analyst did not intend, and the apparent treatment effect can shift depending on which group is implicitly used as a control. @fricke2017identification provides a careful treatment of the identification challenges that arise in this setting, and the principles below distil the practical implications.
<!-- @de2023two provides a **video tutorial** ([YouTube](https://www.youtube.com/watch?v=UHeJoc27qEM&ab_channel=TaylorWright)) and **code** ([Google Drive](https://drive.google.com/file/d/156Fu73avBvvV_H64wePm7eW04V0jEG3K/view)) for implementing multiple-treatment DiD models. -->

**Key Principles When Dealing with Multiple Treatments**

1.  **Always include all treatment groups in a single regression model.**

    -   This ensures proper identification of treatment-specific effects while maintaining a clear comparison against the control group.

2.  **Never use one treated group as a control for the other.**

    -   Running separate regressions for each treatment group can lead to biased estimates because each treatment group may differ systematically from the control group in ways that a separate model cannot fully capture.

3.  **Compare the significance of treatment effects** ($\delta_1$ vs. $\delta_2$).

    -   Instead of assuming equal effects, we should formally test whether the effects of the two treatments are statistically different using an F-test or Wald test:

    $$
    H_0: \delta_1 = \delta_2
    $$

    -   If we reject $H_0$, we conclude that the two treatments have significantly different effects.

------------------------------------------------------------------------

### Multiple Treatment Groups: Model Specification

A properly specified DiD regression model with two treatments takes the following form:

$$
\begin{aligned}
Y_{it} &= \alpha + \gamma_1 Treat1_{i} + \gamma_2 Treat2_{i} + \lambda Post_t  \\
&+ \delta_1(Treat1_i \times Post_t) + \delta_2(Treat2_i \times Post_t) + \epsilon_{it}
\end{aligned}
$$

where:

-   $Y_{it}$ = Outcome variable for individual $i$ at time $t$.
-   $Treat1_i$ = 1 if individual $i$ is in **Treatment Group 1**, 0 otherwise.
-   $Treat2_i$ = 1 if individual $i$ is in **Treatment Group 2**, 0 otherwise.
-   $Post_t$ = 1 for post-treatment period, 0 otherwise.
-   DiD coefficients: 
    -   $\delta_1$ = Effect of Treatment 1.
    -   $\delta_2$ = Effect of Treatment 2.
-   $\epsilon_{it}$ = Error term.

------------------------------------------------------------------------

### Understanding the Control Group in Multiple Treatment DiD

One common concern in multiple-treatment DiD models is how to properly define the control group. A well-specified model ensures that:

-   The control group consists only of untreated individuals, not individuals from another treatment group.
-   The reference category in the regression represents the control group (i.e., individuals with $Treat1_i = 0$ and $Treat2_i = 0$).
-   If $Treat1_i = 1$, then $Treat2_i = 0$ and vice versa.

Failing to correctly specify the control group could lead to incorrect estimates of treatment effects. For example, omitting one of the treatment indicators could unintentionally redefine the control group as a mix of treated and untreated individuals.

------------------------------------------------------------------------

### Alternative Approaches: Separate Regressions vs. One Model

A common question is whether to run one large regression including all treatment groups or to run separate DiD models on subsets of the data. Each approach has implications:

1.  **One Model Approach (Preferred)**

-   Running one comprehensive regression allows for direct comparison between treatment effects in a statistically valid way.
-   The interaction terms ($\delta_1, \delta_2$) ensure that each treatment effect is estimated relative to a common control group.
-   The F-test (or Wald test) enables a formal test of whether the two treatments have significantly different effects.

2.  **Separate Regressions Approach**

-   Running separate DiD models for each treatment group can still be valid, but:
    -   The estimated treatment effects are less efficient because they come from separate samples.
    -   Comparisons become less straightforward, as they rely on confidence interval overlap rather than direct hypothesis testing.
    -   If homoscedasticity holds (i.e., equal error variances across groups), the separate regressions approach is unnecessary. The combined model is more efficient.

Thus, unless there is strong justification for separate regressions (e.g., significant heterogeneity in error variance), the one-model approach is preferred.

------------------------------------------------------------------------

### Handling Treatment Intensity

In some cases, treatments differ not just in type, but also in intensity (e.g., low vs. high treatment exposure). If we observe different levels of treatment intensity, we can model it using a single categorical variable rather than multiple treatment dummies:

Rather than coding separate dummies for each treatment group, we define a multi-valued treatment variable:

$$
Y_{it} = \alpha + \sum_{j=1}^{J} \beta_j (Treatment_j \times Post_t) + \lambda Post_t + \epsilon_{it}
$$

where:

-   $Treatment_j$ is a categorical variable indicating whether an individual belongs to the control group, low-intensity treatment, or high-intensity treatment.
-   This approach allows for cleaner implementation and avoids excessive interaction terms.

This approach has the advantage of:

-   Automatically setting the control group as the reference category.

-   Ensuring correct interpretation of coefficients for different treatment levels.

### Considerations When Individuals Can Move Between Treatment Groups

One potential complication in multiple-treatment DiD settings is when individuals can switch treatment groups over time (e.g., moving from low-intensity to high-intensity treatment after policy implementation).

-   If movement is rare, it may not significantly affect estimates.

-   If movement is frequent, it creates a challenge in causal identification because treatment effects might be confounded by self-selection.

A possible solution is to use an intention-to-treat (ITT) approach, where treatment assignment is based on the initially assigned group, regardless of whether individuals later switch.

### Parallel Trends Assumption in Multiple-Treatment DiD

-   Just as in standard DiD, a key assumption in multiple-treatment DiD models is that the treatment and control groups would have followed parallel trends in the absence of treatment.

-   With multiple treatments, we must check pre-trends separately for each treated group against the control group.

-   If pre-treatment trends are not parallel, we may need to adopt alternative methods such as [synthetic control](#sec-synthetic-control) models or event study analyses.

It is worth noting how multiple-treatment settings interact with the staggered-adoption literature. Once treatments differ in timing as well as in type, the pathologies surveyed in Section \@ref(sec-multiple-periods-and-variation-in-treatment-timing) reappear: each treated arm becomes both a treated unit relative to the never-treated and a potential "forbidden comparison" for some other arm in a [two-way fixed-effects](#sec-two-way-fixed-effects) regression. Reaching for the [group-time average treatment effect](#sec-group-time-average-treatment-effects-callaway2021difference) framework, the [cohort ATT](#sec-cohort-average-treatment-effects-sun2021estimating) estimator, or the [stacked DiD](#sec-stacked-difference-in-differences) design is often more defensible than trying to extend the canonical interaction model to many treatment arms by hand.

------------------------------------------------------------------------

## Mediation Under DiD

Mediation analysis helps determine whether a treatment affects the outcome directly or through an intermediate variable (mediator). In a DiD framework, this allows us to separate:

1.  Direct effects: The effect of the treatment on the outcome independent of the mediator.
2.  Indirect (mediated) effects: The effect of the treatment that operates through the mediator.

This is useful when a treatment consists of multiple components or when we want to understand mechanisms behind an observed effect.

------------------------------------------------------------------------

### Mediation Model in DiD

To incorporate mediation, we estimate two equations:

**Step 1: Effect of Treatment on the Mediator**

$$
M_{it} = \alpha + \gamma Treat_i + \lambda Post_t + \delta (Treat_i \times Post_t) + \epsilon_{it}
$$ where:

-   $M_{it}$ = Mediator variable (e.g., job search intensity, firm investment, police presence).
-   $\delta$ = Effect of the treatment on the mediator (capturing how the treatment changes $M$).

**Step 2: Effect of Treatment and Mediator on the Outcome**

$$
Y_{it} = \alpha' + \gamma' Treat_i + \lambda' Post_t + \delta' (Treat_i \times Post_t) + \theta M_{it} + \epsilon'_{it}
$$ where:

-   $Y_{it}$ = Outcome variable (e.g., employment, crime rate, firm performance).
-   $\theta$ = Effect of the mediator on the outcome.
-   $\delta'$ = Direct effect of the treatment (controlling for the mediator).

------------------------------------------------------------------------

### Interpreting the Results

-   If $\theta$ is statistically significant, it suggests that mediation is occurring (i.e., the treatment affects the outcome partly through the mediator).
-   If $\delta'$ is smaller than $\delta$, this indicates that part of the treatment effect is explained by the mediator. The remaining portion of $\delta'$ represents the direct effect.

Thus, we can decompose the total treatment effect as:

$$
\text{Total Effect} = \delta' + (\theta \times \delta)
$$

where:

-   $\delta'$ = Direct effect (holding the mediator constant).

-   $\theta \times \delta$ = Indirect (mediated) effect.

------------------------------------------------------------------------

### Challenges in Mediation Analysis for DiD

Mediation in a DiD setting introduces several challenges that require careful consideration:

1.  **Potential Confounding of the Mediator**

-   A key assumption is that no unmeasured confounders affect both the mediator and the outcome.
-   If such confounders exist, estimates of $\theta$ may be biased.

2.  **Mediator-Outcome Endogeneity**

-   If the mediator is itself influenced by unobserved factors correlated with the outcome, it introduces endogeneity, making direct OLS estimates of $\theta$ problematic.
-   For example, in a crime policy evaluation:
    -   The number of police officers (mediator) may be influenced by crime rates (outcome), leading to reverse causality.

3.  **Interaction Between Multiple Mediators**

-   If there are multiple mediators (e.g., a policy that increases both police presence and surveillance cameras), they may interact with each other.
-   A useful test is to regress each mediator on treatment and other mediators. If a mediator predicts another, their effects are not independent, complicating interpretation.

------------------------------------------------------------------------

### Alternative Approach: Instrumental Variables for Mediation

One way to address mediator endogeneity is to use **instrumental variables**, where treatment serves as an instrument for the mediator:

**Two-Stage Estimation:**

1.  **First Stage: Predict the Mediator Using the Treatment** $$
    M_{it} = \alpha + \pi Treat_i + \lambda Post_t + \delta (Treat_i \times Post_t) + \nu_{it}
    $$
2.  **Second Stage: Predict the Outcome Using the Instrumented Mediator** $$
    Y_{it} = \alpha' + \gamma' Treat_i + \lambda' Post_t + \phi \hat{M}_{it} + \epsilon'_{it}
    $$

-   Here, $\hat{M}_{it}$ (predicted values from the first stage) replaces $M_{it}$, eliminating endogeneity concerns if the exclusion restriction holds (i.e., treatment only affects $Y$ through $M$).

**Key Limitation of IV Approach**

-   The IV strategy assumes that treatment affects the outcome only through the mediator, which may be too strong of an assumption in complex policy settings.

------------------------------------------------------------------------

## Assumptions

The list of assumptions below is best read as a mapping between identification conditions and diagnostics. Parallel trends is tested visually and via event studies, then stress-tested with [partial identification](#sec-partial-identification-did) when it is suspect. No anticipation is examined in pre-period event-study coefficients and through [placebo tests](#sec-placebo-test-did). Stable composition and exogeneity of shocks justify the design itself and connect to the broader [limitations of quasi-experiments](#sec-limitations-of-quasi-experiments). [Overlap](#sec-overlap-positivity-assumption) and effect homogeneity are the assumptions that, when violated, push the analyst toward [matching methods](#sec-matching-methods) and modern heterogeneity-robust estimators in Section \@ref(sec-modern-estimators-for-staggered-adoption). Each subsection below also notes the failure mode in language that can be cross-referenced back to the corresponding diagnostic later in the chapter.

1.  **Parallel Trends Assumption**

The Difference-in-Differences estimator relies on a key identifying assumption: the parallel trends assumption. This assumption states that, in the absence of treatment, the average outcome for the treated group would have evolved over time in the same way as for the control group.

Let:

-   $Y_{it}(0)$ denote the potential outcome without treatment for unit $i$ at time $t$

-   $D_i = 1$ if unit $i$ is in the treatment group, and $D_i = 0$ if in the control group

Then, the parallel trends assumption can be written as:

$$
E[Y_{it}(0) \mid D_i = 1] - E[Y_{it}(0) \mid D_i = 0] = \Delta_0 \quad \text{for all } t,
$$

where $\Delta_0$ is a constant difference over time in the untreated potential outcomes between the two groups. This assumption does not require the levels of outcomes to be the same between groups, only that the difference remains constant over time.

In other words, the gap between treatment and control groups in the absence of treatment must remain stable. If this holds, any deviation from that stable difference after the treatment is attributed to the causal effect of the treatment.

------------------------------------------------------------------------

It is important to understand how the parallel trends assumption compares to other, stronger assumptions:

-   Same untreated levels across groups: \
    A stronger assumption would require that the treated and control groups have **identical untreated outcomes at all times**:

    $$
    E[Y_{it}(0) \mid D_i = 1] = E[Y_{it}(0) \mid D_i = 0] \quad \text{for all } t
    $$

    This is often unrealistic in observational settings, where baseline characteristics typically differ between groups.

-   No change in untreated outcomes over time: \
    Another strong assumption is that **untreated outcomes remain constant over time**, for both groups:

    $$
    E[Y_{it}(0)] = E[Y_{i,t'}(0)] \quad \text{for all } i \text{ and times } t, t'
    $$

    This implies no secular trends, which is rarely plausible in real-world applications where outcomes (e.g., sales, earnings, health metrics) naturally evolve over time.

The **parallel trends assumption is weaker** than both of these and is generally more defensible, especially when supported by pre-treatment data.

------------------------------------------------------------------------

DiD is appropriate when:

-   You have pre-treatment and post-treatment outcome data

-   You have clearly defined treatment and control groups

-   The parallel trends assumption is plausible

Avoid using DiD when:

-   Treatment assignment is not random or quasi-random

-   Unobserved confounders may cause the groups to evolve differently over time

Testing Parallel Trends: [Prior Parallel Trends Test](#prior-parallel-trends-test).

------------------------------------------------------------------------

2.  **No Anticipation Effect (Pre-Treatment Exogeneity)**

-   Individuals or groups should not change their behavior before the treatment is implemented in expectation of the treatment.

-   If units anticipate the treatment and adjust their behavior beforehand, it can introduce bias in the estimates.

3.  **Exogenous Treatment Assignment**

-   Treatment should not be assigned based on potential outcomes.
-   Ideally, assignment should be as good as random, conditional on observables.

4.  **Stable Composition of Groups (No Attrition or Spillover)**

-   Treatment and control groups should remain stable over time.
-   There should be no selective attrition (where individuals enter/leave due to treatment).
-   No spillover effects: Control units should not be indirectly affected by treatment.

5.  **No Simultaneous Confounding Events (Exogeneity of Shocks)**

-   There should be no other major shocks that affect treatment/control groups differently at the same time as treatment implementation.

6.  **Random Sampling (or Representative Sampling)**

-   Assumption: The units (e.g., individuals, firms, regions) in the treatment and control groups are drawn from the same population, or at least from populations that would follow parallel trends in the absence of treatment.

-   Implication: Ensures external validity. It allows the DID estimator to generalize to the population of interest and prevents selection bias that would arise if treated and control units were systematically different at baseline.

7.  **Overlap (also called Common Support)**

-   Assumption: There is a positive probability of being treated or untreated for all units. Formally, for any covariate profile, there must be both treated and control units.

-   Implication: Without overlap, it becomes impossible to construct a valid counterfactual for some treated units. This assumption is especially important in staggered or generalized DID settings where treatment timing varies.

-   Diagnostic: One can assess this assumption using propensity score distributions or visualizations (e.g., covariate balance plots).

8.  **Effect Additivity (also called Constant Treatment Effect or Additive Structure or Effect Homogeneity)**

-   Assumption: The treatment effect is additive and homogeneous across units and time, meaning that the treatment only shifts the outcome level, not its trend or curvature. In the simplest form: $Y_{it}(1) - Y_{it}(0) = \delta$ for all $i$,$t$

-   Implication: This assumption is not required for identification per se, but it simplifies estimation and interpretation. Violations may bias estimates if treatment effects vary in ways that interact with time or unit characteristics.

-   Alternative Approaches: More recent DID methods (e.g., Callaway & Sant'Anna, Sun & Abraham) relax this assumption by allowing treatment effect heterogeneity.

In a standard DID setup, researchers often *use* the phrases "effect additivity" and "effect homogeneity" interchangeably, because the customary linear-in-levels model embeds *both* ideas. Strictly speaking, however, they are not identical (Table \@ref(tab:did-differences-effect-additivity-homogeneity)).

| Concept                | What it literally requires                                                                                                                                                                                                                  | What it rules out                                                                                   | Relationship                                                                                                                                                                           |
|-----------------|---------------------|-----------------|-----------------|
| **Effect Additivity**  | Treated potential outcome equals untreated potential outcome plus a constant shift: $Y_{it}(1)=Y_{it}(0)+\tau$. The treatment enters the outcome additively and the size of the shift does not depend on the baseline level of the outcome. | Any multiplicative or nonlinear functional form of the effect (e.g., a constant percentage change). | Implies a constant *level* effect. Guarantees homogeneity *and* linearity.                                                                                                             |
| **Effect Homogeneity** | The treatment effect $\tau$ is constant across units and periods: $\tau_i=\tau_t=\tau$. It says nothing about whether that constant is added, multiplied, or otherwise applied.                                                             | Unit-specific or time-varying treatment effects.                                                    | A broader statement. You can have homogeneous *multiplicative* effects ($Y_{it}(1)=Y_{it}(0)\times\lambda$ with constant $\lambda$). Those satisfy homogeneity but violate additivity. |

Table: (\#tab:did-differences-effect-additivity-homogeneity) Differences between effect additivity and effect homogeneity.

Why they get conflated in practice

1.  **Linear DID regression in levels.**\
    The canonical two-way fixed-effects model, $Y_{it}= \alpha_i+\lambda_t+\tau D_{it}+\varepsilon_{it}$ identifies $\tau$ only if (a) the effect is additive and (b) the constant $\tau$ is the same for every $i$, $t$. Hence both additivity and homogeneity are baked in; many textbooks label the joint restriction simply "homogeneous treatment effects"

2.  **Terminology drift.**\
    Recent heterogeneity papers contrast "TWFE with homogeneous effects" against "heterogeneous effects" without separating additivity from constancy, so the two terms are often treated as synonyms even though, conceptually, homogeneity could hold on another scale.

Practical implications

-   If you keep the outcome in **levels**, invoking *homogeneous treatment effects* effectively commits you to *additivity* as well.

-   If the outcome is log-transformed, **homogeneous effects** correspond to a constant percentage change, which is *multiplicative*, so additivity no longer holds.

-   Robustness exercises that re-estimate the model in logs, differences, or growth rates test sensitivity to the additivity choice; stability across scales supports the idea that it is the constancy, not the functional form, that drives the results.

------------------------------------------------------------------------

### Prior Parallel Trends Test {#prior-parallel-trends-test}

The parallel trends assumption ensures that, absent treatment, the treated and control groups would have followed similar outcome trajectories. Testing this assumption involves visualization and statistical analysis.

@marcus2021role discuss pre-trend testing in staggered DiD.

1.  **Visual Inspection: Outcome Trends and Treatment Rollout**

-   Plot raw outcome trends for both groups before and after treatment.
-   Use event-study plots to check for pre-trend violations and anticipation effects.
-   Visualization tools like `ggplot2` or `panelView` help illustrate treatment timing and trends.

2.  **Event-Study Regressions**

A formal test for pre-trends uses the event-study model:

$$ Y_{it} = \alpha + \sum_{k=-K}^{K} \beta_k 1(T = k) + X_{it} \gamma + \lambda_i + \delta_t + \epsilon_{it} $$

where:

-   $1(T = k)$ are time dummies for periods before and after treatment.

-   $\beta_k$ captures deviations in outcomes before treatment; these should be statistically indistinguishable from zero if parallel trends hold.

-   $\lambda_i$ and $\delta_t$ are unit and time fixed effects.

-   $X_{it}$ are optional covariates.

**Violation of parallel trends** occurs if pre-treatment coefficients ($\beta_k$ for $k < 0$) are statistically significant.

3.  **Statistical Test for Pre-Treatment Trend Differences**

Using only pre-treatment data, estimate:

$$
Y = \alpha_g + \beta_1 T + \beta_2 (T \times G) + \epsilon
$$

where:

-   $\beta_2$ measures differences in time trends between groups.

-   If $\beta_2 = 0$, trends are parallel before treatment.

**Considerations:**

-   Alternative functional forms (e.g., polynomials or nonlinear trends) can be tested.
-   If $\beta_2 \neq 0$, potential explanations include:
    -   Large sample size driving statistical significance.
    -   Small deviations in one period disrupting an otherwise stable trend.

While time fixed effects can partially address violations of parallel trends (and are commonly used in modern research), they may also absorb part of the treatment effect, especially when treatment effects vary over time [@wolfers2003business].

------------------------------------------------------------------------

**Debate on Parallel Trends**

-   Levels vs. Trends: @kahn2020promise argue that **similarity in levels** is also crucial. If treatment and control groups start at **different levels**, why assume their trends will be the same?
    -   Solution: 
        -   Plot time series for the treated and control groups.
        -   Use matched samples to improve comparability [@ryan2019now] (useful when parallel trends assumption is questionable).
    -   If levels differ significantly, functional form assumptions become more critical and must be justified.
-   Power of Pre-Trend Tests: 
    -   Pre-trend tests often lack statistical power, making false negatives common [@roth2022pretest].
    -   See: [PretrendsPower](https://github.com/jonathandroth/PretrendsPower) and [pretrends](https://github.com/jonathandroth/pretrends) (for adjustments).
-   Outcome Transformations Matter: 
    -   The parallel trends assumption is specific to both the transformation and units of the outcome variable [@roth2023parallel].
    -   Conduct falsification tests to check whether the assumption holds under different functional forms.

------------------------------------------------------------------------

```{r}
library(tidyverse)
library(fixest)
od <- causaldata::organ_donations %>%
    # Use only pre-treatment data
    filter(Quarter_Num <= 3) %>% 
    # Treatment variable
    dplyr::mutate(California = State == 'California')
```

```{r causalverse-plot-par-trends, fig.cap="Parallel trends for the rate outcome between California and control units.", fig.alt='Line plot showing parallel trends between California (treated) and control units over time quarters, displaying the Rate metric without confidence intervals'}
# use my package
causalverse::plot_par_trends(
    data = od,
    metrics_and_names = list("Rate" = "Rate"),
    treatment_status_var = "California",
    time_var = list(Quarter_Num = "Time"),
    display_CI = F
)
```

```{r fig-parallel-trends-manual, fig.cap='Manual parallel trends plot showing average rates.', fig.alt='Line plot displaying average Rate values across quarters, with separate lines for California (treated) and control groups, using AMA theme styling'}
# do it manually
# always good but plot the dependent out
od |>
    # group by treatment status and time
    dplyr::group_by(California, Quarter) |>
    dplyr::summarize_all(mean) |>
    dplyr::ungroup() |>
    # view()
    
    ggplot2::ggplot(aes(x = Quarter_Num, y = Rate, color = California)) +
    ggplot2::geom_line() +
    causalverse::ama_theme()
```

Figures \@ref(fig:causalverse-plot-par-trends) and \@ref(fig:fig-parallel-trends-manual) show the raw outcomes by periods, while Figure \@ref(fig:fig-prior-trend-coef) shows their coefficient estimates' version.

```{r fig-prior-trend-coef, fig.cap='Coefficient plot for the prior-trend analysis.', fig.alt='Coefficient plot displaying point estimates and confidence intervals from the prior trend regression model, without grid lines'}
# but it's also important to use statistical test
prior_trend <- fixest::feols(Rate ~ i(Quarter_Num, California) |
                                 State + Quarter,
                             data = od)

fixest::coefplot(prior_trend, grid = F)
# fixest::iplot(prior_trend, grid = F)
```

If one of the periods is significantly different from 0, our parallel trends assumption is not plausible.

In cases where the parallel trends assumption is questionable, researchers should consider methods for assessing and addressing potential violations. Some key approaches are discussed in @rambachan2023more:

-   Imposing Restrictions: Constrain how different the post-treatment violations of parallel trends can be relative to pre-treatment deviations.

-   [**Partial Identification**](#sec-partial-identification-did): Rather than assuming a single causal effect, derive bounds on the ATT.

-   Sensitivity Analysis: Evaluate how sensitive the results are to potential deviations from parallel trends.

To implement these approaches, the `HonestDiD` package by @rambachan2023more provides robust statistical tools. The demonstration below builds a panel in which the parallel trends assumption fails in a way that an event study cannot flag: the treated group suffers a one-time dip in the reference period $t = 5$ (an anticipation shock), so every event-study coefficient is measured against a contaminated baseline. The true post-treatment effect is 1, yet the conventional confidence interval for the period 6 effect sits well above it.

```{r honestdid-dgp, warning=FALSE, message=FALSE}
library(fixest)
library(HonestDiD)

set.seed(123)
n_id <- 500
n_t  <- 10

panel <- expand.grid(id = 1:n_id, t = 1:n_t)
panel$g <- as.integer(panel$id <= n_id / 2)

unit_fe <- rnorm(n_id)
time_fe <- rnorm(n_t, 0, 0.1)

# True effect of 1 from t = 6 onward, but the reference period t = 5
# suffers a one-time dip of 1 in the treated group
panel$y <- 1 + 2 * panel$g +
    1 * (panel$t >= 6 & panel$g == 1) -
    1 * (panel$t == 5 & panel$g == 1) +
    unit_fe[panel$id] + time_fe[panel$t] + rnorm(nrow(panel), 0, 5)

es <- feols(y ~ i(t, g, ref = 5) | id + t, cluster = "id", data = panel)
iplot(es, main = "Event study with a contaminated reference period")
```

The event-study coefficients feed directly into `HonestDiD`, which needs only the coefficient vector and its variance-covariance matrix. The conventional confidence set treats the pre-period deviations as zero; the robust confidence sets bound the post-treatment violation by a multiple $\bar{M}$ of the largest pre-treatment violation (the relative magnitudes restriction).

```{r honestdid-rm, warning=FALSE, message=FALSE}
betahat <- coef(es)
sigma   <- vcov(es)
npre  <- 4
npost <- 5

# Conventional CI for the period 6 effect (excludes the true effect of 1)
orig_ci <- constructOriginalCS(
    betahat = betahat, sigma = sigma,
    numPrePeriods = npre, numPostPeriods = npost,
    l_vec = basisVector(1, npost)
)
orig_ci

# Robust CIs: post-treatment violation at most Mbar times the pre-treatment one
rm_ci <- createSensitivityResults_relativeMagnitudes(
    betahat = betahat, sigma = sigma,
    numPrePeriods = npre, numPostPeriods = npost,
    l_vec = basisVector(1, npost),
    Mbarvec = seq(0.5, 1.5, by = 0.5)
)
rm_ci

createSensitivityPlot_relativeMagnitudes(rm_ci, orig_ci)
```

At $\bar{M} = 1$, meaning the post-treatment violation may be as large as the largest pre-treatment violation, the robust confidence set contains the true effect of 1 that the conventional interval excludes. The point at which the robust set first includes zero is the *breakdown value*, a single number summarizing how severe a violation would need to be to erase the finding.

When the worry is a smooth secular trend rather than a discrete shock, the smoothness restriction is the natural alternative: it allows the counterfactual trend to deviate from linearity by at most $M$ per period.

```{r honestdid-smooth, warning=FALSE, message=FALSE}
sm_ci <- createSensitivityResults(
    betahat = betahat, sigma = sigma,
    numPrePeriods = npre, numPostPeriods = npost,
    l_vec = basisVector(1, npost),
    Mvec = seq(0, 0.2, by = 0.1)
)
sm_ci

createSensitivityPlot(sm_ci, orig_ci)
```

Reporting either family of robust confidence sets alongside the conventional event study, together with the breakdown value, has become the standard way to communicate how much the conclusions lean on the parallel trends assumption. Averaging the post-treatment effects instead of isolating one period only requires replacing `l_vec` with a vector of equal weights.

Alternatively, @ban2022generalized propose a method that incorporates pre-treatment covariates as an information set and makes an assumption about the selection bias in the post-treatment period. Specifically, they assume that the selection bias lies within the convex hull of all pre-treatment selection biases. Under this assumption:

-   They identify a set of possible ATT values.

-   With a stronger assumption on selection bias, grounded in policymakers' perspectives, they can estimate a point estimate of ATT.

Another useful tool for assessing parallel trends is the `pretrends` package by @roth2022pretest, which provides formal pre-trend tests:

```{r}
# Install and load the pretrends package
# install.packages("pretrends")
# library(pretrends)

```

------------------------------------------------------------------------

### Placebo Test {#sec-placebo-test-did}

A placebo test is a diagnostic tool used in Difference-in-Differences analysis to assess whether the estimated treatment effect is driven by pre-existing trends rather than the treatment itself. The idea is to estimate a treatment effect in a scenario where no actual treatment occurred. If a significant effect is found, it suggests that the parallel trends assumption may not hold, casting doubt on the validity of the causal inference.

Placebo tests sit alongside the [prior parallel-trends test](#prior-parallel-trends-test) in a complementary role: where the event-study test asks whether deviations are visible *in the data we have*, the placebo test asks whether the estimator generates spurious effects when the treatment label is moved to a unit or period where, by construction, no causal effect can exist. A clean event study without a clean placebo is suggestive but unsatisfying, and a clean placebo with a noisy event study is harder to dismiss than the raw $p$-values alone would imply. Treating the two as a pair helps build the kind of layered defense of parallel trends that complements the partial-identification analysis later in this chapter.

**Types of Placebo DiD Tests**

1.  **Group-Based Placebo Test**

-   Assign treatment to a group that was never actually treated and rerun the DiD model.
-   If the estimated treatment effect is statistically significant, this suggests that differences between groups, not the treatment, are driving results.
-   This test helps rule out the possibility that the estimated effect is an artifact of unobserved systematic differences.

A valid treatment effect should be consistent across different reasonable control groups. To assess this:

-   Rerun the DiD model using an alternative but comparable control group.

-   Compare the estimated treatment effects across multiple control groups.

-   If results vary significantly, this suggests that the choice of control group may be influencing the estimated effect, indicating potential selection bias or unobserved confounding.

2.  **Time-Based Placebo Test**

-   Conduct DiD using only pre-treatment data, pretending that treatment occurred at an earlier period.
-   A significant estimated treatment effect implies that differences in pre-existing trends, not treatment, are responsible for observed post-treatment effects.
-   This test is particularly useful when concerns exist about unobserved shocks or anticipatory effects.

**Random Reassignment of Treatment**

-   Keep the same treatment and control periods but randomly assign treatment to units that were not actually treated.
-   If a significant DiD effect still emerges, it suggests the presence of biases, unobserved confounding, or systematic differences between groups that violate the parallel trends assumption.

------------------------------------------------------------------------

**Procedure for a Placebo Test**

1.  **Using Pre-Treatment Data Only**

A robust placebo test often involves analyzing only pre-treatment periods to check whether spurious treatment effects appear. The procedure includes:

-   Restricting the sample to pre-treatment periods only.

-   Assigning a fake treatment period before the actual intervention.

-   Testing a sequence of placebo cutoffs over time to examine whether different assumed treatment timings yield significant effects.

-   Generating random treatment periods and using randomization inference to assess the sampling distribution of the placebo effect.

-   Estimating the DiD model using the fake post-treatment period (`post_time = 1`).

-   Interpretation: If the estimated treatment effect is statistically significant, this indicates that pre-existing trends (not treatment) might be influencing results, violating the parallel trends assumption.

2.  **Using Control Groups for a Placebo Test**

If multiple control groups are available, a placebo test can also be conducted by:

-   Dropping the actual treated group from the analysis.

-   Assigning one of the control groups as a fake treated group.

-   Estimating the DiD model and checking whether a significant effect is detected.

-   Interpretation: 

    -   If a placebo effect appears (i.e., the estimated treatment effect is significant), it suggests that even among control groups, systematic differences exist over time.

    -   However, this result is not necessarily disqualifying. Some methods, such as [Synthetic Control], explicitly model such differences while maintaining credibility.

------------------------------------------------------------------------

```{r}
# Load necessary libraries
library(tidyverse)
library(fixest)
library(ggplot2)
library(causaldata)

# Load the dataset
od <- causaldata::organ_donations %>%
    # Use only pre-treatment data
    dplyr::filter(Quarter_Num <= 3) %>%
    
    # Create fake (placebo) treatment variables
    dplyr::mutate(
        FakeTreat1 = as.integer(State == 'California' &
                                    Quarter %in% c('Q12011', 'Q22011')),
        FakeTreat2 = as.integer(State == 'California' &
                                    Quarter == 'Q22011')
    )

# Estimate the placebo effects using fixed effects regression
clfe1 <- fixest::feols(Rate ~ FakeTreat1 | State + Quarter, data = od)
clfe2 <- fixest::feols(Rate ~ FakeTreat2 | State + Quarter, data = od)

# Display the regression results
fixest::etable(clfe1, clfe2)

# Extract coefficients and confidence intervals
coef_df <- tibble(
    Model = c("FakeTreat1", "FakeTreat2"),
    Estimate = c(coef(clfe1)["FakeTreat1"], coef(clfe2)["FakeTreat2"]),
    SE = c(summary(clfe1)$coeftable["FakeTreat1", "Std. Error"], 
           summary(clfe2)$coeftable["FakeTreat2", "Std. Error"]),
    Lower = Estimate - 1.96 * SE,
    Upper = Estimate + 1.96 * SE
)
```

```{r fig-placebo-effects, fig.cap='Placebo treatment effects on organ donation rate.', fig.alt='Point plot with error bars showing estimated placebo treatment effects and their confidence intervals for different models. A horizontal dashed red line at zero indicates no effect. Blue points and error bars represent the estimates and their uncertainty.'}
# Plot the placebo effects
ggplot(coef_df, aes(x = Model, y = Estimate)) +
    geom_point(size = 3, color = "blue") +
    geom_errorbar(aes(ymin = Lower, ymax = Upper), width = 0.2, color = "blue") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    theme_minimal() +
    labs(
        title = "Placebo Treatment Effects",
        y = "Estimated Effect on Organ Donation Rate",
        x = "Placebo Treatment"
    )
```

We would like the "supposed" DiD to be insignificant (Figure \@ref(fig:fig-placebo-effects)).

## Robustness Checks

A well-executed Difference-in-Differences analysis requires robustness checks to verify the validity of estimated treatment effects and best practices to ensure methodological rigor. Whereas the [prior parallel-trends test](#prior-parallel-trends-test) and the [placebo test](#sec-placebo-test-did) interrogate the identifying assumption, the checks below interrogate the *modeling choices* layered on top of it: the time window, the functional form for trends, the choice of outcome, and the inclusion of an additional comparison dimension via triple differences. Stability of the estimate across these choices is what allows a reader to attribute the result to the policy variation rather than to a fortunate specification. The broader catalog of [robustness checks for quasi-experimental designs](#sec-robustness-checks-quasi-exp) and the discussion of [coefficient stability bounds](#sec-coefficient-stability-bounds) elsewhere in the book extend the same logic.

------------------------------------------------------------------------

### Robustness Checks to Strengthen Causal Interpretation

Once the parallel trends assumption is assessed, additional robustness tests ensure that treatment effects are not driven by confounding factors or modeling choices.

1.  **Varying the Time Window**

-   Shorter time windows reduce exposure to long-term confounders but risk losing statistical power.
-   Longer time windows capture persistent effects but may introduce unrelated policy changes.
-   Solution: Estimate the DiD model across different time horizons and check if results are stable.

2.  **Higher-Order Polynomial Time Trends**

-   Standard DiD models assume a linear time trend.
-   If trends are nonlinear, this assumption may be too restrictive.
-   Solution: Introduce quadratic or cubic time trends and verify whether results hold.

3.  **Testing Alternative Dependent Variables**

-   The treatment should only affect the expected dependent variable.
-   A robustness check involves running the DiD on unrelated dependent variables.
-   If treatment effects appear where they should not, this signals a possible identification problem.

4.  **Triple-Difference (DDD) Strategy**

A Triple-Difference (DDD) model adds an additional comparison group to address remaining biases:

$$
\begin{aligned}
Y_{ijt} &= \alpha + \gamma Treat_{i} + \lambda Post_t + \theta Group_j + \delta_1 (Treat_i \times Post_t) \\
&+ \delta_2 (Treat_i \times Group_j) + \delta_3 (Post_t \times Group_j) \\
&+ \delta_4 (Treat_i \times Post_t \times Group_j) + \epsilon_{ijt}
\end{aligned}
$$

where:

-   $Group_j$ represents a subgroup within treatment/control (e.g., high- vs. low-intensity exposure).

-   $\delta_4$ captures the DDD effect, which removes residual biases present in the standard DiD model.

------------------------------------------------------------------------

### Best Practices for Reliable DiD Implementation

To improve the credibility and transparency of DiD estimates, researchers should adhere to the following best practices:

1.  **Documenting Treatment Cohorts**

-   Clearly report the number of treated and control units over time.
-   If treatment is staggered, adjust for different exposure durations.

2.  **Checking Covariate Balance & Overlap**

-   Verify whether the distribution of covariates is similar across treatment and control groups.
-   If treatment and control groups differ significantly, consider using matching methods.

3.  **Conducting Sensitivity Analyses for Parallel Trends**

-   Apply alternative weighting schemes (e.g., entropy balancing) to reduce dependence on model assumptions.
-   Use `honestDiD` to test robustness under different parallel trends violations.

## Concerns in DID

The catalogue below collects the issues that recur most often in applied work and pairs each with the diagnostic or estimator that addresses it. Several of them (heterogeneous effects, treatment timing, questionable counterfactuals) feed directly into the modern staggered-adoption literature in Section \@ref(sec-modern-estimators-for-staggered-adoption); others (selection on unobservables, Ashenfelter's dip) connect more naturally to [Rosenbaum bounds](#sec-rosenbaum-bounds) and to the broader [conditional ignorability](#sec-conditional-ignorability-assumption) discussion. The point is not that every analysis must address every concern, but that a credible DiD application explicitly considers which of them are plausible in the context at hand and chooses diagnostics accordingly.

### Limitations and Common Issues

1.  **Functional Form Dependence**

-   If the response to treatment is nonlinear, compare high- vs. low-intensity groups.

2.  **Selection on (Time-Varying) Unobservables**

-   Use [Rosenbaum Bounds] to check the sensitivity of estimates to unobserved confounders.

3.  **Long-Term Effects**

-   Parallel trends are more reliable in short time windows.
-   Over long periods, other confounding factors may emerge.

4.  **Heterogeneous Effects**

-   Treatment intensity (e.g., different doses) may vary across groups, leading to different effects.

5.  **Ashenfelter's Dip** [@ashenfelter1978estimating]

-   Participants in job training programs often experience earnings drops before enrolling, making them systematically different from nonparticipants.
-   Fix: Compute long-run differences, excluding periods around treatment, to test for sustained impact [@proserpio2017online; @heckman1999economics; @jepsen2014labor].

6.  **Lagged Treatment Effects**

-   If effects are not immediate, using a lagged dependent variable $Y_{it-1}$ may be more appropriate [@blundell1998initial].

7.  **Bias from Unobserved Factors Affecting Trends**

-   If external shocks influence treatment and control groups differently, this biases DiD estimates.

8.  **Correlated Observations**

-   Standard errors should be clustered appropriately.

9.  **Incidental Parameters Problem [@lancaster2000incidental]**

-   Always prefer individual and time fixed effects to reduce bias.

10. **Treatment Timing and Negative Weights**

-   If treatment timing varies across units, negative weights can arise in standard DiD estimators when treatment effects are heterogeneous [@athey2022design; @borusyak2024revisiting; @goodman2021difference]. The [Goodman-Bacon decomposition](#sec-goodman-bacon-decomposition) is the standard diagnostic for the presence and severity of these forbidden comparisons.
-   **Fix:** Use estimators from @callaway2021difference and @de2020two (`did` package). Specifically, the [group-time ATT estimator](#sec-group-time-average-treatment-effects-callaway2021difference) and the [switching-DiD estimator](#sec-switching-difference-in-differences-estimator-de2020two) avoid the contaminated comparisons that produce negative weights, and the [stacked DiD](#sec-stacked-difference-in-differences) design provides a transparent alternative when the analyst prefers to keep a single regression.
-   If expecting lags and leads, see @sun2021estimating, whose [cohort ATT](#sec-cohort-average-treatment-effects-sun2021estimating) approach generalizes naturally to event-study settings.

11. **Treatment Effect Heterogeneity Across Groups**

-   If treatment effects vary across groups and interact with treatment variance, standard estimators may be invalid [@gibbons2018broken]. Beyond the staggered-adoption tools above, the [reshaped IPW TWFE estimator](#sec-reshaped-inverse-probability-weighting-twfe-estimator) and the [matrix-completion estimator](#sec-matrix-completion-estimator) are useful when heterogeneity is suspected to load on covariates or on idiosyncratic unit-time shocks rather than purely on cohort.

12. **Endogenous Timing**

If the timing of units can be influenced by strategic decisions in a DID analysis, an instrumental variable approach with a control function can be used to control for endogeneity in timing. See Section \@ref(sec-instrumental-variables) for the IV machinery and Section \@ref(sec-design-based) for design-based alternatives. A distinct and easily missed version of this problem arises when the adoption date is not chosen by the unit at all but is *measured* by applying a detector to the outcome; see [Outcome-Triggered Treatment Timing](#sec-outcome-triggered-timing) below.

13. **Questionable Counterfactuals**

In situations where the control units may not serve as a reliable counterfactual for the treated units, matching methods such as propensity score matching or generalized random forest can be utilized. Additional methods can be found in [Matching Methods]. When parallel trends in levels is implausible but the treated unit can be approximated by a weighted combination of controls, [synthetic control](#sec-synthetic-control) and the hybrid [synthetic DiD](#sec-synthetic-difference-in-differences) approach are natural fallbacks; the [panel-match estimator](#sec-panel-match-did-estimator-with-in-and-out-treatment-conditions) and the family of [counterfactual estimators](#sec-counterfactual-estimators) extend the idea to richer treatment-history settings.

### Outcome-Triggered Treatment Timing {#sec-outcome-triggered-timing}

Every concern above takes the adoption date $G_i$ as something the analyst observes. In a growing class of applications it is not observed but *inferred*: the researcher runs a detector over the data and declares that unit $i$ was treated in the first period the detector fires. Technology-adoption studies date a firm's adoption from the first patent, job posting, or product page mentioning the technology; studies of scientific practice date a researcher's adoption of a tool from the first paper in which the tool is detected; audit and enforcement studies date exposure from the first flagged filing. In each case the detector is applied to a stream of output that *is itself the outcome variable*, or is mechanically tied to it.

@renault2026scientific show that this construction is enough, on its own, to manufacture a large and statistically significant event-study effect where none exists. Their target is @kusumegi2025scientific, who date a researcher's adoption of large language models as the first month in which at least one submitted abstract crosses an LLM-detection threshold, and report sharp productivity gains afterwards. The problem is that a researcher who submits more papers in a month is mechanically more likely to have at least one of them flagged that month, so the estimated adoption date is drawn toward high-output months. @renault2026scientific show that random treatment assignment, neutral-keyword triggers, inverted treatment, and pre-ChatGPT placebo windows all reproduce the reported dynamics.

What makes this failure mode worth its own section is that it defeats the diagnostics the rest of this chapter relies on. It is not a heterogeneity problem, so the staggered-adoption estimators of Section \@ref(sec-modern-estimators-for-staggered-adoption) do not fix it. And, as the derivation below shows, it leaves pre-treatment coefficients *exactly flat in expectation*, so the [prior parallel trends test](#prior-parallel-trends-test) returns a clean bill of health precisely when the design is broken.

#### Setup and Notation

Let $Y_{it}$ be the outcome of unit $i$ in period $t$, $t = 1, \dots, T$, measured in units of output (papers, patents, filings, posts). Adoption is never directly observed. Instead, each of the $Y_{it}$ units of output is passed through a detector that flags it independently with probability $q \in (0,1)$. Write $F_{it}$ for the number of flagged units and

$$
D_{it} = \mathbf{1}\{F_{it} \ge 1\}, \qquad
G_i = \min\{t : D_{it} = 1\},
$$

with $G_i = \infty$ (never-treated) if the detector never fires. Event time is $k = t - G_i$, and the analyst runs the usual event-study regression from the [prior parallel trends test](#prior-parallel-trends-test),

$$
Y_{it} = \alpha + \sum_{k \neq -1} \beta_k \mathbf{1}\{t - G_i = k\} + \lambda_i + \delta_t + \varepsilon_{it},
$$

normalizing $\beta_{-1} = 0$.

**Assumption 1 (No treatment effect).** $Y_{it}$ is i.i.d. across $t$ within unit, with mean $\lambda$, and does not depend on $G_i$ or on any treatment. *Substantively:* the null we are simulating under is the sharpest possible one. There is no effect to find, no trend, no anticipation, and no heterogeneity. Anything the event study reports is an artifact.

**Assumption 2 (Output-increasing detection).** Conditional on $Y_{it}$, the number of flags $F_{it} \sim \text{Binomial}(Y_{it}, q)$, independently across $t$. *Substantively:* the detector is applied unit-by-unit to output and has no other input. It is not biased, not strategic, and not correlated with anything unobserved; it is simply run more times in a period when more output exists. *What would violate it:* a detector applied to a fixed-size sample per period regardless of output, or an adoption date taken from an administrative record independent of the outcome.

Note that Assumption 2 is deliberately benign. There is no strategic timing, no anticipation, and no selection on unobservables — exactly the conditions under which the analyst would ordinarily consider the timing exogenous.

#### The Mechanism

**Proposition.** Under Assumptions 1--2 with $Y_{it} \sim \text{Poisson}(\lambda)$, the expectation of $Y_{it}$ conditional on event time takes three values,

$$
E[Y_{it} \mid t < G_i] = \underbrace{\lambda(1-q)}_{\mu^-}, \qquad
E[Y_{iG_i}] = \underbrace{\frac{\lambda\left[1 - (1-q)e^{-\lambda q}\right]}{1 - e^{-\lambda q}}}_{\mu^+}, \qquad
E[Y_{it} \mid t > G_i] = \lambda,
$$

with $\mu^- < \lambda < \mu^+$. Consequently the population event-study coefficients are

$$
\beta_k = 0 \ \text{ for } k < 0, \qquad
\beta_0 = \mu^+ - \mu^-, \qquad
\beta_k = \lambda q \ \text{ for } k > 0 .
$$

*Derivation.* Fix a unit and condition on $G_i = g$. By definition of a first-hitting time, $G_i = g$ is the event $\{F_{i1} = 0, \dots, F_{i,g-1} = 0, F_{ig} \ge 1\}$, and because the flags are drawn independently across periods given the outputs, this event restricts each period separately.

For $t < g$ the restriction is $F_{it} = 0$. Poisson thinning is what makes this tractable: if $Y \sim \text{Poisson}(\lambda)$ and $F \mid Y \sim \text{Binomial}(Y, q)$, then $F \sim \text{Poisson}(\lambda q)$ and $Y - F \sim \text{Poisson}(\lambda(1-q))$ *independently*. Hence $Y_{it} \mid F_{it} = 0$ is distributed as $Y - F \sim \text{Poisson}(\lambda(1-q))$, giving $E[Y_{it} \mid t < g] = \lambda(1-q) = \mu^-$. Crucially this does not depend on $t$ or on $g$: *every* pre-period is depressed by the same amount.

For $t = g$ the restriction is $F_{ig} \ge 1$, which occurs with probability $1 - P(F = 0) = 1 - e^{-\lambda q}$. Decomposing the unconditional mean,
$$
\lambda = E[Y] = P(F = 0)\, E[Y \mid F = 0] + P(F \ge 1)\, E[Y \mid F \ge 1]
= e^{-\lambda q}\lambda(1-q) + (1 - e^{-\lambda q})\, \mu^+,
$$
and solving for $\mu^+$ gives the expression above.

For $t > g$ there is no restriction at all: once the detector has fired, later periods are unconstrained by the definition of a *first* hitting time. So $E[Y_{it} \mid t > g] = \lambda$.

Because all three values are free of $g$, differencing against the reference period $k = -1$ (which has mean $\mu^-$) delivers the stated $\beta_k$, and the post-treatment plateau simplifies to
$$
\beta_k = \lambda - \lambda(1-q) = \lambda q, \qquad k > 0,
$$
which is exactly the expected number of flagged units of output per period. $\square$

#### Intuition

The event study is not measuring a response to treatment; it is measuring the shadow of the selection rule that defined the treatment date. Pre-treatment periods are, by construction, periods in which the detector *failed* to fire, and a period with unusually low output is a period in which failure was likely. Conditioning on non-detection therefore selects low draws, dragging every pre-period below the unit's own mean. The adoption period is the mirror image: it is the period in which detection *succeeded*, which selects high draws. Post-treatment periods carry no conditioning whatsoever and simply revert to the unconditional mean.

The resulting profile is a step function — depressed, spike, flat-at-normal — and when it is normalized to the last pre-period it reads exactly like the textbook picture of a successful intervention: flat pre-trends, a sharp jump on impact, and a stable positive plateau. Nothing happened. The unit merely returned to its own average after a period of below-average output that the design guaranteed by selecting on it. This is a regression-to-the-mean artifact given a treatment date, and it is closely related to Ashenfelter's dip (item 5 above) with one important difference: here the dip is not a behavioral fact about program participants but a mechanical consequence of how the analyst constructed $G_i$.

#### Simulation

The DGP below implements Assumptions 1--2 exactly: i.i.d. Poisson output, no trend, no treatment effect, and an adoption date defined as the first flagged period. The first check is whether the closed-form event-time means hold.

```{r outcome-triggered-dgp, message = FALSE, warning = FALSE}
library(fixest)
library(data.table)

# Output is i.i.d. Poisson with no trend and no treatment effect. Each unit of
# output is flagged independently with probability q, and the "adoption date" is
# the first period in which at least one unit is flagged.
sim_trigger_panel <- function(N, Tn, lambda, q, seed = NULL) {
    if (!is.null(seed)) set.seed(seed)
    Y  <- matrix(rpois(N * Tn, lambda), N, Tn)
    Fl <- matrix(rbinom(N * Tn, as.vector(Y), q), N, Tn)
    D  <- Fl >= 1
    G  <- max.col(D, ties.method = "first")
    G[rowSums(D) == 0] <- NA_integer_          # detector never fired
    dt <- data.table(
        id     = rep(1:N, each = Tn),
        period = rep(1:Tn, times = N),
        y      = as.vector(t(Y))
    )
    dt[, g := G[id]]
    dt[]
}

lambda <- 4      # mean units of output per period
q      <- 0.05   # per-unit probability of being flagged
Tn     <- 12
panel  <- sim_trigger_panel(N = 20000, Tn = Tn, lambda = lambda, q = q,
                            seed = 20260812)

# Closed-form event-time means from the Proposition
p_flag   <- 1 - exp(-lambda * q)
mu_minus <- lambda * (1 - q)
mu_plus  <- lambda * (1 - (1 - q) * exp(-lambda * q)) / p_flag

theory <- data.frame(
    event_time = c("k < 0", "k = 0", "k > 0"),
    theory     = c(mu_minus, mu_plus, lambda),
    simulated  = c(
        panel[!is.na(g) & period <  g, mean(y)],
        panel[!is.na(g) & period == g, mean(y)],
        panel[!is.na(g) & period >  g, mean(y)]
    )
)
theory$implied_beta <- theory$theory - mu_minus
round_df <- function(d) {
    d[] <- lapply(d, function(x) if (is.numeric(x)) round(x, 3) else x)
    d
}
round_df(theory)
```

The simulated means reproduce $\mu^- = 3.8$, $\mu^+ = 4.90$, and $\lambda = 4$ to three digits. The implied artifacts are a spike of $\mu^+ - \mu^- \approx 1.10$ on impact and a permanent plateau of $\lambda q = 0.2$, which on a base of $4$ is a $5\%$ "productivity gain" produced by a DGP containing no treatment effect.

Fitting the event-study regression confirms that the regression recovers exactly these numbers, and that the pre-trend diagnostic passes.

```{r fig-outcome-triggered-es, fig.cap = "Event-study coefficients under outcome-triggered treatment timing. The data-generating process contains no treatment effect of any kind. Pre-treatment coefficients are flat and insignificant, yet the on-impact and post-treatment coefficients are large and highly significant.", fig.alt = "Coefficient plot of event-study estimates by event time from a simulation with no true treatment effect. Coefficients for event times minus five through minus two lie on zero with confidence intervals covering zero. The coefficient at event time zero jumps to about 1.1, and coefficients from event time one onward sit on a plateau near 0.2, all with confidence intervals excluding zero.", message = FALSE, warning = FALSE}
panel[, rel  := fifelse(is.na(g), 1000L, period - g)]
panel[, relb := pmin(pmax(rel, -5L), 6L)]
panel[is.na(g), relb := -1L]     # never-detected units absorbed into the reference

es_trigger <- feols(y ~ i(relb, ref = -1) | id + period,
                    data = panel, cluster = ~ id)

iplot(es_trigger, main = "",
      xlab = "Event time (periods from first detection)",
      ylab = "Estimated coefficient", ref.line = -1, pt.join = TRUE)

coeftable(es_trigger)
```

```{r outcome-triggered-pretrend, message = FALSE, warning = FALSE}
# Joint test of the four pre-treatment coefficients
pre_test <- fixest::wald(es_trigger,
                         paste0("relb::-", 5:2, "$", collapse = "|"),
                         print = FALSE)
c(F_stat = pre_test$stat, df = pre_test$df1, p_value = pre_test$p)
```

The four pre-treatment coefficients are individually insignificant and jointly indistinguishable from zero, while $\hat\beta_0 \approx 1.10$ and the post-treatment coefficients sit on a plateau near $0.19$, matching the theoretical $1.103$ and $0.200$. An analyst following standard practice would report a passing pre-trend test alongside a highly significant dynamic effect.

To separate the artifact from sampling noise, and to benchmark against the design the book normally assumes, the following Monte Carlo re-runs the experiment $200$ times under two designs. The first is the outcome-triggered rule. The second permutes the adoption dates across units, which preserves the marginal distribution of $G_i$ (and therefore the cohort structure, the share of never-treated units, and the event-time sample sizes) while severing the link between a unit's adoption date and its own output. Because the true effect is zero everywhere, the mean estimate *is* the bias and the rejection rate *is* the size of the test.

```{r outcome-triggered-mc, message = FALSE, warning = FALSE}
one_rep <- function(N, Tn, lambda, q, exogenous = FALSE) {
    dt <- sim_trigger_panel(N, Tn, lambda, q)
    if (exogenous) {                     # break the output-timing link only
        gmap <- sample(dt[, .(g = g[1]), by = id]$g)
        dt[, g := gmap[id]]
    }
    dt[, rel  := fifelse(is.na(g), 1000L, period - g)]
    dt[, relb := pmin(pmax(rel, -5L), 6L)]
    dt[is.na(g), relb := -1L]
    fit <- feols(y ~ i(relb, ref = -1) | id + period, data = dt, cluster = ~ id)
    ct  <- coeftable(fit)
    grab <- function(k, j) {
        r <- paste0("relb::", k)
        if (r %in% rownames(ct)) unname(ct[r, j]) else NA_real_
    }
    w <- fixest::wald(fit, paste0("relb::-", 5:2, "$", collapse = "|"),
                      print = FALSE)
    c(b0 = grab(0, 1), p0 = grab(0, 4),
      b3 = grab(3, 1), p3 = grab(3, 4),
      pre_p = unname(w$p))
}

set.seed(20260812)
R <- 200
mc_trigger <- t(replicate(R, one_rep(1500, Tn, lambda, q, exogenous = FALSE)))
mc_exog    <- t(replicate(R, one_rep(1500, Tn, lambda, q, exogenous = TRUE)))

summarise_mc <- function(m, label) data.frame(
    design          = label,
    bias_k0         = mean(m[, "b0"]),
    rmse_k0         = sqrt(mean(m[, "b0"]^2)),
    reject_k0       = mean(m[, "p0"] < 0.05),
    bias_k3         = mean(m[, "b3"]),
    rmse_k3         = sqrt(mean(m[, "b3"]^2)),
    reject_k3       = mean(m[, "p3"] < 0.05),
    reject_pretrend = mean(m[, "pre_p"] < 0.05)
)

round_df(rbind(
    summarise_mc(mc_trigger, "Outcome-triggered timing"),
    summarise_mc(mc_exog,    "Exogenous timing")
))
```

The contrast is stark. Under exogenous timing the estimator behaves exactly as it should: bias near zero at both event times, and rejection rates of $0.025$ and $0.055$ against a nominal $5\%$. Under outcome-triggered timing the on-impact coefficient is biased by $1.10$ and rejects in $100\%$ of replications, while the $k = 3$ coefficient is biased by $0.18$ and rejects in $52\%$ — a test with nominal size $5\%$ that falsely declares a significant effect in more than half of samples where no effect exists. The final column is the one that should worry practitioners most: the joint pre-trend test rejects in $2.5\%$ of replications under the broken design, no more often than under the valid one. **The pre-trend test has no power against this violation at all**, because the Proposition shows the pre-period means are equal to each other. A flat pre-trend is evidence about the pre-period; it is silent about how the treatment date was constructed.

##### What Does Not Fix It

Because the symptom — significant dynamic effects in a staggered panel — resembles the heterogeneity problems of Section \@ref(sec-modern-estimators-for-staggered-adoption), the natural reflex is to reach for a heterogeneity-robust estimator. That reflex fails here.

```{r outcome-triggered-sunab, message = FALSE, warning = FALSE}
sa_trigger <- feols(y ~ sunab(g, period) | id + period,
                    data = panel[!is.na(g)], cluster = ~ id)

sa_ct <- coeftable(sa_trigger)
round(sa_ct[rownames(sa_ct) %in% paste0("period::", -3:4), ], 3)
```

The @sun2021estimating estimator returns an on-impact coefficient of $1.29$ and a post-treatment plateau near $0.40$ — larger, not smaller, than plain TWFE. This is the expected result: interaction-weighting repairs contaminated *comparisons* between cohorts, but the contamination here is inside every cohort's own pre-period, so no reweighting across cohorts can remove it. The same logic applies to @callaway2021difference and @borusyak2024revisiting: all of them take $G_i$ as given.

The Sun--Abraham pre-period coefficients are, however, informative in a way the TWFE ones are not. They are small but strongly significant and negative ($-0.14$ and $-0.19$ at $k = -3$ and $k = -2$), because this estimator compares each cohort against later-treated and never-treated units whose pre-period depression is differently timed, so the selection artifact no longer differences out. Running both estimators and noticing that the pre-trend verdict flips between them is a useful warning sign, even though neither delivers an unbiased estimate. It also creates a trap: the natural response to a significant negative pre-trend is to add unit-specific linear trends, which fits the artificial pre-period dip and extrapolates it forward, inflating the post-treatment estimates further. The artifact is a level shift plus a step, not a trend, so trend adjustment can only make matters worse. Clustering, likewise, addresses the standard error and not the estimate; here the point estimate itself is the artifact, so tighter or looser standard errors change nothing.

#### Replication

@renault2026scientific work with a reconstruction of the arXiv submission histories underlying @kusumegi2025scientific. That reconstruction is not distributed with the comment and is not part of this book's datasets, so the published empirical result is not reproduced here. The simulation above is instead a self-contained replication of the paper's central claim — that first-detection timing alone generates the reported dynamics — under a DGP transparent enough to admit a closed-form solution.

#### Practical Guidance

The question to ask of any staggered design is not whether the pre-trends are flat but **where the treatment date came from**. If the answer involves running a classifier, keyword search, or threshold rule over a stream that is the outcome or is mechanically tied to it, the design is suspect regardless of what the event-study plot looks like. Three checks are worth running.

*Neutral-trigger placebo.* Re-run the entire pipeline with a detector that fires on something irrelevant — a common word, a random subset of output, a fixed per-unit hazard — while preserving the property that firing probability rises with output. Under a genuine effect the placebo returns nothing; under an outcome-triggered artifact it reproduces the original dynamics. This is the sharpest available diagnostic and it is cheap.

*Pre-period placebo.* Re-run the design on a window ending before the treatment could possibly have existed. @renault2026scientific use a pre-ChatGPT window; the analogue in most settings is a window before the technology, policy, or program was available.

*Independent dating.* The real fix is to date adoption from a source that does not pass through the outcome — an administrative record, a procurement or licensing database, a survey, a disclosure filing. When no such source exists, an alternative is to define $G_i$ from a *rate* rather than a count (the first period in which the flagged share exceeds a threshold), which weakens but does not eliminate the dependence on output, or to condition the analysis on units whose output is stable enough that detection probability is roughly constant.

If none of these is available, the honest response is to report the design as descriptive rather than causal. Note also that the artifact scales with $\lambda q$: it is worst in settings with high output volume and a sensitive detector, and it shrinks as detection approaches certainty ($q \to 1$, where the first period is always flagged) or vanishing rarity.

#### Pitfalls

- **Treating a flat pre-trend as design validation.** The Proposition shows pre-period coefficients are flat *in expectation* under the artifact. A passing pre-trend test is the expected outcome of a broken design here, not evidence against one.
- **Dropping the on-impact coefficient.** Because $\beta_0$ is visibly the largest artifact, a tempting response is to drop $k = 0$ as "contaminated" and read the plateau. But the plateau is an artifact too, of size $\lambda q$, and it is the one that gets interpreted as the long-run effect.
- **Assuming the problem is heterogeneity.** Modern staggered estimators leave the artifact intact, and Sun--Abraham makes it larger in the simulation above.
- **Assuming a rare detector is safe.** Lowering $q$ shrinks the plateau proportionally but also pushes adoption dates later and concentrates them on ever more extreme output months, so the on-impact spike does not disappear.
- **Reading the mirror image as a negative effect.** If the detector fires on *low* output (a first missed deadline, a first failed inspection), every sign flips and the same mechanism manufactures a spurious improvement after treatment.

------------------------------------------------------------------------

### Matching Methods in DID

Matching methods are often used in **causal inference** to balance treated and control units based on **pre-treatment observables**. In the context of Difference-in-Differences, matching helps:

-   Reduce selection bias by ensuring that treated and control units are comparable before treatment.
-   Improve parallel trends validity by selecting control units with similar pre-treatment trajectories.
-   Enhance robustness when treatment assignment is non-random across groups.

**Key Considerations in Matching**

-   **Standard Errors Need Adjustment**
    -   Standard errors should account for the fact that matching reduces variance [@heckman1997matching].
    -   A more robust alternative is Doubly Robust DID [@sant2020doubly], where either matching or regression suffices for unbiased treatment effect identification.
-   **Group Fixed Effects Alone Do Not Eliminate Selection Bias**
    -   Fixed effects absorb time-invariant heterogeneity, but do not correct for selection into treatment.
    -   Matching helps close the "backdoor path" between:
        1.  Propensity to be treated
        2.  Dynamics of outcome evolution post-treatment
-   **Matching on Time-Varying Covariates**
    -   Beware of regression to the mean: extreme pre-treatment outcomes may artificially bias post-treatment estimates [@daw2018matching].
    -   This issue is less concerning for time-invariant covariates.
-   **Comparing Matching vs. DID Performance**
    -   Matching and DID both use pre-treatment outcomes to mitigate selection bias.
    -   Simulations [@chabe2015analysis] show that:
        -   Matching tends to underestimate the true treatment effect but improves with more pre-treatment periods.
        -   When selection bias is symmetric, Symmetric DID (equal pre- and post-treatment periods) performs well.
        -   When selection bias is asymmetric, DID generally outperforms Matching.
-   **Forward DID as a Control Unit Selection Algorithm**
    -   An efficient way to select control units is Forward DID [@li2024frontiers].

------------------------------------------------------------------------

### Control Variables in DID

-   Always report results with and without controls:
    -   If controls are fixed within groups or time periods, they should be absorbed in fixed effects.
    -   If controls vary across both groups and time, this suggests the parallel trends assumption is questionable.
-   $R^2$ is not crucial in causal inference:
    -   Unlike predictive models, causal models do not prioritize explanatory power ($R^2$), but rather unbiased identification of treatment effects.

------------------------------------------------------------------------

### DID for Count Data: Fixed-Effects Poisson Model

For count data, one can use the fixed-effects Poisson pseudo-maximum likelihood estimator (PPML) [@athey2006identification; @puhani2012treatment]. Applications of this method can be found in management [@burtch2018can] and marketing [@he2021end].

This approach offers robust standard errors under over-dispersion [@wooldridge1999quasi] and is particularly useful when dealing with excess zeros in the data.

Key advantages of PPML:

-   Handles zero-inflated data better than log-OLS: A log-OLS regression may produce biased estimates [@o2010not] when heteroskedasticity is present [@silva2006log], especially in datasets with many zeros [@silva2011further].
-   Avoids the limitations of negative binomial fixed effects: Unlike Poisson, there is no widely accepted fixed-effects estimator for the negative binomial model [@allison20027].

------------------------------------------------------------------------

### Handling Zero-Valued Outcomes in DID

Zero-valued outcomes are a special case of the broader functional-form sensitivity that runs through this chapter. In the canonical DiD model, parallel trends is a property of the *outcome scale*, so transforming the outcome (logs, percent changes, normalized counts) silently changes the identifying assumption being made. When a non-trivial fraction of the data is at zero, the choice of scale also collides with the distinction between intensive- and extensive-margin responses. The two approaches below take different stances on that tradeoff, and they should be read alongside the [changes-in-changes](#sec-changes-in-changes) estimator, which provides a fully nonparametric alternative when the analyst is unwilling to commit to either parallel trends in levels or parallel trends in ratios.

When dealing with **zero-valued outcomes**, it is crucial to separate the **intensive margin effect** (e.g., outcome changes from 10 to 11) from the **extensive margin effect** (e.g., outcome changes from 0 to 1).

A common issue is that the treatment coefficient from a **log-transformed regression** cannot be directly interpreted as a percentage change when zeros are present [@chen2024logs]. To address this, we can consider two alternative approaches:

1.  **Proportional Treatment Effects**

We define the percentage change in the treated group's post-treatment outcome as:

$$
\theta_{ATT\%} = \frac{E[Y_{it}(1) \mid D_i = 1, Post_t = 1] - E[Y_{it}(0) \mid D_i = 1, Post_t = 1]}{E[Y_{it}(0) \mid D_i = 1, Post_t = 1]}
$$

Instead of assuming parallel trends in levels, we can rely on a parallel trends assumption in ratios [@wooldridge2023simple].

The Poisson QMLE model is:

$$
Y_{it} = \exp(\beta_0 + \beta_1 D_i \times Post_t + \beta_2 D_i + \beta_3 Post_t + X_{it}) \epsilon_{it}
$$

The treatment effect is estimated as:

$$
\hat{\theta}_{ATT\%} = \exp(\hat{\beta}_1) - 1
$$

To validate the parallel trends in ratios assumption, we can estimate a dynamic Poisson QMLE model:

$$
Y_{it} = \exp(\lambda_t + \beta_2 D_i + \sum_{r \neq -1} \beta_r D_i \times (RelativeTime_t = r))
$$

If the assumption holds, we expect:

$$
\exp(\hat{\beta}_r) - 1 = 0 \quad \text{for} \quad r < 0.
$$

Even if the pre-treatment estimates appear close to zero, we should still conduct a sensitivity analysis [@rambachan2023more] to assess robustness (see [Prior Parallel Trends Test](#prior-parallel-trends-test)).

```{r}
library(fixest)

base_did_log0 <- base_did |> 
    mutate(y = if_else(y > 0, y, 0))

res_pois_es <-
    fepois(y ~ x1 + i(period, treat, 5) | id + period,
           data = base_did_log0,
           vcov = "hetero")

etable(res_pois_es)
```

Figure \@ref(fig:fig-poisson-event-study) shows the parallel trends before treatment and the effect after treatment.

```{r fig-poisson-event-study, fig.cap='Event-study estimates from a Poisson regression model.', fig.alt='Event study plot showing treatment effect coefficients over time from a Poisson regression model with period 5 as the reference. Points represent coefficient estimates with confidence intervals, allowing assessment of pre-treatment parallel trends and post-treatment effects.'}
iplot(res_pois_es)
```

This parallel trend is the "ratio" version as in @wooldridge2023simple :

$$
\frac{E(Y_{it}(0) |D_i = 1, Post_t = 1)}{E(Y_{it}(0) |D_i = 1, Post_t = 0)} = \frac{E(Y_{it}(0) |D_i = 0, Post_t = 1)}{E(Y_{it}(0) |D_i =0, Post_t = 0)}
$$

which means without treatment, the average percentage change in the mean outcome for treated group is identical to that of the control group.

2.  **Log Effects with Calibrated Extensive-Margin Value**

A potential limitation of proportional treatment effects is that they may not be well-suited for heavy-tailed outcomes. In such cases, we may prefer to explicitly model the extensive margin effect.

Following [@chen2024logs, p. 39], we can calibrate the weight placed on the intensive vs. extensive margin to ensure meaningful interpretation of the treatment effect.

If we want to study the treatment effect on a concave transformation of the outcome that is less influenced by those in the distribution's tail, then we can perform this analysis.

Steps:

1.  Normalize the outcomes such that 1 represents the minimum non-zero and positive value (i.e., divide the outcome by its minimum non-zero and positive value).
2.  Estimate the treatment effects for the new outcome

$$
m(y) =
\begin{cases}
\log(y) & \text{for } y >0 \\
-x & \text{for } y = 0
\end{cases}
$$

The choice of $x$ depends on what the researcher is interested in:

| Value of $x$ | Interest                                                                                                                                                                        |
|-------------------|-----------------------------------------------------|
| $x = 0$      | The treatment effect in logs where all zero-valued outcomes are set to equal the minimum non-zero value (i.e., we exclude the extensive-margin change between 0 and $y_{min}$ ) |
| $x>0$        | Setting the change between 0 and $y_{min}$ to be valued as the equivalent of a $x$ log point change along the intensive margin.                                                 |

```{r}
library(fixest)
base_did_log0_cali <- base_did_log0 |> 
    # get min 
    mutate(min_y = min(y[y > 0])) |> 
    
    # normalized the outcome 
    mutate(y_norm = y / min_y)

my_regression <-
    function(x) {
        base_did_log0_cali <-
            base_did_log0_cali %>% mutate(my = ifelse(y_norm == 0,-x,
                                                      log(y_norm)))
        my_reg <-
            feols(
                fml = my ~ x1 + i(period, treat, 5) | id + period,
                data = base_did_log0_cali,
                vcov = "hetero"
            )
        
        return(my_reg)
    }

xvec <- c(0, .1, .5, 1, 3)
reg_list <- purrr::map(.x = xvec, .f = my_regression)


iplot(reg_list, 
      pt.col =  1:length(xvec),
      pt.pch = 1:length(xvec))
legend("topleft", 
       col = 1:length(xvec),
       pch = 1:length(xvec),
       legend = as.character(xvec))


etable(
    reg_list,
    headers = list("Extensive-margin value (x)" = as.character(xvec)),
    digits = 2,
    digits.stats = 2
)
```

We have the dynamic treatment effects for different hypothesized extensive-margin value of $x \in (0, .1, .5, 1, 3, 5)$

The first column is when the zero-valued outcome equal to $y_{min, y>0}$ (i.e., there is no difference between the minimum outcome and zero outcome - $x = 0$)

For this particular example, as the extensive margin increases, we see an increase in the effect magnitude. The second column is when we assume an extensive-margin change from 0 to $y_{min, y >0}$ is equivalent to a 10 (i.e., $0.1 \times 100$) log point change along the intensive margin.

------------------------------------------------------------------------

### Standard Errors

One of the major statistical challenges in DiD estimation is serial correlation in the error terms. This issue is particularly problematic because it can lead to underestimated standard errors, inflating the likelihood of Type I errors (false positives). As discussed in @bertrand2004much, serial correlation arises in DiD settings due to several factors:

1.  Long time series: Many DiD studies use multiple time periods, increasing the risk of correlated errors.
2.  Highly positively serially correlated outcomes: Many economic and business variables (e.g., GDP, sales, employment rates) exhibit strong persistence over time.
3.  Minimal within-group variation in treatment timing: For example, in a state-level policy change, all individuals in a state receive treatment at the same time, leading to correlation within state-time clusters.

To correct for serial correlation, various methods can be employed. However, some approaches work better than others:

-   Avoid standard parametric corrections: A common approach is to model the error term using an autoregressive (AR) process. However, @bertrand2004much show that this often fails in DiD settings because it does not fully account for within-group correlation.
-   Nonparametric solutions (preferred when the number of groups is large): 
    -   Block bootstrap: Resampling entire groups (e.g., states) rather than individual observations maintains the correlation structure and provides robust standard errors.
-   Collapsing data into two periods (Pre vs. Post): 
    -   Aggregating the data into a single pre-treatment and single post-treatment period can mitigate serial correlation issues. This approach is particularly useful when the number of groups is small [@donald2007inference].
    -   Note: While this reduces the power of the analysis by discarding variation across time, it ensures that standard errors are not artificially deflated.
-   Variance-covariance matrix corrections: 
    -   Empirical corrections (e.g., cluster-robust standard errors) and arbitrary variance-covariance matrix adjustments (e.g., Newey-West) can work well, but they are reliable only in large samples.

Overall, selecting the appropriate correction method depends on the sample size and structure of the data. When possible, block bootstrapping and collapsing data into pre/post periods are among the most effective approaches. The serial-correlation problem is also one of the reasons that modern staggered-adoption estimators (Section \@ref(sec-modern-estimators-for-staggered-adoption)) report cluster-bootstrap standard errors by default, and it is closely connected to the inference issues flagged in the [SUTVA](#sec-sutva) discussion when treatment exposure is correlated within clusters.

------------------------------------------------------------------------

### Partial Identification {#sec-partial-identification-did}

Classical DiD delivers a **point-identified** estimate of the [average treatment effect on the treated](#sec-average-treatment-effect-on-the-treated) (ATT) only when the [parallel-trends assumption](#prior-parallel-trends-test) is exact. Yet in practice pre-trends often deviate or post-treatment shocks differ across groups (e.g., the famous @ashenfelter1978estimating). Partial-identification (PI) methods shift the question from *"Is the estimate unbiased?"* to *"How much would the identifying assumption have to fail before I reverse my conclusion?"*

In the diagnostic workflow sketched at the start of this chapter, partial identification is the natural last line of defense once the visual checks, event-study tests, and [placebo tests](#sec-placebo-test-did) have all left some residual doubt. It is also the right tool when pre-trend tests are statistically silent but underpowered: rather than treating "we failed to reject" as proof of parallel trends, the analyst reports the magnitude of violation that the data would still permit. Used this way, PI complements the heterogeneity-robust estimators in Section \@ref(sec-modern-estimators-for-staggered-adoption) and the [coefficient stability bounds](#sec-coefficient-stability-bounds) machinery: the former handles bias from contaminated comparisons, while PI handles bias from a violated identifying assumption itself.

The partial-identification literature that has grown up around DiD reflects a series of distinct philosophical bets about how to make the "how much could parallel trends fail?" question precise. It is helpful to walk through them in order, because each one chooses a different lever to constrain.

The starting point is @manski2018right, who simply ask the analyst to commit to *any* substantively credible bounds on $\delta$, across time, across space, or both, and then propagate those bounds through to the ATT. The approach is conceptually clean but requires the researcher to defend the bounds from outside the data. @keele2019patterns sharpen the question into something more empirically tractable by importing Rosenbaum-style sensitivity analysis from the matching literature: rather than bounding $\delta$ directly, fix a single sensitivity parameter that measures how much hidden bias one would need to reverse the estimated sign. @keele2019bracketing extend this machinery to compound treatments, where the treated group is simultaneously exposed to another event, asking whether the causal claim survives under that more demanding scenario.

A different strategy is to lean on auxiliary control groups rather than auxiliary parameters. @ye2024negative use two control groups exposed to *negatively correlated* shocks to non-parametrically bracket the counterfactual trend, sidestepping parallel trends without committing to a specific bound on $\delta$. @leavitt2020beyond goes Bayesian: a hierarchical prior shrinks extreme violations of parallel trends and yields credible intervals that account for trend heterogeneity directly.

The most influential synthesis is @rambachan2023more, which collapses the question into a researcher-chosen smoothness or magnitude restriction and then reports the *breakdown threshold*, the value of the restriction at which the confidence set first touches zero. Two bounding strategies fall out of this framework, both implemented in the `honestdid` package and discussed in detail later in this section: [relative-magnitude bounds](#sec-relative-magnitude-restriction-honestdid), which constrain $|\delta|$ to be no larger than $M$ times the worst pre-period jump, and [smoothness bounds](#sec-smoothness-restriction-honestdid), which allow the slope of $\delta$ to shift by at most $M$ per period.

------------------------------------------------------------------------

#### Canonical DiD notation and the source of bias

Let

-   $Y_{it}^{1}$ and $Y_{it}^{0}$ be potential outcomes for unit $i$ in period $t$ under treatment and no treatment;
-   $D_i\in{0,1}$ indicate ever-treated units;
-   $\texttt{Post}_t$ mark periods after the policy shock.

For two periods (pre, post) and two groups (T = treated, U = untreated) the simple DiD estimator is

$$
\hat\beta_{DD}= \left[E(Y_T^{1}\mid \texttt{Post})-E(Y_T^{0}\mid \texttt{Pre})\right]-\left[E(Y_U^{0}\mid \texttt{Post})-E(Y_U^{0}\mid \texttt{Pre})\right]
$$

Add and subtract the **counterfactual trend** $E(Y_T^{0}\mid \texttt{Post})$ and rearrange:

$$
\begin{aligned}
\hat\beta_{DD} &= \underbrace{E(Y_T^{1}\mid\texttt{Post})-E(Y_T^{0}\mid\texttt{Post})}_{\text{ATT}} \\
&+
\underbrace{[E(Y_T^{0}\mid\texttt{Post})-E(Y_T^{0}\mid\texttt{Pre})]
-[E(Y_U^{0}\mid\texttt{Post})-E(Y_U^{0}\mid\texttt{Pre})]}_{\delta}
\end{aligned}
$$

Hence

$$
\hat\beta_{DD}= \text{ATT}+\delta .
$$

When [parallel trends](#prior-parallel-trends-test) hold, $\delta=0$; any violation translates **one-for-one** into bias [@frake2025partial].

------------------------------------------------------------------------

#### Bounding the ATT when $\delta\neq 0$

The core PI idea is simple: *derive credible bounds on* $\delta$, then translate those into bounds on the ATT.

Two families of restrictions, both operationalized by @rambachan2023more and implemented in the `honestdid` package, are now widely used:

1.  [Relative-magnitude (RM) bounds](#sec-relative-magnitude-restriction-honestdid): the post-treatment gap $|\delta|$ cannot exceed $M$ times the largest *period-to-period* deviation observed in the *pre-treatment* event-study.
2.  [Smoothness (SM) bounds](#sec-smoothness-restriction-honestdid): the *slope* of the treated-minus-control gap can change by at most $M$ units each period after treatment.

Because $M$ is researcher-chosen, inference is reported across a grid of plausible $M$ values, revealing the *breakdown* threshold where the confidence set first touches zero [@frake2025partial, p. 14].

------------------------------------------------------------------------

#### Relative-Magnitude Restriction {#sec-relative-magnitude-restriction-honestdid}

*Let* $\hat g_t$ be the event-study coefficient at time $t$ (normalized to zero at $t=-1$).

-   Step 1: calibrate the worst pre-trend: \
    $M_0 = \max_{s<0} |\hat g_s-\hat g_{s-1}|$.

-   Step 2: allow proportional violations post-treatment: \
    Impose $|\delta_t|\le M\cdot M_0$ for every $t\ge 0$.

-   **Step 3: construct "honest" confidence sets** for the ATT or for period-specific effects by solving a constrained least-squares problem [@rambachan2023more].

------------------------------------------------------------------------

#### Smoothness Restriction {#sec-smoothness-restriction-honestdid}

When violations evolve gradually (e.g., treated units drift upward over time), the SM bound is more credible.

-   Step 1: estimate the pre-treatment slope: \
    $s_0 = \hat g_{-1}-\hat g_{-2}$.

-   Step 2: restrict post-treatment slopes: \
    The counterfactual slope in period $k$ may vary within\
    $[s_0-M, s_0+M]$.

-   **Step 3: propagate forward** to obtain an admissible path for $\delta_t$; compute the least-favorable path, then derive confidence sets [@rambachan2023more].

------------------------------------------------------------------------

#### Step-by-step Empirical Workflow

See @frake2025partial, p. 15 for a decision tree that guides the choice of method.

1.  **Graph the event-study.** Are any pre-trends visible? Quantify their magnitude and/or slope.
2.  **Choose a restriction family** (RM for episodic shocks, SM for trending deviations). Justify the choice theoretically (e.g., anticipation effects vs. gradual selection).
3.  **Select a grid of** $M$. Always display the break-even $M^{*}$ where the CI first covers zero; interpret $M^{*}$ in real-world units.
4.  **Report the full set, not only the "robust" interval.** Transparency trumps opportunistic suppression.
5.  **Complement with falsification tests** (placebo policy dates, placebo outcomes) to discipline the plausible range of $M$.
6.  **Document software and code.** `honestdid` (R/Stata) includes convenient wrappers for both restrictions and automatic figures.

------------------------------------------------------------------------

#### Worked Example (Stylized)

Suppose pre-policy trends exhibit at most a 0.8-point swing. Setting $M\in{0,0.5,1,1.5}$:

| $M$ | 95 % CI for ATT | Conclusion           |
|-----|-----------------|----------------------|
| 0   | [0.85, 1.25]    | Positive, robust     |
| 0.5 | [0.60, 1.50]    | Positive             |
| 1   | [-0.05, 1.85]   | Sign flips plausible |
| 1.5 | [-0.45, 2.25]   | Cannot rule out zero |

Thus the causal claim survives unless the unobserved counterfactual deviations are $\ge 1 \times$ the worst pre-treatment kink, an interpretation far clearer than a single $p$-value.

------------------------------------------------------------------------

#### Common Pitfalls and Best Practices

-   **Mistaking absence of significant pre-trends for proof of parallel-trends.** Under-powered tests can miss economically large differences.
-   **Choosing** $M$ post-hoc. Decide the grid *before* looking at results.
-   **Ignoring sign.** If theory predicts the treated counterfactual would have **fallen**, impose a one-sided bound; doing otherwise wastes power.
-   **Forgetting heterogeneous timing or staggered adoption.** Apply PI to group-time ATT estimates from modern DiD estimators before aggregation.

Partial-identification transforms DiD from a "take-it-or-leave-it" enterprise into a transparent sensitivity analysis. By explicitly parameterizing and visually communicating, the extent of permissible departures from parallel trends, researchers allow readers to map their own priors onto empirical conclusions. Used thoughtfully, PI techniques enhance credibility without demanding the impossible.

------------------------------------------------------------------------

#### Putting the Pieces Together

A defensible modern DiD analysis rarely relies on a single diagnostic; it layers several. A reasonable template, drawing on the tools surveyed in this chapter and on the broader [quasi-experimental](#sec-quasi-experimental) toolkit, looks like this. Begin with raw outcome plots and an event-study (the [visualization](#sec-visualization-did) and [event studies](#sec-event-studies) sections give the mechanics) to inspect both levels and trends. If the data are staggered, run the [Goodman-Bacon decomposition](#sec-goodman-bacon-decomposition) and re-estimate using a heterogeneity-robust estimator from Section \@ref(sec-modern-estimators-for-staggered-adoption) before drawing any substantive conclusion from a TWFE coefficient. Layer in [placebo tests](#sec-placebo-test-did) (group-based, time-based, and randomization-based) and the robustness checks above to discipline modeling choices. Where parallel trends in levels is suspect, consider [matching](#sec-matching-methods), [synthetic control](#sec-synthetic-control), [synthetic DiD](#sec-synthetic-difference-in-differences), or [changes-in-changes](#sec-changes-in-changes) as design-level alternatives. Finally, use partial identification, with both [relative-magnitude](#sec-relative-magnitude-restriction-honestdid) and [smoothness](#sec-smoothness-restriction-honestdid) restrictions, to communicate how much violation of the identifying assumption the conclusion can absorb. The point of the exercise is not to mechanically tick boxes but to give a reader a concrete sense of which failure modes have been ruled out, which remain plausible, and how much they would have to bite before the headline result reverses. A DiD analysis that survives that gauntlet is rarely "proven", but it is unusually difficult to dismiss.
