# Sensitivity Analysis/ Robustness Check

## Specification curve

-   also known as Specification robustness graph or coefficient stability plot

Resources

-   [In Stata](https://github.com/hhsievertsen/speccurve) or [speccurve](https://github.com/martin-andresen/speccurve)

-   [@simonsohn2020]

### starbility

-   Recommend

Installation

```{r, eval = FALSE}
devtools::install_github('https://github.com/AakaashRao/starbility')
library(starbility)
```

Example by the [package's author](https://htmlpreview.github.io/?https://github.com/AakaashRao/starbility/blob/master/doc/starbility.html)

```{r}
library(tidyverse)
library(starbility)
library(lfe)
data("diamonds")
set.seed(43)
indices = sample(1:nrow(diamonds),
                 replace = F,
                 size = round(nrow(diamonds) / 20))
diamonds = diamonds[indices, ]
```

Plot different combinations of controls

```{r}

# If you want to make the diamond dimensions as base control
base_controls = c(
  'Diamond dimensions' = 'x + y + z' # include all variables under 1 dimension
)


perm_controls = c(
  'Depth' = 'depth',
  'Table width' = 'table'
)

nonperm_fe_controls = c(
  'Clarity FE (granular)' = 'clarity',
  'Clarity FE (binary)' = 'high_clarity'
)

# Adding fixed effects
nonperm_fe_controls = c(
  'Clarity FE (granular)' = 'clarity',
  'Clarity FE (binary)' = 'high_clarity'
)

# Adding instrumental variables 
instruments = 'x+y+z'

# clustering and weights 
diamonds$sample_weights = runif(n = nrow(diamonds))


# robust standard errors 
starb_felm_custom = function(spec, data, rhs, ...) {
  spec = as.formula(spec)
  model = lfe::felm(spec, data=data) %>% broom::tidy()

  row = which(model$term==rhs)
  coef = model[row, 'estimate'] %>% as.numeric()
  se   = model[row, 'std.error'] %>% as.numeric()
  p    = model[row, 'p.value'] %>% as.numeric()
  
  # 99% confidence interval
  z = qnorm(0.995) 
  # one-tailed test
  return(c(coef, p/2, coef+z*se, coef-z*se))
}

plots = stability_plot(
    data = diamonds,
    lhs = 'price',
    rhs = 'carat',
    error_geom = 'ribbon', # make the plot more aesthetics
    # error_geom = 'none', # if you don't want ribbon (i.e., error bar)
    model = starb_felm_custom,
    cluster = 'cut',
    weights = 'sample_weights',
    # iv = instruments,
    perm = perm_controls,
    base = base_controls,
    # perm_fe = perm_fe_controls,
    # nonperm_fe = nonperm_fe_controls, # if you want to include fixed effects sequentially (not all combinations) (e.g., you want to test country or state fixed effect, not both )
    # fe_always = F,  # if you want to have a model without any Fixed Effects
    sort = "asc-by-fe", # sort "asc", "desc", or by fixed effects: "asc-by-fe" or "desc-by-fe"
    
    # if you have less variables and want more aesthetics 
    # control_geom = 'circle',
    # point_size = 2,
    # control_spacing = 0.3,
    
    
    # error_alpha = 0.2, # change alpha of the error geom
    # point_size = 1.5, # change the size of the coefficient points
    # control_text_size = 10, # change the size of the control labels
    # coef_ylim = c(-5000, 35000), # change the endpoints of the y-axis
    # trip_top = 3, # change the spacing between the two panels
    
    rel_height = 0.6
)
plots

# add comments
# replacement_coef_panel = plots[[1]] + 
#   scale_y_reverse() +
#   theme(panel.grid.minor = element_blank()) +
#   geom_vline(xintercept = 41, linetype = 'dashed', alpha = 0.4) +
#   annotate(geom = 'label', x = 52, y = 30000, label = 'What a great\nspecification!', alpha = 0.75)
# 
# combine_plots(replacement_coef_panel, 
#               plots[[2]], 
#               rel_height = 0.6)



```

Note:

-   $p < 0.01$: red
-   $p < 0.05$: green
-   $p < 0.1$: blue
-   $p > 0.1$: black

[More Advanced Stuff](https://htmlpreview.github.io/?https://github.com/AakaashRao/starbility/blob/master/doc/starbility-advanced.html)

```{r}
# Step 1: Control Grid

diamonds$high_clarity = diamonds$clarity %in% c('VS1','VVS2','VVS1','IF')

base_controls = c(
  'Diamond dimensions' = 'x + y + z'
)

perm_controls = c(
  'Depth' = 'depth',
  'Table width' = 'table'
)

perm_fe_controls = c(
  'Cut FE' = 'cut',
  'Color FE' = 'color'
)
nonperm_fe_controls = c(
  'Clarity FE (granular)' = 'clarity',
  'Clarity FE (binary)' = 'high_clarity'
)

grid1 = stability_plot(data = diamonds, 
                      lhs = 'price', 
                      rhs = 'carat', 
                      perm = perm_controls,
                      base = base_controls, 
                      perm_fe = perm_fe_controls, 
                      nonperm_fe = nonperm_fe_controls, 
                      run_to=2)

knitr::kable(grid1 %>% head(10))

# Step 2: Get model expression

grid2 = stability_plot(grid = grid1,
                      data=diamonds, 
                      lhs='price', 
                      rhs='carat', 
                      perm=perm_controls, 
                      base=base_controls,
                      run_from=2,
                      run_to=3)


knitr::kable(grid2 %>% head(10))

# Step 3: Estimate models
grid3 = stability_plot(grid = grid2,
                      data=diamonds, 
                      lhs='price', 
                      rhs='carat', 
                      perm=perm_controls, 
                      base=base_controls,
                      run_from=3,
                      run_to=4)

knitr::kable(grid3 %>% head(10))

# Step 4: Get dataframe to draw
dfs = stability_plot(grid = grid3,
                      data=diamonds, 
                      lhs='price', 
                      rhs='carat', 
                      perm=perm_controls, 
                      base=base_controls,
                      run_from=4,
                      run_to=5)

coef_grid = dfs[[1]]
control_grid = dfs[[2]]

knitr::kable(coef_grid %>% head(10))

# Step 5: plot the sensitivity graph 
panels = stability_plot(data = diamonds, 
                      lhs='price', 
                      rhs='carat', 
                      coef_grid = coef_grid,
                      control_grid = control_grid,
                      run_from=5,
                      run_to=6)

stability_plot(data = diamonds,
               lhs='price', 
               rhs='carat', 
               coef_panel = panels[[1]],
               control_panel = panels[[2]],
               run_from = 6,
               run_to = 7)

```

In step 2, we can modify to use other function (e.g., `glm`)

```{r}
diamonds$above_med_price = as.numeric(diamonds$price > median(diamonds$price))

base_controls = c('Diamond dimensions' = 'x + y + z')

perm_controls = c('Depth' = 'depth',
                  'Table width' = 'table',
                  'Clarity' = 'clarity')
lhs_var = 'above_med_price'
rhs_var = 'carat'

grid1 = stability_plot(
    data = diamonds,
    lhs = lhs_var,
    rhs = rhs_var,
    perm = perm_controls,
    base = base_controls,
    fe_always = F,
    run_to = 2
)

# Create control part of formula
base_perm = c(base_controls, perm_controls)
grid1$expr = apply(grid1[, 1:length(base_perm)], 1,
                   function(x)
                       paste(base_perm[names(base_perm)[which(x == 1)]], collapse = '+'))

# Complete formula with LHS and RHS variables
grid1$expr = paste(lhs_var, '~', rhs_var, '+', grid1$expr, sep = '')

knitr::kable(grid1 %>% head(10))

# customer function for the logit model
starb_logit = function(spec, data, rhs, ...) {
  spec = as.formula(spec)
  model = glm(spec, data=data, family='binomial', weights=data$weight) %>%
    broom::tidy()
  row = which(model$term==rhs)
  coef = model[row, 'estimate'] %>% as.numeric()
  se   = model[row, 'std.error'] %>% as.numeric()
  p    = model[row, 'p.value'] %>% as.numeric()

  return(c(coef, p, coef+1.96*se, coef-1.96*se))
}

stability_plot(grid = grid1,
               data = diamonds, 
               lhs = lhs_var, 
               rhs = rhs_var,
               model = starb_logit,
               perm = perm_controls,
               base = base_controls,
               fe_always = F,
               run_from=3)

```

For getting other specification (e.g., different CI)

```{r}
library(margins)
starb_logit_enhanced = function(spec, data, rhs, ...) {
  # Unpack ...
  l = list(...)
  get_mfx = ifelse(is.null(l$get_mfx), F, T) # Set a default to F
  
  spec = as.formula(spec)
  if (get_mfx) {
    model = glm(spec, data=data, family='binomial', weights=data$weight) %>%
      margins() %>%
      summary
    row = which(model$factor==rhs)
    coef = model[row, 'AME'] %>% as.numeric()
    se   = model[row, 'SE'] %>% as.numeric()
    p    = model[row, 'p'] %>% as.numeric()
  } else {
    model = glm(spec, data=data, family='binomial', weights=data$weight) %>%
      broom::tidy()
    row = which(model$term==rhs)
    coef = model[row, 'estimate'] %>% as.numeric()
    se   = model[row, 'std.error'] %>% as.numeric()
    p    = model[row, 'p.value'] %>% as.numeric()
  }

  z = qnorm(0.995)
  return(c(coef, p, coef+z*se, coef-z*se))
}

stability_plot(grid = grid1,
               data = diamonds, 
               lhs = lhs_var, 
               rhs = rhs_var,
               model = starb_logit_enhanced,
               get_mfx = T,
               perm = perm_controls,
               base = base_controls,
               fe_always = F,
               run_from = 3)
```

To get your customized plot

```{r}
dfs = stability_plot(grid = grid1,
               data = diamonds, 
               lhs = lhs_var, 
               rhs = rhs_var,
               model = starb_logit_enhanced,
               get_mfx = T,
               perm = perm_controls,
               base = base_controls,
               fe_always = F,
               run_from = 3,
               run_to = 5)

coef_grid_logit = dfs[[1]]
control_grid_logit = dfs[[2]]

min_space = 0.5

coef_plot = ggplot2::ggplot(coef_grid_logit, aes(x = model, y = coef, shape=p, group=p)) +
  geom_linerange(aes(ymin = error_low, ymax = error_high), alpha=0.75) +
  geom_point(size=5, aes(col=p, fill=p), alpha=1) +
  viridis::scale_color_viridis(discrete = TRUE, option = "D")+
  scale_shape_manual(values = c(15,17,18, 19)) +
  theme_classic() +
  geom_hline(yintercept=0, linetype='dotted') +
  ggtitle('A custom coefficient stability plot!') +
  labs(subtitle="Error bars represent 99% confidence intervals") +
  theme(axis.text.x = element_blank(),
        axis.title = element_blank(),
        axis.ticks.x = element_blank()) +
  coord_cartesian(xlim=c(1-min_space, max(coef_grid_logit$model)+min_space), 
                  ylim=c(-0.1, 1.6)) +
  guides(fill=F, shape=F, col=F)


control_plot = ggplot(control_grid_logit) +
  geom_point(aes(x = model, y = y, fill=value), shape=23, size=4) +
  scale_fill_manual(values=c('#FFFFFF', '#000000')) +
  guides(fill=F) +
  scale_y_continuous(breaks = unique(control_grid_logit$y), 
                     labels = unique(control_grid_logit$key),
                     limits=c(min(control_grid_logit$y)-1, max(control_grid_logit$y)+1)) +
  scale_x_continuous(breaks=c(1:max(control_grid_logit$model))) +
  coord_cartesian(xlim=c(1-min_space, max(control_grid_logit$model)+min_space)) +
  theme_classic() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.title = element_blank(),
        axis.text.y = element_text(size=10),
        axis.ticks = element_blank(),
        axis.line = element_blank()) 

cowplot::plot_grid(coef_plot, control_plot, rel_heights=c(1,0.5), 
                   align='v', ncol=1, axis='b')
```

To get different model specification (e.g., probit vs. logit)

```{r}
starb_probit = function(spec, data, rhs, ...) {
    # Unpack ...
    l = list(...)
    get_mfx = ifelse(is.null(l$get_mfx), F, T) # Set a default to F
    
    spec = as.formula(spec)
    if (get_mfx) {
        model = glm(
            spec,
            data = data,
            family = binomial(link = 'probit'),
            weights = data$weight
        ) %>%
            margins() %>%
            summary
        row = which(model$factor == rhs)
        coef = model[row, 'AME'] %>% as.numeric()
        se   = model[row, 'SE'] %>% as.numeric()
        p    = model[row, 'p'] %>% as.numeric()
    } else {
        model = glm(
            spec,
            data = data,
            family = binomial(link = 'probit'),
            weights = data$weight
        ) %>%
            broom::tidy()
        row = which(model$term == rhs)
        coef = model[row, 'estimate'] %>% as.numeric()
        se   = model[row, 'std.error'] %>% as.numeric()
        p    = model[row, 'p.value'] %>% as.numeric()
    }
    
    z = qnorm(0.995)
    return(c(coef, p, coef + z * se, coef - z * se))
}

probit_dfs = stability_plot(
    grid = grid1,
    data = diamonds,
    lhs = lhs_var,
    rhs = rhs_var,
    model = starb_probit,
    get_mfx = T,
    perm = perm_controls,
    base = base_controls,
    fe_always = F,
    run_from = 3,
    run_to = 5
)

# We'll put the probit DFs on the left, so we need to adjust the model numbers accordingly
# so the probit and logit DFs don't plot on top of one another!
coef_grid_probit = probit_dfs[[1]] %>% 
    mutate(model = model + max(coef_grid_logit$model))

control_grid_probit = probit_dfs[[2]] %>% 
    mutate(model = model + max(control_grid_logit$model))

coef_grid    = bind_rows(coef_grid_logit, coef_grid_probit)
control_grid = bind_rows(control_grid_logit, control_grid_probit)

panels = stability_plot(
    coef_grid = coef_grid,
    control_grid = control_grid,
    data = diamonds,
    lhs = lhs_var,
    rhs = rhs_var,
    perm = perm_controls,
    base = base_controls,
    fe_always = F,
    run_from = 5,
    run_to = 6
)

coef_plot = panels[[1]] + geom_vline(xintercept = 8.5,
                                     linetype = 'dashed',
                                     alpha = 0.8) +
    annotate(
        geom = 'label',
        x = 4.25,
        y = 1.8,
        label = 'Logit models',
        size = 6,
        fill = '#D3D3D3',
        alpha = 0.7
    ) +
    annotate(
        geom = 'label',
        x = 12.75,
        y = 1.8,
        label = 'Probit models',
        size = 6,
        fill = '#D3D3D3',
        alpha = 0.7
    ) +
    coord_cartesian(ylim = c(-0.5, 1.9))

control_plot = panels[[2]] + geom_vline(xintercept = 8.5,
                                        linetype = 'dashed',
                                        alpha = 0.8)

cowplot::plot_grid(
    coef_plot,
    control_plot,
    rel_heights = c(1, 0.5),
    align = 'v',
    ncol = 1,
    axis = 'b'
)
```

### rdfanalysis

-   Not recommend

Installation

```{r, eval=FALSE}
devtools::install_github("joachim-gassen/rdfanalysis")
```

Example by the [package's author](https://joachim-gassen.github.io/rdfanalysis/)

```{r}
library(rdfanalysis)
load(url("https://joachim-gassen.github.io/data/rdf_ests.RData"))
plot_rdf_spec_curve(ests, "est", "lb", "ub") 
```

Shiny app for readers to explore

```{r, eval=FALSE}
design <- define_design(steps = c("read_data",
                                  "select_idvs",
                                  "treat_extreme_obs",
                                  "specify_model",
                                  "est_model"),
                        rel_dir = "vignettes/case_study_code")

shiny_rdf_spec_curve(ests, list("est", "lb", "ub"),
                     design, "vignettes/case_study_code",
                     "https://joachim-gassen.github.io/data/wb_new.csv")
```

## Coefficient stability

[@oster2017]

-   Coefficient stability can be evident against omitted variable bias.

-   But coeffiicent stability alone can be misleadning, but combing with $R^2$ movement, it can become informative.

Packages

-   `mplot`: graphical Model stability and Variable Selection

-   `robomit`: Robustness checks for omitted variable bias (implementation of

```{r}
library(robomit)

# estimate beta 
o_beta(
  y     = "mpg",       # dependent variable
  x     = "wt",        # independent treatment variable
  con   = "hp + qsec", # related control variables
  delta = 1,           # delta
  R2max = 0.9,         # maximum R-square
  type  = "lm",        # model type
  data  = mtcars       # dataset
) 

```


