### Binary Endogenous Regressors {#sec-binary-endogenous-regressors}

The [control function approach](#sec-control-function-approach) and [Heckman-style corrections](#sec-endogenous-sample-selection) developed elsewhere in this book both lean on a continuous endogenous regressor whose first-stage residual can be plugged into the structural equation. That machinery breaks down once the endogenous variable itself is binary. A binary endogenous regressor cannot be written as a linear projection onto instruments plus a well-behaved continuous error, so the residual used in the standard two-step control function is neither normally distributed nor independent of the regressors in the way the Rivers and Vuong two-step procedure requires [@rivers1988limited; @wooldridge2015control].

Binary endogenous regressors are common in applied work: a treatment dummy in a program evaluation, a union membership indicator, a college attendance flag, an insurance enrollment choice, or a firm's decision to adopt a technology. Each of these is a discrete choice that is itself correlated with the unobserved determinants of the outcome, so ignoring the endogeneity produces the same asymptotic bias that motivates instrumental variables in the continuous case. What changes is the toolkit needed to correct it.

This section works through three configurations that arise depending on whether the outcome and the endogenous regressor are continuous or binary, replicates the classic Heckman (1978) dummy endogenous variable model on real labor supply data, and closes with a discussion of when a simple linear probability model with two stage least squares is the more defensible applied choice.

------------------------------------------------------------------------

#### Three configurations

Let $D_i \in \{0,1\}$ denote the endogenous regressor, $Y_i$ the outcome, and $Z_i$ a vector of instruments excluded from the structural equation for $Y_i$ but relevant for $D_i$.

1.  **Continuous outcome, binary endogenous regressor.** The structural equation is linear,
    $$
    Y_i = \beta_0 + \beta_1 D_i + X_i'\beta_2 + u_i,
    $$
    and $D_i$ is generated by a latent index crossing a threshold,
    $$
    D_i = \mathbb{1}[Z_i'\gamma + X_i'\delta + v_i > 0].
    $$
    This is the setting Heckman (1978) originally analyzed under the label dummy endogenous variables [@heckman1978dummy]. Two consistent estimators are available: linear IV/2SLS treating $D_i$ as any other endogenous regressor, and a control function two-step estimator built from the first-stage probit's generalized residual.

2.  **Binary outcome, continuous endogenous regressor.** This is the Rivers and Vuong (1988) case already introduced in [the control function section](#sec-control-function-approach): a probit outcome with a continuous endogenous regressor whose linear first-stage residual can be added directly to the second-stage probit and tested with a $t$-statistic [@rivers1988limited].

3.  **Binary outcome, binary endogenous regressor.** Both equations are threshold-crossing models,
    $$
    Y_i = \mathbb{1}[\beta_1 D_i + X_i'\beta_2 + u_i > 0], \qquad D_i = \mathbb{1}[Z_i'\gamma + X_i'\delta + v_i > 0],
    $$
    with $(u_i, v_i)$ jointly normal and correlated. This is Amemiya's (1978) generalized probit model, estimated as a recursive bivariate probit [@amemiya1978estimation]. No residual-plug-in control function is valid here: the correlation between $u_i$ and $v_i$ must be estimated jointly with the rest of the model, typically by full information maximum likelihood.

Configuration 2 is already covered under [control functions](#sec-control-function-approach); this section focuses on configurations 1 and 3, since neither has a dedicated treatment elsewhere in the book.

------------------------------------------------------------------------

#### Configuration 1: continuous outcome

##### Two step control function (Heckman 1978)

Under joint normality of $(u_i, v_i)$ with correlation $\rho$, the conditional expectation of the structural error given the endogenous regressor and instruments is
$$
E[u_i \mid Z_i, X_i, D_i] = \rho\sigma_u \left[ D_i \lambda(Z_i'\hat\gamma + X_i'\hat\delta) - (1-D_i)\lambda(-(Z_i'\hat\gamma + X_i'\hat\delta)) \right],
$$
where $\lambda(\cdot)$ is the inverse Mills ratio. The bracketed term is Wooldridge's generalized residual for a binary regressor [@wooldridge2015control]. The two-step estimator is:

1.  Estimate the first-stage probit of $D_i$ on $Z_i$ and $X_i$, and compute the generalized residual $\widehat{gr}_i$ for every observation.
2.  Regress $Y_i$ on $D_i$, $X_i$, and $\widehat{gr}_i$ by OLS. A significant coefficient on $\widehat{gr}_i$ is a test of $H_0: \rho = 0$, i.e., a test of exogeneity of $D_i$.

Standard errors from the second step must be corrected for the generated-regressor problem, either analytically or by bootstrapping both steps jointly.

##### Linear IV/2SLS

The competing estimator ignores the discreteness of $D_i$ entirely and instruments it as if it were continuous:
$$
\hat D_i = \hat\pi_0 + Z_i'\hat\pi_1 + X_i'\hat\pi_2, \qquad Y_i = \beta_0 + \beta_1 \hat D_i + X_i'\beta_2 + \varepsilon_i.
$$
This is consistent under the same exclusion and relevance conditions as any 2SLS estimator, regardless of whether the first-stage functional form is correctly specified, because 2SLS only requires the linear projection of $D_i$ on $(Z_i, X_i)$ to have nonzero correlation with $Z_i$ [@wooldridge2015control].

A tempting hybrid is to estimate a probit (or logit) first stage and then plug its fitted probabilities into the second-stage OLS as if they were first-stage fitted values. This is the *forbidden regression* [@angrist2009mostly]. The nonlinear fitted values $\hat\Phi(Z_i'\hat\gamma + X_i'\hat\delta)$ are not the linear projection of $D_i$ on the instruments, so their residual is not orthogonal to the included regressors, and the usual 2SLS consistency argument collapses; the resulting estimator is inconsistent except under functional-form assumptions that the procedure never tests. Two safe uses of the first-stage probit exist. The first is the generalized-residual control function above, which relies on joint normality explicitly. The second is to use the probit fitted probability *as an instrument* for $D_i$ rather than as a regressor: 2SLS with $\hat\Phi(\cdot)$ in the instrument set remains consistent even if the probit is misspecified, and it attains the efficiency bound when the probit happens to be correct [@wooldridge2015control]. The rule of thumb is that nonlinear first-stage predictions may enter the instrument list, never the regressor list.

------------------------------------------------------------------------

#### Configuration 3: recursive bivariate probit

When both $Y_i$ and $D_i$ are binary, the model is
$$
\begin{aligned}
Y_i^* &= \beta_1 D_i + X_i'\beta_2 + u_i, & Y_i &= \mathbb{1}[Y_i^* > 0] \\
D_i^* &= Z_i'\gamma + X_i'\delta + v_i, & D_i &= \mathbb{1}[D_i^* > 0]
\end{aligned}
$$
with $(u_i, v_i) \sim N(0, 0, 1, 1, \rho)$. Identification requires at least one instrument in $Z_i$ excluded from the $Y_i$ equation, exactly as in the linear case, though the model is technically identified off nonlinearity alone if no exclusion is available; applied practice strongly prefers a genuine exclusion restriction. The log-likelihood is built from the bivariate normal CDF $\Phi_2(\cdot,\cdot;\rho)$ evaluated at the four combinations of $Y_i, D_i \in \{0,1\}$, and $\rho$ is estimated jointly with $(\beta, \gamma, \delta)$ by full information maximum likelihood.

Freedman and Sekhon (2010) show that this joint likelihood is numerically fragile: even with a handful of covariates, standard optimizers can fail to converge or converge to a local optimum, and they argue that the two-step control function is not a safe substitute because it is only valid under the same joint normality assumption the likelihood itself imposes, so a two-step estimator inherits the fragility without inheriting the efficiency of full information maximum likelihood [@freedman2010endogeneity]. Their recommendation, echoed in practice, is to treat convergence diagnostics (multiple starting values, a well conditioned Hessian) as mandatory before reporting a recursive bivariate probit estimate.

------------------------------------------------------------------------

#### Replication: returns to college attendance and labor force participation

We replicate the Heckman (1978) dummy endogenous variable model on Mroz's (1987) data on married women's labor supply, the same dataset used in the book's treatment of [Heckman selection](#sec-endogenous-sample-selection) [@mroz1987sensitivity]. Unlike the selection application, here the binary regressor of interest is the wife's college attendance (`wc`), and the question is whether attending college has a causal effect on (a) log wage, a continuous outcome, and (b) labor force participation itself, a binary outcome. Parental education (`motheduc`, `fatheduc`) instruments college attendance: it plausibly shifts the wife's own educational attainment through household background without directly entering either the wage or participation equation once the wife's own characteristics are controlled for. This is the same identification logic used throughout the intergenerational-schooling literature, and the reader should treat it, as with any single-instrument application, as illustrative rather than beyond challenge.

```{r, message = FALSE, warning = FALSE}
library(sampleSelection) # Mroz87 data
library(tidyverse)
library(modelsummary)
library(VGAM) # recursive bivariate probit via vglm(..., binom2.rho)

data("Mroz87")

mroz <- Mroz87 %>%
    mutate(
        kids = kids5 + kids618,
        # wife attended college (binary endogenous regressor); the factor
        # levels of wifecoll carry a leading space in this dataset
        wc   = as.integer(trimws(wifecoll) == "TRUE")
    )
```

##### Configuration 1: log wage as a continuous outcome

We restrict to working wives, since wage is observed only for participants, and compare naive OLS, linear 2SLS, and the two-step control function.

```{r, message = FALSE, warning = FALSE}
working <- mroz %>% filter(lfp == 1, wage > 0)

# Naive OLS: treats wc as exogenous
ols_fit <- lm(log(wage) ~ wc + exper + I(exper^2) + city, data = working)

# First-stage probit for the generalized residual
first_stage <- glm(wc ~ motheduc + fatheduc + exper + I(exper^2) + city,
                    data = working, family = binomial(link = "probit"))

eta_hat <- predict(first_stage, type = "link")
gr <- ifelse(working$wc == 1,
             dnorm(eta_hat) / pnorm(eta_hat),
             -dnorm(eta_hat) / (1 - pnorm(eta_hat)))

# Two-step control function
cf_fit <- lm(log(wage) ~ wc + exper + I(exper^2) + city + gr, data = working)

# Linear 2SLS
library(fixest)
iv_fit <- feols(log(wage) ~ exper + I(exper^2) + city | wc ~ motheduc + fatheduc,
                 data = working)

modelsummary(
    list("OLS" = ols_fit, "Control function" = cf_fit, "2SLS" = iv_fit),
    stars = TRUE,
    gof_omit = "IC|Log|Adj"
)
```

The coefficient on the generalized residual `gr` in the control function model is the direct test of $H_0: \rho = 0$. If it is not statistically distinguishable from zero, the naive OLS estimate of the return to college attendance is defensible and the more complex correction buys nothing; if it is significant, both the control function and 2SLS estimates should be preferred over OLS, and their similarity (or lack of it) is itself informative about first-stage strength and functional form sensitivity.

##### Configuration 3: labor force participation as a binary outcome

Now let the outcome be labor force participation itself, `lfp`, with college attendance as the endogenous regressor in a second binary equation. Because both variables are binary, the two-step control function of the previous subsection is not valid; we instead fit the joint model by maximum likelihood as a recursive bivariate probit.

```{r, message = FALSE, warning = FALSE}
# Naive probit: ignores endogeneity of wc
naive_probit <- glm(lfp ~ wc + kids + age + huswage,
                     data = mroz, family = binomial(link = "probit"))

# Recursive bivariate probit: wc equation carries the excluded instruments
biprobit_fit <- vglm(
    cbind(wc, lfp) ~ motheduc + fatheduc + kids + age + huswage,
    family = binom2.rho,
    data = mroz
)

summary(biprobit_fit)
```

The `rho` parameter reported by `vglm` is the estimated correlation between the two equations' errors. A confidence interval for `rho` that excludes zero is evidence that `wc` is endogenous in the participation equation, in which case the naive probit's marginal effect for `wc` is inconsistent and the recursive bivariate probit's average partial effect, computed by averaging $\Phi\left(\frac{\hat\beta_1 D + X'\hat\beta_2 + \hat\rho \hat v_i}{\sqrt{1-\hat\rho^2}}\right)$ over the estimation sample, is the estimate to report instead.

------------------------------------------------------------------------

#### Show and tell: robustness

Three checks are standard practice before reporting either estimator.

1.  **First-stage strength.** Report the first-stage probit's pseudo $R^2$ or a likelihood-ratio test of the excluded instruments, in addition to the linear-model $F$-statistic if a linear first stage is also estimated for comparison. A weak first stage undermines the control function residual exactly as it undermines 2SLS.
2.  **Convergence diagnostics for the bivariate probit.** Refit `biprobit_fit` from at least two different starting values (`vglm`'s default and a small perturbation of it) and confirm the log-likelihood and `rho` estimate are stable, following Freedman and Sekhon's caution about numerical fragility [@freedman2010endogeneity].
3.  **Instrument exclusion.** Reestimate configuration 1 using only `motheduc` or only `fatheduc` as the excluded instrument, and reestimate configuration 3 on the subsample of wives with at least one parent's education reported at both extremes of the distribution, to see whether the recovered effect of college attendance is sensitive to which part of the instrument's support is doing the identifying work.

```{r, message = FALSE, warning = FALSE}
# First-stage strength for configuration 1
first_stage_lm <- lm(wc ~ motheduc + fatheduc + exper + I(exper^2) + city, data = working)
car::linearHypothesis(first_stage_lm, c("motheduc = 0", "fatheduc = 0"))

# Single-instrument robustness checks
iv_mother_only <- feols(log(wage) ~ exper + I(exper^2) + city | wc ~ motheduc, data = working)
iv_father_only <- feols(log(wage) ~ exper + I(exper^2) + city | wc ~ fatheduc, data = working)

modelsummary(
    list("Both instruments" = iv_fit, "Mother's educ. only" = iv_mother_only, "Father's educ. only" = iv_father_only),
    stars = TRUE,
    gof_omit = "IC|Log|Adj"
)
```

------------------------------------------------------------------------

#### Extension: the special regressor alternative

Both estimators above rely on joint normality of the error terms, an assumption that is convenient but untestable in the way the linearity assumption behind ordinary 2SLS is not. Lewbel (2000) proposes a semiparametric alternative for binary-outcome, binary-or-continuous-endogenous-regressor models that avoids the normality assumption entirely, at the cost of requiring a continuously distributed, large-support "special regressor" that enters the outcome equation with a coefficient normalized to one and is independent of the structural error [@lewbel2000semiparametric]. When such a regressor is credible, for example a continuous cost or price shifter that plausibly satisfies these conditions, the special regressor estimator is a useful robustness check against the bivariate probit's normality assumption. It is more data hungry than the parametric alternatives, since its convergence rate depends on nonparametric density estimation, and finding a genuine special regressor is often no easier than finding a valid instrument in the first place.

A simpler and frequently underrated robustness check is to ignore the discreteness of both variables altogether and estimate a linear probability model by 2SLS, clustering standard errors by any natural grouping in the data. Comparative studies of these estimators find that the linear probability model with 2SLS rarely gets the direction of the average partial effect wrong, even when the underlying data-generating process is a threshold-crossing model, whereas control function estimators for binary outcomes can be sensitive to first-stage misspecification precisely because they require the parametric normality assumption to hold exactly. The recommended practice is to report the linear probability/2SLS estimate alongside the bivariate probit or control function estimate rather than as a substitute for it, since the two differ in what they assume and a large gap between them is itself diagnostic of assumption failure rather than a nuisance to be resolved by picking one.

------------------------------------------------------------------------

#### Applied relevance

Binary endogenous regressors arise constantly in litigation and policy settings where the object of interest is a discrete choice: whether an employer offered health insurance, whether a firm adopted a disputed pricing algorithm, whether a worker was covered by a union contract, or whether a student attended a specific school type. In each of these settings the analyst rarely has the luxury of a continuous treatment intensity to instrument in the usual way, and the choice between a linear probability 2SLS estimate, a control function estimate, and a full bivariate probit is not merely a technical preference: it determines whether the reported effect is defensible under cross-examination on the grounds that it does not depend on an unverifiable joint-normality assumption. The safest applied posture, consistent with the robustness exercise above, is to lead with the linear probability/2SLS estimate as the assumption-light baseline and present the parametric alternatives as a check on functional form, rather than the reverse.

> Key Takeaway: when the endogenous regressor is binary, the standard control function two-step is a well-defined estimator only when the outcome is continuous. If the outcome is also binary, joint normality must be imposed and estimated by full information maximum likelihood as a recursive bivariate probit, and its numerical fragility means a linear probability/2SLS estimate should always be reported alongside it as a robustness benchmark.
