# Directed Acyclic Graphs {#sec-directed-acyclic-graphs}

Directed acyclic graphs (DAGs) provide a formal and visual framework for representing assumptions about causal structures. In modern data analysis, they are essential tools for understanding, identifying, and validating causal effects [@Pearl_2018; @pearl2014graphical].

A DAG is a graph composed of nodes (representing variables) and directed edges (arrows) showing the direction of causality. *Acyclic* means that the graph contains no feedback loops; one cannot return to the same node by following the direction of the arrows. The arrows encode *structural* assumptions: an arrow from $X$ to $Y$ states that, holding all other parents of $Y$ fixed, intervening on $X$ can change $Y$.

Understanding DAGs helps analysts reason about:

-   Which variables to control for in a regression model.
-   How to avoid collider bias and confounding.
-   What types of data are needed to estimate causal effects.
-   Where causal identification fails due to unobserved variables.

These R packages facilitate DAG creation, visualization, and causal analysis:

-   `dagitty`: Powerful syntax for defining DAGs, checking d-separation, and performing adjustment-set analysis.
-   `ggdag`: A `ggplot2`-based visualization tool for DAGs, compatible with `dagitty`, providing publication-ready DAGs.
-   `dagR`: Focuses on applied epidemiological use of DAGs, particularly for teaching.
-   `r-causal`: Developed by the [Center for Causal Discovery](https://www.ccd.pitt.edu/data-science/). Offers methods for causal discovery from data (also in Python).

Web tools:

-   **Publication-ready DAG editor**: [shinyDAG](https://www.gerkelab.com/project/shinydag/) -- An R-based Shiny app for creating DAGs interactively.
-   **Standalone DAG tool**: [DAG program](https://hsz.dife.de/dag/) by Sven Knüppel -- Suited to beginners needing an intuitive graphical interface.

```{r dag-setup, message=FALSE, warning=FALSE}
library(dagitty)
library(ggdag)
library(ggplot2)
```

## Basic notation and graph structures

Directed acyclic graphs are composed of three elementary building blocks: chains, forks, and inverted forks. Every path in any DAG decomposes into a sequence of these primitives, so the rules below generalise to graphs of arbitrary size.

### Mediators (chains)

$$
X \to Z \to Y
$$

-   Variable $Z$ mediates the effect of $X$ on $Y$.
-   Conditioning on $Z$ blocks the *indirect* effect of $X$ on $Y$ that flows through $Z$.
-   Marketing example: an email promotion ($X$) raises customer interest ($Z$), which raises purchase probability ($Y$). Controlling for interest removes the indirect path and isolates any remaining direct impact of the email.

### Common causes (forks)

$$
X \leftarrow Z \to Y
$$

-   $Z$ is a *confounder* and induces a non-causal association between $X$ and $Y$.
-   To estimate the causal effect of $X$ on $Y$, the path through $Z$ must be blocked, typically by conditioning on $Z$.
-   Finance example: a macroeconomic indicator ($Z$) shifts both investment decisions ($X$) and market returns ($Y$).

> If $Z$ is left uncontrolled, $X$ and $Y$ will appear correlated through their shared cause rather than through any causal link.

### Common effects (colliders)

$$
X \to Z \leftarrow Y
$$

-   $Z$ is a *collider* on the path $X \to Z \leftarrow Y$. The path is blocked by default.
-   Conditioning on $Z$ -- or on any descendant of $Z$ -- *opens* the path and induces a spurious association between $X$ and $Y$.
-   HR analytics example: education ($X$) and experience ($Y$) are independent in the applicant pool but both raise the probability of being hired ($Z$). Among hires only, education and experience become negatively associated even though no causal link exists.

### Other key concepts

-   **Descendants.** A descendant of a node $V$ is any variable reachable by following arrows out of $V$. Conditioning on a descendant of a *collider* is almost as bad as conditioning on the collider itself: it partially opens the same non-causal path, because the descendant carries information about the collider. Conditioning on a descendant of a *confounder*, by contrast, can serve as a proxy adjustment, though usually with residual bias.
-   **d-separation.** A graphical criterion for conditional independence. A path is *blocked* by a conditioning set $\mathbf{S}$ when (i) it contains a chain or fork whose middle node is in $\mathbf{S}$, or (ii) it contains a collider such that neither the collider nor any of its descendants is in $\mathbf{S}$. Two nodes $X$ and $Y$ are *d-separated given* $\mathbf{S}$ when every path between them is blocked; this implies $X \perp\!\!\!\perp Y \mid \mathbf{S}$ in any distribution compatible with the graph.

## Identification: back-door and front-door criteria

Knowing how chains, forks, and colliders behave is half the battle; the other half is turning that knowledge into a recipe that picks out the *right* set of variables to condition on. Pearl's two classical criteria do exactly that. The back-door criterion handles the common case in which all confounders are observable; the front-door criterion rescues identification when they are not, provided a fully mediated pathway is observed instead.

To validly estimate the causal effect of $X$ on $Y$ in a DAG, we need a set $\mathbf{S}$ of measured variables that satisfies one of two classical identification criteria.

### The back-door criterion

A set $\mathbf{S}$ satisfies the **back-door criterion** relative to $(X, Y)$ when:

1.  no node in $\mathbf{S}$ is a descendant of $X$, and
2.  $\mathbf{S}$ blocks every path between $X$ and $Y$ that begins with an arrow into $X$ ("back-door paths").

When such an $\mathbf{S}$ exists,
$$
P(Y \mid \operatorname{do}(X = x)) = \sum_{s} P(Y \mid X = x, \mathbf{S} = s)\, P(\mathbf{S} = s),
$$
the familiar covariate-adjustment formula. The back-door criterion is the workhorse behind regression adjustment, propensity-score methods, and matching [@Pearl_2018].

### The front-door criterion

When every back-door adjustment set contains an *unmeasured* variable, back-door adjustment fails. The **front-door criterion** offers a second route to identification through a fully mediated pathway. Suppose all of the effect of $X$ on $Y$ flows through an observed mediator $M$, and there is no unblocked back-door path from $X$ to $M$ or from $M$ to $Y$ (other than through $X$). Then
$$
P(Y \mid \operatorname{do}(X)) = \sum_m P(M = m \mid X) \sum_{x'} P(Y \mid M = m, X = x')\, P(X = x'),
$$
which is identifiable from observed data even when $X$ and $Y$ share an unmeasured confounder. The canonical illustration is given in Section \@ref(sec-frontdoor) below.

## Confounders, mediators, and colliders side by side

The three primitive structures look superficially similar but call for opposite analytic decisions: condition on a confounder, leave a collider alone, and condition on a mediator only when total effects are not the target. Figure \@ref(fig:dag-three-roles) places them next to one another.

```{r dag-three-roles, fig.cap="Three elementary DAG structures: a confounder (fork), a mediator (chain), and a collider (inverted fork).", fig.alt="Three small DAGs side by side showing X and Y connected through Z as a fork, chain, and inverted fork respectively.", fig.width=8, fig.height=3}
fork <- dagitty("dag { X <- Z -> Y }")
coordinates(fork) <- list(x = c(X = 1, Z = 2, Y = 3), y = c(X = 1, Z = 2, Y = 1))

chain <- dagitty("dag { X -> Z -> Y }")
coordinates(chain) <- list(x = c(X = 1, Z = 2, Y = 3), y = c(X = 1, Z = 1, Y = 1))

collider <- dagitty("dag { X -> Z; Y -> Z }")
coordinates(collider) <- list(x = c(X = 1, Z = 2, Y = 3), y = c(X = 1, Z = 0, Y = 1))

p1 <- ggdag(fork) + theme_dag() + ggtitle("Fork (confounder)")
p2 <- ggdag(chain) + theme_dag() + ggtitle("Chain (mediator)")
p3 <- ggdag(collider) + theme_dag() + ggtitle("Inverted fork (collider)")

# Combine into a single figure so the chunk label resolves to one image
# rather than three separate sub-figures.
gridExtra::grid.arrange(p1, p2, p3, ncol = 3)
```

The decision rule is summarised in Table \@ref(tab:dag-roles).

```{r dag-roles, echo=FALSE}
roles <- data.frame(
  Role = c("Confounder", "Mediator", "Collider"),
  Structure = c("X <- Z -> Y", "X -> Z -> Y", "X -> Z <- Y"),
  `Default state` = c("Path open", "Path open", "Path blocked"),
  `Effect of conditioning on Z` = c(
    "Closes the back-door path; removes confounding bias.",
    "Blocks indirect effect; isolates direct effect only.",
    "Opens the path; induces spurious X--Y association."
  ),
  check.names = FALSE
)
knitr::kable(
  roles,
  booktabs = TRUE,
  caption = "Three roles of a third variable Z and the consequence of conditioning on it."
)
```

## A confounded mediator example

We now build a small DAG that contains *both* a mediator and a true confounder. The exposure $X$ affects $Y$ through a mediator $Z$, and an unobserved variable $U$ confounds the $X \to Y$ relationship by acting as a common cause of $X$ and $Y$. Figure \@ref(fig:dag-mediator-confounder) shows the structure.

```{r dag-mediator-confounder, fig.cap="Confounded mediator: X affects Y both directly and through the mediator Z, while U is an unobserved common cause of X and Y.", fig.alt="DAG with X on the left, Z in the middle, Y on the right, and U above as a parent of both X and Y, with X also pointing directly to Y."}
dag <- dagitty("dag {
  X -> Z -> Y
  X -> Y
  U -> X
  U -> Y
  U [unobserved]
}")

coordinates(dag) <- list(
  x = c(X = 1, Z = 2, Y = 3, U = 2),
  y = c(X = 1, Z = 1, Y = 1, U = 2)
)

ggdag(dag) + theme_dag()
```

This DAG has:

-   A direct causal path $X \to Y$.
-   A mediated causal path $X \to Z \to Y$.
-   A back-door path through an unobserved confounder: $X \leftarrow U \to Y$.

Because $U$ is unobserved, regressing $Y$ on $X$ and $Z$ alone will not identify the total causal effect of $X$ on $Y$. `dagitty` confirms there is no admissible adjustment set built from observed variables:

```{r}
adjustmentSets(dag, exposure = "X", outcome = "Y", type = "all")
```

If $U$ were observed, the singleton $\{U\}$ would close the back-door path and identify the total effect of $X$ on $Y$. Conditioning *also* on the mediator $Z$ would instead identify the controlled direct effect of $X$ on $Y$ holding $Z$ fixed -- a different estimand whose interpretation is much more delicate [@vanderweele2015explanation].

## M-bias: when adjusting for a pre-treatment variable hurts

A common misconception is that adjusting for any pre-treatment covariate can only reduce bias. The **M-bias** structure shows this is false. Consider Figure \@ref(fig:dag-mbias): two unobserved variables $U_1$ and $U_2$ each cause an observed pre-treatment variable $Z$, and they also separately cause $X$ and $Y$. The graph contains no confounding of $X$ on $Y$ -- the only path from $X$ to $Y$ that does not go through $X$'s causal descendants is $X \leftarrow U_1 \to Z \leftarrow U_2 \to Y$, which is blocked at the collider $Z$.

```{r dag-mbias, fig.cap="M-bias structure: Z is a pre-treatment collider on the path X <- U1 -> Z <- U2 -> Y, so conditioning on Z opens a non-causal path between X and Y.", fig.alt="DAG arranged as the letter M with U1 and U2 at the top corners pointing down to Z in the centre and to X and Y at the bottom corners; X also points to Y."}
mbias <- dagitty("dag {
  U1 -> X
  U1 -> Z
  U2 -> Z
  U2 -> Y
  X  -> Y
  U1 [unobserved]
  U2 [unobserved]
}")

coordinates(mbias) <- list(
  x = c(U1 = 1, U2 = 3, Z = 2, X = 1, Y = 3),
  y = c(U1 = 2, U2 = 2, Z = 1.5, X = 1, Y = 1)
)

ggdag(mbias) + theme_dag()
```

If we leave $Z$ alone, no adjustment is needed: the marginal regression of $Y$ on $X$ recovers the causal effect because the only back-door path is already blocked by the collider $Z$. If, however, an analyst routinely "controls for everything pre-treatment" and includes $Z$, conditioning *opens* the path $X \leftarrow U_1 \to Z \leftarrow U_2 \to Y$, and the estimate of $X \to Y$ becomes biased.

```{r}
# No conditioning needed: Z is a pre-treatment collider.
adjustmentSets(mbias, exposure = "X", outcome = "Y")

# But if we naively condition on Z, paths reopen.
isAdjustmentSet(mbias, "Z", exposure = "X", outcome = "Y")
```

A short simulation makes the bias concrete:

```{r}
set.seed(42)
n  <- 5000
U1 <- rnorm(n); U2 <- rnorm(n)
X  <- 0.8 * U1 + rnorm(n)
Z  <- 0.8 * U1 + 0.8 * U2 + rnorm(n)
Y  <- 1.0 * X + 0.8 * U2 + rnorm(n)

coef(lm(Y ~ X))["X"]      # close to the true effect (1.0)
coef(lm(Y ~ X + Z))["X"]  # biased away from 1.0 because Z is a collider
```

The "control for everything" heuristic fails here. M-bias is the prototypical reason that pre-treatment covariates cannot be adjusted for indiscriminately [@cinelli2022crash].

## Berkson's paradox and selection on a collider

A close cousin of M-bias arises through *sample selection*. Whenever the data we analyse are restricted to units satisfying some condition $S$, and $S$ is a descendant of a collider, the analysis effectively conditions on that collider. The result is **Berkson's paradox**: variables that are independent in the population become correlated in the selected sample.

Suppose we want to study whether acting talent ($X$) affects beauty ($Y$) among Hollywood celebrities. In the general population the two are plausibly independent. Both, however, raise the probability of becoming famous ($S$), and our sample contains only the famous. Figure \@ref(fig:dag-berkson) encodes this story.

```{r dag-berkson, fig.cap="Berkson selection: talent and beauty are independent in the population but both raise the probability of selection S, so restricting to S = 1 induces a spurious negative association.", fig.alt="DAG with Talent and Beauty as parents of Selection, no arrow between Talent and Beauty, and Selection drawn with a box indicating that the analysis conditions on it."}
berkson <- dagitty("dag {
  Talent -> Selection
  Beauty -> Selection
  Selection [adjusted]
}")

coordinates(berkson) <- list(
  x = c(Talent = 1, Beauty = 3, Selection = 2),
  y = c(Talent = 2, Beauty = 2, Selection = 1)
)

ggdag(berkson) + theme_dag()
```

A simulation confirms the paradox:

```{r}
set.seed(7)
n      <- 20000
talent <- rnorm(n)
beauty <- rnorm(n)
# Famous if combined score is high enough.
famous <- (talent + beauty + rnorm(n)) > 1.5

cor(talent, beauty)                             # near 0
cor(talent[famous], beauty[famous])             # spuriously negative
```

The lesson generalises: any analysis that filters cases on an outcome-related criterion (hospital records, survey respondents who replied, customers retained for at least 90 days) is implicitly conditioning on a collider, and "obvious" associations among the survivors may be entirely artefactual.

## The front-door criterion: smoking, tar, and cancer {#sec-frontdoor}

Suppose we wish to estimate the effect of smoking ($X$) on lung cancer ($Y$), but suspect an unobserved genetic factor ($U$) affects both. Direct back-door adjustment is impossible because $U$ is unmeasured. If, however, smoking causes cancer *only* through tar deposition in the lungs ($M$), and the genetic factor does not affect tar except through smoking, the front-door criterion identifies the causal effect from $(X, M, Y)$ alone. Figure \@ref(fig:dag-frontdoor) illustrates.

```{r dag-frontdoor, fig.cap="Front-door identification: smoking affects lung cancer only via tar, and the unobserved genotype U confounds smoking and cancer but does not affect tar directly.", fig.alt="DAG with U as an unobserved parent of both Smoking and Cancer, and a chain Smoking -> Tar -> Cancer running below."}
frontdoor <- dagitty("dag {
  Smoking -> Tar -> Cancer
  U -> Smoking
  U -> Cancer
  U [unobserved]
}")

coordinates(frontdoor) <- list(
  x = c(Smoking = 1, Tar = 2, Cancer = 3, U = 2),
  y = c(Smoking = 1, Tar = 1, Cancer = 1, U = 2)
)

ggdag(frontdoor) + theme_dag()
```

`dagitty` knows the front-door logic: it returns no back-door adjustment set built from observed variables, but recognises the front-door identification when asked.

```{r}
adjustmentSets(frontdoor, exposure = "Smoking", outcome = "Cancer", type = "all")
adjustmentSets(frontdoor, exposure = "Smoking", outcome = "Cancer",
               effect = "direct")
```

The mediator $M$ (tar) plays a dual role: it is part of the causal pathway, *and* it is itself unconfounded with the outcome conditional on $X$. The front-door criterion is the historically important demonstration that causal effects can be identified from observational data even in the presence of unmeasured confounders, provided the mediation structure is credible [@Pearl_2018].

## Conditioning on a descendant of a collider

A subtle but practically common error is to condition not on a collider itself but on one of its *descendants*. This still partially opens the non-causal path, because the descendant carries probabilistic information about the collider. Figure \@ref(fig:dag-collider-descendant) shows the canonical case: $X \to Z \leftarrow Y$, with $Z$ generating an observable downstream variable $W$ (for instance, a survey response or an automated flag).

```{r dag-collider-descendant, fig.cap="Conditioning on W, a descendant of the collider Z, reopens the non-causal path X -> Z <- Y and biases the estimated X--Y association.", fig.alt="DAG with X and Y both pointing into Z, and Z pointing down to W; W is highlighted as the conditioning variable."}
descendant <- dagitty("dag {
  X -> Z
  Y -> Z
  Z -> W
  W [adjusted]
}")

coordinates(descendant) <- list(
  x = c(X = 1, Y = 3, Z = 2, W = 2),
  y = c(X = 2, Y = 2, Z = 1, W = 0)
)

ggdag(descendant) + theme_dag()
```

A short simulation shows that the bias from adjusting for $W$ is qualitatively the same as from adjusting for $Z$, only attenuated by the strength of the $Z \to W$ link.

```{r}
set.seed(1)
n <- 5000
X <- rnorm(n); Y <- rnorm(n)         # truly independent
Z <- X + Y + rnorm(n)
W <- Z + rnorm(n)                    # noisy descendant of Z

cor(X, Y)                            # ~ 0
coef(lm(Y ~ X + Z))["X"]             # spurious negative
coef(lm(Y ~ X + W))["X"]             # also spurious, smaller magnitude
```

The takeaway is that the d-separation rule for colliders extends to all descendants: a collider blocks a path only when it *and all of its descendants* are absent from the conditioning set.

## Good and bad controls: a checklist

Cinelli, Forney, and Pearl's "crash course in good and bad controls" works through eighteen canonical DAG configurations and groups them into a small number of decision rules about which variables to adjust for [@cinelli2022crash]. The headline rules, restated graphically, are:

-   **Good controls.** Variables on a back-door path -- whether or not they are direct causes of $X$ -- block confounding when conditioned on. Proxies for unobserved confounders generally help, though residual bias remains.
-   **Neutral controls.** Variables that are causes of $Y$ but not of $X$, and not on any back-door path, do not bias the effect estimate but can improve precision.
-   **Bad controls.** Mediators (when total effects are the target), colliders, descendants of colliders, and "M-bias" pre-treatment colliders should *not* be adjusted for. Adjusting for an instrument of $X$ amplifies bias from any remaining unobserved confounders.

The practical workflow is therefore: draw the DAG first, then ask `dagitty::adjustmentSets()` -- not the other way around.

## Causal discovery

Causal discovery involves algorithmically identifying causal relationships from data under structural assumptions such as faithfulness and causal sufficiency. Key algorithms include:

-   **PC algorithm**: Constraint-based, uses conditional-independence testing.
-   **GES (Greedy Equivalence Search)**: Score-based method.
-   **FCI (Fast Causal Inference)**: Extends PC to handle latent confounders.

> See [@eberhardt2024discovering] for a comprehensive discussion of the assumptions and limitations of discovery algorithms in practice.

## Practical workflow

A defensible DAG-based analysis proceeds in four steps.

1.  **Draw the DAG** from substantive knowledge, including any unobserved variables that domain experts insist on. This step is unavoidable: no statistical procedure can recover what the analyst is unwilling to commit to on paper.
2.  **Check identification.** Use `dagitty::adjustmentSets()` (or the equivalent in `ggdag`) to ask whether the target effect is identifiable from observed data, and if so, with which adjustment sets.
3.  **Check the DAG against data** through testable implications -- conditional independencies the DAG predicts. `dagitty::impliedConditionalIndependencies()` lists them; rejecting them in data is evidence the DAG is misspecified.
4.  **Estimate**, using a method (regression, matching, propensity scores, doubly robust estimators) that is faithful to the chosen adjustment set. The DAG only tells us *what* to adjust for; estimation choices still matter for finite-sample performance.

Following these steps does not guarantee a correct answer -- assumptions can be wrong -- but it makes those assumptions explicit, falsifiable, and reviewable, which is the minimum a Springer-grade causal analysis should offer.
