# Report

Structure

-   Exploratory analysis

    -   plots
    -   preliminary results
    -   interesting structure/features in the data
    -   outliers

-   Model

    -   Assumptions
    -   Why this model/ How is this model the best one?
    -   Consideration: interactions, collinearity, dependence

-   Model Fit

    -   How well does it fit?

    -   Are the model assumptions met?

        -   Residual analysis

-   Inference/ Prediction

    -   Are there different way to support your inference?

-   Conclusion

    -   Recommendation

    -   Limitation of the analysis

    -   How to correct those in the future

This chapter is based on the `jtools` package. More information can be found [here.](https://www.rdocumentation.org/packages/jtools/versions/2.1.0)

## One summary table

Packages for reporting:

Summary Statistics Table:

-   [qwraps2](https://cran.r-project.org/web/packages/qwraps2/vignettes/summary-statistics.html)
-   [vtable](https://cran.r-project.org/web/packages/vtable/vignettes/sumtable.html)
-   [gtsummary](http://www.danieldsjoberg.com/gtsummary/)
-   [apaTables](https://cran.r-project.org/web/packages/apaTables/apaTables.pdf)
-   [stargazer](https://cran.r-project.org/web/packages/stargazer/vignettes/stargazer.pdf)

Regression Table

-   [gtsummary](http://www.danieldsjoberg.com/gtsummary/)
-   [sjPlot,sjmisc, sjlabelled](https://cran.r-project.org/web/packages/sjPlot/vignettes/tab_model_estimates.html)
-   [stargazer](https://cran.r-project.org/web/packages/stargazer/vignettes/stargazer.pdf): recommended ([Example](https://www.jakeruss.com/cheatsheets/stargazer/))
-   [modelsummary](https://github.com/vincentarelbundock/modelsummary#a-simple-example)

```{r}
library(jtools)
data(movies)
fit <- lm(metascore ~ budget + us_gross + year, data = movies)
summ(fit)
summ(
    fit,
    scale = TRUE,
    vifs = TRUE,
    part.corr = TRUE,
    confint = TRUE,
    pvals = FALSE
) # notice that scale here is TRUE

#obtain clsuter-robust SE
data("PetersenCL", package = "sandwich")
fit2 <- lm(y ~ x, data = PetersenCL)
summ(fit2, robust = "HC3", cluster = "firm") 
```

Model to Equation

```{r, eval = FALSE}
# install.packages("equatiomatic") # not available for R 4.2
fit <- lm(metascore ~ budget + us_gross + year, data = movies)
# show the theoretical model
equatiomatic::extract_eq(fit)
# display the actual coefficients
equatiomatic::extract_eq(fit, use_coefs = TRUE)
```

## Model Comparison

```{r}
fit <- lm(metascore ~ log(budget), data = movies)
fit_b <- lm(metascore ~ log(budget) + log(us_gross), data = movies)
fit_c <- lm(metascore ~ log(budget) + log(us_gross) + runtime, data = movies)
coef_names <- c("Budget" = "log(budget)", "US Gross" = "log(us_gross)",
                "Runtime (Hours)" = "runtime", "Constant" = "(Intercept)")
export_summs(fit, fit_b, fit_c, robust = "HC3", coefs = coef_names)
```

Another package is `modelsummary`

```{r}
library(modelsummary)
lm_mod <- lm(mpg ~ wt + hp + cyl, mtcars)
msummary(lm_mod, vcov = c("iid","robust","HC4"))
modelplot(lm_mod, vcov = c("iid","robust","HC4"))
```

Another package is `stargazer`

```{r}
library("stargazer")
stargazer(attitude)
## 2 OLS models
linear.1 <-
    lm(rating ~ complaints + privileges + learning + raises + critical,
       data = attitude)
linear.2 <-
    lm(rating ~ complaints + privileges + learning, data = attitude)
## create an indicator dependent variable, and run a probit model
attitude$high.rating <- (attitude$rating > 70)

probit.model <-
    glm(
        high.rating ~ learning + critical + advance,
        data = attitude,
        family = binomial(link = "probit")
    )
stargazer(linear.1,
          linear.2,
          probit.model,
          title = "Results",
          align = TRUE)
```

```{r eval = FALSE}
# Latex
stargazer(
    linear.1,
    linear.2,
    probit.model,
    title = "Regression Results",
    align = TRUE,
    dep.var.labels = c("Overall Rating", "High Rating"),
    covariate.labels = c(
        "Handling of Complaints",
        "No Special Privileges",
        "Opportunity to Learn",
        "Performance-Based Raises",
        "Too Critical",
        "Advancement"
    ),
    omit.stat = c("LL", "ser", "f"),
    no.space = TRUE
)
```

```{r}
# ASCII text output
stargazer(
    linear.1,
    linear.2,
    type = "text",
    title = "Regression Results",
    dep.var.labels = c("Overall Rating", "High Rating"),
    covariate.labels = c(
        "Handling of Complaints",
        "No Special Privileges",
        "Opportunity to Learn",
        "Performance-Based Raises",
        "Too Critical",
        "Advancement"
    ),
    omit.stat = c("LL", "ser", "f"),
    ci = TRUE,
    ci.level = 0.90,
    single.row = TRUE
)
```

```{r eval = FALSE}
stargazer(
    linear.1,
    linear.2,
    probit.model,
    title = "Regression Results",
    align = TRUE,
    dep.var.labels = c("Overall Rating", "High Rating"),
    covariate.labels = c(
        "Handling of Complaints",
        "No Special Privileges",
        "Opportunity to Learn",
        "Performance-Based Raises",
        "Too Critical",
        "Advancement"
    ),
    omit.stat = c("LL", "ser", "f"),
    no.space = TRUE
)
```

Correlation Table

```{r eval = FALSE}
correlation.matrix <-
    cor(attitude[, c("rating", "complaints", "privileges")])
stargazer(correlation.matrix, title = "Correlation Matrix")
```

## Changes in an estimate

```{r}
coef_names <- coef_names[1:3] # Dropping intercept for plots
plot_summs(fit, fit_b, fit_c, robust = "HC3", coefs = coef_names)
plot_summs(
    fit_c,
    robust = "HC3",
    coefs = coef_names,
    plot.distributions = TRUE
)
```

## Standard Errors

`sandwich` [vignette](cran.r-project.org/web/packages/sandwich/vignettes/sandwich-CL.pdf)

+------------+------------+------------------------------------------------------------------------------------------+--------------------------------+
| Type       | Applicable | Usage                                                                                    | Reference                      |
+============+============+==========================================================================================+================================+
| `const`    |            | Assume constant variances                                                                |                                |
+------------+------------+------------------------------------------------------------------------------------------+--------------------------------+
| `HC` `HC0` | `vcovCL`   | Heterogeneity                                                                            | [@white1980heteroskedasticity] |
|            |            |                                                                                          |                                |
|            |            | White's estimator                                                                        |                                |
|            |            |                                                                                          |                                |
|            |            | All other heterogeneity SE methods are derivatives of this.                              |                                |
|            |            |                                                                                          |                                |
|            |            | No small sample bias adjustment                                                          |                                |
+------------+------------+------------------------------------------------------------------------------------------+--------------------------------+
| `HC1`      | `vcovCL`   | Uses a degrees of freedom-based correction                                               | [@mackinnon1985some]           |
|            |            |                                                                                          |                                |
|            |            | When the number of clusters is small, `HC2` and `HC3` are better [@cameron2008bootstrap] |                                |
+------------+------------+------------------------------------------------------------------------------------------+--------------------------------+
| `HC2`      | `vcovCL`   | Better with the linear model, but still applicable for [Generalized Linear Models]       |                                |
|            |            |                                                                                          |                                |
|            |            | Needs a hat (weighted) matrix                                                            |                                |
+------------+------------+------------------------------------------------------------------------------------------+--------------------------------+
| `HC3`      | `vcovCL`   | Better with the linear model, but still applicable for [Generalized Linear Models]       |                                |
|            |            |                                                                                          |                                |
|            |            | Needs a hat (weighted) matrix                                                            |                                |
+------------+------------+------------------------------------------------------------------------------------------+--------------------------------+
| `HC4`      | `vcovHC`   |                                                                                          | [@cribari2004asymptotic]       |
+------------+------------+------------------------------------------------------------------------------------------+--------------------------------+
| `HC4m`     | `vcovHC`   |                                                                                          | [@cribari2007inference]        |
+------------+------------+------------------------------------------------------------------------------------------+--------------------------------+
| `HC5`      | `vcovHC`   |                                                                                          | [@cribari2011new]              |
+------------+------------+------------------------------------------------------------------------------------------+--------------------------------+

```{r}
data(cars)
model <- lm(speed ~ dist, data = cars)
summary(model)
lmtest::coeftest(model, vcov. = sandwich::vcovHC(model, type = "HC1"))
```

## Coefficient Uncertainty and Distribution

The `ggdist` allows us to visualize uncertainty under both frequentist and Bayesian frameworks

```{r, eval=FALSE}
library(ggdist)
```

## Descriptive Tables

Export APA theme

```{r, eval = FALSE}
data("mtcars")

library(flextable)
theme_apa(flextable(mtcars[1:5,1:5]))
```

Export to Latex

```{r, eval=FALSE}
print(xtable::xtable(mtcars, type = "latex"),
      file = file.path(getwd(), "output", "mtcars_xtable.tex"))

# American Economic Review style
stargazer::stargazer(
    mtcars,
    title = "Testing",
    style = "aer",
    out = file.path(getwd(), "output", "mtcars_stargazer.tex")
)

# other styles include
# Administrative Science Quarterly
# Quarterly Journal of Economics
```

However, the above codes do not play well with notes. Hence, I create my own custom code that follows the AMA guidelines

```{r, eval = FALSE}
ama_tbl <- function(data, caption, label, note, output_path) {
  library(tidyverse)
  library(xtable)
  # Function to determine column alignment
  get_column_alignment <- function(data) {
    # Start with the alignment for the header row
    alignment <- c("l", "l")
    
    # Check each column
    for (col in seq_len(ncol(data))[-1]) {
      if (is.numeric(data[[col]])) {
        alignment <- c(alignment, "r")  # Right alignment for numbers
      } else {
        alignment <- c(alignment, "c")  # Center alignment for other data
      }
    }
    
    return(alignment)
  }
  
  data %>%
    # bold + left align first column 
    rename_with(~paste("\\multicolumn{1}{l}{\\textbf{", ., "}}"), 1) %>% 
    # bold + center align all other columns
    `colnames<-`(ifelse(colnames(.) != colnames(.)[1],
                        paste("\\multicolumn{1}{c}{\\textbf{", colnames(.), "}}"),
                        colnames(.))) %>% 
    
    xtable(caption = caption,
           label = label,
           align = get_column_alignment(data),
           auto = TRUE) %>%
    print(
      include.rownames = FALSE,
      caption.placement = "top",
      
      hline.after=c(-1, 0),
      
       # p{0.9\linewidth} sets the width of the column to 90% of the line width, and the @{} removes any extra padding around the cell.
      
      add.to.row = list(pos = list(nrow(data)), # Add at the bottom of the table
                        command = c(paste0("\\hline \n \\multicolumn{",ncol(data), "}{l} {", "\n \\begin{tabular}{@{}p{0.9\\linewidth}@{}} \n","Note: ", note, "\n \\end{tabular}  } \n"))), # Add your note here
      
      # make sure your heading is untouched (because you manually change it above)
      sanitize.colnames.function = identity,
      
      # place a the top of the page
      table.placement = "h",
      
      file = output_path
    )
}
```

```{r, eval = FALSE}
ama_tbl(
    mtcars,
    caption     = "This is caption",
    label       = "tab:this_is_label",
    note        = "this is note",
    output_path = file.path(getwd(), "output", "mtcars_custom_ama.tex")
)
```

## Visualizations and Plots

You can customize your plots based on your preferred journals. Here, I am creating a custom setting for the American Marketing Association.

American-Marketing-Association-ready theme for plots

```{r}
library(ggplot2)

# check available fonts
# windowsFonts()

# for Times New Roman
# names(windowsFonts()[windowsFonts()=="TT Times New Roman"])
```

```{r}
# Making a theme
amatheme = theme_bw(base_size = 14, base_family = "serif") + # This is Time New Roman
    
    theme(
        # remove major gridlines
        panel.grid.major   = element_blank(),

        # remove minor gridlines
        panel.grid.minor   = element_blank(),

        # remove panel border
        panel.border       = element_blank(),

        line               = element_line(),

        # change font
        text               = element_text(),

        # if you want to remove legend title
        # legend.title     = element_blank(),

        legend.title       = element_text(size = rel(0.6), face = "bold"),

        # change font size of legend
        legend.text        = element_text(size = rel(0.6)),
        
        legend.background  = element_rect(color = "black"),
        
        # legend.margin    = margin(t = 5, l = 5, r = 5, b = 5),
        # legend.key       = element_rect(color = NA, fill = NA),

        # change font size of main title
        plot.title         = element_text(
            size           = rel(1.2),
            face           = "bold",
            hjust          = 0.5,
            margin         = margin(b = 15)
        ),
        
        plot.margin        = unit(c(1, 1, 1, 1), "cm"),

        # add black line along axes
        axis.line          = element_line(colour = "black", linewidth = .8),
        
        axis.ticks         = element_line(),
        

        # axis title
        axis.title.x       = element_text(size = rel(1.2), face = "bold"),
        axis.title.y       = element_text(size = rel(1.2), face = "bold"),

        # axis text size
        axis.text.y        = element_text(size = rel(1)),
        axis.text.x        = element_text(size = rel(1))
    )
```

Example

```{r}
library(tidyverse)
library(ggsci)
data("mtcars")
yourplot <- mtcars %>%
    select(mpg, cyl, gear) %>%
    ggplot(., aes(x = mpg, y = cyl, fill = gear)) + 
    geom_point() +
    labs(title="Some Plot") 

yourplot + 
    amatheme + 
    # choose different color theme
    scale_color_npg() 

yourplot + 
    amatheme + 
    scale_color_continuous()
```

Other pre-specified themes

```{r}
library(ggthemes)


# Stata theme
yourplot +
    theme_stata()

# The economist theme
yourplot + 
    theme_economist()

yourplot + 
    theme_economist_white()

# Wall street journal theme
yourplot + 
    theme_wsj()

# APA theme
yourplot +
    jtools::theme_apa(
        legend.font.size = 24,
        x.font.size = 20,
        y.font.size = 20
    )
```
