# Bookdown cheat sheet

```{r}
# to see non-scientific notation a result
format(12e-17, scientific = FALSE)
```

## Operation

R commands to do derivatives of a defined function Taking derivatives in R involves using the `expression,` `D,` and `eval` functions. You wrap the function you want to take the derivative of in expression(), then use D, then eval as follows.

simple example

```{r}
#define a function
f=expression(sqrt(x))

#take the first derivative
df.dx=D(f,'x')
df.dx

#take the second derivative
d2f.dx2=D(D(f,'x'),'x')
d2f.dx2
```

Evaluate

-   The first argument passed to eval is the expression you want to evaluate
-   the second is a list containing the values of all quantities that are not defined elsewhere.

```{r}
#evaluate the function at a given x
eval(f,list(x=3))

#evaluate the first derivative at a given x
eval(df.dx,list(x=3))

#evaluate the second derivative at a given x
eval(d2f.dx2,list(x=3))
```

## Math Expression/ Syntax

[Full list](http://tug.ctan.org/info/symbols/comprehensive/symbols-a4.pdf)

Aligning equations

```{r eval=FALSE}
\begin{aligned}
a & = b \\
X &\sim {Norm}(10, 3) \\
5 & \le 10
\end{aligned}
```

$$
\begin{aligned}
a & = b \\
X &\sim {Norm}(10, 3) \\
5 & \le 10
\end{aligned}
$$

Cross-reference equation

```{r, eval = FALSE}
\begin{equation} 
a = b
(\#eq:test)
\end{equation}
```

```{=tex}
\begin{equation} 
a = b
(\#eq:test)
\end{equation}
```
to refer in a sentence \@ref(eq:test) (`\@ref(eq:test)`)

+-----------------------------------------------------------+----------------------------------------------------------+
| **Math** Syntax                                           | Notation                                                 |
+===========================================================+==========================================================+
| `$\pm$`                                                   | $\pm$                                                    |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\ge$`                                                   | $\ge$                                                    |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\le$`                                                   | $\le$                                                    |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\neq$`                                                  | $\neq$                                                   |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\equiv$`                                                | $\equiv$                                                 |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$^\circ$`                                                | $^\circ$                                                 |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\times$`                                                | $\times$                                                 |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\cdot$`                                                 | $\cdot$                                                  |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\leq$`                                                  | $\leq$                                                   |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\geq$`                                                  | $\geq$                                                   |
+-----------------------------------------------------------+----------------------------------------------------------+
| `\propto`                                                 | $\propto$                                                |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\subset$`                                               | $\subset$                                                |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\subseteq$`                                             | $\subseteq$                                              |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\leftarrow$`                                            | $\leftarrow$                                             |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\rightarrow$`                                           | $\rightarrow$                                            |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\Leftarrow$`                                            | $\Leftarrow$                                             |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\Rightarrow$`                                           | $\Rightarrow$                                            |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\approx$`                                               | $\approx$                                                |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\mathbb{R}$`                                            | $\mathbb{R}$                                             |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\sum_{n=1}^{10} n^2$`                                   | $\sum_{n=1}^{10} n^2$                                    |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$$\sum_{n=1}^{10} n^2$$`                                 | $$\sum_{n=1}^{10} n^2$$                                  |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$x^{n}$`                                                 | $x^{n}$                                                  |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$x_{n}$`                                                 | $x_{n}$                                                  |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\overline{x}$`                                          | $\overline{x}$                                           |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\hat{x}$`                                               | $\hat{x}$                                                |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\tilde{x}$`                                             | $\tilde{x}$                                              |
+-----------------------------------------------------------+----------------------------------------------------------+
| `\check{}`                                                | $\check{}$                                               |
+-----------------------------------------------------------+----------------------------------------------------------+
| `\underset{\gamma}{\operatorname{argmin}}`                | $\underset{\gamma}{\operatorname{argmin}}$               |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\frac{a}{b}$`                                           | $\frac{a}{b}$                                            |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\frac{a}{b}$`                                           | $\frac{a}{b}$                                            |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\displaystyle \frac{a}{b}$`                             | $\displaystyle \frac{a}{b}$                              |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\binom{n}{k}$`                                          | $\binom{n}{k}$                                           |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$x_{1} + x_{2} + \cdots + x_{n}$`                        | $x_{1} + x_{2} + \cdots + x_{n}$                         |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$x_{1}, x_{2}, \dots, x_{n}$`                            | $x_{1}, x_{2}, \dots, x_{n}$                             |
+-----------------------------------------------------------+----------------------------------------------------------+
| `\mathbf{x} = \langle x_{1}, x_{2}, \dots, x_{n}\rangle$` | $\mathbf{x} = \langle x_{1}, x_{2}, \dots, x_{n}\rangle$ |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$x \in A$`                                               | $x \in A$                                                |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$|A|$`                                                   | $|A|$                                                    |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$x \in A$`                                               | $x \in A$                                                |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$x \subset B$`                                           | $x \subset B$                                            |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$x \subseteq B$`                                         | $x \subseteq B$                                          |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$A \cup B$`                                              | $A \cup B$                                               |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$A \cap B$`                                              | $A \cap B$                                               |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$X \sim Binom(n, \pi)$`                                  | $X \sim Binom(n, \pi)$                                   |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\mathrm{P}(X \le x) = \text{pbinom}(x, n, \pi)$`        | $\mathrm{P}(X \le x) = \text{pbinom}(x, n, \pi)$         |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$P(A \mid B)$`                                           | $P(A \mid B)$                                            |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\mathrm{P}(A \mid B)$`                                  | $\mathrm{P}(A \mid B)$                                   |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\{1, 2, 3\}$`                                           | $\{1, 2, 3\}$                                            |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\sin(x)$`                                               | $\sin(x)$                                                |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\log(x)$`                                               | $\log(x)$                                                |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\int_{a}^{b}$`                                          | $\int_{a}^{b}$                                           |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\left(\int_{a}^{b} f(x) \; dx\right)$`                  | $\left(\int_{a}^{b} f(x) \; dx\right)$                   |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\left[\int_{\-infty}^{\infty} f(x) \; dx\right]$`       | $\left[\int_{-\infty}^{\infty} f(x) \; dx\right]$        |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\left. F(x) \right|_{a}^{b}$`                           | $\left. F(x) \right|_{a}^{b}$                            |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\sum_{x = a}^{b} f(x)$`                                 | $\sum_{x = a}^{b} f(x)$                                  |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\prod_{x = a}^{b} f(x)$`                                | $\prod_{x = a}^{b} f(x)$                                 |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\lim_{x \to \infty} f(x)$`                              | $\lim_{x \to \infty} f(x)$                               |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\displaystyle \lim_{x \to \infty} f(x)$`                | $\displaystyle \lim_{x \to \infty} f(x)$                 |
+-----------------------------------------------------------+----------------------------------------------------------+
| **Greek Letters**                                         |                                                          |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\alpha A$`                                              | $\alpha A$                                               |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\beta B$`                                               | $\beta B$                                                |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\gamma \Gamma$`                                         | $\gamma \Gamma$                                          |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\delta \Delta$`                                         | $\delta \Delta$                                          |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\epsilon \varepsilon E$`                                | $\epsilon \varepsilon E$                                 |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\zeta Z \sigma $`                                       | $\zeta Z \sigma$                                         |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\eta H$`                                                | $\eta H$                                                 |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\theta \vartheta \Theta$`                               | $\theta \vartheta \Theta$                                |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\iota I$`                                               | $\iota I$                                                |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\kappa K$`                                              | $\kappa K$                                               |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\lambda \Lambda$`                                       | $\lambda \Lambda$                                        |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\mu M$`                                                 | $\mu M$                                                  |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\nu N$`                                                 | $\nu N$                                                  |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\xi\Xi$`                                                | $\xi\Xi$                                                 |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$o O$`                                                   | $o O$                                                    |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\pi \Pi$`                                               | $\pi \Pi$                                                |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\rho\varrho P$`                                         | $\rho\varrho P$                                          |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\sigma \Sigma$`                                         | $\sigma \Sigma$                                          |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\tau T$`                                                | $\tau T$                                                 |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\upsilon \Upsilon$`                                     | $\upsilon \Upsilon$                                      |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\phi \varphi \Phi$`                                     | $\phi \varphi \Phi$                                      |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\chi X$`                                                | $\chi X$                                                 |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\psi \Psi$`                                             | $\psi \Psi$                                              |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\omega \Omega$`                                         | $\omega \Omega$                                          |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\cdot$`                                                 | $\cdot$                                                  |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\cdots$`                                                | $\cdots$                                                 |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\ddots$`                                                | $\ddots$                                                 |
+-----------------------------------------------------------+----------------------------------------------------------+
| `$\ldots$`                                                | $\ldots$                                                 |
+-----------------------------------------------------------+----------------------------------------------------------+

Limit `P(\lim_{n\to \infty}\bar{X}_n =\mu) =1`

$$
P(\lim_{n\to \infty}\bar{X}_n =\mu) =1
$$

Matrices

```{r eval = FALSE}
$$\begin{array}
{rrr}
1 & 2 & 3 \\
4 & 5 & 6 \\
7 & 8 & 9
\end{array}
$$
```

$$
\begin{array}
{rrr}
1 & 2 & 3 \\
4 & 5 & 6 \\
7 & 8 & 9
\end{array}
$$

```{r eval=FALSE}
$$\mathbf{X} = \left[\begin{array}
{rrr}
1 & 2 & 3 \\
4 & 5 & 6 \\
7 & 8 & 9
\end{array}\right]
$$
```

$$
\mathbf{X} = \left[\begin{array}
{rrr}
1 & 2 & 3 \\
4 & 5 & 6 \\
7 & 8 & 9
\end{array}\right]
$$

Aligning Equations

Aligning Equations with Comments

```{r eval=FALSE}
\begin{aligned}
    3+x &=4 && \text{(Solve for} x \text{.)}\\
    x &=4-3 && \text{(Subtract 3 from both sides.)}\\
    x &=1   && \text{(Yielding the solution.)}
\end{aligned}
```

$$
\begin{aligned}
    3+x &=4 & &\text{(Solve for} x \text{.)} \\
    x &=4-3 && \text{(Subtract 3 from both sides.)} \\
    x &=1   && \text{(Yielding the solution.)}
\end{aligned}
$$

### Statistics Notation

```{r eval = FALSE}
$$
f(y|N,p) = \frac{N!}{y!(N-y)!}\cdot p^y \cdot (1-p)^{N-y} = {{N}\choose{y}} \cdot p^y \cdot (1-p)^{N-y}
$$
```

$$
f(y|N,p) = \frac{N!}{y!(N-y)!}\cdot p^y \cdot (1-p)^{N-y} = {{N}\choose{y}} \cdot p^y \cdot (1-p)^{N-y}
$$

```{r eval=FALSE}
\begin{cases}
\frac{1}{b-a}&\text{for $x\in[a,b]$}\\
0&\text{otherwise}\\
\end{cases}
```

$$
\begin{cases}
\frac{1}{b-a} & \text{for } x\in[a,b]\\
0 & \text{otherwise}\\
\end{cases}
$$

## Table

```{r eval = FALSE}
+---------------+---------------+--------------------+
| Fruit         | Price         | Advantages         |
+===============+===============+====================+
| *Bananas*     | $1.34         | - built-in wrapper |
|               |               | - bright color     |
+---------------+---------------+--------------------+
| Oranges       | $2.10         | - cures scurvy     |
|               |               | - **tasty**        |
+---------------+---------------+--------------------+
```

+--------------+--------------+----------------------+
| Fruit        | Price        | Advantages           |
+==============+==============+======================+
| *Bananas*    | \$1.34       | -   built-in wrapper |
|              |              | -   bright color     |
+--------------+--------------+----------------------+
| Oranges      | \$2.10       | -   cures scurvy     |
|              |              | -   **tasty**        |
+--------------+--------------+----------------------+

```{r eval=FALSE}
(\mathbf{x}^T\mathbf{x})^{-1}\mathbf{x}^T\mathbf{y}

```

$(\mathbf{x}^T\mathbf{x})^{-1}\mathbf{x}^T\mathbf{y}$
