# Exploratory Data Analysis

This section demonstrates a fast, business-oriented EDA workflow on `ggplot2::txhousing`. We'll generate automated reports, engineer interpretable features for marketing/finance contexts, profile missingness, identify errors/outliers, and provide quick interactive tools your stakeholders will actually use. Where possible, we keep everything reproducible in code and easy to extend for your own datasets.

```{r setup, message = FALSE, warning = FALSE}
# Core packages
# install.packages(c("tidyverse","lubridate","DataExplorer","dataReporter","skimr","rpivotTable","esquisse","dlookr"))
library(ggplot2)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(DataExplorer)
library(skimr)
library(dlookr)

# Data: ggplot2 ships with txhousing
data("txhousing", package = "ggplot2")

# Create a clean working copy with business-friendly engineered features
tx <- txhousing %>%
  mutate(
    date = make_date(year, month, 1),
    # Average price per sale (proxy for ARPU-like metric in sales)
    avg_price = ifelse(sales > 0, volume / sales, NA_real_),
    # Absorption-like metric: sales per listing (how fast inventory turns)
    absorption = ifelse(listings > 0, sales / listings, NA_real_),
    # Months of inventory is already provided as `inventory`
    # Seasonal and calendar features
    quarter = factor(quarter(date), labels = c("Q1", "Q2", "Q3", "Q4")),
    ym = factor(format(date, "%Y-%m")),
    city = as.factor(city)
  )

# A couple of quick sanity checks
stopifnot(all(
    c(
        "city",
        "year",
        "month",
        "sales",
        "volume",
        "median",
        "listings",
        "inventory"
    ) %in% names(tx)
))

# Preview
dplyr::glimpse(tx)
```

------------------------------------------------------------------------

## Data Report

Automate a first-pass profile to see dimensions, types, basic completeness, and distributions.

```{r, message = FALSE, warning = FALSE, results = "hide"}
# High-level overview
introduce(tx)
```

```{r, fig.width=8, fig.height=5}
# Structural intro plots
plot_intro(tx)
```

```{r, fig.width=10, fig.height=6}
# Quick distribution sweep for numerics (great to spot skew and heavy tails)
plot_histogram(tx, ncol = 3L)
```

```{r, fig.width=9, fig.height=6}
# Pairwise correlations (continuous); useful for multicollinearity hints
# Note: if your dataset has many numerics, set cor_args = list(use = "pairwise.complete.obs")
plot_correlation(tx, type = "continuous")
```

```{r, fig.width=9, fig.height=6}
# If you want correlations/associations among discrete variables:
# (city and quarter here; may be high-cardinality)
plot_correlation(tx %>% select(city, quarter), type = "discrete")
```

```{r, fig.width=9, fig.height=6}
# Quick PCA map for continuous variables (after standardizing)
DataExplorer::plot_prcomp(tx %>%
  select(where(is.numeric)) %>%
  na.omit()
)
```

*Tip:* For very wide tables, use `maxcat` in `DataExplorer` functions to cap cardinality for cleaner plots.

------------------------------------------------------------------------

## Feature Engineering

Business-minded transforms often beat fancy models. The goal is interpretability and better signal-to-noise.

```{r}
fe_tx <- tx %>%
  mutate(
    # Year-month numeric for time-aware models
    t_index = as.integer(zoo::as.yearmon(date)),
    # Price momentum: rolling median (by city) as a simple trend proxy
    median_roll3 = dplyr::lag(slider::slide_dbl(median, mean, .before = 2, .complete = TRUE), 0), 
    # Volume-per-listing: marketing pipeline proxy (dollars per active listing)
    vol_per_listing = ifelse(listings > 0, volume / listings, NA_real_),
    # Seasonality flags
    is_peak_summer = month(date) %in% 6:8,
    is_december = month(date) == 12
  )
head(fe_tx)
```

One-hot encode high-utility discrete features (keep top cities to avoid huge design matrices)

**Note**: `dummify()` will create binary indicators and drop the original variable by default.

To avoid exploding dimensions, we lump rare cities first.

```{r, message=FALSE, warning=FALSE}
fe_tx_small <- fe_tx %>%
    mutate(city_lumped = forcats::fct_lump_n(city, n = 10, other_level = "Other"))
tx_dummy <-
    dummify(fe_tx_small %>% select(-city), select = "city_lumped")
names(tx_dummy)[1:15]
```

```{r}
# Separate continuous and discrete columns for targeted transformations/modeling
spl <- split_columns(tx)
names(spl)
```

*Rules of thumb:*

-   Normalize skewed monetary variables (e.g., $\log(\text{volume}+1)$) before linear modeling.
-   Group rare categories and avoid too many dummies relative to $n$.
-   Create business KPIs: `avg_price`, `absorption`, `vol_per_listing`, and simple momentum features.

------------------------------------------------------------------------

## Missing Data

Missingness is information. First detect, then decide: delete, impute, or model it explicitly.

```{r}
# Counts by variable
profile_missing(tx)
```

```{r, fig.width=9, fig.height=5}
plot_missing(tx)
```

```{r}
# Simple tidy summary (counts and proportions)
tx %>%
    summarise(across(everything(),
                     ~ sprintf(
                         "%d (%.1f%%)", sum(is.na(.)), 100 * mean(is.na(.))
                     ),
                     .names = "{.col}")) %>%
    pivot_longer(everything(), names_to = "variable", values_to = "n_pct_na") %>%
    arrange(desc(n_pct_na)) %>%
    head(12)
```

```{r}
# Example: median imputation for a few numerics (for demonstration only)
# (In modeling, prefer imputations *within* resampling using recipes/caret/tidymodels.)
tx_imputed <- tx %>%
  mutate(
    sales     = ifelse(is.na(sales),     median(sales,     na.rm = TRUE), sales),
    listings  = ifelse(is.na(listings),  median(listings,  na.rm = TRUE), listings),
    inventory = ifelse(is.na(inventory), median(inventory, na.rm = TRUE), inventory),
    median    = ifelse(is.na(median),    median(median,    na.rm = TRUE), median)
  )
skimr::skim(tx_imputed %>% select(sales, listings, inventory, median))
```

```{r, message=FALSE}
# dlookr helpers (diagnose and visualize NA concentration)
diagnose(tx)
plot_na_pareto(tx)
```

*Best practice:* Impute inside a modeling pipeline (e.g., `recipes::step_impute_median()`), and consider adding "was missing" flags to retain signal from missingness patterns.

------------------------------------------------------------------------

## Error Identification

Catch anomalies early: impossible values, data entry glitches, or KPI mismatches.

1.  Logical constraints

```{r}

violations <- tx %>%
  mutate(
    v_sales_negative   = sales < 0,
    v_volume_negative  = volume < 0,
    v_listings_negative= listings < 0,
    v_inventory_negative = inventory < 0
  ) %>%
  summarise(across(starts_with("v_"), ~sum(.x, na.rm = TRUE)))
violations
```

2.  Price coherence: median vs. average price (avg_price = volume/sales)

Large gaps are not necessarily errors, but extreme ratios can flag data issues.

```{r}
coherence <- tx %>%
  mutate(avg_price = ifelse(sales > 0, volume / sales, NA_real_),
         ratio = median / avg_price) %>%
  filter(!is.na(ratio), is.finite(ratio)) %>%
  summarise(
    n = n(),
    ratio_p01 = quantile(ratio, 0.01),
    ratio_p99 = quantile(ratio, 0.99),
    extreme = sum(ratio < 0.25 | ratio > 4, na.rm = TRUE)
  )
coherence
```

3.  Outliers (univariate) with `dlookr`

```{r}
out_summary <- diagnose_outlier(tx %>% select(sales, volume, median, listings, inventory, avg_price))
out_summary %>% arrange(desc(outliers_cnt)) %>% head(10)
```

```{r, eval = FALSE}
# Full, human-readable data quality report (HTML) with dataReporter
# install.packages("dataReporter")
library(dataReporter)
makeDataReport(tx)  # Generates an HTML report in your working directory
```

------------------------------------------------------------------------

## Summary statistics

Concise summaries for continuous and categorical variables, plus grouped business KPIs.

```{r}
skim(tx)
```

```{r}
# City-level yearly KPIs
city_year <- tx %>%
    group_by(city, year) %>%
    summarise(
        n_months      = n(),
        sales_total   = sum(sales, na.rm = TRUE),
        volume_total  = sum(volume, na.rm = TRUE),
        median_price  = median(median, na.rm = TRUE),
        avg_inventory = mean(inventory, na.rm = TRUE),
        .groups = "drop"
    ) %>%
    arrange(city, year)
head(city_year, 10)
```

```{r, fig.width=9, fig.height=5}
# Example visual: price over time for a few major cities
top_cities <- tx %>%
    count(city, sort = TRUE) %>%
    slice_head(n = 6) %>%
    pull(city)

tx %>%
    filter(city %in% top_cities) %>%
    ggplot(aes(date, median, color = city)) +
    geom_line(linewidth = 0.7) +
    scale_y_continuous(labels = scales::dollar_format()) +
    labs(
        title = "Median price over time",
        x = NULL,
        y = "Median price",
        color = "City"
    ) +
    theme_minimal(base_size = 12)
```

```{r, fig.width=8, fig.height=5}
# Volume vs. Sales (log-scale to reduce skew)
tx %>%
    filter(sales > 0, volume > 0) %>%
    ggplot(aes(sales, volume)) +
    geom_point(alpha = 0.25) +
    scale_x_log10() +
    scale_y_log10(labels = scales::dollar_format()) +
    geom_smooth(method = "lm",
                se = FALSE,
                color = "steelblue") +
    labs(title = "Sales vs. Volume (log-log)",
         x = "Sales (transactions)",
         y = "Volume (USD)") +
    theme_minimal(base_size = 12)
```

------------------------------------------------------------------------

## Not so code-y process

### Quick and dirty way to look at your data

```{r, eval = FALSE}
# install.packages("rpivotTable")
library(rpivotTable)

# Works best in HTML output; drag-and-drop just like Excel PivotTables
rpivotTable(
  data = tx,
  rows = "city",
  cols = "year",
  vals = "sales",
  aggregatorName = "Sum",
  rendererName = "Heatmap"
)
```

### Code generation and wrangling (visual)

```{r, eval = FALSE}
# install.packages("esquisse")
library(esquisse)
# Launch the GUI to build ggplots and dplyr pipelines interactively
esquisse::esquisser(tx)
```

------------------------------------------------------------------------

## Shiny-app based Tableau style

```{r, eval = FALSE}
# Minimal Shiny-style dashboard idea (skeleton)
# shinyApp(
#   ui = fluidPage(
#     titlePanel("TX Housing Explorer"),
#     sidebarLayout(
#       sidebarPanel(selectInput("city", "City", choices = sort(unique(tx$city)), selected = "Houston")),
#       mainPanel(plotOutput("trend"))
#     )
#   ),
#   server = function(input, output, session) {
#     output$trend <- renderPlot({
#       tx %>%
#         filter(city == input$city) %>%
#         ggplot(aes(date, median)) + geom_line() +
#         labs(x = NULL, y = "Median Price") +
#         theme_minimal()
#     })
#   }
# )
```

------------------------------------------------------------------------

## Customize your daily/automatic report

```{r, eval = FALSE}
# Option A: One-liner automated HTML report (DataExplorer)
DataExplorer::create_report(tx, y = "median")  # sets target variable for focus

# Option B: Knit this Rmd on a schedule (cron/Task Scheduler)
# rmarkdown::render("eda_txhousing.Rmd", output_format = "html_document")
```

```{r, eval = FALSE}
# Optionally explore additional EDA helpers
# install.packages(c("chronicle","descriptr"))
library(chronicle)
library(descriptr)
# Tip: Use these packages selectively; their APIs evolve.
# Prefer stable verbs above; consult vignettes for current functions.
```

------------------------------------------------------------------------

### Appendix: Small "gotchas" to keep in mind

-   For correlation heatmaps, ensure you're not mixing units without scaling; use `scale. = TRUE` in PCA.
-   For discrete correlations, high-cardinality features (like many cities) can make matrices unwieldy. Thus, lump or select top-$k$ first.
-   Imputation should be inside resampling to avoid leakage.
-   KPI checks (e.g., `median` vs `volume/sales`) are *diagnostics*, not proofs of error. Hence, investigate outliers before removal.
