# Sensitivity Analysis and Robustness Checks

> "Do robustness checks until you believe the result."
>
> --- the author, having repeatedly experienced the brief joy of promising results only to watch them collapse under closer inspection, and now treating every finding as too good to be true until proven otherwise.

Sensitivity analysis and robustness checks are critical components of rigorous empirical research. They help researchers assess the stability of their findings across different model specifications, evaluate the potential impact of omitted variable bias, and provide evidence for or against causal interpretations. This chapter provides guidance on implementing various sensitivity analysis techniques, ranging from specification curve analysis to sophisticated methods for quantifying omitted variable bias.

## The Philosophy of Robustness

Before diving into specific techniques, it's worth understanding why robustness checks matter. In empirical research, we rarely have perfect certainty about the "correct" specification. Our choices about which controls to include, how to measure variables, which functional forms to use, and how to address endogeneity can all influence our results. Sensitivity analysis helps us understand whether our main conclusions depend critically on these choices, or whether they remain stable across reasonable alternatives.

A robust finding is one that persists across multiple plausible specifications. This doesn't necessarily mean the coefficient estimate must be identical across all specifications, some variation is expected and often informative. Rather, robustness means that the key substantive conclusion (e.g., the sign, statistical significance, or economic magnitude of an effect) remains consistent despite reasonable variation in modeling choices.

## Specification Curve Analysis {#sec-specification-curve-analysis}

Specification curve analysis (also known as multiverse analysis or the specification robustness graph) provides a systematic way to examine how results vary across a large set of defensible specifications. Rather than reporting a single "preferred" specification, this approach acknowledges that multiple specifications may be equally justifiable and examines the distribution of estimates across all of them.

### Conceptual Foundation

The specification curve approach was formalized by @simonsohn2020specification, though similar ideas have appeared under various names in the literature. The key insight is that researchers make many decisions when specifying a model: which controls to include, which fixed effects to add, how to cluster standard errors, etc. And these decisions can be viewed as creating a "multiverse" of possible specifications. By systematically varying these choices and examining the resulting distribution of estimates, we can assess whether our main conclusion depends on arbitrary specification choices.

A specification curve typically consists of two panels:

1.  **The coefficient panel**: Shows the point estimate and confidence interval for the main variable of interest across all specifications, typically sorted by coefficient magnitude
2.  **The specification panel**: Shows which modeling choices were made in each specification (e.g., which controls were included, which fixed effects were used)

This visualization makes it easy to see whether results are driven by particular specification choices and whether the finding is robust across the specification space.

### The `starbility` Package

The `starbility` package provides a flexible and user-friendly implementation of specification curve analysis. It works seamlessly with various model types and allows for sophisticated customization.

#### Installation and Setup

```{r, eval = FALSE}
# Install from GitHub
# devtools::install_github('https://github.com/AakaashRao/starbility')
library(starbility)

# Load other required packages
library(tidyverse)  # For data manipulation and visualization
library(lfe)        # For fixed effects models
library(broom)      # For tidying model output
library(cowplot)    # For combining plots
```

#### Basic Specification Curve with Multiple Controls

Let's start with an example using the diamonds dataset. This example demonstrates how to systematically vary control variables and visualize the resulting specification curve.

```{r}
library(tidyverse)
library(starbility)
library(lfe)

# Load and prepare data
data("diamonds")
set.seed(43)  # For reproducibility

# Create a subset for computational efficiency
# In practice, you'd use your full dataset
indices = sample(1:nrow(diamonds),
                 replace = FALSE,
                 size = round(nrow(diamonds) / 20))
diamonds = diamonds[indices, ]

# Create additional variables for demonstration
diamonds$high_clarity = diamonds$clarity %in% c('VS1','VVS2','VVS1','IF')
diamonds$log_price = log(diamonds$price)
diamonds$log_carat = log(diamonds$carat)
```

Now let's define our specification universe. The key is to think carefully about which specification choices are defensible and should be explored:

```{r}
# Base controls: These are included in ALL specifications
# Use this for controls that you believe should always be included based on theory
base_controls = c(
  'Diamond dimensions' = 'x + y + z'  # Physical dimensions
)

# Permutable controls: These will be included in all possible combinations
# These are controls where theory doesn't give clear guidance on inclusion
perm_controls = c(
  'Depth' = 'depth',
  'Table width' = 'table'
)

# Permutable fixed effects: Different types of fixed effects to explore
# Useful when you have multiple ways to control for unobserved heterogeneity
perm_fe_controls = c(
  'Cut FE' = 'cut',
  'Color FE' = 'color'
)

# Non-permutable fixed effects: Alternative specifications (only one included at a time)
# Use this when you have mutually exclusive ways of controlling for something
nonperm_fe_controls = c(
  'Clarity FE (granular)' = 'clarity',
  'Clarity FE (binary)' = 'high_clarity'
)

# If you want to explore instrumental variables specifications
instruments = 'x + y + z'

# Add sample weights for robustness
diamonds$sample_weights = runif(n = nrow(diamonds))
```

#### Custom Model Functions

One of the most powerful features of `starbility` is the ability to use custom model functions. This allows you to implement your preferred estimation approach, including custom standard errors, specific inference procedures, or alternative confidence intervals.

```{r}
# Custom function for felm models with robust standard errors
# This function must return a vector of: c(coefficient, p-value, upper CI, lower CI)
starb_felm_custom = function(spec, data, rhs, ...) {
  # Convert specification string to formula
  spec = as.formula(spec)
  
  # Estimate the model using lfe::felm
  # felm is particularly useful for models with multiple fixed effects
  model = lfe::felm(spec, data = data) %>% 
    broom::tidy()  # Convert to tidy format

  # Extract results for the variable of interest (rhs)
  row  = which(model$term == rhs)
  coef = model[row, 'estimate'] %>% as.numeric()
  se   = model[row, 'std.error'] %>% as.numeric()
  p    = model[row, 'p.value'] %>% as.numeric()
  
  # Calculate confidence intervals
  # Here we use 99% CI for more conservative inference
  z = qnorm(0.995)  # 99% confidence level
  upper_ci = coef + z * se
  lower_ci = coef - z * se
  
  # For one-tailed tests, divide p-value by 2
  # Remove this if you want two-tailed p-values
  p_onetailed = p / 2
  
  return(c(coef, p_onetailed, upper_ci, lower_ci))
}

# Alternative: Custom function with heteroskedasticity-robust SEs
starb_lm_robust = function(spec, data, rhs, ...) {
  spec = as.formula(spec)
  
  # Estimate with HC3 robust standard errors
  model = lm(spec, data = data)
  robust_se = sandwich::vcovHC(model, type = "HC3")
  
  # Use lmtest for robust inference
  coef_test = lmtest::coeftest(model, vcov = robust_se)
  
  row  = which(rownames(coef_test) == rhs)
  coef = coef_test[row, 'Estimate']
  se   = coef_test[row, 'Std. Error']
  p    = coef_test[row, 'Pr(>|t|)']
  
  z = qnorm(0.975)  # 95% confidence level
  return(c(coef, p, coef + z * se, coef - z * se))
}
```

#### Creating the Specification Curve

Now we can generate our specification curve with all the bells and whistles (Figure \@ref(fig:robustness-specification-curve))

```{r}
# Generate specification curve
# This will create plots showing how the coefficient varies across specifications
plots = stability_plot(
    data = diamonds,
    lhs = 'price',           # Dependent variable
    rhs = 'carat',           # Main independent variable of interest
    model = starb_felm_custom,  # Use our custom model function
    
    # Clustering and weights
    cluster = 'cut',            # Cluster standard errors by cut
    weights = 'sample_weights', # Use sample weights
    
    # Control variable specifications
    base = base_controls,       # Always included
    perm = perm_controls,       # All combinations
    perm_fe = perm_fe_controls, # All combinations of these FEs
    
    # Alternative: Use non-permutable FE (only one at a time)
    # nonperm_fe = nonperm_fe_controls,
    # fe_always = FALSE,  # Set to FALSE to include specs without any FEs
    
    # Instrumental variables (if needed)
    # iv = instruments,
    
    # Sorting and display options
    sort = "asc-by-fe",  # Options: "asc", "desc", "asc-by-fe", "desc-by-fe"
    
    # Visual customization
    error_geom = 'ribbon',  # Display error bands as ribbons (alternatives: 'linerange', 'none')
    # error_alpha = 0.2,    # Transparency of error bands
    # point_size = 1.5,     # Size of coefficient points
    # control_text_size = 10,  # Size of control labels
    
    # For datasets with fewer specifications, you might want:
    # control_geom = 'circle',  # Use circles instead of rectangles
    # point_size = 2,
    # control_spacing = 0.3,
    
    # Customize the y-axis range if needed
    # coef_ylim = c(-5000, 35000),
    
    # Adjust spacing between panels
    # trip_top = 3,
    
    # Relative height of coefficient panel vs control panel
    rel_height = 0.6
)
```

```{r robustness-specification-curve, fig.cap="Specification curve for the effect of carat on price.", fig.alt="Specification curve: top panel shows point estimates and 95 percent confidence ribbons across model specifications; bottom panel shows a binary grid indicating included controls and fixed effects for each specification."}
# Display the plots
plots
```

The specification curve uses color coding to indicate statistical significance:

-   **Red**: $p < 0.01$ (highly significant)
-   **Green**: $p < 0.05$ (significant)
-   **Blue**: $p < 0.1$ (marginally significant)
-   **Black**: $p > 0.1$ (not significant)

This color scheme makes it easy to see at a glance whether your finding is robust across specifications, or whether significance depends on particular specification choices.

### Advanced Specification Curve Techniques

#### Step-by-Step Control of the Process

For maximum flexibility, `starbility` allows you to control each step of the specification curve generation process. This is useful when you want to modify the grid of specifications, use custom model functions, or create highly customized visualizations.

```{r}
# Ensure high_clarity variable exists
diamonds$high_clarity = diamonds$clarity %in% c('VS1','VVS2','VVS1','IF')

# Redefine controls for this example
base_controls = c(
  'Diamond dimensions' = 'x + y + z'
)

perm_controls = c(
  'Depth' = 'depth',
  'Table width' = 'table'
)

perm_fe_controls = c(
  'Cut FE' = 'cut',
  'Color FE' = 'color'
)

nonperm_fe_controls = c(
  'Clarity FE (granular)' = 'clarity',
  'Clarity FE (binary)' = 'high_clarity'
)

# Step 1: Create the control grid
# This generates all possible combinations of controls
grid1 = stability_plot(
  data = diamonds, 
  lhs = 'price', 
  rhs = 'carat', 
  perm = perm_controls,
  base = base_controls, 
  perm_fe = perm_fe_controls, 
  nonperm_fe = nonperm_fe_controls, 
  run_to = 2  # Stop after creating the grid
)

# Examine the grid structure
head(grid1, 10)
# Each row represents a different specification
# Columns indicate which controls/FEs are included (1 = yes, 0 = no)

# Step 2: Generate model expressions
# This creates the actual formula for each specification
grid2 = stability_plot(
  grid = grid1,  # Use the grid from step 1
  data = diamonds, 
  lhs = 'price', 
  rhs = 'carat', 
  perm = perm_controls, 
  base = base_controls,
  run_from = 2,  # Start from step 2
  run_to = 3     # Stop after generating expressions
)

# View the formulas
head(grid2, 10)
# Now each row has an 'expr' column with the full model formula

# Step 3: Estimate all models
# This runs the actual regressions
grid3 = stability_plot(
  grid = grid2,
  data = diamonds, 
  lhs = 'price', 
  rhs = 'carat', 
  perm = perm_controls, 
  base = base_controls,
  run_from = 3,
  run_to = 4
)

# View estimation results
head(grid3, 10)
# Now includes coefficient estimates, p-values, and confidence intervals

# Step 4: Prepare data for plotting
# This creates the two dataframes needed for visualization
dfs = stability_plot(
  grid = grid3,
  data = diamonds, 
  lhs = 'price', 
  rhs = 'carat', 
  perm = perm_controls, 
  base = base_controls,
  run_from = 4,
  run_to = 5
)

coef_grid = dfs[[1]]      # Data for coefficient panel
control_grid = dfs[[2]]   # Data for control specification panel

head(coef_grid, 10)

# Step 5: Create the plot panels
# This generates the two ggplot objects
panels = stability_plot(
  data = diamonds, 
  lhs = 'price', 
  rhs = 'carat', 
  coef_grid = coef_grid,
  control_grid = control_grid,
  run_from = 5,
  run_to = 6
)

# Step 6: Combine and display
# Final step to create the complete visualization
final_plot = stability_plot(
  data = diamonds,
  lhs = 'price', 
  rhs = 'carat', 
  coef_panel = panels[[1]],
  control_panel = panels[[2]],
  run_from = 6,
  run_to = 7
)
```

#### Specification Curves for Non-Linear Models

Specification curve analysis is not limited to linear models. Here's how to implement it with logistic regression (Figure \@ref(fig:robustness-logit-curve)).

```{r}
# Create binary outcome variable
diamonds$above_med_price = as.numeric(diamonds$price > median(diamonds$price))

# Add sample weights for logit model
diamonds$weight = runif(nrow(diamonds))

# Define controls
base_controls = c('Diamond dimensions' = 'x + y + z')

perm_controls = c(
  'Depth' = 'depth',
  'Table width' = 'table',
  'Clarity' = 'clarity'
)

lhs_var = 'above_med_price'
rhs_var = 'carat'

# Step 1: Create initial grid
grid1 = stability_plot(
    data = diamonds,
    lhs = lhs_var,
    rhs = rhs_var,
    perm = perm_controls,
    base = base_controls,
    fe_always = FALSE,  # Include specifications without FEs
    run_to = 2
)

# Step 2: Manually create formulas for logit model
# The starbility package creates expressions for lm/felm by default
# For glm, we need to create our own formula structure
base_perm = c(base_controls, perm_controls)

# Create control part of formula
grid1$expr = apply(
  grid1[, 1:length(base_perm)], 
  1,
  function(x) {
    paste(
      base_perm[names(base_perm)[which(x == 1)]], 
      collapse = '+'
    )
  }
)

# Complete formula with LHS and RHS variables
grid1$expr = paste(lhs_var, '~', rhs_var, '+', grid1$expr, sep = '')

head(grid1, 10)

# Step 3: Create custom logit estimation function
# This function estimates a logistic regression and extracts results
starb_logit = function(spec, data, rhs, ...) {
  spec = as.formula(spec)
  
  # Estimate logit model with weights, suppressing separation warning
  model = suppressWarnings(
    glm(
      spec, 
      data = data, 
      family = 'binomial',
      weights = data$weight
    )
  ) %>%
    broom::tidy()
  
  # Extract coefficient for variable of interest
  row  = which(model$term == rhs)
  coef = model[row, 'estimate'] %>% as.numeric()
  se   = model[row, 'std.error'] %>% as.numeric()
  p    = model[row, 'p.value'] %>% as.numeric()

  # Return coefficient, p-value, and 95% CI bounds
  return(c(coef, p, coef + 1.96*se, coef - 1.96*se))
}

# Generate specification curve for logit model
logit_curve = stability_plot(
  grid = grid1,
  data = diamonds, 
  lhs = lhs_var, 
  rhs = rhs_var,
  model = starb_logit,  # Use our custom logit function
  perm = perm_controls,
  base = base_controls,
  fe_always = FALSE,
  run_from = 3  # Start from estimation step
)
```

```{r robustness-logit-curve, fig.cap="Specification curve (logit specifications).", fig.alt="Coefficient plot showing the estimated effects of various diamond characteristics (depth, table width, and clarity) on diamond price. Points represent coefficient estimates with error bars indicating 95% confidence intervals. Different colors distinguish between different levels or interactions of the variables. The y-axis shows coefficient values ranging from approximately -5 to 50, while the x-axis displays different combinations of diamond dimension variables."}
logit_curve
```

#### Marginal Effects for Non-Linear Models

For non-linear models like logit or probit, we often want to report marginal effects (average marginal effects, AME) rather than raw coefficients, as they're more interpretable. Here's how to incorporate marginal effects into specification curves (Figure \@ref(fig:robustness-ame-curve)).

```{r}
library(margins)  # For calculating marginal effects

# Enhanced logit function with marginal effects option
starb_logit_enhanced = function(spec, data, rhs, ...) {
  # Extract additional arguments
  l = list(...)
  get_mfx = ifelse(is.null(l$get_mfx), FALSE, TRUE)  # Default to FALSE
  
  spec = as.formula(spec)
  
  if (get_mfx) {
    # Calculate average marginal effects
    model = suppressWarnings(
      glm(
        spec, 
        data = data, 
        family = 'binomial',
        weights = data$weight
      )
    ) %>%
      margins() %>%  # Calculate marginal effects
      summary()
    
    # Extract AME results
    row = which(model$factor == rhs)
    coef = model[row, 'AME'] %>% as.numeric()  # Average Marginal Effect
    se   = model[row, 'SE'] %>% as.numeric()
    p    = model[row, 'p'] %>% as.numeric()
  } else {
    # Return raw coefficients (log-odds)
    model = suppressWarnings(
      glm(
        spec, 
        data = data, 
        family = 'binomial',
        weights = data$weight
      )
    ) %>%
      broom::tidy()
    
    row = which(model$term == rhs)
    coef = model[row, 'estimate'] %>% as.numeric()
    se   = model[row, 'std.error'] %>% as.numeric()
    p    = model[row, 'p.value'] %>% as.numeric()
  }

  # Use 99% confidence intervals for more conservative inference
  z = qnorm(0.995)
  return(c(coef, p, coef + z*se, coef - z*se))
}

# Generate specification curve with marginal effects
ame_curve = stability_plot(
  grid = grid1,
  data = diamonds, 
  lhs = lhs_var, 
  rhs = rhs_var,
  model = starb_logit_enhanced,
  get_mfx = TRUE,  # Request marginal effects
  perm = perm_controls,
  base = base_controls,
  fe_always = FALSE,
  run_from = 3
)
```

```{r robustness-ame-curve, fig.cap='Robustness of diamond-dimension effects measured as average marginal effects.', fig.alt='Specification curve displaying average marginal effects of diamond characteristics including depth, table width, and clarity on diamond price across different model specifications. Points represent marginal effect estimates with error bars indicating confidence intervals. Different colors distinguish between different levels or interactions of the variables. The y-axis shows marginal effect values ranging from approximately negative 0.5 to 1.5, while the x-axis displays different combinations of diamond dimension variables tested across specifications.'}
ame_curve
```

#### Fully Customized Specification Curve Visualizations

When you need complete control over the appearance of your specification curve, you can extract the underlying data and create custom ggplot visualizations (Figure \@ref(fig:fully-customized-spec-curve)).

```{r fully-customized-spec-curve, fig.cap='Specification curve for the effect of carat on above-median diamond price.', fig.alt='Specification curve displaying the effect of diamond carat weight on the probability of being priced above the median across eight different model specifications. The top panel shows average marginal effect estimates represented by different colored shapes including triangles, diamonds, circles, and squares, with error bars indicating confidence intervals. The y-axis ranges from 0 to 1.5, with a dotted horizontal line at zero. The bottom panel uses a matrix of filled and unfilled diamonds to indicate which control variables including diamond dimensions, depth, table width, and clarity are included in each of the eight specifications numbered along the x-axis.'}
# Extract data for custom plotting
dfs = stability_plot(
  grid = grid1,
  data = diamonds, 
  lhs = lhs_var, 
  rhs = rhs_var,
  model = starb_logit_enhanced,
  get_mfx = TRUE,
  perm = perm_controls,
  base = base_controls,
  fe_always = FALSE,
  run_from = 3,
  run_to = 5  # Stop before plotting
)

coef_grid_logit = dfs[[1]]
control_grid_logit = dfs[[2]]

# Define plot parameters
min_space = 0.5  # Space at edges of plot

# Create highly customized coefficient plot
coef_plot = ggplot2::ggplot(
  coef_grid_logit, 
  aes(
    x = model,
    y = coef,
    shape = p,
    group = p
  )
) +
  # Add confidence interval ribbons
  geom_linerange(
    aes(ymin = error_low, ymax = error_high),
    alpha = 0.75,
    size = 0.8
  ) +
  
  # Add coefficient points with color and shape by significance
  geom_point(
    size = 5,
    aes(col = p, fill = p),
    alpha = 1
  ) +
  
  # Use viridis color palette (colorblind-friendly)
  viridis::scale_color_viridis(
    discrete = TRUE,
    option = "D",
    name = "P-value"
  ) +
  
  # Different shapes for different significance levels
  scale_shape_manual(
    values = c(15, 17, 18, 19),
    name = "P-value"
  ) +
  
  # Reference line at zero
  geom_hline(
    yintercept = 0,
    linetype = 'dotted',
    color = 'red',
    size = 0.5
  ) +
  
  # Styling
  theme_classic() +
  theme(
    axis.text.x = element_blank(),
    axis.title = element_blank(),
    axis.ticks.x = element_blank(),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11)
  ) +
  
  # Set axis limits
  coord_cartesian(
    xlim = c(1 - min_space, max(coef_grid_logit$model) + min_space),
    ylim = c(-0.1, 1.6)
  ) +
  
  # Remove redundant legends
  guides(fill = FALSE, shape = FALSE, col = FALSE) +
  
  # Add titles
  ggtitle('Specification Curve: Effect of Carat on Above-Median Price') +
  labs(subtitle = "Error bars represent 99% confidence intervals for average marginal effects")

# Create customized control specification plot
control_plot = ggplot(control_grid_logit) +
  # Use diamond shapes for controls
  geom_point(
    aes(x = model, y = y, fill = value),
    shape = 23,  # Diamond shape
    size = 4
  ) +
  
  # Black for included, white for excluded
  scale_fill_manual(values = c('#FFFFFF', '#000000')) +
  guides(fill = FALSE) +
  
  # Custom y-axis labels showing control names
  scale_y_continuous(
    breaks = unique(control_grid_logit$y), 
    labels = unique(control_grid_logit$key),
    limits = c(
      min(control_grid_logit$y) - 1,
      max(control_grid_logit$y) + 1
    )
  ) +
  
  # X-axis shows specification number
  scale_x_continuous(
    breaks = c(1:max(control_grid_logit$model))
  ) +
  
  coord_cartesian(
    xlim = c(1 - min_space, max(control_grid_logit$model) + min_space)
  ) +
  
  # Minimal theme for control panel
  theme_classic() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    axis.title = element_blank(),
    axis.text.y = element_text(size = 10),
    axis.ticks = element_blank(),
    axis.line = element_blank()
  )

# Combine plots vertically
cowplot::plot_grid(
  coef_plot,
  control_plot,
  rel_heights = c(1, 0.5),  # Coefficient plot gets more space
  align = 'v',
  ncol = 1,
  axis = 'b'  # Align bottom axes
)
```

#### Comparing Multiple Model Types

A powerful extension is to compare results across different model types (e.g., logit vs. probit) in the same specification curve (Figure \@ref(fig:robustness-cowplot-plotgrid)). This helps assess whether your findings are specific to a particular functional form assumption:

```{r}
# Create custom probit estimation function
starb_probit = function(spec, data, rhs, ...) {
    # Extract additional arguments
    l = list(...)
    get_mfx = ifelse(is.null(l$get_mfx), FALSE, TRUE)
    
    spec = as.formula(spec)
    
    if (get_mfx) {
        # Calculate average marginal effects for probit
        model = suppressWarnings(
            glm(
                spec,
                data = data,
                family = binomial(link = 'probit'),  # Probit link
                weights = data$weight
            )
        ) %>%
            margins() %>%
            summary()
        
        row = which(model$factor == rhs)
        coef = model[row, 'AME'] %>% as.numeric()
        se   = model[row, 'SE'] %>% as.numeric()
        p    = model[row, 'p'] %>% as.numeric()
    } else {
        # Return raw probit coefficients
        model = suppressWarnings(
            glm(
                spec,
                data = data,
                family = binomial(link = 'probit'),
                weights = data$weight
            )
        ) %>%
            broom::tidy()
        
        row = which(model$term == rhs)
        coef = model[row, 'estimate'] %>% as.numeric()
        se   = model[row, 'std.error'] %>% as.numeric()
        p    = model[row, 'p.value'] %>% as.numeric()
    }
    
    # 99% confidence intervals
    z = qnorm(0.995)
    return(c(coef, p, coef + z * se, coef - z * se))
}

# Generate probit specification curve
probit_dfs = stability_plot(
    grid = grid1,
    data = diamonds,
    lhs = lhs_var,
    rhs = rhs_var,
    model = starb_probit,
    get_mfx = TRUE,
    perm = perm_controls,
    base = base_controls,
    fe_always = FALSE,
    run_from = 3,
    run_to = 5
)

# Adjust model numbers for probit to plot side-by-side with logit
coef_grid_probit = probit_dfs[[1]] %>% 
    mutate(model = model + max(coef_grid_logit$model))

control_grid_probit = probit_dfs[[2]] %>% 
    mutate(model = model + max(control_grid_logit$model))

# Combine logit and probit results
coef_grid_combined = bind_rows(coef_grid_logit, coef_grid_probit)
control_grid_combined = bind_rows(control_grid_logit, control_grid_probit)

# Generate combined plots
panels = stability_plot(
    coef_grid = coef_grid_combined,
    control_grid = control_grid_combined,
    data = diamonds,
    lhs = lhs_var,
    rhs = rhs_var,
    perm = perm_controls,
    base = base_controls,
    fe_always = FALSE,
    run_from = 5,
    run_to = 6
)

# Add annotations to distinguish model types
coef_plot_combined = panels[[1]] +
  # Vertical line separating logit and probit
  geom_vline(
    xintercept = max(coef_grid_logit$model) + 0.5,
    linetype = 'dashed',
    alpha = 0.8,
    size = 1
  ) +
  
  # Label for logit models
  annotate(
    geom = 'label',
    x = max(coef_grid_logit$model) / 2,
    y = 1.8,
    label = 'Logit models',
    size = 6,
    fill = '#D3D3D3',
    alpha = 0.7
  ) +
  
  # Label for probit models
  annotate(
    geom = 'label',
    x = max(coef_grid_logit$model) + max(coef_grid_probit$model) / 2,
    y = 1.8,
    label = 'Probit models',
    size = 6,
    fill = '#D3D3D3',
    alpha = 0.7
  ) +
  
  coord_cartesian(ylim = c(-0.5, 1.9))

control_plot_combined = panels[[2]] +
  geom_vline(
    xintercept = max(control_grid_logit$model) + 0.5,
    linetype = 'dashed',
    alpha = 0.8,
    size = 1
  )
```

```{r robustness-cowplot-plotgrid, fig.cap='Comparison of average marginal effects across logit and probit model specifications.', fig.alt='Side-by-side specification curves comparing average marginal effects from logit models on the left and probit models on the right, separated by a vertical dashed line. Both panels show the effects of diamond characteristics including depth, table width, and clarity on diamond price across multiple model specifications. Points represent marginal effect estimates with error bars indicating confidence intervals, with different colors distinguishing between variable levels or interactions. The y-axis shows marginal effect values ranging from approximately negative 0.5 to 2.0, while the x-axis displays different combinations of diamond dimension variables. The bottom portion uses gray boxes to indicate which control variables are included in each specification for both model types.'}
# Display combined plot
cowplot::plot_grid(
    coef_plot_combined,
    control_plot_combined,
    rel_heights = c(1, 0.5),
    align = 'v',
    ncol = 1,
    axis = 'b'
)
```

### The `specr` Package

The `specr` package provides an alternative implementation of specification curve analysis that focuses on concise specifications of the model space. Instead of defining a custom estimation function, the analyst specifies the sets of possible outcomes, focal predictors, controls, and sample restrictions. The package then enumerates all admissible specifications, estimates them, and offers summary and plotting methods.

In contrast to `starbility`, which is designed around user-supplied estimation functions and highly customized plotting, `specr` aims for a compact workflow that covers a broad range of standard models.

```{r}
# Load packages
library(specr)
library(tidyverse)   # Data manipulation and plotting
```

#### Preparing the Diamonds Example

To keep the exposition comparable to the `starbility` example, the same `diamonds` data are used. A subsample is created for speed, and log transformed variables are added to illustrate how `specr` can handle multiple outcomes and focal predictors.

```{r}
library(tidyverse)
library(specr)

# Load data
data("diamonds", package = "ggplot2")

set.seed(43)

# Subsample for computational convenience
indices = sample(
  x       = seq_len(nrow(diamonds)),
  size    = round(nrow(diamonds) / 20),
  replace = FALSE
)

diamonds_specr = diamonds[indices, ] |>
  mutate(
    high_clarity = clarity %in% c("VS1", "VVS2", "VVS1", "IF"),
    log_price    = log(price),
    log_carat    = log(carat)
  )
```

The key idea is now to describe the *specification universe* in terms of:

-   possible outcome variables,
-   possible focal predictors,
-   a set of optional controls that may or may not enter the model, and
-   optional sample restrictions.

`specr` will then traverse this design and estimate all corresponding models.

#### Defining and Running the Specification Curve

The central function in the package is typically called via

``` r
specr::specr(
  data,
  y,
  x,
  model,
  controls,
  subsets,
  ...
)
```

where

-   `y` is a vector of outcome variable names,
-   `x` is a vector of focal predictor names,
-   `model` specifies the estimation method (for example `"lm"` for linear regression),
-   `controls` is a vector of candidate control variables, and
-   `subsets` is an optional list that encodes sample restrictions.

The example below sets up a relatively rich specification universe:

-   outcomes: either `price` or `log_price`,
-   focal predictor: either `carat` or `log_carat`,
-   controls: any subset of seven potential covariates.

This already yields a sizable number of specifications and illustrates how quickly the design space expands.

```{r}
specs_diamonds <- specr::setup(
  data     = diamonds_specr,
  y        = c("price", "log_price"),
  x        = c("carat", "log_carat"),
  model    = "lm",
  controls = c("x", "y", "z", "cut", "color")
) |> 
  specr::specr()

# Inspect the resulting object
summary(specs_diamonds)
```

The `specs_diamonds` object stores one row per estimated specification, including the estimated coefficient for the focal predictor, its standard error, confidence interval, and associated $p$ value, plus a record of which modeling choices generated that estimate.

#### Visualizing the Specification Curve

`specr` provides a plot method that produces a specification curve directly from the results object. The exact appearance depends on the package version and plotting options, but the default is typically a curve where each point corresponds to one specification, the vertical axis shows the estimated coefficient, and uncertainty intervals are plotted around each point (Figure \@ref(fig:diamonds-specr-specification-curve)).

```{r diamonds-specr-specification-curve, fig.cap = "Specification-curve analysis of diamond price effects across model choices.", fig.alt = "Two-panel specification curve showing the robustness of diamond price effect estimates across different model specifications. Panel A displays coefficient estimates on the y-axis ranging from negative 10000 to positive 10000, plotted against specification number on the x-axis from 0 to approximately 120. Points are colored red for negative estimates, gray for near-zero estimates, and blue for positive estimates. Panel B shows the analytical choices made for each specification, including the dependent variable transformations at the top (log carat, log price, and linear model indicated by vertical lines), and control variables in the middle section showing various combinations of x, y, z variables with cut, clarity, and color controls. The bottom row indicates whether all controls were included. Red and blue vertical lines in Panel B correspond to the red and blue estimates in Panel A, showing which analytical choices produced negative versus positive results."}
# Basic specification curve plot
plot(specs_diamonds)
```

In this visualization (Figure \@ref(fig:diamonds-specr-specification-curve)), the user can typically read off:

-   the range of estimates across all admissible specifications,
-   how often the effect is statistically significant (for example, $p < 0.05$),
-   whether sign changes occur as controls are added or removed, and
-   how sensitive the effect size is to modeling choices.

Additional plotting options in `specr` generally allow the user to:

-   show separate panels for different outcomes or focal predictors,
-   display heatmaps of significance patterns,
-   or summarize the distribution of coefficients.

The exact arguments are version dependent, so it is good practice to consult `?plot` for the `specr` object to see the current capabilities and defaults.

#### Relation to the `starbility` Workflow

Both `starbility` and `specr` implement the same methodological idea: documenting the full set of defensible specifications and showing how the estimated effect of interest behaves across that universe.

From a practical perspective:

-   `specr` is convenient when:

    -   the relevant models are standard (for example linear regression, generalized linear models),
    -   the specification universe is naturally described in terms of outcome, focal predictor, controls, and simple subsets, and
    -   a concise, high level interface is preferred.

-   `starbility` is preferable when:

    -   fully custom estimation routines are needed,
    -   specialist estimators or complex clustering and weighting schemes are central to the analysis,
    -   or the user wants complete control over how coefficients, $p$ values, and confidence intervals are computed.

In empirical work, it is entirely reasonable to begin with `specr` to get a quick overview of robustness patterns, then move to `starbility` when the analysis requires more specialized modeling choices than `specr` natively supports.

### The `rdfanalysis` Package

While `starbility` is recommended for most applications, the `rdfanalysis` package by Joachim Gassen offers an alternative implementation with some unique features, particularly for research that follows a researcher degrees of freedom (RDF) framework.

#### Installation and Basic Usage

```{r, eval=FALSE}
# Install from GitHub
devtools::install_github("joachim-gassen/rdfanalysis")
```

The `rdfanalysis` package focuses on documenting and visualizing researcher degrees of freedom throughout the research process (Figure \@ref(fig:robustness-rdf-spec-curve)), from data collection to model specification:

```{r}
library(rdfanalysis)

# Load example estimates from the package documentation
load(url("https://joachim-gassen.github.io/data/rdf_ests.RData"))
```

```{r robustness-rdf-spec-curve, fig.cap='Specification curve from a multiverse analysis across analytical protocols.', fig.alt='Two-panel specification curve displaying results from a multiverse analysis across thousands of analytical protocols. The top panel shows effect estimates on the y-axis ranging from approximately negative 0.25 to 0.75, plotted against protocol number on the x-axis from 0 to over 12000. Individual estimates are shown as gray points with a smoothed trend line transitioning from red (negative) through black (zero) to blue (positive). The bottom panel shows the analytical choices made for each protocol using horizontal colored lines and dots. Y-axis labels indicate various methodological choices including missing data handling (na.omit with yes or no), variable transformations (idv and dv options), outlier treatment methods, model types (including lmer and various cutoff options), random effects specifications, and clustering approaches (cluster county and cluster city-year). The dense patterns of colored lines and dots show which combinations of analytical choices were used across the many protocols tested.'}
# Generate specification curve
# The package expects a dataframe with estimates and confidence bounds
plot_rdf_spec_curve(
  ests,      # Dataframe with estimates
  "est",     # Column name for point estimates
  "lb",      # Column name for lower confidence bound
  "ub"       # Column name for upper confidence bound
)
```

This level of transparency is particularly valuable for addressing concerns about p-hacking and researcher degrees of freedom.

------------------------------------------------------------------------

## Coefficient Stability {#sec-coefficient-stability}

Beyond specification curve analysis, another crucial aspect of sensitivity analysis is assessing whether your estimates are robust to potential omitted variable bias [@altonji2005selection]. Even with comprehensive controls, unobserved confounders may threaten causal inference.

### Theoretical Foundation: The Oster (2019) Approach

Oster's [-@oster2019unobservable] influential paper provides a formal framework for assessing omitted variable bias. The key insight is that coefficient stability alone is insufficient, we need to consider both coefficient stability and $R^2$ movement to evaluate the likely impact of unobservables.

The intuition is as follows:

1.  **Coefficient stability**: How much does the coefficient on your treatment variable change when you add controls?
2.  $R^2$ **movement**: How much does the model's explanatory power increase when you add controls?

If adding observed controls moves the $R^2$ substantially but barely affects the coefficient, this suggests that unobservables (which might also affect $R^2$) are unlikely to overturn your result. Conversely, if the coefficient is very sensitive to the controls you add, unobservables might also have large effects.

Oster formalizes this by computing a parameter $\delta$ (delta), which represents how strong the relationship between unobservables and the outcome would need to be (relative to the observables) to explain away the entire treatment effect. Higher values of $\delta$ suggest more robust results.

The formula for calculating the bias-adjusted treatment effect is:

$$
\beta^* = \tilde{\beta} - \delta \times [\dot{\beta} - \tilde{\beta}] \times \frac{R_{max} - \tilde{R}^2}{\tilde{R}^2 - \dot{R}^2}
$$

Where:

-   $\beta^*$ = bias-adjusted treatment effect
-   $\dot{\beta}$ = coefficient from short regression (without controls)
-   $\tilde{\beta}$ = coefficient from full regression (with controls)
-   $\dot{R}^2$ = $R^2$ from short regression
-   $\tilde{R}^2$ = $R^2$ from full regression
-   $R_{max}$ = hypothetical maximum $R^2$ (often set to 1 or a realistic upper bound)
-   $\delta$ = proportional selection on unobservables (assumed relationship)

### The `robomit` Package

The `robomit` package provides a straightforward implementation of Oster's method:

```{r}
library(robomit)

# Calculate bias-adjusted treatment effect using Oster's method
# This function estimates beta* under different assumptions about delta

o_beta_results = o_beta(
  y     = "mpg",        # Dependent variable
  x     = "wt",         # Treatment variable of interest
  con   = "hp + qsec",  # Control variables (covariates)
  delta = 1,            # Proportional selection assumption
                        # delta = 1 means unobservables are as important as observables
                        # delta = 0 means no omitted variable bias
                        # delta > 1 means unobservables are more important than observables
  R2max = 0.9,          # Maximum R-squared achievable
                        # Common choices: 1.0 (theoretical max), 
                        # 1.3*R2_full (30% improvement over full model),
                        # or domain-specific reasonable maximum
  type  = "lm",         # Model type: "lm" for OLS, "logit" for logistic
  data  = mtcars        # Dataset
)
```

The function returns:

-   Coefficient from short model (without controls)

-   Coefficient from full model (with controls)

-   Bias-adjusted coefficient under delta and $R^2$ max assumptions

-   The value of delta needed to drive the coefficient to zero

```{r}
print(o_beta_results)
```

Interpretation:

-   If bias-adjusted beta is still large and significant, this suggests robustness to omitted variable bias

-   If the delta needed to explain away the effect is large (\>1), this suggests you'd need very strong unobservables to overturn the result

For a more comprehensive analysis, you can explore coefficient stability across a range of delta values:

```{r}
# Create a sequence of delta values to explore
delta_values = seq(0, 2, by = 0.1)

beta_results = sapply(delta_values, function(d) {
  result = suppressWarnings(
    o_beta(
      y = "mpg",
      x = "wt",
      con = "hp + qsec",
      delta = d,
      R2max = 0.9,
      type = "lm",
      data = mtcars
    )
  )
  # Extract beta* (first row)
  result$Value[result$Name == "beta*"]
})

# Create visualization
stability_df = data.frame(
  delta = delta_values,
  beta_adjusted = beta_results
)
```

Interpretation guidelines:

-   $|\delta| < 1$: Unobservables would need to be less related to treatment/outcome than observables to overturn the result (less robust)

-   $|\delta| \approx 1$: Unobservables would need to be about as related as observables (moderate robustness)

-   $|\delta| > 1$: Unobservables would need to be MORE related than observables (more robust)

-   $|\delta| >> 1$: Very robust to omitted variable bias

Figure \@ref(fig:robustness-ovb-stability) shows bias-adjusted coefficients against $\delta$

```{r robustness-ovb-stability, fig.cap='Coefficient stability under omitted-variable bias.', fig.alt='Line plot showing the relationship between coefficient stability and the strength of potential confounding due to omitted variables. The x-axis represents delta, measuring the strength of confounding from 0.0 to 2.0, while the y-axis shows the bias-adjusted coefficient (beta*) ranging from approximately negative 6.5 to 0. A horizontal dashed red line is positioned at zero. The black solid line starts at approximately negative 4.5 when delta equals zero, gradually increases to approach negative 1 as delta approaches 1.0, then drops sharply to approximately negative 6.5 just after delta equals 1.0, before leveling off at approximately negative 6.3 for the remainder of the plot. This discontinuity illustrates a critical threshold in confounding strength where coefficient stability changes dramatically.'}
# Plot
library(ggplot2)
ggplot(stability_df, aes(x = delta, y = beta_adjusted)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "Delta (strength of confounding)",
    y = "Bias-adjusted coefficient (beta*)",
    title = "Coefficient stability under omitted variable bias"
  ) +
  theme_minimal()
```

For more sophisticated applications with multiple treatments or different model types:

```{r}
# Example with multiple treatment variables
# Useful when you have several key independent variables of interest

results_multi = lapply(c("wt", "hp"), function(treat) {
  o_beta(
    y = "mpg",
    x = treat,
    con = "qsec + gear + carb",  # More extensive controls
    delta = 1,
    R2max = 1.0,  # Theoretical maximum
    type = "lm",
    data = mtcars
  )
})

names(results_multi) = c("wt", "hp")
print(results_multi)
```

Comparing robustness across treatment variables helps identify which relationships are most stable.

### The `mplot` Package for Graphical Model Stability

The `mplot` package provides complementary tools for visualizing model stability and variable selection:

```{r}
# Install if needed
# install.packages("mplot")
library(mplot)

# Visualize variable importance across models
mplot::vis(
  lm(mpg ~ wt + hp + qsec + gear + carb, data = mtcars),
  B = 100  # Number of bootstrap samples
)
```

This creates plots showing:

1.  Which variables are selected across bootstrap samples
2.  Coefficient stability for each variable
3.  Model fit across different variable combinations

Particularly useful for:

-   Assessing variable importance

-   Understanding multicollinearity effects

-   Evaluating model selection stability

------------------------------------------------------------------------

## Quantifying Omitted Variable Bias

While Oster's approach focuses on bias-adjusted coefficients, the Konfound framework [@narvaiz2024konfound] takes a complementary approach by asking: "How much bias would be needed to invalidate our inference?"

### The `konfound` Package

The `konfound` package implements sensitivity analysis for causal inferences by calculating how much unmeasured confounding would be needed to change your substantive conclusion.

```{r}
library(konfound)

# Basic konfound analysis
# This calculates the amount of bias needed to invalidate your inference
pkonfound(
    est_eff = 5,         # Your estimated effect
    std_err = 2,         # Standard error of the estimate
    n_obs = 1000,        # Number of observations
    n_covariates = 5,    # Number of covariates in your model
    alpha = 0.05,        # Significance level
    tails = 2            # Two-tailed test
)
```

The output provides several key metrics:

1.  Robustness of Inference (RIR): The number of observations that would need to be replaced with observations having null effects to invalidate the inference

2.  Percentage of sample that would need to be replaced: RIR / n_obs \* 100

3.  Impact threshold: The correlation between an omitted variable and both the treatment and outcome needed to invalidate the inference

Interpretation:

-   Higher RIR = more robust inference

-   RIR should be compared to n_obs to assess practical robustness

-   Impact threshold shows how strong confounding needs to be

### Visualizing Sensitivity: The Threshold Plot

In Figure \@ref(fig:robustness-threshold-plot), the threshold plot shows the combination of correlations between a confound and the treatment (horizontal axis) and outcome (vertical axis) that would be needed to overturn your conclusion:

```{r robustness-threshold-plot, fig.cap='Estimated effect size relative to the robustness threshold.', fig.alt='Stacked area chart showing the estimated effect size in relation to a robustness threshold. The y-axis displays the absolute value of the effect ranging from 0 to 5, while the x-axis represents the estimated effect.'}
# Create threshold plot
# This visualizes the "confounding space" that would invalidate inference
pkonfound(
    est_eff = 5,
    std_err = 2,
    n_obs = 1000,
    n_covariates = 5,
    to_return = "thresh_plot"  # Request threshold plot
)
```

Interpretation of the plot:

-   The red line shows the threshold

-   Points above/beyond this line represent confounding strong enough to overturn your inference

-   You can compare this to the strength of known confounds

-   Benchmark correlations (e.g., 0.1, 0.3, 0.5) help assess plausibility

Example interpretation: "An omitted variable would need to be correlated at 0.35 with both the treatment and outcome to invalidate our inference. Given that our strongest observed control is correlated at 0.25 with the outcome, this seems unlikely."

### The Correlation Plot

In Figure \@ref(fig:robustness-correlation-plot), the correlation plot provides another view, showing the required partial correlation of an omitted variable with the outcome, conditional on the treatment and covariates:

```{r robustness-correlation-plot, fig.cap='Sensitivity analysis showing conditions required to invalidate a causal inference.', fig.alt='Path diagram illustrating the requirements to invalidate a causal inference through omitted variable bias. The diagram shows relationships between four variables arranged in a roughly triangular pattern.'}
# Create correlation plot
pkonfound(
    est_eff = 5,
    std_err = 2,
    n_obs = 1000,
    n_covariates = 5,
    to_return = "corr_plot"  # Request correlation plot
)
```

This plot shows:

-   The relationship between bias and the required correlation

-   How much the effect estimate would change for different confound strengths

-   The threshold where the inference would be overturned

### Konfound for Model Objects

You can also apply konfound directly to model objects, which is more convenient for real analyses:

```{r, eval=FALSE}
# Fit your model
model = lm(mpg ~ wt + hp + qsec, data = mtcars)

# Apply konfound to the model
# This automatically extracts the necessary statistics
konfound(model, wt)  # Assess robustness for the 'wt' coefficient
```

------------------------------------------------------------------------

## Cinelli–Hazlett Robustness Value (`sensemakr`)

While Oster's $\delta$ summarizes selection on unobservables relative to observables, and `konfound` rephrases the question in terms of the share of cases that would need to be replaced, @cinelli2020making provide a complementary and now widely adopted answer. They ask: *how strong would an unobserved confounder need to be, in terms of partial $R^2$ with both the treatment and the outcome, before the estimated effect would lose statistical significance, or be reduced to zero?* Their `sensemakr` package operationalizes this question with two interpretable summaries that can be reported alongside any regression-based effect estimate.

### Motivation and the Two Key Statistics

The framework is built around partial $R^2$ values rather than abstract selection ratios. For a hypothesized confounder $Z$, two quantities matter:

-   $R^2_{Y \sim Z \mid D, \mathbf{X}}$, the partial $R^2$ of the confounder with the outcome, given treatment and observed covariates, and
-   $R^2_{D \sim Z \mid \mathbf{X}}$, the partial $R^2$ of the confounder with the treatment, given observed covariates.

A confounder is dangerous only if it is reasonably well associated with *both*. From these two ingredients, @cinelli2020making derive two headline summaries:

1.  **Robustness Value (RV)**. The smallest value $\mathrm{RV}_q$ such that a confounder with partial $R^2$ at least $\mathrm{RV}_q$ on *both* the treatment and the outcome would be sufficient to reduce the estimated coefficient by a fraction $q$. With $q = 1$ this is the strength needed to drive the point estimate to zero; with $q = q^\star$ chosen so that the bound coincides with the conventional $t$-statistic threshold (typically $1.96$), the RV is the strength needed to make the estimate statistically insignificant.
2.  **Partial $R^2$ of treatment with outcome**. This is the worst-case bound: even a confounder that explains *all* residual variation in the treatment can only bias the estimate by an amount governed by $R^2_{Y \sim D \mid \mathbf{X}}$. If this number is small, no plausible confounder, however strong, can flip the sign.

A useful rule of thumb is that an RV above $0.1$–$0.2$ already represents a fairly demanding bar: an unobserved confounder would need to explain at least 10–20% of the residual variation in *both* the treatment and the outcome, after netting out everything in the model.

### The Extreme-Scenario Interpretation

A common objection to sensitivity analysis is that the bounds feel abstract: *what does $R^2 = 0.15$ on the treatment actually mean?* `sensemakr` provides two natural answers.

-   **Benchmarking against an observed covariate**. The analyst picks a covariate (or set of covariates) already in the model and asks what would happen if the unobserved confounder were $kd \times$ as strong as that covariate, where $kd \in \{1, 2, 3\}$ is a multiplier. This converts an abstract partial $R^2$ into the concrete statement *"a confounder $3\times$ as predictive as gender"*.
-   **Extreme-confounding scenario**. The procedure also reports what the estimate becomes under a worst-case confounder that explains *all* residual variation in the outcome, varying only its association with treatment. If the estimate survives this stress test, the result is essentially impervious to omitted variable bias of that variety.

### Worked Example: The `darfur` Study

The canonical illustration in @cinelli2020making uses the `darfur` data shipped with `sensemakr`, drawn from a study of attitudes toward peace among survivors of violence in Darfur. The treatment `directlyharmed` indicates whether the respondent was directly attacked, and the outcome `peacefactor` is a measure of pro-peace attitudes.

```{r}
library(sensemakr)

data(darfur)

# Fit the regression
darfur_model <- lm(
  peacefactor ~ directlyharmed + village + female + age +
                farmer_dar + herder_dar + pastvoted + hhsize_darfur,
  data = darfur
)

# Run the sensitivity analysis, benchmarking against `female`
darfur_sens <- sensemakr(
  model                = darfur_model,
  treatment            = "directlyharmed",
  benchmark_covariates = "female",
  kd                   = 1:3
)

summary(darfur_sens)
```

The summary output reports the unadjusted point estimate and standard error, the robustness value to bring the estimate to zero, the robustness value to bring the $t$-statistic below the conventional significance threshold, and the partial $R^2$ of treatment with outcome. It then tabulates the bias-adjusted estimate, $t$-statistic, and lower confidence bound under hypothetical confounders that are $1\times$, $2\times$, and $3\times$ as strong as the chosen benchmark covariate (`female` in this example).

For the Darfur data, the RV is on the order of $0.14$, meaning that a confounder would need a partial $R^2$ of about 14% with both treatment and outcome to drive the estimate to zero. That is substantially stronger than the role played by `female` in this regression, which lends credibility to the original conclusion.

### The Sensitivity Contour Plot

The most informative output of `sensemakr` is a contour plot in the partial-$R^2$ plane (Fig. \@ref(fig:robustness-sensemakr-contour)). The horizontal axis represents the partial $R^2$ of the confounder with the treatment; the vertical axis represents the partial $R^2$ of the confounder with the outcome. Contour lines show the bias-adjusted estimate (or $t$-statistic) for every combination of the two. Points labeled $1\times$, $2\times$, $3\times$ indicate where confounders of the corresponding multiples of the benchmark covariate would land in this plane. If those points sit far inside the region of "estimates that remain significant", the result is robust.

```{r robustness-sensemakr-contour, fig.cap="Cinelli and Hazlett sensitivity contour plot for the darfur regression.", fig.alt="Contour plot in the partial R squared plane. The horizontal axis is the partial R squared of an unobserved confounder with the treatment, and the vertical axis is the partial R squared of the same confounder with the outcome. Contour lines indicate the bias-adjusted estimated effect for every combination of the two strengths. Three labeled points show where confounders that are one, two, and three times as strong as the benchmark covariate female would lie."}
plot(darfur_sens)
```

The same machinery can be used to plot the bias-adjusted $t$-statistic instead of the point estimate, by passing `sensitivity.of = "t-value"` to `plot`. This is convenient when the practical question is whether a confounder of a given strength would cause the result to lose statistical significance.

### Practical Recommendations

A few practical points are worth keeping in mind:

1.  **Pick benchmarks that are substantively informative**. Choose covariates that are strong predictors of both treatment and outcome and whose role in the system is well understood, so that the $kd$-multiplier statements are meaningful.
2.  **Report RV alongside the point estimate**. A coefficient of `0.10` with `RV = 0.20` and `RV (alpha = 0.05) = 0.13` tells the reader much more than a coefficient and a $p$-value alone.
3.  **Combine with other tools**. RV/`sensemakr` is complementary to Oster's $\delta$ (which gives a single scalar), `konfound` (which expresses the threshold as a share of biased cases), and the matching-based Rosenbaum bounds described next.

------------------------------------------------------------------------

## Rosenbaum Bounds

The methods discussed so far are designed for regression-based estimators. In matched observational studies, where the analyst has explicitly paired treated units with similar controls on observed covariates, a different and arguably even more natural sensitivity tool is available: the @rosenbaum2002overt sensitivity bound. It asks how much the matched pairs would have to differ on an *unobserved* covariate before the conclusions would be overturned.

### Motivation and the $\Gamma$ Parameter

After matching, two units in the same matched set look identical on observed covariates $\mathbf{X}$. If the matching were perfect and treatment assignment were ignorable given $\mathbf{X}$, then within each matched set treatment would be assigned by a coin flip. Rosenbaum's framework asks: *what if the coin is biased, and the bias comes from an unobserved covariate $U$ that we did not match on?*

Formally, for two units $i$ and $j$ in the same matched set with treatment probabilities $\pi_i$ and $\pi_j$, the odds ratio is bounded by

$$
\frac{1}{\Gamma} \;\le\; \frac{\pi_i / (1 - \pi_i)}{\pi_j / (1 - \pi_j)} \;\le\; \Gamma.
$$

The parameter $\Gamma \ge 1$ measures the maximum departure from random assignment that an unobserved confounder could induce.

-   $\Gamma = 1$ corresponds to a randomized experiment: matched units are equally likely to be treated.
-   $\Gamma = 2$ means that, within a matched pair, one unit could be up to twice as likely as its partner to be treated, on the basis of some unobserved difference.
-   $\Gamma = 5$ would mean the unobserved confounder has *enormous* leverage over treatment assignment.

The sensitivity analysis proceeds by computing, at each value of $\Gamma$, an upper bound on the $p$-value (or a lower bound on the test statistic) for the treatment effect under the worst-case allocation of $U$ across pairs. The *critical $\Gamma$* is the value at which the bound first crosses the chosen significance threshold.

A widely cited rule of thumb is that critical $\Gamma$ values above about $2$ generally indicate a robust finding for a binary treatment in social-science contexts: an unobserved confounder would need to roughly double the odds of treatment to overturn the result. Examples of $\Gamma$ thresholds in published studies are tabulated in Table \@ref(tab:rosenbaum-gamma-marketing) of the [matching-methods chapter](#sec-matching-methods).

### Worked Example: Lalonde Job-Training Data

The `lalonde` dataset packaged with `MatchIt` is a standard testbed for matching methods. The treatment is participation in a job-training program, and the outcome `re78` is real earnings in 1978. We construct a nearest-neighbor match on a small set of pre-treatment covariates and then compute Rosenbaum bounds on the matched pairs.

```{r}
library(MatchIt)
library(rbounds)

data("lalonde", package = "MatchIt")

# 1:1 nearest-neighbor matching on observed covariates
m_out <- matchit(
  treat ~ age + educ + race + married + nodegree + re74 + re75,
  data    = lalonde,
  method  = "nearest"
)

matched <- match.data(m_out)

# Wilcoxon signed-rank Rosenbaum bounds
psens(
  matched$re78[matched$treat == 1],
  matched$re78[matched$treat == 0],
  Gamma    = 2,
  GammaInc = 0.1
)
```

The `psens()` function returns, for each value of $\Gamma$ between $1$ and the supplied upper limit, a lower and upper $p$-value bound for the Wilcoxon signed-rank test under the worst-case unobserved confounder. The critical $\Gamma$ is the smallest value at which the upper $p$-value bound exceeds the chosen significance level (often $0.05$).

For Hodges–Lehmann point and interval estimates of the treatment effect under the same family of bounds, `rbounds::hlsens()` provides the analogous sensitivity table:

```{r}
hlsens(
  matched$re78[matched$treat == 1],
  matched$re78[matched$treat == 0],
  Gamma    = 2,
  GammaInc = 0.1
)
```

### Interpreting and Reporting

When reporting Rosenbaum bounds, three numbers are typically of interest:

-   the *unadjusted* test statistic and $p$-value (corresponding to $\Gamma = 1$),
-   the *critical $\Gamma$* at which the upper $p$-value bound reaches the chosen significance threshold, and
-   the bias-adjusted point and interval estimates from `hlsens()` at one or two interpretable values of $\Gamma$ (for example $\Gamma = 1.5$ and $\Gamma = 2$).

It is good practice to interpret the critical $\Gamma$ in domain-specific language: *"to overturn this conclusion, an unobserved confounder would need to make program participants up to $\Gamma$ times more likely to be matched into the treatment group than their observably similar controls, even after conditioning on age, education, race, marital status, schooling, and prior earnings."* Whether such a confounder is plausible is a substantive judgment that the bound itself cannot settle.

### Extensions

The basic Rosenbaum bound has been extended in several directions that are increasingly relevant in modern observational work:

-   **Clustered Rosenbaum bounds** [@hansen2014clustered] adapt the framework to settings where treatment is assigned at a higher level than the unit of analysis (for example, treatment at the school level with student outcomes), so that biases need to be evaluated at the cluster level rather than pair by pair.
-   **Amplification and the $(\Lambda, \Delta)$ parameterization**, due to @rosenbaum2009amplification, decomposes the single sensitivity parameter $\Gamma$ into two more interpretable pieces: $\Lambda$, the maximum impact of $U$ on treatment, and $\Delta$, the maximum impact of $U$ on the outcome. A given $\Gamma$ corresponds to a one-dimensional curve in the $(\Lambda, \Delta)$ plane, which can be plotted alongside the Cinelli–Hazlett contour from the previous section.
-   **Sensitivity for matched comparative studies with multiple control groups, instruments, or regression discontinuity-style designs** is implemented in `sensitivitymv`, `sensitivitymw`, and related packages, which generalize `rbounds` to weighted M-estimators and to studies with more than one matched control per treated unit.

In practice, the Rosenbaum framework and the Cinelli–Hazlett RV framework are usually run side by side. They probe overlapping but distinct aspects of unmeasured confounding: $\Gamma$ speaks the language of treatment-assignment odds within matched sets, while RV speaks the language of partial $R^2$ in the regression that produced the point estimate. Reporting both gives the reader a much more complete picture of how fragile or robust a causal claim really is.

------------------------------------------------------------------------

## Advanced Topics in Sensitivity Analysis

### Sensitivity to Outliers and Influential Observations

Outliers can drive results, so it's important to assess robustness to their inclusion:

```{r}
# Identify influential observations
model = lm(mpg ~ wt + hp + qsec, data = mtcars)

# Calculate influence measures
influence_measures = influence.measures(model)

# Cook's Distance
cooks_d = cooks.distance(model)
influential = cooks_d > 4 / nrow(mtcars)  # Common threshold

# DFBETAS (change in coefficient when observation removed)
dfbetas_vals = dfbetas(model)

# Compare models with/without influential observations
model_full = lm(mpg ~ wt + hp + qsec, data = mtcars)
model_no_influential = lm(
  mpg ~ wt + hp + qsec,
  data = mtcars[!influential, ]
)

# Compare coefficients
compare_df = data.frame(
  Variable = names(coef(model_full)),
  Full_Sample = coef(model_full),
  Excl_Influential = c(
    coef(model_no_influential),
    rep(NA, length(coef(model_full)) - length(coef(model_no_influential)))
  )
)

# Winsorization approach
library(DescTools)
mtcars_winsor = mtcars
mtcars_winsor$wt = DescTools::Winsorize(mtcars$wt, val = c(0.01, 0.99))
mtcars_winsor$hp = DescTools::Winsorize(mtcars$hp, val = c(0.01, 0.99))

model_winsor = lm(mpg ~ wt + hp + qsec, data = mtcars_winsor)

# Compare original vs. winsorized
```

### Sensitivity to Measurement Error

If key variables are measured with error, results may be biased:

```{r}
# Simulate measurement error
# This helps understand potential attenuation bias

library(simex)

# SIMEX (Simulation-Extrapolation) for measurement error correction
# Assumes classical measurement error in covariates

# Fit model with x = TRUE
naive_model <- lm(mpg ~ wt + hp, data = mtcars, x = TRUE)

# Apply SIMEX
# lambda is the factor by which measurement error variance increases
simex_model <- simex(
  naive_model,
  SIMEXvariable = "wt",  # Variable with measurement error
  measurement.error = 0.1,  # Assumed ME variance (proportion of var)
  lambda = seq(0.1, 2, 0.1),  # Extrapolation sequence
  B = 100  # Number of simulations
)


# View results
summary(simex_model)

# Compare naive vs corrected coefficients
data.frame(
  Coefficient = names(coef(naive_model)),
  Naive = coef(naive_model),
  SIMEX_Corrected = coef(simex_model)
)

# SIMEX extrapolates back to zero ME, giving corrected estimate
```

Figure \@ref(fig:robustness-simex) plots SIMEX extrapolation.

```{r robustness-simex, fig.cap='SIMEX extrapolation plots for measurement-error correction.', fig.alt='Three-panel diagnostic plot showing SIMEX (Simulation Extrapolation) results for correcting measurement error in regression coefficients.'}
# Plot SIMEX extrapolation
par(mfrow = c(1, 3))
plot(simex_model)
par(mfrow = c(1, 1))
```

------------------------------------------------------------------------

## Reporting Sensitivity Analysis Results

### Best Practices for Presentation

When presenting sensitivity analyses in your paper:

1.  **Main Text**: Present your primary specification and 1-2 key robustness checks that directly address the most plausible threats to identification

2.  **Tables**: Create comprehensive tables showing:

    -   Coefficient estimates across specifications
    -   Standard errors (in parentheses)
    -   $R^2$ and other fit statistics
    -   Number of observations
    -   Clear column headers describing each specification

3.  **Figures**: Use specification curves for visual impact when you have many specifications

4.  **Appendix**: Place exhaustive robustness checks in appendices with clear organization

## Conclusion and Recommendations

Sensitivity analysis is not optional in modern empirical research, it's essential. The techniques described in this chapter provide a toolkit for assessing the robustness of your findings:

1.  **Start with specification curve analysis** (`starbility`) to visualize how your results vary across defensible specifications

2.  **Apply Oster's method** (`robomit`) to assess sensitivity to omitted variable bias using coefficient stability and $R^2$ movement

3.  **Use konfound analysis** (`konfound`) to quantify how much bias would be needed to overturn your inference

4.  **Conduct additional tests** relevant to your context: placebo tests, subsample analysis, outlier diagnostics, etc.

5.  **Present results clearly**: Use both tables and figures, provide verbal interpretation, and be transparent about which specifications you view as most credible and why

**Remember**: The goal is not to show that your result is robust to *everything*, but to demonstrate that it's robust to *reasonable alternative choices* and *plausible threats to identification*. Be honest about the limitations while making the strongest case possible for your findings.

The mark of rigorous empirical work is not that every robustness check confirms your main result, it's that you've thoughtfully considered the most important threats to validity and provided evidence about whether those threats are likely to overturn your conclusions.

------------------------------------------------------------------------
