## Placebo, Falsification, and Negative-Control Tests {#sec-placebo-falsification}

The sensitivity analyses discussed earlier in this chapter ask how strong an unobserved confounder would have to be before it overturned a result. Placebo and falsification tests ask a complementary question. Instead of quantifying the bias an unmeasured confounder could produce, they look for the footprints such a confounder should leave if it were present. The logic is indirect but powerful. If the identifying assumptions of a study hold, then certain quantities that the analyst can estimate ought to be zero. When those quantities turn out to be large and statistically distinguishable from zero, the design is telling us that something other than the treatment is driving the data.

Researchers use many names for this family of checks, including balance tests, falsification tests, refutability tests, tests for known effects, tests of unconfoundedness, and tests with negative controls. The shared idea is that we construct a setting in which we already know what the answer should be, run the estimator, and see whether it returns that known answer. A placebo test is therefore a deliberate attempt to make the design fail. Passing the test does not prove that the identifying assumptions hold, since no finite set of falsification checks can verify an untestable assumption such as unconfoundedness. Failing the test, by contrast, is informative in the way that a refuted prediction is informative in any empirical science. A significant placebo effect is evidence that the study design has a problem, even when we cannot say exactly which assumption has broken.

### Two Complementary Designs

There are two broad ways to construct a quantity that should be zero, and they correspond to the two columns of a study, the treatment and the outcome.

The first design estimates the causal effect of a treatment that is already understood to have no impact on the outcome of interest. We replace the real treatment with a fake or placebo treatment, run the same estimator, and check that the estimated effect is null. The replacement can be a unit that was never exposed but is analyzed as if it had been, a time period before the policy took effect, or a group that the mechanism could not plausibly reach. In a difference in differences study, for example, assigning a placebo intervention date that precedes the actual policy change should yield no effect if the parallel trends assumption holds, because nothing happened at the placebo date. A nonzero placebo estimate signals that treated and control groups were already diverging for reasons unrelated to the policy.

The second design estimates the causal effect of the real treatment on an outcome that is known to be unaffected by it. Here the treatment is genuine, but the dependent variable is chosen because no causal pathway connects it to the treatment. If the treatment nonetheless appears to move this placebo outcome, the apparent association cannot be causal and must instead reflect confounding, selection, or some other artifact of the design. A classic instance is checking whether a treatment predicts an outcome measured before the treatment was assigned. Since the future cannot cause the past, any such association is spurious by construction and reveals a flaw in the comparison.

Both strategies rest on the same statistical foundation. Each evaluates the null hypothesis that the relevant average causal effect equals zero, and each treats a rejection of that null as a warning rather than as a finding to be celebrated. The interpretive direction is reversed relative to ordinary hypothesis testing. In a placebo test we hope to fail to reject the null, because the null is what the valid design predicts.

### Negative Controls

The negative-control framework, articulated for observational epidemiology by @lipsitch2010negative, gives these ideas a precise vocabulary and a clear causal-diagram interpretation. A negative control is a variable known a priori to have no causal relationship with either the exposure or the outcome through the pathway under study, yet to share the same sources of confounding as the primary analysis. Because it shares the confounding structure but not the causal effect, the negative control isolates bias. If the analysis is unconfounded, the estimated negative-control association should be null. If confounding is present, it should bleed into the negative-control estimate, and its magnitude offers a direct, observable signal of the bias that we otherwise could only bound through sensitivity analysis.

@lipsitch2010negative distinguish two types. A negative-control outcome is a response variable that the exposure cannot affect but that is subject to the same confounders as the true outcome. Detecting an exposure effect on such an outcome implies residual confounding or selection bias. A negative-control exposure is a treatment that cannot affect the outcome but that is associated with the same confounders as the true exposure. Detecting an effect of the negative-control exposure on the outcome again implies that the comparison is contaminated. The well-known observation that influenza vaccination appeared to reduce mortality during periods when influenza was not circulating is a canonical negative-control finding. Because the vaccine could not prevent deaths in the absence of influenza, the apparent protective effect exposed healthy-vaccinee selection, with healthier people both more likely to be vaccinated and less likely to die.

A useful way to read a negative control is through the same causal graph that motivates the main analysis. The control is chosen so that the only open path between it and the comparison variable runs through the suspected confounders. With the causal arrow severed by design, any remaining association must travel along the backdoor path, so the negative-control estimate measures the backdoor association alone. This logic connects placebo testing to the broader literature on confounding bias and underpins the proximal causal inference methods that use negative controls not merely to detect bias but to correct for it, as developed by @miao2018identifying and surveyed by @tchetgen2024proximal.

### The Rosenbaum Tradition

The use of known effects to probe a design has a long history in observational study practice. @rosenbaum2002overt frames falsification through the related notions of known effects [@rosenbaum1989known] and pattern specificity. If a treatment is believed to operate through a specific mechanism, then it should affect outcomes that lie on the mechanism's pathway while leaving outcomes off that pathway untouched. Observing the predicted pattern of present and absent effects, rather than a diffuse association with everything, strengthens a causal claim, because confounding tends to produce broad rather than narrowly targeted associations. A confounder that raises one outcome usually raises many, so a treatment that moves only the theoretically predicted outcome is harder to explain away as confounding.

Rosenbaum also emphasizes that a single comparison rarely settles a causal question, and that elaborate theories with multiple testable implications are more credible precisely because they expose the design to more opportunities to fail. A negative-control outcome is one such implication. So is the requirement that the treatment effect be absent in a subgroup that the mechanism cannot reach, or that a dose-response relationship appear where the theory predicts one. Each falsification check that the design survives makes the alternative explanation, that confounding rather than treatment produced the result, less tenable, even though no check can ever rule it out entirely.

The three devices that @rosenbaum1984association introduced, which he called unaffected responses, essentially equivalent treatments, and unaffected units, turn out to organize almost the entire applied practice. @eggers2024placebo survey every observational study mentioning a placebo, balance, or falsification test in four leading political science journals between 2009 and 2018 and find that these three categories, which they rename placebo outcomes, placebo treatments, and placebo populations, account for nearly all of the 110 tests they catalog. The next several subsections develop their framework, which places placebo tests and the research designs they probe inside a single hypothesis-testing logic and supplies the vocabulary that the rest of this section uses.

### A Hypothesis-Testing Framework for Placebo Tests {#sec-placebo-etd-framework}

The most useful way to think about what a placebo test delivers is as a hypothesis test about the research design itself. Following @eggers2024placebo, let the null hypothesis $H_0$ be that the design's core assumptions hold and let the alternative $H_1$ be that they are violated in some specific, relevant way. A placebo test reports a binary verdict. We say the test fails when it rejects the no-effect null in the placebo analysis and passes when it does not. Write $p_0$ for the probability that the test fails when the core assumptions actually hold, which is its false positive rate or size, and $p_1$ for the probability that it fails when the assumptions are violated, which is its true positive rate or power. A single application of Bayes rule then shows that the test is informative, meaning a failing result is evidence against the design and a passing result is evidence for it, if and only if $p_1 > p_0$. Power must exceed size. The larger the gap between them, the more a single result moves our beliefs, exactly as the diagnostic odds ratio summarizes the value of a medical test. This framing carries an immediate practical lesson. A placebo test that has almost no power, because the placebo estimate is buried in noise, has $p_1$ barely above $p_0$ and therefore teaches us almost nothing, whether it passes or fails. A passing test from an underpowered design is not reassurance, which is the central warning of @hartman2018equivalence and the reason they advocate reversing the null through equivalence testing.

To see what has to be true for power to exceed size, @eggers2024placebo decompose the core estimate into the true effect, the bias, and sampling error, $\hat{\delta} = \delta + b + \varepsilon$. The design rests on two distinct kinds of assumption. The bias assumptions $\mathcal{BA}$, covering identification, estimation, measurement, and sample selection, are what make $b = 0$ so that the estimator is centered on the truth. The distributional assumptions $\mathcal{DA}$ govern the sampling distribution of $\varepsilon$, and they are what make the stated standard errors and the rejection region correct. The placebo analysis produces its own estimate $\hat{\delta}_p = \delta_p + b_p + \varepsilon_p$ by altering one ingredient of the core analysis, and its informativeness rests on a small set of named assumptions that @eggers2024placebo state formally. It is worth listing them, because they are the precise content of the loose phrase "the placebo should mirror the core analysis."

1. No average treatment effect (NATE): $\delta_p = 0$. The treatment has no effect on the outcome in the placebo analysis, which is what justifies the word placebo and guarantees that any signal the test detects is not a real effect.
2. Linked bias assumptions (LBA): $\mathcal{BA} \Rightarrow b_p = 0$. If the bias assumptions hold in the core analysis, they also hold in the placebo analysis, so the placebo inherits the design's freedom from bias.
3. Linked distributional assumptions (LDA): $\mathcal{DA}$ implies the placebo rejection region also has at most nominal mass, so the placebo inherits the design's correct standard errors.
4. Linked violation of bias assumptions (LVBA): if the core analysis is biased in the way contemplated by the alternative, then $b_p \neq 0$, so the placebo inherits the design's bias.
5. Linked violation of distributional assumptions (LVDA): if the core standard errors are wrong in the contemplated way, then the placebo rejection region also carries excess mass, so the placebo inherits the design's miscalibration.
6. Symmetric placebo sampling error (SEPA): the sampling distribution of $\hat\delta_p$ is unimodal and symmetric, a regularity condition ensuring that a nonzero $b_p$ actually pushes the estimate into the rejection region more often.

These assumptions slot together into two clean sufficiency results [@eggers2024placebo]. NATE, LBA, and LDA jointly hold the false positive rate to the nominal level, $p_0 \leq \alpha_p$, whenever the core assumptions are true. Adding LVBA and SEPA makes a test of the bias assumptions informative, lifting $p_1$ above $\alpha_p$ whenever the design is biased, while adding LVDA instead makes a test of the distributional assumptions informative. The two halves capture the same tension in formal dress. The placebo must inherit the design's virtues (LBA and LDA) so that it does not raise false alarms, yet it must also inherit the design's vices (LVBA or LVDA) so that it can sound a true one. The art of constructing a placebo test lies exactly here. The alteration must be large enough to switch off the genuine treatment effect (NATE), yet minimal enough that the placebo analysis still reproduces whatever flaw might be corrupting the core estimate (LVBA or LVDA). An alteration that breaks the link to the treatment but also accidentally removes the confounding satisfies NATE while violating LVBA, producing a powerless test that passes for the wrong reason.

A compact simulation makes the power-versus-size logic concrete. We imagine a placebo outcome test where the placebo outcome shares the design's confounder to a degree controlled by a loading $\lambda$. When $\lambda = 0$ the placebo carries none of the confounding and the test is pure size, rejecting at the nominal five percent. As $\lambda$ grows the test inherits more of the bias and its power climbs. Tracing the rejection rate against $\lambda$ traces the informativeness of the test.

```{r etd-power-size, eval=FALSE}
library(tidyverse)

set.seed(123)

# One placebo-outcome test. The confounder U biases the treatment-outcome
# association; the placebo outcome loads on U with strength `lambda` and is
# never affected by the treatment, so its true effect is zero (NATE holds).
sim_placebo <- function(n = 300, lambda = 1, conf = 0.6) {
  U     <- rnorm(n)
  treat <- rbinom(n, 1, plogis(conf * U))        # U drives treatment (the flaw)
  y_plc <- lambda * U + rnorm(n)                  # placebo outcome: no treat effect
  summary(lm(y_plc ~ treat))$coefficients["treat", "Pr(>|t|)"]
}

# Rejection rate (estimated p1) across placebo loadings; lambda = 0 gives p0.
grid <- expand_grid(lambda = seq(0, 1, by = 0.1), rep = 1:400) |>
  mutate(p = map_dbl(lambda, ~ sim_placebo(lambda = .x))) |>
  group_by(lambda) |>
  summarise(reject_rate = mean(p < 0.05), .groups = "drop")

ggplot(grid, aes(lambda, reject_rate)) +
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  geom_line() + geom_point() +
  labs(x = "Confounder loading on the placebo outcome (lambda)",
       y = "Probability the placebo test fails") +
  causalverse::ama_theme()
```

When the placebo outcome does not share the confounder, the curve sits at the dashed nominal line, the test is all size and no power, and a passing result is uninformative. As the shared loading increases the failure rate rises well above five percent, so a failing test becomes strong evidence of confounding while a passing test becomes meaningful reassurance. The figure is a visual statement of the requirement $p_1 > p_0$, and it shows why reporting the precision of a placebo estimate matters at least as much as reporting its significance.

### A Typology: Placebo Outcomes, Treatments, and Populations {#sec-placebo-typology}

@eggers2024placebo classify placebo tests along two axes. The first asks which assumption is under examination, the bias assumptions that concern the point estimate or the distributional assumptions that concern the standard error. The second asks which component of the core analysis has been altered to build the placebo, the outcome, the treatment, or the population. Crossing these axes organizes the sprawling vocabulary of falsification testing into a small grid, and their survey shows the practical weight of each cell. Of the 110 tests they catalog, 64 alter the outcome, 34 alter the treatment, and 12 alter the population, and the overwhelming majority probe bias rather than distribution.

A placebo outcome test replaces the real outcome with one that the treatment cannot affect but that responds to the same confounders. This is the negative-control outcome of Section \@ref(sec-placebo-falsification), and when the substitute outcome is a pretreatment covariate the test reduces to the familiar balance test, whose validity is essentially automatic because a variable fixed before treatment cannot be moved by it. Post-treatment placebo outcomes are more delicate. When @eggers2024placebo discuss the study by Dube and coauthors of the United States assault weapons ban and homicides in nearby Mexican states, where death by accident and death by suicide serve as placebo outcomes, they stress that the no-effect assumption is no longer guaranteed, since a change in firearm availability could in principle move those rates too. Authors who use a post-treatment placebo outcome owe the reader an argument for why the treatment leaves it untouched.

A placebo treatment test replaces the real treatment with one that does not affect the outcome but that would be subject to the same confounding. The running example in @eggers2024placebo is the study by Peisakhin and Rozenas of Russian television in Ukraine, where the quality of the Russian news signal is the treatment and the quality of the Russian sports and entertainment signal is the placebo treatment. If transmitter placement were correlated with latent pro-Russian sentiment, that correlation should contaminate the sports signal as much as the news signal, yet sports broadcasts are assumed not to move the vote. A significant association between the placebo treatment and the outcome therefore points to confounding in the core design. There is a subtle requirement here that practitioners often miss, and it is worth stating carefully because it cuts against intuition.

The crucial point about a placebo treatment test is that one should usually condition on the real treatment. The placebo treatment and the real treatment frequently share causes that are not themselves confounders of the outcome, such as a common engineering reason for siting two kinds of transmitter together. If the real treatment genuinely affects the outcome and we fail to condition on it, the placebo treatment will inherit a spurious association with the outcome through that shared cause, and the test will fail even when the core design is perfectly valid. Conditioning on the real treatment closes this path. @eggers2024placebo note further that this conditioning is exactly what distinguishes a placebo treatment test from a placebo outcome test. If the placebo variable were related to the real treatment only through confounding, one would simply use it as a placebo outcome. The reason to run a placebo treatment test instead, and to condition on the actual treatment when doing so, is precisely that the placebo and real treatments share non-confounding causes. Their Supporting Information sharpens the point with a striking observation. Even a placebo treatment that the confounder does not touch can still detect that confounder once one conditions on the real treatment, because conditioning on the real treatment opens a collider path that runs through the confounder. The same logic that makes the placebo treatment informative also reveals that the shared cause is a valid instrument, so in some settings one can choose between testing the identification assumption with a placebo treatment and relaxing it with instrumental variables.

A placebo population test repeats the core analysis on a group for whom the treatment cannot operate but in which the same confounding should still be present. In the Russian television study the placebo population is Ukrainians who rely on satellite rather than terrestrial television, for whom terrestrial signal quality cannot drive viewing, so any remaining association between signal and vote choice would expose a violated assumption. Other canonical instances include comparing applicants who received disaster aid before an election with those who applied only after it, since aid awarded after the vote cannot change turnout in that vote, and comparing white Americans in the South, where a historical treatment operated, with white Americans in the North, where it did not. The validity of a placebo population test turns on whether the confounding structure really is the same across the two populations, which is a substantive judgment rather than a statistical one. If the placebo population is richer or more mobile or otherwise different in ways that change the confounding, the test can have inflated size or deflated power.

### Distributional Placebo Tests {#sec-placebo-distributional}

The placebo tests discussed so far probe bias, asking whether the point estimate is centered where the design claims. A second and less common family probes the distributional assumptions instead, asking whether the standard errors are right and therefore whether the nominal false positive rate is honest. The leading example in @eggers2024placebo is the reanalysis by @fowler2018shark of the claim that shark attacks off New Jersey cost Woodrow Wilson votes in 1916. The original finding rests on a sharp divergence between beach and inland counties that is assigned a small p-value under the assumption that idiosyncratic county shocks are independent. Fowler and Hall reconstruct the same comparison of coastal and non-coastal counties across twenty coastal states and more than a century of elections, restricting attention to the hundreds of state-years in which no shark attack occurred. If the original inferential assumptions were sound, the analysis should reject the no-difference null in about five percent of these placebo populations. Instead it rejects in roughly twenty-seven percent of them, which says that coastal and inland areas diverge far more often than chance alone would produce, and that the original standard errors were too small. Here the placebo analysis is not detecting bias in a point estimate. It is estimating the true false positive rate of the procedure directly, by running it many times in settings where the answer is known to be null. This same idea underlies the recommendation, taken up below, that regression discontinuity studies which test arbitrary cutoffs should test many of them rather than a handful, since only a large collection of placebo cutoffs yields a credible estimate of how often the procedure rejects when it should not.

### Testing Versus Relaxing Assumptions {#sec-placebo-test-vs-relax}

A theme that @eggers2024placebo press, and one that is easy to overlook, is that whenever the conditions hold for an informative placebo test there is usually an alternative procedure that relaxes the very assumption the placebo test merely checks. The choice between them is a real methodological decision rather than a foregone conclusion. A variable that qualifies as a placebo outcome, being a confounder or the descendant of one and being unaffected by the treatment, would often also serve as a control variable, in which case one could condition on it and weaken the identification assumption rather than hold it back to test that assumption. A placebo population test has differencing as its natural alternative, since assuming the bias is identical across the two populations turns the placebo comparison into a difference-in-differences estimator that removes the bias instead of testing for it. A distributional placebo test has its analogue in randomization-style inference, where the spread of estimates across placebo outcomes, treatments, or populations is used to build an empirical reference distribution and a corrected p-value for the core analysis, as Fowler and Hall also do when they report that their original-sized effect is exceeded in nearly a third of placebo state-years.

Why ever test rather than relax? @eggers2024placebo give two durable reasons. First, withholding a covariate to use as a placebo outcome supplies evidence about whether the maintained conditioning set was adequate in the first place, evidence that vanishes if every available covariate is simply thrown into the regression. Second, a variable can be useless as a control yet valuable as a placebo, for instance when it has no effect on the outcome but is associated with the treatment exactly when an unobserved confounder is. The deeper caution is that the freedom to choose between testing and relaxing, like the freedom to choose which outcome is real and which is placebo, is one more researcher degree of freedom. Because a researcher can run several candidate placebo tests and report only the comforting ones, or quietly soften a placebo result toward insignificance, the safeguards are the same ones that discipline the main analysis. Pre-register the placebo tests where possible, tether the placebo analysis as closely as possible to the core analysis so there is little room to maneuver, and report the precision of every placebo estimate so that readers can judge power for themselves rather than accept a null at face value.

### A Checklist for Placebo Tests {#sec-placebo-checklist}

@eggers2024placebo close with a set of questions that distill their framework into a usable standard, and they translate directly into a checklist for designing a placebo test or for refereeing one. The overarching question, from which the rest follow, is whether the test is genuinely more likely to fail when a core assumption is violated than when it holds.

1. Which core assumption does the test probe, a bias assumption tied to the point estimate, namely identification, estimation, measurement, or sample selection, or a distributional assumption tied to the standard error?
2. Which specific violation of that assumption is the most relevant threat to this design?
3. Which component of the core analysis has been altered to build the placebo, the outcome, the treatment, or the population?
4. Why should the treatment have no effect on the outcome in the placebo analysis, so that the no-effect benchmark is credible?
5. In what way does the placebo analysis resemble the core analysis closely enough that it would reproduce the relevant flaw, so that power exceeds size?
6. Could the placebo analysis suffer from a problem the core analysis does not have, which would inflate its false positive rate and make a failing result misleading?
7. Does the placebo test have enough statistical precision, judged from its standard error, to detect the violation it is meant to catch?

A placebo test that can answer all seven questions is far more informative than a lone null result presented as vindication, and authors who supply enough information for a reader to answer them raise the standard of the whole literature. The reusable templates in Section \@ref(sec-nc-workflow) operationalize several of these questions, and the precision-reporting discipline they enforce is a direct response to the seventh.

### Worked Logic

A simple simulation makes the reasoning concrete. We generate a confounded observational data set in which a confounder raises both the probability of treatment and the level of two outcomes. The first outcome is the real outcome of interest. The second is a negative-control outcome, constructed so that the treatment has no causal effect on it, yet it responds to the same confounder. A naive comparison of treated and untreated units will find a spurious effect on the negative-control outcome, and that spurious effect is the visible signature of the confounding that also biases the estimate on the real outcome.

```{r, eval = FALSE}
set.seed(42)
n <- 5000

# A single confounder drives both treatment and both outcomes
U <- rnorm(n)

# Treatment is more likely when the confounder is high
treat <- rbinom(n, 1, plogis(0.9 * U))

# Real outcome: treatment has a genuine effect of 1.5, plus confounding
y_real <- 1.5 * treat + 2.0 * U + rnorm(n)

# Negative-control outcome: treatment has NO effect, but shares the confounder
y_placebo <- 0.0 * treat + 2.0 * U + rnorm(n)

# Naive comparisons ignoring the confounder
real_effect <- coef(lm(y_real ~ treat))["treat"]
placebo_effect <- coef(lm(y_placebo ~ treat))["treat"]

real_effect      # biased upward, well above the true 1.5
placebo_effect   # should be 0 by construction, but appears positive
```

The naive estimate on the real outcome exceeds its true value of 1.5 because the confounder inflates it. The naive estimate on the negative-control outcome is positive even though the true effect is exactly zero. That nonzero placebo estimate is the falsification test firing. It tells us, without any knowledge of the true parameter, that the design is confounded and that the estimate on the real outcome cannot be trusted at face value. An analyst who then conditions on the confounder, where it is observed, would see both estimates move toward their correct values, with the placebo estimate collapsing toward zero. The placebo test thus doubles as a diagnostic for whether a proposed adjustment strategy has succeeded.

### Implementation Across Designs

Placebo and falsification tests adapt naturally to the major identification strategies, and each strategy suggests its own characteristic version of the test.

In matching and propensity-score studies, the most familiar falsification check is balance assessment. After matching, covariates that predate treatment should be distributed similarly across treated and control groups, because a valid matching procedure has removed their association with treatment. Standardized mean differences that remain large after matching indicate that the matching has failed to reproduce a randomized comparison. A complementary check estimates the treatment effect on a pretreatment outcome, which by temporal logic must be zero.

In difference in differences and event-study designs, the leading falsification test examines pretreatment periods. The coefficients on lead indicators, which capture differences between treated and control units before treatment begins, should be statistically indistinguishable from zero if the parallel trends assumption holds. Significant pretrends are the canonical placebo failure for these designs, signaling that the groups were already on diverging paths.

In regression discontinuity designs, falsification takes two forms. Covariates measured before the running variable was realized should be continuous at the cutoff, since units just above and just below the threshold should be comparable. A jump in a predetermined covariate at the cutoff suggests sorting or manipulation. Analysts also estimate the treatment effect at placebo cutoffs away from the true threshold, where no discontinuity should appear.

In instrumental-variables analyses, a falsification test checks whether the instrument predicts outcomes among subpopulations for whom the instrument cannot operate through the treatment channel, or whether it predicts predetermined covariates. An instrument correlated with pretreatment characteristics casts doubt on the exclusion restriction and on the assumption that the instrument is as good as randomly assigned.

### Pitfalls and Interpretation

Placebo tests are diagnostic tools, not proofs, and several cautions govern their use. The most important is the asymmetry of evidence already noted. Passing a falsification test cannot establish that the identifying assumption holds, because the assumption is fundamentally untestable and a finite battery of placebos can only probe particular violations. A clean placebo result raises confidence without conferring certainty. The strength of the evidence depends on how stringent the test was and on how plausibly a confounder that biased the main estimate would also have moved the placebo quantity.

A second pitfall is the power of the test. A placebo test that fails to reject the null may do so because the design is sound or merely because the test had too little power to detect a real problem. A negative-control outcome measured with heavy noise, or a placebo comparison based on a tiny subsample, can pass simply because the standard errors are large. Reporting the precision of a placebo estimate, and not only its statistical significance, is therefore essential. A tightly estimated zero is reassuring, while a wide confidence interval centered on zero is nearly uninformative. For the same reason, equivalence testing, which asks whether a placebo effect is small enough to be deemed negligible rather than merely insignificant, is often more appropriate than a conventional test against a point null [@hartman2018equivalence].

A third concern is the validity of the negative control itself. The entire argument depends on the control sharing the confounding structure of the primary analysis while having no causal link to the comparison variable. If the chosen control is in fact affected by the treatment through an unrecognized pathway, then a nonzero estimate is not evidence of confounding but of a real effect, and the test misleads. Conversely, if the control does not share the relevant confounders, a null result provides no reassurance about the main analysis, because there was never any bias for the control to detect. Choosing negative controls is therefore a substantive exercise in subject-matter reasoning rather than a mechanical one, and the credibility of the test rests on the credibility of those judgments.

Finally, multiplicity deserves attention. Researchers often run many falsification checks, and with enough of them some will reject by chance even in a perfectly valid design. The remedy is to specify the falsification tests in advance, to interpret the full pattern of results rather than cherry-picking the comforting ones, and to treat a single isolated rejection among many tests differently from a coherent pattern of failures. Used in this disciplined way, placebo and negative-control tests are among the most credible tools available for stress-testing a causal claim, precisely because they make the design vulnerable to refutation.

### Measurement-Error Diagnostics: Balance Tests vs. Coefficient Stability {#sec-pei-diagnostic}

A recurring question in applied work is how to use an imperfectly measured confounder. Suppose theory identifies a variable that ought to absorb the bias in a comparison, but the available measure of it is noisy. Two diagnostic strategies present themselves. The first, coefficient stability, adds the proxy to the regression and asks whether the treatment coefficient moves, treating a stable coefficient as evidence that confounding is mild. This logic underlies the family of selection-on-observables bounds discussed earlier in connection with @oster2019unobservable. The second, a balance test, puts the proxy on the left-hand side and asks whether the treatment predicts it, treating a null association as evidence that the confounder is balanced across treatment status.

@pei2019poorly show that when the confounder is poorly measured, these two strategies are not equally informative, and the balance test is usually the more powerful of the two. The intuition is a direct consequence of attenuation bias (Section \@ref(sec-proxy-variables)). A noisy proxy on the right-hand side is a weak control. Measurement error attenuates its ability to soak up the omitted variable, so the treatment coefficient can barely move even when confounding is severe, manufacturing a false impression of robustness. The same noisy proxy on the left-hand side suffers no such handicap as a detector. If the treatment is correlated with the latent confounder, and the proxy loads on that confounder, then the treatment will predict the proxy whether or not the proxy is measured precisely. Noise inflates the standard error of the balance coefficient, but it does not bias the test toward a null. A poorly measured confounder is therefore more useful on the left than on the right, which is the sense in which a balance test dominates a coefficient-stability check under measurement error.

The following simulation makes the comparison concrete. A latent confounder $U$ drives both the treatment $A$ and the outcome $Y$, and the analyst observes only a noisy proxy $X = U + \eta$. The treatment has no true effect on the outcome, so any apparent effect is pure confounding. We compute two diagnostics across a range of measurement-error magnitudes: the coefficient-stability check, which asks whether adding $X$ shifts the treatment coefficient, and the balance test, which regresses $X$ on $A$. For each level of noise we record how often each diagnostic correctly fires at the five percent level, that is, its power to detect the confounding we know is present.

```{r nc-pei-power, eval=FALSE}
library(tidyverse)

set.seed(1)

# One draw: latent confounder U, noisy proxy X = U + eta, no true effect of A on Y.
sim_once <- function(n = 2000, gamma = 1, pi = 1,
                     sigma_eps = 1, sigma_nu = 1, sigma_eta = 1) {
  U   <- rnorm(n)
  A   <- pi * U + rnorm(n, sd = sigma_nu)
  Y   <- gamma * U + rnorm(n, sd = sigma_eps)   # true effect of A on Y is zero
  X   <- U + rnorm(n, sd = sigma_eta)           # proxy measured with error

  # Coefficient-stability diagnostic: does adding X move A's coefficient?
  tau_short <- coef(lm(Y ~ A))["A"]
  tau_long  <- coef(lm(Y ~ A + X))["A"]
  diff_obs  <- unname(tau_short - tau_long)

  # Bootstrap reference distribution for the coefficient shift.
  B <- 200
  diffs <- replicate(B, {
    idx <- sample.int(n, n, replace = TRUE)
    ts  <- coef(lm(Y[idx] ~ A[idx]))[2]
    tl  <- coef(lm(Y[idx] ~ A[idx] + X[idx]))[2]
    unname(ts - tl)
  })
  p_diff <- mean(abs(diffs - mean(diffs)) >= abs(diff_obs - mean(diffs)))

  # Balance diagnostic: does A predict the proxy X?
  p_bal <- summary(lm(X ~ A))$coefficients["A", "Pr(>|t|)"]

  tibble(p_diff = p_diff, p_bal = p_bal)
}

power_curve <- function(sigmas, reps = 200, alpha = 0.05) {
  map_dfr(sigmas, function(s) {
    res <- map_dfr(seq_len(reps), ~ sim_once(sigma_eta = s))
    tibble(
      sigma_eta            = s,
      power_coef_stability = mean(res$p_diff < alpha),
      power_balance_test   = mean(res$p_bal  < alpha)
    )
  })
}

sigmas <- seq(0.2, 5, length.out = 10)
pow    <- power_curve(sigmas)

pow_long <- pow |>
  pivot_longer(cols = starts_with("power_"),
               names_to = "Diagnostic", values_to = "power") |>
  mutate(Diagnostic = recode(Diagnostic,
                             power_coef_stability = "Coefficient stability",
                             power_balance_test   = "Balance test"))

ggplot(pow_long, aes(x = sigma_eta, y = power, linetype = Diagnostic)) +
  geom_line() +
  geom_point() +
  labs(x = "Measurement-error SD in the proxy (sigma_eta)",
       y = "Probability of detecting confounding",
       linetype = "Diagnostic") +
  causalverse::ama_theme()
```

As the proxy becomes noisier, the coefficient-stability diagnostic loses power quickly, because a noisy $X$ cannot absorb the confounding and the treatment coefficient stops moving. The balance test degrades far more gracefully, since it only requires that the treatment remain correlated with the latent confounder through the proxy, not that the proxy fully control for it. The practical lesson carries over directly to the falsification tests described above. When an analyst regresses predetermined covariates or lagged outcomes on a treatment or an instrument, that is a balance test, and by the logic of @pei2019poorly it is often a sharper probe of residual confounding than comparing estimates across alternating sets of imperfect controls.

### A Formal Negative-Control Framework {#sec-nc-formal}

The epidemiology literature has developed the negative-control idea into a precise framework that deserves its own treatment, because it supplies both the assumptions that license a falsification test and the bridge to the bias-correction methods that follow. @shi2020selective give a careful synthesis, and their notation makes the assumptions transparent. Write $A$ for the primary treatment, $Y$ for the primary outcome, $X$ for measured covariates, and $U$ for an unmeasured confounder. A negative control outcome, written $W$, is a variable known not to be causally affected by the treatment. A negative control exposure, written $Z$, is a variable known not to causally affect the outcome. The power of the device comes from a second requirement layered on top of these known nulls. To be useful, the negative control should share the confounding mechanism of the primary analysis, so that it serves as a proxy for $U$.

The canonical illustration is the study by @jackson2006evidence of influenza vaccination and hospitalization in the elderly, which @shi2020selective use as their running example. The worry is that people who seek out the flu shot ($A$) are healthier and more health-conscious in unmeasured ways ($U$) that also lower their hospitalization risk ($Y$). Hospitalization for injury or trauma serves as a negative control outcome, since the vaccine cannot plausibly prevent a broken bone, yet health-seeking behavior should still predict it. The data bore this out. Vaccination appeared to cut injury and trauma hospitalization by about the same proportion as influenza hospitalization, a logically impossible protective effect that exposed the residual confounding contaminating the main estimate. Annual wellness visit history, which reflects the same health-seeking disposition but should not itself cause influenza hospitalization, serves as a negative control exposure in the same study.

Rather than invoke the usual no-unmeasured-confounding condition, @shi2020selective work from a weaker latent ignorability assumption, $A \perp Y(a) \mid U, X$, which permits an unmeasured confounder and then asks negative controls to expose or undo its effect. The negative controls are defined by exclusion restrictions: the negative control outcome satisfies $W \perp A \mid U, X$, meaning the treatment does not affect it once we condition on the confounder, and the negative control exposure satisfies the analogous condition that it does not affect $Y$. Layered on these is the U-comparability condition that makes the controls informative, namely $W \not\perp U \mid X$ and $Z \not\perp U \mid A, X$, which states that each control is genuinely associated with the confounder and so can act as its proxy. A control that satisfies the exclusion restriction but is unrelated to $U$ is a valid but useless negative control, since there is no bias for it to detect. One elegant consequence of this framework, noted by @shi2020selective, is that a valid instrumental variable is a special case of a negative control exposure, because an instrument is excluded from the outcome and, through the collider at the treatment, becomes a usable proxy for confounding once one conditions on the treatment.

### Designing and Validating Negative Controls {#sec-nc-design-validate}

Because the negative-control assumptions are causal, they cannot be checked from data alone and must be defended with subject-matter knowledge. @shi2020selective catalog the recurring strategies that applied researchers use to find credible controls, and they reduce to a few reusable templates. The most common exploits time. Since the future cannot cause the past, a future value of the exposure can serve as a negative control exposure, as in air pollution studies that use next period's pollutant level to detect confounding in this period's health effect, and a pre-treatment value of the outcome can serve as a negative control outcome, as when influenza hospitalization measured before the flu season circulates is used to probe a vaccine study. A second strategy exploits space, choosing a control that is too far away for the treatment to reach, such as pollutant levels in a distant city or, in a study of screening sigmoidoscopy, tumors in the proximal colon that lie beyond the instrument's reach. A third strategy chooses an outcome that arises through a mechanism known to be unrelated to the treatment, such as death from causes other than the one under study, or the injury hospitalizations of the flu example. A fourth, common in perinatal epidemiology, uses paternal exposure as a negative control for a maternal intrauterine effect, since the father's smoking shares the family-level confounders but cannot operate through the mother's womb.

Whatever the strategy, @shi2020selective recommend validating a candidate against three criteria. The control must be irrelevant to the causal channel, which is the exclusion restriction that the negative control exposure does not cause the outcome and the negative control outcome is not caused by the treatment. The control must be comparable in its confounding, meaning it is associated with the same unmeasured confounder so that it can proxy for the bias. And the control must have adequate power, since a control that is rare or only weakly tied to the confounder will fail to detect bias except in very large samples, which is the same precision concern that runs through this entire chapter. The framework also clarifies how negative controls fail, and the failure modes are instructive. If the control has no link to the confounder there is nothing for it to detect. If the outcome causes the negative control outcome, the treatment reaches it through the real causal path and the exclusion restriction breaks. If the negative control exposure and outcome both cause the confounder, conditioning patterns can open a collider path that manufactures a spurious association. The diversity of names in the literature, where a negative control outcome has been called a falsification outcome, an unaffected outcome, a control outcome, or a proxy for the confounder, reflects how often this single idea has been rediscovered, and @shi2020selective perform a service simply by unifying the vocabulary. The same device also extends beyond confounding to detect selection bias and measurement bias [@arnold2016negative].

### Detection, Reduction, and Correction {#sec-nc-detect-reduce-correct}

The deepest contribution of @shi2020selective is to organize negative-control methods along a ladder of ambition, from detecting bias, to reducing it, to removing it entirely. Detection is the rung occupied by the falsification tests of this chapter. Under the assumptions above, an association between the treatment and the negative control outcome, or between the negative control exposure and the outcome after adjusting for the treatment, is evidence of residual confounding, since the known null says there should be none. The adjustment for the treatment matters, and it is the same point raised earlier about placebo treatment tests, because without it a negative control exposure can pick up an association that merely reflects the treatment's genuine effect.

The second rung uses the negative control to subtract the bias rather than just signal it. The key idea is equi-confounding. Suppose the unmeasured confounder moves the primary outcome and the negative control outcome by the same amount, a condition @shi2020selective call additive outcome equi-confounding. Then the apparent effect of the treatment on the negative control outcome, which can only be bias since the treatment has no real effect there, equals the bias in the primary analysis, and subtracting one from the other recovers the causal effect. When the negative control outcome is simply the pre-treatment value of the outcome, this subtraction is exactly difference-in-differences, a connection @sofer2016negative develop in detail and a reminder that a familiar workhorse is a negative-control method in disguise. A small simulation shows the detection and reduction steps together. A confounder inflates the treatment-outcome association, a negative control outcome that loads on the same confounder reveals the bias, and subtracting recovers the truth.

```{r nc-detect-correct, eval=FALSE}
set.seed(7)
n <- 5000

U <- rnorm(n)                       # unmeasured confounder
A <- rbinom(n, 1, plogis(U))        # treatment confounded by U
tau <- 1.0                          # true effect of A on Y

Y <- tau * A + 1.5 * U + rnorm(n)   # outcome: U loads with weight 1.5
W <- 0   * A + 1.5 * U + rnorm(n)   # NCO: no treatment effect, same U loading

naive     <- coef(lm(Y ~ A))["A"]            # biased above the truth
nco_test  <- summary(lm(W ~ A))$coefficients["A", ]  # detection
bias_hat  <- coef(lm(W ~ A))["A"]            # treatment "effect" on NCO = the bias
corrected <- naive - bias_hat                # equi-confounding correction

naive       # well above 1.0 because U inflates it
nco_test    # significant: the design is confounded
corrected   # back near the true value of 1.0
```

The treatment cannot really affect the negative control outcome, so the coefficient the regression assigns to it is pure confounding, and it is statistically distinguishable from zero, which is the detection result. Subtracting that estimated bias from the naive estimate returns it to the neighborhood of the true effect, which is the reduction result, valid here because the confounder was constructed to load equally on the two outcomes. Equi-confounding is a strong assumption, and the third rung of the ladder relaxes it by bringing in a second control. With both a negative control exposure and a negative control outcome, the so-called double negative control design, one no longer needs the bias to be equal across outcomes. The negative control outcome reveals the bias up to an unknown scale, and the ratio of the negative control exposure's associations with the outcome and with the negative control outcome recovers that scale, so the true effect is identified nonparametrically. This is the identification result of @miao2018identifying, extended to categorical confounders by @shi2020multiply and surveyed by @tchetgen2024proximal, and it leads directly to the proximal estimator developed next.

### From Detection to Correction: Linear Proximal Estimation {#sec-proximal-correction}

The diagnostics above stop at detection, or reach correction only under the strong equi-confounding assumption. The double negative control design removes that crutch. Proximal causal inference takes the additional step of using a pair of negative controls to correct the bias rather than merely flag it, and the cross-references to the proxy-variable discussion in Section \@ref(sec-proxy-variables) supply the identification theory. In the language of Section \@ref(sec-nc-formal), the treatment-side proxy is a negative control exposure and the outcome-side proxy is a negative control outcome, and the two-stage estimator below is precisely the linear implementation that @shi2020selective present as a simple recipe runnable with standard instrumental-variable software. The essential requirement is a pair of proxies for the unmeasured confounder $U$: a treatment-side proxy $Z$, associated with $U$ and with the treatment but excluded from the outcome except through $U$, and an outcome-side proxy $W$, associated with $U$ and with the outcome but unaffected by the treatment. Under completeness conditions on how these proxies relate to $U$, the causal effect is identified even though $U$ is never observed, a result due to @miao2018identifying and surveyed by @tchetgen2024proximal.

In the linear-Gaussian special case the estimator has a transparent two-stage form that mirrors two-stage least squares. The first stage regresses the outcome-side proxy $W$ on the treatment, the treatment-side proxy $Z$, and the covariates, producing fitted values that purge $W$ of the variation correlated with the treatment-side proxy. The second stage regresses the outcome on the treatment, these fitted values, and the covariates. The coefficient on the treatment in the second stage is consistent for the causal effect. The following simulation shows the correction at work. The treatment loads heavily on a confounder that is omitted from any regression, so a naive estimate is badly biased, yet the two proxies recover the true effect of one.

```{r nc-proximal-linear, eval=FALSE}
library(sandwich)
library(lmtest)

set.seed(2)
n <- 5000

U <- rnorm(n)            # unmeasured confounder
X <- rnorm(n)            # measured covariate

A <- 0.8 * U + 0.3 * X + rnorm(n)             # treatment, confounded by U
Y <- 1.0 * A + 1.2 * U + 0.2 * X + rnorm(n)   # true effect of A on Y is 1.0

Z <- 1.0 * U + 0.2 * X + rnorm(n)   # treatment-inducing proxy (excluded from Y given U)
W <- 1.0 * U + 0.2 * X + rnorm(n)   # outcome-inducing proxy (unaffected by A)

dat <- data.frame(Y, A, X, Z, W)

# Naive regression: U is omitted, so the estimate of A is biased upward.
m_naive <- lm(Y ~ A + X, data = dat)

# Linear proximal two-stage estimator.
stage1    <- lm(W ~ A + Z + X, data = dat)
dat$What  <- predict(stage1)
stage2    <- lm(Y ~ A + What + X, data = dat)

coeftest(m_naive, vcov = vcovHC(m_naive, type = "HC1"))["A", ]
coeftest(stage2,  vcov = vcovHC(stage2,  type = "HC1"))["A", ]
```

The naive coefficient on the treatment sits well above the true value of one, because the treatment carries the confounder with it. The proximal estimate returns to the neighborhood of one, having used the proxy pair to net out the unmeasured confounding. The linear version shown here is the most accessible case, but it is not the most general. When the proxies and confounder are categorical rather than continuous, identification can proceed through matrix and rank conditions, and @shi2020multiply develop multiply robust estimators that remain consistent even if part of the model is misspecified, a property that is valuable when the proxy relationships are difficult to specify exactly. The broader nonparametric machinery, including the bridge-function formulation and its estimation, is developed in Section \@ref(sec-proxy-variables) and is best treated there. The point to carry forward is that the same negative controls that power the falsification tests in this chapter can, under stronger but explicit assumptions, be promoted from diagnostics into a constructive identification strategy.

### Practical Workflow Templates {#sec-nc-workflow}

The tests described throughout this chapter are easy to state and easy to run inconsistently. A small set of reusable templates helps enforce discipline. The templates below use the `fixest` package for fast fixed-effects estimation with clustered standard errors, and they assume a panel data frame with a unit identifier, a time variable, a treatment or instrument, several candidate outcomes, and an indicator marking units that the mechanism could plausibly reach. They are written as functions so that the same code runs every balance test, every subgroup reduced form, and the multiplicity adjustment that ties them together. The chunks are marked non-evaluating because they reference a dataset that is defined by the analyst rather than in the text.

The first template runs a balance test for each candidate outcome, regressing it on the treatment or instrument while absorbing the panel fixed effects, and returns one tidy row per outcome. Predetermined covariates and lagged outcomes belong here, since under a valid design the treatment should not predict anything measured before it was assigned.

```{r nc-balance-template, eval=FALSE}
library(fixest)
library(tidyverse)
library(broom)

# Panel data frame `panel_df` is assumed to contain:
#   id, year, z_treat, several outcome columns, and an `eligible` indicator.

run_balance_tests <- function(df, z, outcomes,
                              fe = "id + year", cluster = "id") {
  map_dfr(outcomes, function(y) {
    fml <- as.formula(paste0(y, " ~ ", z, " | ", fe))
    m   <- feols(fml, data = df, cluster = as.formula(paste0("~", cluster)))
    tidy(m) |>
      filter(term == z) |>
      transmute(outcome = y, estimate, std.error, statistic, p.value)
  })
}

# bal <- run_balance_tests(
#   df = panel_df, z = "z_treat",
#   outcomes = c("pre_cites_1", "pre_cites_2", "pre_cites_3", "pre_pubs")
# )
```

The second template implements the negative-control-outcome test by subgroup. It estimates the reduced form separately among units the mechanism can reach and units it cannot, returning both so they can be compared. The estimate in the ineligible subgroup is the negative control, and it should be null if the exclusion and independence assumptions hold.

```{r nc-subgroup-template, eval=FALSE}
run_subgroup_rf <- function(df, z, y, subgroup_var = "eligible",
                            fe = "id + year", cluster = "id") {
  map_dfr(c(eligible = 1, ineligible = 0), function(g) {
    dfg <- df |> filter(.data[[subgroup_var]] == g)
    fml <- as.formula(paste0(y, " ~ ", z, " | ", fe))
    m   <- feols(fml, data = dfg, cluster = as.formula(paste0("~", cluster)))
    tidy(m) |> filter(term == z) |> mutate(group = g, outcome = y)
  }, .id = "subgroup")
}

# rf_nc <- run_subgroup_rf(panel_df, "z_treat", "y_cites")
```

The third template addresses the multiplicity concern raised in the discussion of pitfalls. Running many falsification checks invites false rejections, so the p-values from the templates above should be collected and adjusted before they are interpreted. The function applies the Benjamini-Hochberg procedure to control the false discovery rate, and reporting both the raw and adjusted p-values lets readers see the effect of the correction.

```{r nc-fdr-template, eval=FALSE}
adjust_fdr <- function(tbl, p_col = "p.value") {
  tbl |> mutate(p_fdr = p.adjust(.data[[p_col]], method = "BH"))
}

# bind_rows(bal, rf_nc) |> adjust_fdr()
```

Reporting the precision of each falsification estimate alongside its adjusted p-value, rather than its significance alone, keeps the interpretation honest. A tightly estimated null is reassuring, a wide interval around zero is not, and a coherent pattern of adjusted rejections is a far stronger signal of trouble than a single isolated one.
