## Publication Bias and p-Hacking {#sec-publication-bias-phacking}

The robustness checks discussed so far concern a single study. Publication bias and p-hacking are problems of the literature as a whole, and they distort the evidence that a reader of published work confronts even when every individual study is internally honest. A sensitivity analysis tells us whether one estimate would survive a plausible confounder. The question here is different and arguably deeper. If we could see all the studies that were ever run on a question, would the published subset paint an accurate picture, or has the process of selection and analysis systematically tilted what we observe? Understanding these distortions, and the methods that have been developed to detect them, is an essential part of reading and producing credible empirical research.

### The File-Drawer Problem

The classical statement of the problem is the file-drawer problem of @rosenthal1979file. Suppose that for a given hypothesis there is a population of studies, some finding statistically significant effects and some finding null results. If journals prefer to publish significant findings, the significant studies enter the literature while the null studies stay in researchers' file drawers. A reader who synthesizes the published record then sees a biased sample, overstating both the probability that the effect is real and its average magnitude. Rosenthal framed the worry quantitatively by asking how many unpublished null results would have to exist to overturn a published significant finding, a fail-safe number that captures the fragility of conclusions drawn from a selected literature.

The bias operates through the relationship between an estimate and its statistical significance. Because significance depends on the ratio of an estimate to its standard error, selection on significance is also selection on large estimates, especially in small or noisy studies where only an unusually large draw can clear the threshold. The published literature therefore tends to exaggerate effect sizes, an effect sometimes called the winner's curse of significance testing. @ioannidis2005why argued that under realistic assumptions about prior plausibility, statistical power, and selective reporting, a substantial fraction of published positive findings may not reflect true effects at all, a claim that helped catalyze the broader concern with research credibility.

Direct evidence that selection operates at the publication stage comes from @franco2014publication, who tracked a large set of social-science experiments from a registry through to eventual publication. They found that null and weak results were far less likely to be written up and published than strong positive results, and that much of the attrition occurred because authors declined to submit unpromising findings rather than because journals rejected them. The file drawer, in other words, is filled partly by editors and partly by researchers anticipating editorial preferences.

### p-Hacking and Researcher Degrees of Freedom

Publication bias selects among completed studies. p-Hacking operates within a study, before any result is reported. Empirical analysis involves a long sequence of defensible choices, including which observations to exclude, which covariates to include, how to define variables, which functional form to adopt, and when to stop collecting data. Each choice is a researcher degree of freedom, and when an analyst tries several and reports the specification that crosses the significance threshold, the reported p-value no longer means what it claims. A nominal five-percent test conducted after searching across many specifications has a true false-positive rate far above five percent.

The distinction between the two phenomena matters for detection. Publication bias censors a distribution of completed estimates, removing the insignificant ones. p-Hacking deforms the distribution from within, pushing estimates that would otherwise have fallen just short of significance over the line. The two leave different signatures in the data, and the most informative detection methods exploit those signatures rather than relying on any single summary statistic.

### Density and Caliper Tests Around Thresholds

If researchers selectively report and selectively specify so as to clear conventional significance cutoffs, then the distribution of published test statistics should be unusually dense just past those cutoffs and unusually sparse just short of them. This observation motivates a family of tests that examine the shape of the test-statistic distribution in a narrow window, or caliper, around a threshold such as the value that corresponds to a five-percent two-sided test.

@gerber2008publication implemented an early version of this idea, comparing the number of published results landing just above the significance threshold with the number landing just below it. Under the null of no selection or manipulation, the distribution of test statistics should be smooth across the threshold, so a sharp excess of results on the significant side and a deficit on the insignificant side is evidence of selection. The caliper test formalizes this comparison within a chosen bandwidth and tests whether the over-representation of marginally significant results exceeds what sampling variation alone would produce.

The most influential application of these ideas in economics is the body of work by Abel Brodeur and coauthors. @brodeur2016star assembled tens of thousands of test statistics from articles in leading economics journals and documented a striking pattern. The distribution of z-statistics exhibited a pronounced dip just below the conventional significance threshold and a corresponding hump just above it, a two-humped shape inconsistent with what selection alone would produce and indicative of specification searching that nudges marginal results across the line. @brodeur2020methods extended this analysis to compare research designs, finding that the telltale excess of marginally significant results was more severe for methods with more discretion, such as instrumental variables and difference in differences, than for randomized controlled trials and regression discontinuity designs, where the analyst has fewer levers to pull. This comparison is itself a kind of placebo logic, using variation in the scope for manipulation across designs to attribute the distortion to researcher behavior rather than to genuine effects.

### p-Curve Analysis

A complementary approach examines the distribution of significant p-values themselves. The p-curve method of @simonsohn2014p starts from a statistical fact. When a tested effect is genuinely null, p-values are uniformly distributed, so among significant results the small p-values near zero and the larger ones near the threshold are equally likely. When a tested effect is real and the study has power, the distribution of significant p-values is right-skewed, with many very small values and few near the threshold. p-Hacking produces the opposite skew. Because hacking tends to generate results that just barely clear the cutoff, a literature built on p-hacked nulls shows a left-skewed p-curve, with results bunched just under the threshold rather than near zero.

By examining only the statistically significant findings on a topic and testing whether their distribution is right-skewed, flat, or left-skewed, the p-curve provides a diagnostic that distinguishes evidential value from its absence. A right-skewed curve indicates that the studies contain real evidential value. A flat or left-skewed curve indicates that the significant findings are consistent with selective reporting of nulls and carry little evidential weight. @simonsohn2014pa extended the method to recover effect-size estimates from significant results while correcting for the selection that publication bias induces, so that the p-curve serves both as a detection tool and as a bias-corrected estimator. A practical caution is that the p-curve assumes the analyst can identify the correct test corresponding to each study's hypothesis and that the studies share a common effect, assumptions that require careful judgment when assembling the curve.

### Formal Tests of the p-Value Distribution

The heuristic diagnostics above were put on rigorous econometric footing by @elliott2022detecting, who derive the sharpest testable implications that the absence of p-hacking places on the distribution of reported p-values. Their starting point is a characterization result: if researchers do not p-hack, then for a wide class of test statistics the density of p-values across studies must be *non-increasing*, and beyond monotonicity it must satisfy explicit upper bounds on its level and derivatives. These restrictions hold regardless of the distribution of true effects across studies, which is what makes them a valid null hypothesis for the whole literature rather than for any single study. p-Hacking that shifts marginal results across a threshold violates them by creating a local increase, a hump, in the p-curve just below the significance cutoff.

The paper turns each restriction into a statistical test. A binomial test compares the mass of p-values in the upper and lower halves of a narrow window such as $[0.04, 0.05]$, since a non-increasing density implies the upper half can hold at most half of the window's mass. Fisher's test aggregates evidence of non-uniformity across the window. A density discontinuity test checks for a jump in the p-curve at the threshold itself. The most powerful members of the family are conditional chi-squared tests in the tradition of Cox and Shi, which test the full set of inequality restrictions (monotonicity together with the bounds on the density and its first derivative) on a histogram of p-values. A final test exploits the fact that the restrictions imply concavity of the p-value CDF and measures the distance between the empirical CDF and its least concave majorant (LCM). Applying this battery to the large collections of test statistics assembled by Brodeur and coauthors, @elliott2022detecting reject the no-p-hacking null decisively in observational research designs, and much less so for randomized experiments, mirroring the discretion gradient documented by @brodeur2020methods.

A minimal version of the binomial test illustrates the logic. We simulate an honest literature, in which p-values inherit a decreasing density from a mix of true effects, and a hacked literature in which some near-misses are nudged just under the threshold, and then ask whether the upper half of the $[0.04, 0.05]$ window is too full.

```{r phacking-binomial-test}
set.seed(2026)
n_tests <- 5000

# Honest literature: z-statistics from studies with true effects
z_honest <- abs(rnorm(n_tests, mean = 1))
p_honest <- 2 * pnorm(-z_honest)

# Hacked literature: nudge 40% of near-misses to just below 0.05
p_hacked <- p_honest
marginal <- which(p_hacked > 0.05 & p_hacked < 0.09)
nudged   <- sample(marginal, floor(0.4 * length(marginal)))
p_hacked[nudged] <- runif(length(nudged), 0.045, 0.0499)

# Binomial test: under a non-increasing density, at most half of the
# p-values in [0.04, 0.05] should lie above the midpoint 0.045
binom_bunching <- function(p, lo = 0.04, hi = 0.05) {
    window <- p[p >= lo & p <= hi]
    k <- sum(window > (lo + hi) / 2)
    binom.test(k, length(window), p = 0.5, alternative = "greater")$p.value
}

c(honest = binom_bunching(p_honest), hacked = binom_bunching(p_hacked))
```

The honest literature survives the test while the hacked one fails it decisively. Two practical lessons from @elliott2022detecting deserve emphasis. First, rounding of reported statistics generates spurious mass points, so the tests should be run on de-rounded data or on a random draw of one test per article to avoid dependence across tests from the same study. Second, the tests differ sharply in power depending on where in the distribution the manipulation occurs, so the battery should be applied jointly rather than relying on any single member.

### Specification Curve and Multiverse Analysis

The within-study counterpart to detecting p-hacking after the fact is to forestall it by reporting the entire distribution of defensible specifications rather than a single chosen one. Specification curve analysis, developed by @simonsohn2020specification and treated in detail in Section \@ref(sec-specification-curve-analysis), enumerates the reasonable analytic choices, estimates the effect under every combination, and displays the resulting set of estimates together with a test of whether the observed distribution of results is more extreme than would be expected by chance. By making the full multiverse of choices visible, the approach removes the analyst's ability to present a favorable specification as if it were the only one and lets readers see how sensitive the conclusion is to discretionary decisions. In the context of publication bias and p-hacking, the specification curve is best understood as a disclosure device. It does not prevent a determined researcher from misreporting, but it raises the cost of selective presentation and converts the researcher degrees of freedom that enable p-hacking into a reported object that readers can scrutinize.

### Model-Based Correction for Selection

The detection methods above diagnose distortion. A separate line of work models the selection process directly and corrects estimates for it. @andrews2019identification develop a framework in which the probability that a study is published is allowed to depend on its results, typically on whether and in which direction it reached significance. By specifying this publication-probability function and combining it with the observed distribution of published estimates, they show how to identify the degree of selection and to recover bias-corrected estimates of effects and of their dispersion. Crucially, their approach can use either replication studies or known features of the original study designs to pin down the selection function, and it yields a formal estimate of how much the published literature overstates the underlying effects. This moves the analysis from a binary diagnosis of whether selection is present toward a quantitative adjustment for how severe it is.

### Detection in Practice

Bringing these methods to bear on a literature follows a common workflow. The analyst collects the relevant test statistics or p-values from published studies, often by extracting coefficients and standard errors from reported tables, and then applies one or more of the tools above. A density or caliper test around the significance threshold checks for the bunching signature of selection. A p-curve checks whether the significant results carry evidential value or merely reflect hacking. A selection model produces a bias-corrected effect estimate. The methods are complementary, and a credible assessment usually combines them, since each is sensitive to a somewhat different form of distortion and each rests on assumptions that may fail in a given application.

A small simulation illustrates the threshold-bunching signature that the caliper and density tests are designed to catch. We simulate a large set of true nulls, retain only those that reach significance as publication bias would, and additionally allow a fraction of marginal results to be hacked just over the threshold. The resulting distribution of z-statistics shows the characteristic excess just above the conventional cutoff.

```{r, eval = FALSE}
set.seed(2024)
n_studies <- 50000

# True effects are null, so z-statistics are standard normal
z <- rnorm(n_studies)

# Publication bias: keep results significant at the 5% two-sided level
threshold <- qnorm(0.975)  # about 1.96
published <- z[abs(z) > threshold]

# p-hacking: take some near-misses and nudge them just over the line
near_miss <- z[abs(z) > 1.6 & abs(z) <= threshold]
hacked <- sample(near_miss, size = floor(0.5 * length(near_miss)))
hacked <- sign(hacked) * (threshold + 0.05)

z_observed <- c(published, hacked)

# Compare density just below vs. just above the threshold (a caliper test)
below <- mean(abs(z_observed) > 1.85 & abs(z_observed) <= threshold)
above <- mean(abs(z_observed) > threshold & abs(z_observed) <= 2.07)
c(just_below = below, just_above = above)
```

The mass just above the threshold exceeds the mass just below it, the asymmetry that a caliper test formalizes and that @brodeur2016star documented in real economics articles. Several R packages support this kind of analysis on real data. The `phack` package, which wraps the detection code accompanying the literature on testing for p-hacking, provides tools for examining the distribution of test statistics, and dedicated p-curve and specification-curve software implements the methods of @simonsohn2014p and @simonsohn2020specification. Because the choice of which test statistic represents each study, which bandwidth to use, and which selection model to assume all affect the conclusions, these tools should be applied with the same transparency that they are meant to enforce, with the analytic choices preregistered or at least reported in full.

### Implications for the Working Researcher

For the producer of research, the lesson of this literature is preventive rather than diagnostic. The distortions arise from selection on significance, so the remedies attack that selection at its source. Preregistration commits the analyst to a specification before seeing the results, shrinking the space of researcher degrees of freedom. Reporting all specifications through a specification curve removes the temptation to present only the favorable one. Valuing and publishing null results, as registered reports and results-blind review aim to do, drains the file drawer. None of these practices is a panacea, and each imposes costs, but together they reduce the gap between the studies that are run and the studies that are seen. The detection methods then serve as an audit of how well a literature has lived up to these standards, and as a quantitative correction when it has not.
