# Structural Models of Auctions {#sec-structural-auctions}

This chapter belongs to the structural econometrics cluster, alongside the treatments of demand estimation, dynamic discrete choice, and structural selection. Auctions occupy a privileged place in that cluster because the mapping from primitives to data is unusually explicit. A bidder holds a valuation that the analyst never sees, the auction rules and the equilibrium of the bidding game turn that valuation into a bid, and the bid is recorded. The structural program reverses this arrow. Given the rules and a behavioral assumption about equilibrium play, the observed distribution of bids is enough to recover the latent distribution of valuations, and from there to answer questions about revenue, efficiency, and optimal design that no reduced form description of bids alone can address.

The reason auctions are a canonical structural setting is that the equilibrium **bid function** is a known mapping from valuations to bids. Once the analyst commits to a format and a paradigm, that mapping is a tight theoretical restriction rather than a free regression specification. Inverting it converts a problem about unobservable preferences into a problem about observable bid distributions, which can be estimated by ordinary nonparametric tools. The valuation distribution is the object the economist wants, and the auction structure is what makes it recoverable.

```{r auc-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(46801)
```

## Why Auctions Are a Canonical Structural Setting {#sec-auction-canonical}

Consider a single indivisible object sold to one of $n$ bidders. Bidder $i$ holds a valuation $v_i$, a private number measuring willingness to pay, drawn from a distribution $F$ with density $f$ on a support $[\underline{v}, \overline{v}]$. The analyst observes the bids $b_i$ that bidders submit but not the valuations $v_i$ that generated them. In equilibrium each bidder follows a strategy $b_i = \beta(v_i)$, a function that the theory pins down given the format and the informational environment. Because $\beta$ is strictly increasing under standard conditions, it can be inverted, so that $v_i = \beta^{-1}(b_i)$. The distribution of bids $G$ and the distribution of valuations $F$ are then linked by the change of variables

$$
G(b) = F\big(\beta^{-1}(b)\big),
$$

and the structural task is to use the observed $G$, together with the equilibrium restriction that fixes $\beta$, to back out the unobserved $F$. The power of the approach is that $F$ is a primitive: it does not change when the auction format or the reserve price changes, so once recovered it supports counterfactuals about designs that were never run. This is the structural payoff that a purely descriptive model of bids cannot deliver.

## Formats and Paradigms {#sec-auction-formats}

The four classic single-object formats divide into two pairs. In the **first-price sealed-bid** auction each bidder submits one sealed bid and the highest bidder wins and pays the bid. In the **Dutch** auction a price descends from a high level until a bidder accepts, which is strategically equivalent to the first-price auction because the only decision is the price at which to commit. In the **second-price sealed-bid** auction, the Vickrey auction, the highest bidder wins but pays the second highest bid, and in the ascending **English** auction the price rises until one bidder remains. The second-price and English formats share the feature that, under private values, bidding one's own valuation is a dominant strategy, so bids reveal valuations directly and the inversion problem is trivial. The first-price and Dutch formats shade bids below valuations by an amount that depends on competition and on beliefs about rivals, which is what makes their structural analysis interesting.

The informational paradigm matters as much as the format. Under **independent private values** (IPV), each bidder knows her own valuation, valuations are independent draws from $F$, and one bidder's valuation carries no information about another's. Under the **common value** paradigm the object is worth the same unknown amount to all bidders, for example the recoverable oil under a tract, and each bidder sees only a private signal correlated with that common value. The intermediate and most general case is the **affiliated values** model of @milgrom1982theory, in which valuations or signals are positively dependent, so that a high signal for one bidder makes high signals for others more likely. Affiliation nests both private and common values and delivers the central comparative result that, on average, the English auction raises more revenue than the second-price auction, which in turn raises more than the first-price auction, a ranking that collapses to revenue equivalence only in the independent private values benchmark.

The common value paradigm introduces the **winner's curse**. The bidder who wins is the one whose signal was most optimistic, so conditioning on winning is bad news about the true value: the winner is precisely the bidder most likely to have overestimated. Rational bidders anticipate this and shade their bids to correct for the adverse selection inherent in winning. A bidder who bids naively, as if her signal were unbiased even conditional on winning, will systematically overpay. The winner's curse is not a behavioral mistake in the equilibrium model but a force that disciplined bidders build into their strategies, and detecting whether bidders account for it is one empirical reason to distinguish common from private values.

## Identification and Estimation {#sec-auction-identification}

### The First-Order Condition of the First-Price Auction {#sec-auction-foc}

Take the symmetric IPV first-price auction with $n$ bidders. A bidder with valuation $v$ choosing a bid $b$ wins when her bid exceeds the highest of the other $n-1$ bids, an event whose probability, under the equilibrium strategy, is $G(b)^{n-1}$ in terms of the bid distribution. She solves

$$
\max_{b}\; (v - b)\, G(b)^{n-1},
$$

trading the surplus $v - b$ she earns on winning against the probability of winning. The first-order condition equates the marginal gain from a higher chance of winning against the marginal cost of a smaller margin,

$$
(v - b)\,(n-1)\, G(b)^{n-2} g(b) - G(b)^{n-1} = 0,
$$

where $g$ is the density of bids. Solving for the valuation yields the relation at the heart of the modern literature,

$$
v = b + \frac{1}{n-1}\,\frac{G(b)}{g(b)}.
$$

This expression says that a bidder's valuation equals her bid plus a markup that is large when bids are sparse near $b$, that is, when $g(b)$ is small, and that vanishes as competition $n$ grows. The valuation on the left is unobserved, but every object on the right, the bid $b$, the bid distribution $G$, and the bid density $g$, is estimable from data on bids alone.

### The Guerre, Perrigne, and Vuong Estimator {#sec-auction-gpv}

@guerre2000optimal turned this observation into a fully nonparametric two-step estimator that requires no parametric assumption on $F$ and no numerical solution of the equilibrium differential equation. The logic is to treat the right-hand side of the first-order condition as a recipe for constructing a **pseudo-value** for every observed bid. In the first step, the analyst estimates the bid distribution $G$ by the empirical distribution function of the bids and the bid density $g$ by a kernel density estimator. In the second step, each bid is mapped to a pseudo-value through

$$
\widehat{v}_i = b_i + \frac{1}{n-1}\,\frac{\widehat{G}(b_i)}{\widehat{g}(b_i)},
$$

and the collection of pseudo-values $\{\widehat{v}_i\}$ is treated as a sample from the valuation distribution $F$, whose density or quantiles are then estimated by any standard method. The estimator is appealing because it sidesteps the equilibrium inversion entirely: rather than solving for $\beta$ and inverting it, it uses the first-order condition to read valuations off bids directly. @guerre2000optimal established that this two-step procedure attains the optimal nonparametric rate for estimating $f$, which is why it has become the default empirical strategy for first-price auctions.

### Identification Results {#sec-auction-identification-theory}

The pseudo-value construction presumes that $F$ is identified from $G$ to begin with, and the general identification theory was settled by @athey2002identification. They asked, format by format and paradigm by paradigm, exactly which features of the model are recoverable from observed bids under the assumption of equilibrium play, treating the number of bidders, the presence of reserve prices, and the dependence structure as known or varying. For the symmetric IPV first-price auction their results confirm that $F$ is nonparametrically identified, which is the formal license for the @guerre2000optimal inversion. They also delineate the boundaries of identification, showing where additional structure or exclusion restrictions, such as exogenous variation in the number of bidders, become necessary to separate the model's components.

The common value case is far harder, a difficulty made precise by @laffont1995econometrics and the literature that followed. The obstacle is that private and common value models can generate identical distributions of observed bids, so that the data cannot, without further restrictions, tell whether bidders hold independent private valuations or correlated signals of a common value. Intuitively, the bid distribution summarizes equilibrium behavior but does not by itself reveal whether the shading in bids reflects private value markups or common value corrections for the winner's curse. Identifying common value models therefore relies on auxiliary variation, such as the response of bids to the number of competitors, since the winner's curse correction sharpens as more rivals are added in a way that pure private value models do not predict.

## Extensions {#sec-auction-extensions}

The baseline estimator extends in several directions that matter for applied work. **Risk aversion** breaks the neutral first-order condition above, because a risk averse bidder values the certainty of winning more and therefore bids more aggressively; the same observed bids are then consistent with a different and less dispersed valuation distribution, so risk attitudes and valuation heterogeneity are entangled and require either functional form restrictions or exclusion variation to separate. **Unobserved heterogeneity** across auctions, where some auctions attract systematically higher or lower valuations for reasons the analyst does not record, contaminates the pooled bid distribution and biases the naive inversion. @krasnokutskaya2011identification showed how to identify and strip out such auction level unobserved heterogeneity using deconvolution, exploiting the multiplicative or additive separability of the unobserved component from the idiosyncratic bidder draw. **Entry** introduces a prior stage in which potential bidders decide whether to incur a cost to participate, so the set of actual bidders is selected and the number of bidders is endogenous, which must be modeled jointly with the bidding stage to avoid confounding selection with valuation differences. Finally, **testing private versus common values** uses the comparative statics of bids in the number of competitors, since adding rivals intensifies the winner's curse correction under common values but only the competition effect under private values, giving a testable distinction grounded in the identification difficulties noted above.

## Policy and Counterfactual Design {#sec-auction-policy}

The reason to recover $F$ rather than merely describe bids is that $F$ supports counterfactual reasoning about designs the seller might choose. The leading application is the **optimal reserve price**. A reserve price excludes low valuations and so risks losing a sale, but it also forces bidders above the reserve to compete against a credible outside option, raising the price they pay. Knowledge of $F$ lets the analyst compute the reserve that maximizes expected revenue, which under regularity is the value $r$ solving $r - (1 - F(r))/f(r) = v_0$, where $v_0$ is the seller's own valuation, exactly the marginal revenue logic of monopoly pricing applied to the bidder with the reserve valuation. **Revenue comparisons** across formats follow the same template: with $F$ in hand the analyst simulates the first-price, second-price, and ascending formats and reports which raises more, recovering the revenue equivalence benchmark under IPV and the affiliation-driven ranking otherwise. More broadly, the recovered primitive underwrites **counterfactual mechanism design**, evaluating bidder subsidies, set-asides, reserve schedules, and entry promotion against the revenue and efficiency they would deliver, none of which is answerable from observed bids alone because each counterfactual changes the equilibrium bidding strategy and hence the bid distribution itself.

## Simulation: Estimating Valuations from First-Price Bids {#sec-auction-simulation}

To make the inversion concrete we simulate a symmetric IPV first-price auction in which valuations are uniform on the unit interval, a case for which the equilibrium bid function has the closed form $\beta(v) = v\,(n-1)/n$. We draw valuations, compute equilibrium bids from this closed form, and then pretend we observe only the bids. Applying the @guerre2000optimal inversion, we estimate the bid distribution by its empirical CDF and the bid density by a kernel, recover a pseudo-value for each bid through the first-order condition, and compare the distribution of recovered values against the known truth.

```{r auc-simulate}
# Symmetric IPV first-price auction with Uniform(0,1) valuations.
# Equilibrium bid function: beta(v) = v * (n - 1) / n.
n_bidders <- 5          # bidders per auction
n_auctions <- 4000      # number of independent auctions

# Draw one representative bidder's valuation per auction, then all bids.
# For the inversion we pool bids across auctions (same n in each).
v_true <- runif(n_auctions * n_bidders)          # latent valuations
b_obs  <- v_true * (n_bidders - 1) / n_bidders    # equilibrium bids

# Step 1: nonparametric estimates of the bid CDF and bid density.
G_hat <- ecdf(b_obs)                              # empirical bid CDF
dens  <- density(b_obs, n = 1024)                 # kernel bid density
g_hat <- approxfun(dens$x, dens$y, rule = 2)      # interpolate density

# Step 2: GPV inversion. Each bid maps to a pseudo-value via the FOC:
#   v = b + (1 / (n - 1)) * G(b) / g(b).
v_hat <- b_obs + (1 / (n_bidders - 1)) * G_hat(b_obs) / g_hat(b_obs)

# Trim a small fraction near the boundaries where the kernel density is
# unreliable, a standard precaution with the GPV estimator.
keep  <- b_obs > quantile(b_obs, 0.02) &
         b_obs < quantile(b_obs, 0.98)
v_hat_trim  <- v_hat[keep]
v_true_trim <- v_true[keep]
```

The pseudo-values in `v_hat` are the recovered valuations. Because the data generating process is known, we can check the recovery directly. The table below compares quantiles of the true valuation distribution against quantiles of the recovered one, and the figure overlays the two densities.

```{r auc-quantile-table}
probs <- c(0.10, 0.25, 0.50, 0.75, 0.90)
auc_tab <- data.frame(
  Quantile        = probs,
  Value_true      = round(quantile(v_true_trim, probs), 3),
  Value_recovered = round(quantile(v_hat_trim,  probs), 3)
)
auc_tab
```

The recovered quantiles sit close to the truth across the distribution. Figure \@ref(fig:auc-density-fig) overlays the true and recovered valuation densities, confirming that the inversion recovers the latent valuations from bids alone without ever using the true valuations in the estimation. The agreement is not mechanical: the estimator saw only `b_obs` and the assumed equilibrium structure.

```{r auc-density-fig, fig.cap="True versus recovered valuation densities for a symmetric IPV first-price auction with five bidders and Uniform(0,1) valuations. The recovered density is constructed by the Guerre, Perrigne, and Vuong inversion applied to observed bids only, with no use of the latent valuations.", fig.width=7, fig.height=4.2}
library(ggplot2)

plot_df <- rbind(
  data.frame(value = v_true_trim, series = "True valuations"),
  data.frame(value = v_hat_trim,  series = "Recovered (GPV)")
)

ggplot(plot_df, aes(value, color = series, fill = series)) +
  geom_density(alpha = 0.15, linewidth = 0.9) +
  labs(
    x = "Valuation",
    y = "Density",
    color = NULL, fill = NULL
  ) +
  causalverse::ama_theme()
```

The two densities are nearly coincident over the interior of the support, with the familiar boundary degradation that motivates the trimming: the kernel density of bids is biased near the edges of its support, which inflates the markup term $G(b)/g(b)$ there and distorts the recovered values at the extreme quantiles. The exercise is a self-contained demonstration of the structural logic of the chapter. The auction rules supplied the first-order condition, the first-order condition supplied the inversion, and the inversion turned a distribution of observed bids into the distribution of unobserved valuations that the seller would need to choose a reserve price or compare formats.

## Summary {#sec-structural-auctions-summary}

Auctions are the cleanest illustration of the structural method because the equilibrium bid function is a known and invertible map from unobserved valuations to observed bids. The format fixes the strategic problem, the informational paradigm fixes the dependence structure, and together they determine whether bids reveal valuations directly, as in the second-price and English auctions under private values, or shade them in a way that must be undone, as in the first-price and Dutch auctions. The @guerre2000optimal two-step estimator performs that undoing nonparametrically by reading valuations off bids through the first-order condition, an inversion licensed by the identification results of @athey2002identification for private values and complicated, in the common value case, by the observational equivalence noted by @laffont1995econometrics. Extensions for risk aversion, unobserved heterogeneity in the manner of @krasnokutskaya2011identification, and endogenous entry adapt the method to the messiness of field data, and the recovered valuation distribution is what makes optimal reserve prices, revenue comparisons, and counterfactual mechanism design tractable. The next chapters in the cluster carry the same recover-the-primitive logic into other equilibrium settings.
