## Penalized (Regularized) Estimators {#penalized-regularized-estimators}

Penalized or regularized estimators are extensions of [Ordinary Least Squares](#ordinary-least-squares) designed to address its limitations, particularly in high-dimensional settings. Regularization methods introduce a penalty term to the loss function to prevent overfitting, handle multicollinearity, and improve model interpretability.

There are three popular regularization techniques (but not limited to):

1.  [Ridge Regression]
2.  [Lasso Regression]
3.  [Elastic Net]

------------------------------------------------------------------------

### Motivation for Penalized Estimators

OLS minimizes the Residual Sum of Squares (RSS):

$$
RSS = \sum_{i=1}^n \left( y_i - \hat{y}_i \right)^2 = \sum_{i=1}^n \left( y_i - x_i'\beta \right)^2,
$$

where:

-   $y_i$ is the observed outcome,

-   $x_i$ is the vector of predictors for observation $i$,

-   $\beta$ is the vector of coefficients.

While OLS works well under ideal conditions (e.g., low dimensionality, no multicollinearity), it struggles when:

-   **Multicollinearity**: Predictors are highly correlated, leading to large variances in $\beta$ estimates.

-   **High Dimensionality**: The number of predictors ($p$) exceeds or approaches the sample size ($n$), making OLS inapplicable or unstable.

-   **Overfitting**: When $p$ is large, OLS fits noise in the data, reducing generalizability.

To address these issues, **penalized regression** modifies the OLS loss function by adding a **penalty term** that shrinks the coefficients toward zero. This discourages overfitting and improves predictive performance.

The general form of the penalized loss function is:

$$
L(\beta) = \sum_{i=1}^n \left( y_i - x_i'\beta \right)^2 + \lambda P(\beta),
$$

where:

-   $\lambda \geq 0$: Tuning parameter controlling the strength of regularization.

-   $P(\beta)$: Penalty term that quantifies model complexity.

Different choices of $P(\beta)$ lead to ridge regression, lasso regression, or elastic net.

------------------------------------------------------------------------

### Ridge Regression

Ridge regression, also known as **L2 regularization**, penalizes the sum of squared coefficients:

$$
P(\beta) = \sum_{j=1}^p \beta_j^2.
$$

The ridge objective function becomes:

$$
L_{ridge}(\beta) = \sum_{i=1}^n \left( y_i - x_i'\beta \right)^2 + \lambda \sum_{j=1}^p \beta_j^2,
$$

where:

-   $\lambda \geq 0$ controls the degree of shrinkage. Larger $\lambda$ leads to greater shrinkage.

Ridge regression has a closed-form solution:

$$
\hat{\beta}_{ridge} = \left( X'X + \lambda I \right)^{-1} X'y,
$$

where $I$ is the $p \times p$ identity matrix.

**Key Features**

-   Shrinks coefficients but does **not set them exactly to zero**.
-   Handles multicollinearity effectively by stabilizing the coefficient estimates [@hoerl1970].
-   Works well when all predictors contribute to the response.

**Example Use Case**

Ridge regression is ideal for applications with many correlated predictors, such as:

-   Predicting housing prices based on a large set of features (e.g., size, location, age of the house).

------------------------------------------------------------------------

### Lasso Regression

Lasso regression, or **L1 regularization**, penalizes the sum of absolute coefficients:

$$
P(\beta) = \sum_{j=1}^p |\beta_j|.
$$

The lasso objective function is:

$$
L_{lasso}(\beta) = \sum_{i=1}^n \left( y_i - x_i'\beta \right)^2 + \lambda \sum_{j=1}^p |\beta_j|.
$$

**Key Features**

-   Unlike ridge regression, lasso can set coefficients to **exactly zero**, performing automatic feature selection.
-   Encourages sparse models, making it suitable for high-dimensional data [@tibshirani1996].

**Optimization**

Lasso does not have a closed-form solution due to the non-differentiability of $|\beta_j|$ at $\beta_j = 0$. It requires iterative algorithms, such as:

-   **Coordinate Descent**,

-   **Least Angle Regression (LARS)**.

**Example Use Case**

Lasso regression is useful when many predictors are irrelevant, such as:

-   Genomics, where only a subset of genes are associated with a disease outcome.

------------------------------------------------------------------------

### Elastic Net

Elastic Net combines the penalties of ridge and lasso regression:

$$
P(\beta) = \alpha \sum_{j=1}^p |\beta_j| + \frac{1 - \alpha}{2} \sum_{j=1}^p \beta_j^2,
$$

where:

-   $0 \leq \alpha \leq 1$ determines the balance between lasso (L1) and ridge (L2) penalties.

-   $\lambda$ controls the overall strength of regularization.

The elastic net objective function is:

$$
L_{elastic\ net}(\beta) = \sum_{i=1}^n \left( y_i - x_i'\beta \right)^2 + \lambda \left( \alpha \sum_{j=1}^p |\beta_j| + \frac{1 - \alpha}{2} \sum_{j=1}^p \beta_j^2 \right).
$$

**Key Features**

-   Combines the strengths of lasso (sparse models) and ridge (stability with correlated predictors) [@zou2005a].
-   Effective when predictors are highly correlated or when $p > n$.

**Example Use Case**

Elastic net is ideal for high-dimensional datasets with correlated predictors, such as:

-   Predicting customer churn using demographic and behavioral features.

------------------------------------------------------------------------

### Tuning Parameter Selection

Choosing the regularization parameter $\lambda$ (and $\alpha$ for elastic net) is critical for balancing model complexity (fit) and regularization (parsimony). If $\lambda$ is too large, coefficients are overly shrunk (or even set to zero in the case of L1 penalty), leading to underfitting. If $\lambda$ is too small, the model might overfit because coefficients are not penalized sufficiently. Hence, a systematic approach is needed to determine the optimal $\lambda$. For elastic net, we also choose an appropriate $\alpha$ to balance the L1 and L2 penalties.

#### Cross-Validation

A common approach to selecting $\lambda$ (and $\alpha$) is $K$-Fold Cross-Validation:

1.  **Partition the data** into $K$ roughly equal-sized "folds."
2.  **Train the model** on $K-1$ folds and **validate** on the remaining fold, computing a validation error.
3.  Repeat this process **for all folds**, and compute the average validation error across the $K$ folds.
4.  **Select** the value of $\lambda$ (and $\alpha$ if tuning it) that **minimizes** the cross-validated error.

This method helps us maintain a good bias-variance trade-off because every point is used for both training and validation exactly once.

#### Information Criteria

Alternatively, one can use **information criteria**---like the Akaike Information Criterion (AIC) or the Bayesian Information Criterion (BIC)---to guide model selection. These criteria reward goodness-of-fit while penalizing model complexity, thereby helping in selecting an appropriately regularized model.

------------------------------------------------------------------------

### Properties of Penalized Estimators

1.  **Bias-Variance Tradeoff**:
    -   Regularization introduces some bias in exchange for reducing variance, often resulting in better predictive performance on new data.
2.  **Shrinkage**:
    -   Ridge shrinks coefficients toward zero but usually retains all predictors.
    -   Lasso shrinks some coefficients exactly to zero, performing inherent feature selection.
3.  **Flexibility**:
    -   Elastic net allows for a continuum between ridge and lasso, so it can adapt to different data structures (e.g., many correlated features or very high-dimensional feature spaces).

------------------------------------------------------------------------

```{r fig-coef-ridge, fig.cap="Coefficient Paths for Ridge Regression", fig.alt="Plot showing coefficient paths for ridge regression. The X-axis represents Log Lambda, ranging from -4 to 4, and the Y-axis represents Coefficients, ranging from -0.15 to 0.15. Multiple colored lines depict the change in coefficients as Log Lambda varies, converging towards zero as Log Lambda increases. The title reads 'Coefficient Paths for Ridge Regression.'", out.width="100%", fig.align='center'}
# Load required libraries
library(glmnet)

# Simulate data
set.seed(123)
n <- 100   # Number of observations
p <- 20    # Number of predictors
X <- matrix(rnorm(n * p), nrow = n, ncol = p)  # Predictor matrix
y <- rnorm(n)                                 # Response vector

# Ridge regression (alpha = 0)
ridge_fit <- glmnet(X, y, alpha = 0)
plot(ridge_fit, xvar = "lambda", label = TRUE)
title("Coefficient Paths for Ridge Regression")
```

-   In this plot, each curve represents a coefficient's value as a function of $\lambda$.

-   As $\lambda$ increases (moving from left to right on a log-scale by default), coefficients shrink toward zero but typically stay non-zero.

-   Ridge regression tends to shrink coefficients but does not force them to be exactly zero.

```{r fig-coef-lasso, fig.cap="Coefficient Paths for Lasso Regression", fig.alt="Graph showing coefficient paths for Lasso Regression. The x-axis represents Log Lambda, ranging from -7 to -2, and the y-axis represents Coefficients, ranging from -0.15 to 0.15. Multiple colored lines depict the paths of different coefficients as Log Lambda changes, illustrating how coefficients shrink towards zero as regularization increases.", out.width="100%", fig.align='center'}
# Lasso regression (alpha = 1)
lasso_fit <- glmnet(X, y, alpha = 1)
plot(lasso_fit, xvar = "lambda", label = TRUE)
title("Coefficient Paths for Lasso Regression")
```

Here, as $\lambda$ grows, several coefficient paths **hit zero exactly**, illustrating the variable selection property of lasso.

```{r fig-coef-elastic, fig.cap="Coefficient Paths for Elastic Net", fig.alt="Plot titled 'Coefficient Paths for Elastic Net (alpha = 0.5)' showing multiple colored lines representing coefficient paths against the x-axis labeled 'Log Lambda' and y-axis labeled 'Coefficients.' The lines converge towards zero as Log Lambda increases, illustrating the effect of regularization on coefficients in an Elastic Net model.", out.width="100%", fig.align='center'}
# Elastic net (alpha = 0.5)
elastic_net_fit <- glmnet(X, y, alpha = 0.5)
plot(elastic_net_fit, xvar = "lambda", label = TRUE)
title("Coefficient Paths for Elastic Net (alpha = 0.5)")
```

-   Elastic net combines ridge and lasso penalties. At $\lambda = 0.5$, we see partial shrinkage and some coefficients going to zero.

-   This model is often helpful when you suspect both group-wise shrinkage (like ridge) and sparse solutions (like lasso) might be beneficial.

We can further refine our choice of $\lambda$ by performing cross-validation on the lasso model:

```{r fig-mse-log-lambda, fig.cap="Cross-Validation Curve for Lasso Regression: Mean-Squared Error by Log Lambda", fig.alt="A line chart displaying the relationship between Log(lambda) on the x-axis and Mean-Squared Error on the y-axis. The chart features a series of red dots representing data points, which form a downward trend from left to right. Vertical gray lines indicate error bars for each data point. The x-axis ranges from -7 to -2, while the y-axis ranges from 0.75 to 1.05. The top of the chart shows numbers from 19 to 0, likely indicating additional data or context.", out.width="100%", fig.align='center'}
cv_lasso <- cv.glmnet(X, y, alpha = 1)
plot(cv_lasso)
best_lambda <- cv_lasso$lambda.min
best_lambda
```

-   The plot displays the cross-validated error (often mean-squared error or deviance) on the y-axis versus $\log(\lambda)$ on the x-axis.

-   Two vertical dotted lines typically appear:

    1.  $\lambda.min$: The $\lambda$ that achieves the minimum cross-validated error.

    2.  $\lambda.1se$: The largest $\lambda$ such that the cross-validated error is still within one standard error of the minimum. This is a more conservative choice that favors higher regularization (simpler models).

-   `best_lambda` above prints the numeric value of $\lambda.min$. This is the $\lambda$ that gave the lowest cross-validation error for the lasso model.

**Interpretation**:

-   By using `cv.glmnet`, we systematically compare different values of $\lambda$ in terms of their predictive performance (cross-validation error).

-   The selected $\lambda$ typically balances having a smaller model (due to regularization) with retaining sufficient predictive power.

-   If we used real-world data, we might also look at performance metrics on a hold-out test set to ensure that the chosen $\lambda$ generalizes well.

------------------------------------------------------------------------
