## Application

### Nonlinear Estimation Using Gauss-Newton Algorithm

This section demonstrates nonlinear parameter estimation using the Gauss-Newton algorithm and compares results with `nls()`. The model is given by:

$$
y_i = \frac{\theta_0 + \theta_1 x_i}{1 + \theta_2 \exp(0.4 x_i)} + \epsilon_i
$$

where

-   $i = 1, \dots ,n$
-   $\theta_0$, $\theta_1$, $\theta_2$ are the unknown parameters.
-   $\epsilon_i$ represents errors.

1.  Loading and Visualizing the Data

```{r}
library(dplyr)
library(ggplot2)
library(investr)

# Load the dataset
my_data <- read.delim("images/S21hw1pr4.txt", header = FALSE, sep = "") %>%
  dplyr::rename(x = V1, y = V2)
```

```{r fig-observed-data-nonlinear, fig.cap="Observed Data", fig.alt="Scatter plot titled 'Observed Data' showing a distribution of blue data points. The X-axis is labeled 'X' and the Y-axis is labeled 'Y.' The data points form a downward curve initially, then level out and slightly rise, indicating a potential trend or pattern in the data.", out.width="100%", fig.align='center'}
# Plot data
ggplot(my_data, aes(x = x, y = y)) +
  geom_point(color = "blue") +
  labs(title = "Observed Data", x = "X", y = "Y") +
  theme_minimal()
```

2.  Deriving Starting Values for Parameters

Since nonlinear optimization is sensitive to starting values, we estimate reasonable initial values based on model interpretation.

Finding the Maximum $Y$ Value

```{r}
max(my_data$y)
my_data$x[which.max(my_data$y)]
```

-   When $y = 2.6722$, the corresponding $x = 0.0094$.

-   From the model equation: $\theta_0 + 0.0094 \theta_1 = 2.6722$

Estimating $\theta_2$ from the Median $y$ Value

-   The equation simplifies to: $1 + \theta_2 \exp(0.4 x) = 2$

```{r}
# find mean y
mean(my_data$y) 

# find y closest to its mean
my_data$y[which.min(abs(my_data$y - (mean(my_data$y))))] 


# find x closest to the mean y
my_data$x[which.min(abs(my_data$y - (mean(my_data$y))))] 
```

-   This yields the equation: $83.58967 \theta_2 = 1$

Finding the Value of $\theta_0$ and $\theta_1$

```{r}
# find value of x closet to 1
my_data$x[which.min(abs(my_data$x - 1))] 

# find index of x closest to 1
match(my_data$x[which.min(abs(my_data$x - 1))], my_data$x) 

# find y value
my_data$y[match(my_data$x[which.min(abs(my_data$x - 1))], my_data$x)]
```

-   This provides another equation: $\theta_0 + \theta_1 \times 0.9895 - 2.164479 \theta_2 = 1.457$

Solving for $\theta_0, \theta_1, \theta_2$

```{r}
library(matlib)

# Define coefficient matrix
A = matrix(
  c(0, 0.0094, 0, 0, 0, 83.58967, 1, 0.9895, -2.164479),
  nrow = 3,
  ncol = 3,
  byrow = T
)

# Define constant vector
b <- c(2.6722, 1, 1.457)

# Display system of equations
showEqn(A, b)

# Solve for parameters
theta_start <- Solve(A, b, fractions = FALSE)
theta_start
```

3.  Implementing the Gauss-Newton Algorithm

Using these estimates, we manually implement the **Gauss-Newton optimization**.

**Defining the Model and Its Derivatives**

```{r}
# Starting values
theta_0_strt <- as.numeric(gsub(".*=\\s*", "", theta_start[1]))
theta_1_strt <- as.numeric(gsub(".*=\\s*", "", theta_start[2]))
theta_2_strt <- as.numeric(gsub(".*=\\s*", "", theta_start[3]))

# Model function
mod_4 <- function(theta_0, theta_1, theta_2, x) {
  (theta_0 + theta_1 * x) / (1 + theta_2 * exp(0.4 * x))
}

# Define function expression
f_4 = expression((theta_0 + theta_1 * x) / (1 + theta_2 * exp(0.4 * x)))

# First derivatives
df_4.d_theta_0 <- D(f_4, 'theta_0')
df_4.d_theta_1 <- D(f_4, 'theta_1')
df_4.d_theta_2 <- D(f_4, 'theta_2')
```

Iterative Gauss-Newton Optimization

```{r}
# Initialize
theta_vec <- matrix(c(theta_0_strt, theta_1_strt, theta_2_strt))
delta <- matrix(NA, nrow = 3, ncol = 1)
i <- 1

# Evaluate function at initial estimates
f_theta <- as.matrix(eval(f_4, list(
  x = my_data$x,
  theta_0 = theta_vec[1, 1],
  theta_1 = theta_vec[2, 1],
  theta_2 = theta_vec[3, 1]
)))

repeat {
  # Compute Jacobian matrix
  F_theta_0 <- as.matrix(cbind(
    eval(df_4.d_theta_0, list(
      x = my_data$x,
      theta_0 = theta_vec[1, i],
      theta_1 = theta_vec[2, i],
      theta_2 = theta_vec[3, i]
    )),
    eval(df_4.d_theta_1, list(
      x = my_data$x,
      theta_0 = theta_vec[1, i],
      theta_1 = theta_vec[2, i],
      theta_2 = theta_vec[3, i]
    )),
    eval(df_4.d_theta_2, list(
      x = my_data$x,
      theta_0 = theta_vec[1, i],
      theta_1 = theta_vec[2, i],
      theta_2 = theta_vec[3, i]
    ))
  ))
  
  # Compute parameter updates
  delta[, i] = (solve(t(F_theta_0)%*%F_theta_0))%*%t(F_theta_0)%*%(my_data$y-f_theta[,i])
    
  
  # Update parameter estimates
  theta_vec <- cbind(theta_vec, theta_vec[, i] + delta[, i])
  theta_vec[, i + 1] = theta_vec[, i] + delta[, i]
  
  # Increment iteration counter
  i <- i + 1
  
  # Compute new function values
  f_theta <- cbind(f_theta, as.matrix(eval(f_4, list(
    x = my_data$x,
    theta_0 = theta_vec[1, i],
    theta_1 = theta_vec[2, i],
    theta_2 = theta_vec[3, i]
  ))))
  
  delta = cbind(delta, matrix(NA, nrow = 3, ncol = 1))
  
  # Convergence criteria based on SSE
  if (abs(sum((my_data$y - f_theta[, i])^2) - sum((my_data$y - f_theta[, i - 1])^2)) / 
      sum((my_data$y - f_theta[, i - 1])^2) < 0.001) {
    break
  }
}

# Final parameter estimates
theta_vec[, ncol(theta_vec)]
```

4.  Checking Convergence and Variance

```{r}
# Final objective function value (SSE)
sum((my_data$y - f_theta[, i])^2)

sigma2 <- 1 / (nrow(my_data) - 3) * 
  (t(my_data$y - f_theta[, ncol(f_theta)]) %*% 
   (my_data$y - f_theta[, ncol(f_theta)]))  # p = 3

# Asymptotic variance-covariance matrix
as.numeric(sigma2)*as.matrix(solve(crossprod(F_theta_0)))
```

5.  Validating with `nls()`

```{r}
nlin_4 <- nls(
  y ~ mod_4(theta_0, theta_1, theta_2, x),
  start = list(
    theta_0 = as.numeric(gsub(".*=\\s*", "", theta_start[1])),
    theta_1 = as.numeric(gsub(".*=\\s*", "", theta_start[2])),
    theta_2 = as.numeric(gsub(".*=\\s*", "", theta_start[3]))
  ),
  data = my_data
)
summary(nlin_4)
```

### Logistic Growth Model

A classic logistic growth model follows the equation:

$$
P = \frac{K}{1 + \exp(P_0 + r t)} + \epsilon
$$

where:

-   $P$ = population at time $t$

-   $K$ = carrying capacity (maximum population)

-   $r$ = population growth rate

-   $P_0$ = initial population log-ratio

However, R's built-in `SSlogis` function uses a slightly different parameterization:

$$
P = \frac{asym}{1 + \exp\left(\frac{xmid - t}{scal}\right)}
$$

where:

-   $asym$ = carrying capacity ($K$)

-   $xmid$ = the $x$-value at the inflection point of the curve

-   $scal$ = scaling parameter

This gives the parameter relationships:

-   $K = asym$

-   $r = -1 / scal$

-   $P_0 = -r \cdot xmid$

```{r}
# Simulated time-series data
time <- c(1, 2, 3, 5, 10, 15, 20, 25, 30, 35)
population <- c(2.8, 4.2, 3.5, 6.3, 15.7, 21.3, 23.7, 25.1, 25.8, 25.9)
```

```{r fig-logistic-growth-example, fig.cap='Logistic Growth Model', fig.alt="Scatter plot titled Logistic Growth Model showing population growth over time. The x-axis shows time and the y-axis shows population. Data points indicate an initial slow increase in population, followed by a rapid rise and then a plateau, illustrating logistic growth", out.width="100%", fig.align='center'}
# Plot data points
plot(time,
     population,
     las = 1,
     pch = 16,
     main = "Logistic Growth Model")
```

```{r}
# Fit the logistic growth model using programmed starting values
logisticModelSS <- nls(population ~ SSlogis(time, Asym, xmid, scal))

# Model summary
summary(logisticModelSS)

# Extract parameter estimates
coef(logisticModelSS)
```

To fit the model using an alternative parameterization ($K, r, P_0$), we convert the estimated coefficients:

```{r}
# Convert parameter estimates to alternative logistic model parameters
Ks <- as.numeric(coef(logisticModelSS)[1])  # Carrying capacity (K)
rs <- -1 / as.numeric(coef(logisticModelSS)[3])  # Growth rate (r)
Pos <- -rs * as.numeric(coef(logisticModelSS)[2])  # P_0

# Fit the logistic model with the alternative parameterization
logisticModel <- nls(
    population ~ K / (1 + exp(Po + r * time)),
    start = list(Po = Pos, r = rs, K = Ks)
)

# Model summary
summary(logisticModel)

```

Visualizing the Logistic Model Fit

```{r fig-logistic-model-fit, fig.cap="Logistic Growth Model Fit", fig.alt="Chart titled Logistic Growth Model Fit showing population growth over time. The x-axis shows time and the y-axis shows population. Black data points are plotted with a red curve that begins low, rises steeply, and levels off to illustrate logistic growth", out.width="100%",    fig.align="center"}
# Plot original data
plot(time,
     population,
     las = 1,
     pch = 16,
     main = "Logistic Growth Model Fit")

# Overlay the fitted logistic curve
lines(time,
      predict(logisticModel),
      col = "red",
      lwd = 2)
```

### Nonlinear Plateau Model

This example is based on [@Schabenberger_2001] and demonstrates the use of a **plateau model** to estimate the relationship between soil nitrate ($NO_3$) concentration and relative yield percent (RYP) at two different depths (30 cm and 60 cm).

```{r fig-nonlinear-plateau, fig.cap='Nonlinear Plateau Plot', fig.alt='Scatter plot showing the relationship between soil NO3 concentration and relative yield percent. Data points are color-coded by depth: red for 30 cm and blue for 60 cm. The x-axis represents soil NO3 concentration, ranging from 0 to 100, and the y-axis represents relative yield percent, ranging from 30 to 110. Most data points cluster at higher yield percentages with lower NO3 concentrations.', out.width="100%", fig.align='center'}
# Load data
dat <- read.table("images/dat.txt", header = TRUE)

# Plot NO3 concentration vs. relative yield percent, colored by depth
library(ggplot2)
dat.plot <- ggplot(dat) + 
  geom_point(aes(x = no3, y = ryp, color = as.factor(depth))) +
  labs(color = 'Depth (cm)') + 
  xlab('Soil NO3 Concentration') + 
  ylab('Relative Yield Percent') +
  theme_minimal()

# Display plot
dat.plot
```

The **suggested nonlinear plateau model** is given by:

$$
E(Y_{ij}) = (\beta_{0j} + \beta_{1j}N_{ij})I_{N_{ij}\le \alpha_j} + (\beta_{0j} + \beta_{1j}\alpha_j)I_{N_{ij} > \alpha_j}
$$

where:

-   $N_{ij}$ represents the soil nitrate ($NO_3$) concentration for observation $i$ at depth $j$.

-   $i$ indexes individual observations.

-   $j = 1, 2$ corresponds to depths **30 cm** and **60 cm**.

This model assumes a **linear increase** up to a threshold ($\alpha_j$), beyond which the response **levels off (plateaus).**

Defining the Plateau Model as a Function

```{r}
# Define the nonlinear plateau model function
nonlinModel <- function(predictor, b0, b1, alpha) {
  ifelse(predictor <= alpha, 
         b0 + b1 * predictor,  # Linear growth below threshold
         b0 + b1 * alpha)      # Plateau beyond threshold
}
```

Creating a Self-Starting Function for `nls`

Since the model is **piecewise linear**, we can estimate starting values using:

1.  A **linear regression** on the **first half** of sorted predictor values to estimate $b_0$ and $b_1$.

2.  The **last predictor value** used in the regression as the plateau threshold ($\alpha$)

```{r}
# Define initialization function for self-starting plateau model
nonlinModelInit <- function(mCall, LHS, data) {
  # Sort data by increasing predictor value
  xy <- sortedXyData(mCall[['predictor']], LHS, data)
  n <- nrow(xy)
  
  # Fit a simple linear model using the first half of the sorted data
  lmFit <- lm(xy[1:(n / 2), 'y'] ~ xy[1:(n / 2), 'x'])
  
  # Extract initial estimates
  b0 <- coef(lmFit)[1]  # Intercept
  b1 <- coef(lmFit)[2]  # Slope
  alpha <- xy[(n / 2), 'x']  # Last x-value in the fitted linear range
  
  # Return initial parameter estimates
  value <- c(b0, b1, alpha)
  names(value) <- mCall[c('b0', 'b1', 'alpha')]
  value
}
```

Combining Model and Self-Start Function

```{r}
# Define a self-starting nonlinear model for nls
SS_nonlinModel <- selfStart(nonlinModel,
                            nonlinModelInit,
                            c('b0', 'b1', 'alpha'))
```

The `nls` function is used to estimate parameters separately for **each soil depth (30 cm and 60 cm).**

```{r}
# Fit the model for depth = 30 cm
sep30_nls <- nls(ryp ~ SS_nonlinModel(predictor = no3, b0, b1, alpha),
                  data = dat[dat$depth == 30,])

# Fit the model for depth = 60 cm
sep60_nls <- nls(ryp ~ SS_nonlinModel(predictor = no3, b0, b1, alpha),
                  data = dat[dat$depth == 60,])
```

We generate separate plots for **30 cm** and **60 cm** depths, showing both **confidence** and **prediction intervals.**

```{r fig-yield-depth-comparison, fig.cap="Fitted Curves for Relative Yield Response by Soil NO3 Concentration at 30 cm and 60 cm Depths", fig.alt="Two XY charts comparing relative yield percent to soil NO3 concentration at different depths. The left chart, titled Results at 30 cm Depth, shows data points clustered at higher yield percentages with a blue shaded confidence interval. The right chart, titled Results at 60 cm Depth, shows a similar pattern with a red shaded confidence interval. Both charts show a positive relationship between soil NO3 and relative yield percent",  out.width="100%", fig.align="center"}
# Set plotting layout
par(mfrow = c(1, 2))

# Plot model fit for 30 cm depth
investr::plotFit(
  sep30_nls,
  interval = "both",
  pch = 19,
  shade = TRUE,
  col.conf = "skyblue4",
  col.pred = "lightskyblue2",
  data = dat[dat$depth == 30,],
  main = "Results at 30 cm Depth",
  ylab = "Relative Yield Percent",
  xlab = "Soil NO3 Concentration",
  xlim = c(0, 120)
)

# Plot model fit for 60 cm depth
investr::plotFit(
  sep60_nls,
  interval = "both",
  pch = 19,
  shade = TRUE,
  col.conf = "lightpink4",
  col.pred = "lightpink2",
  data = dat[dat$depth == 60,],
  main = "Results at 60 cm Depth",
  ylab = "Relative Yield Percent",
  xlab = "Soil NO3 Concentration",
  xlim = c(0, 120)
)
```

```{r}
summary(sep30_nls)
summary(sep60_nls)
```

**Modeling Soil Depths Together and Comparing Models**

Instead of fitting separate models for different soil depths, we first fit a **combined model** where all observations share a **common slope, intercept, and plateau**. We then test whether modeling the two depths separately provides a significantly better fit.

1.  **Fitting a Reduced (Combined) Model**

The reduced model assumes that **all soil depths follow the same nonlinear relationship**.

```{r reduce-model}
# Fit the combined model (common parameters across all depths)
red_nls <- nls(
  ryp ~ SS_nonlinModel(predictor = no3, b0, b1, alpha), 
  data = dat
)

# Display model summary
summary(red_nls)
```

```{r fig-res-combined-mod, fig.cap="Results for Combined Model", fig.alt="Scatter plot titled Results for Combined Model showing the relationship between Soil NO3 concentration on the x-axis and relative yield percent on the y-axis. Data points are scattered with a trend line that increases rapidly and then levels off. A shaded band around the line represents variability or confidence intervals", out.width="100%", fig.align='center'}
# Visualizing the combined model fit
par(mfrow = c(1, 1))
plotFit(
  red_nls,
  interval = "both",
  pch = 19,
  shade = TRUE,
  col.conf = "lightblue4",
  col.pred = "lightblue2",
  data = dat,
  main = 'Results for Combined Model',
  ylab = 'Relative Yield Percent',
  xlab = 'Soil NO3 Concentration'
)
```


2.  **Examining Residuals for the Combined Model**

Checking residuals helps diagnose potential **lack of fit**.

```{r}
library(nlstools)

# Residual diagnostics using nlstools
resid <- nlsResiduals(red_nls)
```


```{r fig-resid-nonlinear, fig.cap="Residual Plot Nonlinear Example", fig.alt="Four panel figure showing residual analysis. Top left shows residuals versus fitted values labeled Residuals. Top right shows standardized residuals versus fitted values labeled Standardized Residuals. Bottom left shows autocorrelation of residuals i versus residuals i+1 labeled Autocorrelation. Bottom right shows a normal QQ plot comparing sample quantiles to theoretical quantiles labeled Normal QQ Plot of Standardized Residuals. Each plot includes a horizontal reference line", out.width="100%", fig.align="center"}
# Plot residuals
plot(resid)
```


If there is a **pattern** in the residuals (e.g., systematic deviations based on soil depth), this suggests that a **separate model for each depth** may be necessary.

3.  **Testing Whether Depths Require Separate Models**

To formally test whether soil depth significantly affects the model parameters, we introduce a **parameterization where depth-specific parameters are increments from a baseline model** (30 cm depth):

$$
\begin{aligned}
\beta_{02} &= \beta_{01} + d_0 \\
\beta_{12} &= \beta_{11} + d_1 \\
\alpha_{2} &= \alpha_{1} + d_a
\end{aligned}
$$

where:

-   $\beta_{01}, \beta_{11}, \alpha_1$ are parameters for **30 cm depth**.

-   $d_0, d_1, d_a$ represent **depth-specific differences** for **60 cm depth**.

-   If $d_0, d_1, d_a$ are significantly **different from 0**, the **two depths should be modeled separately**.

4.  **Defining the Full (Depth-Specific) Model**

```{r full-model, eval=TRUE}
nonlinModelF <- function(predictor, soildep, b01, b11, a1, d0, d1, da) {
  
  # Define parameters for 60 cm depth as increments from 30 cm parameters
  b02 <- b01 + d0
  b12 <- b11 + d1
  a2 <- a1 + da
  
  # Compute model output for 30 cm depth
  y1 <- ifelse(
    predictor <= a1, 
    b01 + b11 * predictor, 
    b01 + b11 * a1
  )
  
  # Compute model output for 60 cm depth
  y2 <- ifelse(
    predictor <= a2, 
    b02 + b12 * predictor, 
    b02 + b12 * a2
  )
  
  # Assign correct model output based on depth
  y <- y1 * (soildep == 30) + y2 * (soildep == 60)
  
  return(y)
}


```

5.  **Fitting the Full (Depth-Specific) Model**

The starting values are taken from the **separately fitted models** for each depth.

```{r full-model-fit}
Soil_full <- nls(
  ryp ~ nonlinModelF(
    predictor = no3,
    soildep = depth,
    b01,
    b11,
    a1,
    d0,
    d1,
    da
  ),
  data = dat,
  start = list(
    b01 = 15.2,   # Intercept for 30 cm depth
    b11 = 3.58,   # Slope for 30 cm depth
    a1 = 23.13,   # Plateau cutoff for 30 cm depth
    d0 = -9.74,   # Intercept difference (60 cm - 30 cm)
    d1 = 2.11,    # Slope difference (60 cm - 30 cm)
    da = -6.85    # Plateau cutoff difference (60 cm - 30 cm)
  )
)

# Display model summary
summary(Soil_full)

```

6.  **Model Comparison: Does Depth Matter?**

-   If $d_0, d_1, d_a$ **are significantly different from 0**, the depths should be modeled separately.

-   The **p-values** for these parameters indicate whether depth-specific modeling is necessary.

------------------------------------------------------------------------
