---
title: "LATE, ITT, and IV: Why ITT divided by Complier Share Must Match 2SLS, and Why It Often Does Not"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)

set.seed(12345)
options(scipen = 999)
```

In the standard binary instrument case, we often learn the mnemonic:

$$
\text{LATE} = \frac{\text{ITT}}{\text{Share of compliers}}.
$$

Then compute:

-   ITT as a mean difference in outcomes across assignment ($Z$)
-   share of compliers as some compliance rate
-   and compare to an IV or 2SLS estimate

They often do not match.

**Core message:** In the simple exactly identified design, the IV
coefficient equals a Wald ratio, and the denominator is the first stage
difference in treatment rates, not the take up rate in the treated
assignment group. Once you add controls, fixed effects, weights,
multiple instruments, or sample differences, you must compute the
adjusted versions of the same objects.

Let

-   $Z_i \in {0,1}$ is assignment, encouragement, or instrument
-   $D_i \in {0,1}$ is actual treatment uptake
-   $Y_i$ is the outcome

Potential treatment uptake:

-   $D_i(1)$ is treatment status if assigned $Z=1$
-   $D_i(0)$ is treatment status if assigned $Z=0$

Potential outcomes:

-   $Y_i(1)$ is outcome if treated
-   $Y_i(0)$ is outcome if untreated

Observed data:

$$
D_i = D_i(Z_i), \quad Y_i = Y_i(D_i).
$$

Define compliance type by $(D_i(1), D_i(0))$:

-   Complier: $(1,0)$
-   Always taker: $(1,1)$
-   Never taker: $(0,0)$
-   Defier: $(0,1)$

The complier share is $P(D_i(1)=1, D_i(0)=0)$.

For the LATE interpretation, we need the following assumptions.

1.  Random assignment (independence)

$$
Z_i \perp \left(Y_i(1), Y_i(0), D_i(1), D_i(0)\right).
$$

This holds when $Z$ is randomized, or as-if random conditional on
controls.

2.  Exclusion restriction

Assignment affects outcomes only through treatment:

$$
Y_i(d, z) = Y_i(d), 
$$ so $Y_i$ depends on $z$ only via $d$.

3.  Monotonicity (no defiers)

$$
D_i(1) \ge D_i(0) \forall i.
$$

This rules out defiers.

4.  Relevance

$$
\mathbb{E}[D_i \mid Z_i=1] \ne \mathbb{E}[D_i \mid Z_i=0].
$$

Without relevance, the denominator is zero and IV is undefined.

# Deriving the identity: LATE = ITT / first stage

## Step 1. The reduced form (ITT on outcomes)

Define the reduced form effect:

$$
ITT_Y = \mathbb{E}[Y \mid Z=1] - \mathbb{E}[Y \mid Z=0].
$$

Under A1 and A2:

$$
ITT_Y = \mathbb{E}[Y(D(1)) - Y(D(0))].
$$

Now decompose by compliance type. Under monotonicity:

-   Never takers have $D(1)=0, D(0)=0$, no change in treatment, so
    contribution is zero
-   Always takers have $D(1)=1, D(0)=1$, no change, contribution is zero
-   Compliers have $D(1)=1, D(0)=0$, treatment changes, contribution is
    $Y(1)-Y(0)$

So:

$$
ITT_Y = P(\text{complier}) \cdot \mathbb{E}[Y(1)-Y(0)
\mid \text{complier}].
$$

The second term is the Local Average Treatment Effect:

$$
LATE = \mathbb{E}[Y(1)-Y(0) \mid \text{complier}].
$$

Thus:

$$
ITT_Y = P(\text{complier}) \cdot LATE.
$$

## Step 2. The first stage on treatment uptake

Define:

$$
ITT_D = \mathbb{E}[D \mid Z=1] - \mathbb{E}[D \mid Z=0].
$$

Using potential treatments:

$$
ITT_D = \mathbb{E}[D(1) - D(0)].
$$

Under monotonicity:

-   Compliers contribute $1-0 = 1$
-   Always takers contribute $1-1 = 0$
-   Never takers contribute $0-0 = 0$

So:

$$
ITT_D = P(\text{complier}).
$$

## Step 3. Ratio form

Combine:

$$
LATE = \frac{ITT_Y}{ITT_D}.
$$

This is the Wald estimator.

**Important:** the denominator is not $P(D=1 \mid Z=1)$. It is the
difference $P(D=1 \mid Z=1) - P(D=1 \mid Z=0)$.

# The most common mistake: wrong "complier share"

Many people incorrectly use:

$$
\widehat{P(\text{complier})} \approx \bar D_{Z=1}.
$$

That is only correct if $\bar D_{Z=0} \approx 0$, meaning no always
takers and no treatment uptake in the control assignment group.

In general:

-   $\bar D_{Z=1} = P(\text{always}) + P(\text{complier})$
-   $\bar D_{Z=0} = P(\text{always})$

So:

$$
P(\text{complier}) = \bar D_{Z=1} - \bar D_{Z=0}.
$$

# Simulation 1: Exactly identified case, constant only

This simulation does three things:

1.  Shows Wald LATE equals 2SLS exactly (up to numerical tolerance)
2.  Shows using $\bar D_{Z=1}$ as "complier share" gives the wrong
    number
3.  Uses a data generating process with always takers present, so the
    mistake is visible

## Data generating process

We generate individuals with a compliance type, a randomized assignment
$Z$, treatment uptake consistent with the type, and outcomes where
treatment has an effect only for compliers. That makes the causal
estimand transparent.

```{r}
simulate_binary_iv <- function(
N = 20000,
p_complier = 0.40,
p_always = 0.10,
tau_c = 2.0,
sigma = 1.0
) {
stopifnot(p_complier >= 0, p_always >= 0, p_complier + p_always <= 1)

# Compliance types

types <- sample(
x = c("C", "A", "N"),
size = N,
replace = TRUE,
prob = c(p_complier, p_always, 1 - p_complier - p_always)
)

# Random assignment

Z <- rbinom(N, 1, 0.5)

# Potential treatment uptake

D0 <- ifelse(types == "A", 1, 0)                    # if Z=0
D1 <- ifelse(types == "A", 1, ifelse(types == "C", 1, 0))  # if Z=1
D  <- ifelse(Z == 1, D1, D0)

# Baseline unobserved factor (creates endogeneity if you regress Y on D)

U <- rnorm(N)

# Potential outcomes: baseline depends on U

Y0 <- 1.0 + 1.5 * U + rnorm(N, sd = sigma)

# Treatment effect only for compliers

tau <- ifelse(types == "C", tau_c, 0.0)
Y1  <- Y0 + tau

# Observed outcome

Y <- ifelse(D == 1, Y1, Y0)

data.frame(Y = Y, D = D, Z = Z, type = types, U = U)
}

wald_late <- function(df) {
ITT_Y <- mean(df$Y[df$Z == 1]) - mean(df$Y[df$Z == 0])
FS    <- mean(df$D[df$Z == 1]) - mean(df$D[df$Z == 0])
LATE  <- ITT_Y / FS
list(ITT_Y = ITT_Y, FS = FS, LATE = LATE)
}

## Compute Wald, 2SLS, and the wrong denominator

df1 <- simulate_binary_iv()

w1 <- wald_late(df1)

# 2SLS with intercept only:

# stage 1: D ~ Z

# stage 2: Y ~ Dhat

stage1 <- lm(D ~ Z, data = df1)
df1$Dhat <- fitted(stage1)
stage2 <- lm(Y ~ Dhat, data = df1)
beta_2sls <- coef(stage2)[["Dhat"]]

# Wrong "complier share" many people use:

takeup_assigned <- mean(df1$D[df1$Z == 1])
late_wrong <- w1$ITT_Y / takeup_assigned

# True shares in the realized sample

share_C <- mean(df1$type == "C")
share_A <- mean(df1$type == "A")


out1 <- data.frame(
  quantity = c(
    "ITT_Y = E[Y|Z=1] - E[Y|Z=0]",
    "First stage = E[D|Z=1] - E[D|Z=0] (equals complier share under monotonicity)",
    "Wald LATE = ITT_Y / First stage",
    "2SLS coefficient (intercept only)",
    "Wrong denominator: Pr(D=1|Z=1)",
    "ITT_Y / Pr(D=1|Z=1) (not LATE)",
    "Realized complier share in sample",
    "Realized always taker share in sample"
  ),
  value = c(
    w1$ITT_Y,
    w1$FS,
    w1$LATE,
    beta_2sls,
    takeup_assigned,
    late_wrong,
    share_C,
    share_A
  )
)

out1
```

## Interpretation

-   The Wald ratio equals the 2SLS coefficient because this is the
    exactly identified, constant only case.
-   Using $Pr(D=1|Z=1)$ as the denominator produces a different number
    because it includes always takers.
-   The correct denominator is the first stage difference in treatment
    rates.

# When you add controls: you must compare adjusted objects

The second most common reason for mismatch is that your IV regression
includes controls or fixed effects, but your manual ITT and first stage
are computed as raw mean differences.

In that case, 2SLS is no longer the raw Wald ratio. It is a residualized
Wald ratio.

## Algebra you must use with controls

Suppose you estimate:

-   First stage: $D = \gamma Z + X\delta + u$
-   Second stage: $Y = \beta D + X\theta + \varepsilon$

Then:

$$
\hat\beta_{2SLS} =
\frac{\widehat{\text{Cov}}(\tilde Z,\tilde Y)}{\widehat{\text{Cov}}(\tilde Z,\tilde D)}
$$

where tildes denote residuals after partialling out the constant and
$X$.

If you compute ITT and first stage using unconditional means, you are
estimating a different object than the regression is estimating.

# Simulation 2: Z correlated with X, controls fix the mismatch

Here we generate a covariate $X$ that affects both the probability of
assignment $Z$ and the outcome $Y$. Then:

-   naive ITT and naive first stage are confounded by $X$
-   2SLS with $X$ recovers the intended effect
-   residualized Wald matches 2SLS exactly

```{r}
simulate_with_covariate <- function(
N = 30000,
p_complier = 0.35,
p_always = 0.15,
tau_c = 2.0,
sigma = 1.0,
z_strength = 0.8
) {
stopifnot(p_complier >= 0, p_always >= 0, p_complier + p_always <= 1)

types <- sample(
x = c("C", "A", "N"),
size = N,
replace = TRUE,
prob = c(p_complier, p_always, 1 - p_complier - p_always)
)

X <- rnorm(N)

# Non random assignment: higher X increases Pr(Z=1)

pZ <- 1 / (1 + exp(-z_strength * X))
Z <- rbinom(N, 1, pZ)

D0 <- ifelse(types == "A", 1, 0)
D1 <- ifelse(types == "A", 1, ifelse(types == "C", 1, 0))
D  <- ifelse(Z == 1, D1, D0)

U <- rnorm(N)

# X affects outcomes, so unconditional mean differences are biased

Y0 <- 1.0 + 1.0 * X + 1.3 * U + rnorm(N, sd = sigma)
tau <- ifelse(types == "C", tau_c, 0.0)
Y1 <- Y0 + tau
Y  <- ifelse(D == 1, Y1, Y0)

data.frame(Y = Y, D = D, Z = Z, X = X)
}

residualize <- function(v, Xmat) {
fit <- lm.fit(x = Xmat, y = v)
v - Xmat %*% fit$coefficients
}

df2 <- simulate_with_covariate()

# Naive Wald using unconditional group means

w2 <- wald_late(df2)

# 2SLS with X as control (manual 2SLS)

fs2 <- lm(D ~ Z + X, data = df2)
df2$Dhat <- fitted(fs2)
ss2 <- lm(Y ~ Dhat + X, data = df2)
beta_2sls_X <- coef(ss2)[["Dhat"]]

# Residualized Wald ratio

Xmat <- cbind(1, df2$X)
Ztil <- as.numeric(residualize(df2$Z, Xmat))
Ytil <- as.numeric(residualize(df2$Y, Xmat))
Dtil <- as.numeric(residualize(df2$D, Xmat))

beta_resid_wald <- (sum(Ztil * Ytil) / length(Ztil)) / (sum(Ztil * Dtil) / length(Ztil))

out2 <- data.frame(
estimator = c(
"Naive ITT_Y / naive first stage using group means (no X)",
"2SLS with X in both stages (manual 2SLS)",
"Residualized Wald: Cov(Z~,Y~) / Cov(Z~,D~) after partialling out X"
),
value = c(w2$LATE, beta_2sls_X, beta_resid_wald)
)

out2
```

If you run IV with controls, you must compute ITT and first stage in a
way that matches that regression. The clean way is:

-   reduced form coefficient on Z from $Y \sim Z + X$
-   first stage coefficient on Z from $D \sim Z + X$
-   ratio of those coefficients

That ratio matches 2SLS.

# What exactly is the "share of compliers" in practice?

Under monotonicity and random assignment, the complier share is
identified by the first stage:

$$
P(\text{complier}) = \mathbb{E}[D \mid Z=1] - \mathbb{E}[D \mid Z=0].
$$

So in code, you almost always want:

```{r, eval = FALSE}
complier_share_hat <- mean(D[Z == 1]) - mean(D[Z == 0])
```

If you used `mean(D[Z==1])`, you estimated `always_takers + compliers`,
not compliers.

# Relationship between ITT and LATE: magnitude and sign

A common statement is:

-   "LATE is always larger than ITT"

That statement needs conditions. From:

$$
ITT_Y = P(\text{complier}) \cdot LATE,
$$

we get:

$$
LATE = \frac{ITT_Y}{P(\text{complier})}.
$$

If $0 < P(\text{complier}) \le 1$, then:

-   If $ITT_Y > 0$, then $LATE \ge ITT_Y$
-   If $ITT_Y < 0$, then $LATE \le ITT_Y$ (more negative)
-   In absolute value, $|LATE| \ge |ITT_Y|$ whenever
    $0 < P(\text{complier}) < 1$

Also note: if the instrument reduces treatment uptake, then the first
stage is negative, and the ratio flips sign. That is not a "bug", it is
what the estimator implies.

# Standard errors: why SE(LATE) approx SE(ITT)/complier share is only a rule of thumb

The ratio estimator has variance that depends on the uncertainty in both
numerator and denominator.

Let:

-   $\hat\pi$ be the reduced form coefficient on $Z$
-   $\hat\gamma$ be the first stage coefficient on $Z$
-   $\hat\beta = \hat\pi / \hat\gamma$

A delta method approximation gives:

\$\$ \text{Var}(\hat\beta) \approx \frac{1}{\hat\gamma^2}
\text{Var}(\hat\pi)

-   \frac{\hat\pi^2}{\hat\gamma^4} \text{Var}(\hat\gamma)

-   2 \frac{\hat\pi}{\hat\gamma^3} \text{Cov}(\hat\pi,\hat\gamma). \$\$

The simple rule:

$$
SE(\hat\beta) \approx \frac{SE(\hat\pi)}{|\hat\gamma|}
$$

drops the last two terms. It can be reasonable when the first stage is
very precisely estimated and far from zero. It can be misleading when
instruments are weak.

## Quick Monte Carlo illustration: the rule works when the first stage is strong

Here we vary the complier share and see how the dispersion of
$\hat\beta$ scales.

```{r}
mc_once <- function(N = 3000, p_complier = 0.4, p_always = 0.1, tau_c = 2.0) {
df <- simulate_binary_iv(N = N, p_complier = p_complier, p_always = p_always, tau_c = tau_c)
w <- wald_late(df)
w$LATE
}

mc_grid <- function(p_grid, R = 500, N = 3000) {
out <- lapply(p_grid, function(pC) {
est <- replicate(R, mc_once(N = N, p_complier = pC))
data.frame(
p_complier = pC,
sd_late = sd(est),
mean_late = mean(est)
)
})
do.call(rbind, out)
}

p_grid <- c(0.15, 0.25, 0.35, 0.50, 0.70)
mc_res <- mc_grid(p_grid, R = 300, N = 3000)
mc_res
```

A pattern you should see: as the first stage gets smaller (fewer
compliers), the estimator becomes noisier, and its standard deviation
increases, roughly scaling like $1 / P(\text{complier})$.

# Small sample bias and weak instruments

## What "2SLS bias in small samples" means

Even when the instrument is valid, 2SLS can be biased in finite samples,
especially with weak instruments (small first stage). The bias is
typically toward OLS in many designs.

This has two implications:

1.  Your point estimate may wander substantially in small samples, even
    if the population estimand is clear.
2.  Your normal approximation based inference can be unreliable with
    weak instruments.

## Minimal diagnostics you should always report

-   The first stage estimate and its standard error
-   An F statistic or partial F statistic for instrument relevance
-   Sensitivity to alternative estimators (LIML, Fuller) if weak

Below is a simple simulation where we weaken relevance by shrinking the
complier share, and we compare OLS and IV.

```{r}
run_compare <- function(N = 800, p_complier = 0.10, p_always = 0.10, tau_c = 2.0) {
df <- simulate_binary_iv(N = N, p_complier = p_complier, p_always = p_always, tau_c = tau_c)

ols <- lm(Y ~ D, data = df)
w <- wald_late(df)

data.frame(
N = N,
p_complier = p_complier,
ITT_Y = w$ITT_Y,
first_stage = w$FS,
wald_late = w$LATE,
ols_coef = coef(ols)[["D"]]
)
}

grid <- expand.grid(
p_complier = c(0.05, 0.10, 0.20, 0.40),
rep = 1:200
)

sim_tab <- do.call(rbind, lapply(1:nrow(grid), function(i) {
run_compare(p_complier = grid$p_complier[i])
}))

agg <- aggregate(
cbind(wald_late, ols_coef, first_stage) ~ p_complier,
data = sim_tab,
FUN = function(x) c(mean = mean(x), sd = sd(x))
)

agg
```

Interpretation:

-   As (p\_{\text{complier}}) falls, the first stage shrinks.
-   IV becomes much noisier.
-   OLS remains stable but is not estimating LATE, and can be biased
    when D is endogenous.

For weak instruments, consider LIML or Fuller, and consider weak
instrument robust inference. Bayesian corrections exist in the
literature, but the practical first step is to diagnose instrument
strength and avoid overconfident inference.

# Practical debugging checklist

If your manual ratio does not match IV, check the following in this
order.

## 1. Are you using the correct denominator?

Correct:

```{r}
first_stage <- mean(D[Z == 1]) - mean(D[Z == 0])
late <- itty / first_stage
```

Incorrect in general:

```{r}
late_wrong <- itty / mean(D[Z == 1])
```

## 2. Are you estimating the same specification?

If your 2SLS includes controls, fixed effects, or weights, your manual
objects must match that.

A safe route:

-   Fit reduced form: `lm(Y ~ Z + controls)`
-   Fit first stage: `lm(D ~ Z + controls)`
-   Take ratio of the Z coefficients

```{r}
rf <- lm(Y ~ Z + X1 + X2, data = df)
fs <- lm(D ~ Z + X1 + X2, data = df)
late <- coef(rf)[["Z"]] / coef(fs)[["Z"]]
```

This ratio matches 2SLS under the same linear control specification.

## 3. Is your sample identical across steps?

If you drop missing values differently in different regressions, you are
comparing different samples.

Force a common analysis dataset:

```{r}
vars <- c("Y", "D", "Z", "X1", "X2")
dfA <- df[complete.cases(df[, vars]), vars]
```

## 4. Are you exactly identified?

If you have multiple instruments, 2SLS is not the simple Wald ratio.
Your manual computation must reflect the multi-instrument formula, or
you should use the ratio of reduced form to first stage coefficients
from the same set of instruments in the correct matrix form.

# Summary

-   In the exactly identified binary instrument setup, LATE equals ITT
    on outcomes divided by the first stage.
-   The complier share is the first stage difference in treatment rates,
    not the treated group take up rate.
-   With controls, 2SLS equals a residualized Wald ratio, so you must
    compute adjusted ITT and adjusted first stage to match.
-   The standard error scaling rule is a rough approximation, and it can
    fail when instruments are weak.
-   Small sample behavior can be unstable with weak instruments, so
    diagnose relevance and consider robust alternatives.

# Appendix: Optional packages for IV output

If you want conventional IV output tables, you can use `AER::ivreg`.

```{r}
if (!requireNamespace("AER", quietly = TRUE)) install.packages("AER")
library(AER)

# Constant only:

iv1 <- ivreg(Y ~ D | Z, data = df1)
summary(iv1)

# With controls:

iv2 <- ivreg(Y ~ D + X | Z + X, data = df2)
summary(iv2)
```
