## Johnson-Neyman 2.0

<https://papers.ssrn.com/sol3/papers.cfm?abstract_id=4861381>

<https://papers.ssrn.com/sol3/papers.cfm?abstract_id=4861381>

JN 2.0 is a two-step workflow for probing interactions that replaces linear probing with a Generalized Additive Model, then translates the visual GAM result into a simple numerical test.

Step 1, fit a GAM and visualize either GAM simple slopes or a GAM Johnson-Neyman curve across moderator values.

Step 2, identify the qualitative pattern you claim from that plot, for example, the treatment is negative below $z_0$, then confirm it with a conventional t test or linear regression run on exactly that subset of the data.

The method preserves interpretability, gives you a single estimate and p value, and avoids the strong and often wrong linearity assumption behind classic spotlight and linear J-N.

### Why this matters compared to classic Johnson-Neyman

Classic J-N is linear, it reports where the marginal effect of $x$ on $y$ is statistically different from zero as $z$ varies. It works, but only under the linearity assumption. When the true relationship is curved or changes shape, linear probing can produce qualitatively wrong conclusions, including sign reversals and spurious significance. GAM probing estimates the shape rather than assuming it, and JN 2.0 then converts that flexible result into a familiar test on a relevant subset so that you can report a clean number.

### Assumptions and caveats

1.  Independence for causal interpretation: JN 2.0 is most defensible when $x$ and $z$ are independent. For example, $x$ is randomly assigned and $z$ is measured. If $x$ and $z$ are correlated, nonlinearity can bias linear regressions that you run in Step 2 on subsets, so treat those follow-ups as descriptive and emphasize the GAM plot rather than the subset regression.
2.  If the true interaction is linear, GAM collapses to linear, you do not lose much efficiency.

```{r}
# packages
library(mgcv)          # GAMs
library(dplyr)
library(ggplot2)
library(interactions)  # classic linear J-N

set.seed(123)

# 1) simulate data: x independently randomized, nonlinear interaction in truth
n  <- 1500
x  <- rbinom(n, 1, 0.5)
z  <- rnorm(n, 0, 1)
# true effect of x varies with z in a nonlinear way
tau_z <- -0.6 + 0.9 * exp(-z ^ 2) - 0.4 * pmax(z - 0.5, 0)
y  <- 0.2 * z + tau_z * x + rnorm(n, 0, 1)
dat <- tibble(y, x = factor(x), z)

# 2) classic linear J-N for comparison
m_lin <- lm(y ~ x * z, data = dat)
jn_lin <-
    johnson_neyman(m_lin,
                   pred = x,
                   modx = z,
                   control.fdr = FALSE)
# jn_lin$bounds gives the linear J-N region, and plot(jn_lin) draws the curve

# 3) JN 2.0: probe with GAM, then confirm with a simple test on the relevant subset
# GAM with a smooth for z and a smooth-by interaction of z varying by x
m_gam <-
    gam(y ~ s(z) + s(z, by = x) + x, data = dat, method = "REML")

# compute the marginal effect of x across a grid of z:
grid <- tibble(z = seq(min(z), max(z), length.out = 401))
pred_x1 <-
    predict(m_gam, newdata = mutate(grid, x = factor(1, levels = levels(dat$x))),
            se.fit = TRUE)
pred_x0 <-
    predict(m_gam, newdata = mutate(grid, x = factor(0, levels = levels(dat$x))),
            se.fit = TRUE)
eff   <- pred_x1$fit - pred_x0$fit
se    <- sqrt(pred_x1$se.fit ^ 2 + pred_x0$se.fit ^ 2)
crit  <- qnorm(0.975)
df_g  <-
    grid %>% mutate(effect = eff,
                    lo = eff - crit * se,
                    hi = eff + crit * se)

# visualize the GAM Johnson-Neyman style curve
ggplot(df_g, aes(z, effect)) +
    geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15) +
    geom_line(size = 0.9) +
    geom_hline(yintercept = 0, linetype = 2) +
    labs(x = "Moderator z", y = "Marginal effect of x on y",
         title = "GAM probing of interaction, JN 2.0 Step 1")

# identify a qualitative conclusion to test in Step 2.
# Example: effect is negative for z below the smallest z where the CI includes 0
crosses <- with(df_g, z[which(lo <= 0 & hi >= 0)])
z0 <- if (length(crosses))
    min(crosses)
else
    min(df_g$z) - 1

# Step 2: confirm with a simple and interpretable test on the subset
subset_neg <- dat %>% filter(z < z0)
tt <- t.test(y ~ x, data = subset_neg)  # difference in means
tt$estimate            # point estimate of treatment effect in that region
tt$p.value             # p value for that qualitative claim

# If the visual suggests monotonic growth or decline within a region,
# you can confirm with a linear regression slope on that subset:
lm_slope <- lm(y ~ x + z, data = subset_neg)
summary(lm_slope)$coef
```

Interpretation pattern

1.  Use the GAM plot to describe where the effect is positive, negative, or near zero, and where the confidence band crosses zero.
2.  Translate that pattern into one or two simple numeric statements from Step 2. For example, among observations with $z$ below 0.70, the average effect is −0.42, $p = 0.004$.

### Notes on implementation choices

-   Model specification: `y ~ s(z) + s(z, by = x) + x` lets the shape in $z$ differ by treatment. If x is continuous, use a tensor interaction like `ti(x, z)`.

-   Uncertainty for the $x$ effect: the code above uses the difference of two predictions and the delta method for the standard error, which mirrors a GAM Johnson-Neyman plot.

-   Multiple regions: if the confidence band crosses zero more than once, define each region and confirm each claim separately. Report adjusted p values if needed.

-   Multilevel or non-Gaussian outcomes: GAM supports these through families and random effects as needed. The same two-step logic applies.

### Using `interacting` package

Package created by \@Montealegre

```{r, eval = FALSE}
library(itneracting)
```

# JN 2.0 reusable functions

```{r}
# Dependencies
suppressPackageStartupMessages({
  library(mgcv)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(purrr)
  library(tibble)
})

# Helper: contiguous runs of a logical vector
.contiguous_runs <- function(x) {
  r <- rle(x)
  ends <- cumsum(r$lengths)
  starts <- c(1, head(ends, -1) + 1)
  tibble(val = r$values, start = starts, end = ends)
}

# Core: fit GAM, compute effect of x across z, detect regions, confirm with subset tests
jn2_probe <- function(
  data,
  y,                # unquoted column name
  x,                # unquoted column name, binary, 0 or 1 or two-level factor
  z,                # unquoted column name, continuous
  grid_n = 401,     # resolution of the z grid
  ref_level = NULL, # if x is factor, which level is control, defaults to first
  family = gaussian(),
  k = -1,           # let mgcv select basis size using REML unless you set it
  method = "REML",
  alpha = 0.05,
  confirm_lm = TRUE,# also run lm on the subset as a robustness check
  adjust_p = "BH"   # p adjust method for multiple regions
) {
  y <- rlang::ensym(y); x <- rlang::ensym(x); z <- rlang::ensym(z)
  df <- data %>% drop_na(!!y, !!x, !!z)

  # Coerce x to a two-level factor with levels 0 and 1 semantics
  xvec <- df %>% dplyr::pull(!!x)
  if (is.numeric(xvec)) {
    if (!all(xvec %in% c(0, 1))) stop("x must be binary 0 or 1 if numeric.")
    df[[rlang::as_name(x)]] <- factor(xvec, levels = c(0, 1))
  } else {
    fx <- factor(xvec)
    if (nlevels(fx) != 2) stop("x must have exactly two levels.")
    if (!is.null(ref_level)) fx <- relevel(fx, ref = ref_level)
    df[[rlang::as_name(x)]] <- fx
  }
  levs <- levels(df[[rlang::as_name(x)]])
  ctrl <- levs[1]
  trt  <- levs[2]

  # Build smooth terms, allow k override if provided
  s_main <- if (k > 0) bquote(s(.(z), k = .(k))) else bquote(s(.(z)))
  s_by   <- if (k > 0) bquote(s(.(z), by = .(x), k = .(k))) else bquote(s(.(z), by = .(x)))

  f <- as.formula(bquote(.(y) ~ .(s_main) + .(s_by) + .(x)))

  m_gam <- gam(f, data = df, family = family, method = method)

  # Grid of z
  zseq <- seq(min(df[[rlang::as_name(z)]]), max(df[[rlang::as_name(z)]]), length.out = grid_n)
  grid <- tibble(!!z := zseq)

  # Predictions at treatment and control, delta method for effect and SE
  new_trt  <- grid %>% mutate(!!x := factor(trt, levels = levs))
  new_ctrl <- grid %>% mutate(!!x := factor(ctrl, levels = levs))

  p_trt  <- predict(m_gam, newdata = new_trt, se.fit = TRUE, type = "link")
  p_ctrl <- predict(m_gam, newdata = new_ctrl, se.fit = TRUE, type = "link")

  eff  <- p_trt$fit - p_ctrl$fit
  se   <- sqrt(p_trt$se.fit^2 + p_ctrl$se.fit^2)

  crit <- qnorm(1 - alpha / 2)
  curve_df <- grid %>%
    mutate(effect = eff, se = se, lo = eff - crit * se, hi = eff + crit * se) %>%
    rename(z = !!z)

  # Identify sign regions by whether CI is entirely above or below zero
  above <- curve_df$lo > 0
  below <- curve_df$hi < 0
  runs_above <- .contiguous_runs(above) %>% filter(val)
  runs_below <- .contiguous_runs(below) %>% filter(val)

  idx_to_ranges <- function(runs) {
    if (nrow(runs) == 0) return(tibble(z_lo = numeric(0), z_hi = numeric(0), sign = character(0)))
    tibble(
      z_lo = curve_df$z[runs$start],
      z_hi = curve_df$z[runs$end],
      sign = NA_character_
    )
  }

  regions <- bind_rows(
    idx_to_ranges(runs_below) %>% mutate(sign = "negative"),
    idx_to_ranges(runs_above) %>% mutate(sign = "positive")
  ) %>% arrange(z_lo, z_hi)

  # Confirmatory tests in each region
  confirm_tbl <- map_dfr(seq_len(nrow(regions)), function(i) {
    r <- regions[i, ]
    sub <- df %>% filter(dplyr::between(!!z, r$z_lo, r$z_hi))
    n_all <- nrow(sub)
    n0 <- sum(sub[[rlang::as_name(x)]] == ctrl)
    n1 <- sum(sub[[rlang::as_name(x)]] == trt)

    if (n0 < 5 || n1 < 5) {
      return(tibble(
        z_lo = r$z_lo, z_hi = r$z_hi, sign = r$sign,
        n = n_all, n_ctrl = n0, n_trt = n1,
        est = NA_real_, p_ttest = NA_real_, est_lm = NA_real_, p_lm = NA_real_
      ))
    }

    # t test on difference in means
    tt <- t.test(sub[[rlang::as_name(y)]] ~ sub[[rlang::as_name(x)]])
    est <- as.numeric(diff(tt$estimate))  # trt minus ctrl

    # simple regression on subset
    if (confirm_lm) {
      lm_fit <- lm(as.formula(paste(rlang::as_name(y), "~", rlang::as_name(x), "+", rlang::as_name(z))), data = sub)
      co <- summary(lm_fit)$coef
      est_lm <- unname(co[rownames(co) == paste0(rlang::as_name(x), trt), "Estimate"])
      p_lm   <- unname(co[rownames(co) == paste0(rlang::as_name(x), trt), "Pr(>|t|)"])
    } else {
      est_lm <- NA_real_; p_lm <- NA_real_
    }

    tibble(
      z_lo = r$z_lo, z_hi = r$z_hi, sign = r$sign,
      n = n_all, n_ctrl = n0, n_trt = n1,
      est = est, p_ttest = tt$p.value,
      est_lm = est_lm, p_lm = p_lm
    )
  })

  # P value adjustment across detected regions
  if (nrow(confirm_tbl) > 0) {
    if (!all(is.na(confirm_tbl$p_ttest))) {
      confirm_tbl$p_ttest_adj <- p.adjust(confirm_tbl$p_ttest, method = adjust_p)
    } else confirm_tbl$p_ttest_adj <- NA_real_
    if (!all(is.na(confirm_tbl$p_lm))) {
      confirm_tbl$p_lm_adj <- p.adjust(confirm_tbl$p_lm, method = adjust_p)
    } else confirm_tbl$p_lm_adj <- NA_real_
  }

  # Plot
  p <- ggplot(curve_df, aes(z, effect)) +
    geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15) +
    geom_line(size = 0.9) +
    geom_hline(yintercept = 0, linetype = 2) +
    labs(x = rlang::as_name(z), y = paste("Effect of", rlang::as_name(x), "on", rlang::as_name(y)),
         title = "Johnson Neyman 2.0, GAM probing of interaction")

  list(
    model = m_gam,
    curve = curve_df,
    regions = regions,
    confirm = confirm_tbl,
    plot = p,
    meta = list(
      x_levels = levs, control = ctrl, treatment = trt,
      alpha = alpha, p_adjust = adjust_p,
      formula = f
    )
  )
}

# Convenience: classic linear Johnson Neyman via interactions, if installed
jn_linear_wrapper <- function(lm_fit, pred, modx, control_fdr = FALSE) {
  if (!requireNamespace("interactions", quietly = TRUE)) {
    stop("Package 'interactions' is required for classic linear J N. Install it first.")
  }
  interactions::johnson_neyman(lm_fit, pred = {{pred}}, modx = {{modx}}, control.fdr = control_fdr)
}
```

# End to end example

```{r}
set.seed(1)
n  <- 1500
x  <- rbinom(n, 1, 0.5)
z  <- rnorm(n)
tau_z <- -0.5 + 0.8 * exp(-z^2) - 0.35 * pmax(z - 0.6, 0)
y  <- 0.2*z + tau_z * x + rnorm(n, 0, 1)
dat <- tibble(y, x, z)

out <- jn2_probe(dat, y, x, z, grid_n = 501)

# Plot the GAM Johnson Neyman style curve
out$plot

# Regions and confirmatory tests
out$regions
out$confirm %>% mutate(across(starts_with("p_"), ~signif(.x, 3)))
```

# Reporting template

1.  Describe the shape using the GAM curve: for example, the marginal effect of x on y is negative for low z, crosses zero near z equals 0.65, and becomes positive for high z.
2.  Translate to numbers using the confirmatory tests: for example, among observations with z between 0.66 and 1.85, the average effect is 0.41, p equals 0.003, BH adjusted p equals 0.006. Also report the sign and range for negative regions if present.
3.  If x and z are correlated or if the design is observational, add a caution that subset regressions are descriptive. Emphasize the shape from the GAM plot.

# Options and extensions

-   Family and links: jn2_probe accepts any mgcv family, for example binomial for a binary outcome. The effect calculation remains a difference on the link scale. If you want the effect on the response scale, I can add simulation based transformation.

-   Basis size: pass k to control smooth complexity, otherwise REML selects it.

-   Continuous x: use a tensor product ti(x, z) and compute the derivative dE[y]/dx across z. I can add this variant if you want continuous doses or prices.

-   Multiple crossings: the function already handles multiple disjoint regions and applies Benjamini Hochberg correction.

-   Classic linear J N: jn_linear_wrapper runs interactions::johnson_neyman for a fitted lm to include in an appendix.
