## Simulation Workshop: What Fails Where (and Why)

We'll simulate three regimes:

1.  **S1: No true effect + heavy *p*-hacking** (optional stopping + choose-best-of-$m$).
2.  **S2: Small true effect + publication bias** (selection on significance).
3.  **S3: Heterogeneous effects + selection + hacking** (worst case).

For each, we apply multiple tools and compare **bias**, **RMSE**, and (where meaningful) **coverage** across replicates.

```{r}
suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(purrr); library(ggplot2); 
})

# ---------- Utility: meta-analytic estimators (minimal) ----------
est_re_ivw <- function(yi, sei) {
  # DerSimonian-Laird tau^2 (simple) then IVW; robust enough for demo
  vi <- sei^2
  w_fixed <- 1/vi
  theta_fixed <- sum(w_fixed*yi)/sum(w_fixed)
  Q <- sum(w_fixed*(yi - theta_fixed)^2)
  c_val <- sum(w_fixed) - sum(w_fixed^2)/sum(w_fixed)
  tau2 <- max(0, (Q - (length(yi)-1))/c_val)
  w_re <- 1/(vi + tau2)
  theta_re <- sum(w_re*yi)/sum(w_re)
  se_re <- sqrt(1/sum(w_re))
  list(theta = theta_re, se = se_re, tau2 = tau2)
}

est_petse <- function(yi, sei, type = c("PET","PEESE")) {
  type <- match.arg(type)
  if (type == "PET") {
    fit <- lm(yi ~ sei, weights = 1/sei^2)
  } else {
    fit <- lm(yi ~ I(sei^2), weights = 1/sei^2)
  }
  co <- coef(summary(fit))
  list(theta = unname(co[1,"Estimate"]), se = unname(co[1,"Std. Error"]))
}

# Weight-function toy: two bins (p<=.05 vs p>.05) fixed-effect theta
est_weightfun_toy <- function(yi, sei, p_two_sided) {
  negloglik <- function(par) {
    theta <- par[1]; logw <- par[2]; w1 <- exp(logw)
    z <- yi/sei
    prob_sig <- pnorm(-1.96 - theta/sei) + (1 - pnorm(1.96 - theta/sei))
    wi <- ifelse(p_two_sided <= .05, w1, 1)
    num <- dnorm((yi - theta)/sei)/sei * wi
    denom <- w1 * prob_sig + (1 - prob_sig)
    -sum(log(num) - log(denom))
  }
  fit <- optim(c(mean(yi), 0), negloglik, method = "BFGS", control = list(reltol = 1e-9))
  list(theta = fit$par[1], weight_sig = exp(fit$par[2]), ok = fit$convergence == 0)
}

# p-curve-style conditional MLE (diagnostic estimator; one-sided transform)
est_pcurve_mle <- function(p_sig, alpha = 0.05, map_se = NULL) {
  if (length(p_sig) < 3) return(list(mu = NA, theta = NA))
  z <- qnorm(1 - p_sig)
  pi_mu <- function(mu, alpha) 1 - pnorm(qnorm(1 - alpha) - mu)
  ll <- function(mu) sum(mu*z - 0.5*mu^2 - log(pi_mu(mu, alpha)))
  opt <- try(optimize(function(m) -ll(m), c(-6,6)), silent = TRUE)
  if (inherits(opt, "try-error")) return(list(mu = NA, theta = NA))
  muhat <- opt$minimum
  theta <- if (is.null(map_se)) NA else muhat * mean(map_se)  # rough map
  list(mu = muhat, theta = theta)
}

# z-curve & RoBMA wrappers if installed
safe_zcurve <- function(p) {
  if (!requireNamespace("zcurve", quietly = TRUE)) return(list(EDR = NA, ERR = NA))
  fit <- zcurve::zcurve(p)
  s <- summary(fit)
  list(EDR = as.numeric(s$EDR[1]), ERR = as.numeric(s$ERR[1]))
}
safe_RoBMA <- function(yi, sei) {
  if (!requireNamespace("RoBMA", quietly = TRUE)) return(list(mu = NA))
  fit <- RoBMA::RoBMA(yi, sei, priors_bias = c("none","pet_peese","selection"))
  s <- summary(fit)
  list(mu = as.numeric(s$mu[1]))
}

# ---------- DGPs ----------
one_study_optional_stopping <- function(theta, se, alpha = 0.05, max_n_looks = 20) {
  # Sequential looks; increments mimic accumulating data (toy)
  mu <- theta/se
  for (j in 1:max_n_looks) {
    z <- rnorm(1, mu, 1)
    p <- 2*(1 - pnorm(abs(z)))
    if (p <= alpha) return(c(z=z, p=p))
  }
  c(z=z, p=p)  # return last
}

one_study_best_of_m <- function(theta, se, m = 5) {
  mu <- theta/se
  z <- rnorm(m, mu, 1)
  p <- 2*(1 - pnorm(abs(z)))
  j <- which.min(p)
  c(z=z[j], p=p[j])
}


simulate_meta <-
    function(K,
             theta_bar,
             tau,
             sel = c(0.9, 0.2),
             hack = c("none", "optional", "best_of_m"),
             m = 5,
             alpha = 0.05) {
        n <- sample(40:300, K, TRUE)
        se <- sqrt(2 / n)
        theta_i <- rnorm(K, theta_bar, tau)
        out <- lapply(1:K, function(i) {
            if (hack[1] == "optional") {
                zp <- one_study_optional_stopping(theta_i[i], se[i], alpha = alpha)
            } else if (hack[1] == "best_of_m") {
                zp <- one_study_best_of_m(theta_i[i], se[i], m = m)
            } else {
                mu <- theta_i[i] / se[i]
                z <-
                    rnorm(1, mu, 1)
                p <- 2 * (1 - pnorm(abs(z)))
                zp <- c(z = z, p = p)
            }
            yi <- zp["z"] * se[i]
            sel_prob <- ifelse(zp["p"] <= alpha, sel[1], sel[2])
            published <- rbinom(1, 1, sel_prob) == 1
            c(
                yi = yi,
                se = se[i],
                z = zp["z"],
                p = zp["p"],
                published = published
            )
        })
        as.data.frame(do.call(rbind, out))
    }
```

### 3.1 Define scenarios

-   **S1 (Null + heavy hacking)**: $\theta=0$, use **optional stopping** and **best-of-**$m$ (we'll run both sub-scenarios).
-   **S2 (Small effect + selection)**: $\theta=0.20$, $\tau=0.10$, selection $(0.9,0.2)$, no hacking.
-   **S3 (Heterogeneity + selection + hacking)**: $\theta=0.20$, $\tau=0.30$, selection $(0.95,0.05)$, **best-of-**$m$ with $m=10$.

We'll run **R=200 replicates** for speed.

```{r}
set.seed(2026)
R <- 200

run_once <- function(scen) {
  if (scen=="S1_optional") {
    dat <- simulate_meta(200, 0, 0, sel=c(0.9,0.9), hack="optional", alpha=0.05) # no selection, only hacking
  } else if (scen=="S1_bestm") {
    dat <- simulate_meta(200, 0, 0, sel=c(0.9,0.9), hack="best_of_m", m=10, alpha=0.05)
  } else if (scen=="S2") {
    dat <- simulate_meta(200, 0.20, 0.10, sel=c(0.9,0.2), hack="none", alpha=0.05)
  } else if (scen=="S3") {
    dat <- simulate_meta(200, 0.20, 0.30, sel=c(0.95,0.05), hack="best_of_m", m=10, alpha=0.05)
  } else stop("Unknown scenario")
  
  dat_pub <- subset(dat, published==1)
  if (nrow(dat_pub) < 8) return(NULL)
  # Estimators
  re  <- est_re_ivw(dat_pub$yi, dat_pub$se)
  pet <- est_petse(dat_pub$yi, dat_pub$se, "PET")
  pse <- est_petse(dat_pub$yi, dat_pub$se, "PEESE")
  wf  <- est_weightfun_toy(dat_pub$yi, dat_pub$se, p_two_sided = dat_pub$p)
  pc  <- est_pcurve_mle(dat_pub$p[dat_pub$p<=0.05], alpha=0.05, map_se = dat_pub$se)
  
  # Diagnostics
  alpha <- 0.05
  p_sig <- dat_pub$p[dat_pub$p<=alpha]
  right_skew_p <- if (length(p_sig)>=5) binom.test(sum(p_sig<alpha/2), length(p_sig), 0.5, "greater")$p.value else NA_real_
  z <- dat_pub$z; c <- 1.96; h <- 0.10
  Np <- sum(z>c & z<=c+h); Nm <- sum(z>c-h & z<=c)
  caliper_p <- if (Np+Nm>0) binom.test(Np, Np+Nm, 0.5, "greater")$p.value else NA_real_
  
  # Optional: z-curve & RoBMA if available
  zc  <- safe_zcurve(dat_pub$p[dat_pub$p<=0.10]) # many users pass all p<=.10
  rb  <- safe_RoBMA(dat_pub$yi, dat_pub$se)
  
  tibble(
    n_pub = nrow(dat_pub),
    RE = re$theta, PET = pet$theta, PEESE = pse$theta,
    WF = wf$theta, WF_w = wf$weight_sig, PC_mu = pc$mu, PC_theta = pc$theta,
    p_right = right_skew_p, p_caliper = caliper_p, EDR = zc$EDR, ERR = zc$ERR, RoBMA = rb$mu
  )
}

scenarios <- c("S1_optional","S1_bestm","S2","S3")
res <- lapply(scenarios, function(s) {
  tmp <- purrr::map_dfr(1:R, ~run_once(s))
  cbind(scenario = s, tmp)
}) %>% bind_rows()

summary_res <- res %>%
  group_by(scenario) %>%
  summarise(across(c(RE, PET, PEESE, WF, PC_theta, RoBMA), 
                   list(mean=~mean(., na.rm=TRUE), sd=~sd(., na.rm=TRUE)), .names="{.col}_{.fn}"),
            n_pub_mean = mean(n_pub, na.rm=TRUE),
            p_right_sig = mean(p_right < 0.05, na.rm=TRUE),
            p_caliper_sig = mean(p_caliper < 0.05, na.rm=TRUE),
            EDR_mean = mean(EDR, na.rm=TRUE), ERR_mean = mean(ERR, na.rm=TRUE))

kable(summary_res, "html", digits = 3) %>% kable_styling(full_width = TRUE)
```

#### Reading the table

-   **S1_optional / S1_bestm (true** $\theta=0$):

    -   **RE** should show **positive bias** (fake discoveries).
    -   **PET** often shrinks more toward 0; **PEESE** can overshoot or undershoot depending on pattern.
    -   **WF** (toy selection model) should attenuate toward 0 if $p$-bunching is present.
    -   **p-curve right-skew** might **wrongly detect evidential value** under some hacking regimes (optional stopping can create skew), highlighting a key **blogosphere caution**.

-   **S2 (small effect)**:

    -   **RE** overestimates if selection strong; **PET/PEESE** vary.
    -   **WF** should pull toward the truth if selection matches the model form.
    -   **p-curve right-skew** should be significant (real effect present).

-   **S3 (heterogeneity + hacking + selection)**:

    -   This is the **hard case**: disagreements between PET/PEESE vs WF vs PC are common.
    -   If available, **z-curve EDR** often flags low expected discovery rate relative to observed significance (selection signal).
    -   **RoBMA** (if installed) typically shrinks compared to RE.

### 3.2 Visuals: p-histograms and caliper checks per scenario

```{r}
viz_one <- function(dat_pub, title="") {
  g1 <- ggplot(dat_pub, aes(p)) + 
    geom_histogram(binwidth = 0.01, boundary=0, fill="#3182bd", color="white") +
    geom_vline(xintercept = 0.05, linetype=2, color="red") +
    labs(title=paste("p-histogram:", title), x="p", y="Count")
  z <- dat_pub$z; c <- 1.96; h <- 0.1
  Np <- sum(z>c & z<=c+h); Nm <- sum(z>c-h & z<=c)
  p_cal <- if (Np+Nm>0) binom.test(Np, Np+Nm, 0.5, "greater")$p.value else NA_real_
  list(plot=g1, Np=Np, Nm=Nm, caliper_p=p_cal)
}

set.seed(3)
ex_S1 <- simulate_meta(400, 0, 0, sel=c(0.9,0.9), hack="best_of_m", m=10) %>% subset(published==1)
ex_S2 <- simulate_meta(400, 0.2, 0.1, sel=c(0.9,0.2), hack="none") %>% subset(published==1)
ex_S3 <- simulate_meta(400, 0.2, 0.3, sel=c(0.95,0.05), hack="best_of_m", m=10) %>% subset(published==1)

v1 <- viz_one(ex_S1, "S1 best-of-m (theta=0)")
v2 <- viz_one(ex_S2, "S2 selection (theta=0.2)")
v3 <- viz_one(ex_S3, "S3 heterogeneity + hacking")

v1$plot; v2$plot; v3$plot
rbind(
  S1=c(N_above=v1$Np, N_below=v1$Nm, caliper_p=v1$caliper_p),
  S2=c(N_above=v2$Np, N_below=v2$Nm, caliper_p=v2$caliper_p),
  S3=c(N_above=v3$Np, N_below=v3$Nm, caliper_p=v3$caliper_p)
)
```

------------------------------------------------------------------------

## 4. Revisiting Specific "Beef" With Simulations

### 4.1 p-curve as estimator vs. diagnostic

-   **Claim** (critiques): With heterogeneity and dependence, p-curve MLE is biased/under-covers.
-   **Demo**: Compare p-curve MLE ($\hat\mu$ mapped to $\hat\theta$) vs. **WF** and **RE/PET/PEESE** in **S3**. Expect divergence and instability of p-curve's $\hat\theta$.

```{r}
res_S3 <- subset(res, scenario=="S3")
summary_S3 <- res_S3 %>% 
  summarise(across(c(RE, PET, PEESE, WF, PC_theta), list(m=~mean(., na.rm=TRUE), sd=~sd(., na.rm=TRUE))))
kable(summary_S3, "html", digits = 3)
```

**Interpretation**: If **PC_theta** bounces around or drifts compared to WF/PET/PEESE, that illustrates the critique. Use p-curve primarily to test **right-skew** (evidential value) instead.

### 4.2 Optional stopping (S1_optional) and right-skew

-   **Claim** (blogs): Optional stopping can generate right-skew among "significant" $p$---p-curve can misread **manipulation** as **evidential value**.
-   **Demo**: Inspect **p_right** in **S1_optional**.

```{r}
res %>%
  filter(scenario=="S1_optional") %>%
  summarise(prop_right_skew_sig = mean(p_right < 0.05, na.rm=TRUE))
```

If the proportion is high, that's a cautionary tale.

### 4.3 Calipers (Brodeur et al. "Star Wars") and bunching

-   **Claim**: Bunching near 1.96 suggests thresholds.
-   **Counter**: Heterogeneity can also cause local density differences; need triangulation.
-   **Demo**: **S2** often shows both small right-skew and caliper significance.

```{r}
res %>%
  group_by(scenario) %>%
  summarise(caliper_sig = mean(p_caliper < 0.05, na.rm=TRUE),
            right_skew_sig = mean(p_right < 0.05, na.rm=TRUE))
```

------------------------------------------------------------------------

## 5. Toolbox: Code Recipes You Can Drop Into Your Meta-Analyses

### 5.1 Multi-method report (with graceful fallbacks)

```{r}
multi_method_report <- function(yi, sei, p, title="") {
  out <- list(title=title)
  # Naive RE
  re <- est_re_ivw(yi, sei)
  out$RE <- re
  
  # PET & PEESE
  out$PET <- est_petse(yi, sei, "PET")
  out$PEESE <- est_petse(yi, sei, "PEESE")
  
  # Weight-function toy
  out$WF <- est_weightfun_toy(yi, sei, p)
  
  # p-curve diagnostics
  alpha <- 0.05
  p_sig <- p[p<=alpha]
  out$pcurve <- list(
    right_skew_p = if (length(p_sig)>=5) binom.test(sum(p_sig<alpha/2), length(p_sig), 0.5, "greater")$p.value else NA_real_,
    mle = est_pcurve_mle(p_sig, alpha=0.05, map_se = sei)
  )
  
  # z-curve if available
  out$zcurve <- safe_zcurve(p)
  # RoBMA if available
  out$RoBMA <- safe_RoBMA(yi, sei)
  
  out
}

# Demo on S3 example
rep_S3 <- multi_method_report(ex_S3$yi, ex_S3$se, ex_S3$p, "S3 demo")
rep_S3
```

### 5.2 Specification Curve (multiverse) for your own single-study analysis

```{r}
set.seed(101)
n <- 600
X <- data.frame(x1 = rnorm(n), x2 = rnorm(n), x3 = rnorm(n))
D <- rbinom(n, 1, 0.5)
Y <- 0.2*D + 0.3*X$x1 - 0.1*X$x2 + rnorm(n, sd=1)
specs <- list(
  ~ D,
  ~ D + x1,
  ~ D + x2,
  ~ D + x3,
  ~ D + x1 + x2,
  ~ D + x1 + x3,
  ~ D + x2 + x3,
  ~ D + x1 + x2 + x3
)
est <- purrr::map_dfr(specs, function(fml) {
  f <- as.formula(paste("Y ~", deparse(fml)[2]))
  m <- lm(f, data=cbind(Y, D, X))
  co <- summary(m)$coef["D",]
  tibble(spec = deparse(f)[2], beta = co["Estimate"], se = co["Std. Error"], p = 2*(1 - pnorm(abs(co["Estimate"]/co["Std. Error"]))))
})
ggplot(est, aes(x = reorder(spec, beta), y = beta)) +
  geom_hline(yintercept = 0, linetype=2, color="gray50") +
  geom_point(aes(color = p <= 0.05)) +
  geom_errorbar(aes(ymin = beta - 1.96*se, ymax = beta + 1.96*se), width = 0) +
  scale_color_manual(values = c("FALSE"="#636363","TRUE"="#e34a33")) +
  coord_flip() + labs(x="Specification", y="Effect of D", title="Specification Curve (Multiverse)")
```

------------------------------------------------------------------------

## What to Assign (Exercises)

1.  **Replicate S1--S3** and add **FDR**-control post-hoc on the published set. Does FDR flag the hacked regimes?
2.  **Swap** the weight-function from a single threshold (.05) to a **3-step** threshold (.10, .05, .01); compare estimated weights and adjusted $\hat\theta$.
3.  **Dependence stress**: Generate 5 estimates per study with within-study $\rho=0.5$; evaluate how PET/PEESE and p-curve diagnostics change (hint: use cluster-robust SE or multilevel meta-analysis).
4.  **Finance-flavored**: Simulate 100 candidate "factors," keep the most significant per dataset, run z-curve and RoBMA if available, and compute **out-of-sample** predictive $R^2$.

------------------------------------------------------------------------

## 2. Mathematical Core (Expanded)

### 2.1 Selection on significance vs. p-hacking

Let test statistic $Z \sim \mathcal{N}(\mu,1)$; two-sided $p=2[1-\Phi(|Z|)]$. A **selection function** $s(z)$ yields the observed $z$-density $$
f_{\mathrm{obs}}(z) \propto s(z) \int \varphi(z-\mu), dG(\mu),
$$ where $G$ is the mixing distribution of noncentralities $\mu$ and $\varphi$ is the standard normal density.

-   **Publication bias**: $s(z)$ jumps at $|z|=z_{1-\alpha/2}$.
-   **p-hacking**: $s(z)$ has local spikes/bunching near thresholds (e.g., **bunching** just above $1.96$).

**Identifiability**: Without restrictions, $s$ and $G$ are not point-identified. Restrictions include: monotonicity/stepwise $s$, parametric mixtures for $G$, or conditioning on significance (curve methods).

### 2.2 PET-PEESE (precision meta-regression)

Let $y_i$ and $s_i$ be effect and SE. Under $y_i=\theta + \beta s_i^\delta + \varepsilon_i$ with weights $1/s_i^2$,

-   PET: $\delta=1$; $\hat\theta=\hat\beta_0$ as $s_i \to 0$.
-   PEESE: $\delta=2$; often used if PET rejects $\theta=0$.

Assumes bias is **smooth in precision**; conflates genuine small-study heterogeneity with bias if the relation is not linear in $s_i^\delta$.

### 2.3 Weight-function selection models

For $p$-bins ${\mathcal{I}*b}$, let $w_b>0$ be selection weights. Fixed effects: $$
L(\theta,\mathbf{w}) \propto \prod*{i=1}^k \frac{\phi\big((y_i-\theta)/s_i\big) s_i^{-1} , w_{b(i)}}
{\sum_{b} w_b , \Pr_\theta\big(p(y)\in \mathcal{I}_b ,\big|; s_i\big)}.
$$ Random effects integrate $\theta_i \sim \mathcal{N}(\theta,\tau^2)$ in numerator and denominator. $\mathbf{w}$ is identified **up to scale** (fix one bin to 1).

### 2.4 p-curve and p-uniform\*

Given $P=1-\Phi(Z)$ (one-sided) with power $\pi(\mu,\alpha)=1-\Phi(z_{1-\alpha}-\mu)$, the **conditional density** given $P \le \alpha$ is $$
f_\alpha(p \mid \mu) = \frac{\exp{\mu z - \tfrac{1}{2}\mu^2}}{\pi(\mu,\alpha)}, \quad z=\Phi^{-1}(1-p).
$$

-   **p-curve**: diagnostic tests of right-skew; sometimes MLE for $\mu$.
-   **p-uniform**: MLE for effect under **homogeneous** $\mu$;
-   **p-uniform**\*: extends to random effects; better finite-sample properties under heterogeneity.

**Key caveat**: All **curve** methods condition on $p\le\alpha$; if selection is not (mostly) on this threshold or if $p$'s are **not independent** (e.g., many specs from one dataset), properties degrade.

### 2.5 z-curve 2.0

Model $Z \mid |Z|\ge z_{1-\alpha/2}$ as a **finite mixture** $\sum_k \pi_k ,\mathcal{N}(\mu_k,1)$ with truncation. Estimate via EM with truncation corrections. Compute $$
\mathrm{EDR} = \int \Pr(|Z| \ge z_{1-\alpha/2} \mid \mu), d\hat G(\mu), \qquad
\mathrm{ERR} = \mathbb{E}\big[\Pr(|Z^\star|\ge z_{1-\alpha/2} \mid \mu)\mid |Z|\ge z_{1-\alpha/2}\big].
$$ EDR acts as a **power-weighted prevalence** of significance; ERR is a model-based **replication** probability.

### 2.6 Copas--Shi sensitivity

Assume latent selection driven by a correlation between the probability of publication and the **study precision** (or its proxy). The Copas model posits a latent selection variable influencing whether a study is observed; sensitivity is performed over a grid of selection parameters to project bias-adjusted effects.

------------------------------------------------------------------------

## 3. The "Beef": Improvements, Disagreements, and What They Mean

-   **Curve methods vs. selection models**

    -   *Curve advocates* emphasize robustness when selection is heavily on the $p \le \alpha$ threshold and when nonsignificant studies are missing entirely.
    -   *Selection-model advocates* argue for **explicit modeling** of selection and the use of **all** studies (sig + nsig), producing **directly adjusted** effects under transparent assumptions about $w(p)$.

-   **p-uniform vs. p-uniform**\*

    -   Original p-uniform can be **biased** under heterogeneity and low power; p-uniform\* introduces random effects and better coverage, but remains sensitive to **mis-specification of test families** and **dependence** among $p$'s.

-   **z-curve vs. others**

    -   z-curve's **mixture flexibility** handles heterogeneity well; **EDR/ERR** are intuitive to users.
    -   Critics point to **component number selection**, **truncation identification**, and **sensitivity** to the $p$-value curation (e.g., inclusion criteria, dependence).

-   **PET-PEESE vs. weight-functions**

    -   PET-PEESE is **transparent and simple** but may **overcorrect** when small-study effects reflect **real** heterogeneity;
    -   Weight-functions can be **unstable** if the stepwise form is misspecified, but they directly target the mechanism producing asymmetry.

-   **Excess significance**

    -   Intuitive, but depends on a (possibly biased) effect used to compute **power**; heterogeneity and **researcher degrees of freedom** can inflate significance beyond what the test uses as "expected".

-   **On the statistical properties of p-curve (Morey & Davis-Stober critique)**

    -   Recent critiques (e.g., "On the poor statistical properties of the p-curve meta-analytic procedure") argue that p-curve's **effect estimator** can exhibit **poor calibration**, **bias** and **under-coverage**, especially under heterogeneity, misspecified test families, and dependent $p$'s.
    -   Conceptual points include: conditioning on $p\le\alpha$ discards information, the estimator may be **non-invariant** to certain reparameterizations, and finite-sample behavior can be **fragile**.
    -   **Implication**: Treat p-curve primarily as a **diagnostic** for evidential value; for estimation, prefer **p-uniform**\*, **selection models**, or **model-averaged Bayesian** methods (RoBMA).

------------------------------------------------------------------------

## 4. Simulating p-Hacking & Selection, Then Diagnosing With Many Tools

We'll simulate a meta-analytic corpus with:

1.  **True heterogeneity**, 2) **publication bias** favoring $p\le .05$, and 3) **p-hacking** via a simple "add a covariate" retry if initial $p> .05$.

```{r}
set.seed(2025)

# --- Simulation design ---
K         <- 150                      # number of studies
theta_bar <- 0.20                     # grand mean
tau       <- 0.15                     # heterogeneity SD
n_i       <- sample(30:200, K, TRUE)  # per-study sample sizes
se_i      <- sqrt(2/n_i)              # SE for standardized mean difference (toy)
theta_i   <- rnorm(K, theta_bar, tau) # random effects

# Generate study z and two-sided p under a simple z-test
z_raw     <- rnorm(K, mean = theta_i / se_i, sd = 1)
p_raw     <- 2 * (1 - pnorm(abs(z_raw)))

# Simple p-hacking: if p>.05, "try a covariate" with correlated boost in z
# (This is stylized; real p-hacking is more complex.)
hack      <- runif(K) < 0.5           # only some attempt hacking
z_hack    <- z_raw + rnorm(K, 0.35, 0.20) # small optimistic bump
p_hacked  <- 2 * (1 - pnorm(abs(ifelse(p_raw > .05 & hack, z_hack, z_raw))))

# Publication selection: favor p<=.05 with higher selection probability
sel_prob  <- ifelse(p_hacked <= .05, 0.9, 0.2)
published <- rbinom(K, 1, sel_prob) == 1

dat <- data.frame(
  study = seq_len(K),
  n = n_i,
  se = se_i,
  yi = (ifelse(p_raw > .05 & hack, z_hack, z_raw)) * se_i,  # back to effect scale
  zi = ifelse(p_raw > .05 & hack, z_hack, z_raw),
  p = p_hacked,
  published = published
)

dat_pub <- subset(dat, published)

summary(list(
  K_total = nrow(dat),
  K_published = nrow(dat_pub),
  prop_significant = mean(dat_pub$p <= .05),
  mean_effect_published = mean(dat_pub$yi),
  sd_effect_published = sd(dat_pub$yi)
))
```

### 4.1 Visual diagnostics: p-histogram & caliper test

```{r}
library(ggplot2)
p_sig <- dat_pub$p[dat_pub$p <= .10]  # focus near .10 for visualization
plt_phist <- ggplot(data.frame(p = dat_pub$p), aes(p)) +
  geom_histogram(binwidth = 0.01, boundary = 0, color = "white", fill = "#3182bd") +
  geom_vline(xintercept = 0.05, linetype = 2, color = "red") +
  labs(title = "Histogram of Published p-values", subtitle = "Dashed line at 0.05", x = "p", y = "Count")

# Caliper around z=1.96 (two-sided .05)
zvals <- dat_pub$zi
c <- 1.96; h <- 0.1
N_plus  <- sum(zvals > c & zvals <= c+h)
N_minus <- sum(zvals > c-h & zvals <= c)
p_caliper <- binom.test(N_plus, N_plus + N_minus, 0.5, alternative = "greater")$p.value

plt_phist
list(N_above = N_plus, N_below = N_minus, caliper_p = p_caliper)
```

### 4.2 Classical random-effects meta-analysis, funnel plot, and Egger

```{r}
library(metafor)

# Random-effects (REML)
rma_re <- rma.uni(yi = dat_pub$yi, sei = dat_pub$se, method = "REML")
egger   <- regtest(rma_re, model = "lm")  # Egger's test of funnel asymmetry

# Funnel plot
funnel_data <- data.frame(yi = dat_pub$yi, sei = dat_pub$se, wi = 1/dat_pub$se^2)
plot(rma_re, main = "Funnel Plot (metafor)")

list(RE_theta = c(est = rma_re$b[1], se = rma_re$se[1]),
     tau2 = rma_re$tau2,
     Egger_intercept = egger$beta[1], Egger_p = egger$pval)
```

### 4.3 PET-PEESE

```{r}
# PET
pet <- rma.uni(yi = dat_pub$yi, vi = dat_pub$se^2, mods = ~ I(dat_pub$se), method = "REML")
# PEESE
peese <- rma.uni(yi = dat_pub$yi, vi = dat_pub$se^2, mods = ~ I(dat_pub$se^2), method = "REML")

list(
  PET_intercept = coef(pet)[1], PET_p = coef(summary(pet))[1,4],
  PEESE_intercept = coef(peese)[1], PEESE_p = coef(summary(peese))[1,4]
)
```

### 4.4 Trim-and-fill

```{r}
tf <- trimfill(rma_re)
list(TF_theta = c(est = tf$b[1], se = tf$se[1]), n_imputed = tf$k0)
```

### 4.5 Copas--Shi sensitivity (metasens)

```{r}
if (requireNamespace("metasens", quietly = TRUE)) {
  library(metasens)
  cop <- copas(TE = dat_pub$yi, seTE = dat_pub$se)
  list(Copas_theta_adj = cop$adj, Copas_note = "Interpret across sensitivity grid")
} else {
  "metasens not installed; skip Copas."
}
```

### 4.6 Weight-function selection model (weightr)

```{r}
if (requireNamespace("weightr", quietly = TRUE)) {
  library(weightr)
  # We need two-sided p-values; use dat_pub$p
  wf <- weightfunct(effect = dat_pub$yi, v = dat_pub$se^2,
                    steps = c(0.025, 0.05, 0.10), table = TRUE)
  list(weightr_theta = wf$adj.est, weights = wf$weights.table)
} else {
  "weightr not installed; skip weight-function."
}
```

### 4.7 p-curve diagnostics (custom): right-skew & conditional MLE

```{r}
p_sig <- dat_pub$p[dat_pub$p <= 0.05]
alpha <- 0.05
n_sig <- length(p_sig)

# Right-skew binomial test: P(p < alpha/2 | p <= alpha) = 0.5 under null of no evidential value
below <- sum(p_sig < alpha/2)
p_right_skew <- binom.test(below, n_sig, 0.5, alternative = "greater")$p.value

# Conditional MLE for one-sided z (diagnostic; do not over-interpret as grand mean)
z_one <- qnorm(1 - p_sig) # treat as one-sided to match the transform
pi_mu <- function(mu, alpha) { 1 - pnorm(qnorm(1 - alpha) - mu) }
loglik_mu <- function(mu, z, alpha) sum(mu * z - 0.5 * mu^2 - log(pi_mu(mu, alpha)))
opt <- optimize(function(m) -loglik_mu(m, z_one, alpha), c(-5, 5))
mu_hat <- opt$minimum
theta_hat_conditional <- mean(dat_pub$se) * mu_hat  # rough mapping for illustration

list(
  n_sig = n_sig,
  right_skew_p = p_right_skew,
  pcurve_mu_hat = mu_hat,
  pcurve_theta_on_avg_SE = theta_hat_conditional
)
```

### 4.8 p-uniform / p-uniform\* (puniform)

```{r}
if (requireNamespace("puniform", quietly = TRUE)) {
  library(puniform)
  # puniform expects effect sizes and variances; focuses on significant studies by default
  pu <- puniform(yi = dat_pub$yi, vi = dat_pub$se^2, method = "ML", side = "two.sided")
  # p-uniform* (random-effects) via argument method = "ML" and type = "puniform*" in newer versions
  # (API differs by version; we report whatever is available.)
  list(puniform_est = pu$est, puniform_ci = pu$ci)
} else {
  "puniform not installed; skip."
}
```

### 4.9 z-curve 2.0 (EDR/ERR)

```{r}
if (requireNamespace("zcurve", quietly = TRUE)) {
  library(zcurve)
  # zcurve accepts p-values (two-sided) of significant results or all with method; common use includes p <= .05
  p_for_zcurve <- dat_pub$p
  fit_zc <- zcurve(p_for_zcurve, method = "two.sided", ci = 0.95)
  summ_zc <- summary(fit_zc)
  list(EDR = summ_zc$EDR, ERR = summ_zc$ERR, components = summ_zc$Coef)
} else {
  "zcurve not installed; skip."
}
```

### 4.10 Robust Bayesian meta-analysis (RoBMA)

```{r}
if (requireNamespace("RoBMA", quietly = TRUE)) {
  library(RoBMA)
  fit_rb <- RoBMA(dat_pub$yi, dat_pub$se,
                  priors_effect = c("spike_and_slab"),          # effect present/absent
                  priors_heterogeneity = c("informed","spike"), # RE vs FE
                  priors_bias = c("none","pet_peese","selection"))
  summ_rb <- summary(fit_rb)
  list(RoBMA_post_mean = summ_rb$mu, BF_H1_H0 = summ_rb$`BF (H1/H0)`, 
       posterior_model_prob = summ_rb$`post. prob.`)
} else {
  "RoBMA not installed; skip."
}
```

------------------------------------------------------------------------

## 5. Comparative Summary Table (from the run above)

```{r}
# Collect whatever we computed into a compact table (robust to missing packages)
summ <- list()

# Base RE and PET-PEESE
summ$RE_theta     <- tryCatch(as.numeric(rma_re$b[1]), error = function(e) NA)
summ$RE_tau2      <- tryCatch(as.numeric(rma_re$tau2), error = function(e) NA)
summ$Egger_p      <- tryCatch(as.numeric(regtest(rma_re, model="lm")$pval), error=function(e) NA)
summ$PET          <- tryCatch(as.numeric(coef(pet)[1]), error = function(e) NA)
summ$PEESE        <- tryCatch(as.numeric(coef(peese)[1]), error = function(e) NA)

# Trim-and-fill
summ$TrimFill     <- tryCatch(as.numeric(trimfill(rma_re)$b[1]), error = function(e) NA)

# Copas (if available)
if (exists("cop")) summ$Copas_adj <- tryCatch(as.numeric(cop$adj), error=function(e) NA) else summ$Copas_adj <- NA

# Weight-function
if (exists("wf"))  summ$Weightr   <- tryCatch(as.numeric(wf$adj.est), error=function(e) NA) else summ$Weightr <- NA

# p-curve diag
summ$pcurve_right_skew_p <- tryCatch(p_right_skew, error=function(e) NA)
summ$pcurve_mu_hat       <- tryCatch(mu_hat, error=function(e) NA)

# p-uniform
if (exists("pu")) {
  summ$puniform_est <- tryCatch(as.numeric(pu$est), error=function(e) NA)
} else summ$puniform_est <- NA

# z-curve
if (exists("summ_zc")) {
  summ$zcurve_EDR <- tryCatch(as.numeric(summ_zc$EDR[1]), error=function(e) NA)
  summ$zcurve_ERR <- tryCatch(as.numeric(summ_zc$ERR[1]), error=function(e) NA)
} else { summ$zcurve_EDR <- NA; summ$zcurve_ERR <- NA }

# RoBMA
if (exists("summ_rb")) {
  summ$RoBMA_mu <- tryCatch(as.numeric(summ_rb$mu[1]), error=function(e) NA)
} else summ$RoBMA_mu <- NA

tbl <- tibble::enframe(summ, name = "Statistic", value = "Estimate")
kable(tbl, "html") %>% kable_styling(full_width = FALSE)
```

**Interpretation guide** (what to expect in this simulation):

-   **Funnel asymmetry** and **Egger** should often be **significant**.
-   **Trim-and-fill** typically moves the pooled effect **toward zero**.
-   **Weightr** (weight-function) should give an **attenuated** adjusted estimate and reveal **higher weights** for $p\le .05$.
-   **Copas** will provide a **range** of adjusted effects under different selection strengths.
-   **p-curve** right-skew test likely **rejects** uniformity (evidential value present), but the **effect MLE** can be **fragile**.
-   **p-uniform**\* often yields **attenuation** vs. naĂŻve RE.
-   **z-curve** EDR could be **below** the observed discovery rate (signaling selection); ERR quantifies replication probability.
-   **RoBMA** should **shrink** toward zero when bias models get posterior weight.

------------------------------------------------------------------------

## 6. Deep Dive: p-Curve's Statistical Properties (and Critiques)

Let $P_i$ be **dependent** across multiple specifications from the same dataset (violating independence). Conditioning on $P_i\le\alpha$ and maximizing $\prod_i f_\alpha(p_i\mid \mu)$ yields an estimator $\hat\mu$ that is:

-   **Biased** under heterogeneity ($\mu_i$ vary), because $$
    \log L(\mu) = \sum_i \big[\mu z_i - \tfrac{1}{2}\mu^2 - \log \pi(\mu,\alpha)\big]
    $$ is not the log-likelihood of any convex mixture unless $\mu_i\equiv \mu$.
-   **Under-covering** for CIs that ignore mixture/selection uncertainty.
-   **Non-invariant** to the exact test family if $p$'s are pooled from $t$, $F$, $\chi^2$, etc. without harmonization.
-   **Sensitive** to inclusion rules: "p-hacked" $p$'s do not follow $f_\alpha(p\mid \mu)$, invalidating calibration.

These points are central to critiques such as **Morey & Davis-Stober**, which marshal formal results and simulations showing **poor finite-sample behavior** and **miscalibration** for p-curve as a **meta-analytic estimator**. The practical upshot: use p-curve to **diagnose** evidential value and complement it with **p-uniform**\*, **weight-function models**, or **RoBMA** for effect estimation.

------------------------------------------------------------------------

## 7. Extended Worked Example: Heterogeneity & Bias Stress Test

We stress-test methods by increasing heterogeneity and the strength of selection.

```{r}
set.seed(11)
simulate_corpus <- function(K=200, theta_bar=0.15, tau=0.25, hack_bump=0.5, sel_sig=0.95, sel_nsig=0.1) {
  n_i   <- sample(40:250, K, TRUE)
  se_i  <- sqrt(2/n_i)
  th_i  <- rnorm(K, theta_bar, tau)
  z0    <- rnorm(K, th_i/se_i, 1)
  p0    <- 2*(1 - pnorm(abs(z0)))
  hack  <- runif(K) < 0.5
  z1    <- z0 + rnorm(K, hack_bump, 0.25)
  z_use <- ifelse(p0 > .05 & hack, z1, z0)
  p_use <- 2*(1 - pnorm(abs(z_use)))
  sel   <- rbinom(K, 1, ifelse(p_use <= .05, sel_sig, sel_nsig)) == 1
  data.frame(yi = z_use * se_i, se = se_i, zi = z_use, p = p_use, published = sel)
}

# Stressful scenario
dat2 <- subset(simulate_corpus(tau=0.30, hack_bump=0.6, sel_sig=0.98, sel_nsig=0.05), published)

# Run a subset of methods
rma2 <- rma.uni(yi = dat2$yi, sei = dat2$se, method = "REML")
tf2  <- tryCatch(trimfill(rma2), error=function(e) NULL)
pet2 <- rma.uni(yi = dat2$yi, vi = dat2$se^2, mods = ~ I(dat2$se), method = "REML")
peese2 <- rma.uni(yi = dat2$yi, vi = dat2$se^2, mods = ~ I(dat2$se^2), method = "REML")

out2 <- list(
  RE = c(est = rma2$b[1], tau2 = rma2$tau2),
  TF = if(!is.null(tf2)) c(est = tf2$b[1]) else NA,
  PET = coef(pet2)[1],
  PEESE = coef(peese2)[1]
)

if (requireNamespace("weightr", quietly = TRUE)) {
  library(weightr)
  wf2 <- weightfunct(dat2$yi, dat2$se^2, steps = c(.025,.05,.10))
  out2$Weightr <- wf2$adj.est
}

if (requireNamespace("zcurve", quietly = TRUE)) {
  library(zcurve)
  zc2 <- zcurve(dat2$p, method = "two.sided")
  out2$zcurve_EDR <- tryCatch(summary(zc2)$EDR[1], error=function(e) NA)
  out2$zcurve_ERR <- tryCatch(summary(zc2)$ERR[1], error=function(e) NA)
}

if (requireNamespace("RoBMA", quietly = TRUE)) {
  library(RoBMA)
  rb2 <- RoBMA(dat2$yi, dat2$se, priors_bias = c("none","pet_peese","selection"))
  out2$RoBMA_mu <- tryCatch(summary(rb2)$mu[1], error=function(e) NA)
}

out2
```

**What to look for**: As stress increases,

-   naĂŻve RE inflates; PET/PEESE may overshoot/undershoot;
-   trim-and-fill attenuates somewhat;
-   weightr often gives stronger attenuation with explicit weights;
-   z-curve's **EDR** often falls **below** the observed discovery rate (diagnosing selection);
-   RoBMA shrinks via posterior mass on bias models.

------------------------------------------------------------------------

## 8. Practical Recommendations (PhD-Level)

1.  **Diagnostic triad**: (i) Funnel+Egger, (ii) Curve method (p-uniform\* or p-curve diagnostics), (iii) Selection model (weight-function and/or Copas).
2.  **Primary estimator**: Prefer **RoBMA** or **weight-function** as the headline adjusted effect; cross-check with **p-uniform**\* and **PET-PEESE**.
3.  **Heterogeneity first**: Always model random effects (or mixtures). Report $\hat\tau^2$ and sensitivity to heterogeneity priors (Bayesian).
4.  **Dependence and multiverse**: If multiple estimates per study, use **multilevel meta-analysis** or **robust variance**; report **specification curves**.
5.  **Report selection function**: With weight-function models, print the $\hat w_b$ table; with z-curve, report EDR/ERR.
6.  **Do not over-interpret p-curve MLE**: Treat p-curve as a **diagnostic** unless strong conditions are met; cite the critiques.
7.  **Transparency beats ex post correction**: Pre-registration, registered reports, shared code/materials reduce the need for ex post bias surgery.

------------------------------------------------------------------------

## 9. Annotated References (selected; align with methods above)

-   **Funnel & PET-PEESE**: Egger et al. (1997), Stanley & Doucouliagos (2014).
-   **Trim-and-fill**: Duval & Tweedie (2000).
-   **Selection models**: Iyengar & Greenhouse (1988); Hedges (1984); Vevea & Hedges (1995); Vevea & Woods (2005).
-   **Copas sensitivity**: Copas & Shi (2001).
-   **p-curve**: Simonsohn, Nelson & Simmons (2014).
-   **p-uniform / p-uniform**\*: van Assen, van Aert & Wicherts (2015); van Aert & van Assen (2021).
-   **z-curve 2.0**: BartoĹˇ, Maier, Wagenmakers & Schimmack (2021+).
-   **RoBMA**: Maier, BartoĹˇ, Wagenmakers & colleagues (Robust Bayesian Meta-Analysis).
-   **Excess significance**: Ioannidis & Trikalinos (2007).
-   **Forensics**: Nuijten et al. (2016) statcheck; GRIM/GRIMMER (Brown & Heathers; Heathers et al.).
-   **p-curve critiques**: Morey & Davis-Stober, "On the poor statistical properties of the p-curve meta-analytic procedure" (preprint/article); also broader debates in Carter et al. (2019) comparing bias-correction methods.

*(Please replace with exact bibliographic details in your reference manager; the narrative above reflects the mainstream literature up to mid-2024.)*

------------------------------------------------------------------------

## 10. Appendix: Minimal Code to Run Forensics (statcheck) on a Text String

```{r}
if (requireNamespace("statcheck", quietly = TRUE)) {
  library(statcheck)
  txt <- "We found a significant effect, t(38) = 2.10, p = .025, and another t(24) = 1.50, p = .045."
  res <- statcheck(txt, OneFile = TRUE, quietly = TRUE)
  res[, c("Statistic", "df1", "df2", "Value", "Reported.Comparison", "Reported.p", "Computed", "Consistent")]
} else {
  "statcheck not installed; skip."
}
```

------------------------------------------------------------------------

## 11. Appendix: Hand-rolled Excess Significance Test (Poisson-Binomial)

```{r}
# Estimate power per study under the RE pooled effect (caution: circularity)
theta_hat <- as.numeric(rma_re$b[1])
z_alpha   <- qnorm(1 - 0.05/2)

# Using per-study noncentrality mu_i = theta_hat / se_i (two-sided test)
power_i <- pnorm(-z_alpha - theta_hat/dat_pub$se) + (1 - pnorm(z_alpha - theta_hat/dat_pub$se))
O_obs   <- sum(dat_pub$p <= 0.05)

# Approximate p-value using normal approximation to Poisson-binomial
E <- sum(power_i); V <- sum(power_i*(1-power_i))
z_os <- (O_obs - E) / sqrt(V + 1e-9)
p_excess <- 1 - pnorm(z_os)

list(observed_signif = O_obs, expected_signif = E, p_excess_sig = p_excess)
```

------------------------------------------------------------------------

## 12. Closing Notes for Practitioners

-   **Triangulate** across methods; **report assumptions**; perform **sensitivity**.
-   Prefer **model-averaged** or **explicit selection** estimators for decisions; use curve methods to **diagnose**, not to anchor estimates.
-   In business/finance/marketing applications, heterogeneity and researcher degrees of freedom are **large**---treat independence as the exception, not the norm.

## 2. Mathematical Core (Expanded)

### 2.1 Selection on significance vs. p-hacking

Let test statistic $Z \sim \mathcal{N}(\mu,1)$; two-sided $p=2[1-\Phi(|Z|)]$. A **selection function** $s(z)$ yields the observed $z$-density $$
f_{\mathrm{obs}}(z) \propto s(z) \int \varphi(z-\mu), dG(\mu),
$$ where $G$ is the mixing distribution of noncentralities $\mu$ and $\varphi$ is the standard normal density.

-   **Publication bias**: $s(z)$ jumps at $|z|=z_{1-\alpha/2}$.
-   **p-hacking**: $s(z)$ has local spikes/bunching near thresholds (e.g., **bunching** just above $1.96$).

**Identifiability**: Without restrictions, $s$ and $G$ are not point-identified. Restrictions include: monotonicity/stepwise $s$, parametric mixtures for $G$, or conditioning on significance (curve methods).

### 2.2 PET-PEESE (precision meta-regression)

Let $y_i$ and $s_i$ be effect and SE. Under $y_i=\theta + \beta s_i^\delta + \varepsilon_i$ with weights $1/s_i^2$,

-   PET: $\delta=1$; $\hat\theta=\hat\beta_0$ as $s_i \to 0$.
-   PEESE: $\delta=2$; often used if PET rejects $\theta=0$.

Assumes bias is **smooth in precision**; conflates genuine small-study heterogeneity with bias if the relation is not linear in $s_i^\delta$.

### 2.3 Weight-function selection models

For $p$-bins ${\mathcal{I}*b}$, let $w_b>0$ be selection weights. Fixed effects: $$
L(\theta,\mathbf{w}) \propto \prod*{i=1}^k \frac{\phi\big((y_i-\theta)/s_i\big) s_i^{-1} , w_{b(i)}}
{\sum_{b} w_b , \Pr_\theta\big(p(y)\in \mathcal{I}_b ,\big|; s_i\big)}.
$$ Random effects integrate $\theta_i \sim \mathcal{N}(\theta,\tau^2)$ in numerator and denominator. $\mathbf{w}$ is identified **up to scale** (fix one bin to 1).

### 2.4 p-curve and p-uniform\*

Given $P=1-\Phi(Z)$ (one-sided) with power $\pi(\mu,\alpha)=1-\Phi(z_{1-\alpha}-\mu)$, the **conditional density** given $P \le \alpha$ is $$
f_\alpha(p \mid \mu) = \frac{\exp{\mu z - \tfrac{1}{2}\mu^2}}{\pi(\mu,\alpha)}, \quad z=\Phi^{-1}(1-p).
$$

-   **p-curve**: diagnostic tests of right-skew; sometimes MLE for $\mu$.
-   **p-uniform**: MLE for effect under **homogeneous** $\mu$;
-   **p-uniform**\*: extends to random effects; better finite-sample properties under heterogeneity.

**Key caveat**: All **curve** methods condition on $p\le\alpha$; if selection is not (mostly) on this threshold or if $p$'s are **not independent** (e.g., many specs from one dataset), properties degrade.

### 2.5 z-curve 2.0

Model $Z \mid |Z|\ge z_{1-\alpha/2}$ as a **finite mixture** $\sum_k \pi_k ,\mathcal{N}(\mu_k,1)$ with truncation. Estimate via EM with truncation corrections. Compute $$
\mathrm{EDR} = \int \Pr(|Z| \ge z_{1-\alpha/2} \mid \mu), d\hat G(\mu), \qquad
\mathrm{ERR} = \mathbb{E}\big[\Pr(|Z^\star|\ge z_{1-\alpha/2} \mid \mu)\mid |Z|\ge z_{1-\alpha/2}\big].
$$ EDR acts as a **power-weighted prevalence** of significance; ERR is a model-based **replication** probability.

### 2.6 Copas--Shi sensitivity

Assume latent selection driven by a correlation between the probability of publication and the **study precision** (or its proxy). The Copas model posits a latent selection variable influencing whether a study is observed; sensitivity is performed over a grid of selection parameters to project bias-adjusted effects.

------------------------------------------------------------------------

## 3. The "Beef": Improvements, Disagreements, and What They Mean

-   **Curve methods vs. selection models**

    -   *Curve advocates* emphasize robustness when selection is heavily on the $p \le \alpha$ threshold and when nonsignificant studies are missing entirely.
    -   *Selection-model advocates* argue for **explicit modeling** of selection and the use of **all** studies (sig + nsig), producing **directly adjusted** effects under transparent assumptions about $w(p)$.

-   **p-uniform vs. p-uniform**\*

    -   Original p-uniform can be **biased** under heterogeneity and low power; p-uniform\* introduces random effects and better coverage, but remains sensitive to **mis-specification of test families** and **dependence** among $p$'s.

-   **z-curve vs. others**

    -   z-curve's **mixture flexibility** handles heterogeneity well; **EDR/ERR** are intuitive to users.
    -   Critics point to **component number selection**, **truncation identification**, and **sensitivity** to the $p$-value curation (e.g., inclusion criteria, dependence).

-   **PET-PEESE vs. weight-functions**

    -   PET-PEESE is **transparent and simple** but may **overcorrect** when small-study effects reflect **real** heterogeneity;
    -   Weight-functions can be **unstable** if the stepwise form is misspecified, but they directly target the mechanism producing asymmetry.

-   **Excess significance**

    -   Intuitive, but depends on a (possibly biased) effect used to compute **power**; heterogeneity and **researcher degrees of freedom** can inflate significance beyond what the test uses as "expected".

-   **On the statistical properties of p-curve (Morey & Davis-Stober critique)**

    -   Recent critiques (e.g., "On the poor statistical properties of the p-curve meta-analytic procedure") argue that p-curve's **effect estimator** can exhibit **poor calibration**, **bias** and **under-coverage**, especially under heterogeneity, misspecified test families, and dependent $p$'s.
    -   Conceptual points include: conditioning on $p\le\alpha$ discards information, the estimator may be **non-invariant** to certain reparameterizations, and finite-sample behavior can be **fragile**.
    -   **Implication**: Treat p-curve primarily as a **diagnostic** for evidential value; for estimation, prefer **p-uniform**\*, **selection models**, or **model-averaged Bayesian** methods (RoBMA).

------------------------------------------------------------------------

## 4. Simulating p-Hacking & Selection, Then Diagnosing With Many Tools

We'll simulate a meta-analytic corpus with:

1.  **True heterogeneity**, 2) **publication bias** favoring $p\le .05$, and 3) **p-hacking** via a simple "add a covariate" retry if initial $p> .05$.

```{r}
set.seed(2025)

# --- Simulation design ---
K         <- 150                      # number of studies
theta_bar <- 0.20                     # grand mean
tau       <- 0.15                     # heterogeneity SD
n_i       <- sample(30:200, K, TRUE)  # per-study sample sizes
se_i      <- sqrt(2/n_i)              # SE for standardized mean difference (toy)
theta_i   <- rnorm(K, theta_bar, tau) # random effects

# Generate study z and two-sided p under a simple z-test
z_raw     <- rnorm(K, mean = theta_i / se_i, sd = 1)
p_raw     <- 2 * (1 - pnorm(abs(z_raw)))

# Simple p-hacking: if p>.05, "try a covariate" with correlated boost in z
# (This is stylized; real p-hacking is more complex.)
hack      <- runif(K) < 0.5           # only some attempt hacking
z_hack    <- z_raw + rnorm(K, 0.35, 0.20) # small optimistic bump
p_hacked  <- 2 * (1 - pnorm(abs(ifelse(p_raw > .05 & hack, z_hack, z_raw))))

# Publication selection: favor p<=.05 with higher selection probability
sel_prob  <- ifelse(p_hacked <= .05, 0.9, 0.2)
published <- rbinom(K, 1, sel_prob) == 1

dat <- data.frame(
  study = seq_len(K),
  n = n_i,
  se = se_i,
  yi = (ifelse(p_raw > .05 & hack, z_hack, z_raw)) * se_i,  # back to effect scale
  zi = ifelse(p_raw > .05 & hack, z_hack, z_raw),
  p = p_hacked,
  published = published
)

dat_pub <- subset(dat, published)

summary(list(
  K_total = nrow(dat),
  K_published = nrow(dat_pub),
  prop_significant = mean(dat_pub$p <= .05),
  mean_effect_published = mean(dat_pub$yi),
  sd_effect_published = sd(dat_pub$yi)
))
```

### 4.1 Visual diagnostics: p-histogram & caliper test

```{r}
library(ggplot2)
p_sig <- dat_pub$p[dat_pub$p <= .10]  # focus near .10 for visualization
plt_phist <- ggplot(data.frame(p = dat_pub$p), aes(p)) +
  geom_histogram(binwidth = 0.01, boundary = 0, color = "white", fill = "#3182bd") +
  geom_vline(xintercept = 0.05, linetype = 2, color = "red") +
  labs(title = "Histogram of Published p-values", subtitle = "Dashed line at 0.05", x = "p", y = "Count")

# Caliper around z=1.96 (two-sided .05)
zvals <- dat_pub$zi
c <- 1.96; h <- 0.1
N_plus  <- sum(zvals > c & zvals <= c+h)
N_minus <- sum(zvals > c-h & zvals <= c)
p_caliper <- binom.test(N_plus, N_plus + N_minus, 0.5, alternative = "greater")$p.value

plt_phist
list(N_above = N_plus, N_below = N_minus, caliper_p = p_caliper)
```

### 4.2 Classical random-effects meta-analysis, funnel plot, and Egger

```{r}
library(metafor)

# Random-effects (REML)
rma_re <- rma.uni(yi = dat_pub$yi, sei = dat_pub$se, method = "REML")
egger   <- regtest(rma_re, model = "lm")  # Egger's test of funnel asymmetry

# Funnel plot
funnel_data <- data.frame(yi = dat_pub$yi, sei = dat_pub$se, wi = 1/dat_pub$se^2)
plot(rma_re, main = "Funnel Plot (metafor)")

list(RE_theta = c(est = rma_re$b[1], se = rma_re$se[1]),
     tau2 = rma_re$tau2,
     Egger_intercept = egger$beta[1], Egger_p = egger$pval)
```

### 4.3 PET-PEESE

```{r}
# PET
pet <- rma.uni(yi = dat_pub$yi, vi = dat_pub$se^2, mods = ~ I(dat_pub$se), method = "REML")
# PEESE
peese <- rma.uni(yi = dat_pub$yi, vi = dat_pub$se^2, mods = ~ I(dat_pub$se^2), method = "REML")

list(
  PET_intercept = coef(pet)[1], PET_p = coef(summary(pet))[1,4],
  PEESE_intercept = coef(peese)[1], PEESE_p = coef(summary(peese))[1,4]
)
```

### 4.4 Trim-and-fill

```{r}
tf <- trimfill(rma_re)
list(TF_theta = c(est = tf$b[1], se = tf$se[1]), n_imputed = tf$k0)
```

### 4.5 Copas--Shi sensitivity (metasens)

```{r}
if (requireNamespace("metasens", quietly = TRUE)) {
  library(metasens)
  cop <- copas(TE = dat_pub$yi, seTE = dat_pub$se)
  list(Copas_theta_adj = cop$adj, Copas_note = "Interpret across sensitivity grid")
} else {
  "metasens not installed; skip Copas."
}
```

### 4.6 Weight-function selection model (weightr)

```{r}
if (requireNamespace("weightr", quietly = TRUE)) {
  library(weightr)
  # We need two-sided p-values; use dat_pub$p
  wf <- weightfunct(effect = dat_pub$yi, v = dat_pub$se^2,
                    steps = c(0.025, 0.05, 0.10), table = TRUE)
  list(weightr_theta = wf$adj.est, weights = wf$weights.table)
} else {
  "weightr not installed; skip weight-function."
}
```

### 4.7 p-curve diagnostics (custom): right-skew & conditional MLE

```{r}
p_sig <- dat_pub$p[dat_pub$p <= 0.05]
alpha <- 0.05
n_sig <- length(p_sig)

# Right-skew binomial test: P(p < alpha/2 | p <= alpha) = 0.5 under null of no evidential value
below <- sum(p_sig < alpha/2)
p_right_skew <- binom.test(below, n_sig, 0.5, alternative = "greater")$p.value

# Conditional MLE for one-sided z (diagnostic; do not over-interpret as grand mean)
z_one <- qnorm(1 - p_sig) # treat as one-sided to match the transform
pi_mu <- function(mu, alpha) { 1 - pnorm(qnorm(1 - alpha) - mu) }
loglik_mu <- function(mu, z, alpha) sum(mu * z - 0.5 * mu^2 - log(pi_mu(mu, alpha)))
opt <- optimize(function(m) -loglik_mu(m, z_one, alpha), c(-5, 5))
mu_hat <- opt$minimum
theta_hat_conditional <- mean(dat_pub$se) * mu_hat  # rough mapping for illustration

list(
  n_sig = n_sig,
  right_skew_p = p_right_skew,
  pcurve_mu_hat = mu_hat,
  pcurve_theta_on_avg_SE = theta_hat_conditional
)
```

### 4.8 p-uniform / p-uniform\* (puniform)

```{r}
if (requireNamespace("puniform", quietly = TRUE)) {
  library(puniform)
  # puniform expects effect sizes and variances; focuses on significant studies by default
  pu <- puniform(yi = dat_pub$yi, vi = dat_pub$se^2, method = "ML", side = "two.sided")
  # p-uniform* (random-effects) via argument method = "ML" and type = "puniform*" in newer versions
  # (API differs by version; we report whatever is available.)
  list(puniform_est = pu$est, puniform_ci = pu$ci)
} else {
  "puniform not installed; skip."
}
```

### 4.9 z-curve 2.0 (EDR/ERR)

```{r}
if (requireNamespace("zcurve", quietly = TRUE)) {
  library(zcurve)
  # zcurve accepts p-values (two-sided) of significant results or all with method; common use includes p <= .05
  p_for_zcurve <- dat_pub$p
  fit_zc <- zcurve(p_for_zcurve, method = "two.sided", ci = 0.95)
  summ_zc <- summary(fit_zc)
  list(EDR = summ_zc$EDR, ERR = summ_zc$ERR, components = summ_zc$Coef)
} else {
  "zcurve not installed; skip."
}
```

### 4.10 Robust Bayesian meta-analysis (RoBMA)

```{r}
if (requireNamespace("RoBMA", quietly = TRUE)) {
  library(RoBMA)
  fit_rb <- RoBMA(dat_pub$yi, dat_pub$se,
                  priors_effect = c("spike_and_slab"),          # effect present/absent
                  priors_heterogeneity = c("informed","spike"), # RE vs FE
                  priors_bias = c("none","pet_peese","selection"))
  summ_rb <- summary(fit_rb)
  list(RoBMA_post_mean = summ_rb$mu, BF_H1_H0 = summ_rb$`BF (H1/H0)`, 
       posterior_model_prob = summ_rb$`post. prob.`)
} else {
  "RoBMA not installed; skip."
}
```

------------------------------------------------------------------------

## 5. Comparative Summary Table (from the run above)

```{r}
# Collect whatever we computed into a compact table (robust to missing packages)
summ <- list()

# Base RE and PET-PEESE
summ$RE_theta     <- tryCatch(as.numeric(rma_re$b[1]), error = function(e) NA)
summ$RE_tau2      <- tryCatch(as.numeric(rma_re$tau2), error = function(e) NA)
summ$Egger_p      <- tryCatch(as.numeric(regtest(rma_re, model="lm")$pval), error=function(e) NA)
summ$PET          <- tryCatch(as.numeric(coef(pet)[1]), error = function(e) NA)
summ$PEESE        <- tryCatch(as.numeric(coef(peese)[1]), error = function(e) NA)

# Trim-and-fill
summ$TrimFill     <- tryCatch(as.numeric(trimfill(rma_re)$b[1]), error = function(e) NA)

# Copas (if available)
if (exists("cop")) summ$Copas_adj <- tryCatch(as.numeric(cop$adj), error=function(e) NA) else summ$Copas_adj <- NA

# Weight-function
if (exists("wf"))  summ$Weightr   <- tryCatch(as.numeric(wf$adj.est), error=function(e) NA) else summ$Weightr <- NA

# p-curve diag
summ$pcurve_right_skew_p <- tryCatch(p_right_skew, error=function(e) NA)
summ$pcurve_mu_hat       <- tryCatch(mu_hat, error=function(e) NA)

# p-uniform
if (exists("pu")) {
  summ$puniform_est <- tryCatch(as.numeric(pu$est), error=function(e) NA)
} else summ$puniform_est <- NA

# z-curve
if (exists("summ_zc")) {
  summ$zcurve_EDR <- tryCatch(as.numeric(summ_zc$EDR[1]), error=function(e) NA)
  summ$zcurve_ERR <- tryCatch(as.numeric(summ_zc$ERR[1]), error=function(e) NA)
} else { summ$zcurve_EDR <- NA; summ$zcurve_ERR <- NA }

# RoBMA
if (exists("summ_rb")) {
  summ$RoBMA_mu <- tryCatch(as.numeric(summ_rb$mu[1]), error=function(e) NA)
} else summ$RoBMA_mu <- NA

tbl <- tibble::enframe(summ, name = "Statistic", value = "Estimate")
kable(tbl, "html") %>% kable_styling(full_width = FALSE)
```

**Interpretation guide** (what to expect in this simulation):

-   **Funnel asymmetry** and **Egger** should often be **significant**.
-   **Trim-and-fill** typically moves the pooled effect **toward zero**.
-   **Weightr** (weight-function) should give an **attenuated** adjusted estimate and reveal **higher weights** for $p\le .05$.
-   **Copas** will provide a **range** of adjusted effects under different selection strengths.
-   **p-curve** right-skew test likely **rejects** uniformity (evidential value present), but the **effect MLE** can be **fragile**.
-   **p-uniform**\* often yields **attenuation** vs. naĂŻve RE.
-   **z-curve** EDR could be **below** the observed discovery rate (signaling selection); ERR quantifies replication probability.
-   **RoBMA** should **shrink** toward zero when bias models get posterior weight.

------------------------------------------------------------------------

## 6. Deep Dive: p-Curve's Statistical Properties (and Critiques)

Let $P_i$ be **dependent** across multiple specifications from the same dataset (violating independence). Conditioning on $P_i\le\alpha$ and maximizing $\prod_i f_\alpha(p_i\mid \mu)$ yields an estimator $\hat\mu$ that is:

-   **Biased** under heterogeneity ($\mu_i$ vary), because $$
    \log L(\mu) = \sum_i \big[\mu z_i - \tfrac{1}{2}\mu^2 - \log \pi(\mu,\alpha)\big]
    $$ is not the log-likelihood of any convex mixture unless $\mu_i\equiv \mu$.
-   **Under-covering** for CIs that ignore mixture/selection uncertainty.
-   **Non-invariant** to the exact test family if $p$'s are pooled from $t$, $F$, $\chi^2$, etc. without harmonization.
-   **Sensitive** to inclusion rules: "p-hacked" $p$'s do not follow $f_\alpha(p\mid \mu)$, invalidating calibration.

These points are central to critiques such as **Morey & Davis-Stober**, which marshal formal results and simulations showing **poor finite-sample behavior** and **miscalibration** for p-curve as a **meta-analytic estimator**. The practical upshot: use p-curve to **diagnose** evidential value and complement it with **p-uniform**\*, **weight-function models**, or **RoBMA** for effect estimation.

------------------------------------------------------------------------

## 7. Extended Worked Example: Heterogeneity & Bias Stress Test

We stress-test methods by increasing heterogeneity and the strength of selection.

```{r}
set.seed(11)
simulate_corpus <- function(K=200, theta_bar=0.15, tau=0.25, hack_bump=0.5, sel_sig=0.95, sel_nsig=0.1) {
  n_i   <- sample(40:250, K, TRUE)
  se_i  <- sqrt(2/n_i)
  th_i  <- rnorm(K, theta_bar, tau)
  z0    <- rnorm(K, th_i/se_i, 1)
  p0    <- 2*(1 - pnorm(abs(z0)))
  hack  <- runif(K) < 0.5
  z1    <- z0 + rnorm(K, hack_bump, 0.25)
  z_use <- ifelse(p0 > .05 & hack, z1, z0)
  p_use <- 2*(1 - pnorm(abs(z_use)))
  sel   <- rbinom(K, 1, ifelse(p_use <= .05, sel_sig, sel_nsig)) == 1
  data.frame(yi = z_use * se_i, se = se_i, zi = z_use, p = p_use, published = sel)
}

# Stressful scenario
dat2 <- subset(simulate_corpus(tau=0.30, hack_bump=0.6, sel_sig=0.98, sel_nsig=0.05), published)

# Run a subset of methods
rma2 <- rma.uni(yi = dat2$yi, sei = dat2$se, method = "REML")
tf2  <- tryCatch(trimfill(rma2), error=function(e) NULL)
pet2 <- rma.uni(yi = dat2$yi, vi = dat2$se^2, mods = ~ I(dat2$se), method = "REML")
peese2 <- rma.uni(yi = dat2$yi, vi = dat2$se^2, mods = ~ I(dat2$se^2), method = "REML")

out2 <- list(
  RE = c(est = rma2$b[1], tau2 = rma2$tau2),
  TF = if(!is.null(tf2)) c(est = tf2$b[1]) else NA,
  PET = coef(pet2)[1],
  PEESE = coef(peese2)[1]
)

if (requireNamespace("weightr", quietly = TRUE)) {
  library(weightr)
  wf2 <- weightfunct(dat2$yi, dat2$se^2, steps = c(.025,.05,.10))
  out2$Weightr <- wf2$adj.est
}

if (requireNamespace("zcurve", quietly = TRUE)) {
  library(zcurve)
  zc2 <- zcurve(dat2$p, method = "two.sided")
  out2$zcurve_EDR <- tryCatch(summary(zc2)$EDR[1], error=function(e) NA)
  out2$zcurve_ERR <- tryCatch(summary(zc2)$ERR[1], error=function(e) NA)
}

if (requireNamespace("RoBMA", quietly = TRUE)) {
  library(RoBMA)
  rb2 <- RoBMA(dat2$yi, dat2$se, priors_bias = c("none","pet_peese","selection"))
  out2$RoBMA_mu <- tryCatch(summary(rb2)$mu[1], error=function(e) NA)
}

out2
```

**What to look for**: As stress increases,

-   naĂŻve RE inflates; PET/PEESE may overshoot/undershoot;
-   trim-and-fill attenuates somewhat;
-   weightr often gives stronger attenuation with explicit weights;
-   z-curve's **EDR** often falls **below** the observed discovery rate (diagnosing selection);
-   RoBMA shrinks via posterior mass on bias models.

------------------------------------------------------------------------

## 8. Practical Recommendations (PhD-Level)

1.  **Diagnostic triad**: (i) Funnel+Egger, (ii) Curve method (p-uniform\* or p-curve diagnostics), (iii) Selection model (weight-function and/or Copas).
2.  **Primary estimator**: Prefer **RoBMA** or **weight-function** as the headline adjusted effect; cross-check with **p-uniform**\* and **PET-PEESE**.
3.  **Heterogeneity first**: Always model random effects (or mixtures). Report $\hat\tau^2$ and sensitivity to heterogeneity priors (Bayesian).
4.  **Dependence and multiverse**: If multiple estimates per study, use **multilevel meta-analysis** or **robust variance**; report **specification curves**.
5.  **Report selection function**: With weight-function models, print the $\hat w_b$ table; with z-curve, report EDR/ERR.
6.  **Do not over-interpret p-curve MLE**: Treat p-curve as a **diagnostic** unless strong conditions are met; cite the critiques.
7.  **Transparency beats ex post correction**: Pre-registration, registered reports, shared code/materials reduce the need for ex post bias surgery.

------------------------------------------------------------------------

## 9. Annotated References (selected; align with methods above)

-   **Funnel & PET-PEESE**: Egger et al. (1997), Stanley & Doucouliagos (2014).
-   **Trim-and-fill**: Duval & Tweedie (2000).
-   **Selection models**: Iyengar & Greenhouse (1988); Hedges (1984); Vevea & Hedges (1995); Vevea & Woods (2005).
-   **Copas sensitivity**: Copas & Shi (2001).
-   **p-curve**: Simonsohn, Nelson & Simmons (2014).
-   **p-uniform / p-uniform**\*: van Assen, van Aert & Wicherts (2015); van Aert & van Assen (2021).
-   **z-curve 2.0**: BartoĹˇ, Maier, Wagenmakers & Schimmack (2021+).
-   **RoBMA**: Maier, BartoĹˇ, Wagenmakers & colleagues (Robust Bayesian Meta-Analysis).
-   **Excess significance**: Ioannidis & Trikalinos (2007).
-   **Forensics**: Nuijten et al. (2016) statcheck; GRIM/GRIMMER (Brown & Heathers; Heathers et al.).
-   **p-curve critiques**: Morey & Davis-Stober, "On the poor statistical properties of the p-curve meta-analytic procedure" (preprint/article); also broader debates in Carter et al. (2019) comparing bias-correction methods.

*(Please replace with exact bibliographic details in your reference manager; the narrative above reflects the mainstream literature up to mid-2024.)*

------------------------------------------------------------------------

## 10. Appendix: Minimal Code to Run Forensics (statcheck) on a Text String

```{r}
if (requireNamespace("statcheck", quietly = TRUE)) {
  library(statcheck)
  txt <- "We found a significant effect, t(38) = 2.10, p = .025, and another t(24) = 1.50, p = .045."
  res <- statcheck(txt, OneFile = TRUE, quietly = TRUE)
  res[, c("Statistic", "df1", "df2", "Value", "Reported.Comparison", "Reported.p", "Computed", "Consistent")]
} else {
  "statcheck not installed; skip."
}
```

------------------------------------------------------------------------

## 11. Appendix: Hand-rolled Excess Significance Test (Poisson-Binomial)

```{r}
# Estimate power per study under the RE pooled effect (caution: circularity)
theta_hat <- as.numeric(rma_re$b[1])
z_alpha   <- qnorm(1 - 0.05/2)

# Using per-study noncentrality mu_i = theta_hat / se_i (two-sided test)
power_i <- pnorm(-z_alpha - theta_hat/dat_pub$se) + (1 - pnorm(z_alpha - theta_hat/dat_pub$se))
O_obs   <- sum(dat_pub$p <= 0.05)

# Approximate p-value using normal approximation to Poisson-binomial
E <- sum(power_i); V <- sum(power_i*(1-power_i))
z_os <- (O_obs - E) / sqrt(V + 1e-9)
p_excess <- 1 - pnorm(z_os)

list(observed_signif = O_obs, expected_signif = E, p_excess_sig = p_excess)
```

------------------------------------------------------------------------
