# Sampling

## Simple Sampling

Simple (random) Sampling

```{r}
library(dplyr)
iris_df <- iris
set.seed(1)
sample_n(iris_df, 10)
```

```{r}
library(sampling)
# set unique id number for each row 
iris_df$id = 1:nrow(iris_df)

# Simple random sampling with replacement
srswor(10, length(iris_df$id))

# Simple random sampling without replacement (sequential method)
srswor1(10, length(iris_df$id))

# Simple random sampling with replacement
srswr(10, length(iris_df$id))



```

```{r}
library(survey)
data("api")
srs_design <- svydesign(data = apistrat,
                        weights = ~pw, 
                        fpc = ~fpc, 
                        id = ~1)
```

```{r}
library(sampler)
rsamp(albania,
      n = 260,
      over = 0.1, # desired oversampling proportion
      rep = F)
```

Identify missing points between sample and collected data

```{r, eval = FALSE}
alsample <- rsamp(df = albania, 544)
alreceived <- rsamp(df = alsample, 390)
rmissing(sampdf = alsample,
         colldf = alreceived,
         col_name = qvKod)
```

## Stratified Sampling

A stratum is a subset of the population that has at least one common characteristic.

Steps:

1.  Identify relevant stratums and their representation in the population.
2.  Randomly sample to select a sufficient number of subjects from each stratum.

Stratified sampling reduces sampling error.

```{r}
library(dplyr)
# by number of rows
sample_iris <- iris %>%
    group_by(Species) %>%
    sample_n(5)
sample_iris

# by fraction
sample_iris <- iris %>%
    group_by(Species) %>%
    sample_frac(size = .15)
sample_iris

```

```{r}
library(sampler)
# Stratified sample using proportional allocation without replacement
ssamp(df=albania, n=360, strata=qarku, over=0.1)


```

Identify number of missing points by strata between sample and collected data

```{r, eval = FALSE}
alsample <- rsamp(df = albania, 544)
alreceived <- rsamp(df = alsample, 390)
smissing(
    sampdf = alsample,
    colldf = alreceived,
    strata = qarku,
    col_name = qvKod
)
```

## Unequal Probability Sampling

```{r, eval = FALSE}
UPbrewer()
UPmaxentropy()
UPmidzuno()
UPmidzunopi2()
UPmultinomial()
UPpivotal()
UPrandompivotal()
UPpoisson()
UPsampford()
UPsystematic()
UPrandomsystematic()
UPsystematicpi2()
UPtille()
UPtillepi2()
```

## Balanced Sampling

-   Purpose: to get the same means in the population and the sample for all the auxiliary variables

-   Balanced sampling is different from purposive selection

Balancing equations

$$
\sum_{k \in S} \frac{\mathbf{x}_k}{\pi_k} = \sum_{k \in U} \mathbf{x}_k
$$

where $\mathbf{x}_k$ is a vector of auxiliary variables

### Cube

-   flight phase

-   landing phase

```{r, eval = FALSE}
samplecube()
fastflightcube()
landingcube()

```

### Stratification

-   Try to replicate the population based on the original multivariate histogram

```{r}
library(survey)
data("api")
srs_design <- svydesign(data = apistrat,
                        weights = ~pw, 
                        fpc = ~fpc, 
                        strata = ~stype,
                        id = ~1)
```

```{r, eval = FALSE}
balancedstratification()

```

### Cluster

```{r}
library(survey)
data("api")
srs_design <- svydesign(data = apiclus1,
                        weights = ~pw, 
                        fpc = ~fpc, 
                        id = ~dnum)
```

```{r, eval = FALSE}
balancedcluster()
```

### Two-stage

```{r}
library(survey)
data("api")
srs_design <- svydesign(data = apiclus2, 
                        fpc = ~fpc1 + fpc2, 
                        id = ~ dnum + snum)
```

```{r, eval = FALSE}
balancedtwostage()
```
