# Stochastic Frontier

-   From stochastic production frontier models [@aigner1977]

-   "Cost frontier" (cost efficiency) using non-negative cost-inefficiency component is added to the function [@kumbhakar2000]

-   "Profit frontier analysis" : similar to the production frontier

-   Three-component error structure: two inefficiency components from both parties in a transaction can be modeled [@polachek1987]

-   Non- and semi-parametric approaches: [@parmeter2014] [@park2014]

The production frontier model

$$
\begin{aligned}
y_i &= f(x_i; \beta) \times TE_i \times \exp(v_i) \\
&= f(x_i; \beta) \times \exp(-u_i)\times \exp(v_i) 
\end{aligned}
$$

where

-   $y_i$ is the observed output scalar

-   $x_i$ i s matrix of inputs

-   $f(.)$ is the production frontier

-   $\beta$ is the technology parameters

-   $TE_i$ is the technical efficiency = the ratio of observed output to maxim feasible output

    -   $TE_i =1$ is from the firm with the maximum feasible output

    -   $TE_i = \exp(-u_i)$ where $u_i \ge 0$

-   $\exp(v_i)$ is the random stochastic shock

Assume that the production frontier function $f(.)$ is log-linear Cobb-Douglas form, then

$$
\ln y_i = \beta_0 + \sum_n \beta_n \ln x_{ni} + v_i - u_i
$$

where

-   $v_i$ can take a two-sided normally distributed form

-   $u_i$ is the non-negative technological inefficiency component

-   $v_i + u_i$ = compound (composed) error term with a pre-determined distribution

`frontier`

`sfaR`

Try Frontier again after finishing the M&A project

```{r}
# install.packages("frontier")
library(frontier)

# example by the package author
data(front41Data)
head(front41Data)

# Cobb-Douglas production frontier
cb <- sfa(log(output) ~ log(capital) + log(labour),
                    data = front41Data )
summary(cb)
```

-   

```{r}
# install.packages("sfaR")
library(sfaR)
```

# Data Envelopment Analysis

`DEA` Data Envelopment Analysis

-   Non-parametric technique to evaluate the efficiency

-   Types of model

    -   CRS: Constant returns to scale

    -   VRS: Variable returns to scale

```{r}
install.packages("rDEA")
library(rDEA)
```
